# Directed Indirect Connections

```{r setup, include=FALSE}
    require(ggraph)
    require(tidygraph)
    require(igraph)
```

What if the relationship we are studying is not symmetric like a similarity or a co-location but it is asymmetric like an interaction or an exchange (e.g., communication or advice) and we need to represent the resulting network using a directed graph? How do indirect connections work in the directed case? Well, it's just like in the undirected case we saw in @sec-indirect but with a couple of differences. 

- First, we have to *respect the direction of the path* (follow  the arrows). 

- Second, just like asymmetric ties in directed graphs, directed paths can be non-reciprocal. 

That is, there can be a directed path between two nodes $A$ and $B$ where $A$ is the origin node and $B$ is the destination node, but that does *not* mean that there is a directed path going $B$ to $A$, with $B$ as the origin and $A$ as the destination node.

```{r}
#| label: fig-dirpath1
#| fig-cap: "A directed graph showing a directed path of length 3 between nodes A and B."
#| fig-width: 8
#| fig-cap-location: margin

    fr <- c("A", "A", "B", "C", "C", "D", "D", "E", "E")
    to <- c("C", "D", "E", "D", "E", "E", "C", "B", "D")
    dp <- c("y", "n", "n", "n", "y", "n", "n", "y", "n")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr <- tbl_graph(edges = edge.dat, nodes = node.dat) %>%       
      activate(edges) %>% 
      mutate(path = dp)
    cols <- c("n" = "steelblue", "y" = "red")
    wdts <- c("n" = 1, "y" = 2)
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_parallel(aes(color = path, width = path), arrow = arrow(length = unit(5, 'mm')),
                                      sep = unit(8, 'mm'), end_cap = circle(10, 'mm')) 
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() + theme(legend.position = "none") + ylim(-0.6,0.6) + xlim(-1.3, 1.6)
    p
``` 

@fig-dirpath1 shows an example of this situation. The highlighted red path goes from origin node $A$ to destination node $B$. However, note that if we were to start with node $B$ and try to get a message or a package to node $A$, there is no way we could do it by a path of any length in this graph, as long as we are forced to follow the direction of the arrows. 

That means that while node $A$ can **reach** node $A$ by a *directed shortest path* of $l_{AB} = 3$, node $B$ cannot **reach** node $A$ ($l_{BA} = \infty$). The **reachability** relation in directed graphs is therefore asymmetric. 

```{r}
#| label: fig-dirpath2
#| fig-cap: "A directed graph showing a directed path of length 4 between nodes A and B."
#| fig-width: 8
#| fig-cap-location: margin

    fr <- c("A", "A", "B", "C", "C", "D", "D", "E", "E")
    to <- c("C", "D", "E", "D", "E", "E", "C", "B", "D")
    cw <- c("n", "y", "n", "n", "y", "n", "y", "y", "n")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr <- tbl_graph(edges = edge.dat, nodes = node.dat) %>%       
      activate(edges) %>% 
      mutate(path = cw)
    cols <- c("n" = "steelblue", "y" = "red")
    wdts <- c("n" = 1, "y" = 2)
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_parallel(aes(color = path, width = path), arrow = arrow(length = unit(5, 'mm')),
                                      sep = unit(8, 'mm'), end_cap = circle(10, 'mm')) 
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() + theme(legend.position = "none") + ylim(-0.6,0.6) + xlim(-1.3, 1.6)
    p
``` 

This also brings up another difference between undirected and directed graphs when it comes to reachability. In the undirected case, as we will see in a future lesson, the only way that a node cannot be reachable other nodes in the graph is for the graph to be **disconnected**. In the directed case, on the other hand, a graph can be fully connected while at the same time some nodes being incapable of reaching others. 

Note like before, there can be multiple directed paths of a given length connecting two nodes. In the case of $A$ and $B$, there is at least one other path of length 3 with origin node $A$ and destination node $B$. Can you see it? 

```{r}
#| label: fig-dirpath3
#| fig-cap: "A directed graph showing two nodes B and C, mutually reachable via a path of length 3."
#| fig-width: 8
#| fig-cap-location: margin

    fr <- c("A", "A", "B", "C", "C", "D", "D", "E", "E")
    to <- c("C", "D", "E", "D", "E", "E", "C", "B", "D")
    cw <- c("n", "n", "p", "r", "n", "r", "p", "r", "p")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr <- tbl_graph(edges = edge.dat, nodes = node.dat) %>%       
      activate(edges) %>% 
      mutate(path = cw)
    cols <- c("n" = "steelblue", "r" = "red", "p" = "purple")
    wdts <- c("n" = 1, "r" = 2, "p" = 2)
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_parallel(aes(color = path, width = path), arrow = arrow(length = unit(5, 'mm')),
                                      sep = unit(8, 'mm'), end_cap = circle(10, 'mm')) 
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() + theme(legend.position = "none") + ylim(-0.6,0.6) + xlim(-1.3, 1.6)
    p
``` 

Also like we saw in @sec-directed, there can be directed paths of different lengths connecting pairs of nodes. For instance @fig-dirpath2 shows a highlighted red path of length $l_{AB} = 4$ with origin node $A$ and destination node $B$ and edges $(AD, DC, CE, EB)$. In the case of $A$ and $B$, there is at least one other path of length 4 with origin node $A$ and destination node $B$. Can you see it? 

```{r}
#| label: fig-dirpath4
#| fig-cap: "A directed graph showing two nodes B and C, mutually reachable via a paths of different lengths."
#| fig-width: 8
#| fig-cap-location: margin

    fr <- c("A", "A", "B", "C", "C", "D", "D", "E", "E")
    to <- c("C", "D", "E", "D", "E", "E", "C", "B", "D")
    cw <- c("n", "n", "p", "n", "r", "n", "p", "r", "p")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr <- tbl_graph(edges = edge.dat, nodes = node.dat) %>%       
      activate(edges) %>% 
      mutate(path = cw)
    cols <- c("n" = "steelblue", "r" = "red", "p" = "purple")
    wdts <- c("n" = 1, "r" = 2, "p" = 2)
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_parallel(aes(color = path, width = path), arrow = arrow(length = unit(5, 'mm')),
                                      sep = unit(8, 'mm'), end_cap = circle(10, 'mm')) 
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() + theme(legend.position = "none") + ylim(-0.6,0.6) + xlim(-1.3, 1.6)
    p
``` 

## Mutual Reachability
In directed graphs, some pairs of nodes can be **mutually reachable**. That is, there can be a directed path going from one node to the other, and vice versa. @fig-dirpath3 shows an example of this case involving nodes $B$ and $C$. Starting with origin node $B$ we can get to $C$ via the highlighted purple path of length $l_{BC} = 3$ formed by the edges $(BE, ED, DC)$. In the same way, starting with node $C$, we can get to node $B$ via the separate, highlighted red path of length $l_{CB} = 3$ formed by the edges $(CD, DE, EB)$.

For mutual reachability between two nodes to happen in a directed graph, the two paths do not have to be the same length in fact, in the case of nodes $B$ and $C$ node $C$ can reach $B$ via path shorter than length 3. This is shown in @fig-dirpath4), where we can see that $C$ can reach $B$ via the shortest path possible short of being directly connected $(l_{CB} = 2)$, formed by the edges $(CE, EB)$ (highlighted in red). On the other hand, $B$ cannot reach $C$ via a path shorter than 3. 


```{r}
#| label: fig-dircycle
#| fig-cap: "A directed graph showing a directed cycle starting and ending with node C."
#| fig-width: 8
#| fig-cap-location: margin

    fr <- c("A", "A", "B", "C", "C", "D", "D", "E")
    to <- c("C", "D", "E", "D", "E", "E", "C", "D")
    cw <- c("n", "n", "n", "n", "y", "n", "y", "y")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr <- tbl_graph(edges = edge.dat, nodes = node.dat) %>%       
      activate(edges) %>% 
      mutate(path = cw)
    cols <- c("n" = "steelblue", "y" = "red")
    wdts <- c("n" = 1, "y" = 2)
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_parallel(aes(color = path, width = path), arrow = arrow(length = unit(5, 'mm')),
                                      sep = unit(8, 'mm'), end_cap = circle(10, 'mm')) 
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() + theme(legend.position = "none") + ylim(-0.6,0.6) + xlim(-1.3, 1.6)
    p
``` 

## Directed Cycles

Just like in the undirected case, a directed path that begins and ends with the same node is called a **directed cycle**. We refer to different cycles by the number of directed edges they include. For instance a 3-directed cycle is a directed cycle with three directed edges, a 4-directed cycle is a directed cycle with four edges and so forth. For instance, in @fig-dircycle, the directed 3-cycle features node $C$ as both the origin and destination node involving the directed edges $(CE, ED, DC)$ is highlighted in red. 

A  directed graph that does not contain any cycles is a special kind of directed graph (composed of anti-symmetric ties) called a **directed acyclic graph**. 

```{r}
#| label: fig-dirpath5
#| fig-cap: "A directed graph showing two weakly connected nodes A and B."
#| fig-width: 8
#| fig-cap-location: margin

    fr <- c("A", "A", "B", "C", "C", "D", "D", "E")
    to <- c("C", "D", "E", "D", "E", "E", "C", "D")
    cw <- c("y", "n", "y", "n", "y", "n", "n", "n")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr <- tbl_graph(edges = edge.dat, nodes = node.dat) %>%       
      activate(edges) %>% 
      mutate(path = cw)
    cols <- c("n" = "steelblue", "y" = "red")
    wdts <- c("n" = 1, "y" = 2)
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_parallel(aes(color = path, width = path), arrow = arrow(length = unit(5, 'mm')),
                                      sep = unit(8, 'mm'), end_cap = circle(10, 'mm')) 
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() + theme(legend.position = "none") + ylim(-0.6,0.6) + xlim(-1.3, 1.6)
    p
``` 

## Types of Indirect Connections in Directed Graphs
In a directed graph, when pairs of nodes are mutually reachable they are also said to be **strongly connected**. Otherwise if nodes in a directed graph are connected via at least one directed path that only goes in one direction (like nodes *A* and *B* in @fig-dirpath3)), they are said to be **unilaterally connected**. Two nodes are said to be **recursively connected** when they are strongly connected *and* and at least one of the pairs of directed paths going in both directions use the same nodes (like paths $(BE, ED, DC)$ and $(CD, DE, EB)$ connecting nodes *B* and *C* in @fig-dirpath3)). 

Finally, in a directed graph, two nodes are **weakly connected** if we can trace a path from one to the other, but only by ignoring the direction of the arrows! For instance, @fig-dirpath5 shows a directed graph in which nodes *A* and *B* have a weak connection via the $(BE, CE, AC)$ path (highlighted in red), although this is not the only weak connection they share.

Can you trace other weakly connected paths between nodes *A* and *B* in  @fig-dirpath5)?