# Types of Ties and Their Graphs {#sec-ties}

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
  require(ggraph)
  require(tidygraph)
  require(igraph)
  require(kableExtra)
  require(patchwork)
```


## Symmetric Ties and Undirected Graphs 
Nodes and edges are indeed the building blocks of a graph. However, types of relationships that the edges represent can change both how we understand the network conceptually and also what mathematical techniques we can apply to the graph when we compute *graph metrics* (the subject of a future lesson). The basic idea is that when we do network analysis, we want to map our understanding of the nature of the **social relationships** we are studying to the **types of graphs** we use to represent the network formed by the **concatenation** of those relationships. 

```{r}
#| label: fig-undirected
#| fig-cap: "A undirected graph"
#| fig-cap-location: margin
#| fig-height: 8
#| fig-width: 8

  gr <- create_empty(9, directed = FALSE)
  gr <- gr %>% 
    bind_edges(data.frame(from = 1, to = 2:5)) %>% 
    bind_edges(data.frame(from = 2, to = 3:4)) %>% 
    bind_edges(data.frame(from = 3, to = 4:5)) %>% 
    bind_edges(data.frame(from = 4, to = 5:6)) %>% 
    bind_edges(data.frame(from = 6, to = c(7:9))) %>%
    bind_edges(data.frame(from = 7, to = c(8:9))) %>%
    bind_edges(data.frame(from = 8, to = c(9))) %>%
    mutate(name = toupper(letters[1:9]))
  p <- ggraph(gr, layout = 'kk') 
  p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
  p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
  p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
  p <- p + theme_graph()
  p
```

Let us assume that @fig-undirected represents a network of people who spend time together. One way of building this network would be to ask people on your dorm room floor who are the people that they spend some amount of time (e.g., more than an hour a week) hanging out with. By definition the relation "spending time together" lacks any inherent directionality. **Mutuality** (or **reciprocity**) is built in by construction. It would be nonsensical for a person (say *A*) to claim that they spend time with another person (say *B*) and for *B* to say that they do not spend time with *A*. In social network analysis these types of ties are called **symmetric ties**. 

Accordingly, two people being in the same place at the same time (co-location), even if they do not one another, is an example of a symmetric tie. You also have the symmetric tie "being in the same class as" every other student that is also taking your Social Networks seminar this term. Note that, in this sense, all **co-memberships** (e.g., being in the same club or organization or being part of the same family) create symmetric ties among all actors involved (we will revisit this topic when talking about **two-mode networks** in another lesson). If I am a member of your family, you are also my family member; if we are both members of the soccer club, we are considered teammates. Social networks composed of symmetric ties are represented using **undirected graphs** like the one shown in @fig-undirected. When an undirected graph has no **loops** (edges connecting a node to itself), and there is only one edge connecting adjacent vertices to one another (the graph has no **multiedges**), it is called a **simple undirected graph**.

Networks composed of symmetric ties have some interesting properties. If we know that the relationship (**R**) linking two nodes *A* and B is **symmetric**, then only a single edge exists that links them, and it does not matter whether we call this edge *AB* or *BA*. The order does not matter. In this way, we can formally define as **symmetric tie** as one that lacks directionality; if a tie is symmetric, then if we know that A is related to B (the AB edge is part of the edge set of the graph), then we know by necessity that B is related to A. 

Can you think of other examples of symmetric ties? Is friendship, as culturally defined in the contemporary world, a symmetric tie?

## Asymmetric Ties and Directed Graphs 
In contrast to spending time together, being members of the same family, or being in the same place at the same time, some social ties allow for inherent **directionality**. Edges in these graphs are are called **asymmetric ties**. That is, one member of the pair can claim to have a particular type of social relationship with the other, but it is possible (although not necessary) that the other person fails to have the same relationship with the first. 

Helping or **social support** relations, are like this. For instance, you can help someone with their homework, or given them personal advice, but this does not *necessarily* mean that that person will return the favor. They may, or they may not. The point is that, in contrast to symmetric tie, **mutuality** or **reciprocity** is not built in by definition, but must happen as an empirical event in the world. We need to ask the other person to find out (or check their email logs). Can you think of other examples of asymmetric social ties?

```{r}
#| label: fig-directed
#| fig-cap: "A directed graph"
#| fig-cap-location: margin
#| fig-height: 8
#| fig-width: 8

  fr <- c("A", "A", "B", "B", "B", "D", "E", "E", "F", "G", "G")
  to <- c("B", "F", "A", "D", "C", "B", "C", "D", "A", "F", "D")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  p <- ggraph(gr, layout = 'kk')
  p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
              arrow = arrow(length = unit(7, 'mm')),
              end_cap = circle(10, 'mm'), 
              sep = unit(8, 'mm'))
  p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
  p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
  p <- p + theme_graph() 
  p
```

**Reciprocity** is an important concept in social network analysis. Some have said it is perhaps *the most* important concept for understanding human society [@gouldner60], which may be a bit of an exaggeration. Only asymmetric ties may have the property of being non-reciprocal or having more or less reciprocity. If I think you are my friend, I very much hope that you also think you are my friend. That said, sociologists have found that in many natural social settings this is not the case. Sometimes people think they are friends with others, but those other people disagree [@carley_krackhardt96]. For this reason, sociologists typically ask: if I do you a favor, would you do me a favor in the future? Additionally, sociologists often ask: if I treat you with respect, will you also treat me with respect? If I text you, will you text me back? If this is true, we have a level of reciprocity in our relationship. 

For some ties, such advice or support, or friendship relations, reciprocity is all or none; it either exists or it does not. For instance, the friendship offer you extend to someone may be reciprocated (or not). In the same way, you can like someone and they may like you back (or not), like the notes you passed around in middle school. For other ties, such as communication ties (e.g., those defined by the amount of texting, or calling), reciprocity is a matter of degree, there may be more or less. For instance, you can text someone 10 times a day, but they may text you back only half of those instances. In all cases, reciprocity is at a maximum when the content of the relationship is equally exchanged between actors. 

Can you think of relationships in your life characterized by more or less reciprocity?

Just like symmetric ties are represented using a particular type of graph (namely, an **undirected graph**), social networks composed of asymmetric ties are best represented by a type of graph called a **directed graph**.^[Directed graphs are also called **digraphs**] @fig-directed shows the point and line diagram picture of a directed graph. What were simple lines for in the undirected graph shown in @fig-undirected have been replaced with arrows indicating directionality. A node **sends** a relationship to the node that the arrow points to. Up to two directed arrows may link nodes going both ways. 

When an undirected graph has no **loops** (edges connecting a node to itself), and there is only one edge connecting adjacent vertices to one another (the graph has no **multiedges**) like @fig-directed, it is called a **simple directed graph**. If @fig-directed were an **advice network** [@cross01], on the other hand, we could say that *H* seeks advice from *D*, but *D* does not seek advice from *H*. This may be because *D* is higher in the office hierarchy or is more experienced than *H*, in which case lack of reciprocity may be indicative of an authority relationship between the two nodes. 

In a directed graph, for every edge, there is a **source node** and a **destination node**. So in the case of "A helps B" the source node is *A* and the destination node is *B*. In the case of "B helps A" the source node is *B* and the destination node is *A*. This means that in a directed graph, in contrast to a undirected one, *the order in which you list the nodes when you name the edges matters*. Thus, the edge *AB* is a different one from the edge *BA*. For instance, the first one may exist but the second one may not exist.^[Edges in a direted graph are sometimes also called **arcs**.] 

One must always be careful when examining a directed network to make sure one properly understands the direction of the underlying social relationships!


```{r}
#| label: fig-symmdirected
#| fig-cap: "A symmetric directed graph"
#| fig-cap-location: margin
#| fig-height: 8
#| fig-width: 8

  fr <- c("A", "B", "D", "A", "B", "C", "E", "C", "F", "A", "G", "D", "C", "D", "E", "G")
  to <- c("B", "A", "A", "D", "C", "B", "C", "E", "A", "F", "D", "G", "D", "C", "G", "E")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  p <- ggraph(gr, layout = 'kk')
  p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
              arrow = arrow(length = unit(7, 'mm')),
              end_cap = circle(10, 'mm'), 
              sep = unit(8, 'mm'))
  p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
  p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
  p <- p + theme_graph() 
  p
```

There are also some special types of directed graphs. For instance, if in directed graph, every pair of adjacent nodes is connected via a bi-directional, reciprocal relationship, as in @fig-symmdirected, then it is called as **symmetric directed graph**.  

### Node Neighborhoods in Directed Graphs
Just like in undirected (simple) graphs, each node in a directed graph has a **node neighborhood**. However, because now each node can be the **source** or **destination** for a asymmetric edges, this means that we have to differentiate the neighborhood of a node depending on whether the node is the **sender** or the **recipient** of a given link. 

So, we say that a node *j* is an an **in-neighbor** of a node *i* if there is a directed link with *j* as the source and *i* as the destination node. For instance, in @fig-directed, *E* is an in-neighbor of *C*, because there's a asymmetric edge with *E* as the source and *C* as the destination. 

In the same way, we say that a node *i* is an **out-neighbor** of a node *j* if there is a directed link with *i* as the source and *j* as the destination. For instance, in @fig-directed, *F* is an in-neighbor of *G*, because there's a asymmetric edge with *G* as the source and *F* as the destination

For each node, the full set of in-neighbors forms the **in-neighborhood** of that node. This is written $N^{in}(v)$, where $v$ is the label corresponding to the node. For instance, in @fig-directed, the node set $N^{in}(D) = \{B, E, G\}$ is the in-neighborhood of node *D*. 

In the same way, the full set of in-neighbors defines the **out-neighborhood** of that node. This is written $N^{out}(v)$, where $v$ is the label corresponding to the node. For instance, in @fig-directed, the node set $N^{out}(B) = \{A, C, D\}$ is the out-neighborhood of node *B*. 

Note that typically, the set of in-neighbors and out-neighbors of a given node will not be exactly the same, and sometimes the two sets will be completely **disjoint** (they won't share any members). 

Nodes will only show up in both the in and out-neighborhood set when there are **reciprocal** or mutual ties between the nodes. For instance, in @fig-directed, the out-neighborhood of node *F* is $\{A\}$ and the in-neighborhood is $\{A, G\}$. Here node *A* shows up in both the in and out-neighborhood sets because *A* has a reciprocal tie with *F*.

### Node Degree in Directed Graphs
Because in a directed graph, each node has two distinct set of neighbors, we can compute two versions of degree for the same node. 

- in a directed graph, for any node *i*, we can count the number of edges that have a given node $v$ as their destination node. This is also the cardinality of the in-neighborhood set of that node. This is called a node's **indegree** and it is written $k^{in}_i$, where *i* is the label corresponding to that node.

- Additionally, in a directed graph, for any node *i*, we can count the number of edges that have a given node $i$ as their source node. This is also the cardinality of the out-neighborhood set of that node. This is called that node's **outdegree** and it is written as $k^{out}_i$, , where *i* is the label corresponding to that node.

For instance, in @fig-directed, $k^{out}_B = 3$ and $k^{in}_B = 2$. Node *B* has *three* **outgoing ties** (from nodes *A*, *C*, and *D*) and three **incoming ties** (from nodes *A* and *D*).

Can you calculate what the indegree and outdegree of node *D* in @fig-directed is?

The graph theoretic ideas of indegree and outdegree have clear sociological interpretations. In a social network, for instance, a node having a large outdegree could indicate a *sociable person* (a person that likes to connect with others), while having a large indegree can indicate a *popular person* (e.g., a person lots of other people want to be friends with).^[In a later lesson we will see how to use a directed graph's **asymmetric adjacency matrix** to readily compute the outdegree and indegree in real social networks.]

This means that in a directed graph, there will typically be three types of (non-isolate) nodes [@Harary65]: 

- First, there will be nodes that receive ties but don't send them. These are called **receivers** (like node *C* in @fig-directed). For receiver nodes $k_{in} > 0$ and $k_{out} = 0$.
- Second, there will be nodes that receive ties and also send out ties. These are called **carriers** (like nodes *A* and *B* in @fig-directed. For carrier nodes, $k_{in} > 0$ and $k_{out} > 0$.
- Finally, there will be nodes that send ties but don't receive them. These are called **transmitters** (like nodes *E* and *G* in @fig-directed). For transmitter nodes, $k_{in} = 0$ and $k_{out} > 0$. 

## Anti-Symmetric Ties and Oriented Graphs {#sec-oriented}

There is a particular type of directed relationship that has the property of only going in one direction. These are called **anti-symmetric ties**. Like asymmetric ties, anti-symmetric ties have a directionality (and thus source and destination nodes), but reciprocity is forbidden by definition. That means that if *A* is anti-symmetrically connected to *B*, then *B* *cannot* send the *same type of tie* back to *A* (although *B* may be connected, and typically is, to *A* via some other type of tie in a different network). 

A common example of anti-symmetric ties in political sociology are *patron-client ties* [@martin09]. Patrons can have many clients, but it is impossible for client of a patron to also be a patron to the same person. Other types of anti-symmetric ties include **authority relations** such as hierarchical ties at work, or **cross-generational relations** like kinship ties in families. 

In the first (authority relations) case, your boss is your boss, while you cannot be your boss’ boss. The same goes for armies and other **command and control structures**, *giving orders to* is an anti-symmetric relation. An officer who gives orders to another officer (and thus commands them) *cannot* by definition also receive orders from them. Thus, @fig-tree could be a network in which the anti-symmetric links are directed "gives orders to" (in an army or an office) relations, where the **source node** directs commands toward the **destination node**. So *A* is the top boss and commands *B*, *C*, and *D*. Node *B*, in their turn, gives orders to *E*, who is at the lowest level of the command and control structure, not commanding anybody in turn. 

```{r}
#| label: fig-oriented
#| fig-cap: "An oriented directed graph"
#| fig-cap-location: margin
#| fig-height: 8
#| fig-width: 8

  fr <- c("A", "D", "C", "C", "A", "D", "D", "G", "C", "B", "G")
  to <- c("B", "A", "F", "E", "F", "G", "C", "E", "A", "F", "A")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  p <- ggraph(gr, layout = 'kk')
  p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
              arrow = arrow(length = unit(7, 'mm')),
              end_cap = circle(10, 'mm'), 
              sep = unit(8, 'mm'))
  p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
  p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
  p <- p + theme_graph() 
  p
```

In the second (cross-generational) case, your parents are your parents (but you can only be a son or daughter to your parents), and your grandparents are their parents, and so forth. "Being the parent of" thus counts as an anti-symmetric relation as we define here; it only goes way (from parents to children) but it cannot come back from children towards parents. Teacher-student, coach-athlete, buyer-seller, parent-child are all examples of anti-symmetric relationships that can be depicted as tree graphs.  


```{r}
#| label: fig-oriented-tree
#| fig-cap: "An oriented directed graph drawn as a tree."
#| fig-cap-location: margin
#| fig-height: 8
#| fig-width: 8

  fr <- c("A", "D", "C", "C", "A", "D", "D", "G", "C", "B", "G")
  to <- c("B", "A", "F", "E", "F", "G", "C", "E", "A", "F", "A")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  p <- ggraph(gr, layout = 'tree')
  p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
              arrow = arrow(length = unit(7, 'mm')),
              end_cap = circle(10, 'mm'), 
              sep = unit(8, 'mm'))
  p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
  p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
  p <- p + theme_graph() 
  p
```

What happens if we build a network composed of pure anti-symmetric ties? The resulting graph is a special type of directed graph called an **oriented graph**, like the one in @fig-oriented. An oriented graph is kind of the opposite of the kind of symmetric directed graph depicted in @fig-symmdirected. Whereas in @fig-symmdirected every connected dyad is linked by a bi-directional reciprocal link, in the oriented graph shown in @fig-oriented, every connected dyad is linked *only* by a one-way non-reciprocal link, the link that would turn the asymmetrically connected dyads in @fig-oriented into a reciprocal dyad is forbidden by the nature of the tie.

```{r}
#| label: fig-tree
#| fig-cap: "A tree graph with thirteen nodes."
#| fig-cap-location: margin
#| fig-width: 10
#| fig-height: 10

  gr <- create_tree(n = 13, children = 3) 
  gr <- mutate(gr, name = toupper(letters[1:13])) 
  p <- ggraph(gr, layout = 'tree')
  p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25,
              arrow = arrow(length = unit(6, 'mm')),
              end_cap = circle(10, 'mm'))
  p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
  p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
  p <- p + theme_graph() 
  p
```


As we will see in @sec-indirect, when considering connections between people that involve intermediaries, there is a special pattern called a **cycle**. This involves a node sending a directed tie to another node, which sends a directed tie to a third node. What makes it a cycle is that third node then sends a directed tie to the first node we began with. Recall from @sec-dirtriads, that the three-person subgraph *AAA(C)* shown in @fig-AAAC forms a cycle. For instance, in @fig-oriented, the sequence of directed edges $ED, DG, GE$ forms a cycle starting and ending in node $E$. Can you spot any other cycles in @fig-oriented?

One feature of a network composed of only anti-symmetric relations is that its corresponding graph can always be drawn from top to bottom, starting (at the top) with the node that only sends but does receive any ties and ending (at the bottom) with nodes that only receive, but do not send, ties. This is called a **tree layout**. For instance, @fig-oriented tree shows the same graph as in @fig-oriented, but laid out as a tree. 

However, a **tree graph** is not the same as an oriented graph drawn as a tree. Instead, a tree graph is a special kind of oriented graph, oberying a few restrictions:

1. Every node is the source of only *one* directed edge and/or the destination of only *one* directed edge. 

1. A node can only send a tie to a node that does not receive ties from any other node. 

1. If a node sends an anti-symmetric link to a second node, and that node sends an anti-symmetric link to a third node, this last node cannot send an anti-symmetric link back to the first node. 

An oriented graph obeying these three restrictions, and thus counting as a tree graph, is shown in @fig-tree. As per conditions (1) and (2), the indegrees of each node in the graph are restricted to be either zero or one. Transmitter nodes have indegree zero but oudegree equal to one or more, like $A$ in @fig-tree ($k_{in}(A) = 0, k_{out}(A) = 3$). Receiver nodes have indegree one but outdegree zero like $\{E, F, G, H, I, J, K, L, M\}$ in @fig-tree. Finally carrier nodes indegree equal to one like and outdegree of one or more like $\{B, C, D\}$ in @fig-tree.

As noted, your family tree is an example of a tree graph composed of anti-symmetric kin ties going from people in an older generation to those in a younger generation. For instance, *A* could be your grandmother, and *B*, *C*, and *D* could be her three daughters. If *B* was you mom, then you could be *E* (along with your siblings *F* and *G*) and your cousins *H*, *I*, *J*, *K*, *L*, *M*. 

Given condition (3), a key property of tree graphs composed of anti-symmetric ties is that *they can't have any cycles*. No node that sends a tie to another node can receive a tie back from a third node their neighbor points to! This is the reason why tree graphs are said to to be "acyclic" (they lack cycles). 

In this case, tree graphs are a special case of a directed graph without cycles. These special type of directed graphs are also called **directed acyclic graphs** which is sometimes abbreviated as "DAG." DAGs have applications in certain fields of statistics, causal analysis, and machine learning and have been made famous by [Judea Pearl](http://bayes.cs.ucla.edu/jp_home.html), a Professor of Computer science right here at UCLA.


## References {.unnumbered}