---
title: "Homework V: Affiliation Networks"
sidebar: false
navbar: false
toc: false
---
```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

```{r}
#| label: fig-grex1
#| fig-cap: "A bipartite graph."
#| fig-cap-location: bottom
#| fig-height: 12
#| fig-width: 12
    set.seed(325689)
    a <- sample(8:10, 1)
    b <- sample(12:13, 1) 
    c <- sample(35:40, 1) 
    gr <- sample_bipartite_gnm(n1 = a, n2 = b, m = c)
    V(gr)$type <- bipartite_mapping(gr)$type
    V(gr)$color <- ifelse(V(gr)$type, "red", "tan2")
    V(gr)$shape <- ifelse(V(gr)$type, "circle", "square")
    V(gr)$size <- ifelse(V(gr)$type, 26, 24)
    V(gr)$name <- c(LETTERS[1:a], paste("E", 1:b, sep = ""))
    E(gr)$color <- "steelblue"
    p <- ggraph(gr, layout = 'sugiyama') 
    p <- p + geom_edge_link(color = "steelblue", edge_width = 1.3) 
    p <- p + geom_node_point(size = V(gr)$size, color = V(gr)$color, shape = V(gr)$shape)
    p <- p + geom_node_text(aes(label = V(gr)$name), color = "white", 
                            size = c(rep(10, a), rep(8, b)), 
                            fontface = "bold")
    p <- p + theme_graph() 
    p
```

@fig-grex1 shows a **bipartite graph** corresponding to the attendance of different *people* (the tan square nodes) at different *events* (the red circle nodes).

## From Bipartite Graph to Affiliation Matrix

- Write down the **affiliation matrix** corresponding to the **bipartite graph** shown in @fig-grex1. 

```{r}
    A <- as.matrix(as_biadjacency_matrix(gr))
    kbl(A, format = "html", padding = 1, align = "c") %>% 
        kable_classic(lightable_options = "striped") %>% 
        row_spec(1:a, color = "black", extra_css = "border-bottom: 1px solid;") %>% 
        column_spec(1:b+1, color = "black", border_right = TRUE, border_left = TRUE) 
```

## Bipartite Node Neighborhoods

1. Write down the neighborhood of node *G*
<br>
`r V(gr)$name[neighbors(gr, "G")]`
<br>

1. Write down the neighborhood of event *E11*
<br>
`r V(gr)$name[neighbors(gr, "E11")]`
<br>

1. What is the intersection of the neighborhoods nodes *A* and *E*?
<br>
`r intersect(V(gr)$name[neighbors(gr, "A")], V(gr)$name[neighbors(gr, "E")])`
<br>

1. What is the intersection of the neighborhoods of events *E11* and *E6*?
<br>
`r intersect(V(gr)$name[neighbors(gr, "E11")], V(gr)$name[neighbors(gr, "E6")])`
<br>

1. What is the union of the neighborhoods nodes *B* and *C*?
<br>
`r union(V(gr)$name[neighbors(gr, "B")], V(gr)$name[neighbors(gr, "C")])`
<br>

1. What is the union of the neighborhoods of events *E9* and *E5*?
<br>
`r union(V(gr)$name[neighbors(gr, "E9")], V(gr)$name[neighbors(gr, "E5")])`
<br>

## Bipartite Degrees

1. Write down the person degree sequence
<br>
<br>
`r kable(t(sort(degree(gr)[1:a], decreasing = TRUE)))`
<br>
<br>

1. Write down the event degree sequence
<br>
<br>
```{r} 
    c <- a + 1
    d <- a + b
    kable(t(sort(degree(gr)[c:d], decreasing = TRUE)))
```
<br>
<br>

1. What is the bipartite graph's sum of degrees?
<br>
<br>
`r sum(degree(gr))/2`
<br>
<br>


## Bipartite Graph Metrics

1. Which person(s) attend(s) the most events?
<br>
`r names(which(rowSums(A) == max(rowSums(A))))`
<br>

1. Which person(s) attend(s) the least events?
<br>
`r names(which(rowSums(A) == min(rowSums(A))))`
<br>

1. What is the average number of events attended by people?
<br>
`r round(mean(rowSums(A)), 3)`
<br>

1. Which is(are) the largest event(s)?
<br>
`r names(which(colSums(A) == max(colSums(A))))`
<br>

1. Which is(are) the smallest event(s)?
<br>
`r names(which(colSums(A) == min(colSums(A))))`
<br>

1. What is the average event size?
<br>
`r round(mean(colSums(A)), 3)`
<br>

- Write down the **transpose** of the **affiliation matrix** corresponding to the bipartite graph.

```{r}
    At <- t(as.matrix(as_biadjacency_matrix(gr)))
    kbl(At, format = "html", padding = 1, align = "c") %>% 
        kable_classic(lightable_options = "striped") %>% 
        row_spec(1:b, color = "black", extra_css = "border-bottom: 1px solid;") %>% 
        column_spec(1:a+1, color = "black", border_right = TRUE, border_left = TRUE) 
```

Using the information in the previous tables and what you know about **matrix multiplication**:

- Write down the entries corresponding to the **co-membership matrix** for the different people.

```{r}
    c.m <- A %*% t(A)
    rownames(c.m) <- LETTERS[1:a]
    colnames(c.m) <- LETTERS[1:a]
    kbl(c.m, format = "html", padding = 1, align = "c") %>% 
        kable_classic(lightable_options = "striped") %>% 
        row_spec(1:a, color = "black", extra_css = "border-bottom: 1px solid;") %>% 
        column_spec(1:a+1, color = "black", border_right = TRUE, border_left = TRUE)
```

- Write down the entries corresponding to the **group overlap matrix** for the different events.

```{r}
    g.o <- t(A) %*% A 
    rownames(g.o) <- paste("E", 1:b, sep = "")
    colnames(g.o) <- paste("E", 1:b, sep = "")
    kbl(g.o, format = "html", padding = 1, align = "c") %>% 
        kable_classic(lightable_options = "striped") %>% 
        row_spec(1:b, color = "black", extra_css = "border-bottom: 1px solid;") %>% 
        column_spec(1:b+1, color = "black", border_right = TRUE, border_left = TRUE)
```
