---
title: "Homework VII: Ego-Network Metrics"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(colorspace)
    require(patchwork)
```

```{r}
    set.seed(389)
    s <- 10
    c1 <- sample(rainbow_hcl(4), s, replace = TRUE)
    c2 <- sample(rainbow_hcl(5), s, replace = TRUE)
    c3 <- sample(rainbow_hcl(4), s, replace = TRUE)
    c4 <- sample(rainbow_hcl(4), s, replace = TRUE)

```

## Size, Effective Size, and Clustering


```{r}
#| label: fig-ego
#| fig-cap: "Ego Network ties in a hypothetical ego-network."
#| fig-cap-location: margin  
#| fig-subcap:
#|   - "Ego to alter ties."
#|   - "Alter to alter ties."
#| layout-ncol: 2 
#| fig-height: 8
#| fig-width: 8

    gr <- create_star(s + 1, directed = FALSE) %>% 
        mutate(name = c("Ego", toupper(letters[1:s])))
    p <- ggraph(gr, layout = 'star') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.5) 
    p <- p + geom_node_point(aes(x = x, y = y), size = c(30, rep(22, s)), color = c("red", rep("tan2", s))) 
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p1 <- p + theme_graph()
    p1
    
     set.seed(179)
    w <- 0
    while (w == 0) {
      d <- sample(0:s-4, s, replace = TRUE)
      w <- as.numeric(is.graphical.degree.sequence(d))
      }      
    gr <- play_degree(d, method = "simple.no.multiple")
    gr <- mutate(gr, name = toupper(letters[1:length(d)]))
    p <- ggraph(gr, layout = 'circle') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p2 <- p + theme_graph() 
    p2
```


- Consider @fig-ego-1. This shows the **ego-to-alter** ties in a hypothetical ego-network.

- Consider @fig-ego-2. This shows the **alter-to-alter** ties in the hypothetical ego-network shown in @fig-ego-1.

1. What is the **size** of ego's network?
<br>
<br>
<br>
<br>
<br>
<br>

1. What is the **effective size** of ego's network?
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

1. What is ego's **clustering coefficient**?
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

## Composition, Diversity, and Homophily


```{r}
#| label: fig-altersdem
#| fig-cap: "Socio-demographic Characteristics of Alters"
#| fig-subcap:
#|   - "Gender Identification of Alters."
#|   - "Ethnoracial identification of alters."
#|   - "Role relations of alters with Ego."
#|   - "Employment status of alters."
#| layout-ncol: 2 
#| fig-cap-location: margin  
#| fig-height: 10
#| fig-width: 10

    gr <- create_empty(s, directed = FALSE) %>% 
        mutate(name = toupper(letters[1:s]))
    p <- ggraph(gr, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c1), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p1 <- p + scale_color_discrete(labels = c("Women", "Men", "Genderfluid", "Non-binary"))
    p1

    gr <- create_empty(s) %>% 
        mutate(name = toupper(letters[1:s]))
    p <- ggraph(gr, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c2), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p2 <- p + scale_color_discrete(labels = c("Asian", "Black", "White", "Native", "Hispanic"))
    p2

    gr <- create_empty(s, directed = FALSE) %>% 
        mutate(name = toupper(letters[1:s]))
    cl <- sample(rainbow_hcl(4), 11, replace = TRUE)
    p <- ggraph(gr, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c3), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p3 <- p + scale_color_discrete(labels = c("Neighbor", "Kin", "Acquaintance", "Friend"))
    p3

    gr <- create_empty(s, directed = FALSE) %>% 
        mutate(name = toupper(letters[1:s]))
    cl <- sample(rainbow_hcl(4), 11, replace = TRUE)
    p <- ggraph(gr, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c4), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p4 <- p + scale_color_discrete(labels = c("Employed Outside the Home", "Retired", "Student", "Full-Time Parent"))
    p4
```
@fig-altersdem-1-@fig-altersdem-4 show some sociodemographic and role-relational characteristics of alters in ego's network. Ego is a *UCLA student who identifies as Asian and as a Woman*. Using this information:

1. Compute Ego's EI **homophily** index with respect to gender identity.
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute Ego's EI **homophily** index with respect to Ethnoracial identification.
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute the **proportion** of neighbors in Ego's network.
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute the **proportion** of people who identify as Hispanic in Ego's network.
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute the **proportion** of fellow students in Ego's network.
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute the **diversity** of Ego's Network with respect to ethnoracial identity.
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute the **diversity** of Ego's Network with respect to employment status. 
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute the **diversity** of Ego's Network with respect to gender identity.
<br>
<br>
<br>
<br>
<br>
<br>

1. Does ego have a stronger **same-group preference** with respect to race than with respect to gender? (Yes/No, show your reasoning)
<br>
<br>
<br>
<br>
<br>
<br>

1. Is the **diversity** of ego's network larger when it comes to race than when it comes to employment status  (Yes/No, show your reasoning)
<br>
<br>
<br>
<br>
<br>
<br>

1. What is the **proportion** of ego's network composed of friends who are also Genderfluid?
<br>
<br>
<br>
<br>
<br>
<br>

1. Compute the **proportion** of people who identify as either Men or Women in ego's network. 
<br>
<br>
<br>
<br>
<br>
<br>
