---
title: "Homework VI: Centrality"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(kableExtra)
    require(tidygraph)
    require(igraph)
```

```{r}
#| label: fig-grex1
#| fig-cap: "An undirected graph."
#| fig-cap-location: margin   
#| fig-height: 12
#| fig-width: 12
    set.seed(45279)
    w <- 0
    n <- 10
    while (w == 0) {
      d <- sample(1:n, n, replace = TRUE)
      w <- as.numeric(is_graphical(d, allowed.edge.types = "simple"))
      }    
    gr <- sample_degseq(d, method = "configuration")
    gr <- simplify(gr)
    V(gr)$name <- LETTERS[1:length(d)]
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 14, color = "white")
    p <- p + theme_graph() 
    p
```

## Degree Centrality in Undirected Graphs

Rank all the nodes in the graph shown in @fig-grex1 by their **degree centrality**.
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
    a <- matrix(" ", nrow = n, ncol = 2)
    colnames(a) <- c("Node", "Degree")
    kable(a, format = "html", align = "c", full_width = F)  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:2, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()

```

```{r}
#| label: fig-grex2
#| fig-cap: "A directed graph."
#| fig-cap-location: margin   
#| fig-height: 10
#| fig-width: 12
    set.seed(234)
    w <- 0
    n <- 11
    a <- n - 4
    b <- n - 5
    while (w == 0) {
      d1 <- sample(1:a, n, replace = TRUE)
      d2 <- sample(1:b, n, replace = TRUE)
      w <- as.numeric(is_graphical(d1, d2, allowed.edge.types = "simple"))
        }    
    gr2 <- sample_degseq(d1, d2, method = "configuration")
    gr2 <- simplify(gr2)
    V(gr2)$name <- LETTERS[1:length(d1)]
    p <- ggraph(gr2, layout = 'fr') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(7, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))    
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() 
    p
```

Rank all the nodes in the graph shown in @fig-grex2 by their **outdegree centrality** and **indegree centrality**.
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
    a <- matrix(" ", nrow = n, ncol = 4)
    colnames(a) <- c("Node", "Outdegree", "Node", "Indegree")
    kable(a, format = "html", align = "c", full_width = F)  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:4, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()

```

## Closeness Centrality in Directed Graphs
- In the matrix below, write down the cell entries **geodesic distance matrix** corresponding to the graph shown in @fig-grex2:

```{r}
     b <- matrix(" ", nrow = n, ncol = n)
     rownames(b) <- LETTERS[1:length(d1)]
     colnames(b) <- LETTERS[1:length(d1)]
     diag(b) <- "----"
     kbl(b, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```

Using the information in the above matrix, rank all the nodes in the graph shown in @fig-grex2 by their **closeness centrality**.

<br>
<br>
<br>
<br>
<br>
<br>

```{r}
    a <- matrix(" ", nrow = n, ncol = 2)
    colnames(a) <- c("Node", "Closeness")
    kable(a, format = "html", align = "c", full_width = F)  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:2, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```

## Relative Betweenness Centrality

Go back to @fig-grex1. 

What is the **relative betweenness centrality** of node *C* with respect to nodes *A* and *I*? (show your work)
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


## Degree Centralization
Compute the **degree centralization** of the graph shown in @fig-grex1.
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>