---
title: "Homework VII: Ego-Network Metrics"
fig-cap-location: bottom
tbl-cap-location: bottom
---
```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(colorspace)
    require(patchwork)
```

```{r}
    set.seed(49374857)
    s <- 10
    c1 <- sample(rainbow_hcl(4), s, replace = TRUE)
    c2 <- sample(rainbow_hcl(5), s, replace = TRUE)
    c3 <- sample(rainbow_hcl(4), s, replace = TRUE)
    c4 <- sample(rainbow_hcl(4), s, replace = TRUE)
    
    c1_fac <- factor(c1, levels = sort(rainbow_hcl(4)))
    c2_fac <- factor(c2, levels = sort(rainbow_hcl(5)))
    c3_fac <- factor(c3, levels = sort(rainbow_hcl(4)))
    c4_fac <- factor(c4, levels = sort(rainbow_hcl(4)))
    
    I_gender <- sum(as.numeric(c1_fac) == 2, na.rm = TRUE)
    E_gender <- s - I_gender
    EI_gender <- (E_gender - I_gender) / s
    
    I_race <- sum(as.numeric(c2_fac) == 2, na.rm = TRUE)
    E_race <- s - I_race
    EI_race <- (E_race - I_race) / s
    
    prop_neighbors <- sum(as.numeric(c3_fac) == 1, na.rm = TRUE) / s
    prop_hispanic <- sum(as.numeric(c2_fac) == 5, na.rm = TRUE) / s
    prop_high_school <- sum(as.numeric(c4_fac) == 2, na.rm = TRUE) / s
    
    H_race <- round(1 - sum((table(c2_fac)/s)^2), 2)
    H_edu <- round(1 - sum((table(c4_fac)/s)^2), 2)
    H_gender <- round(1 - sum((table(c1_fac)/s)^2), 2)
    
    pref_race_vs_gender <- if (EI_race < EI_gender) "Yes" else "No"
    div_race_vs_edu <- if (H_race > H_edu) "Yes" else "No"
    
    prop_friend_women <- sum(as.numeric(c3_fac) == 4 & as.numeric(c1_fac) == 1, na.rm = TRUE) / s
    prop_men_women <- sum(as.numeric(c1_fac) %in% c(1, 2), na.rm = TRUE) / s

```

## Size, Effective Size, and Clustering


```{r}
#| label: fig-ego
#| fig-cap: "Ego Network ties in a hypothetical ego-network."
#| fig-cap-location: bottom
#| fig-subcap:
#|   - "Ego to alter ties."
#|   - "Alter to alter ties."
#| layout-ncol: 2 
#| fig-height: 8
#| fig-width: 8

    gr <- create_star(s + 1, directed = FALSE) %>% 
        mutate(name = c("Ego", LETTERS[1:s]))
    p <- ggraph(gr, layout = 'star') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.5) 
    p <- p + geom_node_point(aes(x = x, y = y), size = c(32, rep(22, s)), color = c("red", rep("tan2", s))) 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p1 <- p + theme_graph()
    p1
    
     set.seed(179)
    w <- 0
    while (w == 0) {
      d <- sample(0:s-4, s, replace = TRUE)
      w <- as.numeric(is.graphical.degree.sequence(d))
      }      
    gr <- play_degree(d, method = "simple.no.multiple")
    gr <- mutate(gr, name = LETTERS[1:length(d)])
    p <- ggraph(gr, layout = 'circle') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p2 <- p + theme_graph() 
    p2
```


- Consider @fig-ego-1. This shows the **ego-to-alter** ties in a hypothetical ego-network.

- Consider @fig-ego-2. This shows the **alter-to-alter** ties in the hypothetical ego-network shown in @fig-ego-1.

1. What is the **size** of ego's network?

ANSWER: `r s`

2. What is the **effective size** of ego's network?

ANSWER: `r s - mean(degree(gr))` 

3. What is ego's **clustering coefficient**?

ANSWER: `r round(edge_density(gr), 3)`

## Composition, Diversity, and Homophily


```{r}
#| label: fig-altersdem
#| fig-cap: "Socio-demographic Characteristics of Alters"
#| fig-subcap:
#|   - "Gender Identification of Alters."
#|   - "Ethnoracial identification of alters."
#|   - "Role relations of alters with Ego."
#|   - "Educational attainment of alters."
#| layout-ncol: 2 
#| fig-cap-location: bottom
#| fig-height: 10
#| fig-width: 10
    gr1 <- create_empty(s, directed = FALSE) %>% 
        mutate(name = LETTERS[1:s])
    p <- ggraph(gr1, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c1), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p1 <- p + scale_color_discrete(labels = c("Women", "Men", "Genderfluid", "Non-binary"))
    p1

    gr2 <- create_empty(s) %>% 
        mutate(name = LETTERS[1:s])
    p <- ggraph(gr2, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c2), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p2 <- p + scale_color_discrete(labels = c("Asian", "Black", "White", "Native", "Hispanic"))
    p2

    gr3 <- create_empty(s, directed = FALSE) %>% 
        mutate(name = LETTERS[1:s])
    cl <- sample(rainbow_hcl(4), 11, replace = TRUE)
    p <- ggraph(gr3, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c3), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p3 <- p + scale_color_discrete(labels = c("Neighbor", "Kin", "Acquaintance", "Friend"))
    p3

    gr4 <- create_empty(s, directed = FALSE) %>% 
        mutate(name = LETTERS[1:s])
    cl <- sample(rainbow_hcl(4), 11, replace = TRUE)
    p <- ggraph(gr4, layout = 'circle') 
    p <- p + geom_node_point(aes(x = x, y = y, color = c4), size = 24)
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
    p <- p + theme_graph() + theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 18))
    p <- p + guides(color = guide_legend(override.aes = list(size = 14))) 
    p4 <- p + scale_color_discrete(labels = c("College", "High School", "Associate Degree", "Professional Degree"))
    p4
```
@fig-altersdem-1-@fig-altersdem-4 show some sociodemographic and role-relational characteristics of alters in ego's network. Ego is a *UCLA student who identifies as Black and as a Man*. Using this information:

1. Compute Ego's EI **homophily** index with respect to gender identity.

ANSWER: `r EI_gender`

2. Compute Ego's EI **homophily** index with respect to Ethnoracial identification.

ANSWER: `r EI_race`


3. Compute the **proportion** of neighbors in Ego's network.

ANSWER: `r prop_neighbors`


4. Compute the **proportion** of people who identify as Hispanic in Ego's network.

ANSWER: `r prop_hispanic`

5. Compute the **proportion** of people with a high school degree in Ego's network.

ANSWER: `r prop_high_school`


6. Compute the **diversity** of Ego's Network with respect to ethnoracial identity.

ANSWER: `r H_race`

7. Compute the **diversity** of Ego's Network with respect to education. 

ANSWER: `r H_edu`


8. Compute the **diversity** of Ego's Network with respect to gender identity.

ANSWER: `r H_gender`


9. Does ego have a stronger **same-group preference** with respect to race than with respect to gender? (Yes/No, show your reasoning)

ANSWER: `r pref_race_vs_gender` (Gender EI = `r EI_gender`, versus Race EI = `r EI_race`)


1. Is the **diversity** of ego's network larger when it comes to race than when it comes to education  (Yes/No, show your reasoning)

ANSWER: `r div_race_vs_edu` (Race H = `r H_race`, versus Education H = `r H_edu`)


1. What is the **proportion** of ego's network composed of friends who are also Women?

ANSWER: `r prop_friend_women`


1. Compute the **proportion** of people who identify as either Men or Women in ego's network. 

ANSWER: `r prop_men_women`

