---
title: "Homework IX: Subgroups, Graph Connectivity, and Structural Similarity"
sidebar: false
navbar: false
toc: false
fig-cap-location: bottom
tbl-cap-location: bottom
---
```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    library(ggraph)
    library(tidygraph)
    library(igraph)
    library(kableExtra)
```

## Cliques and n-Cliques

```{r}
#| label: fig-grex1
#| fig-width: 12
#| fig-height: 12
#| fig-cap: "An undirected graph."
#| fig-cap-location: bottom
    set.seed(34758935)
    n <- 13
    names <- LETTERS[sample(1:n)]
    gr1 <- create_complete(4) |> 
        mutate(name = names[1:4])
    gr2 <- create_complete(4) |>
        mutate(name = names[5:8])
    gr3 <- create_complete(5) |>
        mutate(name = names[9:13])
    gr4 <- play_islands(n_islands = 4, size_islands = 4, 
                        p_within = 0.45, m_between = 1) |>
         mutate(name = LETTERS[1:16])
    gr <- bind_graphs(gr1) |>
         bind_graphs(gr2) |>
         bind_graphs(gr3) |>
         graph_join(gr4)
    gr <- as_undirected(gr)
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.15) 
    p <- p + geom_node_point(color = "tan2", size = 16) 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() 
    p
    
    # Save the first graph for subsequent sections
    gr_fig1 <- gr
```

1. List the nodes in all the **cliques** of *size four* in @fig-grex1.

```{r}
    c4 <- cliques(gr, min = 4, max = 4)
    c4_matrix <- matrix(vertex_attr(gr, "name")[unlist(c4)], ncol = 4, byrow = TRUE)
    c4_matrix <- t(apply(c4_matrix, 1, sort))
    kbl(c4_matrix)
```

1. List the nodes in all the **cliques** of *size five* in @fig-grex1.

```{r}
    c5 <- cliques(gr, min = 5, max = 5)
    c5_matrix <- matrix(vertex_attr(gr, "name")[unlist(c5)], ncol = 5, byrow = TRUE)
    c5_matrix <- t(apply(c5_matrix, 1, sort))
    kbl(c5_matrix)
```

  
1. How many **cliques** of *size four* does node *A* belong to?

```{r, include = FALSE}
    c4 <- cliques(gr, min = 4, max = 4)
    count_A <- sum(sapply(c4, function(x) { "A" %in% vertex_attr(gr, "name")[x] }))
```

ANSWER: `r count_A`


1. List the nodes in one of the **2-cliques** in @fig-grex1:

```{r, include = FALSE}
    dist_gr <- distances(gr)
    adj_2clique <- (dist_gr <= 2) * 1
    diag(adj_2clique) <- 0
    g_2clique <- graph_from_adjacency_matrix(adj_2clique, mode = "undirected")
    cl_2clique <- max_cliques(g_2clique)
```

ANSWER: `r paste(sort(vertex_attr(gr, "name")[cl_2clique[[1]]]), collapse = ", ")`


1. List the nodes in the *five cliques of size four* nested inside the clique of size five in @fig-grex1.

```{r}
    c5 <- cliques(gr, min = 5, max = 5)
    c5_nodes <- vertex_attr(gr, "name")[c5[[1]]]
    c4 <- cliques(gr, min = 4, max = 4)
    nested_c4 <- c4[sapply(c4, function(x) { all(vertex_attr(gr, "name")[x] %in% c5_nodes) })]
    nested_matrix <- matrix(vertex_attr(gr, "name")[unlist(nested_c4)], ncol = 4, byrow = TRUE)
    nested_matrix <- t(apply(nested_matrix, 1, sort))
    kbl(nested_matrix)
```

## Structural Equivalence and Structural Similarity

```{r}
#| label: fig-grex2
#| fig-cap: "An undirected graph."
#| fig-cap-location: bottom
#| fig-height: 8
#| fig-width: 8

    set.seed(5678)
    pm <- matrix(c( .15, .11, .12,  1, 
                    .11, .5, .12,   0,
                    .14, .14, .5,  0,
                     1,  0,  0,   0), nrow=4)
    pm[lower.tri(pm)] = t(pm)[lower.tri(pm)]
    gr <- play_blocks(n = 12, size_blocks = c(2, 2, 4, 4), 
                      p_between = pm)
    V(gr)$name <- sample(LETTERS[1:length(V(gr))])
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.07) 
    p <- p + geom_node_point(color = "tan2", size = 14) 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() 
    p
```

- In the matrix below, write down the cell entries for the **adjacency matrix** corresponding to the graph shown in @fig-grex2:

```{r}
    a <- as.matrix(as_adjacency_matrix(gr))
    a <- a[LETTERS[1:nrow(a)], LETTERS[1:nrow(a)]]
    kbl(a, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (2:length(V(gr)), 
                      border_left = T, border_right = T) %>%
         row_spec(1:length(V(gr)), 
                  extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```

- Using the information from the previous matrix, in the matrix below, write down the cell entries for the **structural similarity matrix** corresponding to the graph shown in @fig-grex2 using the **Euclidean Distance**:

```{r}
    a <- as.matrix(round(dist(as.matrix(as_adjacency_matrix(gr))), 2))
    a <- a[LETTERS[1:nrow(a)], LETTERS[1:nrow(a)]]
    kbl(a, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (2:length(V(gr)), 
                      border_left = T, border_right = T) %>%
         row_spec(1:length(V(gr)), 
                  extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```

1. Using the Information in the previous matrix, write down the sets of nodes that are **structurally equivalent** (e.g., connected to the same set of neighbors) in the graph shown in @fig-grex2.

```{r}
    a <- as.matrix(round(dist(as.matrix(as_adjacency_matrix(gr))), 2))
    a <- (a == 0) * 1
    diag(a) <- 0
    a[lower.tri(a)] <- 0
    se <- graph_from_adjacency_matrix(a, mode = "undirected")
    kbl(as_edgelist(se))
```

1. Out of the nodes that are not structurally equivalent, which pairs of nodes have the highest **structural similarity** in the graph shown in @fig-grex2?

```{r}
    a <- as.matrix(round(dist(as.matrix(as_adjacency_matrix(gr))), 2))
    a <- (a == min(a[a!=0])) * 1
    a[lower.tri(a)] <- 0
    se <- graph_from_adjacency_matrix(a, mode = "undirected")
    kbl(as_edgelist(se))
```

1. Using the **Jaccard Similarity** metric, compute the structural similarity between nodes *J* and *M* in the graph shown in @fig-grex1.

```{r, include = FALSE}
    A_fig1 <- as.matrix(as_adjacency_matrix(gr_fig1))
    M_fig1 <- A_fig1 %*% A_fig1
    d_fig1 <- rowSums(A_fig1)
```

ANSWER: `r round(M_fig1["J", "M"]/(d_fig1["J"] + d_fig1["M"] - M_fig1["J", "M"]), 3)`

1. Using the **Dice Similarity** metric, compute the structural similarity between nodes *E* and *N* in the graph shown in @fig-grex1.

ANSWER: `r round((2 * M_fig1["E", "N"])/(d_fig1["E"] + d_fig1["N"]), 3)`

1. Using the **Cosine Similarity** metric, compute the structural similarity between nodes *D* and *H* in the graph shown in @fig-grex1.

ANSWER: `r round((M_fig1["D", "H"])/sqrt(d_fig1["D"] * d_fig1["H"]), 3)`

## Graph Connectivity

```{r, include = FALSE}
    gr <- gr_fig1
```

1. Write down a **node cut set** for the graph shown in @fig-grex1.

```{r, include = FALSE}
    seps <- min_separators(gr)
```

ANSWER: `r paste(vertex_attr(gr, "name")[seps[[1]]], collapse = ", ")`


1. Write down an **edge cut set** for the graph shown in @fig-grex1.

```{r, include = FALSE}
    res_cut <- min_cut(gr, value.only = FALSE)
```

ANSWER: `r paste(gsub("\\|", "--", as_ids(res_cut[['cut']])), collapse = ", ")`

1. What is the **k-connectivity** of the graph in @fig-grex1?

ANSWER: `r vertex_connectivity(gr)`

1. What is the **edge connectivity** of the graph in @fig-grex1?

ANSWER: `r edge_connectivity(gr)`

1. What is the **pairwise k-connectivity** between nodes *E* and *N* in @fig-grex1?

ANSWER: `r vertex_connectivity(gr, source = "E", target = "N")`
