---
title: "Homework IV: Matrix Operations"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

Consider the graphs shown in @fig-grex1:

```{r}
#| label: fig-grex1
#| fig-cap: "Two graphs representing two different set of network ties between the same people."
#| fig-subcap:
#|   - "Studying with network."
#|   - "Going out with network."
#| layout-ncol: 2 
#| fig-height: 10
#| fig-width: 10
    set.seed(123)
    n <- 8
    m <- 13
    g1 <- play_gnm(n = n, m = m, directed = FALSE, loops = FALSE)
    g1 <- mutate(g1, name = LETTERS[1:n])
    p <- ggraph(g1, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white", face = "bold")
    p <- p + theme_graph() 
    p
    n <- 8
    m <- 12
    g2 <- play_gnm(n = n, m = m, directed = FALSE, loops = FALSE)
    g2 <- mutate(g2, name = LETTERS[1:n])
    p <- ggraph(g2, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white", face = "bold")
    p <- p + theme_graph() 
    p
```

1. Write down the **adjacency matrix** corresponding to each network:




```{r}
#| label: tbl-adj
#| tbl-cap: "Two graphs representing two different set of network ties between the same people."
#| tbl-subcap:
#|   - "Studying with network."
#|   - "Going out with network."
#| layout-ncol: 2 

    a <- as.matrix(as_adjacency_matrix(g1))
    b <- as.matrix(as_adjacency_matrix(g2))
    a1 <- a
    b1 <- b
    diag(a1) <- "-"
    kbl(a1, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
    diag(b1) <- "-"
    kbl(b1, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```




1. Write down the **matrix sum** and the **element-wise product** of the **adjacency matrices** you filled out above:


```{r}
#| label: tbl-sumprod
#| tbl-cap: "The sum and element-wise product of the adjacency matrices representing two different set of network ties between the same people."
#| tbl-subcap:
#|   - "Matrix Sum."
#|   - "Element-Wise Product."
#| layout-ncol: 2 
    c <- a + b
    d <- a * b
    c1 <- c
    d1 <- d
    diag(c1) <- "-"
    diag(d1) <- "-"
    kbl(c1, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()

    kbl(d1, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```


1. Write the down the **dyads** that are connected via **multiplex ties** (dyads sharing more than one relationship at a time):

```{r}
    el <- graph_from_adjacency_matrix(c, weighted = TRUE)
    el <- data.frame(as_edgelist(el), E(el)$weight)
    el <- el[el[3] == 2,  ][1:2]
    m <- apply(el, 1, paste, collapse = "")
    sorted_strings <- vapply(strsplit(m, ""), function(chars) {
    paste(sort(chars), collapse = "")
    }, character(1))
    result <- m[!duplicated(sorted_strings)]
    kbl(result, col.names = "", row.names = F)
```


1. Write the down the **dyads** that are connected via **uniplex ties** (dyads sharing only one relationship at a time):

```{r}
    el <- graph_from_adjacency_matrix(c, weighted = TRUE)
    el <- data.frame(as_edgelist(el), E(el)$weight)
    el <- el[el[3] == 1,  ][1:2]
    m <- apply(el, 1, paste, collapse = "")
    sorted_strings <- vapply(strsplit(m, ""), function(chars) {
    paste(sort(chars), collapse = "")
    }, character(1))
    result <- m[!duplicated(sorted_strings)]
    kbl(result, col.names = "", row.names = F)
```

Consider the graphs shown in @fig-dir:


```{r}
#| label: fig-dir
#| fig-cap: "A directed graph."
#| fig-cap-location: margin   
#| fig-height: 12
#| fig-width: 12
    n <- 7
    gr <- sample_gnp(n = n, p = 0.45, directed = TRUE, loops = FALSE)
    V(gr)$name <- LETTERS[1:n]
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(5, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))    
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph()
    p
```

1. Fill out the four matrices below:

```{r}
#| label: tbl-dirmats
#| tbl-cap: "Some Matrices Derived From a Directed Graph."
#| tbl-subcap:
#|   - "Adjacency Matrix (A)."
#|   - "Adjacency Matrix Transpose (A')."
#|   - "One Minus Transpose (1 - A')"
#|   - "Mutual Dyads Matrix (A * A')."
#|   - "Asymmetric Dyads Matrix (A * 1-A')"
#| layout-ncol: 2 

    a <- as.matrix(as_adjacency_matrix(gr))
    a1 <- a
    rownames(a1) <- LETTERS[1:n]
    colnames(a1) <- LETTERS[1:n]
    diag(a1) <- "-"
    kbl(a1, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()

    kbl(t(a), format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
    kbl(1-t(a), format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()

    kbl(a * t(a), format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()

    kbl(a * (1- t(a)), format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (1:n+1, border_left = T, border_right = T) %>%
         row_spec(1:n, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```


1. How many **mutual dyads** in the graph?

```{r}
    kbl(sum(a * t(a))/2, col.names = "")
```

1. How many **asymmetric dyads** in the graph?

```{r}
    kbl(sum(a * (1- t(a))), col.names = "")
```

1. What is the **graph reciprocity**?

```{r}
    kbl((sum(a * t(a))/2)/sum(a * (1- t(a))), col.names = "")
```