---
title: "Homework II: Directed Graphs"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

```{r}
#| label: fig-grex1
#| fig-cap: "A directed graph."
#| fig-cap-location: margin   
#| fig-height: 12
#| fig-width: 12
    set.seed(12456734)
    gr <- sample_gnp(n = 12, p = 0.2, directed = TRUE, loops = FALSE)
    V(gr)$name <- LETTERS[1:12]
    gr <- gr - edge("D|L")
    gr <- gr - edge("E|L")
    gr <- gr + edge("L", "D")
    gr <- gr + edge("L", "E")
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(5, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))    
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph()
    p
```



Consider the graph shown in @fig-grex1:


1. Write down the **edge set** of the graph:
<br>
<br>
<br>
<br>
<br>

## Directed Graph Neighborhoods

1. Who are the **out-neighbors** of node *A*?
<br>
<br>
<br>

1. Who are the **in-neighbors** of node *E*?
<br>
<br>
<br>

1. What is the **intersection** of the **out-neighborhoods** of nodes *A* and *I*?
<br>
<br>
<br>

1. What is the **intersection** of the **in-neighborhoods** of nodes *H* and *E*?
<br>
<br>
<br>

1. What is the **union** of the **out-neighborhoods** of nodes *G* and *K*?
<br>
<br>
<br>

1. What is the **union** of the **in-neighborhoods** of nodes *B* and *I*?
<br>
<br>
<br>

## Directed Graph Degrees

1. What is the **out-degree** of node *H*?
<br>
<br>
<br>

1. What is the **in-degree** of node *E*?
<br>
<br>
<br>

1. What is(are) the node(s) with the largest **out-degree**?
<br>
<br>
<br>

1. What is(are) the node(s) with the largest **in-degree**?
<br>
<br>
<br>

1. What is(are) the node(s) with the smallest **out-degree**?
<br>
<br>
<br>

1. What is(are) the node(s) with the smallest **in-degree**?
<br>
<br>
<br>

## Directed Graph Node Types

1. Which is(are) the **receiver node(s)** in the graph?
<br>
<br>
<br>

1. Which is(are) the **transmitter node(s)** in the graph?
<br>
<br>
<br>

## Directed Graph Metrics

1. What is the graph's **in-degree sequence**?
<br>
<br>
<br>

1. What is the graph's **out-degree range**?
<br>
<br>
<br>

1. What is the graph's **in-degree range**?
<br>
<br>
c

1. What is the graph's **sum of degrees**?
<br>
<br>
<br>

1. What is the **size** of the graph?
<br>
<br>
<br>

1. What is the graph's **maximum size**?
<br>
<br>
<br>

1. Compute the **density** of the graph:
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

## From Graph to Matrix

- In the matrix below, write down the cell entries for the **adjacency matrix** corresponding to the graph shown in @fig-grex1:

```{r}
    a <- matrix(" ", nrow = 12, ncol = 12)
    rownames(a) <- LETTERS[1:12]
    colnames(a) <- LETTERS[1:12]
    diag(a) <- "----"
    kbl(a, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (2:13, border_left = T, border_right = T) %>%
         row_spec(1:12, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```
