---
title: "Homework I: Graph Theory"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
```

## Vertex and edge sets


```{r}
#| label: fig-grex1
#| fig-cap: "A simple graph."
#| fig-height: 10
#| fig-width: 10
    set.seed(489)
    n <- 11
    gr <- play_gnm(n = n, m = 19, directed = FALSE, loops = FALSE)
    gr <- mutate(gr, name = LETTERS[1:n])
    p <- ggraph(gr, layout = 'dh') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 12, color = "white", face = "bold")
    p <- p + theme_graph() 
    p
```

Consider the graph shown in @fig-grex1:

1. Write down the **vertex set** of the graph: 
<br>
<br>
<br>
<br>

1. Write down the **edge set** of the graph:
<br>
<br>
<br>
<br>

## Node Neighborhoods

1. Write down the **neighborhood** of node **C**
<br>
<br>
<br>
<br>

1. Write down the **neighborhood** of node **K**
<br>
<br>
<br>
<br>

1. What is the **intersection** of the neighborhoods of nodes **C** and **K**?
<br>
<br>
<br>
<br>

1. What is the **intersection** of the neighborhoods of nodes **G** and **H**?
<br>
<br>
<br>
<br>

1. What is the **union** of the neighborhoods of nodes **F** and **A**? 
<br>
<br>
<br>
<br>

## Graph Metrics

1. What is the **order** of the graph?
<br>
<br>
<br>
<br>

1. What is the **size** of the graph?
<br>
<br>
<br>
<br>

1. Write down the graph's **degree sequence**:
<br>
<br>
<br>
<br>

1. Write down the graph's **maximum Degree**:
<br>
<br>
<br>
<br>

1. Write down the graph's **minimum degree**:
<br>
<br>
<br>
<br>

1. What is the graph's **degree range**?
<br>
<br>
<br>
<br>

1. What is the graph's **sum of degrees**?
<br>
<br>
<br>
<br>

1. What is the graph's **average degree**?
<br>
<br>
<br>
<br>

1. What is the graph's **maximum size**?
<br>
<br>
<br>
<br>

1. Compute the **density** of the graph:
<br>
<br>
<br>
<br>


## Subgraphs

1. Go back to @fig-grex1. Draw the **node-deleted** subgraph of this graph that *excludes* nodes *F*, and *H*.
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

1. List the **isolate** nodes in the node-deleted subgraph you obtained in the previous step.

<br>
<br>
<br>
<br>



