---
title: "Homework III: Indirect Connections and Matrices"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(kableExtra)
    require(tidygraph)
    require(igraph)
```

## Undirected Paths

Consider the graph shown in @fig-grex1:

```{r}
#| label: fig-grex1
#| fig-cap: "An undirected graph."
#| fig-cap-location: margin   
#| fig-height: 12
#| fig-width: 12
    set.seed(25978457)
    w <- 0
    while (w == 0) {
      d <- sample(2:10, 12, replace = TRUE)
      w <- as.numeric(is_graphical(d))
      }    
    gr <- play_degree(d, method = "simple")
    gr <- mutate(gr, name = LETTERS[1:length(d)])
    gr <- simplify(as.igraph(gr))
    gr <- gr - edge("J|K")
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 20, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() 
    p
```

1. Write down *all* the **paths** of length four ($l = 4$) connecting node *B* and node *E*:
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

1. Write down *all* the **paths** of length two ($l = 2$) featuring node *D* as the **inner node**:
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

1. Write down *all* the **shortest paths** connecting nodes *A* and *D*:
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


1. Write down *all* the **cycles** of length *three* that start and end with node *H*:
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


## Directed Paths

Consider the graph shown in @fig-grex2:

```{r}
#| label: fig-grex2
#| fig-cap: "A directed graph."
#| fig-cap-location: margin   
#| fig-height: 12
#| fig-width: 16
    # Function to sample a weakly connected directed graph
    sample_weakly_directed <- function(n, m) {
    set.seed(4538503)
    repeat {
        g <- sample_gnm(n, m, directed = TRUE)
        if (is_connected(g, mode = "weak")) {
            return(g)
            }
        }
    }
    # Example: 10 nodes and 12 edgesn
    n <- 10
    m <- 22
    gr2 <- sample_weakly_directed(n, m)
    V(gr2)$name <- LETTERS[1:n]
    #ifelse(is.finite(distances(gr2, mode = "out")), 1, 0)
    p <- ggraph(gr2, layout = 'auto') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(7, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() 
    p
```

1. Write down *all* the **directed path(s)** of *any* length going *from* node *F* *to* node *A*:
<br>
<br>
<br>
<br>
<br>
<br>


1. What is the length of the **shortest directed path(s)** going *from* node *I* *to* node *H*?
<br>
<br>
<br>

1. If *B* wanted to send a message to *C* in the most efficient way, how many intermediaries would *B* have to use?
<br>
<br>
<br>


## Matrices

- In the matrix below, write down the cell entries **geodesic distance matrix** corresponding to the graph shown in @fig-grex1:

```{r}
    n <- vcount(gr)
    m <- n - 1
    a <- matrix(" ", nrow = n, ncol = n)
    rownames(a) <- V(gr)$name
    colnames(a) <- V(gr)$name
    diag(a) <- "----"
    kbl(a, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (2:n, border_left = T, border_right = T) %>%
         row_spec(1:m, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```

- What is the **diameter** of the graph shown in @fig-grex1?
<br>
<br>
<br>


- What is the **eccentricity** of each node graph shown in @fig-grex1?
```{r}
    n <- vcount(gr)
    a <- matrix(" ", nrow = 1, ncol = n)
    colnames(a) <- V(gr)$name
    kbl(a, format = "html", align = "c")  |> 
         column_spec(1, bold = TRUE) |> 
         column_spec (1:n, border_left = T, border_right = T) |> 
         row_spec(1, extra_css = "border-bottom: 1px solid; height: 50px") |> 
         kable_styling()
```

- In the matrix below, write down the cell entries for the **reachability matrix** corresponding to the graph shown in @fig-grex2:

```{r}
    n <- vcount(gr2)
    m <- n - 1
    b <- matrix(" ", nrow = n, ncol = n)
    rownames(b) <- sort(V(gr2)$name)
    colnames(b) <- sort(V(gr2)$name)
    diag(b) <- "----"
    kbl(b, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (2:n, border_left = T, border_right = T) %>%
         row_spec(1:m, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```

- What flavor of connectivity is the graph in @fig-grex2?
<br>
<br>
<br>

