---
title: "Homework III: Indirect Connections and Matrices"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(kableExtra)
    require(tidygraph)
    require(igraph)
```

## Undirected Paths

Consider the graph shown in @fig-grex1:

```{r}
#| label: fig-grex1
#| fig-cap: "An undirected graph."
#| fig-cap-location: margin   
#| fig-height: 12
#| fig-width: 12
    set.seed(25978457)
    w <- 0
    while (w == 0) {
      d <- sample(2:10, 12, replace = TRUE)
      w <- as.numeric(is_graphical(d))
      }    
    gr <- play_degree(d, method = "simple")
    gr <- mutate(gr, name = LETTERS[1:length(d)])
    gr <- simplify(as.igraph(gr))
    gr1 <- gr - edge("J|K")
    p <- ggraph(gr1, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 20, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() 
    p
```




1. Write down *all* the **paths** of length four ($l = 4$) connecting node *B* and node *E*:




```{r}
    paths <- all_simple_paths(gr, from = "B", to = "E")
    p <- paths[which(sapply(paths, length) == 5)]
    kbl(toupper(matrix(letters[unlist(p)], ncol = 5, byrow = TRUE)))
```




1. Write down *all* the **paths** of length two ($l = 2$) featuring node *D* as the **inner node**:




```{r}
    paths <- list()
    for (i in 1:vcount(gr)) {
      p <- all_simple_paths(gr, from = i, cutoff = 2)
      p <- p[which(sapply(p, length) == 3)]
      paths[[i]] <- p
      }
    m <- matrix(LETTERS[unlist(paths)], ncol = 3, byrow = TRUE)
    m <- m[which(m[, 2] == "D"), ]
    m <- m[which(m[, 1] != m[, 3]), ]
    m <- apply(m, 1, paste, collapse = "")
    sorted_strings <- vapply(strsplit(m, ""), function(chars) {
    paste(sort(chars), collapse = "")
    }, character(1))
    result <- m[!duplicated(sorted_strings)]
    kbl(result, col.names = "")
```



1. Write down *all* the **shortest paths** connecting nodes *A* and *D*:
```{r}
    p <- all_shortest_paths(gr, from = "A", to = "D")
    p <- lapply(p$vpaths, names)
    kbl(matrix(unlist(p), nrow = length(p), ncol = length(p[[1]]), byrow = TRUE))
```




1. Write down *all* the **cycles** of length *three* that start and end with node *H*:




```{r}
    k <- 4
    c1 <- lapply(all_simple_paths(gr, "D", "H"), function(p) c("G",p))
    c2 <- lapply(all_simple_paths(gr, "G", "H"), function(p) c("G",p))
    c3 <- lapply(all_simple_paths(gr, "I", "H"), function(p) c("G",p))
    c4 <- lapply(all_simple_paths(gr, "A", "H"), function(p) c("G",p))
    c5 <- lapply(all_simple_paths(gr, "F", "H"), function(p) c("G",p))
    c <- c(c1, c2, c3, c4, c5)
    c <- c[which(lapply(c, length) == k)]
    m <- matrix(unlist(lapply(c, names)), ncol = k, byrow = TRUE)
    m[, 1] <- rep("H", nrow(m))
    m <- apply(m, 1, paste, collapse = "")
    sorted_strings <- vapply(strsplit(m, ""), function(chars) {
    paste(sort(chars), collapse = "")
    }, character(1))
    result <- m[!duplicated(sorted_strings)]
    kbl(result, col.names = "")
```




## Directed Paths

Consider the graph shown in @fig-grex2:

```{r}
#| label: fig-grex2
#| fig-cap: "A directed graph."
#| fig-cap-location: margin   
#| fig-height: 12
#| fig-width: 16
    # Function to sample a weakly connected directed graph
    sample_weakly_directed <- function(n, m) {
    set.seed(4538503)
    repeat {
        g <- sample_gnm(n, m, directed = TRUE)
        if (is_connected(g, mode = "weak")) {
            return(g)
            }
        }
    }
    # Example: 10 nodes and 12 edgesn
    n <- 10
    m <- 22
    gr2 <- sample_weakly_directed(n, m)
    V(gr2)$name <- LETTERS[1:n]
    #ifelse(is.finite(distances(gr2, mode = "out")), 1, 0)
    p <- ggraph(gr2, layout = 'auto') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(7, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() 
    p
```




1. Write down *all* the **directed path(s)** of *any* length going *from* node *F* *to* node *A*:

```{r}
    paths <- all_simple_paths(gr2, from = "F", to = "A", mode = "out")
    paths <- lapply(paths, names)

    df_collapsed <- data.frame(
    Paths = sapply(paths, paste, collapse = ", ")
    )

    kbl(df_collapsed)
```


1. What is the length of the **shortest directed path(s)** going *from* node *I* *to* node *H*?

```{r}
    paths <- all_shortest_paths(gr2, from = "I", to = "H", mode = "out")
    length(paths[[1]]) + 1
```

1. If *B* wanted to send a message to *C* in the most efficient way, how many intermediaries would *B* have to use?

```{r}
    paths <- all_shortest_paths(gr2, from = "B", to = "C", mode = "out")
    length(paths$res[[1]]) - 2
```

## Matrices

- In the matrix below, write down the cell entries **geodesic distance matrix** corresponding to the graph shown in @fig-grex1:

```{r}
    m <- as.matrix(distances(gr1))
    kbl(m, format = "html", align = "c")  |> 
         column_spec(1, bold = TRUE) |> 
         column_spec (1:nrow(m), border_left = T, border_right = T) |> 
         row_spec(1, extra_css = "border-bottom: 1px solid; height: 50px") |> 
         kable_styling()
```

- What is the **diameter** of the graph shown in @fig-grex1?

`r max(distances(gr))`


- What is the **eccentricity** of each node graph shown in @fig-grex1?
```{r}
    e <- apply(m, 1, max)
    kbl(t(e), format = "html", align = "c")  |> 
         column_spec(1, bold = TRUE) |> 
         column_spec (1:length(e), border_left = T, border_right = T) |> 
         row_spec(1, extra_css = "border-bottom: 1px solid; height: 50px") |> 
         kable_styling()
```

- In the matrix below, write down the cell entries for the **reachability matrix** corresponding to the graph shown in @fig-grex2:

```{r}
    m <- as.matrix(distances(gr2, mode = "out"))
    r <- nrow(m)
    s <- r-1
    m <- (1/m > 0) * 1
    kbl(m[order(rownames(m)), order(colnames(m))], format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (2:r, border_left = T, border_right = T) %>%
         row_spec(1:s, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```


- In the matrix below, write down the cell entries **geodesic distance matrix** corresponding to the graph shown in @fig-grex2:

```{r}
    b <- as.matrix(distances(gr2, mode = "out"))
    b <- b[sort(rownames(b)), sort(colnames(b))]
    r <- nrow(m)
    s <- r-1
    kbl(b, format = "html", align = "c")  %>% 
         column_spec(1, bold = TRUE) %>% 
         column_spec (2:r+1, border_left = T, border_right = T) %>%
         row_spec(1:s, extra_css = "border-bottom: 1px solid") %>%
         kable_styling()
```

- What flavor of connectivity is the graph in @fig-grex2?

Weakly Connected

