# Types of Ties and Their Graphs {#sec-ties}

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
  library(ggplot2)
  library(ggraph)
  library(tidygraph)
  library(igraph)
  library(patchwork)
```

## Introduction

Nodes and edges are the fundamental building blocks of a graph. However, the types of relationships that the edges represent can change both how we understand the network conceptually and what mathematical techniques we can apply to the graph when computing *graph metrics* (the subject of @sec-graphmetrics). When doing social network analysis, we map our understanding of the nature of the **social relationships** we are studying to the **types of graphs** we use to represent the network formed by the **concatenation** of those relationships [@martin09]. 

This chapter integrates the sociological classifications of ties with their graph-theoretic representations, demonstrating how different social relations dictate the mathematical properties and structures of the networks they form.

---

## Symmetric Ties and Undirected Graphs 

By definition, some relations lack any inherent directionality. **Mutuality** (or **reciprocity**) is built in by construction. For example, "spending time together" or "being siblings" are relations where mutuality is logically and necessarily implied. It would be nonsensical for a person (say *A*) to claim that they spend time with another person (say *B*) and for *B* to say that they do not spend time with *A*. In social network analysis, these are called **symmetric ties** [@heider46]. Formally, if a tie is symmetric, then the existence of a tie from *A* to *B* logically implies that *B* is also tied to *A* ($AB \implies BA$).

Social networks composed of symmetric ties are represented as **undirected graphs**, such as the one shown in @fig-undirected. When an undirected graph has no **loops** (edges connecting a node to itself), and there is only one edge connecting adjacent vertices to one another (the graph has no **multiedges**), it is called a **simple undirected graph**.

```{r}
#| label: fig-undirected
#| fig-cap: "An undirected graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  gr <- create_empty(9, directed = FALSE)
  gr <- gr %>% 
    bind_edges(data.frame(from = 1, to = 2:5)) %>% 
    bind_edges(data.frame(from = 2, to = 3:4)) %>% 
    bind_edges(data.frame(from = 3, to = 4:5)) %>% 
    bind_edges(data.frame(from = 4, to = 5:6)) %>% 
    bind_edges(data.frame(from = 6, to = c(7:9))) %>%
    bind_edges(data.frame(from = 7, to = c(8:9))) %>%
    bind_edges(data.frame(from = 8, to = c(9))) %>%
    mutate(name = toupper(letters[1:9]))
    
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_link(color = "steelblue", width = 1.25) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

### Real-World Examples of Symmetric Ties

1.  **Similarity and Propinquity**: Two people being in the same place at the same time (co-location), or having similar socio-demographic attributes (e.g., sharing nationality or gender).
2.  **Co-membership / Affiliations**: "Being in the same class as" every other student in your seminar. All **co-memberships** create symmetric ties among all involved actors (we will revisit this in @sec-affiliation). If *A* is a teammate of *B* in a soccer club, *B* is necessarily a teammate of *A*.

In undirected graphs, the order in which we name the nodes in an edge does not matter ($AB$ and $BA$ refer to the exact same edge).

---

## Asymmetric Ties and Directed Graphs 

In contrast to co-membership or spending time together, many social relationships allow for inherent **directionality**. These are represented by **asymmetric ties** [@heider46]. For these ties, one member of a pair can claim a relationship with the other, but it is possible (though not necessary) that the other person does not send the same relationship back. 

Helping, texting, and **social support** relations are prime examples. You can help someone with their homework or give them personal advice, but this does not *necessarily* mean that that person will return the favor. **Mutuality** or **reciprocity** is not built-in by definition; instead, it must happen as an empirical event in the world. 

Social networks composed of asymmetric ties are represented by **directed graphs** (or **digraphs**). @fig-directed shows a point-and-line diagram of a digraph, where simple undirected lines are replaced by directed arrows (also called **arcs**).

```{r}
#| label: fig-directed
#| fig-cap: "A directed graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "A", "B", "B", "B", "D", "E", "E", "F", "G", "G")
  to <- c("B", "F", "A", "D", "C", "B", "C", "D", "A", "F", "D")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

In a directed graph, for every edge, there is a **source node** and a **destination node**. Therefore, *the order in which you list the nodes when you name the edges matters*. The edge $AB$ is a different entity from the edge $BA$. When a digraph has no loops and at most one directed edge connecting a sender to a receiver in each direction, it is called a **simple directed graph**.

### Reciprocity as an Empirical Concept

In directed networks, reciprocity is a major analytical focus. For some ties (such as liking, trusting, or friendship), reciprocity is binary; it either exists or it does not. For other ties (such as communication networks built on text volume or call duration), reciprocity is a matter of degree. For instance, you can text someone 10 times a day, but they may text you back only half the time. 

If @fig-directed were an **advice network** [@cross01], we could say that *H* seeks advice from *D*, but *D* does not seek advice from *H*. This lack of reciprocity is common in formal hierarchies, where advice-seeking often flows upward to more experienced or authoritative actors, reflecting power differentials.

### Types of Nodes in Directed Graphs {#sec-typedir}

In a directed graph, we can categorize (non-isolate) nodes based on their sending and receiving behavior [@Harary65]: 

*   **Receivers**: Nodes that receive ties but do not send them (e.g., node *C* in @fig-directed). For receiver nodes, indegree $k_{in} > 0$ and outdegree $k_{out} = 0$.
*   **Carriers**: Nodes that both receive and send ties (e.g., nodes *A* and *B* in @fig-directed). For carrier nodes, $k_{in} > 0$ and $k_{out} > 0$.
*   **Transmitters**: Nodes that send ties but do not receive them (e.g., nodes *E* and *G* in @fig-directed). For transmitter nodes, $k_{in} = 0$ and $k_{out} > 0$.

---

## Visual and Sociological Taxonomies

To systematically synthesize the types of ties, we can classify them both graph-theoretically and sociologically.

### Graph-Theoretic Classification of Ties

In graph theory, we categorize directed and undirected ties based on their inherent directionality. Figure @fig-graph-theoretic-ties provides a visual summary of these types.

```{r}
#| label: fig-graph-theoretic-ties
#| fig-cap: "A Graph-Theoretic Classification of Ties"
#| echo: false
#| fig-width: 8
#| fig-height: 6

# Function to create a clean, minimalist 2-node graph plot
plot_tie_type <- function(nodes, edges, title, subtitle, directed = TRUE, has_cross = FALSE) {
  g <- tbl_graph(nodes = nodes, edges = edges)
  
  p <- ggraph(g, layout = "manual", x = x, y = y)
  
  if (directed) {
    p <- p + geom_edge_arc(aes(strength = strength, color = color, linetype = linetype),
                           arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                           end_cap = circle(8, 'mm'),
                           start_cap = circle(8, 'mm'),
                           width = 1.1,
                           show.legend = FALSE) +
             scale_edge_color_identity() +
             scale_edge_linetype_identity()
  } else {
    p <- p + geom_edge_link(color = "#2c3e50",
                            width = 1.1,
                            end_cap = circle(8, 'mm'),
                            start_cap = circle(8, 'mm'),
                            show.legend = FALSE)
  }
  
  # Add nodes
  p <- p + geom_node_point(size = 14, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
    geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "#2c3e50")
  
  # Add red "forbidden" cross if needed
  if (has_cross) {
    p <- p + annotate("text", x = 1.5, y = 0.88, label = "✕", color = "#e74c3c", size = 8, fontface = "bold")
  }
  
  p <- p + 
    labs(title = title, subtitle = subtitle) +
    theme_graph() +
    theme(
      plot.title = element_text(size = 12, face = "bold", color = "#2c3e50", hjust = 0.5),
      plot.subtitle = element_text(size = 9.5, color = "#7f8c8d", hjust = 0.5, lineheight = 1.1),
      plot.margin = margin(10, 10, 10, 10)
    ) +
    coord_cartesian(xlim = c(0.6, 2.4), ylim = c(0.7, 1.3), clip = "off")
  
  return(p)
}

# Define Nodes
nodes_2 <- data.frame(name = c("A", "B"), x = c(1, 2), y = c(1, 1))

# 1. Symmetric (Undirected)
edges_sym <- data.frame(from = 1, to = 2)
p1 <- plot_tie_type(nodes_2, edges_sym, "Symmetric", "Undirected tie\ne.g., 'is sibling of'", directed = FALSE)

# 2. Asymmetric: Reciprocal
edges_rec <- data.frame(from = c(1, 2), to = c(2, 1), strength = c(0.2, 0.2), color = c("#2c3e50", "#2c3e50"), linetype = c("solid", "solid"))
p2 <- plot_tie_type(nodes_2, edges_rec, "Asymmetric: Reciprocal", "Mutual directed ties\ne.g., 'A likes B' & 'B likes A'", directed = TRUE)

# 3. Asymmetric: Non-Reciprocal
edges_nonrec <- data.frame(from = 1, to = 2, strength = 0, color = "#2c3e50", linetype = "solid")
p3 <- plot_tie_type(nodes_2, edges_nonrec, "Asymmetric: Non-Reciprocal", "One-way directed tie\ne.g., 'A sends email to B'", directed = TRUE)

# 4. Asymmetric: Anti-Reciprocal
edges_anti <- data.frame(
  from = c(1, 2), 
  to = c(2, 1), 
  strength = c(0.2, 0.2), 
  color = c("#2c3e50", "#e74c3c"), 
  linetype = c("solid", "dashed")
)
p4 <- plot_tie_type(nodes_2, edges_anti, "Asymmetric: Anti-Reciprocal", "Reverse tie is impossible\ne.g., 'A is boss of B'", directed = TRUE, has_cross = TRUE)

# Combine using patchwork
combined <- (p1 + p2) / (p3 + p4) + 
  plot_annotation(
    title = "Graph-Theoretic Classification of Ties",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", color = "#2c3e50", hjust = 0.5, margin = margin(b = 15))
    )
  )

combined
```

*   **Symmetric (Undirected)**: Lack meaningful directionality (e.g., "is sibling of"). Represented by undirected graphs.
*   **Asymmetric (Directed)**: Direction-dependent relations. They split into three flavors:
    *   **Reciprocal**: Edges exist in both directions, indicating mutual interaction (e.g., mutual liking or active email exchange).
    *   **Non-Reciprocal**: An edge exists in only one direction (e.g., sending an email without getting a reply).
    *   **Anti-Reciprocal**: The presence of a tie in one direction logically implies the absence of a tie in the opposite direction (e.g., hierarchical relations like "is boss of").

### Sociological Classification of Social Ties

Sociologists categorize social ties into four main types: **Similarities**, **Social Relations**, **Interactions**, and **Flows**. These four types can be further grouped into **states** (static situations that persist over time, like similarities and social relations) and **events** (repeated behavioral processes that unfold over time, like interactions and flows). Table @tbl-sociological-ties provides a summary.

| Category | Type | Description | Examples |
|:---------|:---------------------|:----------------------------------------------------------------------------|:---------------------------------------------------------------------------|
| **States** | **Similarities** | Connections based on having something in common; typically symmetric. | Propinquity (same school), Memberships (same club), Attributes (same gender). |
| | **Social Relations** | Connections based on a culturally recognized role or an affective/cognitive attitude. | **Roles**: Kinship (sibling, parent), Non-Kin (friend, boss). <br> **Attitudes**: Affective (like, hate), Cognitive (knows, is aware of). |
| **Events** | **Interactions** | Regular communication or behavior directed at another person; usually asymmetric. | Communications (texting, calling), Social Support (advice, help), Dominance (bullying, fighting). |
| | **Flows** | Transmission of something from one person to another; inherently directed. | Information (gossip, influence), Beliefs, Resources (economic transactions). |

: A Sociological Classification of Social Ties {#tbl-sociological-ties}

#### Similarities
Similarities represent static states where individuals share a common attribute or context. Ties based on similarities are typically symmetric. Examples include *propinquities* (being co-located in a neighborhood), *memberships* (belonging to the same professional organization), and *demographic attributes* (sharing the same age or ethnicity).

#### Social Relations
Social relations describe culturally recognized roles or cognitive/affective attitudes. These are also static states. They can be symmetric (friendship, sibling) or asymmetric (boss-subordinate, parent-child). Sibling ties are symmetric, while parent/child ties are anti-reciprocal (if *A* is the parent of *B*, *B* cannot be the parent of *A*). Affective ties (liking, trusting, hating) and cognitive ties (knowing of someone, being aware of someone) are inherently asymmetric.

#### Interactions
Interactions are behavioral events occurring in time. They are asymmetric by default but can be reciprocal. Typical examples include *communications* (texting, calling), *social support* (exchanging advice or lending money), and *dominance* or negative interactions (bullying, physical fighting).

#### Flows
Flows involve the directed movement of some entity (information, resources, or beliefs) from a sender to a receiver. They are dynamic events that unfold over time and are inherently directional. Examples include the diffusion of *information* (gossip, news, influence) and the transfer of *resources* (economic transactions, disease transmission).

---

## Special Classes of Directed Graphs

By imposing structural restrictions on reciprocity, degree, or edge directionality, we can define special classes of directed graphs.

### Symmetric Directed Graphs

If, in a directed graph, every pair of connected nodes is linked by a bi-directional, reciprocal relationship, the graph is a **symmetric directed graph**. 

```{r}
#| label: fig-symm
#| fig-cap: "A symmetric directed graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "B", "D", "A", "B", "C", "E", "C", "F", "A", "G", "D", "C", "D", "E", "G")
  to <- c("B", "A", "A", "D", "C", "B", "C", "E", "A", "F", "D", "G", "D", "C", "G", "E")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

@fig-symm shows a symmetric directed graph with eight nodes and sixteen bi-directional edges (thirty-two directed arcs). For every edge from $u$ to $v$, there exists a corresponding edge from $v$ to $u$.

### Functional Graphs

A **functional graph** is a special kind of digraph built by imposing the restriction that each node can only have a *single out-neighbor* (i.e., the outdegree of every node is exactly equal to one, $k_{out} = 1$). Functional graphs have important applications in the anthropological analysis of kinship exchange systems in traditional societies [@hageharary96].

```{r}
#| label: fig-func
#| fig-cap: "A functional directed graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "B", "C", "D", "E", "F", "G", "H")
  to <- c("C", "G", "G", "H", "A", "A", "B", "C")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

@fig-func shows a functional graph with eight nodes. If a graph $G$ is a functional graph, then the number of nodes $V$ is always equal to the number of directed edges $E$ ($V = E = 8$). This is mathematically constrained because each of the $V$ nodes sends out exactly one edge.

### Oriented Graphs {#sec-oriented}

An **oriented graph** is a directed graph containing no symmetric (reciprocal) dyads. If there is a directed link from $u$ to $v$, there cannot be a link from $v$ to $u$. It is the structural opposite of a symmetric directed graph.

```{r}
#| label: fig-oriented
#| fig-cap: "An oriented directed graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "D", "C", "C", "A", "D", "D", "G", "C", "B", "G")
  to <- c("B", "A", "F", "E", "F", "G", "C", "E", "A", "F", "A")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

In the oriented graph shown in @fig-oriented, every connected pair is linked *only* by a one-way directed tie; reciprocal dyads are forbidden by the nature of the relationship (e.g., strict hierarchy or non-reversible flows).

### Anti-Symmetric Ties and Tree Graphs

As we saw earlier, an asymmetric relation is **anti-reciprocal** if reciprocity is strictly forbidden. If *A* is anti-reciprocally connected to *B*, then *B* *cannot* send the *same type of tie* back to *A*. When these relations are concatenated across multiple nodes, they can form a **directed tree graph** (for a more general consideration of tree graphs see @sec-trees). 

Consider the relationship "being the father of." You have a father, your father has a father, and so on. This creates a generational lineage represented as a tree. 

```{r}
#| label: fig-tree
#| fig-cap: "A tree graph with thirteen nodes"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  gr <- create_tree(n = 13, children = 3) 
  gr <- mutate(gr, name = toupper(letters[1:13])) 
  
  p <- ggraph(gr, layout = 'tree') +
    geom_edge_link(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(4, 'mm')),
                end_cap = circle(8, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

In @fig-tree, node $A$ is the earliest common ancestor, and all other nodes are anti-reciprocally related to it. Similarly, formal organizational structures where subordinates "take orders from" a single supervisor generate tree graphs. A notable property of tree graphs is that every node (except the root) has exactly one in-neighbor (one supervisor or one father), but can have multiple out-neighbors (children or subordinates).

---

## Real-World Structural Overlays

In real-world settings, simple graph models are often combined to capture the rich social environments of families, organizations, and communities.

*   **Hierarchies in Organizations**: Formal organizations overlay distinct tie types:
    *   **Same-level ties** (e.g., coworkers, office mates) are typically **symmetric**, implying parity and equality.
    *   **Cross-level ties** (e.g., supervisor-subordinate) are **anti-reciprocal** and **oriented**, representing vertical authority. These ties are often transitive: if *A* is the boss of *B*, and *B* is the boss of *C*, then *A* is also effectively the boss of *C*.
*   **Kinship Networks in Families**: Family structures mirror this logic:
    *   **Same-generation ties** (e.g., siblings, cousins) are **symmetric**.
    *   **Cross-generation ties** (e.g., parent-of, uncle-of) are **anti-reciprocal** and transitive.

### Multiplexity in Social Relations

In actual social networks, actors are rarely linked by only one relationship. **Multiplexity** refers to the presence of multiple, distinct types of ties linking the same pair of nodes simultaneously. 

For example, two individuals in an office might be linked by a formal, anti-reciprocal role relationship (supervisor and employee) and also by an informal, asymmetric sentiment tie (mutual friendship or liking). Multiplexity adds rich, multi-layered complexity to networks, as the behavior or pattern in one relation (e.g., friendship) often heavily influences or moderates another (e.g., work productivity).

### The Growth of Social Networks

Social networks are highly dynamic. They grow and evolve as new ties form, often linking previously separate network structures. 

A classic example is the formation of a new family unit. A romantic tie (an asymmetric, non-kin relation) between two individuals from separate families can culminate in a marriage tie (a symmetric, kin relation). This single event triggers a cascade of new "in-law" ties (both symmetric same-generation ties and anti-reciprocal cross-generation ties). This effectively merges two separate kinship trees into a highly complex, interconnected network structure.

---

## References {.unnumbered}
