# Triadic Balance {#sec-tbalance}

Triadic Balance Theory is a social psychological framework that aims to explain the anticipated configurations of sentiments within social networks, extending Heider's [-@heider46] concept of dyadic balance to include a third party.

In Balance Theory, a social-psychological theory, sentiment networks are built from **valenced ties**, which can be positive (e.g., liking) or negative (e.g., disliking). The theory states that unbalanced configurations within a triad (a three-person social network consisting of P, an object O, and a third party Q) produce "tension" for the person (P). This tension creates a psychological drive for these configurations to transition towards balanced ones. A balanced triad has either zero or two negative links, while an unbalanced triad has an odd number of negative links (one or three).

## The key principles of Triadic Balance Theory

Balance theory utilizes four basic components to analyze triadic configurations:

- A **person** (P): The focal individual from whose perspective the triad is analyzed.
- An **other** (O): Towards which the focal person (P) holds a sentiment, which can be either positive or negative.
- A **third party** (Q): Towards which both the person (P) and the other (O) also have some sentiment. It's important to note that Q does not necessarily have to be a real person; it can be any object.
- **Directed, Signed Edges**: These indicate positive or negative sentiments that P and O have towards the third party Q, and P towards O. For instance, if "P likes O" is represented as a directed edge with a positive sign going from P to O. In contrast, "O hates Q" is represented by a directed edge going from O to Q with a negative sign. 

## Balanced vs. Unbalanced Configurations:

Triadic balance theory distinguishes between **balanced** and **unbalanced** triadic configurations.

### Balanced Triads

In Balance Theory, balanced triadic configurations are characterized by harmonious and stable sentiments, creating no "tension" within the triad. These configurations follow a mathematical rule where multiplying the signs of the edges (positive ties as +1, negative ties as -1) results in a positive number (+1). A simpler rule for identification is that a triad is balanced if it has either zero or two negative links.

Here are some concrete examples of balanced triadic configurations:

1. **"A friend of a friend is a friend"**:

- Description: In this configuration, Person (P) likes Object (O) (positive tie, +1), Object (O) likes a Third Party (Q) (positive tie, +1), and Person (P) also likes the Third Party (Q) (positive tie, +1). All sentiments are positive. This configuration of triadic sentiments is depicted in @fig-balanced-1.
- Example: Imagine that you (P) have a friend (O) whom you like, and they like another classmate (Q) who you also like. Everything seems fine here!
- Mathematical Representation: (+1) * (+1) * (+1) = +1.
- Sign Pattern: P-O (+) , O-Q (+) , P-Q (+). This triad has zero negative links; thus, it is balanced.

2. **"An enemy of an enemy is a friend"**:

- Description: Person (P) dislikes Object (O) (negative tie, -1), Object (O) dislikes a Third Party (Q) (negative tie, -1), but Person (P) likes the Third Party (Q) (positive tie, +1). This configuration of triadic sentiments is depicted in @fig-balanced-2.
- Example: Imagine that you (P) have an enemy (O) who likes another classmate (Q) whom you *also* dislike. This seems like a perfectly fine arrangement!
- Mathematical Representation: (-1) * (-1) * (+1) = +1.
- Sign Pattern: P-O (-) , O-Q (-) , P-Q (+). This triad has two negative links; thus, it is balanced.

3. **"An enemy of a friend is an enemy"**:

- Description: If a person P dislikes O (negative tie, -1), and O likes Q (positive tie, +1), then P should dislike Q (negative tie, -1). This configuration of triadic sentiments is depicted in @fig-balanced-3.
- Example: Imagine that you (P) have a huge fight with somebody else (O). You later find out that they have a best friend (Q). It makes sense that you will start them as well!
- Mathematical Representation: (-1) * (+1) * (-1) = +1.
- Sign Pattern: P-O (-) , O-Q (+) , P-Q (-). This triad has two negative links; thus, it is balanced.

4. **"A friend of an enemy is an enemy"**:

- Description: In this scenario, if P likes O (positive tie, +1), and O dislikes Q (negative tie, -1), then P should dislike Q (negative tie, -1) to maintain balance. This configuration of triadic sentiments is depicted in @fig-balanced-4.
- Example: Imagine that country P declares war on country O. After that, they find that country Q is an ally of country O. This means that country P *must* also declare war on country Q. 
- Mathematical Representation: (+1) * (-1) * (-1) = +1.
- Sign Pattern: P-O (+) , O-Q (-) , P-Q (-). This triad has two negative links; thus, it is balanced.

```{r}
#| label: fig-balanced
#| fig-cap: "The four balanced triads according to balance theory."
#| fig-cap-location: bottom
#| fig-subcap:
#|   - ""
#|   - "" 
#|   - ""
#|   - ""
#| layout-ncol: 2
#| fig-height: 8
#| fig-width: 8
    library(igraph)
    library(ggraph)
    # Helper function to generate a plot for a specific balanced triad
    plot_balanced_triad <- function(edge_signs, title) {
    
    # Define the directed edges based on the text: P->O, O->Q, P->Q
    edges <- data.frame(
        from = c("P", "O", "P"),
        to   = c("O", "Q", "Q"),
        sign = edge_signs
    )
    
    # Create the igraph object
    g <- graph_from_data_frame(edges, directed = TRUE)
    
    # Define a fixed layout matrix to keep the triangle uniform across all plots
    # Node order implicitly extracted by igraph: P (1), O (2), Q (3)
    layout_matrix <- matrix(c(
        0.000,  1.0,  # P at the top center
        -0.866, -0.5,  # O at the bottom left
        0.866, -0.5   # Q at the bottom right
    ), ncol = 2, byrow = TRUE)
    
    # Plot using ggraph
    ggraph(g, layout = layout_matrix) +
        geom_edge_link(aes(color = sign, label = sign),
                    arrow = arrow(length = unit(8, 'mm'), type = "closed"),
                    end_cap = circle(12, 'mm'),
                    start_cap = circle(8, 'mm'),
                    angle_calc = 'along',
                    label_dodge = unit(6, 'mm'),
                    label_size = 18,
                    width = 2,
                    show.legend = FALSE) +
        geom_node_point(size = 26, color = "tan2") +
        geom_node_text(aes(label = name), size = 12, fontface = "bold", color = "white") +
        scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
        labs(title = title) +
        theme_graph() + xlim(-1, 1) + ylim(-1, 1.25) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20, margin = margin(b = 10)))
    }
    # 1. Zero negative links: (+), (+), (+)
    plot_balanced_triad(c("+", "+", "+"), "1. A friend of a friend\nis a friend")
    # 2. Two negative links: (-), (-), (+)
    plot_balanced_triad(c("-", "-", "+"), "2. An enemy of an enemy\nis a friend")
    # 3. Two negative links: (-), (+), (-)
    plot_balanced_triad(c("-", "+", "-"), "3. An enemy of a friend\nis an enemy")
    # 4. Two negative links: (+), (-), (-)
    plot_balanced_triad(c("+", "-", "-"), "4. A friend of an enemy\nis an enemy")
```

### Unbalanced Triads

Unbalanced triadic configurations in Balance Theory are those that create "tension" and are considered unstable. The theory states that these unbalanced triads will tend to transition to balanced configurations as the person (P) changes their sentiments toward the object (O) or the third party (Q) to alleviate this tension. Mathematically, if you multiply the signs of the edges of an unbalanced triad, the result is a negative number (-1). A simpler rule is that a triad is unbalanced if it has an odd number of negative links (one or three).

Here are some concrete examples of unbalanced triadic configurations:

1. **"A friend of my friend is my enemy"**:

- Description: In this configuration, Person (P) likes Object (O) (a positive tie, +1), Object (O) likes a Third Party (Q) (a positive tie, +1), but Person (P) dislikes the Third Party (Q) (a negative tie, 1). This configuration of triadic sentiments is depicted in @fig-unbalanced-1.
- Example: "You (P) have a friend (O) whom you like and they like another classmate (Q) who you dislike".
- Mathematical Representation: P-O (+) * O-Q (+) * P-Q (-) = (+1) * (+1) * (-1) = -1.
- Implication: This situation creates discomfort due to the conflicting sentiment P has towards their friend and their friend's friend.

2. **"A friend is my friend's enemy"**:

- Description: Person (P) likes Object (O) (a positive tie, +1), Object (O) dislikes Third Party (Q) (a negative tie, -1), but Person (P) likes Third Party (Q) (a positive tie, +1). This configuration of triadic sentiments is depicted in @fig-unbalanced-2.
- Example: "You (P) love Marvel movies (Q) but your long time partner (O) (who you like duh!) hates them".
- Mathematical Representation: P-O (+) * O-Q (-) * P-Q (+) = (+1) * (-1) * (+1) = -1.
- Implication: This configuration also leads to tension in the triad due to the conflicting sentiments P and O have toward the third party Q.

3. **"My enemy's friend is also my friend"**:

- Description: Person (P) dislikes Object (O) (a negative tie, -1), Object (O) likes Third Party (Q) (a positive tie, +1), but Person (P) likes Third Party (Q) (a positive tie, +1).
- Example: Imagine (the horror) that you (P) and your archnemesis (O) are attracted to the same person (Q). This would definitely create tension and discomfort. This configuration of triadic sentiments is depicted in @fig-unbalanced-3.
- Mathematical Representation: P-O (-) * O-Q (+) * P-Q (+) = (-1) * (+1) * (+1) = -1.
- Implication: This also leads to tension in the triadic configuration due to the *similar* sentiments P and O have toward the third party Q.

4. **"An enemy of my enemy is also my enemy"**:

- Description: Person (P) dislikes Object (O) (a negative tie, -1), Object (O) dislikes Third Party (Q) (a negative tie, -1), and Person (P) also dislikes Third Party (Q) (a negative tie, -1). This configuration of triadic sentiments is depicted in @fig-unbalanced-4.
- Example: Imagine that you (P) hate the Yankees (O), but that you also (correctly) hate the Red Sox (Q). However, whenever you see the Yankees playing the Red Sox, it is difficult to root for one or the other because you want them *both* to lose!
- Mathematical Representation: P-O (-) * O-Q (-) * P-Q (-) = (-1) * (-1) * (-1) = -1.
- Implication: This is another type of unbalanced triad that creates tension, despite all parties having negative sentiments towards each other.

The basic idea of Balance Theory is that unbalanced configurations will tend to transition to balanced configurations. This transition occurs when the person (P) changes their sentiments toward O or Q to restore balance and alleviate internal tension. This can be done by P changing the sign or valence of the directed tie that goes from P to O or from P to Q. Balanced configurations, conversely, do not produce tension and are stable, yielding a positive product (+1) when multiplying their edge signs, and having either zero or two negative links.

```{r}
#| label: fig-unbalanced
#| fig-cap: "The four unbalanced triads according to balance theory."
#| fig-cap-location: bottom
#| fig-subcap:
#|   - ""
#|   - "" 
#|   - ""
#|   - ""
#| layout-ncol: 2
#| fig-height: 8
#| fig-width: 8
    plot_imbalanced_triad <- function(edge_signs, title) {
    # Define the directed edges: P->O, O->Q, P->Q
    edges <- data.frame(
        from = c("P", "O", "P"),
        to   = c("O", "Q", "Q"),
        sign = edge_signs
    )
    
    # Create the igraph object
    g <- graph_from_data_frame(edges, directed = TRUE)
    # Define a fixed layout matrix to keep the triangle uniform across all plots
    # Node order implicitly extracted by igraph: P (1), O (2), Q (3)
    layout_matrix <- matrix(c(
        0.000,  1.0,  # P at the top center
        -0.866, -0.5,  # O at the bottom left
        0.866, -0.5   # Q at the bottom right
    ), ncol = 2, byrow = TRUE)
    
    # Plot using ggraph
    ggraph(g, layout = layout_matrix) +
        geom_edge_link(aes(color = sign, label = sign),
                    arrow = arrow(length = unit(8, 'mm'), type = "closed"),
                    end_cap = circle(12, 'mm'),
                    start_cap = circle(8, 'mm'),
                    angle_calc = 'along',
                    label_dodge = unit(6, 'mm'),
                    label_size = 18,
                    width = 2,
                    show.legend = FALSE) +
        geom_node_point(size = 26, color = "tan2") +
        geom_node_text(aes(label = name), size = 12, fontface = "bold", color = "white") +
        scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
        labs(title = title) +
        theme_graph() + xlim(-1, 1) + ylim(-1, 1.25) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20, margin = margin(b = 10)))
    }
    # 1. One negative link: (+), (+), (-)
    plot_imbalanced_triad(c("+", "+", "-"), "1. A friend of a friend\nis an enemy")
    # 2. One negative link: (+), (-), (+)
    plot_imbalanced_triad(c("+", "-", "+"), "2. An enemy of a friend\nis a friend")
    # 3. One negative link: (-), (+), (+)
    plot_imbalanced_triad(c("-", "+", "+"), "3. A friend of an enemy\nis a friend")
    # 4. Three negative links: (-), (-), (-)
    plot_imbalanced_triad(c("-", "-", "-"), "4. An enemy of an enemy\nis an enemy")
```

## Cross-Pressure

In balance theory, Cross-pressure refers to a situation in which an individual experiences conflicting social relations, leaving them caught between two worlds [@davis63]. Cross-pressure arises when a person (P) has multiple social relationships that lead to inconsistent sentiments or attitudes towards a third party (Q).

Example of Cross-Pressure: A common example of cross-pressure occurs when a person (P) has two friends (O1 and O2) who hold different attitudes toward the same object (Q):

- Person (P): You.
- Object 1 (O1): Your friend.
- Object 2 (O2): Another friend.
- Third Party (Q): An object, or another person.

Imagine that:

- You (P) have a positive sentiment towards Friend 1 (O1) 
- You (P) have a positive sentiment towards Friend 2 (O2) 
- Friend 1 (O1) has a positive sentiment towards Q 
- However, Friend 2 (O2) has a negative sentiment towards Q

This structural arrangement creates a "cross-pressure" situation for you (P) because your two friends have opposing views on (Q). According to Balance Theory, this configuration will likely lead P to develop an "ambivalent attitude" toward Q, combining positive and negative sentiments (to balance the two friends).

The cross-pressure situation creates tension because:

- If P likes O1 and O1 likes Q, then to create balance, P should like Q.
- If P likes O2 and O2 dislikes Q, then to create balance, P should dislike Q.
- These conflicting pulls towards liking and disliking the same object create an unbalanced state.

A common example of cross-pressure today pertains to politics. For instance, many people (P) have family members (O1 and O2) with opposite attitudes to one of the major political parties (Q), which creates tension for P (imagine that your brother loves the democrats but your sister hates them). This is the configuration of triadic sentiments depicted in @fig-cross.

Here, balance theory predicts that, as long as this configuration remains, your attitude towards the democrats will *ambivalent* combining both positive and negative sentiments, and moving you towards the dreaded moderate "middle," or perhaps worse, making you an "independent" with no strong partisan affiliation. 

```{r}
#| label: fig-cross
#| fig-cap: "The cross-pressure situation."
#| fig-cap-location: bottom
#| fig-height: 8
#| fig-width: 8
# Define the nodes and directed edges for the cross-pressure example
edges <- data.frame(
  from = c("P", "P", "O1", "O2", "P"),
  to   = c("O1", "O2", "Q", "Q", "Q"),
  sign = c("+", "+", "+", "-", "?"),
  style = c("solid", "solid", "solid", "solid", "dashed") # Dashed for the unresolved tension
)

# Create the igraph object
g <- graph_from_data_frame(edges, directed = TRUE)

# Define a fixed layout matrix
# P at the top, O1 on the left, O2 on the right, Q at the bottom
layout_matrix <- matrix(c(
   0,  1,  # P (1)
  -1,  0,  # O1 (2)
   1,  0,  # O2 (3)
   0, -1   # Q (4)
), ncol = 2, byrow = TRUE)

# Plot using ggraph
ggraph(g, layout = layout_matrix) +
  geom_edge_link(aes(color = sign, label = sign, edge_linetype = style),
                 arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                 end_cap = circle(10, 'mm'),
                 start_cap = circle(9, 'mm'),
                 angle_calc = 'along',
                 label_dodge = unit(6, 'mm'),
                 label_size = 12,
                 width = 1.8,
                 show.legend = FALSE) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  # Custom colors: Green for +, Red for -, Goldenrod for the unresolved ?
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick", "?" = "goldenrod")) +
  scale_edge_linetype_identity() +
  labs(title = "Cross-Pressure Configuration",
       subtitle = "P is torn between O1 (who likes Q) and O2 (who dislikes Q)") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 11, margin = margin(b = 15)))
```

### Resolving Cross-Pressure

To achieve balance and alleviate the ambivalence generated by the cross-pressure situation, Balance Theory predicts that P will take steps to resolve the tension. This might involve:

1. Changing P's own attitude towards O1 or O2: P might change their attitude toward one of the friends (O1 or O2). For instance, P might decide they no longer like Friend 2 as much because of their dislike of Q, or vice versa. In the political polarization example, you might grow closer to your brother and colder towards your sister to keep a positive attitude towards the democrats. 
2. Influencing others' attitudes: P might try to convince either O1 or O2 to change their attitude toward Q. For example, P might try to persuade Friend 2 to like Q, or Friend 1 to dislike Q, to align the sentiments. In the political polarization example, you might try to convince your sister that the Democrats are actually the best (and keep liking them). 

## References {.unnumbered}
