# The "Big Three" Centrality Metrics {#sec-bigthree}

```{r setup, include=FALSE}
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

So what is centrality? In graph theory, a **centrality metric** is a summary index of a node's position in a graph, based on (sometimes weighted) *sums* or *averages* of one of three things [@borgatti_everett06]:

1. The *number* of **edges** that are directly **incident** upon that node.
1. The *length* of (usually **shortest**) **paths**^[Or trails, or walks.] that have that node as one of their **end nodes**.
1. The *proportion* of  (usually **shortest**) **paths**^[Or trails, or walks.] that have that node as one their **inner nodes**.^[If you need a refresher on these graph theory concepts, see the lesson indirect connectivity in social networks.]

The basic idea behind all centrality metrics is based on the **more/more principle**. This is the idea that *the higher any of the things above is for a node, then the more central that node is in the graph*. So a node with more nodes incident upon it (the size of the neighbor set) should be more central. In the same way, a node who can **reach** many other nodes via the smallest shortest paths, should be more central. Finally, a node standing in the middle of the largest percentage of paths between other nodes should be more central, and so forth. 

We can refer to all these things as **network goods**. The more/principle has a basic measurement implication, which is that centrality should never *decrease* when a node gains access to more network goods (e.g., one an additional connection) in the network. The **more/more centrality principle** was described by legendary social network theorist [**Linton Freeman**](https://www.socsci.uci.edu/newsevents/news/2018/2018-08-23-freeman.php) in a classic paper [@freeman79]. 

We have also seen that, soiologically, the notion of centrality is tied to such concepts as "status," "prestige," "importance," "influence," "popularity," "reachability," "brokerage," and many more. So the more central nodes have higher status, are more popular, influential, broker more deals, and so forth. However, these English language descriptors map onto distinct metrics. For instance, "popularity" clearly has to do with the number of connections a node has, what in @sec-degcent we referred to as degree centralities. *Brokerage*, on the other hand, has more to do with a node stand in-between paths linking others nodes, what in @sec-pathcent we referred to as **betweenness centrality**. Influence, on the other hand, seems to have more to do with the capacity to reach others, so it is more like **closeness** and so forth.

Accordingly, all these sociological concepts related to centrality have to be kept distinct from any one centrality metric (remember the **three-step shuffle** in social network analysis discussed earlier), because the same concept can be measured by more than one centrality metric, and the same metric can point to many distinct sociological concepts, depending on the context, the type of social ties the network is constructed from, and the way either "goods" (e.g., information, advice, support, etc.) or "bads" (e.g., disease, gossip, violence etc.) "flow" through the network [@borgatti05]. 

## The "big three" centrality metrics
Linton @freeman79 defined the "big three" classic centrality metrics, roughly corresponding to the extent that a node accumulates one of the three network goods mentioned above. 
- So the **degree centrality** metric deal with nodes that have more edges directly incident upon them [@nieminen74].
- The **closeness centrality** metric has to do with nodes that can reach more nodes via smallest **shortest paths** and thus accumulate as many of these paths in which they figure as the **origin node** as possible [@sabidussi66].
- Finally, the **betweenness centrality** metric has to do with a node's accumulation of the largest share of shortest paths in which they intermediate between two other nodes, and thus featuring them as one of the **inner nodes** in the paths between others [@freeman77]. 

Other centrality metrics can be seen as generalizations or special cases of any of these three basic notions [@borgatti05]. 

Importantly, Freeman showed that the big three centrality metrics reach their theoretical *maximum* for the central node in a **star graph**, such as the one shown in @fig-star). 


```{r}
#| label: fig-star
#| fig-cap: "A star graph"
#| fig-cap-location: margin
#| fig-width: 7
#| fig-height: 7

    gr <- create_star(7) %>% 
       mutate(name = toupper(letters[1:7]))
    p <- ggraph(gr, layout = 'star') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() + ylim(-1.1, 1.1)
    p
```

A star graph is a graph containing a **central** or **inner** node (in @fig-star, node *A*), who is connected to all the other nodes in the graph, called the **satellite** or **outer** nodes (in @fig-star, nodes *B* through *F*). These nodes in contrast have only *one* connection and that is to the central node, none among themselves. 

Because of these restrictions, it is easy to see that if $G = (V, E)$ is a star graph of order $n$, then we know that that graph size $m = |E|$ (the size of the edge set), has to be $n-1$. So in the example shown in @fig-star, $n =7$ and $m = n-1 = 7-1=6$. Neat!

## The Big Three Centralities in the Star Graph
Degree, Closeness, and Betweenness centralities have an interesting property that provides a conceptual connection between them [@freeman79]. Consider the star graph shown in @fig-star with central node *A*. The degree, closeness, and betweenness centralities of the different nodes are shown in @tbl-startab). 

Of course, by definition, we know beforehand that the central node in a star graph has to have the highest degree, since the degree of peripheral nodes is fixed to one and the degree of the central node is always $n-1$, where $n$ is the graph order. 

However, note also that the central node has to have the highest closeness, since it is connected by a path of length one (an edge) to every peripheral node, but each peripheral node can only reach other peripheral nodes in the graph by a path of length two. They are farther away from other nodes than the central node. 

Finally, note that the central node in the star will also always have the highest betweenness because each of the paths of length two connecting every pair of peripheral nodes to one another *has* to include the central node as an inner node in the path. So it serves as the intermediary between any communication between peripheral nodes. 

```{r startab}
#| label: tbl-startab 
#| tbl-cap: "Centralities in a star graph of order 7."
#| tbl-cap-location: margin

    gr <- create_star(7) %>% 
       mutate(name = toupper(letters[1:7]))
    d <- degree(gr)
    c <- distances(gr)
    c <- t(round(length(c)*1/rowSums(c),1))
    b <- t(round(betweenness(gr), 1))
    c.tab <- rbind(d, c, b)
    rownames(c.tab) <- c("Degree", "Closeness", "Betwenness")
    kbl(c.tab, format = "pipe", align = "c")
```

The mathematical sociologist **Linton @freeman79** thus thinks that the "big three" centrality measures are the big three precisely because they are maximized for the central node in a star graph. 


## References {.unnumbered}

<!-- 




## K-Path Centrality

Sade (1989) introduced the idea of k-path centrality. Logically, nodes which are near lots of other nodes, even they are not directly connected, are more closely connected to the network than nodes which are loosely connected through other nodes. It is useful for researchers to be able to examine this easily if their research question calls for measuring this.

**K-Path Centrality**
$$ 
$$
C_k(j) = \sum_{j = 1}^{n}A^k_{ij}
$$
$$

To do this, we multiply the matrix by itself equivalent to the path length we are interested in solving for. For example, looking for a k-path of k=2 would find all nodes two steps away. To solve, one would simply square the matrix. If you remember from chapter 3, this is the same thing as a reachability matrix! Now, we've simply shifted our thinking about it, to realize that it is interpretable as a measure of centrality. With some manipulation of the graph, we can easily calculate how far away a given node is from alters to which they are not directly connected.  

Note that k-path centrality is actually a generalization of degree centrality. Note that this formula reduces to equivalence wtih degree centrality for k=1, as k=1 is simply the condition which asks if two nodes are adjacent or not. 

```{r kpathcent, echo = FALSE, warning = FALSE, message=FALSE, fig.height=5}
from <- c("A", "B", "B", "C", "C", "D", "D")
to <- c("B", "C", "D", "D", "E", "E", "F")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'star')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 5)
p <- p + theme_graph() + theme(plot.margin = unit(c(0,0,0,0), "cm"))
p <- p + expand_limits(x = c(-1.5,1.5), y = c(-1.5, 1.5))
p
```
**Figure 4.1**

Figure 4.1 provides a simple example of how k-path centrality of two provides insights that degree centrality alone fails to provide. Nodes A and F both have a degree centrality of one. However, node F has a 2-path centrality of 3, while node A has a 2-path centrality of 2. We might then think of node F as better connected to the network than node A.   

-->