# Directed Structural Balance {#sec-dirbalance}

Social relations are rarely completely symmetrical. In real-world social networks, one individual might extend a positive feeling (e.g., liking or friendship) to another who does not reciprocate it, or who might even hold a negative sentiment in return. To study structural balance under these more realistic conditions, we must extend the theory of structural balance from undirected graphs to **signed directed graphs** (or **signed digraphs**).

This extension, first formalized by Dorwin Cartwright and Frank Harary [-@cartwright_harary56] and detailed by Stanley Wasserman and Katherine Faust [-@wasserman_faust94], requires that we adapt our definitions of paths, cycles, and balance to account for the directed nature of social ties.

## Why Directed Cycles are Too Restrictive

In a standard directed graph (digraph), a **directed cycle** is a closed sequence of nodes and directed edges (arcs) in which all arrows point in the same direction (e.g., $n_1 \rightarrow n_2 \rightarrow n_3 \rightarrow n_1$). For a detailed refresher on directed cycles and how they are defined, see @sec-dircycles.

If we restricted our study of balance to only directed cycles, we would miss most instances of psychological and social tension in a network. In sentiment networks, social tension can exist even if there is no perfect, closed directed loop of arrows. 

To illustrate, imagine the following three-person configuration (depicted in @fig-unbalanced-digraph):
- **You** ($n_1$) like **Alex** ($n_3$).
- **You** ($n_1$) also like **Jordan** ($n_2$).
- **Alex** ($n_3$) strongly dislikes **Jordan** ($n_2$).

Mathematically, this configuration does not contain a directed cycle because the arcs $n_1 \rightarrow n_2$ and $n_3 \rightarrow n_2$ both point toward Jordan ($n_2$). There is no way to follow the arrows in a single direction to return to your starting node.

Yet, despite the lack of a directed cycle, you still experience a palpable **social tension**. Your friend Alex dislikes Jordan, but you are friendly with both. You realize that your friendliness with Jordan is inconsistent with your friend Alex's unfriendliness with Jordan. This psychological pressure is exactly what Heider's balance theory predicts, regardless of the direction of the arrows.

## Semipaths and Semicycles

To resolve this limitation and formally define balance in directed signed graphs, Cartwright and Harary [-@cartwright_harary56] introduced the concepts of **semipaths** and **semicycles**. For their initial introduction in the context of weakly connected graphs, see @sec-indirectconnectivity.

- **Semipath:** A sequence of distinct nodes and arcs where we *ignore* the direction of the arrows. We only care that an arc exists between adjacent nodes in the sequence, whether it points from the previous node to the next or vice versa.
- **Semicycle:** A closed semipath in which the first and last nodes are identical. In other words, it is a cycle in the underlying undirected graph.

In our high school café example, the sequence $n_1, n_2, n_3, n_1$, along with the arcs connecting them, forms a **semicycle** of length 3, even though it is not a directed cycle.

### Defining the Sign of a Semicycle

Just as with undirected cycles (covered in @sec-sbalance), the **sign of a semicycle** is determined by multiplying the signs of all the arcs that compose it:

$$\text{Sign of Semicycle} = \prod \text{Signs of Arcs}$$

Alternatively, a semicycle is **balanced** if it has an **even number of negative signs** (or zero negative signs), resulting in a positive product ($+1$). It is **unbalanced** if it has an **odd number of negative signs**, resulting in a negative product ($-1$).

With these concepts, we can state the fundamental definition of balance in directed networks:

::: {.callout-note}
### Definition: Balanced Signed Digraph
A signed directed graph (digraph) is **balanced** if and only if **all of its semicycles have positive signs** (i.e., contain an even number of negative arcs).
:::

---

## Graphical Examples of Directed Balance and Imbalance

Let's look at the two primary configurations representing triadic states in signed directed graphs.

### 1. Unbalanced Signed Digraph (Tension-Producer)

The first example (@fig-unbalanced-digraph) represents the classic unbalanced directed triad described by Wasserman and Faust [-@wasserman_faust94] (equivalent to Heider's triad with one negative tie). 

```{r}
#| label: fig-unbalanced-digraph
#| fig-cap: "An unbalanced signed directed triad (Wasserman & Faust, Fig. 6.4)."
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 6

library(igraph)
library(ggraph)

# Define nodes n1, n2, n3
nodes <- data.frame(name = c("n1", "n2", "n3"))

# Define directed edges with signs: n1->n3 (+), n1->n2 (+), n3->n2 (-)
edges <- data.frame(
  from = c("n1", "n1", "n3"),
  to   = c("n3", "n2", "n2"),
  sign = c("+", "+", "-"),
  type = c("Positive (+)", "Positive (+)", "Negative (-)")
)

# Create signed digraph
g <- graph_from_data_frame(edges, directed = TRUE, vertices = nodes)

# Define triangle layout
layout_matrix <- matrix(c(
   0.0,  1.0,  # n1 (top center)
  -0.8, -0.5,  # n2 (bottom left)
   0.8, -0.5   # n3 (bottom right)
), ncol = 2, byrow = TRUE)

ggraph(g, layout = layout_matrix) +
  geom_edge_link(
    aes(color = type, label = sign),
    arrow = arrow(length = unit(6, 'mm'), type = "closed"),
    end_cap = circle(10, 'mm'),
    start_cap = circle(8, 'mm'),
    angle_calc = 'along',
    label_dodge = unit(5, 'mm'),
    label_size = 14,
    width = 1.8,
    show.legend = FALSE
  ) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  scale_edge_color_manual(values = c("Positive (+)" = "forestgreen", "Negative (-)" = "firebrick")) +
  theme_graph() + 
  coord_cartesian(clip = "off")
```

The semicycle $n_1 n_2 n_3 n_1$ in @fig-unbalanced-digraph has a sign product of $(+1) \times (+1) \times (-1) = -1$. Because this semicycle is negative, the entire digraph is **unbalanced**, and the actors will experience a psychological drive to change their sentiments and resolve the tension.

### 2. Balanced Signed Digraph (Stable Equilibrium)

The second example (@fig-balanced-digraph) represents a balanced directed triad ("An enemy of an enemy is a friend"). 

- $n_1$ likes $n_2$ ($+$)
- $n_1$ dislikes $n_3$ ($-$)
- $n_3$ dislikes $n_2$ ($-$)

```{r}
#| label: fig-balanced-digraph
#| fig-cap: "A balanced signed directed triad."
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 6

# Define directed edges with signs: n1->n2 (+), n1->n3 (-), n3->n2 (-)
edges_bal <- data.frame(
  from = c("n1", "n1", "n3"),
  to   = c("n2", "n3", "n2"),
  sign = c("+", "-", "-"),
  type = c("Positive (+)", "Negative (-)", "Negative (-)")
)

g_bal <- graph_from_data_frame(edges_bal, directed = TRUE, vertices = nodes)

ggraph(g_bal, layout = layout_matrix) +
  geom_edge_link(
    aes(color = type, label = sign),
    arrow = arrow(length = unit(6, 'mm'), type = "closed"),
    end_cap = circle(10, 'mm'),
    start_cap = circle(8, 'mm'),
    angle_calc = 'along',
    label_dodge = unit(5, 'mm'),
    label_size = 14,
    width = 1.8,
    show.legend = FALSE
  ) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  scale_edge_color_manual(values = c("Positive (+)" = "forestgreen", "Negative (-)" = "firebrick")) +
  theme_graph() + 
  coord_cartesian(clip = "off")
```

The semicycle $n_1 n_2 n_3 n_1$ in @fig-balanced-digraph contains exactly **two negative links**, resulting in a positive sign product: $(+1) \times (-1) \times (-1) = +1$. Because the semicycle is positive, the digraph is in **stable equilibrium**, and there is no social pressure on any actor to alter their relationships.

---

## Checking for Balance in Signed Digraphs

Checking for balance in large undirected signed graphs can be done using powers of the adjacency matrix (as described in @sec-sbalance). However, doing this for a **directed signed graph** is more complex due to the presence of asymmetric ties and the requirement to check *semicycles* rather than standard directed cycles.

To check for balance in a signed digraph, Harary, Norman, and Cartwright [-@Harary65] proposed constructing a **symmetric valency matrix** $\mathbf{V}$ from the signed adjacency matrix $\mathbf{X}$.

1. **Constructing the Valency Matrix:** For each pair of nodes $i$ and $j$, we define the entry $v_{ij}$ in $\mathbf{V}$ based on the presence and signs of the reciprocal arcs between them:
   - If only $i \rightarrow j$ exists with sign $s$, then $v_{ij} = v_{ji} = s$.
   - If both $i \rightarrow j$ and $j \rightarrow i$ exist with the same sign $s$, then $v_{ij} = v_{ji} = s$.
   - If both exist with different signs, or if no tie exists, then special rules are used to capture the mixed-sentiment state (e.g., using algebraic symbols).

2. **Raising the Valency Matrix to Powers:** We can then analyze the diagonal of the powers of this valency matrix ($\mathbf{V}^p$) up to $p = g$ (the size of the network). 
   - If the diagonal entries of $\mathbf{V}^p$ are positive for all $p$, then all semicycles in the digraph are positive, and the directed network is **perfectly balanced**.

---

## Directed Balance, Hierarchies, and Status

While Heider's original balance theory was designed for purely horizontal, symmetrical sentiment structures, directed balance theory serves as a vital bridge to study **vertical status hierarchies**.

As we will see in our discussion of **Ranked Clusterability** (Davis & Leinhardt, covered in @sec-ranked), real-world groups often have hierarchies. In a ranked clusterable network:
- Positive directed ties tend to point **up** the hierarchy (lower-status individuals choosing higher-status individuals as friends).
- Negative or null directed ties tend to point **down** the hierarchy.

By using semicycles and directed signed networks, we can model how these psychological preferences for balance interact with sociological realities of dominance, prestige, and status (which we will cover in @sec-dominance).
