# Structural Holes and Brokerage {#sec-sht}

The theory of structural holes (SHT), largely associated with the work of Ronald @burt95, is a social network theory that explains how certain network structures can be advantageous to individuals. It describes a form of **social capital** derived from access to unique or valuable information. In this way, SHT builds upon and extends Granovetter's [-@granovetter73] Strength of Weak Ties theory. While Granovetter's theory emphasizes the value of weak ties for accessing novel information, Burt's SHT focuses on how individuals can strategically leverage gaps in social networks to gain advantages and control information flow.

## Definition and Core Concept

A **structural hole** refers to the absence of a link between two parts of a social structure. In graph theoretic terms, these are "two-paths that are not full triangles". For instance, if A is connected to B and C but B and C are not connected to one another, A bridges a structural hole between B and C. Essentially, structural holes represent gaps in the flow of information or control within a network. These "non-ties" are significant in creating opportunities for exchange. 

As we saw in @sec-egonets, when an individual's network is characterized by **clustering** (where contacts are connected to one another), the **effective size** of their network is lower than its observed size, indicating that the ego is connected to redundant contacts. In SHT, this is referred to as a network with high **constraint**. 

The concept of network constraint is the inverse of having structural holes. An individual who has high constraint lives in a social world in which most or all of their contacts are also connected to one another. This also means that the individual is heavily influenced by the strong expectations of a single group and has to rely on others for access to information, rather than being an independent source.

Conversely, low constraint ego networks are those that contain many structural holes (contacts not connected to one another), which facilitate intermediation through other individuals (alters). Networks rich in structural holes, often described as **open networks**, offer increased opportunities to acquire diverse information and ideas, as well as greater mediation in social relationships.

SHT fundamentally aims to explain how to configure one's network to control information. Individuals who position themselves to "fill" these holes act as **brokers**. An example of a graph with a structural hole would show two separate clusters that are not connected except via the broker, as in @fig-sh. This lack of connection is significant because it prevents information or content from passing from one part of the network (blue nodes) to another, except (green nodes) as mediated by the broker (red node). 

```{r}
#| label: fig-sh
#| fig-cap: "An actor (red) bridging a structural hole between two groups (blue and green)."
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 6
library(igraph)
library(ggraph)
# 1. Define the Nodes
# We have one broker and two distinct groups
nodes <- data.frame(
  name = c("Broker", "A1", "A2", "A3", "A4", "B1", "B2", "B3", "B4"),
  role = c("Broker", rep("Group A", 4), rep("Group B", 4))
)

# 2. Define the Edges
# Fully connect Group A internally to show dense cohesion
edges_A <- subset(expand.grid(from = paste0("A", 1:4), to = paste0("A", 1:4)), from != to)
edges_A$type <- "Internal Tie"

# Fully connect Group B internally to show dense cohesion
edges_B <- subset(expand.grid(from = paste0("B", 1:4), to = paste0("B", 1:4)), from != to)
edges_B$type <- "Internal Tie"

# Create the bridging ties that span the structural hole
edges_bridge <- data.frame(
  from = c("Broker", "Broker"),
  to   = c("A1", "B1"),
  type = rep("Bridging Tie", 2)
)

# Combine all edges
edges <- rbind(edges_A, edges_B, edges_bridge)

# 3. Create the igraph object
g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)

# Simplify graph to remove duplicate undirected edges generated by expand.grid
g <- simplify(g, remove.multiple = TRUE, remove.loops = TRUE, edge.attr.comb = "first")

# 4. Plot using ggraph
# The Fruchterman-Reingold ('fr') layout naturally pushes disconnected components apart,
# illustrating the "hole" perfectly, held together only by the broker.
set.seed(42) # For reproducible layout
ggraph(g, layout = 'fr') + 
  geom_edge_link(aes(color = type, edge_width = type), alpha = 0.8) +
  geom_node_point(aes(color = role, size = role)) +
  geom_node_text(aes(label = name), vjust = 0.5, hjust = 0.5, 
                 color = "white", fontface = "bold", size = 3.5) +
  
  # Custom aesthetics to highlight the broker and the bridging ties
  scale_edge_color_manual(values = c("Bridging Tie" = "firebrick", "Internal Tie" = "steelblue")) +
  scale_edge_width_manual(values = c("Bridging Tie" = 1.2, "Internal Tie" = 1.2)) +
  scale_color_manual(values = c("Broker" = "firebrick", "Group A" = "steelblue", "Group B" = "forestgreen")) +
  scale_size_manual(values = c("Broker" = 16, "Group A" = 11, "Group B" = 11)) +
  
  # Theming and Labels
  labs(title = "Spanning a Structural Hole",
       subtitle = "The Broker acts as the sole conduit of information between Group A and Group B.",
       color = "Actor Role", 
       size = "Actor Role", 
       edge_color = "Tie Type", 
       edge_width = "Tie Type") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
        plot.subtitle = element_text(size = 11, hjust = 0.5, margin = margin(b = 15)),
        legend.position = "bottom",
        legend.box = "vertical")
```

## Benefits of Structural Holes

Structural holes provide two primary types of benefits to the individuals who bridge them:

1. **Passive Exposure Effect**: Brokers gain access to diverse information from different parts of the network that would otherwise be inaccessible. Because the groups they connect to are not directly linked, the information flowing through the broker is likely to be novel and non-redundant. This contrasts with strong ties, which often provide redundant information because closely connected individuals tend to know the same people and share similar information. This strategic position allows them to synthesize diverse information from various sources to create novel combinations and generate better ideas.

2. **Active Control Effect**: This benefit grants the broker power to decide whether certain individuals receive specific information, influence how others perceive various individuals, and potentially play people against one another. In the same way, individuals occupying positions that bridge structural holes are less constrained by the norms and expectations of a single group. They are freer to navigate the social structure, allowing them to adapt their roles or presentations to different audiences.

Empirical studies, often conducted within corporate settings, suggest tangible benefits for individuals with networks rich in structural holes (i.e., low constraint). These individuals tend to receive higher salaries, experience faster promotions, and obtain better performance evaluations. They also demonstrate an ability to generate better ideas by combining disparate elements from various sources to create novel combinations [@burt04].

In essence, Burt's Structural Holes theory emphasizes that having diverse contacts who are not connected to each other (i.e., operating across structural holes) provides unique benefits in terms of information access, control, and ultimately, social capital.

## Types of Brokerage

Brokerage can occur both within and between groups. Analysts have identified distinct types of brokerage that can occur in social networks [@gould_fernandez89]:

1. *Within-Group Brokers*: These individuals operate within a single defined group.
    - **Coordinators**: A coordinator connects two people within the same group who would otherwise not be connected. The coordinator themselves is also a member of this group. This type of brokerage helps facilitate connections and potentially new syntheses within an existing cluster.
    - **Itinerant Brokers**: Similar to coordinators, an itinerant broker connects two individuals within a group, but the broker themselves is not part of that specific group. This role suggests a more external, consultative function in facilitating internal group ties.
2. Between-Group Brokers: These individuals span across different, often disconnected, groups.
    - **Gatekeepers**: A gatekeeper grants access from someone outside their group to someone within their own group. They control the flow of information or resources into their group.
    - **Representatives**: A representative connects people from their own group to individuals outside their group. They act as an outward-facing link for their group.
    **Liaisons**: A liaison connects three actors, each belonging to a different group. This is the most expansive form of between-group brokerage, involving connections across multiple distinct social spheres.

These types of brokerage can be measured by counting the number of triads of each type, normalizing for the probability of such counts arising from random matching, and requiring the sorting of nodes into groups.

```{r}
#| label: fig-gf
#| fig-cap: "The four types of brokerage according to Gould and Fernandez."
#| fig-cap-location: bottom
#| fig-subcap:
#|   - ""
#|   - "" 
#|   - ""
#|   - ""
#| layout-ncol: 2
#| fig-height: 8
#| fig-width: 8

# Helper function to generate a plot for a specific Gould & Fernandez brokerage role
plot_brokerage <- function(node_groups, title) {
  
  # Define the directed edges: A -> B -> C (B is the Broker)
  edges <- data.frame(
    from = c("A", "B"),
    to   = c("B", "C")
  )
  
  nodes <- data.frame(
    name = c("A", "B", "C"),
    group = node_groups
  )
  
  # Create the igraph object
  g <- graph_from_data_frame(edges, directed = TRUE, vertices = nodes)
  
  # Define a fixed layout matrix for a uniform V-shape
  # A (bottom left), B (top center), C (bottom right)
  layout_matrix <- matrix(c(
    -1.0,   0.0,  # A
     0.0,   0.8,  # B
     1.0,   0.0   # C
  ), ncol = 2, byrow = TRUE)
  
  # Plot using ggraph
  ggraph(g, layout = layout_matrix) +
    geom_edge_link(arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                   end_cap = circle(12, 'mm'),
                   start_cap = circle(10, 'mm'),
                   width = 2, 
                   color = "steelblue") +
    geom_node_point(aes(color = group), size = 26) +
    geom_node_text(aes(label = name), size = 12, fontface = "bold", color = "white") +
    # Assign distinct colors to different groups
    scale_color_manual(values = c("Group 1" = "tan2", 
                                  "Group 2" = "firebrick", 
                                  "Group 3" = "forestgreen")) +
    labs(title = title) +
    theme_graph() + xlim(-1.2, 1.2) + ylim(-1, 1.25) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20, margin = margin(b = 10)),
          legend.position = "none") # Hide legend to keep individual plots clean
}

# 1. Representative: Source (A) and Broker (B) are in the same group. Target (C) is in a new group.
# B represents their group to the outside world.
plot_brokerage(c("Group 1", "Group 1", "Group 2"), 
                     "1. Representative\n(A & B in Grp 1 -> C in Grp 2)")

# 2. Gatekeeper: Source (A) is in an outside group. Broker (B) and Target (C) are in the same group.
# B controls outside access into their group.
plot_brokerage(c("Group 1", "Group 2", "Group 2"), 
                     "2. Gatekeeper\n(A in Grp 1 -> B & C in Grp 2)")

# 3. Itinerant / Consultant: Source (A) and Target (C) are in the same group. Broker (B) is an outsider.
# B facilitates contact between two members of the same group.
plot_brokerage(c("Group 1", "Group 2", "Group 1"), 
                     "3. Itinerant (Consultant)\n(A & C in Grp 1, B in Grp 2)")

# 4. Liaison: Source (A), Broker (B), and Target (C) are all in different groups.
# B links two completely separate groups together without belonging to either.
plot_brokerage(c("Group 1", "Group 2", "Group 3"), 
                     "4. Liaison\n(All actors in different groups)")
```
### Forms of Brokering Process

Beyond the structural roles, brokerage can also be understood as a *process* with different forms of action [@obstfeld14]. Structural holes represent opportunities for brokerage, but realizing these opportunities requires motivation and action. Three forms of brokering are identified:

1. **Conduit Forms**: This involves the straightforward transfer of information, ideas, or solutions. It's about problem-solving through synthesis. Structural hole brokerage is a primary example of conduit brokerage. This can also occur within groups, leading to "within group syntheses," even without a "hole" in the traditional sense.
2. **Tertius Gaudens** ("the third who benefits"): In this form, the broker exploits the relationship between two other parties (alters) by keeping them disconnected or actively encouraging conflict between them ("divide and conquer"). This grants the broker control over their interaction.
3. **Tertius Lungens** ("the third who joins"): This form describes a broker who actively connects or facilitates interactions between alters. This can involve closing off a structural hole by bringing previously disconnected individuals into direct contact, or by coordinating interactions among those who are already connected but not necessarily interacting efficiently.

## References {.unnumbered}
