# Graphs and their Subgraphs {#sec-subgraphs}
```{r setup, include=FALSE}
      library(ggraph)
      library(tidygraph)
      library(igraph)
```

## Graphs and Subgraphs
Consider the graph shown in @fig-subgraph-1. If all the actors that you are interested in studying are included here, we would refer to it as the **whole network**. However, sometimes, even when we collect data on a large number of actors, we may be interested in analyzing not the whole network, but only some parts of it. How do we do that?

```{r}
#| label: fig-subgraph-1
#| fig-cap: "Original graph for subgraph examples."
#| fig-cap-location: bottom

set.seed(367)
gr <- play_gnm(n = 10, m = 20, directed = FALSE, loops = FALSE) %>% 
  mutate(name = LETTERS[1:10])
l <- ggraph(gr, layout = 'kk')
l <- as.matrix(l$data[, 1:2])
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p1 <- p + theme_graph()  + coord_cartesian(clip = "off")
p1
```

Well, good thing a graph is actually a pair of sets. If you remember your high school set theory, you can always take a set and consider only a **subset** of the original members.

Since graphs are sets, we can do the same thing. A subset of the original nodes (or edges) of a graph is called a **subgraph**. So if $G =\{E,V\}$ is the original graph, the subgraph $G' = \{E',V'\}$ is a subset of $G$, which is written $G' \subset G$, with the understanding that $E' \subset E$ and $V' \subset V$. In mathematics, "$\subset$" is the symbol for subset. Thus, $A \subset B$ is read as "set A is a subset of set B."

For instance, let us say we are interested in just analyzing actors *A*, *B*, *C*, *D*, and *E* in the graph shown in @fig-subgraph-1. They seem to be a close-knit group. In that case, as noted earlier, if we call the original graph $G$ with vertex and edge sets $\{E, V\}$we can define a new subgraph $G'$, whose node subset $V'$ only includes the actors we are interested in studying, in this case $V' = \{A, B, C, D, E\}$, where $V' \subset V$.

The subgraph $G'$ is shown in @fig-subgraph-2. It looks exactly like we wanted, capturing the relations between an interconnected subgroup of actors in the original graph. Note that the edge set of the subgraph $E'$ only includes those edges that are **incident** to the other nodes in the subgraph (as defined in @sec-graphtheory) and omits those in the original graph that are incident to nodes that are not in the subgraph, so $E' \subset E$. As we will see in a later lesson, well-connected subgroups of actors in an original graph are called **cohesive subsets**.

```{r}
#| label: fig-subgraph-2
#| fig-cap: "A vertex-induced subgraph based on nodes {A, B, C, D, E}."
gr2 <- induced_subgraph(gr, toupper(letters[1:5]))
p <- ggraph(gr2, layout = l[1:5, ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p2 <- p + theme_graph() + coord_cartesian(clip = "off") 
p2
```

## Vertex and Edge-Induced Subgraphs
For any graph, we can define a subgraph based on *any* old random subset of the original node set. It is completely up to us. For instance, we could define a new subgraph $G''$ of the original graph shown in @fig-subgraph-1, that includes the node set $V'' = \{D, E, G, I\}$. That is shown in @fig-subgraph-3. That subgraph is weird (composed of two standalone connected dyads) and probably not very useful, but it is a subgraph of the original graph anyways!

```{r}
#| label: fig-subgraph-3
#| fig-cap: "A vertex-induced subgraph based on nodes {D, E, G, I}."
gr3 <- induced_subgraph(gr, c("G", "D", "E", "I"))
p <- ggraph(gr3, layout = l[c(4, 5, 7, 9), ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p3 <- p + theme_graph() + coord_cartesian(clip = "off") 
p3
```

Just like we can define subgraphs based on the node set of a graph, we can define subgraphs based on subsets of the original edge set. For instance, we could pick the edges $E' = \{AB, AC, BJ\}$ and define a subgraph based on them, as shown in @fig-subgraph-edge-induced. This necessarily includes the node set $V' = \{A, B, C, J\}$.

```{r}
#| label: fig-subgraph-edge-induced
#| fig-cap: "An edge-induced subgraph based on edges {AB, AC, BJ}."
# The edges AB, AC, BJ correspond to edge indices 1, 2, and 16 from the original `gr`
gr_edge_induced <- gr |>
  activate(edges) |>
  slice(c(1, 2, 16)) |>
  activate(nodes) |>
  filter(!node_is_isolated())

# We need to get the correct node indices for the layout and keep it as a matrix
edge_induced_nodes <- V(gr_edge_induced)$name
edge_induced_layout <- l[which(LETTERS[1:10] %in% edge_induced_nodes), ]

ggraph(gr_edge_induced, layout = edge_induced_layout) +
  geom_edge_link(color = "steelblue", width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```

When a subgraph is defined by selecting a subset of nodes to retain (the common case), it is called a **vertex-induced subgraph** of the original graph, as in @fig-subgraph-2. When a subgraph is defined by picking a subset of edges from the original graph to keep, it is called (you guessed it) an **edge-induced subgraph** of the original graph.

## Creating Subgraphs by Deleting Nodes and Edges

Another way to think about creating subgraphs is by *deleting* elements from the original graph. This approach is common in social network analysis when we want to see what a network would look like if certain actors or relationships were removed, which can help reveal how important they are for holding the network together.

### Vertex-Deleted Subgraphs

It is possible to define a subgraph by deleting nodes. This is written as $G' = G - \{a, b, c\}$, where $\{a, b, c\}$ is the set of nodes being deleted. When you delete nodes, a critical rule applies: **all edges incident to those nodes are automatically removed as well**.

For example, starting with the graph in @fig-subgraph-1, if we delete the node set $\{F, G, H, I, J\}$, the resulting subgraph is identical to the one shown in @fig-subgraph-2. This shows that a **vertex-deleted subgraph** can be equivalent to a **vertex-induced subgraph**. The action of deleting $\{F, G, H, I, J\}$ *induces* a subgraph on the remaining nodes $\{A, B, C, D, E\}$.

Two special cases of vertex deletion are:
- The **null graph**, which results from removing *all* nodes from the original graph.
- The **singleton graph** (or **trivial graph**), which results from removing all nodes *except for one*.

### Edge-Deleted Subgraphs

Just as we can delete nodes, we can also create subgraphs by removing edges. This is written as $G' = G - \{e_1, e_2, ...\}$, where $\{e_1, e_2, ...\}$ is the set of edges being deleted. For instance, the **edge-deleted subgraph** in @fig-subgraph-4 results from removing the six edges $\{AB, AE, AC, CI, CE, GJ\}$ from the graph in @fig-subgraph-1.

When you delete edges, the nodes that were connected by those edges remain in the graph. A subgraph created by removing *only* edges while leaving all original nodes intact is called a **spanning subgraph**.

A special case of edge deletion is the **empty graph**, which results from removing *all* edges from the original graph, leaving only the set of isolated nodes.

```{r}
#| label: fig-subgraph-4
#| fig-cap: "An edge-deleted subgraph, which is also a spanning subgraph."
gr4 <- gr - edge("A|B") - edge("G|J") - edge("C|I") - edge("A|E") - edge("A|C") - edge("C|E")
p <- ggraph(gr4, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p4 <- p + theme_graph() + coord_cartesian(clip = "off") 
p4
```

As we will see later, subgraphs (as well as vertex and edge deletion) are a useful concept for discussing levels at an "in-between" level, above the node level but "below" the whole network level: subgroups. However, subgraphs are also useful for network concepts at the node level, because there is a special type of subgraph, called the **ego graph** that is defined by picking a central node and the nodes that are connected to it, along with the edges connecting the nodes surrounding ego. 
