# Directed Graphs {#sec-directedgraphs}

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    library(ggplot2)
    library(ggraph)
    library(tidygraph)
    library(igraph)
    library(kableExtra)
```

## Introduction to Directed Graphs (Digraphs)

While simple undirected graphs are useful for representing symmetric relationships (where connection is naturally mutual), many social relationships are inherently directional (see @sec-ties). In these relationships, one member of a pair can claim a tie with the other, but the other person may or may not send the same tie back. Examples include giving advice, sending emails, lending money, or nominating friends.

To represent directional relations, we use **directed graphs** (also called **digraphs**). In a directed graph, the edges are *ordered pairs* of nodes, meaning the order in which we list the nodes in an edge matters. The connection goes from a **source node** (the sender) to a **destination node** (the receiver). Therefore, the edge $AB$ is a different entity from the edge $BA$. 

In directed graphs, simple undirected lines are replaced by directed arrows (often called **arcs**). When a digraph has no loops (edges connecting a node to itself) and at most one directed edge connecting a sender to a receiver in each direction, it is called a **simple directed graph**.

@fig-directed shows a point-and-line diagram of a digraph representing a simple directed social network.

```{r}
#| label: fig-directed
#| fig-cap: "A simple directed graph."
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "A", "B", "B", "B", "D", "E", "E", "F", "G", "G")
  to <- c("B", "F", "A", "D", "C", "B", "C", "D", "A", "F", "D")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

In directed networks, reciprocity is a major analytical focus. For some ties (such as liking, trusting, or friendship), reciprocity is binary; it either exists or it does not. If @fig-directed were an **advice network** [@cross01], we could say that *H* seeks advice from *D*, but *D* does not seek advice from *H*. This lack of reciprocity is common in formal hierarchies, where advice-seeking often flows upward to more experienced or authoritative actors, reflecting power differentials. 

---

## Node Neighborhoods in Directed Graphs

Just like in undirected (simple) graphs, each node in a directed graph has a **node neighborhood**. However, because each node can be the **source** or **destination** for an asymmetric edge, we have to differentiate the neighborhood of a node depending on whether the node is the **sender** or the **recipient** of a given link.

### In-Neighborhoods
We say that a node *j* is an **in-neighbor** of a node *i* if there is a directed link with *j* as the source and *i* as the destination node ($j \rightarrow i$). For each node, the full set of in-neighbors forms the **in-neighborhood** of that node. This is written:

$$
  N^{in}(v)
$$

where $v$ is the label corresponding to the node. For instance, in @fig-directed, the node set $N^{in}(D) = \{B, E, G\}$ is the in-neighborhood of node *D*.

### Out-Neighborhoods
In the same way, we say that a node *i* is an **out-neighbor** of a node *j* if there is a directed link with *j* as the source and *i* as the destination ($j \rightarrow i$). For each node, the full set of out-neighbors defines the **out-neighborhood** of that node. This is written:

$$
  N^{out}(v)
$$

where $v$ is the label corresponding to the node. For instance, in @fig-directed, the node set $N^{out}(B) = \{A, C, D\}$ is the out-neighborhood of node *B*.

Nodes will only appear in both the in- and out-neighborhood sets when there are **reciprocal** or mutual ties between the nodes. For instance, in @fig-directed, the out-neighborhood of node *F* is $\{A\}$ and the in-neighborhood is $\{A, G\}$. Here, node *A* appears in both the in- and out-neighborhood sets because *A* has a reciprocal tie with *F*.

### Neighborhood Intersection in Directed Graphs

Just as in the undirected case, the sets of neighbors for any two nodes in a directed graph can share members in common. However, because directed neighborhoods are differentiated by directionality, we can examine two types of intersection.

**In-Neighborhood Intersection** ($N^{in}(u) \cap N^{in}(v)$): Represents the set of common "nominators" or senders—nodes that direct an edge to *both* $u$ and $v$. In a social network, a high in-neighborhood intersection suggests that $u$ and $v$ share similar sources of prestige, influence, or advice.

**Out-Neighborhood Intersection** ($N^{out}(u) \cap N^{out}(v)$): Represents the set of common targets—nodes that *both* $u$ and $v$ direct an edge to. In a social network, this indicates shared sociability targets (e.g., $u$ and $v$ both sending advice to the same set of people).

Let's find these two types of intersections for nodes $C$ and $D$ in @fig-directed:

For **In-Neighborhoods**: $N^{in}(C) = \{B, E\}$ and $N^{in}(D) = \{B, E, G\}$. The intersection is:
$$
  N^{in}(C) \cap N^{in}(D) = \{B, E\}
$$
  
This means both $C$ and $D$ receive incoming connections from nodes $B$ and $E$.

For **Out-Neighborhoods**: $N^{out}(C) = \emptyset$ and $N^{out}(D) = \{B\}$. The intersection is:
$$
  N^{out}(C) \cap N^{out}(D) = \emptyset
$$
  
Since node $C$ sends no outgoing ties, there is no overlap in their outgoing targets. If we instead intersect $E$ and $G$ (where $N^{out}(E) = \{C, D\}$ and $N^{out}(G) = \{F, D\}$), we find:
$$
  N^{out}(E) \cap N^{out}(G) = \{D\}
$$
 
This shows that both $E$ and $G$ share $D$ as a common out-neighbor.

### Neighborhood Union in Directed Graphs

Similarly, we can compute the union of neighborhood sets to identify the total set of unique nodes connected to either $u$ or $v$ (or both):

**In-Neighborhood Union** ($N^{in}(u) \cup N^{in}(v)$): The total set of unique nodes that send a directed edge to either $u$ or $v$.

**Out-Neighborhood Union** ($N^{out}(u) \cup N^{out}(v)$): The total set of unique nodes that receive a directed edge from either $u$ or $v$.

Using our examples from @fig-directed, the union of the in-neighborhoods of $C$ and $D$ is:
$$
  N^{in}(C) \cup N^{in}(D) = \{B, E, G\}
$$

And the union of the out-neighborhoods of $E$ and $G$ is:
$$
  N^{out}(E) \cup N^{out}(G) = \{C, D, F\}
$$

---

## Node Degree in Directed Graphs

Because in a directed graph each node has two distinct sets of neighbors, we can compute two versions of degree for the same node, corresponding directly to the cardinalities of their neighborhood sets.

### Indegree

For any node $i$, we can count the number of directed edges that have node $i$ as their destination node. Since each incoming arrow comes from an in-neighbor of node $i$, the indegree (denoted $k^{in}_i$) is mathematically defined as the **cardinality of the in-neighborhood set** of that node:

$$
  k^{in}_i = |N^{in}(i)|
$$ {#eq-indegree-card}

### Outdegree

Additionally, for any node $i$, we can count the number of directed edges that have node $i$ as their source node. Since each outgoing arrow goes to an out-neighbor of node $i$, the outdegree (denoted $k^{out}_i$) is mathematically defined as the **cardinality of the out-neighborhood set** of that node:

$$
  k^{out}_i = |N^{out}(i)|
$$ {#eq-outdegree-card}

### Concrete Examples

Let's see how this works in practice by calculating the indegree and outdegree as neighborhood cardinalities for nodes $B$ and $D$ in @fig-directed.

#### Node B
* **Neighborhood Sets**:
  - In-neighborhood set: $N^{in}(B) = \{A, D\}$
  - Out-neighborhood set: $N^{out}(B) = \{A, C, D\}$
* **Degree Calculation**:
  - **Indegree**: $k^{in}_B = |N^{in}(B)| = |\{A, D\}| = 2$. This tells us that node $B$ has two incoming connections.
  - **Outdegree**: $k^{out}_B = |N^{out}(B)| = |\{A, C, D\}| = 3$. This tells us that node $B$ has three outgoing connections.

#### Node D
* **Neighborhood Sets**:
  - In-neighborhood set: $N^{in}(D) = \{B, E, G\}$
  - Out-neighborhood set: $N^{out}(D) = \{B\}$
* **Degree Calculation**:
  - **Indegree**: $k^{in}_D = |N^{in}(D)| = |\{B, E, G\}| = 3$. This tells us that node $D$ has three incoming connections.
  - **Outdegree**: $k^{out}_D = |N^{out}(D)| = |\{B\}| = 1$. This tells us that node $D$ has one outgoing connection.

Can you calculate the indegree and outdegree as neighborhood cardinalities for node $C$ in @fig-directed?

The graph-theoretic concepts of indegree and outdegree have clear sociological interpretations. In a social network, for instance, a node having a large outdegree (large out-neighborhood cardinality) indicates a *sociable person* (someone active in sending out connections), while having a large indegree (large in-neighborhood cardinality) can indicate a *popular person* or an actor of high prestige (someone whom many others want to connect with). 

### Types of Nodes in Directed Graphs

In a directed graph, we can categorize (non-isolate) nodes based on their sending and receiving behavior [@Harary65], and thus based on their indegree and outdegree:

-  **Receivers**: Nodes that receive ties but do not send them. For receiver nodes, indegree $k^{in} > 0$ and outdegree $k^{out} = 0$. For instance, in @fig-directed, node *C* has an indegree of 2 and an outdegree of 0, making it a receiver node.
- **Carriers**: Nodes that both receive and send ties. For carrier nodes, $k^{in} > 0$ and $k^{out} > 0$. For instance, in @fig-directed, node *A* has an indegree of 1 and an outdegree of 2, while node *B* has an indegree of 2 and an outdegree of 3, making both nodes carriers.
- **Transmitters**: Nodes that send ties but do not receive them. For transmitter nodes, $k^{in} = 0$ and $k^{out} > 0$. For instance, in @fig-directed, node *E* has an indegree of 0 and an outdegree of 2, making it a transmitter node.

### Directed Degree Sets

Computing the degrees of all nodes in a directed network yields two separate vectors (called **directed degree sets**), each of length $n$: the **indegree set** (written $\mathbf{k}^{in}$) and the **outdegree set** (written $\mathbf{k}^{out}$). 

For the directed graph in @fig-directed, these sets list each node's incoming and outgoing degrees:

$$
  \mathbf{k}^{in} = (k^{in}_A, k^{in}_B, \dots, k^{in}_G) = (2, 2, 2, 3, 0, 2, 0)
$$

$$
  \mathbf{k}^{out} = (k^{out}_A, k^{out}_B, \dots, k^{out}_G) = (2, 3, 0, 1, 2, 1, 2)
$$

### Directed Degree Sequences

When we arrange the elements of a degree set in decreasing order, from largest to smallest, we obtain the **degree sequence**. In directed graphs, this gives us two distinct sequences:

The **Indegree Sequence** ($\mathbf{d}^{in}$): For @fig-directed, ordering the indegree set yields:
$$
  \mathbf{d}^{in} = (3, 2, 2, 2, 2, 0, 0)
$$

The **Outdegree Sequence** ($\mathbf{d}^{out}$): Ordering the outdegree set yields:
$$
  \mathbf{d}^{out} = (3, 2, 2, 2, 1, 1, 0)
$$

### Minimum and Maximum Degree in Directed Graphs

We can examine the upper and lower bounds of connectivity in a directed network by taking the maximum and minimum values of the directed degree sets. This yields four distinct metrics:

the **Maximum Indegree** ($k^{in}_{max}$): The highest number of incoming ties received by any node:
$$
  k^{in}_{max} = \max(\mathbf{k}^{in}) = 3 \quad (\text{Node } D)
$$

The **Minimum Indegree** ($k^{in}_{min}$): The lowest number of incoming ties received by any node:
$$
  k^{in}_{min} = \min(\mathbf{k}^{in}) = 0 \quad (\text{Nodes } E \text{ and } G)
$$

The **Maximum Outdegree** ($k^{out}_{max}$): The highest number of outgoing ties sent by any node:
$$
  k^{out}_{max} = \max(\mathbf{k}^{out}) = 3 \quad (\text{Node } B)
$$

The **Minimum Outdegree** ($k^{out}_{min}$): The lowest number of outgoing ties sent by any node:
$$
  k^{out}_{min} = \min(\mathbf{k}^{out}) = 0 \quad (\text{Node } C)
$$

### Indegree and Outdegree Range

The arithmetic difference between the maximum and minimum degrees represents the heterogeneity range of connectivity in the digraph:

The **Indegree Range** ($k^{in}_r$): The gap between the most popular and least popular nodes:
$$
  k^{in}_r = k^{in}_{max} - k^{in}_{min} = 3 - 0 = 3
$$

The **Outdegree Range** ($k^{out}_r$): The gap between the most active sender and the least active sender:
$$
  k^{out}_r = k^{out}_{max} - k^{out}_{min} = 3 - 0 = 3
$$

---

## Directed Graph Metrics

Several basic graph-level metrics differ conceptually or mathematically when applied to directed graphs.

### Directed Graph Maximum Size

As we saw in @sec-graphtheory, the **graph size** is the number of edges (or arcs) $m$ in the graph. The **maximum size** represents the maximum possible number of edges that can exist in a graph with $n$ nodes. 

In an undirected graph, loops are not allowed, meaning the maximum possible number of edges is $\frac{n(n - 1)}{2}$. However, in a directed graph, edges have directionality, which means there can be two directed edges (arcs) between any pair of nodes—one in each direction (e.g., $u \rightarrow v$ and $v \rightarrow u$).

Therefore, the maximum possible number of directed edges in a directed graph of order $n$ is exactly twice that of an undirected graph:

$$
  m_{max} = n(n - 1)
$$ {#eq-dirmaxsize}

For our directed graph in @fig-directed of order $n = 7$:

$$
  m_{max} = 7(7 - 1) = 7(6) = 42
$$

This tells us that in a network of 7 nodes, there are 42 possible directed pathways.

### Directed Graph Density

The **density** $d(G)^d$ of a directed graph measures *how many directed ties exist compared to how many directed ties are possible*. It is computed as the ratio of the observed number of directed edges $m$ to the maximum possible number of directed edges:

$$
  d(G)^d = \frac{m}{n(n-1)}
$$ {#eq-dirgraphedens}

If we apply this to the directed graph in @fig-directed, where we observe $m = 11$ directed edges among $n = 7$ nodes:

$$
  d(G)^d = \frac{11}{7(6)} = \frac{11}{42} \approx 0.262 \quad (26.2\%)
$$

This indicates that approximately 26.2% of all possible directed edges in the network are actually observed. Another way to interpret density is as the *probability* that, if we were to select two random nodes $u$ and $v$ from the network, the directed dyad from $u$ to $v$ would be connected by an arrow.

### Average Indegree and Outdegree

We can compute the average (mean) connectivity across nodes in a directed graph. Because each node has two distinct degrees, we can calculate both the **average indegree** ($\bar{k}^{in}$) and the **average outdegree** ($\bar{k}^{out}$):

The **Average Indegree** ($\bar{k}^{in}$): The sum of all indegrees divided by the number of nodes:
$$
  \bar{k}^{in} = \frac{\sum_i k^{in}_i}{n}
$$
* **Average Outdegree** ($\bar{k}^{out}$): The sum of all outdegrees divided by the number of nodes:
$$
  \bar{k}^{out} = \frac{\sum_i k^{out}_i}{n}
$$

An important result in graph theory is that **the average indegree and the average outdegree are always mathematically identical**:

$$
  \bar{k}^{in} = \bar{k}^{out} = \bar{k} = \frac{m}{n}
$$ {#eq-diravgdeg}

This identity exists because every directed edge (arc) must start at exactly one node (contributing to that node's outdegree) and end at exactly one node (contributing to that node's indegree). Consequently, the sum of all indegrees and the sum of all outdegrees are both exactly equal to the total number of edges $m$ ($\sum_i k^{in}_i = \sum_i k^{out}_i = m$).

For our directed graph in @fig-directed ($m = 11$, $n = 7$):

$$
  \bar{k} = \frac{m}{n} = \frac{11}{7} \approx 1.57
$$

## Degree Variance in Directed Graphs

Once we have calculated the average degree, we can measure the **heterogeneity** (or inequality) in connectivity across nodes by computing the degree variance. In directed graphs, because each node has two distinct degrees, we must compute two separate variances: **indegree variance** and **outdegree variance**. 

Just like in the undirected case, the degree variance represents the *average squared deviation* of each node's degree from the graph's average degree.

### Indegree Variance ($\mathcal{v}^{in}(G)$)
The indegree variance measures the inequality in incoming connections (popularity/prestige) across nodes in the network:

$$
  \mathcal{v}^{in}(G) = \frac{\sum_i (k^{in}_i - \bar{k})^2}{n}
$${#eq-dirindegvar}

### Outdegree Variance ($\mathcal{v}^{out}(G)$)
The outdegree variance measures the inequality in outgoing connections (sociability) across nodes:

$$
  \mathcal{v}^{out}(G) = \frac{\sum_i (k^{out}_i - \bar{k})^2}{n}
$$ {#eq-diroutdegvar}

Note that because $\bar{k}^{in} = \bar{k}^{out} = \bar{k} = m/n$, both formulas use the same average degree, but the individual node values $k^{in}_i$ and $k^{out}_i$ can differ, yielding distinct variances.

Let's compute both variances for the directed graph in @fig-directed (where $n = 7$, $m = 11$, and $\bar{k} = 11/7 \approx 1.57$). 

@tbl-dirdegvar lists each node's indegree, outdegree, and squared deviations from the average degree.

| Node | Indegree ($k^{in}$) | Squared Deviation $(k^{in} - 1.57)^2$ | Outdegree ($k^{out}$) | Squared Deviation $(k^{out} - 1.57)^2$ |
|:---:|:---:|:---:|:---:|:---:|
| **A** | 2 | 0.18 | 2 | 0.18 |
| **B** | 2 | 0.18 | 3 | 2.04 |
| **C** | 2 | 0.18 | 0 | 2.47 |
| **D** | 3 | 2.04 | 1 | 0.33 |
| **E** | 0 | 2.47 | 2 | 0.18 |
| **F** | 2 | 0.18 | 1 | 0.33 |
| **G** | 0 | 2.47 | 2 | 0.18 |
| **Sum**| **11** | **7.71** | **11** | **5.71** |

: Squared deviations of indegree and outdegree from the average degree {#tbl-dirdegvar}

Using the sums of the squared deviations from @tbl-dirdegvar, we calculate the variances:

**Indegree Variance**:
$$
  \mathcal{v}^{in}(G) = \frac{7.71}{7} \approx 1.10
$$

**Outdegree Variance**:
  
$$
  \mathcal{v}^{out}(G) = \frac{5.71}{7} \approx 0.82
$$

In this network, the indegree variance ($1.10$) is higher than the outdegree variance ($0.82$). Sociologically, this indicates that there is more inequality in the popularity/prestige of nodes (receiving ties) than in their sociability (sending ties). In other words, incoming connections are more heavily concentrated on a few "popular" nodes (like node $D$) while others are left as complete non-receivers (like $E$ and $G$).

## Special Classes of Directed Graphs

By imposing structural restrictions on reciprocity, degree, or edge directionality, we can define special classes of directed graphs.

### Symmetric Directed Graphs

If, in a directed graph, every pair of connected nodes is linked by a bi-directional, reciprocal relationship, the graph is a **symmetric directed graph**. 

```{r}
#| label: fig-symm
#| fig-cap: "A symmetric directed graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "B", "D", "A", "B", "C", "E", "C", "F", "A", "G", "D", "C", "D", "E", "G")
  to <- c("B", "A", "A", "D", "C", "B", "C", "E", "A", "F", "D", "G", "D", "C", "G", "E")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

@fig-symm shows a symmetric directed graph with eight nodes and sixteen bi-directional edges (thirty-two directed arcs). For every edge from $u$ to $v$, there exists a corresponding edge from $v$ to $u$.

### Functional Graphs

A **functional graph** is a special kind of digraph built by imposing the restriction that each node can only have a *single out-neighbor* (i.e., the outdegree of every node is exactly equal to one, $k_{out} = 1$). Functional graphs have important applications in the anthropological analysis of kinship exchange systems in traditional societies [@hageharary96].

```{r}
#| label: fig-func
#| fig-cap: "A functional directed graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "B", "C", "D", "E", "F", "G", "H")
  to <- c("C", "G", "G", "H", "A", "A", "B", "C")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

@fig-func shows a functional graph with eight nodes. If a graph $G$ is a functional graph, then the number of nodes $V$ is always equal to the number of directed edges $E$ ($V = E = 8$). This is mathematically constrained because each of the $V$ nodes sends out exactly one edge.

### Oriented Graphs {#sec-oriented}

An **oriented graph** is a directed graph containing no symmetric (reciprocal) dyads. If there is a directed link from $u$ to $v$, there cannot be a link from $v$ to $u$. It is the structural opposite of a symmetric directed graph.

```{r}
#| label: fig-oriented
#| fig-cap: "An oriented directed graph"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "D", "C", "C", "A", "D", "D", "G", "C", "B", "G")
  to <- c("B", "A", "F", "E", "F", "G", "C", "E", "A", "F", "A")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

In the oriented graph shown in @fig-oriented, every connected pair is linked *only* by a one-way directed tie; reciprocal dyads are forbidden by the nature of the relationship (e.g., strict hierarchy or non-reversible flows).

### Anti-Symmetric Ties and Directed Tree Graphs

As we saw in @sec-ties, a relation is **anti-symmetric** if reciprocity is strictly forbidden. If *A* is anti-symmetrically connected to *B*, then *B* *cannot* send the *same type of tie* back to *A*. When these relations are concatenated across multiple nodes, they can form a **directed tree graph** (for a more general consideration of tree graphs see @sec-trees). 

Consider the relationship "being the father of." You have a father, your father has a father, and so on. This creates a generational lineage represented as a tree. 

```{r}
#| label: fig-tree
#| fig-cap: "A tree graph with thirteen nodes"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  gr <- create_tree(n = 13, children = 3) 
  gr <- mutate(gr, name = toupper(letters[1:13])) 
  
  p <- ggraph(gr, layout = 'tree') +
    geom_edge_link(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(4, 'mm')),
                end_cap = circle(8, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

In @fig-tree, node $A$ is the earliest common ancestor, and all other nodes are anti-symmetrically related to it. Similarly, formal organizational structures where subordinates "take orders from" a single supervisor generate tree graphs. A notable property of tree graphs is that every node (except the root) has exactly one in-neighbor (one supervisor or one father), but can have multiple out-neighbors (children or subordinates).

---

## References {.unnumbered}
