# The Adjacency Matrix {#sec-adjmat}
```{r setup, include=FALSE}
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

As we saw in the @sec-introtomat, a matrix is a two-dimensional array of numbers. A matrix is two-dimensional because it arranges numbers according to the intersection of a **row dimension** and a **column dimension**. 

Matrices, therefore, are perfect for storing network information, because we can put the nodes in the row and column dimensions and then record characteristics of each **dyad** in the network (like whether they are adjacent in the corresponding graph) in the corresponding **cell** of the matrix. 

Recall that each *ij* cell of a matrix corresponds to the intersection of a particular row object *i* and a particular column object *j*, so they are perfect for recording dyadic information, like adjacency. A matrix that stores adjacency relations between pairs of nodes in a graph is called an **adjacency matrix**.

As we will see later, we can, of course, record all kinds of relationships between dyads in matrix form beyond adjacency. These include the length of the shortest paths linking two nodes, the number of shortest paths, the number of common neighbors, whether two nodes are reachable from one another, and so forth. 

Regardless, the adjacency matrix is the most important, most commonly used way of representing graphs in network analysis. As we saw in @sec-graphtheory, two actors are connected in the network if they share an edge (are **adjacent**) in the graph; otherwise, they are not connected (are **non-adjacent**). 

The adjacency matrix indicates whether two nodes are directly connected, and thus, as its name implies, it records the adjacency relations between every pair of nodes in the network.

## Symmetric Adjacency Matrices for Undirected Graphs

The first step in building the adjacency matrix that represents the graph is to list all the nodes $\{A, B, C, D, E, F, G, H, I\}$ as both a row and a column entry for each node. Next, one goes sequentially across the rows and columns, asking the question "does actor i have the relationship I am examining with actor j?" If the question asks about the presence or absence of a relationship, 0s and 1s are used. If A has a relationship with B, the value 1 is marked. Otherwise, 0.

```{r}
#| label: tbl-symmadjmat
#| tbl-cap: "Symmetric adjacency matrix corresponding to an undirected graph."
#| tbl-cap-location: bottom

    gr <- create_empty(9, directed = FALSE)
    gr <- gr %>% 
        bind_edges(data.frame(from = 1, to = 2:5)) %>% 
        bind_edges(data.frame(from = 2, to = 3:4)) %>% 
        bind_edges(data.frame(from = 3, to = 4:5)) %>% 
        bind_edges(data.frame(from = 4, to = 5:6)) %>% 
        bind_edges(data.frame(from = 6, to = c(7:9))) %>%
        bind_edges(data.frame(from = 7, to = c(8:9))) %>%
        bind_edges(data.frame(from = 8, to = c(9))) %>%
        mutate(name = toupper(letters[1:9]))
    A <- as.matrix(as_adjacency_matrix(gr))
    diag(A) <- "--"
    kbl(A, format = "pipe", align = "c") 
```

As we can see in @tbl-symmadjmat, *A* indeed has a relationship with *B*, so the corresponding cell for the row corresponding to *A* and the column corresponding to *B* is marked 1. In fact, *A* has a relationship with *B*, *C*, *D*, and E, and has 1's in each of the cells corresponding to these actors, but not with *F*, *G*, *H*, or *I*, and so 0's are in these cells.

But what do we do about the cells where we are theoretically supposed to ask if A has a relationship with A? As we have seen, for most sociological applications, it makes sense to just put a dash there, thus **blocking the diagonals**. It's not sociologically meaningful for A to have a relationship with itself. 

For example, asking "Is A friends with A?" does not make much sense, but there are rare cases when it does, such as when A is a group rather than a single person, and the relationship we are looking at might occur both within and between groups. As we saw in @sec-graphtheory, these are called **reflexive-ties** or **loops**. But if the network is represented as a **simple graph**, it should contain no loops.

After completing the first row, we ask whether actor B has a relationship with actor A? Well, yes, it does. In fact, we can know this without even looking, because, if you recall, this network is defined ahead of time as reciprocal: if A is friends with B, then B is friends with A. We can remember this because the graph we are using is undirected. This means the resulting matrix will be symmetric. **Symmetric matrices** are those that, when flipped along the diagonal (as shown in Figure @tbl-symmadjmat), the two sides of the matrix will be mirror images of each other.

## Asymmetric Adjacency Matrices for Directed Graphs
Conversely, a directed graph describing a network of asymmetric or anti-symmetric ties will create an **asymmetric matrix**. Saying a matrix is asymmetric means that the values contained in the **upper and lower triangles** of the matrix do not mirror each other. In other words, in an asymmetric matrix, the cell values are not necessarily the same (the relationship is not necessarily equivalent) between every pair of cases.

@fig-directed shows an example directed graph. The corresponding **asymmetric adjacency matrix** is shown in @tbl-asymmadjmat. Note that while some relationships (such as between node A and B) are reciprocated, not all connections in the network are reciprocated. Node G sends ties to D and F but receives none in return. 

In the resulting matrix, A to B and B to A each have a 1, while G to D and G to F also have a 1. However, the cells corresponding to F to G and D to G each have a value of 0 because the ties are unreciprocated. These unreciprocated ties make the resulting matrix asymmetric. The two halves across the diagonal are no longer mirror images, but contain different entries.

```{r}
#| label: tbl-asymmadjmat
#| tbl-cap: "Asymmetric adjacency matrix corresponding to an directed graph."
#| tbl-cap-location: bottom

    fr <- c("A", "A", "B", "B", "C", "D", "E", "E", "F", "G", "G")
    to <- c("B", "F", "A", "D", "B", "B", "C", "D", "A", "F", "D")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
    A <- as.matrix(as_adjacency_matrix(gr))
    diag(A) <- "--"
    kbl(A, format = "pipe", align = "c") %>% 
      kable_styling(bootstrap_options = c("hover", "condensed", "responsive"))
```

Why are the ties not reciprocated? You might remember from our lessons on types of ties and types of graphs, but it is because of the type of data that the graph and matrix are representing. For example, the matrix in @tbl-asymmadjmat and the graph shown in @fig-directed could represent an intramural basketball club in which members ask one another who they would like to have as a teammate. 
Not everyone would agree that they like having one another as teammates, and the matrix and graph in @fig-directed would represent that. 

In this case, a node like G or E looks really lonely since no one wants to play with them. However, if the tie were to be about advice, such that actually G gives advice to D and F, but does not take their advice back, G (and E) now look like respected figures in the network.


## Weighted Adjacency Matrices and Binarization

In our discussion of adjacency matrices so far, we have assumed that ties are **binary**—meaning they are either present ($1$) or absent ($0$). While binary representations are mathematically simple and clean, they fail to capture the nuanced realities of social life, where relationships have varying degrees of **strength**, **intensity**, or **volume**. 

To represent relationships with non-binary values, we use **weighted (or valued) graphs**, which are stored in **weighted adjacency matrices**. In a weighted adjacency matrix, each cell $W_{ij}$ contains a real number representing the magnitude or value of the tie between node $i$ and node $j$, rather than a simple $0$ or $1$. Some classic examples of weighted ties include:
- **Communication Volume**: The number of emails, text messages, or phone calls exchanged between two individuals in a week.
- **Economic Flow**: The dollar value of goods exported from country $i$ to country $j$ in a year.
- **Co-authorship Strength**: The number of scientific papers co-authored by researcher $i$ and researcher $j$.

In a weighted adjacency matrix, a cell value of $0$ still represents the absence of a tie, but any non-zero value represents the presence of a tie with its corresponding weight. Binary graphs can be thought of as a special, constrained case of weighted graphs where all non-zero weights are equal to $1$.

### The Technique of Binarization

Sometimes, a weighted network can become so densely interconnected that it becomes difficult to analyze or visualize (resembling a chaotic "hairball" of ties of varying strengths). To manage this complexity, network analysts often **binarize** a weighted matrix. 

Binarization is the process of converting a weighted adjacency matrix ($\mathbf{W}$) into a binary adjacency matrix ($\mathbf{A}$) by applying a threshold cutoff, which we denote as $\Theta$ (the Greek letter Theta). Under a threshold binarization rule:
- We set $A_{ij} = 1$ if the tie weight $W_{ij}$ is strictly greater than our threshold: $W_{ij} > \Theta$.
- We set $A_{ij} = 0$ otherwise: $W_{ij} \le \Theta$.

For example, in a network of corporate advice, we might ask employees how often they seek help from one another on a scale of 1 (rarely) to 5 (multiple times a day). To isolate only the most significant, high-volume advice channels, we might choose a threshold of $\Theta = 3$. This binarizes our matrix so that cell entries are only marked $1$ if employees seek help from one another at least several times a week, effectively pruning out weak, casual, or occasional interactions.

## Multiplexity and Multigraph Adjacency Matrices

In the real world, people rarely interact through just a single, isolated relationship. Your co-worker might also be your close friend, your workout partner, and your neighbor. In sociology, when two actors share multiple, distinct types of social ties, we say that their relationship exhibits **multiplexity**. 

In graph theory, a network containing multiple distinct types of edges between the same set of vertices is called a **multigraph**. To represent multiplexity and multigraphs in matrix form, we cannot simply use a single, two-dimensional square matrix, because a single cell can only hold one number at a time! 

Instead, network analysts typically represent a multigraph as a set of **stacked adjacency matrices**. If we are studying a group of actors with $K$ different relation types (e.g., friendship, advice, and kinship), we construct $K$ separate, parallel adjacency matrices ($\mathbf{A}^{(1)}, \mathbf{A}^{(2)}, \ldots, \mathbf{A}^{(K)}$), each of dimensions $N \times N$. 

In computer science and data science, this stacked arrangement is represented as a **three-dimensional matrix** (or a **tensor**) of dimensions $N \times N \times K$. This multi-layered matrix structure allows us to perform advanced operations, such as checking whether having an advice tie (Matrix 1) predicts subsequently forming a friendship tie (Matrix 2), providing a robust mathematical foundation for analyzing multiplex social systems.

## Bipartite Adjacency Matrices (Two-Mode Networks)

Finally, there is a class of social networks in which ties do not connect actors to other actors, but rather connect actors to **groups, events, or organizations**. For example, we might study which corporate executives sit on which boards of directors, which authors co-attend which conferences, or which political activists participate in which social movements. In sociology, these are known as **affiliation networks** or **two-mode networks**.

In graph theory, these networks are modeled using **bipartite graphs**, which consist of two completely disjoint sets of vertices (e.g., a set of $N$ actors and a set of $M$ events). Edges in a bipartite graph can only connect a vertex in the first set to a vertex in the second set; there are no within-set edges (actors cannot directly connect to other actors, and events cannot directly connect to other events).

Because the two sets of vertices are disjoint and of different sizes, a bipartite graph is represented mathematically not by a square matrix, but by a **rectangular matrix** of dimensions $N \times M$. This is known as a **bipartite adjacency matrix** (or an **affiliation matrix**, $\mathbf{C}$):
- Each row represents a specific actor $i$ (from the set of size $N$).
- Each column represents a specific event or group $j$ (from the set of size $M$).
- Cell $C_{ij} = 1$ if actor $i$ is affiliated with event $j$, else $0$.

As we will see in our dedicated lesson on affiliation networks, we can multiply this rectangular bipartite matrix by its transpose to project the two-mode data into two separate, one-mode square matrices, allowing us to see how actors are linked via shared events, and how events are linked via shared actors.


## References {.unnumbered}
