# Basic Matrix Operations {#sec-matops}

```{r setup, include=FALSE}
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
    require(expm)
```

Since matrices are just arrays of *numbers*, one of the most powerful things we can do is perform arithmetic, algebraic, and summary operations on them. These include calculating row and column totals, adding or subtracting matrices representing different relations, multiplying them, taking powers, and transposing them. In this lesson, we cover these basic matrix operations and explain their direct relevance to the analysis of social networks.

Our primary running example for multi-matrix operations will be the pair of graphs shown in @fig-rel. Suppose this is a network of twelve people who work at a (very) small company. We have recorded two types of relationships for each pair of persons: Whether *they hang out together* after work (a "hang out" relation), and whether they get assigned to *work together in team projects* (a "co-working" relation).

@fig-rel-1, composed of nodes joined by blue edges, represents hanging out relations; @fig-rel-2 represents co-working relations. The corresponding **adjacency matrices** for the @fig-rel networks are shown in @tbl-rel; @tbl-rel-1 (let's call it $\mathbf{H}$) records hanging out relations and @tbl-rel-2 (let's call it $\mathbf{C}$) records co-working relations.

```{r}
#| label: fig-rel
#| fig-cap: "Undirected graphs representing different social relationships."
#| fig-cap-location: bottom
#| fig-subcap:
#|   - "A graph representing hanging out relations (blue)."
#|   - "A graph representing co-working relations (red)." 
#|   - "A graph representing both hanging out and co-working relations (purple)."
#|   - "A graph representing either hanging out and co-working relations but not both (green)."
#| layout-ncol: 2
#| fig-width: 10
#| fig-height: 10

set.seed(859)

   gr1 <- play_islands(3, 4, m = 1, p_within = 0.8) %>% 
     activate(nodes) %>% 
     mutate(name = toupper(letters[1:12]))
     l <- ggraph(gr1, layout = 'tree')
     l <- as.matrix(l$data[, 1:2])
     
p1 <- ggraph(gr1, layout = l) +
       geom_edge_link(color = "steelblue", width = 1.15) +
       geom_node_point(size = 24, color = "tan2") +
       geom_node_text(aes(label = name), size = 12, color = "white") +
       theme_graph()
p1
     
   gr2 <- play_islands(4, 3, m = 1, p_within = 1) %>% 
     activate(nodes) %>% 
     mutate(name = toupper(letters[1:12]))
p2 <- ggraph(gr2, layout = l) +
       geom_edge_link(color = "firebrick", width = 1.15) +
       geom_node_point(size = 24, color = "tan2") +
       geom_node_text(aes(label = name), size = 12, color = "white") +
       theme_graph()
p2
     
     A1 <- as.matrix(as_adjacency_matrix(gr1))
     A2 <- as.matrix(as_adjacency_matrix(gr2))
     A_mult <- A1 * A2
     gr_mult <- graph_from_adjacency_matrix(A_mult)
p3 <- ggraph(gr_mult, layout = l) +
       geom_edge_link(color = "purple", width = 2.3) +
       geom_node_point(size = 24, color = "tan2") +
       geom_node_text(aes(label = name), size = 12, color = "white") +
       theme_graph()
p3
     
     A_uni <- abs(A1 - A2)
     gr_uni <- graph_from_adjacency_matrix(A_uni)
p4 <- ggraph(gr_uni, layout = l) +
       geom_edge_link(color = "forestgreen", width = 1.15) +
       geom_node_point(size = 24, color = "tan2") +
       geom_node_text(aes(label = name), size = 12, color = "white") +
       theme_graph()
p4
```

```{r}
#| label: tbl-rel
#| tbl-cap: "Adjacency matrices corresponding to networks defined by different relationships." 
#| tbl-subcap:
#|   - "Hanging out relationships."
#|   - "Co-working relationships."
#| layout-ncol: 1

     A1_display <- A1
     diag(A1_display) <- "--"
     kbl(A1_display, format = "html", align = "c")  %>% 
     column_spec(1, bold = TRUE) %>% 
     kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
    
     A2_display <- A2
     diag(A2_display) <- "--"
     kbl(A2_display, format = "html", align = "c")  %>% 
     column_spec(1, bold = TRUE) %>% 
     kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

---

## Row and Column Sums

Perhaps the most basic operation we can do on a single matrix is to compute its **row sums** and its **column sums**. Let's first look at how this works on a general numeric matrix before exploring its powerful network properties.

```{r}
#| label: tbl-exmat
#| tbl-cap: "An Example Matrix."
#| tbl-cap-location: bottom
     set.seed(123)
     m <- matrix(sample(1:9, 25, replace = TRUE), 
                 ncol = 5, byrow = TRUE)
     kbl(m, format = "html", align = "c") %>% 
          kable_styling(bootstrap_options = c("hover", "condensed", "responsive"))
```

Consider the matrix shown in @tbl-exmat. This matrix is of dimensions $5 \times 5$, meaning it has five rows and five columns. Let's call the matrix $\mathbf{B}$. 

### Row Sums
The **row sums** of $\mathbf{B}$ is written in sigma notation as:

$$
\sum_j b_{ij}
$$ {#eq-rowsums}

In this formula, $i$ refers to the rows of the matrix and $j$ refers to the columns. So the formula says, "to get the row sums, pick a row $i$, and sum the cells across the columns $j$." So for instance, if $i = 2$ (the second row), then @eq-rowsums turns into:

$$
\sum_{j=1}^{5} b_{2j} = 4 + 6 + 9 + 5 + 3 = 27
$$ {#eq-rowex}

The numbers in the sum are just the numbers in row two of @tbl-exmat. If we calculate the row sums of all the rows in the matrix, the result is a **row sum vector**, containing the total sum for the numbers in each row. For @tbl-exmat, the row sum vector is:

```{r}
     rs <- rowSums(m)
     kbl(t(rs), format = "html", align = "c") %>% 
          kable_styling(bootstrap_options = 
                             c("hover", "condensed", "responsive"))
```

### Column Sums
The **column sums** of a matrix work similarly. In sigma notation, the column sum for matrix $\mathbf{B}$ is written as:

$$
\sum_i b_{ij}
$$ {#eq-colsums}

Note that the subscript under the $\sum$ symbol specifies $i$ (meaning we are summing over the rows). @eq-colsums says: "to get the column sums, pick a column $j$, and sum each number down the rows $i$." So for instance, if $j = 3$ (the third column), then @eq-colsums turns into:

$$
\sum_{i=1}^{5} b_{i3} = 2 + 9 + 9 + 3 + 7 = 30
$$ {#eq-colex}

The column sum vector for @tbl-exmat is:

```{r}
     cs <- colSums(m)
     kbl(t(cs), format = "html", align = "c") %>% 
          kable_styling(bootstrap_options = 
                             c("hover", "condensed", "responsive"))
```

---

### Row and Column Sums of the Symmetric Adjacency Matrix

What happens if we calculate the row and column sum vectors of a symmetric adjacency matrix corresponding to an undirected graph? 

Let us go back to the undirected graph example, whose symmetric adjacency matrix $\mathbf{H}$ (from hanging out relations) is shown in @tbl-rel-1. The row and column sums for this symmetric matrix are identical:

```{r}
#| label: tbl-symmsums 
#| tbl-cap: "Row and column sum vectors of the symmetric adjacency matrix for an undirected graph."
#| tbl-cap-location: bottom
#| tbl-subcap: 
#|   - "Row sums of the symmetric adjacency matrix."
#|   - "Column sums of the symmetric adjacency matrix."
#| layout-nrow: 1

     rs_sym <- rowSums(A1)
     kbl(t(rs_sym), format = "html", align = "c") %>% 
          kable_styling(bootstrap_options = 
                             c("hover", "condensed", "responsive"))
     
     cs_sym <- colSums(A1)
     kbl(t(cs_sym), format = "html", align = "c") %>% 
          kable_styling(bootstrap_options = 
                             c("hover", "condensed", "responsive"))
```

For any symmetric adjacency matrix, the row and column sum vectors are identical because $a_{ij} = a_{ji}$. But what do these numbers represent? If you match them to the graph in @fig-rel-1, you will see that they are nothing but each node's **degree**! 

Thus, for an undirected graph, the row *or* column sums of the symmetric adjacency matrix give us the **degrees** of each node!

---

### Row and Column Sums of the Asymmetric Adjacency Matrix

Now, what happens if we compute the row and column sums for an asymmetric adjacency matrix representing a directed graph?

Let's construct a simple directed graph and calculate its row and column sums:

```{r}
#| label: tbl-asymmsums 
#| tbl-cap: "Row and column sum vectors of the asymmetric adjacency matrix for a directed graph."
#| tbl-cap-location: bottom
#| tbl-subcap: 
#|   - "Row sums of the asymmetric adjacency matrix (Outdegrees)."
#|   - "Column sums of the asymmetric adjacency matrix (Indegrees)."
#| layout-nrow: 1

    fr <- c("A", "A", "B", "B", "C", "D", "E", "E", "F", "G", "G")
    to <- c("B", "F", "A", "D", "B", "B", "C", "D", "A", "F", "D")
    edge.dat <- data.frame(fr, to)
    node.dat <- data.frame(name = union(fr, to))
    gr_dir <- tbl_graph(edges = edge.dat, nodes = node.dat)
    a_dir <- as.matrix(as_adjacency_matrix(gr_dir))
    
    rs_dir <- rowSums(a_dir)
    kbl(t(rs_dir), format = "html", align = "c") %>% 
         kable_styling(bootstrap_options = 
                            c("hover", "condensed", "responsive"))
     
    cs_dir <- colSums(a_dir)
    kbl(t(cs_dir), format = "html", align = "c") %>% 
         kable_styling(bootstrap_options = 
                            c("hover", "condensed", "responsive"))
```

In a directed graph, the row sum vector and column sum vector are *different*:
* **Row Sums of a Directed Adjacency Matrix**: Represents the directed graph's **outdegree set**. It sums up all the outgoing ties from node $i$ to all other nodes $j$.
* **Column Sums of a Directed Adjacency Matrix**: Represents the directed graph's **indegree set**. It sums up all incoming ties from all other nodes $i$ to node $j$.

So in the directed case, row and column sums capture separate, crucial graph properties: a node's **outdegree** and **indegree** respectively!

---

## Matrix Addition and Subtraction

When we have two or more matrices of the same dimensions, we can perform cell-by-cell arithmetic on them. To **add** two matrices, we simply add up the corresponding entries in each cell. In matrix notation:

$$
\mathbf{H} + \mathbf{C} = h_{ij} + c_{ij}
$$ {#eq-matadd}

Where $h_{ij}$ is the corresponding entry for nodes $i$ and $j$ in the hanging out adjacency matrix $\mathbf{H}$, and $c_{ij}$ is the same entry in the co-working adjacency matrix $\mathbf{C}$.

Why would we want to do this? If we were studying the network shown in @fig-rel, we might be interested in which dyads have **uniplex** (single-stranded) relations and which have **multiplex** (multi-stranded) relations. While some actors in the network *either* hang out together or work together, some of them do *both*. Adding up the adjacency matrices shown in @tbl-rel will tell us who these are. The result is shown in @tbl-mult.

```{r}
#| label: tbl-mult 
#| tbl-cap: "Uniplex and Multiplex adjacency matrix (H + C)."
#| tbl-cap-location: bottom

     A <- A1 + A2
     diag(A) <- "--"
     kbl(A, format = "html", align = "c")  %>% 
     column_spec(1, bold = TRUE) %>% 
     kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

@tbl-mult shows that the $BC$ dyad has a multiplex relation (there is a "2" in the corresponding cell entry) and so do the $AC$, $FH$, $GH$, $EF$, and $JL$ dyads. A graph showing the nodes linked only by multiplex relations (hanging out and co-working) is shown in @fig-rel-3.

Note that **matrix subtraction** works exactly the same way:

$$
\mathbf{H} - \mathbf{C} = h_{ij} - c_{ij}
$$ {#eq-matsub}

To subtract two matrices, we simply subtract the corresponding entries in each cell. Why would we ever want to do this? Suppose we want to focus on dyads in the network connected by a single, special-purpose **uniplex** tie, while disregarding both disconnected and multiplex-connected dyads.

Taking the **absolute** value of each of the subtracted cell entries (which turns negative entries into positive ones), written as $|h_{ij} - c_{ij}|$, will result in a binary matrix that contains a 1 *only* for people who are hangout buddies or workmates but not both. Such a matrix is shown in @tbl-uni and corresponds to the green graph shown in @fig-rel-4.

```{r}
#| label: tbl-uni
#| tbl-cap: "Uniplex-only adjacency matrix (|H - C|)."
#| tbl-cap-location: bottom

     A_uni_mat <- abs(A1 - A2)
     diag(A_uni_mat) <- "--"
     kbl(A_uni_mat, format = "html", align = "c")  %>% 
     column_spec(1, bold = TRUE) %>% 
     kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

---

## The Matrix Dot Product {#sec-dotproduct}

Another way of figuring out which pairs of people in a network have multiplex ties is to compute the **matrix dot product** (symbol: $\cdot$), also called the **Hadamard product** (symbol: $\circ$). We find the matrix dot product by *multiplying* the corresponding entries in each of the matrices. In matrix format:

$$
\mathbf{H} \circ \mathbf{C} = h_{ij} \times c_{ij}
$$ {#eq-matdot}

If we take the dot product of two adjacency matrices like $\mathbf{H}$ and $\mathbf{C}$, then the resulting matrix will have a one in a given cell *only if* $h_{ij} = 1$ and $c_{ij} = 1$. Otherwise, it will have a zero. This means that the dot product of two adjacency matrices will retain *only* the **multiplex ties** and erase all the other ones. The result of the dot product of the adjacency matrices shown in @tbl-rel is shown in @tbl-dot.

```{r}
#| label: tbl-dot
#| tbl-cap: "Multiplex relationship matrix (H o C)."
#| tbl-cap-location: bottom

     A_dot_mat <- A1 * A2
     diag(A_dot_mat) <- "--"
     kbl(A_dot_mat, format = "html", align = "c")  %>% 
     column_spec(1, bold = TRUE) %>% 
     kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

As we can see, the only dyads that have non-zero entries in @tbl-dot are the multiplex dyads. The resulting network, comprising the combined "hanging + co-working" relationship, is shown in @fig-rel-3. Note that this network is much more sparse than either of the other two, since there's an edge between nodes only when they are adjacent in *both* networks.

---

## The Matrix Transpose {#sec-transpose}

One final basic operation we can do with a matrix is "turn it 90 degrees" so that the rows of the new matrix correspond to the columns of the original matrix, and the columns of the new matrix correspond to the rows of the original matrix. This is called the **matrix transpose** (symbol: $^T$).

For instance, if we have a matrix $\mathbf{A}_{4 \times 5}$ of dimensions $4 \times 5$ (four rows and five columns), then the transpose $\mathbf{A}^T_{5 \times 4}$ will have five rows and four columns, with the respective entries given by the formula:

$$
a_{ij} = a^T_{ji}
$$

That is, the number that appears in the original matrix in the $i^{th}$ row and $j^{th}$ column now appears in the transposed version of the matrix in the $j^{th}$ row and $i^{th}$ column.

An example of a matrix and its transpose is shown in @tbl-trans.

```{r}
#| label: tbl-trans 
#| tbl-cap: A matrix and its transpose
#| tbl-subcap: 
#|   - "Original Matrix."
#|   - "Transposed Matrix."
#| layout-ncol: 2

A_mat <- matrix(c(3, 7, 4, 5, 2, 4, 9, 6, 3, 5, 5, 3, 2, 4, 4), nrow = 5)
B_mat <- t(A_mat)

kbl(A_mat, align = "c", format = "html") %>%
    kable_styling(full_width = TRUE,
                  bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
    column_spec(1:3, extra_css = "border-right: 0.75px solid;") %>% 
    row_spec(1:5, extra_css = "border-bottom: 0.75px solid;") %>%
    column_spec(1, extra_css = "border-left: 0.75px solid;") %>% 
    row_spec(1, extra_css = "border-top: 0.75px solid;") 

kbl(B_mat, align = "c", format = "html") %>%
    kable_styling(full_width = TRUE,
                  bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
    column_spec(1:5, extra_css = "border-right: 0.75px solid;") %>% 
    row_spec(1:3, extra_css = "border-bottom: 0.75px solid;") %>%
    column_spec(1, extra_css = "border-left: 0.75px solid;") %>% 
    row_spec(1, extra_css = "border-top: 0.75px solid;")
```

The original matrix in @tbl-trans-1 has five rows and three columns. The transposed matrix has three rows and five columns. We can find the same numbers in the original and transposed matrices by switching the rows and columns. Thus, in the original matrix, the number in the third row, second column is 4 ($a_{32} = 4$). In the transposed matrix, the same 4 appears in the second row, third column ($a^T_{23} = 4$).

Thus, the transposed version of a matrix contains the exact same numerical information but organized along the opposite dimension. In network analysis, the transpose is invaluable for studying two-mode networks (flipping person-by-group matrices) and analyzing structural equivalence and blockmodels (as we will see in @sec-affiliation and @sec-blockmodeling).

---

## References {.unnumbered}
