# Directed Graphs {#sec-directedgraphs}

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    library(ggplot2)
    library(ggraph)
    library(tidygraph)
    library(igraph)
    library(kableExtra)
```

## Introduction to Directed Graphs (Digraphs)

While simple undirected graphs are useful for representing symmetric relationships (where connection is naturally mutual), many social relationships are inherently directional (see @sec-ties). For these relationships, one member of a pair can claim a tie with the other, but the other person may or may not send the same tie back. Examples include giving advice, sending emails, lending money, or nominating friends.

To represent directional relations, we use **directed graphs** (also called **digraphs**). In a directed graph, the edges are *ordered pairs* of nodes, meaning the order in which we list the nodes in an edge matters. The connection goes from a **source node** (the sender) to a **destination node** (the receiver). Therefore, the edge $AB$ is a different entity from the edge $BA$. 

In directed graphs, simple undirected lines are replaced by directed arrows (often called **arcs**). When a digraph has no loops (edges connecting a node to itself) and at most one directed edge connecting a sender to a receiver in each direction, it is called a **simple directed graph**.

@fig-directed shows a point-and-line diagram of a digraph representing a simple directed social network.

```{r}
#| label: fig-directed
#| fig-cap: "A simple directed graph."
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 8

  fr <- c("A", "A", "B", "B", "B", "D", "E", "E", "F", "G", "G")
  to <- c("B", "F", "A", "D", "C", "B", "C", "D", "A", "F", "D")
  edge.dat <- data.frame(fr, to)
  node.dat <- data.frame(name = union(fr, to))
  gr <- tbl_graph(edges = edge.dat, nodes = node.dat)
  
  p <- ggraph(gr, layout = 'kk') +
    geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                arrow = arrow(length = unit(5, 'mm')),
                end_cap = circle(8, 'mm'), 
                sep = unit(6, 'mm')) +
    geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

In directed networks, reciprocity is a major analytical focus. For some ties (such as liking, trusting, or friendship), reciprocity is binary; it either exists or it does not. If @fig-directed were an **advice network** [@cross01], we could say that *H* seeks advice from *D*, but *D* does not seek advice from *H*. This lack of reciprocity is common in formal hierarchies, where advice-seeking often flows upward to more experienced or authoritative actors, reflecting power differentials. 

---

## Node Neighborhoods in Directed Graphs

Just like in undirected (simple) graphs, each node in a directed graph has a **node neighborhood**. However, because each node can be the **source** or **destination** for an asymmetric edge, we have to differentiate the neighborhood of a node depending on whether the node is the **sender** or the **recipient** of a given link.

### In-Neighborhoods
We say that a node *j* is an **in-neighbor** of a node *i* if there is a directed link with *j* as the source and *i* as the destination node ($j \rightarrow i$). For each node, the full set of in-neighbors forms the **in-neighborhood** of that node. This is written:

$$
  N^{in}(v)
$$

where $v$ is the label corresponding to the node. For instance, in @fig-directed, the node set $N^{in}(D) = \{B, E, G\}$ is the in-neighborhood of node *D*.

### Out-Neighborhoods
In the same way, we say that a node *i* is an **out-neighbor** of a node *j* if there is a directed link with *j* as the source and *i* as the destination ($j \rightarrow i$). For each node, the full set of out-neighbors defines the **out-neighborhood** of that node. This is written:

$$
  N^{out}(v)
$$

where $v$ is the label corresponding to the node. For instance, in @fig-directed, the node set $N^{out}(B) = \{A, C, D\}$ is the out-neighborhood of node *B*.

Nodes will only appear in both the in- and out-neighborhood sets when there are **reciprocal** or mutual ties between the nodes. For instance, in @fig-directed, the out-neighborhood of node *F* is $\{A\}$ and the in-neighborhood is $\{A, G\}$. Here, node *A* appears in both the in- and out-neighborhood sets because *A* has a reciprocal tie with *F*.

---

## Node Degree in Directed Graphs

Because in a directed graph each node has two distinct sets of neighbors, we can compute two versions of degree for the same node, corresponding directly to the cardinalities of their neighborhood sets.

### Indegree
For any node *i*, we can count the number of edges that have a given node *i* as their destination node. This is also the cardinality of the in-neighborhood set of that node. This is called a node's **indegree** and is denoted:

$$
  k^{in}_i
$$

### Outdegree
Additionally, for any node *i*, we can count the number of edges that have a given node *i* as their source node. This is also the cardinality of the out-neighborhood set of that node. This is called the node's **outdegree** and is denoted by:

$$
  k^{out}_i
$$

For instance, in @fig-directed, $k^{out}_B = 3$ and $k^{in}_B = 2$. Node *B* has *three* **outgoing ties** (to nodes *A*, *C*, and *D*) and *two* **incoming ties** (from nodes *A* and *D*).

The graph-theoretic concepts of indegree and outdegree have clear sociological interpretations. In a social network, for instance, a node having a large outdegree could indicate a *sociable person* (a person who likes to connect with others), while having a large indegree can indicate a *popular person* (e.g., a person lots of other people want to be friends with). 

### Types of Nodes in Directed Graphs

In a directed graph, we can categorize (non-isolate) nodes based on their sending and receiving behavior [@Harary65], and thus based on their indegree and outdegree:

*   **Receivers**: Nodes that receive ties but do not send them. For receiver nodes, indegree $k^{in} > 0$ and outdegree $k^{out} = 0$. For instance, in @fig-directed, node *C* has an indegree of 2 and an outdegree of 0, making it a receiver node.
*   **Carriers**: Nodes that both receive and send ties. For carrier nodes, $k^{in} > 0$ and $k^{out} > 0$. For instance, in @fig-directed, node *A* has an indegree of 1 and an outdegree of 2, while node *B* has an indegree of 2 and an outdegree of 3, making both nodes carriers.
*   **Transmitters**: Nodes that send ties but do not receive them. For transmitter nodes, $k^{in} = 0$ and $k^{out} > 0$. For instance, in @fig-directed, node *E* has an indegree of 0 and an outdegree of 2, making it a transmitter node.

---

## Directed Graph Metrics

Several basic graph-level metrics differ conceptually or mathematically when applied to directed graphs.

### Average Degree in Directed Graphs

We can use a variation of the undirected average degree procedure to compute the average degree in directed graphs. Take the directed graph shown in @fig-directed, which has seven nodes and 11 edges. In this case, there is no need to "double" the number of lines as in the case of an undirected graph.

We could compute either the average **indegree** or the average **outdegree** of each node, and it would give us the same answer because each link serves a single purpose in determining either the outdegree or the indegree of each node. Accordingly, the average degree in the directed case is:

$$
  \bar{k} = \frac{m}{|V|} = \frac{m}{n} = \frac{11}{7} \approx 1.6
$$ {#eq-diravgdeg}

### Density in Directed Graphs

To compute the density of a directed graph, we compare the observed number of directed edges $m$ to the maximum possible number of directed edges that could exist among $n$ nodes. In a directed graph, edges can go in both directions between any pair of nodes, meaning the maximum possible number of directed ties is $n(n - 1)$. 

Therefore, there is no need to multiply the numerator by two as we do in undirected graphs, and the equation for computing the density of a directed graph is:

$$
  d(G)^d = \frac{m}{n(n-1)}
$$ {#eq-dirgraphedens}

If we apply this to the directed graph in @fig-directed ($m = 11$, $n = 7$):

$$
  d(G)^d = \frac{11}{7(6)} = \frac{11}{42} \approx 0.262 \quad (26.2\%)
$$

This indicates that approximately 26.2% of all possible directed edges in the network are actually observed.

### Degree Variance in Directed Graphs

Once we have calculated the average degree, we can measure the **heterogeneity** (or inequality) in connectivity across nodes by computing the degree variance. In directed graphs, because each node has two distinct degrees, we must compute two separate variances: **indegree variance** and **outdegree variance**. 

Just like in the undirected case, the degree variance represents the *average squared deviation* of each node's degree from the graph's average degree.

#### Indegree Variance ($\mathcal{v}^{in}(G)$)
The indegree variance measures the inequality in incoming connections (popularity/prestige) across nodes in the network:

$
  \mathcal{v}^{in}(G) = \frac{\sum_i (k^{in}_i - \bar{k})^2}{n}
$ {#eq-dirindegvar}

#### Outdegree Variance ($\mathcal{v}^{out}(G)$)
The outdegree variance measures the inequality in outgoing connections (sociability) across nodes:

$
  \mathcal{v}^{out}(G) = \frac{\sum_i (k^{out}_i - \bar{k})^2}{n}
$ {#eq-diroutdegvar}

Note that because $\bar{k}^{in} = \bar{k}^{out} = \bar{k} = m/n$, both formulas use the same average degree, but the individual node values $k^{in}_i$ and $k^{out}_i$ can differ, yielding distinct variances.

#### Example Calculation
Let's compute both variances for the directed graph in @fig-directed (where $n = 7$, $m = 11$, and $\bar{k} = 11/7 \approx 1.57$). 

@tbl-dirdegvar lists each node's indegree, outdegree, and squared deviations from the average degree.

| Node | Indegree ($k^{in}$) | Squared Deviation $(k^{in} - 1.57)^2$ | Outdegree ($k^{out}$) | Squared Deviation $(k^{out} - 1.57)^2$ |
|:---:|:---:|:---:|:---:|:---:|
| **A** | 2 | 0.18 | 2 | 0.18 |
| **B** | 2 | 0.18 | 3 | 2.04 |
| **C** | 2 | 0.18 | 0 | 2.47 |
| **D** | 3 | 2.04 | 1 | 0.33 |
| **E** | 0 | 2.47 | 2 | 0.18 |
| **F** | 2 | 0.18 | 1 | 0.33 |
| **G** | 0 | 2.47 | 2 | 0.18 |
| **Sum**| **11** | **7.71** | **11** | **5.71** |

: Squared deviations of indegree and outdegree from the average degree {#tbl-dirdegvar}

Using the sums of the squared deviations from @tbl-dirdegvar, we calculate the variances:

* **Indegree Variance**:
  $
    \mathcal{v}^{in}(G) = \frac{7.71}{7} \approx 1.10
  $
* **Outdegree Variance**:
  $
    \mathcal{v}^{out}(G) = \frac{5.71}{7} \approx 0.82
  $

In this network, the indegree variance ($1.10$) is higher than the outdegree variance ($0.82$). Sociologically, this indicates that there is more inequality in the popularity/prestige of nodes (receiving ties) than in their sociability (sending ties). In other words, incoming connections are more heavily concentrated on a few "popular" nodes (like node $D$) while others are left as complete non-receivers (like $E$ and $G$).

---

## References {.unnumbered}
