# Basic Graph Metrics {#sec-graphmetrics}
```{r setup, include=FALSE}
    library(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

## Computing Metrics on Graphs

As graphs are really just representations of real-world networks, they can be as varied as different networks are from one another. Some social networks are small (composed of just a few people), while others are large. In some social networks, there is little connectivity, while other networks feature thousands, if not millions, of ties among people, like some social media platforms. Other networks, such as the network of all the neurons in the human brain, feature billions of nodes and trillions of edges. Some networks are very dense, while others are very sparse. Some networks are very centralized, while others are more decentralized. Some networks are very hierarchical, while others are more decentralized. Some networks are very clustered (featuring dense connections between people), while others are more sparse.

We can quantify a lot of these properties of real-world networks by representing them as **graphs** and then computing **graph metrics**. This is a fancy word for computing various quantities via the simple arithmetic operations of summing, multiplying, and dividing, most of which are functions of the number of nodes and edges in a graph. In this lesson, we will review the most important graph metrics in social networks.

As a running example, we will use the simple undirected graph shown in @fig-ref-graph.

```{r}
#| label: fig-ref-graph
#| fig-cap: "An undirected reference graph with 7 nodes and 9 edges."
#| fig-cap-location: bottom
#| fig-height: 5
#| fig-width: 7

  edges_df <- data.frame(
    from = c(1, 1, 2, 2, 3, 3, 4, 5, 6),
    to   = c(2, 3, 3, 4, 4, 5, 6, 6, 7)
  )
  gr <- tbl_graph(nodes = data.frame(name = toupper(letters[1:7])), edges = edges_df, directed = FALSE)
  
  set.seed(42)
  p <- ggraph(gr, layout = 'nicely') +
    geom_edge_link(color = "steelblue", width = 1.25) +
    geom_node_point(size = 18, color = "tan2") +
    geom_node_text(aes(label = name), size = 8, color = "white") +
    theme_graph() +
    scale_x_continuous(expand = expansion(mult = 0.15)) +
    scale_y_continuous(expand = expansion(mult = 0.15)) +
    coord_cartesian(clip = "off")
  p
```

### Graph Order
The **order** of a graph (typically written as $n$) is the *number of nodes* in the graph. Technically speaking, keeping in mind that the graph is really a set of nodes and edges, the order is the cardinality of the node set $n = |V|$. **Cardinality** is simply a technical term from set theory indicating *how many elements* there are in a set. Thus, our reference graph in @fig-ref-graph has an order of 7.

### Graph Size

Likewise, the **size** of a graph (typically written as $m$) is the *number of edges* in the graph, which is (like with nodes) the **cardinality** of the edge set $m = |E|$. Thus, the **size** of the reference graph in @fig-ref-graph is 9.

This means that if you say the **size** of the graph is 9, we will all understand that you are referring to the number of edges, as you would use the term **order** if you were talking about the number of nodes (which is 7 in this case).

## The Graph Maximum Size

In some cases, we may be interested in knowing how many edges there *could* be in a graph if everyone were related to everyone else in the network. This is the *maximum possible number of edges* that could exist in a network of order $n$.

Recall that, as we discussed previously, the number of edges in a graph is called the **graph size**. Additionally, a graph in which all possible edges are present is called a **complete graph**. The **maximum size** of a graph of order $n$ is the number of edges that would exist in that graph if the graph were complete. A graph in which *no* edges exist (every node is an isolate) is called an **empty graph**.

While this may seem complicated, it is actually given by a simple (and important) formula for an **undirected graph**:

$$
  Max(E)=\frac{n(n-1)}{2}
$${#eq-graphmax1}

For our example graph in @fig-ref-graph, which has an order of $n = 7$, the maximum size of the graph is:

$$
  Max(E) = \frac{7(7-1)}{2} = 
  \frac{7(6)}{2} = 
  \frac{42}{2} = 
  21 
$$  

This means that if every node in the graph were connected to every other node, there would be 21 edges in the graph. Since our reference graph only has 9 edges, we can see that it is not very close to being complete.

## Graph Maximum Size Explainer

Where does this formula for maximum graph size come from? A graph would be complete if every node were connected to every other node. It is easy to see that, for each node, this would be every other node in the graph except that node itself. So in a complete graph of order $n$, the degree of each node *has* to be $n-1$. So if we call the degree of the first node in the graph $(n-1)_1$, then the **degree set** ($\mathbf{k}^c$) of a complete graph would be:

$$
  \mathbf{k}^{c} = [(n-1)_1, (n-1)_2, (n-1)_3 \dots (n-1)_n]  
$${#eq-graphmax3}

It is easy to see that adding up the degree of each node in the complete graph will give us a sum of degrees where each edge is counted twice (once for each endpoint). So we can write that sum of degrees as:

$$
  \sum k_i^{c} = \sum_i^{n} n-1 
$${#eq-graphmax4}

Essentially, we add up $n-1$ to $n-1$ to $n-1$, etc. $n$ times, once for each node in the graph. From basic arithmetic, we know that adding the same number $n$ number of times is the same as multiplying that number by $n$. Thus, the sum of degrees for a complete graph of order $n$ is $n \times (n-1) = n(n-1)$.

Because each undirected edge connects two nodes, it contributes to the degree of both. This means that each edge is counted twice in the degree sum. Therefore, we divide the sum of degrees by 2 to eliminate double-counting, resulting in $\frac{n(n-1)}{2}$ as the formula for the maximum size of an undirected graph of order $n$.

### Applying your knowledge

You can now look like a genius. If somebody were to ask you:

**Question**: In a classroom with 10 kindergartners, how many total pairs of friends would exist if every kid were friends with every other kid?

Using your graph theory metrics knowledge, you can now pull out your phone calculator and compute:

$$
  \frac{10(10-1)}{2} = 
  \frac{10(9)}{2} = 
  \frac{90}{2} = 45
$$

Impressing your audience immensely!

## Graph Density
In a classic study, the British social anthropologist Elizabeth Bott made a classic distinction between two types of network structure. According to Bott, some networks are **tight-knit** and others are **loose-knit** [@bott57]. Tight-knit networks feature many connections, and actors can reach others through multiple pathways. Loose-knit networks are more sparsely connected, and actors are only reachable to one another via a very restricted set of pathways.

It was soon realized that a very simple graph metric, called the **network density**, could serve as a useful index of Bott's ideas about tight versus loose-knit networks. Tight-knit networks are **dense**, featuring many interconnections among actors, while loose-knit networks are less dense [@barnes69]. How can we think of this in terms of graph theory concepts?

The density $d(G)$ of a graph is a measure of *how many ties between actors exist compared to how many ties between actors are possible*, given the **graph size** (number of nodes) and the **graph order** (number of links). As such, the density of an undirected graph is quite simply calculated as the ratio between the observed number of edges $m$ (the cardinality of the edge set), and the graph maximum size as defined using @eq-graphmax1.

$$
  d(G)=\frac{m}{\frac{n(n-1)}{2}}=m \times \frac{2}{n(n-1)}=\frac{2m}{n(n-1)}
$$ {#eq-dens1}

Where $m$ is the number of edges (graph size) and $n$ is the number of nodes (graph order) in the network.

For our reference graph shown in @fig-ref-graph, of size $m = 9$ and order $n = 7$, we can use @eq-dens1 to compute the graph density as follows:

$$
  d(G) = \frac{2m}{n(n-1)} = 
  \frac{2 \times 9}{7 \times (7-1)}=
  \frac{18}{7 \times 6} = 
  \frac{18}{42} = 
  0.43
$$

The estimated density of the network, $d = 0.43$ tells us that 43% of the total possible number of edges are actually observed. Another way to think about density is as the *probability* that, if we were to choose two random nodes in the network, the resulting dyad would be **connected** (as opposed to **null**).

### Why is density important?

The density of a network property is important to consider for two reasons. First, (which is the definition of density!), it can help us understand how connected the network is compared to how connected it might be. Second, when comparing two networks with the same number of nodes and the same types of relationships, it can tell us how they differ.

For example, let us imagine two organizations, each with 10 people. One organization has high density, and the other has low density in terms of the interactions among the members. What might be some of the underlying social differences between the two organizations? While we would need more information, we could posit that the one issue might be that information does not transmit very efficiently within the low-density organization because it has to go from member to member, rather than diffusing rapidly from one member to all the others. Another issue might be the "hit by a bus" problem, where if one or two members are taken out of the network, you can suffer a breakdown because they are no longer there to coordinate the different parts that don't talk to each other. Denser networks are less vulnerable to disruption when key nodes are removed.


## References {.unnumbered}
