# Structural Balance {#sec-sbalance}

The concept of balance can be extended from individual triads to the entire social network, which is analyzed at the **whole network** level. This extension is known as **Structural Balance Theory**. 

In its original formulation, structural balance theory applies to **complete signed graphs**. A **signed graph** is a special kind of graph featuring *one* set of vertices (nodes) and *two* disjoint sets of edges: positive links (E+) and negative links (E-). 

A **complete signed graph** is one in which every pair of nodes is connected by either a positive or a negative edge. In an **adjacency matrix** representation of a signed graph, positive links are typically represented by +1, negative links by -1, and non-adjacent nodes by 0, though for structural balance, it is often applied to complete graphs where all connections exist.

## The Fundamental Theorem of Structural Balance 

The **Fundamental Theorem of Structural Balance** was discovered by mathematicians Dorwin Cartwright and Frank Harary in 1956 [@cartwright_harary56]. This theorem states that if a complete signed graph contains only balanced triads (i.e., every triad in the network is balanced as defined in @sec-tbalance), then the graph is balanced. 

Furthermore, such a balanced graph can be divided into two groups such that:

1. There are only positive links within each group.
2. There are only negative links between the two groups.

This theorem implies that structural balance is equivalent to perfect **group polarization**. For example, in a balanced network, nodes A, B, and C (in blue in @fig-balance) might form one group with positive ties (green solid edges) among themselves, while nodes D, E, and F (in purple in @fig-balance) form another group with positive ties among themselves, and all ties *between* these two groups would be negative (in red dashed edges). Note that in @fig-balance, *every* triangle is balanced, going by the sign multiplication rule covered in @sec-tbalance. In a balanced graph, the polarized structure of these two distinct groups can be recovered by rearranging the rows and columns of the signed adjacency matrix such that nodes in the same group are near one another.

```{r}
#| label: fig-balance
#| fig-cap: "Complete Balanced Signed Graph Illustrating the Fundamental Theorem of Structural Balance."
#| fig-cap-location: bottom
#| fig-height: 10
#| fig-width: 10

    library(igraph)
    library(ggraph)
    library(dplyr)
    # 1. Define nodes and assign them to two mutually exclusive factions
    nodes <- data.frame(
    name = LETTERS[1:6],
    faction = c("Faction 1", "Faction 1", "Faction 1", 
                "Faction 2", "Faction 2", "Faction 2")
        )
    # 2. Create edges for a complete (fully connected) graph
    # expand.grid creates all pairs, filter(from < to) ensures undirected, no self-loops
    edges <- expand.grid(from = LETTERS[1:6], to = LETTERS[1:6], stringsAsFactors = FALSE) %>%
    filter(from < to) 
    # 3. Assign signs based on the Fundamental Theorem of Structural Balance:
    # (+) Positive if in the same faction, (-) Negative if in different factions
    edges <- edges %>%
    mutate(
        from_faction = nodes$faction[match(from, nodes$name)],
        to_faction = nodes$faction[match(to, nodes$name)],
        sign = ifelse(from_faction == to_faction, "+", "-"),
        type = ifelse(from_faction == to_faction, "Positive Tie (+)", "Negative Tie (-)")
        )
    # 4. Create the igraph object
    g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
    # 5. Define a custom manual layout to clearly separate the two factions visually
    layout_matrix <- matrix(c(
    -1.5,  1,   # Node 1 (Faction 1)
    -1.5, -1,   # Node 2 (Faction 1)
    -2.5,  0,   # Node 3 (Faction 1)
    1.5,  1,   # Node 4 (Faction 2)
    1.5, -1,   # Node 5 (Faction 2)
    2.5,  0    # Node 6 (Faction 2)
    ), ncol = 2, byrow = TRUE)
    # 6. Plot using ggraph
    ggraph(g, layout = layout_matrix) +
    # Draw edges: solid green for positive, dashed red for negative
    geom_edge_link(aes(color = type, linetype = type), width = 1.8, alpha = 0.8) +
    # Draw nodes colored by their faction
    geom_node_point(aes(fill = faction), shape = 21, size = 26, color = "white", stroke = 1.5) +
    geom_node_text(aes(label = name), fontface = "bold", color = "white", size = 12) +
    # Custom scales
    scale_edge_color_manual(values = c("Positive Tie (+)" = "forestgreen", 
                            "Negative Tie (-)" = "firebrick")) +
    scale_edge_linetype_manual(values = c("Positive Tie (+)" = "solid", 
                            "Negative Tie (-)" = "dashed")) +
    scale_fill_manual(values = c("Faction 1" = "steelblue", 
                            "Faction 2" = "darkorchid")) +
    theme_graph() + theme(legend.position = "none")
```

A remarkable property of a structurally balanced graph is its resilience to change in its overall polarized structure. If new nodes are added to a structurally balanced graph, the existing group polarization pattern does not change, provided that the new nodes' relationships to every other existing node are also balanced.

Essentially, any new node entering such a network will fall into one of the two existing polarized groups, maintaining internal positive ties and external negative ties, thereby keeping the network polarized.

### Balance Does Not Necessarily Imply a "Good" Social State

Despite the term "balanced," this state does not inherently denote a harmonious or desirable social situation. On the contrary, structural balance describes a stable state of maximal polarization and potential conflict, such as seen in **political polarization** (e.g., Republicans versus Democrats), as in the network plot of Twitter bloggers shown in @fig-twitter, with yellow lines indicating negative links between accounts (e.g., someone posting a negative comment about the other). 

![(Formerly) Twitter network of political bloggers during the 2004 election *via*  [https://blogs.cornell.edu/info2040/2012/09/05/5291/](https://blogs.cornell.edu/info2040/2012/09/05/5291/)](images/twitter.png){#fig-twitter}


This theoretical framework helps explain how factions or polarized groups can form and persist within larger social networks. It is a mathematical property indicating a specific configuration of relationships, not a judgment of social well-being. This nuance is crucial for understanding the implications of balance theory in real-world social networks.

## Cycles in a Balanced Signed Graph

The direct consequence of this two-group partitioning for **cycles** within a balanced signed graph is as follows: *All cycles in a balanced signed graph must have a positive product of signs*.

1. *Cycles entirely within one group*: If a cycle exists entirely within one of the two polarized groups (e.g., within Group 1 or Group 2), then all the edges forming that cycle must be positive, as per the theorem. The product of any number of positive signs is always positive. Therefore, these cycles are inherently "balanced" with respect to their sign product.

2. *Cycles spanning between two groups*: If a cycle involves nodes from both polarized groups, it must cross between the groups an even number of times to return to its starting node. For instance, to go from Group 1 to Group 2 and then back to Group 1, it must cross the "boundary" twice. Each time a link crosses between the two groups, it carries a negative sign. If there's an even number of such negative links in the cycle, then the product of all the signs in the cycle will be positive (because an even number of negative signs multiplied together results in a positive product).

## Paths in a Balanced Signed Graph

Recall that a path is a sequence of distinct vertices and edges connecting a start node to an end node. The *sign* of the path is the product of the sign of each edge in the path. In a balanced graph, the **end nodes** of a path could belong to the same group, or each could belong to different groups in the polarized structure.

1. *Paths with End Nodes in the Same Group*: If a path begins and ends within the same polarized group (e.g., both start and end nodes are in Group 1, or both are in Group 2), *the product of the signs of its edges will always be positive*.

Reasoning: To start and end in the same group, the path must either stay entirely within that group (meaning all its links are positive, and thus their product is positive) or it must cross between the two groups an even number of times. Each time the path crosses from one group to the other, it traverses a negative link (by definition of the two-group partition). Since an even number of negative links multiplied together results in a positive sign (-1 * -1 = +1), and any within-group links are positive, the overall product of signs for the entire path will be positive. 

2. *Paths with End Nodes in Different Groups*: If a path begins in one polarized group and ends in the other (e.g., starts in Group 1 and ends in Group 2), *the product of the signs of its edges will always be negative*.

Reasoning: To connect nodes in different groups, the path must necessarily cross the boundary between the two groups an odd number of times. Each time it crosses, it utilizes a negative link. Since an odd number of negative links multiplied together results in a negative sign (+1 * -1 = -1), the overall product of signs for the entire path will be negative. 

In summary, the two-group partitioning, which is a hallmark of structural balance in complete signed graphs, creates a predictable environment for path signs. Paths that "stay" within their original group's alignment (by ending in the same group) retain a positive sign, while paths that "cross" the network's fundamental division (by ending in a different group) acquire a negative sign, reflecting the inherent polarization.

## From Balance to Clustering
As we have seen, Cartwright and Harary's fundamental theorem of structural balance shows that a balanced network necessarily splits into exactly *two* polarized factions. While this works beautifully for representing two opposing armies or a strict two-party political divide, it is often too restrictive for some other real-world social networks. In 1967, Davis introduced his **Clusterability Theorem** to generalize structural balance so that a network could fragment into multiple groups, rather than just two.

```{r}
#| label: fig-davis
#| fig-cap: A clusterable graph according to Davis
#| cap-location: bottom
#| fig-height: 10
#| fig-width: 10

library(igraph)
library(ggraph)
library(dplyr)

library(igraph)
library(ggraph)
library(dplyr)

# 1. Define nodes and assign them to three mutually exclusive clusters
nodes <- data.frame(
  name = LETTERS[1:9],
  faction = c("Cluster 1", "Cluster 1", "Cluster 1", 
              "Cluster 2", "Cluster 2", "Cluster 2",
              "Cluster 3", "Cluster 3", "Cluster 3")
)

# 2. Create edges for a complete (fully connected) graph
# expand.grid creates all pairs, filter(from < to) ensures undirected edges with no self-loops
edges <- expand.grid(from = LETTERS[1:9], to = LETTERS[1:9], stringsAsFactors = FALSE) %>%
  filter(from < to) 

# 3. Assign signs based on Davis's Clusterability Theorem:
# (+) Positive if in the same cluster, (-) Negative if in different clusters
edges <- edges %>%
  mutate(
    from_faction = nodes$faction[match(from, nodes$name)],
    to_faction = nodes$faction[match(to, nodes$name)],
    sign = ifelse(from_faction == to_faction, "+", "-"),
    type = ifelse(from_faction == to_faction, "Positive Tie (+)", "Negative Tie (-)")
  )

# 4. Create the igraph object
g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)

# 5. Define a custom layout to clearly separate the three clusters visually (a triangle formation)
layout_matrix <- matrix(c(
  # Cluster 1 (Top)
   -0.35,  2.0,   # Node A
  -0.65,  1.5,   # Node B
   0,  1.5,   # Node C
  
  # Cluster 2 (Bottom Left)
  -1.0, 0.5,   # Node D
  -1, 0,   # Node E
  -0.65, -0.2,   # Node F
  
  # Cluster 3 (Bottom Right)
   0.5, -0.5,   # Node G
   0, -0.5,   # Node H
   0.5,  0    # Node I
), ncol = 2, byrow = TRUE)

# 6. Plot using ggraph
ggraph(g, layout = layout_matrix) +
  # Draw edges: solid green for positive, dashed red for negative
  geom_edge_link(aes(color = type, linetype = type, width = type), alpha = 0.6) +
  
  # Draw nodes colored by their assigned cluster
  geom_node_point(aes(fill = faction), shape = 21, size = 22, color = "white", stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", color = "white", size = 10) +
  
  # Custom scales mapping
  scale_edge_color_manual(values = c("Positive Tie (+)" = "forestgreen", 
                                     "Negative Tie (-)" = "firebrick")) +
  scale_edge_linetype_manual(values = c("Positive Tie (+)" = "solid", 
                                        "Negative Tie (-)" = "dashed")) +
  scale_edge_width_manual(values = c("Positive Tie (+)" = 2, 
                                        "Negative Tie (-)" = 1)) +
  scale_fill_manual(values = c("Cluster 1" = "steelblue", 
                               "Cluster 2" = "darkorchid",
                               "Cluster 3" = "goldenrod")) +
  theme_graph() + 
  theme(legend.position = "none")

```

Davis realized that by relaxing the rules of balance slightly, he could account for a wider variety of social clustering. Specifically, classical balance theory considers a triad with three negative ties (where everyone hates everyone else) to be unbalanced. Davis argued that in a highly factionalized network, it makes perfect sense for three people from three completely different, competing groups to dislike one another, so it is possible to group this triad with the other balanced configurations.

Accordingly, under Davis's model, a graph is **clusterable** if it contains no triads with exactly one negative tie. That is, the only triads that exist are the four balanced triads in @sec-tbalance, and the all-negative-ties triad. If this condition is met, the network can be partitioned into any number of distinct clusters, not just two, as in the example graph shown in @fig-davis. Just like in the Cartwright and Harary model, all the ties within a specific cluster (e.g., D, E, F) are positive, but the ties between any two different clusters are entirely negative. As you can also see, all the triads in the graph either contain no negative links (the all positive triads within clusters) or if they contain negative links they contain two of them (+, -, -) or three of them (-, -, -).

## References {.unnumbered}
