---
title: "Balance Theory and Signed Graphs"
subtitle: "The Structure and Dynamics of Valenced Networks"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(patchwork)

# Custom transparent theme for slides
theme_slides <- function(base_size = 12) {
  theme_graph() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.box.background = element_rect(fill = "transparent", color = NA),
    legend.text = element_text(color = "#2c3e50", size = 11, face = "bold"),
    plot.title = element_text(size = base_size * 1.2, face = "bold", hjust = 0.5, color = "#2c3e50"),
    plot.subtitle = element_text(size = base_size * 1.0, hjust = 0.5, color = "#7f8c8d"),
    plot.margin = margin(10, 10, 10, 10)
  )
}

# Set global chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dev = "png",
  dev.args = list(bg = "transparent")
)
```

# Part 1: The Relational Basis of Valenced Networks & Dyadic Balance

---

## Balance Theory and Signed Graphs
### The Local and Global Dynamics of Sentiment Networks

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- In social life, relationships are not just present or absent; they are also **emotional** and **evaluative**.
- We study these using **sentiment networks**, which model liking, disliking, conflict, and cooperation.
- **Balance Theory** is the foundational framework linking:
  - Local, psychological micro-motives.
  - Global, macro-structural network patterns (like polarization).
- Today we will cover: Dyadic Balance, Triadic Balance, Structural Balance, and Clusterability.
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 5
#| fig.height: 5
edges <- data.frame(from = c("A", "B", "C"), to = c("B", "C", "A"), sign = c("+", "+", "-"))
g <- graph_from_data_frame(edges, directed = FALSE)
lay <- matrix(c(0, 1, -0.86, -0.5, 0.86, -0.5), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), width = 2.5, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 20, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## The Anatomy of Sentiment Networks

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- **Valenced ties** are links representing positive ($+$) or negative ($-$) feelings, evaluations, or behaviors.
  - *Positive valence*: Liking, helping, admiring, loving.
  - *Negative valence*: Hating, bullying, fighting, distrusting.
- A **signed graph** features a set of vertices $V$ and two disjoint sets of edges:
  - Positive links ($E^+$) and negative links ($E^-$).
- These networks represent stable social attitudes rather than fleeting events.
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 5
#| fig.height: 4.5
edges <- data.frame(from = c("Ego", "Ego"), to = c("A1", "A2"), sign = c("+", "-"))
g <- graph_from_data_frame(edges, directed = TRUE)
lay <- matrix(c(0, 0, -1, -1, 1, -1), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(4, "mm"), type = "closed"), 
                 end_cap = circle(12, "mm"), start_cap = circle(12, "mm"), width = 2.5) +
  geom_node_point(aes(fill = name), size = 24, color = "#2c3e50", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 5.5, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  scale_fill_manual(values = c("Ego" = "#fff", "A1" = "#d4edda", "A2" = "#f8d7da")) +
  coord_fixed(xlim = c(-1.5, 1.5), ylim = c(-1.4, 0.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## Fritz Heider's Dyadic Balance

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- Developed by social psychologist **Fritz Heider** (1946) to examine how ties cohere within a single pair.
- Distinction between:
  - **Symmetric ties**: Relations implied by similarity or co-membership (e.g., sharing a workspace).
  - **Asymmetric ties**: Directional sentiments (e.g., liking).
- **First Proposition**: Similarity breeds liking. Sharing co-membership or space breeds positive asymmetric ties ($aSb \implies aLb$).
- Forms a **positive feedback loop**: We spend more time with those we like ($aLb \implies aSb$).
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 5.5
#| fig.height: 4.5
edges <- data.frame(
  from = c("a1", "a2"), 
  to = c("b1", "b2"), 
  type = c("Symmetric Co-membership", "Asymmetric Liking")
)
g <- graph_from_data_frame(edges, directed = TRUE)
lay <- matrix(c(-0.8, 1, 0.8, 1, -0.8, -0.8, 0.8, -0.8), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = type), data = function(layout) { edges <- get_edges()(layout); edges[edges$type == "Symmetric Co-membership", ] },
                 end_cap = circle(10, "mm"), start_cap = circle(10, "mm"), width = 2) +
  geom_edge_link(aes(color = type), data = function(layout) { edges <- get_edges()(layout); edges[edges$type == "Asymmetric Liking", ] },
                 arrow = arrow(length = unit(4, "mm"), type = "closed"),
                 end_cap = circle(10, "mm"), start_cap = circle(10, "mm"), width = 2) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = c("a", "b", "a", "b")), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("Symmetric Co-membership" = "#2980b9", "Asymmetric Liking" = "forestgreen")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-1.2, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "bottom", legend.title = element_blank())
```
:::
:::

---

## Proximity, Similarity, and Multiplexity

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- Over time, people accumulate multiple types of ties, leading to **multiplex ties** ("multistranded" connections).
- **Principle of Propinquity** (Newcomb 1956): Sharing physical space or social contexts automatically facilitates acquaintance and liking.
  - Proximity breeds interaction, which breeds similarity, which in turn breeds liking, support, and advice.
- Relationships transition from uniplex (single-stranded) to multiplex.
- Multiple/reciprocal arrows use **parallel edges** to prevent overlapping.
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 5.5
#| fig.height: 4.5
edges <- data.frame(
  from = c("A", "A", "B", "A"),
  to   = c("B", "B", "A", "B"),
  type = c("Co-workers", "Liking", "Liking", "Lunch Partner")
)
g <- graph_from_data_frame(edges, directed = TRUE)
lay <- matrix(c(0, 0, 2, 0), ncol = 2, byrow = TRUE)
rownames(lay) <- c("A", "B")
ggraph(g, layout = lay) +
  geom_edge_parallel(aes(color = type, linetype = type), width = 2, sep = unit(4, "mm"),
                     arrow = arrow(length = unit(3.5, "mm"), type = "closed"), 
                     end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 20, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 7, color = "#2c3e50") +
  scale_edge_color_manual(values = c("Co-workers" = "#7f8c8d", "Liking" = "forestgreen", "Lunch Partner" = "#d35400")) +
  coord_fixed(xlim = c(-0.5, 2.5), ylim = c(-0.6, 0.6), clip = "off") +
  theme_slides() + theme(legend.position = "bottom", legend.title = element_blank())
```
:::
:::

---

## Dyadic Balance and Cognitive Tension

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- **Second Proposition**: We strive to maintain positive asymmetric ties with those we are similar to.
- Situations where similarity is paired with dislike ($aSb \land aDb$) are **unbalanced**.
  - Creates **cognitive inconsistency** and an unpleasant state of **affective tension** (e.g., hating a forced co-worker).
- **Resolution of Imbalance**:
  - Break the similarity tie (e.g., quit the job).
  - Transform the dislike into liking (e.g., adapt to like the co-worker).
- Multiple/reciprocal arrows use **parallel edges** to prevent overlapping.
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 5
#| fig.height: 4.5
edges <- data.frame(from = c("a", "a"), to = c("b", "b"), type = c("Similarity S", "Dislike D"))
g <- graph_from_data_frame(edges, directed = TRUE)
lay <- matrix(c(0, 0, 2, 0), ncol = 2, byrow = TRUE)
rownames(lay) <- c("a", "b")
ggraph(g, layout = lay) +
  geom_edge_parallel(aes(color = type, linetype = type), data = function(layout) { edges <- get_edges()(layout); edges[edges$type == "Similarity S", ] }, width = 2, sep = unit(4, "mm"),
                     end_cap = circle(12, "mm"), start_cap = circle(12, "mm")) +
  geom_edge_parallel(aes(color = type, linetype = type), data = function(layout) { edges <- get_edges()(layout); edges[edges$type == "Dislike D", ] }, width = 2, sep = unit(4, "mm"),
                     arrow = arrow(length = unit(3.5, "mm"), type = "closed"),
                     end_cap = circle(12, "mm"), start_cap = circle(12, "mm")) +
  geom_node_point(size = 22, fill = "#fdfd96", color = "#f8e71c", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 8, color = "#2c3e50") +
  scale_edge_color_manual(values = c("Similarity S" = "#3498db", "Dislike D" = "firebrick")) +
  scale_edge_linetype_manual(values = c("Similarity S" = "solid", "Dislike D" = "33")) +
  coord_fixed(xlim = c(-0.5, 2.5), ylim = c(-0.6, 0.6), clip = "off") +
  theme_slides() + theme(legend.position = "bottom", legend.title = element_blank())
```
:::
:::

---

# Part 2: Triadic Balance Theory

---

## Fritz Heider’s P-O-Q Triad

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- To analyze triads, Heider extended dyadic balance to three distinct elements: the **P-O-Q Triad**.
- **Person (P)**: The focal individual from whose cognitive perspective the triad is analyzed.
- **Other (O)**: Towards whom P holds a directed sentiment.
- **Third Party / Object (Q)**: Another actor, group, organization, or abstract cultural idea.
- Sentiments are modeled via **directed, signed edges** ($+$ or $-$) going from P to O, P to Q, and O to Q.
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 5
#| fig.height: 5
edges <- data.frame(from = c("P", "P", "O"), to = c("O", "Q", "Q"), type = c("sentiment", "sentiment", "sentiment"))
g <- graph_from_data_frame(edges, directed = TRUE)
lay <- matrix(c(0, 1, -0.86, -0.5, 0.86, -0.5), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(arrow = arrow(length = unit(4, "mm"), type = "closed"), width = 1.8, 
                 color = "#7f8c8d", end_cap = circle(12, "mm"), start_cap = circle(12, "mm")) +
  geom_node_point(aes(fill = name), size = 24, color = "#2c3e50", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 7, color = "#2c3e50") +
  scale_fill_manual(values = c("P" = "#d4edda", "O" = "#ecf0f1", "Q" = "#f8d7da")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## The Mathematics of Triadic Balance

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- A triad is **balanced** if the product of its three edge signs is positive:
  $$\text{Sign}(P, O) \times \text{Sign}(O, Q) \times \text{Sign}(P, Q) = +1$$
- **Simplified Identification Rule**: A triad is balanced if it has an **even number of negative links** (exactly zero or two).
- Balanced configurations are stable and represent **harmonious sentiments**, creating no cognitive tension.
:::
:::

::: {.column width="50%"}
| Edge P-O | Edge O-Q | Edge P-Q | Product | State |
| :---: | :---: | :---: | :---: | :---: |
| $+$ | $+$ | $+$ | $+1$ | **Balanced** |
| $-$ | $-$ | $+$ | $+1$ | **Balanced** |
| $+$ | $-$ | $-$ | $+1$ | **Balanced** |
| $-$ | $+$ | $-$ | $+1$ | **Balanced** |
| $+$ | $+$ | $-$ | $-1$ | **Unbalanced** |
| $-$ | $+$ | $+$ | $-1$ | **Unbalanced** |
| $+$ | $-$ | $+$ | $-1$ | **Unbalanced** |
| $-$ | $-$ | $-$ | $-1$ | **Unbalanced** |
:::
:::

---

## Balanced Triads: The 4 Stable Configurations

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
1. **"A friend of a friend is a friend"** ($+ \times + \times + = +1$)
   - *Example*: You (P) and friend (O) both like the same classmate (Q).
2. **"An enemy of an enemy is a friend"** ($- \times - \times + = +1$)
   - *Example*: You (P) and ally (Q) form a bond against a mutual rival (O).
3. **"A friend of an enemy is an enemy"** ($- \times + \times - = +1$)
   - *Example*: Your rival (O) has a close friend (Q); you start to dislike Q.
4. **"An enemy of a friend is an enemy"** ($+ \times - \times - = +1$)
   - *Example*: Your country ally (O) declares war on country (Q); you also oppose Q.
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 6
#| fig.height: 5
# Uses patchwork to keep layouts independent and perfectly uncropped
edges1 <- data.frame(from = c("P", "O", "P"), to = c("O", "Q", "Q"), sign = c("+", "+", "+"))
g1 <- graph_from_data_frame(edges1, directed = TRUE)
edges2 <- data.frame(from = c("P", "O", "P"), to = c("O", "Q", "Q"), sign = c("-", "-", "+"))
g2 <- graph_from_data_frame(edges2, directed = TRUE)
lay <- matrix(c(0, 1, -0.86, -0.5, 0.86, -0.5), ncol = 2, byrow = TRUE)

p1 <- ggraph(g1, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(4, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("Friend of Friend")

p2 <- ggraph(g2, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(4, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("Enemy of Enemy")

p1 + p2 + plot_layout(nrow = 1)
```
:::
:::

---

## Unbalanced Triads: The 4 Unstable Configurations

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
1. **"A friend of my friend is my enemy"** ($+ \times + \times - = -1$)
   - *Example*: Friend (O) likes (Q), whom you (P) dislike.
2. **"An enemy of a friend is a friend"** ($+ \times - \times + = -1$)
   - *Example*: Partner (O) hates Marvel movies (Q), which you (P) love.
3. **"A friend of an enemy is a friend"** ($- \times + \times + = -1$)
   - *Example*: You (P) and your rival (O) are both attracted to the same person (Q).
4. **"An enemy of my enemy is also my enemy"** ($- \times - \times - = -1$)
   - *Example*: You (P) dislike both the Yankees (O) and Red Sox (Q).
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 6
#| fig.height: 5
# Uses patchwork to keep layouts independent and perfectly uncropped
edges1 <- data.frame(from = c("P", "O", "P"), to = c("O", "Q", "Q"), sign = c("+", "+", "-"))
g1 <- graph_from_data_frame(edges1, directed = TRUE)
edges2 <- data.frame(from = c("P", "O", "P"), to = c("O", "Q", "Q"), sign = c("-", "-", "-"))
g2 <- graph_from_data_frame(edges2, directed = TRUE)
lay <- matrix(c(0, 1, -0.86, -0.5, 0.86, -0.5), ncol = 2, byrow = TRUE)

p1 <- ggraph(g1, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(4, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("Friend is Enemy")

p2 <- ggraph(g2, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(4, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("All Enemies")

p1 + p2 + plot_layout(nrow = 1)
```
:::
:::

---

## Tension and Change

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- Unbalanced configurations produce psychological discomfort, making them **socially unstable**.
- This discomfort acts as an **engine of change**, driving social networks to transition toward balanced configurations.
- Transition occurs when the focal actor (P) changes the **sign/valence** of their ties to resolve tension.
- Balanced configurations, having no tension, are stable **equilibrium states**.
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 6
#| fig.height: 5
edges1 <- data.frame(from = c("P", "O", "P"), to = c("O", "Q", "Q"), sign = c("+", "+", "-"))
g1 <- graph_from_data_frame(edges1, directed = TRUE)

edges2 <- data.frame(from = c("P", "O", "P"), to = c("O", "Q", "Q"), sign = c("+", "+", "+"))
g2 <- graph_from_data_frame(edges2, directed = TRUE)

lay <- matrix(c(0, 1, -0.86, -0.5, 0.86, -0.5), ncol = 2, byrow = TRUE)

p1 <- ggraph(g1, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(3.5, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("Unbalanced (Tension)")

p2 <- ggraph(g2, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(3.5, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = c("P", "O", "Q")), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("Resolved (Balance)")

p1 + p2 + plot_layout(nrow = 1)
```
:::
:::

---

## Cross-Pressure

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- **Cross-pressure** (Davis 1963) occurs when an individual experiences conflicting social relations, leaving them caught between opposing forces.
- When P is positively tied to O1 and O2 ($+$, $+$), but O1 likes Q and O2 dislikes Q.
- According to balance theory, P is forced to develop an **ambivalent attitude** ($+/-$) toward Q.
- To resolve this tension, P must either distance themselves from one friend or convince them to reconcile.
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 5
#| fig.height: 5
edges <- data.frame(
  from = c("P", "P", "O1", "O2", "P"),
  to   = c("O1", "O2", "Q", "Q", "Q"),
  sign = c("+", "+", "+", "-", "?")
)
g <- graph_from_data_frame(edges, directed = TRUE)
lay <- matrix(c(0, 1, -1, 0, 1, 0, 0, -1), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(4, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(12, "mm"), start_cap = circle(12, "mm")) +
  geom_node_point(size = 22, fill = "#fdfd96", color = "#f8e71c", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick", "?" = "goldenrod")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "dashed", "?" = "dotted")) +
  coord_fixed(xlim = c(-1.5, 1.5), ylim = c(-1.4, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## Cross-Pressure in Real-Life: Politics

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- Political preferences provide a classic illustration of cross-pressure.
- Imagine you (P) have a brother (O1) who is a passionate Democrat ($+$) and a sister (O2) who strongly opposes them ($-$).
- Because you have positive bonds with both, you find yourself cross-pressured regarding the Democrats (Q).
- This pressure produces **social ambivalence**, often leading individuals to identify as moderates or **independents** with no strong partisan affiliation.
- Labels inside nodes are kept short (**You**, **Bro**, **Sis**, **Dems**) for a perfect fit.
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 5
#| fig.height: 5
edges <- data.frame(
  from = c("You", "You", "Bro", "Sis", "You"),
  to   = c("Bro", "Sis", "Dems", "Dems", "Dems"),
  sign = c("+", "+", "+", "-", "?")
)
g <- graph_from_data_frame(edges, directed = TRUE)
lay <- matrix(c(0, 1, -1, 0, 1, 0, 0, -1), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), arrow = arrow(length = unit(4, "mm"), type = "closed"), 
                 width = 2, end_cap = circle(14, "mm"), start_cap = circle(14, "mm")) +
  geom_node_point(size = 24, fill = "#fdfd96", color = "#f8e71c", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 5.5, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick", "?" = "goldenrod")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "dashed", "?" = "dotted")) +
  coord_fixed(xlim = c(-1.5, 1.5), ylim = c(-1.4, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

# Part 3: Structural Balance Theory & Whole-Network Polarization

---

## Structural Balance at the Network Level

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- Balance theory can be scaled up from individual cognitive triads to the **whole network** level.
- This extension is known as **Structural Balance Theory**.
- In its original formulation, it applies to **complete signed graphs**, where every pair of actors is connected by either a positive or negative link.
- In a **signed adjacency matrix** representation of a signed graph:
  - Positive links are represented by $+1$.
  - Negative links are represented by $-1$.
  - Non-adjacent nodes are represented by $0$.
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 5
#| fig.height: 5
set.seed(422)
g <- play_erdos_renyi(n = 5, p = 1, directed = FALSE)
E(g)$sign <- c("+", "-", "+", "+", "-", "+", "-", "+", "-", "+")
ggraph(g, layout = "circle") +
  geom_edge_link(aes(color = sign, linetype = sign), width = 2) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = 1:5), fontface = "bold", size = 5.5, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  coord_fixed(xlim = c(-1.4, 1.4), ylim = c(-1.4, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## The Fundamental Theorem of Structural Balance

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- Discovered by mathematicians Dorwin **Cartwright** and Frank **Harary** in 1956.
- **Theorem**: If a complete signed graph contains *only* balanced triads, then the entire graph is balanced.
- **The Partition Rule**: Any completely balanced signed graph can be divided into exactly **two polarized factions** such that:
  1. Only positive links exist *within* each faction.
  2. Only negative links exist *between* the two factions.
- Shows that structural balance is equivalent to perfect **group polarization**.
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 6
#| fig.height: 5
nodes <- data.frame(name = LETTERS[1:6], faction = c(rep("1", 3), rep("2", 3)))
edges <- expand.grid(from = LETTERS[1:6], to = LETTERS[1:6], stringsAsFactors = FALSE) %>%
  filter(from < to) %>%
  mutate(
    f_from = nodes$faction[match(from, nodes$name)],
    f_to = nodes$faction[match(to, nodes$name)],
    sign = ifelse(f_from == f_to, "+", "-")
  )
g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
lay <- matrix(c(-1.5, 0.8, -1.5, -0.8, -2.5, 0, 1.5, 0.8, 1.5, -0.8, 2.5, 0), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), width = 2, alpha = 0.8) +
  geom_node_point(aes(fill = faction), size = 22, color = "white", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "white") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  scale_fill_manual(values = c("1" = "steelblue", "2" = "darkorchid")) +
  coord_fixed(xlim = c(-3.0, 3.0), ylim = c(-1.3, 1.3), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## Signed Matrices & Faction Detection

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- How can we mathematically recover and identify factions in a balanced network?
- Rearrange the rows and columns of the **signed adjacency matrix** such that nodes in the same faction are adjacent.
- This permutation yields a **block-diagonal signed matrix**.
- Inside the diagonal blocks, cells are $+1$ (internal cohesion), whereas off-diagonal blocks are entirely $-1$ (external conflict).
:::
:::

::: {.column width="50%"}
### Faction Block-Diagonal Matrix
$$M = \begin{pmatrix}
  \begin{array}{ccc|ccc}
    \mathbf{+} & \mathbf{+} & \mathbf{+} & - & - & - \\
    \mathbf{+} & \mathbf{+} & \mathbf{+} & - & - & - \\
    \mathbf{+} & \mathbf{+} & \mathbf{+} & - & - & - \\ \hline
    - & - & - & \mathbf{+} & \mathbf{+} & \mathbf{+} \\
    - & - & - & \mathbf{+} & \mathbf{+} & \mathbf{+} \\
    - & - & - & \mathbf{+} & \mathbf{+} & \mathbf{+}
  \end{array}
\end{pmatrix}$$
:::
:::

---

## Cycles and Paths in Balanced Graphs

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- Partitioning has direct, rigid consequences for paths and cycles in complete graphs:
- **Cycles**: *All cycles in a balanced signed graph must have a positive sign product.*
  - Staying inside a group utilizes only positive links.
  - Spanning across groups requires crossing the boundary an *even* number of times, multiplying an even number of negative signs to yield a positive product.
- **Paths**:
  - Paths ending in the *same* group have a **positive product**.
  - Paths ending in *different* groups have a **negative product** (due to odd boundary crossings).
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 6
#| fig.height: 5
edges <- data.frame(
  from = c("Start", "Node B", "Node C", "End"),
  to   = c("Node B", "Node C", "End", "Start"),
  sign = c("+", "-", "+", "-")
)
g <- graph_from_data_frame(edges, directed = FALSE)
lay <- matrix(c(-1.5, 0.8, -1.5, -0.8, 1.5, -0.8, 1.5, 0.8), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), width = 2.5) +
  geom_node_point(aes(fill = name %in% c("Start", "Node B")), size = 20, color = "#2c3e50", shape = 21, stroke = 1.8) +
  geom_node_text(aes(label = name), fontface = "bold", size = 4.5, color = "#2c3e50") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  scale_fill_manual(values = c("TRUE" = "#93c5fd", "FALSE" = "#c084fc")) +
  coord_fixed(xlim = c(-2.0, 2.0), ylim = c(-1.3, 1.3), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## Empirical Case Study: Political Blogs

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- Balanced states are **not** inherently "harmonious" or socially desirable.
- They describe stable **polarization and conflict**.
- **Adamic & Glance (2005)** examined the hyperlinking network of political bloggers during the 2004 election.
- Discovered two massive, highly connected factions (Liberals and Conservatives).
- Bloggers linked supportive-positive links inside their group, but hostile-negative links across the political divide, establishing an **echo chamber** structure.
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 6
#| fig.height: 5
set.seed(422)
g1 <- play_erdos_renyi(n = 6, p = 0.8, directed = FALSE)
g2 <- play_erdos_renyi(n = 6, p = 0.8, directed = FALSE)
g <- disjoint_union(g1, g2)
V(g)$faction <- c(rep("Liberal", 6), rep("Conservative", 6))
g <- add_edges(g, c(1,7, 3,9, 5,11))
E(g)$sign <- c(rep("+", ecount(g)-3), rep("-", 3))
ggraph(g, layout = "kk") +
  geom_edge_link(aes(color = sign, linetype = sign), width = 2, alpha = 0.8) +
  geom_node_point(aes(fill = faction), size = 14, color = "white", shape = 21, stroke = 1.5) +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  scale_fill_manual(values = c("Liberal" = "#3b82f6", "Conservative" = "#ef4444")) +
  coord_fixed(clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

# Part 4: Clusterability and Multi-Group Partitioning

---

## Beyond Two Factions: Davis's Theorem

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- Cartwright & Harary's model splits networks into exactly *two* opposing factions.
- This is too restrictive for many real-world networks, such as school cliques, gangs, or multi-party politics.
- In 1967, sociologist **James Davis** introduced his **Clusterability Theorem**.
- He generalized structural balance so that a network could fragment into **any number of distinct clusters** ($k \ge 1$), rather than just two.
:::
:::

::: {.column width="45%"}
```{r}
#| fig.width: 6
#| fig.height: 5
nodes <- data.frame(name = LETTERS[1:9], faction = c(rep("1", 3), rep("2", 3), rep("3", 3)))
edges <- expand.grid(from = LETTERS[1:9], to = LETTERS[1:9], stringsAsFactors = FALSE) %>%
  filter(from < to) %>%
  mutate(
    f_from = nodes$faction[match(from, nodes$name)],
    f_to = nodes$faction[match(to, nodes$name)],
    sign = ifelse(f_from == f_to, "+", "-")
  )
g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
lay <- matrix(c(
  -0.35, 2.0, -0.65, 1.5, 0, 1.5,  # Cluster 1
  -1.0, 0.5, -1, 0, -0.65, -0.2,   # Cluster 2
  0.5, -0.5, 0, -0.5, 0.5, 0       # Cluster 3
), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), width = 1.8, alpha = 0.7) +
  geom_node_point(aes(fill = faction), size = 20, color = "white", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 5, color = "white") +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  scale_fill_manual(values = c("1" = "steelblue", "2" = "darkorchid", "3" = "goldenrod")) +
  coord_fixed(xlim = c(-1.4, 0.9), ylim = c(-0.9, 2.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## The All-Negative Triad Rule

::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- Classical balance considers a triad with three negative ties (everyone hating everyone else) to be *unbalanced*.
- Davis argued that in highly factionalized contexts, it is perfectly stable for three individuals from completely different, competing groups to dislike one another.
- **The Clustering Criterion**: A graph is **clusterable** if and only if it contains **no triads with exactly one negative tie**.
- Permitted triads under Davis: all-positive ($+1$), two-negatives ($+1$), or all-negatives ($-1$).
:::
:::

::: {.column width="50%"}
```{r}
#| fig.width: 5
#| fig.height: 5
edges <- data.frame(from = c("A", "B", "C"), to = c("B", "C", "A"), sign = c("-", "-", "-"))
g <- graph_from_data_frame(edges, directed = FALSE)
lay <- matrix(c(0, 1, -0.86, -0.5, 0.86, -0.5), ncol = 2, byrow = TRUE)
ggraph(g, layout = lay) +
  geom_edge_link(aes(color = sign, linetype = sign), width = 2.5, end_cap = circle(10, "mm"), start_cap = circle(10, "mm")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), fontface = "bold", size = 6, color = "#2c3e50") +
  scale_edge_color_manual(values = c("-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("-" = "33")) +
  coord_fixed(xlim = c(-1.3, 1.3), ylim = c(-0.9, 1.4), clip = "off") +
  theme_slides() + theme(legend.position = "none")
```
:::
:::

---

## Summary: Structural Balance vs. Clusterability

::: {.columns}
::: {.column width="40%"}
::: {.incremental}
- **Structural Balance (Cartwright-Harary 1956)**:
  - Local rule: Triads cannot have an odd number of negative ties.
  - Global topology: Splits the network into exactly **two factions**.
- **Clusterability (Davis 1967)**:
  - Local rule: Relaxes constraints by permitting all-negative triads (no triads with exactly one negative tie).
  - Global topology: Splits the network into **any number of clusters ($k \ge 1$)**.
:::
:::

::: {.column width="60%"}
```{r}
#| fig.width: 8
#| fig.height: 5
# Visual summary comparing structural balance vs clusterability side-by-side
nodes_b <- data.frame(name = LETTERS[1:6], faction = c(rep("1", 3), rep("2", 3)))
edges_b <- expand.grid(from = LETTERS[1:6], to = LETTERS[1:6], stringsAsFactors = FALSE) %>%
  filter(from < to) %>%
  mutate(
    f_from = nodes_b$faction[match(from, nodes_b$name)],
    f_to = nodes_b$faction[match(to, nodes_b$name)],
    sign = ifelse(f_from == f_to, "+", "-")
  )
g_b <- graph_from_data_frame(edges_b, directed = FALSE, vertices = nodes_b)
lay_b <- matrix(c(-1.5, 0.8, -1.5, -0.8, -2.5, 0, 1.5, 0.8, 1.5, -0.8, 2.5, 0), ncol = 2, byrow = TRUE)

p_bal <- ggraph(g_b, layout = lay_b) +
  geom_edge_link(aes(color = sign, linetype = sign), width = 1.2, alpha = 0.6) +
  geom_node_point(aes(fill = faction), size = 14, color = "white", shape = 21, stroke = 1.2) +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  scale_fill_manual(values = c("1" = "steelblue", "2" = "darkorchid")) +
  coord_fixed(xlim = c(-3.0, 3.0), ylim = c(-1.3, 1.3), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("Structural Balance (2 Factions)")

nodes_c <- data.frame(name = LETTERS[1:9], faction = c(rep("1", 3), rep("2", 3), rep("3", 3)))
edges_c <- expand.grid(from = LETTERS[1:9], to = LETTERS[1:9], stringsAsFactors = FALSE) %>%
  filter(from < to) %>%
  mutate(
    f_from = nodes_c$faction[match(from, nodes_c$name)],
    f_to = nodes_c$faction[match(to, nodes_c$name)],
    sign = ifelse(f_from == f_to, "+", "-")
  )
g_c <- graph_from_data_frame(edges_c, directed = FALSE, vertices = nodes_c)
lay_c <- matrix(c(
  -0.35, 2.0, -0.65, 1.5, 0, 1.5,  # Cluster 1
  -1.0, 0.5, -1, 0, -0.65, -0.2,   # Cluster 2
  0.5, -0.5, 0, -0.5, 0.5, 0       # Cluster 3
), ncol = 2, byrow = TRUE)

p_clust <- ggraph(g_c, layout = lay_c) +
  geom_edge_link(aes(color = sign, linetype = sign), width = 1.2, alpha = 0.6) +
  geom_node_point(aes(fill = faction), size = 14, color = "white", shape = 21, stroke = 1.2) +
  scale_edge_color_manual(values = c("+" = "forestgreen", "-" = "firebrick")) +
  scale_edge_linetype_manual(values = c("+" = "solid", "-" = "33")) +
  scale_fill_manual(values = c("1" = "steelblue", "2" = "darkorchid", "3" = "goldenrod")) +
  coord_fixed(xlim = c(-1.4, 0.9), ylim = c(-0.9, 2.4), clip = "off") +
  theme_slides() + theme(legend.position = "none") + ggtitle("Clusterability (k Clusters)")

p_bal + p_clust + plot_layout(nrow = 1)
```
:::
:::
