---
title: "Graph Theory: Blockmodeling"
subtitle: "Role Structure and Block Relations"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(corrplot)

# Custom correlation function from textbook
corr.dist <- function(x) {
  r <- nrow(x)
  c <- ncol(x)
  r.c <- matrix(0, r, r)
  r.m <- rowMeans(x)
  for (i in 1:r) {
    for (j in 1:r) {
      r.x <- x[i, ] - r.m[i]
      r.y <- x[j, ] - r.m[j]
      r.xy <- r.x * r.y
      r.xx <- r.x^2
      r.yy <- r.y^2
      r.num <- sum(r.xy)
      r.den <- sqrt(sum(r.xx)) * sqrt(sum(r.yy))
      r.c[i, j] <- round(r.num / r.den, 2)
    }
  }
  rownames(r.c) <- rownames(x)
  colnames(r.c) <- rownames(x)
  return(r.c)
}

# Standard 6-node undirected graph
gr_undir <- create_empty(6, directed = FALSE) %>%
  bind_edges(data.frame(from = c(1, 1, 2, 2, 3, 3, 5, 6), to = c(3, 4, 3, 4, 1, 2, 3, 5))) %>%
  mutate(name = toupper(letters[1:6]))
gr_undir <- simplify(gr_undir)

# Big 22-node network for CONCOR demonstration
fr <- c(rep(1, 5), rep(2, 5), rep(3, 5), rep(4, 3), rep(20, 3))
to <- c(5:9, 10:14, 15:19, 1:3, 5, 21, 22)
edge_dat <- data.frame(fr, to)
node_dat <- data.frame(name = toupper(letters[union(fr, to)]))
gr_big <- tbl_graph(edges = edge_dat, nodes = node_dat, directed = FALSE)
gr_big <- as_tbl_graph(simplify(gr_big))
```

# Part 1: The Correlation Distance & CONCOR

---

## The Correlation Distance ($d^{corr}_{k,l}$)

::: {.incremental}
- Beyond Euclidean distance, we can measure structural similarity using the **Pearson Correlation Coefficient** ($r_{kl}$ or $d^{corr}_{kl}$) between node vectors.
- Compares row (or column) vectors and returns a value between **$-1.0$ and $+1.0$**:
  - **$+1.0$**: Nodes are perfectly structurally equivalent.
  - **Positive ($> 0.0$)**: Nodes are structurally similar.
  - **Negative ($< 0.0$)**: Nodes are structurally dissimilar.
  - **$-1.0$**: Nodes are completely different, pointing to completely disjoint sets of neighbors.
- The formula is:
  - $$d^{corr}_{k, l} = \frac{\sum_j (a_{(k)j} - \overline{a}_{(k)j}) \times (a_{(l)j} - \overline{a}_{(l)j})}{\sqrt{\sum_j (a_{(k)j} - \overline{a}_{(k)j})^2 \times \sum_j (a_{(l)j} - \overline{a}_{(l)j})^2}}$$
:::

---

## Meaning of Row Means ($\overline{a}_{(k)j}$)

::: {.incremental}
- What does $\overline{a}_{(k)j}$ represent?
  - It is simply the **average of the row elements** for node $k$.
  - $$\overline{a}_{(k)j} = \frac{\sum_j a_{kj}}{N}$$
  - Where $N$ is the number of columns (the order of the graph).
- For a binary adjacency matrix, the row mean is simply the **node degree divided by the number of nodes**:
  - Node $A$ has ties to $\{C, D\}$ in a 6-node network:
  - Vector $a_{(A)j} = (0, 0, 1, 1, 0, 0)$.
  - Average: $\overline{a}_{(A)j} = \frac{2}{6} \approx 0.33$.
:::

---

## Mean Differences Vector

- Subtracting the row mean $\overline{a}_{(A)j} \approx 0.33$ from each entry of $A$'s row vector $a_{(A)j} = (0, 0, 1, 1, 0, 0)$ gives the Mean Differences vector:

```{r}
#| echo: false
A_entries <- c(0, 0, 1, 1, 0, 0)
diff_lbl <- c("(0 - 0.33)", "(0 - 0.33)", "(1 - 0.33)", "(1 - 0.33)", "(0 - 0.33)", "(0 - 0.33)")
diff_val <- c(-0.33, -0.33, 0.67, 0.67, -0.33, -0.33)
mean_diff_mat <- matrix(c(as.character(A_entries), diff_lbl, as.character(diff_val)), nrow = 3, byrow = TRUE)
rownames(mean_diff_mat) <- c("Row A", "A - Mean", "Difference")
colnames(mean_diff_mat) <- c("A", "B", "C", "D", "E", "F")
kbl(mean_diff_mat, format = "html", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

- We compute these difference vectors for all pairs of nodes, multiply them in the numerator, normalize them in the denominator, and obtain the correlation matrix.

---

## Correlation Distance Matrix ($S^{corr}$)

```{r}
#| echo: false
adj_mat_6 <- matrix(as_adjacency_matrix(gr_undir), nrow = length(V(gr_undir)))
rownames(adj_mat_6) <- V(gr_undir)$name
colnames(adj_mat_6) <- V(gr_undir)$name
adj_mat_6 <- adj_mat_6[order(row.names(adj_mat_6)), order(row.names(adj_mat_6))]
cor_mat_6 <- corr.dist(adj_mat_6)
diag(cor_mat_6) <- "--"

kbl(cor_mat_6, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"))
```

- Structurally equivalent pairs ($\{A, B\}$ and $\{C, D\}$) have a correlation of **1.00**.
- Completely disjoint pairs ($\{C, E\}$ and $\{D, E\}$) are perfectly negatively correlated (**-1.00**).
- Similar but non-equivalent nodes ($C$ and $F$) have a correlation of **0.45**.

---

## Iterated Correlation Distances: CONCOR

::: {.incremental}
- What happens if we take the correlation distance of a correlation distance matrix?
- **CONCOR** (Convergence of Iterated Correlations; Breiger, Boorman, & Arabie, 1975):
  - If we recursively calculate the correlation distance matrix of the previous correlation matrix, the values mathematically **converge to only $+1.0$ and $-1.0$**.
  - This slices the network into two distinct, highly clear-cut blocks of structurally equivalent or similar nodes.
- Let's look at this mathematical convergence in action.
:::

---

## Iterative Convergence (2nd & 3rd Iterations)

:::: {.columns}

::: {.column width="50%"}
### 2nd Iteration Matrix
- Values begin polarizing toward $+1$ and $-1$.
```{r}
#| echo: false
diag_clean <- function(mat) {
  diag(mat) <- "--"
  return(mat)
}
cor_6_2 <- corr.dist(corr.dist(adj_mat_6))
kbl(diag_clean(cor_6_2), format = "html", align = "c") %>%
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 12)
```
:::

::: {.column width="50%"}
### 3rd Iteration Matrix
- Every cell reaches **exactly $+1.00$ or $-1.00$**!
```{r}
#| echo: false
cor_6_3 <- corr.dist(cor_6_2)
kbl(diag_clean(cor_6_3), format = "html", align = "c") %>%
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 12)
```
:::

::::

---

## Structural Positions via CONCOR

```{r}
#| echo: false
kbl(cor_6_3, format = "html", align = "c") %>%
  row_spec(1:2, background = "salmon", bold = TRUE) %>%
  row_spec(3:4, background = "lightgreen", bold = TRUE) %>%
  row_spec(5, background = "cyan", bold = TRUE) %>%
  row_spec(6, background = "violet", bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"))
```

::: {.incremental}
- By analyzing the rows of $+1$ and $-1$, we extract perfect structurally equivalent positions:
  - Position 1 (Salmon): Nodes $\{A, B\}$
  - Position 2 (Green): Nodes $\{C, D\}$
  - Position 3 (Blue): Node $\{E\}$
  - Position 4 (Purple): Node $\{F\}$
:::

# Part 2: Blockmodeling a 22-Node Network

---

## Uncovering Secret Patterns in Large Networks

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Let's apply CONCOR to a larger, messier 22-node network on the right.
- It contains:
  - Three high-degree hub nodes ($A$, $B$, $C$).
  - Sets of identical peripheral leaf nodes connected to each hub.
  - A highly central node $D$ bridging hubs.
  - Two peripheral nodes $U, V$ connected only to $S$.
- Let's see how CONCOR recursively partitions this graph.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_big, layout = 'kk') + 
  geom_edge_link(color = "steelblue", edge_width = 1.15) + 
  geom_node_point(color = "tan2", size = 16) +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Step 1: First Split (Two Giant Blocks)

- Running CONCOR on the whole 22-node network splits it into two massive blocks:
  - **Block 1**: $\{V, U, S, H, G, F, E, T, C, A, B\}$ (Colored Red)
  - **Block 2**: $\{R, Q, P, O, N, M, L, K, J, D, I\}$ (Colored Blue)

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 4.8
node_color_split1 <- c(rep("firebrick",3), "darkblue", "firebrick", rep("firebrick", 4), rep("darkblue", 10), "firebrick", rep("firebrick", 2))
ggraph(gr_big, layout = 'kk') + 
  geom_edge_link(color = "steelblue", alpha = 0.5, edge_width = 1.0) + 
  geom_node_point(color = node_color_split1, size = 12) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## Step 2: Going Deeper (The Hierarchical Partition)

::: {.incremental}
- Running CONCOR recursively on the induced subgraphs of each block yields nested subdivisions:
  - **Block 1 Splits**:
    - **Block 1a** (Core hub/bridge): $\{A, B, C, S, U, V\}$
    - **Block 1b** (A's peripheral leaves): $\{E, F, G, H, T\}$
  - **Block 2 Splits**:
    - **Block 2a** (B's peripheral leaves): $\{I, J, K, L, M\}$
    - **Block 2b** (C's peripheral leaves): $\{N, O, P, Q, R\}$
    - **Block 2c** (Unique central bridge): $\{D\}$ (Occupies a block entirely by itself!)
- Further splits on Block 1a isolate $\{U, V\}$, and divide the remaining $\{A, B, C, S\}$ into $\{A, S\}$ and $\{B, C\}$.
:::

---

## Final Seven-Block Partition

- The final, elegant block partition reveals how role structures are naturally isolated:

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5.5
node_color_final <- c("firebrick", rep("darkblue", 2), "purple", rep("tan1", 5), rep("darkgreen", 5), rep("#CC79A7", 5), "firebrick", rep("#56B4E9", 2))
ggraph(gr_big, layout = 'kk') + 
  geom_edge_link(color = "steelblue", alpha = 0.5, edge_width = 1.0) + 
  geom_node_point(color = node_color_final, size = 14) +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## The Blocked Adjacency Matrix

```{r}
#| echo: false
adj_mat_big <- matrix(as_adjacency_matrix(gr_big), nrow = length(V(gr_big)))
rownames(adj_mat_big) <- V(gr_big)$name
colnames(adj_mat_big) <- V(gr_big)$name
final_blocks_order <- c("I", "J", "K", "L", "M", "N", "O", "P", "Q", "R", "T", "E", "F", "G", "H", "B", "C", "A", "S", "U", "V", "D")
blocked_adj <- adj_mat_big[final_blocks_order, final_blocks_order]

kbl(blocked_adj, align = "c", format = "html") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(5, extra_css = "border-bottom: 2px solid black;") %>%  
  row_spec(10, extra_css = "border-bottom: 2px solid black;") %>%  
  row_spec(15, extra_css = "border-bottom: 2px solid black;") %>%  
  row_spec(17, extra_css = "border-bottom: 2px solid black;")  %>% 
  row_spec(19, extra_css = "border-bottom: 2px solid black;")  %>% 
  row_spec(21, extra_css = "border-bottom: 2px solid black;") %>% 
  column_spec(6, extra_css = "border-right: 2px solid black;") %>% 
  column_spec(11, extra_css = "border-right: 2px solid black;") %>%  
  column_spec(16, extra_css = "border-right: 2px solid black;") %>%  
  column_spec(18, extra_css = "border-right: 2px solid black;") %>%  
  column_spec(20, extra_css = "border-right: 2px solid black;") %>% 
  column_spec(22, extra_css = "border-right: 2px solid black;")  
```

::: {.incremental}
- **Thick black lines** divide the matrix into block submatrices.
- Diagonal squares represent **within-block** ties; off-diagonal rectangles represent **between-block** ties.
- We identify **Zero Blocks** (filled only with 0s) and **One Blocks** (containing some 1s).
:::

# Part 3: Image Matrices & Reduced Graphs

---

## Reducing Complexity to an Image Matrix

::: {.incremental}
- The blocked adjacency matrix can be simplified to a **reduced image matrix** ($\mathbf{B}$) of dimensions $b \times b$ (where $b = 7$ is the number of blocks).
- **Rule of Reduction**:
  - We record a **0** if the relation block is a **Zero Block** (completely disconnected).
  - We record a **1** if the relation block is a **One Block** (contains ties).
- This yields a highly simplified map of role interactions:
  - $$
    \mathbf{B} = 
    \begin{pmatrix}
    \text{Zero} & \text{Zero} & \text{Zero} & \text{One} & \text{Zero} & \text{Zero} & \text{Zero} \\
    \text{Zero} & \text{Zero} & \text{Zero} & \text{One} & \text{Zero} & \text{Zero} & \text{Zero} \\
    \text{Zero} & \text{Zero} & \text{Zero} & \text{Zero} & \text{One} & \text{Zero} & \text{Zero} \\
    \text{One} & \text{One} & \text{Zero} & \text{Zero} & \text{Zero} & \text{Zero} & \text{One} \\
    \text{Zero} & \text{Zero} & \text{One} & \text{Zero} & \text{Zero} & \text{One} & \text{One} \\
    \text{Zero} & \text{Zero} & \text{Zero} & \text{Zero} & \text{One} & \text{Zero} & \text{Zero} \\
    \text{Zero} & \text{Zero} & \text{Zero} & \text{One} & \text{One} & \text{Zero} & \text{Zero}
    \end{pmatrix}
    $$
:::

---

## Image Matrix Table

```{r}
#| echo: false
image_mat <- matrix(c(
  0, 0, 0, 1, 0, 0, 0,
  0, 0, 0, 1, 0, 0, 0,
  0, 0, 0, 0, 1, 0, 0,
  1, 1, 0, 0, 0, 0, 1,
  0, 0, 1, 0, 0, 1, 1,
  0, 0, 0, 0, 1, 0, 0,
  0, 0, 0, 1, 1, 0, 0
), nrow = 7, byrow = TRUE)
block_names <- c("Leaves_I", "Leaves_N", "Leaves_T", "Hubs_BC", "Hubs_AS", "Leaves_UV", "Bridge_D")
rownames(image_mat) <- block_names
colnames(image_mat) <- block_names

kbl(image_mat, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

- High-level representation of block relations.
- Notice that Leaf blocks do not connect to themselves (0s on diagonal), but connect specifically to their respective Hub blocks.

---

## Reduced Graph Representation

:::: {.columns align=center}

::: {.column width="45%"}
::: {.incremental}
- We can plot the image matrix as a **reduced graph**.
- **The Power of Blockmodeling**:
  - Distills a sprawling, complex network of 22 nodes down to a highly intuitive role-interaction model of just **7 block-nodes**.
  - Revealing the structural blueprint of the system: peripheral elements (leaves) feeding into coordinated brokers (hubs), bridged by a highly central node ($D$).
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6.5
#| fig-height: 5.5
gr_image <- graph_from_adjacency_matrix(image_mat) %>% 
  as_tbl_graph() %>% 
  activate(nodes) %>% 
  mutate(names = block_names)

ggraph(gr_image, layout = 'kk') + 
  geom_edge_link(color = "steelblue", edge_width = 1.15) + 
  geom_node_label(aes(label = names), size = 4, fontface = "bold") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Blockmodeling: A Structural Summary

::: {.incremental}
- **Blockmodeling Distills Structure**:
  - Slices complex social structures down to high-level system blueprints.
- **CONCOR algorithm**:
  - Recursively polarizes Pearson correlations to group similar actors.
- **Zero vs. One Blocks**:
  - Blocked adjacency matrices cleanly isolate patterns of connection (One blocks) and absolute separation (Zero blocks).
- **Reduced Graph Models**:
  - Image matrices represent relations *between roles*, creating simplified reduced graphs that make networks easy to read, interpret, and model.
:::
