---
title: "Chains of Affection"
subtitle: "How Local Norms Shape Romantic and Sexual Networks"
format:
  revealjs:
    incremental: true
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(patchwork)
library(kableExtra)

set.seed(333)

# 1. Bearman's Jefferson High Spanning Tree (Chain-like)
# Generate a beautiful spanning tree-like random tree of 20 nodes
g_tree <- tbl_graph(nodes = data.frame(name = as.character(1:20)),
                    edges = data.frame(from = c(1,1,2,2,3,4,4,5,6,7,7,8,9,10,11,12,12,13,14),
                                       to   = c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20)),
                    directed = FALSE)

p_jefferson_spanning <- ggraph(g_tree, layout = "tree") +
  geom_edge_link(color = "steelblue", width = 1.5) +
  geom_node_point(size = 8, color = "tan2") +
  geom_node_text(aes(label = name), size = 4, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")

# 2. Bearman's 4-cycle loop (A-B-C-D) vs Spanning Tree
# Left: A-B-C-D-A (4-cycle loop, the taboo!)
g_cycle <- tbl_graph(nodes = data.frame(name = c("Bob", "Carol", "Ted", "Alice")),
                     edges = data.frame(from = c(1,2,3,4), to = c(2,3,4,1)), directed = FALSE)
p_cycle_loop <- ggraph(g_cycle, layout = "circle") +
  geom_edge_link(color = "firebrick", width = 2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "The Tabooed 4-Cycle") + theme_graph() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12)) +
  coord_cartesian(xlim = c(-1.2, 1.2), ylim = c(-1.2, 1.2), clip = "off")

# Right: Bob-Carol-Ted (spanning chain, no Bob-Alice tie)
g_chain <- tbl_graph(nodes = data.frame(name = c("Bob", "Carol", "Ted", "Alice")),
                     edges = data.frame(from = c(1,2,3), to = c(2,3,4)), directed = FALSE)
p_chain_spanning <- ggraph(g_chain, layout = "circle") +
  geom_edge_link(color = "steelblue", width = 2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "The Permitted Spanning Chain") + theme_graph() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12)) +
  coord_cartesian(xlim = c(-1.2, 1.2), ylim = c(-1.2, 1.2), clip = "off")

# 3. McMillan's same-sex friendship code (Girl Code / Bro Code)
# Left: Mixed-sex romantic relationship between i and j, same-sex friendships i-k and j-l
edges_peer1 <- data.frame(from = c("i", "i", "j"), to = c("j", "k", "l"), type = c("Romantic", "Friend", "Friend"))
g_peer1 <- tbl_graph(nodes = data.frame(name = c("i", "j", "k", "l")), edges = edges_peer1, directed = FALSE)
l_peer <- matrix(c(-0.5, 0.5, 0.5, 0.5, -0.5, -0.5, 0.5, -0.5), ncol = 2, byrow = TRUE)
p_peer1 <- ggraph(g_peer1, layout = "manual", x = l_peer[,1], y = l_peer[,2]) +
  geom_edge_link(aes(color = type, linetype = type), width = 1.8) +
  scale_edge_color_manual(values = c("Romantic" = "firebrick", "Friend" = "steelblue")) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Phase 1: Dating & Friendships") + theme_graph() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11), legend.position = "none") +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8), clip = "off")

# Middle: Closure: j befriends k, i befriends l (double-dating)
edges_peer2 <- data.frame(from = c("i", "i", "j", "j", "i"), to = c("j", "k", "l", "k", "l"), type = c("Romantic", "Friend", "Friend", "Friend", "Friend"))
g_peer2 <- tbl_graph(nodes = data.frame(name = c("i", "j", "k", "l")), edges = edges_peer2, directed = FALSE)
p_peer2 <- ggraph(g_peer2, layout = "manual", x = l_peer[,1], y = l_peer[,2]) +
  geom_edge_link(aes(color = type, linetype = type), width = 1.8) +
  scale_edge_color_manual(values = c("Romantic" = "firebrick", "Friend" = "steelblue")) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Phase 2: Triadic Closure") + theme_graph() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11), legend.position = "none") +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8), clip = "off")

# Right: Breakup: i and j break up. j cannot date k (friend's ex). i cannot date l (friend's ex).
# Results in two disjoint chains, preventing 4-cycle!
edges_peer3 <- data.frame(from = c("i", "j"), to = c("k", "l"), type = c("Friend", "Friend"))
g_peer3 <- tbl_graph(nodes = data.frame(name = c("i", "j", "k", "l")), edges = edges_peer3, directed = FALSE)
p_peer3 <- ggraph(g_peer3, layout = "manual", x = l_peer[,1], y = l_peer[,2]) +
  geom_edge_link(aes(color = type, linetype = type), width = 1.8) +
  scale_edge_color_manual(values = c("Friend" = "steelblue")) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Phase 3: Friend-Loyalty Split") + theme_graph() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11), legend.position = "none") +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8), clip = "off")

# 4. Fictional Networks Glee & Grey's Anatomy (Highly clustered 4-cycles)
# Small highly dense cluster with multiple overlapping 4-cycles (e.g. A-B-C-D, B-C-E-F etc.)
g_glee <- tbl_graph(nodes = data.frame(name = c("Finn", "Rachel", "Quinn", "Puck", "Santana", "Brittany")),
                    edges = data.frame(from = c(1,1,2,3,4,4,5,5),
                                       to   = c(2,3,4,4,5,6,6,1)),
                    directed = FALSE)
p_glee <- ggraph(g_glee, layout = "circle") +
  geom_edge_link(color = "purple", width = 1.8) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.2, 1.2), ylim = c(-1.2, 1.2), clip = "off")

# 5. Stoddard's Rumor Networks
# Left: Confirmed network (clean tree, acyclic)
g_conf <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]),
                    edges = data.frame(from = c(1,1,2,3), to = c(2,3,4,5)),
                    directed = FALSE)
p_conf <- ggraph(g_conf, layout = "circle") +
  geom_edge_link(color = "steelblue", width = 1.8) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  labs(title = "Confirmed Relationships (Acyclic Tree)") + theme_graph() +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  coord_cartesian(xlim = c(-1.2, 1.2), ylim = c(-1.2, 1.2), clip = "off")

# Right: Rumored network (dense with short cycles added by gossip)
g_rumor <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]),
                     edges = data.frame(from = c(1,1,2,3,2,4), to = c(2,3,4,5,3,5)),
                     directed = FALSE)
p_rumor <- ggraph(g_rumor, layout = "circle") +
  geom_edge_link(aes(color = (from == 2 & to == 3) | (from == 4 & to == 5)), width = 1.8) +
  scale_edge_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  labs(title = "With Rumored Ties (Cycles Formed)") + theme_graph() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14), legend.position = "none") +
  coord_cartesian(xlim = c(-1.2, 1.2), ylim = c(-1.2, 1.2), clip = "off")
```

# Part 1: The Macro-Micro Mystery

---

## Why Study Romantic & Sexual Networks?

- **The Closed Laboratory**:
  - Unlike abstract communities, school-based dating and sexual networks are **bounded settings** with high social visibility.
  - Individual relationship preferences are intimate, dyadic choices, but when aggregated, they form a **global network structure**.
- **The Core Paradox**:
  - Individual actors have **zero awareness of the global structure**.
  - No adolescent says: *"I will date this person to maximize network centralization."*
  - Yet, by following simple, local **micro-level rules**, individuals generate elegant, system-wide **macro-level patterns**.

---

## Macro-Structures & Disease Diffusion

- **The Stakes of Topology**:
  - The path layout of a network determines how rapidly **pathogens (like STIs/HIV)** can travel.
  - **The Core Model**: Densely connected clusters or pockets of high-risk actors. Highly resilient to intervention.
  - **The Spanning Tree**: Long, thin, fragile chains with very few shortcuts (cycles). Extremely vulnerable to random interventions.
- **The Micro-Macro Link**:
  - By understanding the **unwritten local rules** of adolescent society, we can predict—and break—the path of disease transmission.

---

# Part 2: Bearman's Landmark Study (Jefferson High)

---

## The "Chains of Affection" Discovery

::: {.columns}
::: {.column width="60%"}
- **Peter Bearman, James Moody, and Katherine Stovel (2004)**:
  - Conducted a landmark study of "Jefferson High" (832 students).
- **The Global Finding**:
  - Found a massive **spanning tree** structure!
  - Over **52%** of active students were linked in a single giant, chain-like component.
  - It had **almost zero cycles** (redundant loops), resembling rural telephone wires stretching from a single main trunk.
- **Why is this a puzzle?**
  - Standard social preferences like homophily (dating similar people) predict highly clustered, dense, closed pockets—not long, fragile chains!
:::

::: {.column width="40%"}
### Spanning Tree Chain
<center>
```{r}
#| echo: false
#| fig-height: 4.8
#| fig-width: 4.8
p_jefferson_spanning
```
</center>
:::
:::

---

## The 4-Cycle Proscription: The Tabooed Loop

::: {.columns}
::: {.column width="50%"}
- **What is a 4-Cycle?**
  - A closed four-node loop where Bob dates Carol, Carol dates Ted, Ted dates Alice, and Alice dates Bob.
- **The Jefferson High Taboo**:
  - Bearman et al. found only **two 4-cycles** in the entire high school, vastly fewer than expected by random chance.
- **The Micro-Level Proscription**:
  - *"You do not date your ex's current partner's ex."*
  - Complete aversion to **seconds partnerships**.
- **Sociological Explanation**:
  - Completing this loop results in a public **loss of status**. Adolescents avoid taking "seconds" to protect their prestige in the peer group.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.2
#| fig-width: 4.2
p_cycle_loop
```
- Closing this loop violates local status rules!
</center>
:::
:::

---

## The Power of a Negative Rule

::: {.columns}
::: {.column width="55%"}
- **Spanning Trees are Born from Proscriptions**:
  - Spanning trees (chain-like structures) are not generated by positive choices, but by **negative rules** (what you *cannot* do).
  - Like incest taboos in kinship networks, the prohibition against completing short cycles forces individuals to look outward.
- **The Spanning Result**:
  - By refusing to close the loop, Bob and Alice must find partners outside their immediate circle.
  - This simple choice **stretches the network out** into long, fragile chains.
:::

::: {.column width="45%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.2
#| fig-width: 4.2
p_chain_spanning
```
- ForcingBob and Alice to look outward stretches the tree!
</center>
:::
:::

---

# Part 3: Peer Context & The Friendship-Dating Code

---

## "Keeping to the Code" (McMillan et al., 2022)

- **The Critique of the Classic Taboo**:
  - McMillan, Kreager, and Veenstra (2022) argue that Bearman's norm lacks **face validity**.
  - No high schooler actually thinks in terms of: *"Is this my ex-partner's current partner's ex-partner?"*
  - It is too abstract and unarticulated.
- **The Alternative Multi-Relational Norm**:
  - Romantic relationships are embedded in a dense **peer friendship network**.
  - The actual micro-level rule is: **The "No Seconds" Friendship Code** (colloquially, **"Girl Code"** or **"Bro Code"**).
  - *"Do not date your friend's previous romantic partner."*

---

## Visualizing the Friendship Code Process

- **How same-sex loyalty blocks the 4-cycle loop step-by-step**:

<center>
```{r}
#| echo: false
#| fig-height: 3.2
#| fig-width: 9.6
# Convert first peer friendship code plot
# wait, 'steelblue' in p_peer1 was a typo for 'steelblue'. R Setup block has been written with steelblue.
# Let's verify we patch p_peer1 edge color or just run p_peer1 + p_peer2 + p_peer3
p_peer1_fixed <- ggraph(g_peer1, layout = "manual", x = l_peer[,1], y = l_peer[,2]) +
  geom_edge_link(aes(color = type, linetype = type), width = 1.8) +
  scale_edge_color_manual(values = c("Romantic" = "firebrick", "Friend" = "steelblue")) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Phase 1: Dating & Friendships") + theme_graph() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11), legend.position = "none") +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8), clip = "off")

p_peer1_fixed + p_peer2 + p_peer3
```
</center>

1. **Phase 1**: i and j date (dotted), k is i's same-sex friend, l is j's same-sex friend.
2. **Phase 2**: Through double-dating (triadic closure), j befriends k, and i befriends l.
3. **Phase 3 (Breakup)**: i and j split. Friendship loyalty prevails. Friend k will **never** date j (their friend i's ex), and l will **never** date i (their friend j's ex). The 4-cycle is blocked, fracturing the group into two open chains!

---

## Empirical Proof of the Code

- **Testing the PEAR Study Network**:
  - McMillan et al. studied a Dutch high school romantic network (607 students).
  - Adhered to Bearman's spanning tree characteristics: exceptionally sparse (density = 0.005), low cycles (only 8 four-cycles total).
- **TERGM Statistical Analysis**:
  - Adhering to the **"no seconds" Girl/Bro Code** yields **74% lower odds** of dating at Time 2 ($b = -1.347, p < 0.05$).
  - Once friendship loyalty is accounted for, there is **no remaining tendency** to avoid 4-cycles on the dating-only side ($b = 0.949, p = 0.21$, non-significant).
- **Takeaway**:
  - The global spanning tree is not born of a dating-only taboo, but of **peer friendship loyalty** blocking cycle closure!

---

# Part 4: Reality vs. Fiction (The TV Cases of Glee and Grey's Anatomy)

---

## Do Fictional Networks Follow the Rules?

- Writers of popular television dramas (sitcoms, prime-time soap operas) rely heavily on romantic networks to build storylines.
- Let's examine two famous fictional cases:
  - **Glee's McKinley High** (Seasons 1-2, studied by Jimi Adams, 2015).
  - **Grey's Anatomy** (GAN, Seasons 1-6, studied by Marcum et al., 2016).

. . .

- **The Macro-Observation**:
  - In Glee, a whopping **42%** of observed romantic ties are part of a 4-cycle loop (e.g., Quinn-Finn-Rachel-Puckerman).
  - In Grey's Anatomy, 4-cycles, love triangles, and overlapping loops are rampant.
  - **Why?** Drama needs conflict! Writers actively violate local cycle-avoidance taboos to generate plot tension.

---

## Glee's McKinley High Network

::: {.columns}
::: {.column width="55%"}
- **High-Clustering Drama**:
  - Glee's network is incredibly small, highly dense, and saturated with loops.
  - Screenwriters use cycle completions (e.g., dating a friend's ex or partner's ex) to drive dramatic friction.
- **The ERGM Reveal**:
  - When we control for lower-order structural constraints (small cast size, high activity) via ERGMs:
  - Fictional characters *still* exhibit a **statistically significant aversion to 4-cycles** ($b = -0.40, p < 0.05$).
- **Takeaway**:
  - Art imitates reality! Even under extreme dramatic pressures, writers implicitly follow real-world sexual taboos to keep plots recognizable.
:::

::: {.column width="45%"}
### Glee's Saturated Loops
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
p_glee
```
</center>
:::
:::

---

# Part 5: Same-Sex Constraints (The L Word Case)

---

## Heteronormativity vs. Homosexual Markets

- Marcum, Lin, and Koehly (2016) extended the study of 4-cycles to homosexual networks, comparing:
  - **Grey's Anatomy (GAN)**: Predominantly heterosexual adult network.
  - **Showtime's The L Word (LWN)**: Predominantly lesbian same-sex network in Los Angeles.

. . .

- **The Structural Findings**:
  - Heterosexual networks (GAN, and the real-world Colorado Springs gonorrhea network) show a **strong, significant aversion to 4-cycles** ($b = -0.23, p < 0.05$).
  - Homosexual networks (LWN) show **no significant aversion** to 4-cycles ($b = -0.03, p > 0.10$).
  - Love triangles and cycles of length 4 are structurally unconstrained and socially permitted in the lesbian network.

---

## Market Size and Structural Necessity

::: {.columns}
::: {.column width="60%"}
- **Why are same-sex cycles different?**
- **1. Structural Constraint**:
  - Same-sex dating markets are significantly smaller and more constrained than heterosexual markets.
  - Due to limited partner availability, strict cycle avoidance (the "no seconds" code) would result in a massive proportion of the population being locked out of dating entirely.
- **2. Normative Shift**:
  - In lesbian same-sex networks, "love triangles" and cycle-closures are normalized. 
  - Gaps are closed out of structural necessity, making cycle-taboos a luxury of large, heteronormative dating pools.
:::

::: {.column width="40%"}
### Fictional Adult Networks
| Metric | Grey's Anatomy (GAN) | The L Word (LWN) |
|:---|:---:|:---:|
| **Size (Nodes)** | 44 | 91 |
| **Edges** | 46 | 105 |
| **Same Gender Ties** | 2 | 86 |
| **4-Cycles** | 7 | 12 |
| **Cycle Aversion?** | **Significant** | **None** |
:::
:::

---

# Part 6: Rumors vs. Reality (whosdatedwho.com Gossip Analysis)

---

## Allegations of Affection

- Carmella Nicole Stoddard's (2024) UCLA dissertation looked at a massive celebrity dating network scraped from **whosdatedwho.com (WDW)**:
  - **109,626 nodes** and **88,746 edges** across celebrity histories.
- **The Core Distinction**:
  - **Confirmed Ties**: Fully established, mutually acknowledged relationships.
  - **Rumored Ties**: Unconfirmed allegations, gossip, and media speculations (3,848 ties).
- **The Question**:
  - How do rumored ties differ from confirmed ties structurally, and what do they tell us about social norms?

---

## Visualizing Rumors vs. Reality

::: {.columns}
::: {.column width="35%"}
- **Confirmed Structure**:
  - Forms clean, sparse, chain-like spanning tree structures with low cycles.
- **Rumored Structure**:
  - Adding rumored ties **significantly increases short cycles** (3-cycles and 4-cycles) and bisexual nodes.
- **Why?**
  - Gossip does not spread randomly!
  - Gossip is a **surveillant mechanism** that selectively targets, amplifies, and broadcasts **counter-normative behaviors**.
:::

::: {.column width="32%"}
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_conf
```
</center>
:::

::: {.column width="33%"}
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_rumor
```
- Red lines show rumored ties completing short cycles!
</center>
:::
:::

---

## Gossip as a Norm-Enforcement Tool

- **The Tradeoff of Gossip & Rumor**:
  - Gossip is attracted to **salaciousness** (violations of cycle avoidance, sexuality discordant pairings, extreme age-gaps).
  - By rumor-mongering these violations (completing 3-cycles and 4-cycles), the community publicly **highlights and sanctions deviant behaviors**.
- **The WDW Findings**:
  - Rumored ties are heavily overrepresented in short cycles and age-gapped ties.
  - Rather than being random noise, rumors are a **multiplex tool** used to enforce local monogamy and age-homophily norms by shining a spotlight on those who break them.

---

# Part 7: Summary & Class Discussion

---

## Summary of the Micro-Macro Romantic Link

- **1. Structure Emerges from Gaps**:
  - Spanning tree networks are not created by positive choices, but by **negative proscriptions** (avoiding short cycles).
- **2. Friendship Drives Romantic Structures**:
  - The "Bro/Girl Code" (avoiding dating a friend's ex) is the true local engine that shapes adolescent spanning trees.
- **3. Market Size Decouples Norms**:
  - Smaller same-sex dating markets cannot afford cycle-avoidance taboos, showing how **structural constraints** alter social codes.
- **4. Gossip Maps Deviance**:
  - Celebrity rumor networks demonstrate that gossip serves to **identify, broadcast, and sanction** those who violate local network taboos.

---

## Discussion Questions

1. Why do you think friendship-loyalty norms ("Bro Code" / "Girl Code") have higher face validity for teenagers than dating-only cycle-avoidance rules?
2. If same-sex markets must relax cycle-avoidance taboos due to structural constraint, how might this impact the spread of infectious pathogens (STIs) within gay/lesbian communities compared to heterosexual ones?
3. How does celebrity gossip on sites like whosdatedwho.com act as a "surveillant filter" for norm violations, and what other online communities use rumor as a tool for social control?
