---
title: "Ties, Dyads, and Triads"
subtitle: "The Motifs of Social Network Analysis"
format:
  revealjs:
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)

# Layout for a 3-node grid for triad drawings (perfect triangle)
l_triad <- matrix(c(0, 2, 1, 0, 0, 1.2), ncol = 2)
```

# Part 1: The Taxonomy of Ties

---

## The Relational Basis of Networks

::: {.incremental}
- In social network analysis, a **graph** is more than an abstract structure.
- It represents concrete **relationships** (ties) between actors.
- Not all relationships are created equal!
  - Some ties are naturally mutual (e.g., being siblings).
  - Others are directional (e.g., seeking advice).
  - Some strictly forbid mutuality (e.g., being a supervisor).
- To study networks, we must separate the **mathematical property of the relationship** (the tie type) from its **empirical realization** (the dyad or triad).
:::

---

## The Four Criteria for Social Ties

::: {.incremental}
- How do we decide when two actors share a connection? Network analysts use **four main criteria** ([Borgatti et al. 2009](https://doi.org/10.1126/science.1165821)):
  1. **Similarities**: Having something in common (e.g., co-location, co-membership in a club, shared attributes like gender/nationality).
  2. **Social Relations**: Acknowledging one another via culturally recognized labels or sentiments (e.g., friend, sibling, boss, spouse, liking/disliking).
  3. **Interactions**: Regular communication or behavior directed at another (e.g., texting, talking, helping, bullying).
  4. **Flows**: Transmission of information, resources, or states from one to another (e.g., information flow, virus infection, lending/borrowing).
:::

---

## States vs. Events in Social Networks

::: {.incremental}
- We can divide these four types of ties into two fundamental clusters:
- **States (Static Situations)**:
  - **Similarities** & **Social Relations**.
  - They represent continuous, long-standing conditions.
  - *Examples*: Being in the same class, being siblings, being coworker/boss.
- **Events (Discrete Activities)**:
  - **Interactions** & **Flows**.
  - They occur at specific points in time and are transient.
  - *Examples*: Texting someone, lending money, transmitting a virus.
:::

---

## 1. Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: A relation where connection is built-in and mutual by construction:
  - If a tie exists from $A$ to $B$, it *must* exist from $B$ to $A$.
  - Formally: $AB \iff BA$
- **Representation**: Represented by **undirected graphs**.
- **Examples**:
  - Siblinghood, marriage, co-membership, co-presence.
  - *Similarities are necessarily symmetric*:
    - If you are in the same place at the same time as someone, they are also there with you.
    - If you are zooming with someone, they are zooming with you.
    - If you go to UCLA with someone, they also go to UCLA with you.
- **Structural Constraint**: 
  - Pairs of actors can only be **disconnected** or **connected** by a single undirected edge.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_sym <- create_empty(2, directed = FALSE) %>% 
  bind_edges(data.frame(from = 1, to = 2)) %>% 
  mutate(name = toupper(letters[1:2]))
ggraph(gr_sym, layout = 'linear') +
  geom_edge_link(color = "steelblue", width = 2) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## 2. Asymmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: A relation with natural directionality where reciprocity is possible but *not guaranteed*:
  - If a tie exists from $A$ to $B$, nothing is assumed about $B$ to $A$.
  - Formally: $AB \implies$ nothing about $BA$
- **Representation**: Represented by **directed graphs** (digraphs).
- **Examples**:
  - Friendship nominations, advice-seeking, texting.
  - *Interactions/sentiments are naturally asymmetric*:
    - You can follow someone on Twitter, but they don't have to follow you back.
    - You can send a text to someone, but they don't have to respond.
    - You can seek advice from someone without them seeking advice from you.
- **Structural Constraint**:
  - Reciprocity is an **empirical event** (it may or may not occur), yielding three possible dyad configurations.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_asym <- create_empty(2, directed = TRUE) %>% 
  bind_edges(data.frame(from = 1, to = 2)) %>% 
  mutate(name = toupper(letters[1:2]))
ggraph(gr_asym, layout = 'linear') +
  geom_edge_link(color = "steelblue", edge_width = 1.5,
                  arrow = arrow(length = unit(5, 'mm'), type = "closed"),
                  end_cap = circle(8, 'mm')) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## 3. Anti-Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: An inherently directional relation where reciprocity is **strictly forbidden**:
  - If a tie exists from $A$ to $B$, a tie *cannot* exist from $B$ to $A$.
  - Formally: $AB \implies \neg BA$
- **Representation**: Represented by **directed graphs** (specifically oriented graphs, DAGs, or trees).
- **Examples**:
  - Father-of, boss-of, lending money.
  - *Hierarchical role relations are anti-symmetric*:
    - Kinship: If you are someone's parent, they cannot be your parent.
    - Formal: If you are someone's boss, they cannot be your boss at the same time.
- **Structural Constraint**:
  - Reciprocity is structurally impossible, yielding only two possible dyad configurations.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_anti <- create_empty(2, directed = TRUE) %>% 
  bind_edges(data.frame(from = 1, to = 2)) %>% 
  mutate(name = toupper(letters[1:2]))
ggraph(gr_anti, layout = 'linear') +
  geom_edge_link(color = "steelblue", edge_width = 1.5,
                  arrow = arrow(length = unit(5, 'mm'), type = "closed"),
                  end_cap = circle(8, 'mm')) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Mapping Social Ties to Graph Representations

| Sociological Category | Relational Property | Allowed Dyads | Visual Graph representation |
|:---|:---|:---|:---|
| **Co-membership, Propinquity** | **Symmetric** | • Disconnected<br>• Connected | Undirected Graph ($A - B$) |
| **Sentiments, Advice-seeking** | **Asymmetric** | • Disconnected<br>• Non-reciprocal<br>• Reciprocal | Directed Graph ($A \rightarrow B$ or $A \leftrightarrow B$) |
| **Formal Hierarchies, Kinship** | **Anti-symmetric** | • Disconnected<br>• Non-reciprocal | Oriented Graph/Tree ($A \rightarrow B$, no cycles) |

---

## Structure and Symmetry in Organizations

::: {.incremental}
- Formal organizations are a mixture of symmetric, asymmetric, and anti-symmetric relations:
- **Same-Level Ties (Symmetric)**:
  - Relations like "coworker" or "officemate" represent parity.
  - If $A$ is a coworker of $B$, then $B$ is a coworker of $A$ (implying equality).
- **Cross-Level Ties (Anti-symmetric)**:
  - Authority relations like "boss of" or "supervisor of" represent hierarchy.
  - If $A$ has authority over $B$, $B$ cannot have authority over $A$ (implying inequality).
- **Informal Ties (Asymmetric)**:
  - Sentiment/interaction ties like "liking" or "advice-seeking" exist alongside formal ones.
:::

---

## Structure and Symmetry in Families

::: {.incremental}
- Family kinship networks display similar mathematical structures:
- **Same-Generation Ties (Symmetric)**:
  - Relations like "sibling" or "cousin" represent horizontal equality.
  - If $A$ is a sibling of $B$, $B$ is necessarily a sibling of $A$.
- **Cross-Generation Ties (Anti-symmetric)**:
  - Relations like "parent of" or "uncle of" represent vertical lineage.
  - If $A$ is the parent of $B$, $B$ cannot be the parent of $A$.
- **Kinship Network Growth**:
  - Marriage merges separate family trees, generating a cascade of new symmetric (sibling-in-law) and anti-symmetric (parent-in-law) ties.
:::

---

## Transitivity in Hierarchical Ties

::: {.incremental}
- **Transitivity** is a key property of anti-symmetric hierarchical relations:
  - Formally: If $A \to B$ and $B \to C$, then $A \to C$.
- **In Organizations (Chain of Command)**:
  - If the Executive ($A$) has authority over the Manager ($B$), and the Manager has authority over the Staff member ($C$):
  - Then the Executive ($A$) has authority over the Staff member ($C$).
- **In Families (Lineage)**:
  - If $A$ is the parent of $B$, and $B$ is the parent of $C$:
  - Then $A$ is the grandparent of $C$ (establishing transitive kinship).
:::

---

## Multiplexity in Social Relations

::: {.incremental}
- In real-world social networks, actors are rarely linked by only one relationship.
- **Multiplexity** refers to the presence of multiple, distinct types of ties linking the same pair of nodes simultaneously.
  - e.g., $A$ and $B$ are coworkers (**symmetric** tie) AND $A$ is the supervisor of $B$ (**anti-symmetric** tie) AND they are friends (**asymmetric** tie).
- Multiplexity adds rich, multi-layered complexity to networks, as the behavior or pattern in one relation heavily influences and moderates others.
:::

---

## The Growth of Social Networks

::: {.incremental}
- Social networks are highly dynamic, growing and evolving over time.
- A classic example is the formation of a new family unit:
  1. A romantic tie (an asymmetric, non-kin relation) forms between two individuals.
  2. This culminates in a marriage tie (a symmetric, kin relation).
  3. This single event triggers a cascade of new "in-law" ties (both symmetric same-generation ties and anti-symmetric cross-generation ties).
  4. This merges two separate kinship trees into a single, highly complex, interconnected network.
:::

# Part 2: Dyads and Their Realizations

---

## What is a Dyad?

::: {.incremental}
- A **dyad** is any pair of nodes in a graph, whether they are joined by an edge or not.
- Mathematically, a dyad is *any subgraph of order two* ($n = 2$).
- Crucially, a dyad represents the **realized configuration** of ties between two nodes.
- **Sociological Insight**: Ties that do *not* exist (disconnected dyads) are often just as socially meaningful as those that do!
- The possible dyadic states are strictly constrained by the relational type of the tie.
:::

---

## Dyads under Symmetric Ties

::: {.incremental}
- Because undirected graphs do not allow directionality, a symmetric relation can only take on **two** dyadic states:
  1. **Disconnected Dyads** (historically called **null dyads**): The two nodes do not share an edge (size zero).
  2. **Connected Dyads**: The two nodes share an edge (size one).
- Every pair of actors in an undirected graph is either connected or disconnected.
:::

---

## Dyads under Asymmetric Ties

::: {.incremental}
- In directed graphs, reciprocity is possible but optional.
- Thus, asymmetric ties can yield **three** distinct kinds of dyads:
  1. **Disconnected Dyads** (Null): Neither node sends a directed tie to the other (size zero).
  2. **Non-Reciprocal Dyads** (Asymmetric): One node sends a tie to another, but it is not reciprocated (size one).
  3. **Reciprocal Dyads** (Mutual): Both nodes send directed ties to each other (size two).
- In standard SNA, this three-fold classification is called the **MAN classification**: **M**utual, **A**symmetric, **N**ull.
:::

---

## Separating Tie Properties from Dyad Realizations

::: {.incremental}
- It is critical to keep the property of the **tie** separate from the property of the **dyad**:
  - **Asymmetric Tie**: Describes the type of *relation* (directionality is allowed; reciprocity is optional).
  - **Non-Reciprocal Dyad**: Describes a realized *configuration* of a pair of nodes where reciprocity failed to occur empirically.
  - **Reciprocal Dyad**: Describes an empirical configuration composed of a *pair* of directed edges under an asymmetric relation.
- Confusing "asymmetric tie" with "asymmetric dyad" is a common conceptual error!
:::

---

## Dyads under Anti-Symmetric Ties

::: {.incremental}
- Because reciprocity is strictly forbidden under anti-symmetric ties, mutual (reciprocal) dyads cannot occur.
- Therefore, anti-symmetric ties can only yield **two** types of dyads:
  1. **Disconnected Dyads**: The two nodes do not share any directed edge.
  2. **Non-Reciprocal Dyads**: Only a single one-way directed tie exists between them ($A \rightarrow B$).
- This simplifies the dyad census of hierarchical networks: every pair of nodes is either completely disconnected or connected by a strictly non-reciprocal arrow.
:::

# Part 3: Triads and Motifs

---

## What is a Triad?

::: {.incremental}
- A **triad** is any set of three nodes, whether joined by edges or not.
- Mathematically, a triad is *any subgraph of order three* ($n = 3$).
- Triads have incredibly powerful sociological properties (e.g., structural balance, brokerage, coalitions).
- A triad can be thought of as a **concatenation of three dyads**: $\{A, B\}$, $\{B, C\}$, and $\{A, C\}$.
- Just like dyads, the allowed configurations of a triad depend entirely on the relational properties of its ties.
:::

---

## Triads under Symmetric Ties

::: {.incremental}
- Since undirected dyads can only be connected or disconnected, there are only **four** possible undirected triad configurations:
  1. **Null Triad**: 0 connected dyads ($DDD$).
  2. **Disconnected Triad**: 1 connected dyad ($CDD$).
  3. **Open Triad**: 2 connected dyads ($CCD$).
  4. **Closed Triad**: 3 connected dyads ($CCC$).
- These constitute the classic undirected subgraphs shown in @fig-undir-triads.
:::

---

## The Four Undirected Triads

```{r}
#| label: fig-undir-triads
#| fig-cap: "The four types of undirected triads."
#| fig-cap-location: bottom
#| fig-height: 5.5
#| fig-width: 11

    gr <- create_empty(3, directed = FALSE) %>% mutate(name = toupper(letters[1:3]))
    p1 <- ggraph(gr, layout = l_triad) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Null") + coord_cartesian(clip = "off")
    
    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        mutate(name = toupper(letters[1:3]))
    p2 <- ggraph(gr, layout = l_triad) +
        geom_edge_link(color = "steelblue", width = 1.25) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Disconnected") + coord_cartesian(clip = "off")

    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        bind_edges(data.frame(from = 2, to = 3)) %>%
        mutate(name = toupper(letters[1:3]))
    p3 <- ggraph(gr, layout = l_triad) +
        geom_edge_link(color = "steelblue", width = 1.25) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Open") + coord_cartesian(clip = "off")
    
    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        bind_edges(data.frame(from = 2, to = 3)) %>%
        bind_edges(data.frame(from = 3, to = 1)) %>%
        mutate(name = toupper(letters[1:3]))
    p4 <- ggraph(gr, layout = l_triad) +
        geom_edge_link(color = "steelblue", width = 1.25) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Closed") + coord_cartesian(clip = "off")
    
    library(patchwork)
    p1 + p2 + p3 + p4 + plot_layout(nrow = 1)
```

---

## Triads under Anti-Symmetric Ties

::: {.incremental}
- Because reciprocity is strictly forbidden, any triad containing a mutual connection is impossible.
- This prunes the directed triad census from **16 down to exactly seven** allowed configurations:
  1. **Null**: $NNN$ (all disconnected)
  2. **One-tie**: $NNA$ (single one-way tie)
  3. **Out-Star**: $AAN(O)$ ($A \leftarrow B \rightarrow C$, open)
  4. **In-Star**: $AAN(I)$ ($A \rightarrow B \leftarrow C$, open)
  5. **Directed Line**: $AAN(L)$ ($A \rightarrow B \rightarrow C$, open)
  6. **Cycle**: $AAA(C)$ ($A \rightarrow B \rightarrow C \rightarrow A$, closed)
  7. **Transitive Hierarchy**: $AAA(I/O)$ ($A \rightarrow B \rightarrow C$ and $A \rightarrow C$, closed)
:::

---

## Anti-Symmetric Triads and Command Chains

::: {.incremental}
- This mathematical pruning simplifies the analysis of hierarchies.
- In formal organization command structures:
  - We expect to see **transitive triads** ($AAA(I/O)$), representing clean, unified chains of command (e.g., "the boss of my boss is my boss").
  - **Cycles** ($AAA(C)$) are structurally avoided because they generate circular authority paradoxes, where everyone takes orders from someone else in a loop.
:::

---

## Triads under Asymmetric Ties

::: {.incremental}
- When relations are asymmetric, reciprocity is allowed but optional.
- Because each dyadic slot can take on any of the **three** MAN states (Mutual, Asymmetric, Null), we can generate all **sixteen** directed triad configurations.
- This is the classic **MAN Triad Census**, which categorizes triads based on:
  - The number of **M**utual (reciprocal) dyads.
  - The number of **A**symmetric (non-reciprocal) dyads.
  - The number of **N**ull (disconnected) dyads.
- Let's look at all sixteen configurations together.
:::

---

## The Complete 16 Directed Triads

```{r}
#| label: fig-dirtriads-slides
#| fig-cap: "All Sixteen Directed Triad Configurations (MAN Triad Census)"
#| fig-cap-location: bottom
#| fig-height: 6
#| fig-width: 12

    gr_c <- create_empty(3, directed = TRUE)
    # Define plotting function to save slide code space
    plot_tr <- function(edges, title) {
      g <- gr_c %>% mutate(name = toupper(letters[1:3]))
      if (nrow(edges) > 0) {
        g <- g %>% bind_edges(edges)
      }
      p <- ggraph(g, layout = l_triad) +
        geom_edge_link(color = "steelblue", edge_width = 1.1,
                       arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                       end_cap = circle(5, 'mm')) +
        geom_node_point(size = 12, color = "tan2") +
        geom_node_text(aes(label = name), size = 5, color = "white") +
        theme_graph() + ggtitle(title) + coord_cartesian(clip = "off") +
        theme(plot.title = element_text(size = 10, face="bold", hjust=0.5))
      return(p)
    }
    
    p1 <- plot_tr(data.frame(), "NNN")
    p2 <- plot_tr(data.frame(from=1, to=2), "NNA")
    p3 <- plot_tr(data.frame(from=c(1,2), to=c(2,1)), "NNM")
    p4 <- plot_tr(data.frame(from=c(1,2), to=c(2,3)), "AAN(L)")
    p5 <- plot_tr(data.frame(from=c(1,1), to=c(2,3)), "AAN(O)")
    p6 <- plot_tr(data.frame(from=c(2,3), to=c(1,1)), "AAN(I)")
    
    p7 <- plot_tr(data.frame(from=c(2,1,3), to=c(1,2,1)), "MAN1")
    p8 <- plot_tr(data.frame(from=c(2,1,1), to=c(1,2,3)), "MAN2")
    p9 <- plot_tr(data.frame(from=c(2,1,3,2), to=c(1,2,2,3)), "MMN")
    
    p10 <- plot_tr(data.frame(from=c(1,2,3), to=c(2,3,1)), "AAA(C)")
    p11 <- plot_tr(data.frame(from=c(1,1,2), to=c(2,3,3)), "AAA(I/O)")
    
    p12 <- plot_tr(data.frame(from=c(1,2,2,3), to=c(2,1,3,1)), "MAA(C)")
    p13 <- plot_tr(data.frame(from=c(1,2,3,3), to=c(2,1,1,2)), "MAA(O)")
    p14 <- plot_tr(data.frame(from=c(1,2,1,2), to=c(2,1,3,3)), "MAA(I)")
    
    p15 <- plot_tr(data.frame(from=c(2,1,3,2,1), to=c(1,2,2,3,3)), "MMA")
    p16 <- plot_tr(data.frame(from=c(2,1,3,2,1,3), to=c(1,2,2,3,3,1)), "MMM")
    
    (p1 + p2 + p3 + p4) / (p5 + p6 + p7 + p8) / (p9 + p10 + p11 + p12) / (p13 + p14 + p15 + p16)
```

---

## Triad Enumeration Summary

* **0 connected dyads**:
  - **NNN** (panel 1)
* **1 connected dyad**:
  - **NNA** (panel 2), **NNM** (panel 3)
* **2 connected dyads (Open)**:
  - Non-reciprocal: **AAN(L)**, **AAN(O)**, **AAN(I)** (panels 4-6)
  - Mixed: **MAN1**, **MAN2** (panels 7-8)
  - Reciprocal: **MMN** (panel 9)
* **3 connected dyads (Closed)**:
  - Non-reciprocal: **AAA(C)**, **AAA(I/O)** (panels 10-11)
  - Mixed: **MAA(C)**, **MAA(O)**, **MAA(I)** (panels 12-14)
  - Reciprocal: **MMA** (panel 15), **MMM** (panel 16)

---

## Conclusion & Core Takeaways

::: {.incremental}
- **Ties define dyads, and dyads define triads**:
  - The mathematical type of relationship strictly controls the possible micro-structures in a network.
- **Symmetric Ties** $\implies$ connected vs. disconnected dyads $\implies$ 4 undirected triads.
- **Anti-Symmetric Ties** $\implies$ disconnected vs. non-reciprocal dyads $\implies$ 7 directed triads.
- **Asymmetric Ties** $\implies$ disconnected, non-reciprocal, or reciprocal dyads $\implies$ 16 directed triads (Triad Census).
- Understanding these micro-motifs is the key to decoding the macro-structures of power, friendship, and organization in society.
:::

---

## References {.unnumbered}
