---
title: "Graph Theory"
subtitle: "The Basics of Social Network Analysis"
format: revealjs
---

```{r setup, include=FALSE}
library(ggraph)
library(tidygraph)
library(igraph)
library(kableExtra)
```

# Introduction to Graphs

---

## What is a Graph?

::: {.incremental}
- A **graph** ($G$) is an abstract mathematical object consisting of two main parts:
- A **vertex set**, $V$, which contains all the nodes (actors) in the network.
  - e.g., $V = \{A, B, C, D\}$
- An **edge set**, $E$, which contains all the ties (relationships) between the nodes.
  - e.g., $E = \{AB, AC\}$
- Together, they define the graph: $G = \{V, E\}$
:::

---

## The Building Blocks: Nodes


::: {.incremental}
- In social networks, **nodes** represent the actors
  - Usually drawn as circles or points in a diagram
  - Often individuals (people) or organizations (groups, companies)
- In wider applications of **network science**:
  - Nodes can represent any entities in a system
  - Examples: power generation stations, servers/computers, animals in an ecosystem, towns

:::
---

## The Building Blocks: Edges

::: {.incremental}
- **Edges** represent a connection or a **social tie** between two nodes
  - Can be a permanent relationship (e.g., "brother of")
  - Or a fleeting interaction (e.g., a text message, co-presence)
- In set theory:
  - The set of edges is a collection of pairs of nodes
  - If node $A$ is linked to node $B$, then $AB \in E$ (or written as $V_A V_B$)
:::

---

## Edges as Channels

::: {.incremental}
- Edges act as channels for **flows of content**:
  - Positive/Neutral: influence, advice, information, support, power, traffic
  - Negative: disease, bullying, gossip, conflict
:::
---

## What is a Graph? (An Example) (1)


::: {.incremental}
- Let's define a graph $G$ with 4 nodes ($A, B, C, D$) and 2 edges ($AB, AC$):
  - $V = \{A, B, C, D\}$
  - $E = \{AB, AC\}$
  - $G = \{V, E\}$
- The **cardinality** (number of members in a set) is denoted by $|N|$:
  - Vertex set cardinality: $|V| = n = 4$
  - Edge set cardinality: $|E| = m = 2$
:::

---

## What is a Graph? (An Example) (2)


::: {.incremental}
- This graph can be represented by the signature $G(n, m) = G(4, 2)$
- Note the lack of edges:
  - No edge between $B$ and $C$ means no relationship between them.
  - Node $D$ is not connected to any other node.
:::
---

## Variations of $G(4, 2)$

- The signature $G(4, 2)$ (4 nodes, 2 edges) does not uniquely define a single graph.
- It can describe multiple topologically distinct graphs.

---

## G(4, 2) Variation 1

```{r}
#| label: fig-g42-1
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = 1, to = 2:3)) %>% 
    mutate(name = letters[1:4])
l <- as.matrix(ggraph(gr, layout = 'tree')$data[, 1:2])
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + ylim(-0.2, 1.1) + xlim(-1.1, 1.65)
```

---

## G(4, 2) Variation 2

```{r}
#| label: fig-g42-2
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 4), to = c(2, 3))) %>% 
    mutate(name = letters[1:4])    
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + ylim(-0.2, 1.1) + xlim(-1.1, 1.65)
```

---

## G(4, 2) Variation 3

```{r}
#| label: fig-g42-3
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 2), to = c(4, 3))) %>% 
    mutate(name = letters[1:4])    
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + ylim(-0.2, 1.1) + xlim(-1.1, 1.65)
```

---

## G(4, 2) Variation 4 (Unlabeled)

```{r}
#| label: fig-g42-4
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = 1, to = 2:3))  
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  theme_graph() + ylim(-0.2, 1.1) + xlim(-1.1, 1.65)
```

---

## Graph Labeling


::: {.incremental}
- **Labeled Graph**:
  - The nodes have letters or numbers that function as "names" (e.g., node $A$, node $B$, or $V_1, V_2$).
  - Allows us to refer to specific actors.
- **Unlabeled Graph**:
  - The names are not specified because they do not matter for structural representation (e.g., the last graph in the previous slide).
- *Note*: Graph metrics are mathematically identical regardless of whether a graph is labeled or unlabeled.

:::
---

## Trivial Graphs


::: {.incremental}
- **Trivial Graph**:
  - A $G(1, 0)$ graph: a single isolated node with zero connections.
  - Social equivalent: An isolated person with no connections (e.g., the Japanese *Hikikomori*).
:::

---

## Non-Trivial Graphs


::: {.incremental}
- **Non-Trivial Graphs**:
  - Any graph larger than $G(1, 0)$ used for social network analysis.
  - **Connected Dyad**: The $G(2, 1)$ graph. Two nodes connected by a single link. The "hydrogen atom" of society.
  - **Disconnected Dyad**: A $G(2, 0)$ graph. Two nodes with no link (e.g., two people on a deserted island not speaking).
:::
---

## Basic Graph Properties

:::: {.columns}
::: {.column width="60%"}
::: {.incremental}
- **Adjacent**: Joined by an edge (e.g., $A$ and $B$).
- **Nonadjacent**: Not joined by an edge (e.g., $A$ and $C$).
- **Isolate**: Node with zero incident edges (e.g., $D$).
- **Incident**: An edge "touches" a node (e.g., edge $AB$ is incident on $A$).
- **End Vertices**: Nodes at the ends of an edge (e.g., $A$ and $B$ for edge $AB$).
:::
:::

::: {.column width="40%"}
```{r}
#| label: fig-basic-props
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6

gr <- create_empty(4, directed = FALSE) %>%
  bind_edges(data.frame(from = 1:2, to = 2:3)) %>%
  mutate(name = c("A", "B", "C", "D"))

ggraph(gr, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::
::::

---

## The Graph Complement


::: {.incremental}
- For any graph $G$, its **graph complement** $G'$ is its "evil twin" which satisfies two conditions:
  1. Every pair of nodes adjacent in $G$ is non-adjacent in $G'$.
  2. Every pair of nodes non-adjacent in $G$ is adjacent in $G'$.
- All connections in $G$ become disconnections in $G'$, and vice versa.
- **Usage**: Useful when we want to count or analyze *non-relations* in a network.

:::
---

## Graph for Complement Example

```{r}
#| label: fig-complement-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

gr <- create_empty(6, directed = FALSE) %>%
      bind_edges(data.frame(from = 1, to = c(2, 4, 6))) %>%
      bind_edges(data.frame(from = 6, to = c(3, 5))) %>%
      mutate(name = toupper(letters[1:6])) 
p <- ggraph(gr, layout = 'kk') 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p1 <- p + theme_graph()
p1
```

---

## The Graph Complement

```{r}
#| label: fig-complement-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

gr.c <- complementer(gr)
l <- as.matrix(p$data[, 1:2])
p <- ggraph(gr.c, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p2 <- p + theme_graph()
p2
```

---

## Simple Graphs

:::: {.columns}
::: {.column width="60%"}
::: {.incremental}
- Most social networks are represented as **simple graphs**.
- A graph is simple if it meets two requirements:
  1. **No Multiple Edges**: Only one edge can join any pair of nodes (unlike $A$ and $B$).
  2. **No Loops**: No edges connect a node to itself (unlike node $C$).
- Loops and multiple edges do not make sense for most typical sociological applications.
:::
:::

::: {.column width="40%"}
```{r}
#| label: fig-non-simple
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6

gr_non_simple <- create_empty(3, directed = FALSE) %>%
  bind_edges(data.frame(from = c(1, 1, 2, 3), to = c(2, 2, 3, 3))) %>%
  mutate(name = c("A", "B", "C"))

ggraph(gr_non_simple, layout = 'circle') +
  geom_edge_fan(color = "steelblue", edge_width = 1.5, strength = 1.5) +
  geom_edge_loop(color = "tomato", edge_width = 1.5) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::
::::
---

## Graphs are Not Pictures!


::: {.incremental}
- Do not confuse a mathematical "graph" (a set of two sets of objects) with its visual representation or picture.
- There is no single "true" picture of a graph:
  - The positioning of points and lines in two-dimensional space is completely arbitrary.
  - The same mathematical graph (same vertex set, same edge set) can be drawn in countless different ways.

:::
---

## Same Graph, Different Drawings (1)

```{r}
#| label: fig-random-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

set.seed(678)
l <- layout_randomly(gr)
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p <- p + theme_graph() + coord_cartesian(clip = "off")
p
```

---

## Same Graph, Different Drawings (2)

```{r}
#| label: fig-random-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

set.seed(789)
l <- layout_randomly(gr)
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p <- p + theme_graph() + coord_cartesian(clip = "off")
p
```

# Graphs and Subgraphs

---

## Graphs and Subgraphs


::: {.incremental}
- Sometimes we are interested in analyzing only some parts of a **whole network**.
- A **subgraph** $G' = \{V', E'\}$ is a subset of the original graph $G = \{V, E\}$, which is written as $G' \subset G$:
  - The node set $V'$ is a subset of $V$ ($V' \subset V$).
  - The edge set $E'$ is a subset of $E$ ($E' \subset E$).
  - Edges in the subgraph must have both endpoints in the node subset $V'$.

:::
---

## Original Graph

```{r}
#| label: fig-subgraph-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

set.seed(367)
gr <- play_gnm(n = 10, m = 20, directed = FALSE, loops = FALSE) %>% 
  mutate(name = LETTERS[1:10])
l <- ggraph(gr, layout = 'kk')
l <- as.matrix(l$data[, 1:2])
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p1 <- p + theme_graph() + ylim(-1,1)
p1
```

---

## A Subgraph of the Original

:::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- This graph is a **subgraph** of the original graph from the previous slide.
- It was created by taking a **subset of nodes**: 
  - $V' = \{A, B, C, D, E\}$
- We retained only the edges from the original graph that connect these five nodes to each other.
- All other nodes ($F, G, H, I, J$) and any edges connected to them were removed.
:::
:::

::: {.column width="50%"}
```{r}
#| label: fig-subgraph-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 6

gr2 <- induced_subgraph(gr, toupper(letters[1:5]))
p <- ggraph(gr2, layout = l[1:5, ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p2 <- p + theme_graph() + ylim(-1,1)
p2
```
:::
::::

---

## Vertex-Induced Subgraphs


::: {.incremental}
- A subgraph can be induced by selecting a subset of vertices to retain.
- **Vertex-induced subgraph**: Keep the selected nodes and all edges from the original graph that connect these nodes.
- Useful for identifying cohesive subsets or local cliques.

:::
---

## Vertex-Induced Subgraph Example

```{r}
#| label: fig-subgraph-induced
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

gr3 <- induced_subgraph(gr, c("G", "D", "E", "I"))
p <- ggraph(gr3, layout = l[c(4, 5, 7, 9), ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p3 <- p + theme_graph() + ylim(-1,1)
p3
```

---

## Vertex-Deleted Subgraphs (1)

:::: {.columns}
::: {.column width="60%"}
::: {.incremental}
- We define a subgraph by *deleting* nodes: $G' = G - \{F, G, H, I, J\}$.
- **CRITICAL RULE**: When we delete nodes, all edges incident to those nodes **automatically disappear**.
- Removing nodes is useful for seeing how important a set of actors is for holding the network together.
- For instance, deleting $\{F, G, H, I, J\}$ from our 10-node graph yields the subgraph on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| label: fig-subgraph-deleted-node
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 6

gr_del <- gr - c("F", "G", "H", "I", "J")
p <- ggraph(gr_del, layout = l[1:5, ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p + theme_graph() + ylim(-1,1) + coord_cartesian(clip = "off")
```
:::
::::

---

## Vertex-Deleted Subgraphs (2)

::: {.incremental}
- **Null Graph**: The subgraph that results from removing *all* the nodes of the original graph (cardinality of the node set is zero).
- **Singleton Graph** (or **Trivial Graph**): The subgraph that results from removing all the nodes of the original graph *except for one* (cardinality of the node set is one).
:::
---

## Edge-Deleted Subgraphs

:::: {.columns}
::: {.column width="50%"}
::: {.incremental}
- We can create subgraphs by removing edges: $G'' = G - \{e_1, e_2\}$.
- **CRITICAL RULE**: When we delete edges, the nodes incident to them **remain intact**!
- **Spanning Subgraph**: A subgraph created by removing edges but leaving all nodes of the original graph.
- The example on the right removes 6 edges from the original 10-node graph. Notice all 10 nodes are still there!
:::
:::

::: {.column width="50%"}
```{r}
#| label: fig-subgraph-deleted-edges
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 6

gr4 <- gr - edge("A|B") - edge("G|J") - edge("C|I") - edge("A|E") - edge("A|C") - edge("C|E")

p <- ggraph(gr4, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p + theme_graph() + ylim(-1,1) + coord_cartesian(clip = "off")
```
:::
::::

---

## Special Subgraphs


::: {.incremental}
- **Null Graph**: The subgraph resulting from deleting *all* nodes (vertex set cardinality = 0).
- **Empty Graph**: The subgraph resulting from deleting *all* edges (edge set cardinality = 0).
- **Singleton (Trivial) Graph**: A subgraph consisting of exactly one node ($G(1,0)$).
- **Ego Graph**: A special subgraph centered around a single node ("ego"), containing the ego, its immediate neighbors ("alters"), and all edges among them.

# Vertex Fusion & Set Operations

:::
---

## Vertex Fusion (1)


::: {.incremental}
- We can construct a new graph from a graph by **fusing** two vertices.
- **Vertex Fusion**:
  - Replace two original vertices with a single vertex.
  - Keep the combined adjacency relations of both original vertices.
:::

---

## Vertex Fusion (2)


::: {.incremental}
- Vertex fusion is useful in social network analysis for:
  - Collapsing equivalence classes.
  - Analyzing blockmodels.
  - Merging groups or roles.
:::
---

## Set Operations on Graphs: Union

::: {.incremental}
- Because graphs are defined as sets, we can perform standard set operations on them.
- **Union** ($\cup$): Combine the vertices and edges of two graphs.
  - $V(G_1 \cup G_2) = V(G_1) \cup V(G_2)$
  - $E(G_1 \cup G_2) = E(G_1) \cup E(G_2)$
:::

---

## Set Operations on Graphs: Intersection

::: {.incremental}
- **Intersection** ($\cap$): Keep only the vertices and edges present in *both* graphs.
  - $V(G_1 \cap G_2) = V(G_1) \cap V(G_2)$
  - $E(G_1 \cap G_2) = E(G_1) \cap E(G_2)$
:::

---

# Node Neighborhoods
---

## Node Neighborhoods

::: {.columns}
::: {.column width="50%"}
- In a graph $G$, the **neighbors** of node $v$ are the nodes adjacent to it.
- **Neighborhood** ($\mathcal{N}(v)$): The subset of all vertices adjacent to $v$.
  - $\mathcal{N}(v) \subset V$
  - Neighbors of an isolate is the empty set $\emptyset$.
  - In the graph on the right: $\mathcal{N}(A) = \{B, C, D, F\}$.
:::

::: {.column width="50%"}
```{r}
#| label: fig-simple2-slides
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6

set.seed(329)
gr <- play_erdos_renyi(n = 6, m = 8, directed = FALSE, loops = FALSE)
gr <- mutate(gr, name = toupper(letters[1:6]))
p <- ggraph(gr, layout = 'auto') 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p <- p + theme_graph() + ylim(-1, 1)
p
```
:::
:::

---

## Neighborhood Intersection (1)


::: {.incremental}
- **Neighborhood Intersection** ($\mathcal{N}(u) \cap \mathcal{N}(v)$):
  - The set of nodes connected to both $u$ and $v$ (common neighbors).
  - Can be computed for connected or disconnected dyads.
- Cardinality of this set is the number of common neighbors.
- Disjoint neighborhood sets have an intersection of the empty set $\emptyset$ (0 common neighbors).
:::

---

## Neighborhood Intersection (2)


::: {.incremental}
- In our graph:
  - $\mathcal{N}(A) = \{B, C, D, F\}$
  - $\mathcal{N}(D) = \{A, B, F\}$
- Intersection: $\mathcal{N}(A) \cap \mathcal{N}(D) = \{B, F\}$.
:::
---

## Neighborhood Union


::: {.incremental}
- **Neighborhood Union** ($\mathcal{N}(u) \cup \mathcal{N}(v)$):
  - The set of nodes connected to either $u$ or $v$.
  - In our graph: $\mathcal{N}(A) \cup \mathcal{N}(D) = \{B, C, F\}$.
- Neighborhood intersection and union form the basis of mathematical measures of **structural similarity** between nodes (e.g., Jaccard similarity).

:::
---

## Neighborhoods in Directed Graphs (1)

::: {.incremental}
- Because links are directional, nodes have two neighborhoods.
- **In-Neighborhood** ($N^{in}(v)$):
  - Set of source nodes pointing *to* $v$.
  - In previous slide's directed graph: $N^{in}(D) = \{B, E, G\}$.
:::

---

## Neighborhoods in Directed Graphs (2)

::: {.incremental}
- **Out-Neighborhood** ($N^{out}(v)$):
  - Set of destination nodes that $v$ points *to*.
  - In previous slide's directed graph: $N^{out}(B) = \{A, C, D\}$.
- A node appears in both neighborhoods only if there is a **reciprocal** tie.
:::

---

# Node Degree
---

## Node Degree in Undirected Graphs


::: {.incremental}
- The **degree** ($k_i$) of a node is the number of connections it has.
- Formally, it can be defined in two ways:
  1. The **cardinality** of its neighborhood set:
     $$k_i = |\mathcal{N}(i)|$$
  2. The number of edges **incident** upon that node.
- **Degree Centrality**: The simplest measure of node position (connectedness).
- **Endpoints**: Nodes with degree equal to one (having a single neighbor).

:::
---

## Node Degree in Directed Graphs: Indegree

::: {.incremental}
- Directed graphs have two distinct degrees for each node.
- **Indegree** ($k^{in}_i$):
  - Number of edges pointing to node $i$ (cardinality of $N^{in}(i)$).
  - Sociological meaning: **Popularity, Prestige, or Status**.
:::

---

## Node Degree in Directed Graphs: Outdegree

::: {.incremental}
- **Outdegree** ($k^{out}_i$):
  - Number of edges pointing from node $i$ (cardinality of $N^{out}(i)$).
  - Sociological meaning: **Sociability or Activity**.
:::

---

# Basic Graph Metrics
---

## Graph Order and Size (1)


::: {.incremental}
- **Graph Order** ($n$):
  - The number of nodes in the graph ($n = |V|$).
  - Cardinality of the vertex set.
- **Graph Size** ($m$):
  - The number of edges in the graph ($m = |E|$).
  - Cardinality of the edge set.
:::

---

## Graph Order and Size (2)


::: {.incremental}
- In our undirected graph (9 nodes, 16 edges): $n = 9$, $m = 16$.
- In our directed graph (7 nodes, 11 edges): $n = 7$, $m = 11$.
:::
---

## Graph Maximum Size


::: {.incremental}
- The maximum possible number of edges if a graph of order $n$ were complete.
- **Undirected Graph**:
  
  $$Max(E)^u = \frac{n(n-1)}{2}$$

- **Directed Graph**:
  
  $$Max(E)^d = n(n-1)$$

- *Example*: In a classroom of 10 kids, if everyone is friends with everyone else, the number of friendships is:
  $$\frac{10(10-1)}{2} = 45 \text{ friendships}$$

:::
---

## Graph Density


::: {.incremental}
- Network density $d(G)$ is the ratio of observed edges $m$ to maximum possible edges. Represents tight-knit vs loose-knit networks.
- **Undirected Density**:
  
  $$d(G)^u = \frac{2m}{n(n-1)}$$

- **Directed Density**:
  
  $$d(G)^d = \frac{m}{n(n-1)}$$

- *Probability interpretation*: The probability that two randomly chosen nodes will be connected.
- Denser networks are less vulnerable to breakdowns when key nodes are removed.
:::

---

# Degree-Based Graph Metrics

---

## Graph Degree Set and Sequence


::: {.incremental}
- **Graph Degree Set** ($\mathbf{k}$):
  - A vector containing the degree of each node. Length of vector equals graph order $n$.
  - In our undirected graph example: $\mathbf{k} = (4, 3, 3, 4, 4, 4, 3, 3, 4)$.
- **Graph Degree Sequence** ($\mathbf{d}$):
  - The graph degree set sorted in decreasing order.
  - In our undirected graph example: $\mathbf{d} = (4, 4, 4, 4, 4, 3, 3, 3, 3)$.

:::
---

## Degree Range & Sum


::: {.incremental}
- **Degree Range** ($k_r$):
  - The difference between maximum and minimum degrees: $k_r = k_{max} - k_{min}$.
  - Measures heterogeneity of node connectivity.
- **Sum of Degrees** ($\sum k_i$):
  - **The First Theorem of Graph Theory**: In an undirected graph, the sum of degrees is twice the number of edges:
    
    $$\sum_i k_i = 2m$$

  - Each edge is incident to two nodes, so it is counted twice in the sum.

:::
---

## Graphical Degree Sequences


::: {.incremental}
- Because $\sum_i k_i = 2m$, the sum of degrees in an undirected graph **must be an even number**.
- **Graphical Sequence**: A sequence of integers that can represent the degree sequence of a real graph.
- **Handshaking Lemma**:
  - A degree sequence is graphical only if the number of nodes with **odd degrees** (odd nodes) is an **even number**.
  - Example: $\{4, 3, 3, 3, 2, 1, 1\}$ has 5 odd nodes, so it is **not graphical** (no such graph can exist).

:::
---

## Average Degree


::: {.incremental}
- The average connectivity of the nodes in a graph:
- **Undirected Average Degree**:
  
  $$\bar{k} = \frac{\sum_i k_i}{n} = \frac{2m}{n}$$

- **Directed Average Degree**:
  
  $$\bar{k} = \frac{m}{n}$$

- **Connection to Density**:
  
  $$d(G)^u = \frac{\bar{k}}{n-1}$$

:::
---

## Degree Variance


::: {.incremental}
- Captures the inequality or heterogeneity in connectivity across nodes.
- **Degree Variance** ($\mathcal{v}(G)$):
  
  $$\mathcal{v}(G) = \frac{\sum_i (k_i - \bar{k})^2}{n}$$

- The higher the variance, the higher the inequality (a few popular stars, many isolates).

:::
---

## Regular Graphs

::: {.columns}
::: {.column width="50%"}
- **Regular Graph**: A graph where every node has the exact same degree ($k$).
  - Denoted as a *$k$-regular graph of order $n$*.
  - Because $k_i = \bar{k}$ for all nodes, the degree variance of a regular graph is **always zero**.
  - The example on the right is a **3-regular graph of order 10**.
:::

::: {.column width="50%"}
```{r}
#| label: fig-regular-slides
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 6

set.seed(123)
gr <- sample_k_regular(10, 3)
gr <- gr %>% 
  as_tbl_graph() %>% 
  mutate(name = toupper(letters[1:10])) 
p <- ggraph(gr, layout = 'kk') 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p <- p + theme_graph()
p
```
:::
:::

---

## Feld's Friendship Paradox


::: {.incremental}
- **Average Nearest Neighbor Degree** ($\bar{k}_{nn(i)}$):
  - The average degree of node $i$'s neighbors:
    $$\bar{k}_{nn(i)} = \frac{1}{k_i} \sum_{j \in \mathcal{N}(i)} k_j$$
- **The Friendship Paradox (Scott Feld, 1991)**:
  - In almost all networks (except regular graphs), the average degree of a person's friends ($\bar{k}_{nn}$) is **greater** than the average degree of the people in the network ($\bar{k}$).
  - **Your friends have more friends than you do!**
:::

---

# Directed Graph Elements & Metrics

---

## Directed Degree Sets & Sequences


::: {.incremental}
- Directed graphs have two distinct degree sets (In-Degree set and Out-Degree set) and two distinct sequences.
- **Sum of Degrees**:
  - The sum of in-degrees is always equal to the sum of out-degrees, which is equal to the size $m$ (number of edges):
    
    $$\sum k^{in}_i = \sum k^{out}_i = m$$

  - Since both sum to $m$, directed graphs have only **one** average degree: $\bar{k} = m/n$.

:::
---

## Average Nearest Neighbor Degree

::: {.incremental}
- **Average Nearest Neighbor Degree** ($\bar{k}_{nn(i)}$):
  - The average degree of node $i$'s neighbors:
    $$\bar{k}_{nn(i)} = \frac{1}{k_i} \sum_{j \in \mathcal{N}(i)} k_j$$
- This metric gives us the expected number of "friends of friends" for an average person in the network.
:::

---

## Feld's Friendship Paradox

::: {.incremental}
- **The Friendship Paradox** (Scott Feld, 1991):
  - In almost all social networks, the average degree of a person's friends ($\bar{k}_{nn}$) is **greater** than the average degree of the people in the network ($\bar{k}$).
  - This means, on average, **your friends have more friends than you do!**
  - This is not a personal failing, but a structural property of most social networks.
:::
