---
title: "Graph Theory"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Base graph and layout for neighborhood slides
gr_base <- tbl_graph(
  nodes = data.frame(name = 1:7),
  edges = data.frame(
    from = c(1, 4, 4, 2, 5, 6, 3),
    to   = c(3, 3, 7, 5, 6, 2, 2)
  ),
  directed = FALSE
)

set.seed(42)
layout_fr <- create_layout(gr_base, layout = "fr")

# Reusable G(4, 2) graph and layout for intro slides
gr_g42 <- create_empty(4) %>% 
    bind_edges(data.frame(from = 1, to = 2:3)) %>% 
    mutate(name = LETTERS[1:4])
l_g42 <- as.matrix(ggraph(gr_g42, layout = 'tree')$data[, 1:2])
```

# Introduction to Graphs

---

## What is a Graph?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **graph** ($G$) is an abstract mathematical object consisting of two main parts:
- A **vertex set**, $V$, which contains all the nodes (actors) in the network.
  - e.g., $V = \{A, B, C, D\}$
- An **edge set**, $E$, which contains all the ties (relationships) between the nodes.
  - e.g., $E = \{AB, AC\}$
- Together, they define the graph: $G = \{V, E\}$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
ggraph(gr_g42, layout = 'manual', x=l_g42[,1], y=l_g42[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 30, color = "tan2") +
  geom_node_text(aes(label = name), size = 14, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## The Building Blocks: Nodes

::: {.incremental}
- In social networks, **nodes** represent the actors
  - Usually drawn as circles or points in a diagram
  - Often individuals (people) or organizations (groups, companies)
- In wider applications of **network science**:
  - Nodes can represent any entities in a system
  - Examples: power generation stations, proteins, websites
:::

---

## The Building Blocks: Edges

::: {.incremental}
- **Edges** represent a connection or a **social tie** between two nodes
  - Can be a permanent relationship (e.g., "brother of")
  - Or a fleeting interaction (e.g., a text message, co-presence)
- In set theory:
  - The set of edges is a collection of pairs of nodes
  - If node $A$ is linked to node $B$, then $AB \in E$ (or written as $V_A V_B$)
  - Edges act as channels for **flows of content**:
    - Positive/Neutral: influence, advice, information, support, power, traffic
    - Negative: disease, bullying, gossip, conflict
:::

---

## What is a Graph? (An Example) (1)

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- Let's define a graph $G$ with 4 nodes ($A, B, C, D$) and 2 edges ($AB, AC$):
  - $V = \{A, B, C, D\}$
  - $E = \{AB, AC\}$
  - $G = \{V, E\}$
:::
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
ggraph(gr_g42, layout = 'manual', x=l_g42[,1], y=l_g42[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 30, color = "tan2") +
  geom_node_text(aes(label = name), size = 14, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Graph? (An Example) (2)

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- For the graph $G = \{V, E\}$:
- The **cardinality** (number of members in a set) is denoted by $|N|$:
  - Vertex set cardinality: $|V| = n = 4$
  - Edge set cardinality: $|E| = m = 2$
:::
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
ggraph(gr_g42, layout = 'manual', x=l_g42[,1], y=l_g42[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 30, color = "tan2") +
  geom_node_text(aes(label = name), size = 14, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Graph? (An Example) (3)

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- This graph can be represented by the signature $G(n, m) = G(4, 2)$
- Note the lack of edges:
  - No edge between $B$ and $C$ means no relationship between them.
  - Node $D$ is not connected to any other node.
:::
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
ggraph(gr_g42, layout = 'manual', x=l_g42[,1], y=l_g42[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 30, color = "tan2") +
  geom_node_text(aes(label = name), size = 14, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Variations of $G(4, 2)$

- The signature $G(4, 2)$ (4 nodes, 2 edges) does not uniquely define a single graph.
- It can describe multiple topologically distinct graphs.

---

## G(4, 2) Variation 1

```{r}
#| label: plt-g42-1
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = 1, to = 2:3)) %>% 
    mutate(name = LETTERS[1:4])
l <- as.matrix(ggraph(gr, layout = 'tree')$data[, 1:2])
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## G(4, 2) Variation 2

```{r}
#| label: plt-g42-2
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 4), to = c(2, 3))) %>% 
    mutate(name = LETTERS[1:4])    
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## G(4, 2) Variation 3

```{r}
#| label: plt-g42-3
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 2), to = c(4, 3))) %>% 
    mutate(name = LETTERS[1:4])    
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## G(4, 2) Variation 4 (Unlabeled)

```{r}
#| label: plt-g42-4
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = 1, to = 2:3))  
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## Graph Labeling

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Labeled Graphs**:
  - Distinctions are made between nodes (e.g., A, B, C are different people).
  - Swapping node names mathematically changes the graph.
- **Unlabeled Graphs**:
  - Nodes are interchangeable or anonymous.
  - The focus is strictly on the abstract structure of connections.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4
gr_lab <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 2), to = c(4, 3))) %>% 
    mutate(name = LETTERS[1:4])
p1 <- ggraph(gr_lab, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, color = "white") +
  theme_graph() + ggtitle("Labeled") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust=0.5))

gr_unlab <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 2), to = c(4, 3)))
p2 <- ggraph(gr_unlab, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 12, color = "tan2") +
  theme_graph() + ggtitle("Unlabeled") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust=0.5))

p1 + p2
```
:::

::::

---

## Trivial Graphs

::: {.incremental}
- **Trivial Graph**:
  - A $G(1, 0)$ graph: a single isolated node with zero connections.
  - Social equivalent: An isolated person with no connections to others.
- **Non-Trivial Graphs**:
  - Any graph larger than $G(1, 0)$ used for social network analysis.
  - **Connected Dyad**: The $G(2, 1)$ graph. Two nodes connected by a single link. The "hydrogen atom" of society.
  - **Disconnected Dyad**: A $G(2, 0)$ graph. Two nodes with no link (e.g., two people on a deserted island not speaking).
:::

---

## Basic Relationships Between Nodes and Edges in Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Two nodes are:
  - **Adjacent**: If they are joined by an edge in the graph (e.g., $A$ and $B$).
  - **Nonadjacent**: If they are *not* joined by an edge in the graph (e.g., $A$ and $C$).
  - **End Vertices** with respect to and edge: if the lie at the ends of an edge (e.g., $A$ and $B$ for edge $AB$).
:::
:::

::: {.column width="40%"}
```{r}
#| label: plt-basic-props-1
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6

gr <- create_empty(4, directed = FALSE) %>%
  bind_edges(data.frame(from = 1:2, to = 2:3)) %>%
  mutate(name = c("A", "B", "C", "D"))

ggraph(gr, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 30, color = "tan2") +
  geom_node_text(aes(label = name), size = 14, fontface = "bold", color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Basic Relationships Between Nodes and Edges in Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- An edge is:
  - **Incident** to a node: If it connects to that node (e.g., edge $AB$ is incident to nodes $A$ and $B$).
  - **Non-incident** to a node: If it does not connect to that node (e.g., edge $AB$ is non-incident to node $C$).
- A node is an **isolate**: if it has no incident edges or is not adjacent to another node (e.g., $D$).
:::
:::

::: {.column width="40%"}
```{r}
#| label: plt-basic-props-2
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6

gr <- create_empty(4, directed = FALSE) %>%
  bind_edges(data.frame(from = 1:2, to = 2:3)) %>%
  mutate(name = c("A", "B", "C", "D"))

ggraph(gr, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 30, color = "tan2") +
  geom_node_text(aes(label = name), size = 14, fontface = "bold", color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## The Graph Complement

::: {.incremental}
- The **complement** ($G'$) of a graph $G$ contains an edge between two nodes *if and only if* they are disconnected in $G$.
- It represents the "negative" of the original graph, showing all the ties that *could* exist but don't.
:::

---

## The Graph Complement: Example

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- Let's start with a simple graph $G$ containing 6 nodes and 5 edges.
- Nodes $A$ and $F$ act as central connectors.
- This original graph has many missing edges (e.g., $B-C$, $D-E$).
:::
:::

::: {.column width="50%"}
```{r}
#| label: plt-complement-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5

gr <- create_empty(6, directed = FALSE) %>%
      bind_edges(data.frame(from = 1, to = c(2, 4, 6))) %>%
      bind_edges(data.frame(from = 6, to = c(3, 5))) %>%
      mutate(name = toupper(letters[1:6])) 
p <- ggraph(gr, layout = 'kk') 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p1 <- p + theme_graph() + coord_cartesian(clip = "off")
p1
```
:::

::::

---

## The Graph Complement: Example

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- The complement graph $G'$ on the right connects all pairs that were *not* connected in $G$.
- Every edge present in the original graph is now missing.
- Every edge missing in the original graph is now present.
:::
:::

::: {.column width="50%"}
```{r}
#| label: plt-complement-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5

gr.c <- complementer(gr)
l <- as.matrix(p$data[, 1:2])
p <- ggraph(gr.c, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p2 <- p + theme_graph() + coord_cartesian(clip = "off")
p2
```
:::

::::

---

## Simple Graphs

::: {.incremental}
- Most social networks are represented as **simple graphs**.
- A graph is simple if it meets two requirements:
  1. **No Multiple Edges**: Only one edge can join any pair of nodes.
  2. **No Loops**: No edges connect a node to itself.
- Loops and multiple edges do not make sense for most typical sociological applications.
:::

---

## Non-Simple Graphs

:::: {.columns}

::: {.column width="55%"}
::: {.incremental}
- Graphs that do not meet these requirements are **non-simple** (or multigraphs).
- The diagram on the right shows:
  - **Multiple Edges**: Two edges connect nodes $A$ and $B$.
  - **Loop**: An edge connects node $C$ to itself.
:::
:::

::: {.column width="45%"}
```{r}
#| label: plt-non-simple
#| echo: false
#| fig-align: "center"
#| fig-height: 5.5
#| fig-width: 5.5

gr_non_simple <- create_empty(3, directed = FALSE) %>%
  bind_edges(data.frame(from = c(1, 1, 2, 3), to = c(2, 2, 3, 3))) %>%
  mutate(name = c("A", "B", "C"))

ggraph(gr_non_simple, layout = 'circle') +
  geom_edge_fan(color = "steelblue", edge_width = 1.5, strength = 1.5) +
  geom_edge_loop(color = "tomato", edge_width = 1.5) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Graphs are Not Pictures!

::: {.incremental}
- Do not confuse a mathematical "graph" (a set of two sets) with its "drawing" (a point and line plot).
- The same graph can have many different, and even misleading, drawings.
- The drawing is just a visual aid; the underlying structure is what matters.
:::

---

## Same Graph, Different Drawings (1)

```{r}
#| label: plt-random-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5

set.seed(678)
l <- layout_randomly(gr)
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p <- p + theme_graph() + coord_cartesian(clip = "off")
p
```

---

## Same Graph, Different Drawings (2)

```{r}
#| label: plt-random-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5

set.seed(789)
l <- layout_randomly(gr)
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p <- p + theme_graph() + coord_cartesian(clip = "off")
p
```

# Graphs and Subgraphs

---

## Graphs and Subgraphs

::: {.incremental}
- Sometimes we are interested in analyzing only some parts of a **whole network**.
- A **subgraph** $G' = \{V', E'\}$ is a subset of the original graph $G = \{V, E\}$, which is written as $G' \subset G$:
  - The node set $V'$ is a subset of $V$ ($V' \subset V$).
  - The edge set $E'$ is a subset of $E$ ($E' \subset E$).
  - Edges in the subgraph must have both endpoints in the node subset $V'$.
:::

---

## Original Graph

```{r}
#| label: plt-subgraph-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5

set.seed(367)
gr <- play_gnm(n = 10, m = 20, directed = FALSE, loops = FALSE) %>% 
  mutate(name = LETTERS[1:10])
l <- ggraph(gr, layout = 'kk')
l <- as.matrix(l$data[, 1:2])
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p1 <- p + theme_graph() + coord_cartesian(clip = "off")
p1
```

---

## A Subgraph of the Original

:::: {.columns}

::: {.column width="50%"}
::: {.incremental}
- Created by keeping a subset of nodes: $V' = \{A, B, C, D, E\}$.
- Retains only edges connecting these five nodes.
- All other nodes ($F, G, H, I, J$) and their incident edges are removed.
:::
:::

::: {.column width="50%"}
```{r}
#| label: plt-subgraph-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5

gr2 <- induced_subgraph(gr, toupper(letters[1:5]))
p <- ggraph(gr2, layout = l[1:5, ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p2 <- p + theme_graph() + coord_cartesian(clip = "off")
p2
```
:::

::::

---

## Vertex-Induced Subgraphs

::: {.incremental}
- A **vertex-induced subgraph** is created by selecting a subset of nodes ($V' \subset V$) and including *all* the edges from the original graph that connect nodes within that subset.
- This is the most common method for creating subgraphs in social network analysis.
:::

---

## Vertex-Induced Subgraphs: Example

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Let's select the subset of nodes:
  $$V' = \{D, E, G, I\}$$
- Keeping all original edges between these nodes yields the vertex-induced subgraph on the right.
:::
:::

::: {.column width="45%"}
```{r}
#| label: plt-subgraph-induced
#| echo: false
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5

gr3 <- induced_subgraph(gr, c("G", "D", "E", "I"))
p <- ggraph(gr3, layout = l[c(4, 5, 7, 9), ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p3 <- p + theme_graph() + coord_cartesian(clip = "off")
p3
```
:::

::::

---

## Vertex-Deleted Subgraphs

::: {.incremental}
- A subgraph can be defined by deleting a set of nodes:
  $$G' = G - \{a, b, c\}$$
- **CRITICAL RULE**: Deleting nodes automatically removes all edges incident to them (edges that connect to the deleted nodes).
:::

---

## Vertex-Deleted Subgraphs: Example

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Example: Deleting $\{F, G, H, I, J\}$ from the original graph.
- This removes those 5 nodes and all their incident edges, yielding the subgraph on the right.
- This is equivalent to the vertex-induced subgraph on $\{A, B, C, D, E\}$.
:::
:::

::: {.column width="45%"}
```{r}
#| label: plt-subgraph-deleted-node
#| echo: false
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5

gr_del <- gr - c("F", "G", "H", "I", "J")
p <- ggraph(gr_del, layout = l[1:5, ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p + theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Special Cases of Vertex Deletion

::: {.incremental}
- Two special cases of vertex deletion are:
  - **Null Graph**: The subgraph that results from removing *all* nodes from the original graph.
  - **Singleton Graph** (or **Trivial Graph**): The subgraph that results from removing all nodes *except for one*.
:::

---

## Edge-Deleted Subgraphs

::: {.incremental}
- A subgraph can be defined by deleting a set of edges:
  $$G' = G - \{e_1, e_2, ...\}$$
- **CRITICAL RULE**: Nodes remain in the graph when their edges are deleted (unlike vertex deletion, which automatically deletes incident edges).
:::

---

## Edge-Deleted Subgraphs: Example

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Example: Deleting six specific edges from the original graph:
  - $\{AB, GJ, CI, AE, AC, CE\}$
- This removes the edges while keeping all 10 nodes intact.
- The resulting edge-deleted subgraph is shown on the right.
:::
:::

::: {.column width="45%"}
```{r}
#| label: plt-subgraph-deleted-edges
#| echo: false
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5

gr4 <- gr - edge("A|B") - edge("G|J") - edge("C|I") - edge("A|E") - edge("A|C") - edge("C|E")

p <- ggraph(gr4, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p + theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Spanning & Empty Subgraphs

::: {.incremental}
- Two key concepts in edge deletion:
  - **Spanning Subgraph**: A subgraph that leaves all original nodes intact, removing only edges (like the example we just saw).
  - **Empty Graph**: The subgraph that results from removing all edges, leaving only isolated nodes.
:::

---

## Special Subgraphs: Ego Graph

:::: {.columns}

::: {.column width="50%"}
::: {.incremental}
- **Ego Graph**: Subgraph centered on a node (ego), containing the ego, its neighbors (alters), and all ties among them.
- Particularly useful for actor-level network analysis.
- Example: Ego graph for node **D** (in red) and its alters **A**, **B**, and **G** (in tan).
:::
:::

::: {.column width="50%"}
```{r}
#| label: plt-subgraph-ego
#| echo: false
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5

g_ego <- make_ego_graph(gr, order = 1, nodes = "D")[[1]]
l_ego <- l[match(V(g_ego)$name, LETTERS[1:10]), ]

ggraph(g_ego, layout = l_ego) +
  geom_edge_link(color = "steelblue", width = 1.25) +
  geom_node_point(aes(color = (name == "D")), size = 18, show.legend = FALSE) +
  geom_node_text(aes(label = name), size = 8, color = "white") +
  scale_color_manual(values = c("TRUE" = "tomato", "FALSE" = "tan2")) +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Vertex Fusion (1)

:::: {.columns}

::: {.column width="45%"}
::: {.incremental}
- We can construct a new graph from a graph by **fusing** two vertices.
- **Vertex Fusion**:
  - Replace two original vertices with a single vertex.
  - Keep the combined adjacency relations of both original vertices.
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5
gr_combined <- tbl_graph(
  nodes = data.frame(
    name = c("A", "B", "C", "D", "A", "BC", "D")
  ),
  edges = data.frame(
    from = c(1, 1, 2, 2, 5, 6),
    to   = c(2, 3, 3, 4, 6, 7)
  ),
  directed = FALSE
)

layout_combined <- matrix(c(
  1, 2,
  2, 3,
  2, 1,
  3, 2,
  5, 2,
  6, 2,
  7, 2
), ncol = 2, byrow = TRUE)

ggraph(gr_combined, layout = 'manual', x = layout_combined[,1], y = layout_combined[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, color = "white", fontface = "bold") +
  annotate("text", x = 2, y = 3.6, label = "Original", size = 5, fontface = "bold", color = "grey30") +
  annotate("text", x = 6, y = 3.6, label = "Fused (B & C)", size = 5, fontface = "bold", color = "grey30") +
  annotate("segment", x = 3.7, xend = 4.3, y = 2, yend = 2, 
           arrow = arrow(type = "closed", length = unit(0.1, "inches")), 
           color = "grey50", size = 1.2) +
  theme_graph() + 
  coord_cartesian(clip = "off")
```
:::

::::

---

## Vertex Fusion (2)

::: {.incremental}
- Vertex fusion is useful in social network analysis for:
  - Collapsing equivalence classes.
  - Analyzing blockmodels.
  - Merging groups or roles.
:::

---

## Set Operations on Graphs: Union

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- Because graphs are defined as sets, we can perform standard set operations on them.
- **Union** ($\cup$): Combine the vertices and edges of two graphs.
  - $V(G_1 \cup G_2) = V(G_1) \cup V(G_2)$
  - $E(G_1 \cup G_2) = E(G_1) \cup E(G_2)$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 4.5
gr_union_plt <- tbl_graph(
  nodes = data.frame(
    name = c("A", "B", "C", "B", "C", "D", "A", "B", "C", "D")
  ),
  edges = data.frame(
    from = c(1, 2, 4, 5, 7, 8, 9),
    to   = c(2, 3, 5, 6, 8, 9, 10)
  ),
  directed = FALSE
)

layout_union <- matrix(c(
  1.0, 2.0,
  2.0, 3.0,
  2.0, 1.0,
  4.0, 3.0,
  4.0, 1.0,
  5.0, 2.0,
  7.0, 2.0,
  8.0, 3.0,
  8.0, 1.0,
  9.0, 2.0
), ncol = 2, byrow = TRUE)

ggraph(gr_union_plt, layout = 'manual', x = layout_union[,1], y = layout_union[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, color = "white", fontface = "bold") +
  annotate("text", x = 1.5, y = 3.6, label = "G1", size = 5, fontface = "bold", color = "grey30") +
  annotate("text", x = 3.0, y = 2.0, label = "\u222a", size = 6, fontface = "bold", color = "grey30") +
  annotate("text", x = 4.5, y = 3.6, label = "G2", size = 5, fontface = "bold", color = "grey30") +
  annotate("text", x = 6.0, y = 2.0, label = "=", size = 6, fontface = "bold", color = "grey30") +
  annotate("text", x = 8.0, y = 3.6, label = "G1 \u222a G2", size = 5, fontface = "bold", color = "grey30") +
  theme_graph() + 
  theme(plot.margin = margin(l = 15, r = 10, t = 10, b = 10, unit = "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Set Operations on Graphs: Intersection

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- **Intersection** ($\cap$): Keep only the vertices and edges present in *both* graphs.
  - $V(G_1 \cap G_2) = V(G_1) \cap V(G_2)$
  - $E(G_1 \cap G_2) = E(G_1) \cap E(G_2)$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 4.5
gr_intersect_plt <- tbl_graph(
  nodes = data.frame(
    name = c("A", "B", "C", "B", "C", "D", "B", "C")
  ),
  edges = data.frame(
    from = c(1, 2, 4, 5, 7),
    to   = c(2, 3, 5, 6, 8)
  ),
  directed = FALSE
)

layout_intersect <- matrix(c(
  1.0, 2.0,
  2.0, 3.0,
  2.0, 1.0,
  4.0, 3.0,
  4.0, 1.0,
  5.0, 2.0,
  7.5, 3.0,
  7.5, 1.0
), ncol = 2, byrow = TRUE)

ggraph(gr_intersect_plt, layout = 'manual', x = layout_intersect[,1], y = layout_intersect[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, color = "white", fontface = "bold") +
  annotate("text", x = 1.5, y = 3.6, label = "G1", size = 5, fontface = "bold", color = "grey30") +
  annotate("text", x = 3.0, y = 2.0, label = "\u2229", size = 6, fontface = "bold", color = "grey30") +
  annotate("text", x = 4.5, y = 3.6, label = "G2", size = 5, fontface = "bold", color = "grey30") +
  annotate("text", x = 6.0, y = 2.0, label = "=", size = 6, fontface = "bold", color = "grey30") +
  annotate("text", x = 7.5, y = 3.6, label = "G1 \u2229 G2", size = 5, fontface = "bold", color = "grey30") +
  theme_graph() + 
  theme(plot.margin = margin(l = 15, r = 10, t = 10, b = 10, unit = "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::


