---
title: "Graph Theory: Connectivity & Components"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Base 5-node undirected graph (House layout pointing right)
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# Disconnected graph showing components and giant component
nodes_disconnected <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 1, 1, 0, -1, -1, 3, 4, 3.5),
  y    = c(1, 0.5, -0.5, -1, -0.5, 0.5, 0.5, 0, -1)
)
edges_disconnected <- data.frame(
  from = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 7, 7, 8),
  to   = c(2, 6, 3, 5, 4, 5, 5, 6, 6, 8, 9, 9)
)
gr_disconnected <- tbl_graph(nodes = nodes_disconnected, edges = edges_disconnected, directed = FALSE)

# 7-node directed graphs for connectivity flavors (large, non-toy)
nodes_7 <- data.frame(name = LETTERS[1:7])

edges_large_strong <- data.frame(
  from = c(1, 2, 3, 4, 5, 6, 7, 1, 3, 6), 
  to   = c(2, 3, 4, 5, 6, 7, 1, 4, 6, 2)
)
gr_large_strong <- tbl_graph(nodes = nodes_7, edges = edges_large_strong, directed = TRUE)

edges_large_uni <- data.frame(
  from = c(1, 2, 3, 4, 5, 6, 1, 3, 2), 
  to   = c(2, 3, 4, 5, 6, 7, 4, 6, 5)
)
gr_large_uni <- tbl_graph(nodes = nodes_7, edges = edges_large_uni, directed = TRUE)

edges_large_weak <- data.frame(
  from = c(1, 1, 2, 2, 2, 4, 5, 5, 6, 7, 7), 
  to   = c(2, 6, 1, 4, 3, 2, 3, 4, 1, 6, 4)
)
gr_large_weak <- tbl_graph(nodes = nodes_7, edges = edges_large_weak, directed = TRUE)

# 7-node directed graph for directed components slide
edges_dir_7 <- data.frame(
  from = c("A", "A", "B", "B", "B", "D", "E", "E", "F", "G", "G"),
  to   = c("B", "F", "A", "D", "C", "B", "C", "D", "A", "F", "D")
)
node_dir_7 <- data.frame(name = c("A", "B", "C", "D", "E", "F", "G"))
gr_dir_7 <- tbl_graph(nodes = node_dir_7, edges = edges_dir_7, directed = TRUE)

# Local bridge graph
nodes_bridge_local <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 0.5, 1, 2, 2.5, 3, 0.5, 1.5, 2.5),
  y    = c(0.5, -0.5, 0, 0, -0.5, 0.5, -1.2, -1.5, -1.2)
)
edges_bridge_local <- data.frame(
  from = c(1, 1, 2, 3, 4, 4, 5, 2, 7, 8, 9),
  to   = c(2, 3, 3, 4, 5, 6, 6, 7, 8, 9, 5)
)
gr_bridge_local <- tbl_graph(nodes = nodes_bridge_local, edges = edges_bridge_local, directed = FALSE)
```

# Part 1: Graph Connectivity & Components

---

## Connected vs. Disconnected Graphs

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Connected Graph (Single Component)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5

    gr_l_conn <- tbl_graph(
        nodes = data.frame(name = as.character(1:9)),
        edges = data.frame(
            from = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 3, 2),
            to   = c(2, 3, 4, 5, 6, 7, 8, 9, 1, 5, 7, 8)
        ),
        directed = FALSE
    )
    
    ggraph(gr_l_conn, layout = "circle") +
        geom_edge_link(color = "steelblue", edge_width = 1.0) +
        geom_node_point(size = 14, color = "tan2") +
        geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
        theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph (Two Components)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5

    gr_l_disconn <- tbl_graph(
        nodes = data.frame(name = as.character(1:9)),
        edges = data.frame(
            from = c(1, 2, 3, 4, 5, 6, 1, 7, 8, 9),
            to   = c(2, 3, 4, 5, 6, 1, 4, 8, 9, 7)
        ),
        directed = FALSE
    )
    
    ggraph(gr_l_disconn, layout = "circle") +
        geom_edge_link(color = "steelblue", edge_width = 1.0) +
        geom_node_point(size = 14, color = "tan2") +
        geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
        theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

- **Connected**: A path of some finite length exists between *every* pair of nodes in the graph (left).
- **Disconnected**: At least one pair of nodes is unreachable; the graph is split into separate pieces (right).

---

## Components & The Giant Component

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph with 2 Components**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
gr_disconnected_colors <- gr_disconnected |>
  activate(edges) |>
  mutate(highlight = ifelse(from <= 6 & to <= 6, "giant", "small"))

ggraph(gr_disconnected_colors, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("giant" = "red", "small" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("giant" = 1.8, "small" = 0.8), guide = "none") +
  geom_node_point(aes(color = ifelse(name %in% c("G", "H", "I"), "small", "giant")), size = 14) +
  scale_color_manual(values = c("giant" = "tan2", "small" = "skyblue"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: flex-start;"}
- **Component**: A maximally connected subgraph within a larger disconnected graph.
  - The graph on the left contains exactly **two components**.
- **Giant Component**: The largest connected component in a disconnected graph.
  - The red component $\{A, B, C, D, E, F\}$ is the **giant component** (of order 6).
  - The blue component $\{G, H, I\}$ is a smaller component (of order 3).
:::

::::

---

# Part 2: Edge and Node Connectivity (Cutsets & Bridges)

---

## Edge Cutsets & Connectivity

::: {.incremental}
- We can disconnect a connected graph by removing a subset of its edges.
- **Edge Cutset**:
  - Any set of edges whose removal disconnects the graph into two or more components.
- **Minimum Edge Cutset**:
  - An edge cutset containing the minimum possible number of edges needed to disconnect the graph.
- **Edge-Connectivity** ($\lambda(G)$):
  - The number of edges in a minimum edge cutset.
- **Bridge**:
  - A single edge whose removal disconnects the graph (a minimum edge cutset of size 1, so $\lambda(G) = 1$).
:::

---

## Edge Cutsets: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- In our 5-node graph on the right:
  - Edges **A-C** and **B-C** are highlighted in red (Before).
- **Removing both edges**:
  - Disconnects $\{C, D, E\}$ from $\{A, B\}$ (After).
  - This is an edge cutset of size **2**.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_cutset <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

p_before <- ggraph(gr_cutset, layout = 'kk') +
  geom_edge_link(aes(color = ifelse((from == 1 & to == 3) | (from == 2 & to == 3), "yes", "no")), edge_width = 1.0) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_cutset_after <- gr_cutset |>
  activate(edges) |>
  filter(!((from == 1 & to == 3) | (from == 2 & to == 3)))

p_after <- ggraph(gr_cutset_after, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Bridges: A Cutset of Size 1

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **Bridge**:
  - A single edge whose removal disconnects the graph (so $\lambda(G) = 1$).
- In our graph on the right, the edge **C-D** is highlighted in red.
  - **Removing edge C-D** completely disconnects $\{A, B, C\}$ from $\{D, E\}$.
  - Bridges represent the most critical, yet vulnerable transmission links in any network.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_bridge <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

p_before <- ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = ifelse(from == 3 & to == 4, "yes", "no")), edge_width = 1.0) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Bridge)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_bridge_after <- gr_bridge |>
  activate(edges) |>
  filter(!(from == 3 & to == 4))

p_after <- ggraph(gr_bridge_after, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Bridge Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::
---

## Special Properties of Bridges

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In a connected graph, a **bridge** (such as the red **C-D** edge on the right) possesses two interesting mathematical properties ([Buckley and Harary 1990](https://doi.org/10.1007/978-3-642-75242-1)):
- **1. The No-Cycle Rule**:
  - A bridge **never lies on any cycle** of the graph.
  - *If it did*, removing it would still leave the rest of the cycle intact, failing to disconnect the graph.
- **2. The Intermediation Bottleneck**:
  - The bridge is an edge on **every path** connecting at least one pair of nodes (e.g., any node in $\{A, B, C\}$ to any node in $\{D, E\}$).
  - *Every packet of information flowing between these clusters must pass through this single bottleneck.*
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = ifelse(from == 3 & to == 4, "yes", "no")), edge_width = 1.2) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::


---

## Local Bridges

::: {.incremental}
- In a classic paper on "The Strength of Weak Ties" (1973), Mark Granovetter developed the concept of a **local bridge**:
  - An edge that, if removed, does *not* disconnect the entire graph, but significantly increases the shortest path length between its end nodes.
- **Sociological Power**:
  - Local bridges act as crucial "shortcuts" that facilitate efficient, cross-group communication.
  - They are the primary channels through which novel information diffuses between separate social clusters.
:::

---

## Local Bridge: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Let's look at the edge **C-D** highlighted in red:
  - It connects the left cluster $\{A, B, C\}$ to the right cluster $\{D, E, F\}$.
- **Is it a bridge?** No. If removed, the graph remains connected via the bottom path ($B-G-H-I-E$).
- **Is it a local bridge?** Yes! 
  - Observed distance: $d(C, D) = 1$.
  - If removed, the distance increases to:
    $$d(C, D) = 6 \quad (C-B-G-H-I-E-D)$$
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_bridge_local |>
  activate(edges) |>
  mutate(highlight = ifelse(from == 3 & to == 4, "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Node Cutsets & Connectivity

::: {.incremental}
- We can also disconnect a graph by **removing nodes** (vertices).
- Removing a node also removes all edges incident to it—nodes "take their links with them"!
- **Node Cutset (Vertex Cutset)**:
  - Any set of nodes whose removal disconnects the remaining graph into two or more components.
- **Minimum Node Cutset**:
  - A node cutset containing the minimum possible number of nodes needed to disconnect the graph.
- **Node-Connectivity** ($\kappa(G)$):
  - The number of nodes in a minimum node cutset.
- **Articulation Node (Cutpoint)**:
  - A single node whose removal disconnects the graph (so $\kappa(G) = 1$).
:::

---

## Articulation Nodes: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- In our house graph on the right, node **E** is highlighted in red.
- **Removing node E**:
  - Removes all of its incident edges: $B-E, E-C, E-D$.
  - Completely disconnects node $B$ from the rest of the network.
  - This is a node cutset of size **1** (Node $E$ is an **articulation node**).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "E", "yes", "no")), size = 14) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Node Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_node <- gr_undir |>
  filter(name != "E")

p_after <- ggraph(gr_undir_after_node, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Node Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Special Properties of Articulation Nodes

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- An **articulation node** (or cutpoint, such as **E** on the right) has unique properties mirroring those of bridges:
- **1. Node Equivalent of a Bridge**:
  - A single actor upon whom the entire connectivity of the network depends ($\kappa(G) = 1$).
- **2. The Path Intermediator**:
  - It stands as an **inner node** on *every single path* connecting some node $v$ to the rest of the network (e.g., node $E$ is on every path connecting $B$ to $\{A, C, D\}$).
- **3. The End-Node Rule**:
  - A node of degree $1$ (an **end-node**, like $B$) can *never* be an articulation node; removing it always leaves the remaining graph connected.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(aes(color = ifelse(name == "E", "yes", "no")), size = 18) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Node Cutset: An Edge Cutset Comparison

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Let's look at node $A$ in our house graph on the right.
- **Separating $A$ with an edge cutset** requires removing **2 edges**: $C-A$ and $D-A$ (red).
- **Removing nodes can be more economical**: deleting the single node $E$ destroys three edges at once ($B-E$, $E-C$, $E-D$) and disconnects node $B$ from the rest of the network.
- **Key insight**: Node removals are doubly destructive — a node "takes its links with it," so one node removal can do the work of several edge removals.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 3 & to == 5) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Edge Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_edges <- gr_undir |>
  activate(edges) |>
  filter(!((from == 3 & to == 5) | (from == 4 & to == 5)))

p_after <- ggraph(gr_undir_after_edges, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Edges Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## k-Connected Graphs & Robustness

::: {.incremental}
- **$k$-Connected Graph**:
  - A graph is $k$-connected (or $k$-node connected) if the minimum number of nodes required to disconnect it is $k$.
  - **1-Connected**: Removing 1 node disconnects the graph (contains an articulation point).
  - **2-Connected**: Requires removing at least 2 nodes to disconnect it.
  - **3-Connected**: Requires removing at least 3 nodes to disconnect it.
- **Menger's Theorem (1927)**:
  - If a graph is $k$-connected, there are at least $k$ node-independent paths (paths that do not share any common inner nodes) connecting each non-adjacent pair of vertices.
- **Social Cohesion**: Higher $k$-connectivity indicates a more robust, cohesive, and redundant communication structure.
:::

---

## 1-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **1-Connected Graph**:
  - A graph where removing **exactly 1 node** can disconnect the network.
- **Before** (Left):
  - A 7-node symmetric bow-tie graph, with the central bridge node (node 7) highlighted in red.
- **After** (Right):
  - The central node 7 is removed, leaving two completely separate connected components (triangles).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_k1 <- tbl_graph(
  nodes = data.frame(
    name = as.character(1:7),
    x = c(-2, -2, -1, 1, 2, 2, 0),
    y = c(1, -1, 0, 0, 1, -1, 0)
  ),
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 3, 4),
    to   = c(2, 3, 1, 5, 6, 4, 7, 7)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k1, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "7", "yes", "no")), size = 14) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (7-Node Bow-Tie with Cutpoint 7)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_k1_after <- gr_k1 |> filter(name != "7")
p_after <- ggraph(gr_k1_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Node 7 Removed, Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## 2-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **2-Connected Graph**:
  - A graph where it requires removing at least **2 nodes** to disconnect it.
- **Before** (Left):
  - A 6-node cycle with a diagonal chord (edge 2-5), with nodes 2 and 5 highlighted in red.
- **After** (Right):
  - Removing nodes 2 and 5 leaves the graph split into two disconnected components: {1, 6} and {3, 4}.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_k2 <- tbl_graph(
  nodes = data.frame(name = as.character(1:6)),
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 2),
    to   = c(2, 3, 4, 5, 6, 1, 5)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k2, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name %in% c("2", "5"), "yes", "no")), size = 14) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Before (6-Node Ring with Chord, Cutset {2, 5})") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

gr_k2_after <- gr_k2 |> filter(!name %in% c("2", "5"))
p_after <- ggraph(gr_k2_after, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("After (Nodes 2 & 5 Removed, Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## 3-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **3-Connected Graph**:
  - A graph where it requires removing at least **3 nodes** to disconnect it.
- **Before** (Left):
  - A 3-regular 6-node prism graph (two concentric triangles), with nodes 1, 2, and 6 highlighted in red.
- **After** (Right):
  - Removing nodes 1, 2, and 6 leaves the single edge 4-5 and completely isolates node 3 (disconnected).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
nodes_prism <- data.frame(
  name = as.character(1:6),
  x = c(0, -0.43, 0.43, 0, -1.04, 1.04),
  y = c(0.5, -0.25, -0.25, 1.2, -0.6, -0.6)
)
gr_k3 <- tbl_graph(
  nodes = nodes_prism,
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 1, 2, 3),
    to   = c(2, 3, 1, 5, 6, 4, 4, 5, 6)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k3, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name %in% c("1", "2", "6"), "yes", "no")), size = 14) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Before (6-Node Prism, Cutset {1, 2, 6})") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

gr_k3_after <- gr_k3 |> filter(!name %in% c("1", "2", "6"))
p_after <- ggraph(gr_k3_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("After (Nodes Removed, Node 3 Isolated)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Graph Cutsets: Quantitative Cohesion

::: {.incremental}
- **Cohesion** can be directly translated to graph-theoretic cutsets:
  - **Cohesive Groups**: Have large edge-cutsets and node-cutsets. Many connections or individuals must be removed to disrupt or fragment the group.
  - **Fragmented Groups**: Have small cutsets (such as bridges or single cutpoints). They are vulnerable to rapid disruption if key facilitators are removed.
- Any network can be decomposed into hierarchical subgraphs with specific levels of $k$-connectivity.
:::

---

## The Connectivity Inequality

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- For any undirected graph $G$, node-connectivity ($\kappa$), edge-connectivity ($\lambda$), and minimum degree ($\delta$) obey a fundamental structural hierarchy:
  $$\kappa(G) \le \lambda(G) \le \delta(G)$$
- **A Strict Inequality Example ($1 < 2 < 3$)**:
  - In our 9-node "double-clique" graph on the right:
    - **Minimum Degree ($\delta = 3$)**: The least connected nodes ($3, 4, 7, 8$) have exactly 3 edges.
    - **Edge-Connectivity ($\lambda = 2$)**: Removing $\{1-9, 2-9\}$ separates the left clique from the waist. No single bridge exists.
    - **Node-Connectivity ($\kappa = 1$)**: Removing the central cutpoint (node 9) completely disconnects the graph.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_ineq <- tbl_graph(
  nodes = data.frame(
    name = as.character(1:9),
    x = c(-1, -1, -2, -2, 1, 1, 2, 2, 0),
    y = c(0.5, -0.5, 0.5, -0.5, 0.5, -0.5, 0.5, -0.5, 0)
  ),
  edges = data.frame(
    from = c(1, 1, 1, 2, 2, 3, 5, 5, 5, 6, 6, 7, 1, 2, 5, 6),
    to   = c(2, 3, 4, 3, 4, 4, 6, 7, 8, 7, 8, 8, 9, 9, 9, 9)
  ),
  directed = FALSE
)

ggraph(gr_ineq, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

# Part 3: Directed Connectivity & Components

---

## Pairwise Connectivity in Directed Graphs

- Before we can define whether a **whole graph** has a certain flavor of connectivity, we must define the connectivity state of **individual pairs of nodes (pairwise connectivity)**.
- Two nodes $u$ and $v$ can be connected in four distinct pairwise flavors depending on the directed pathways between them:
  1. **Pairwise Weakly Connected**
  2. **Pairwise Unilaterally Connected**
  3. **Pairwise Strongly Connected**
  4. **Pairwise Recursively Connected**
- Let's examine these four pairwise states step-by-step.

---

## 1. Pairwise Weak & Unilateral Connectivity

- **Pairwise Weakly Connected**:
  - Node $u$ and $v$ are connected by at least one **semipath** (connected if we ignore the arrow directions).
  - **Sociological Meaning**: They are part of the same general network, but there is no directed path starting from either node that can reach the other.
- **Pairwise Unilaterally Connected**:
  - A directed path exists from $u \rightarrow v$ **OR** from $v \rightarrow u$ (one-way reachability).
  - **Sociological Meaning**: Information or influence can flow from one person to the other, but not in both directions (no mutual exchange).

---

## 2. Pairwise Strong & Recursive Connectivity

- **Pairwise Strongly Connected**:
  - A directed path exists from $u \rightarrow v$ **AND** from $v \rightarrow u$ (mutual reachability).
  - **Sociological Meaning**: Both nodes can reach each other, but the paths in each direction may use completely different intermediate nodes or channels.
- **Pairwise Recursively Connected**:
  - A directed path exists from $u \rightarrow v$ and $v \rightarrow u$, **AND** the directed path in both directions shares the exact same set of intermediate nodes (sharing both vertices and edges).
  - **Sociological Meaning**: The relationship can be traversed back-and-forth along the same closed, mutually symmetric cycle.

---

## Node-Level Pairwise vs. Graph-Level Connectivity

- A **whole graph**'s flavor of connectivity is defined by the **weakest pairwise connectivity state** present in the network:

::: {.compact-table}
| Whole Graph State | Defining Pairwise Condition |
|:---|:---|
| **Strongly Connected** | **every** pair of nodes is *pairwise strongly connected*. |
| **Unilaterally Connected** | **every** pair is *at least* unilaterally connected, and **at least one** pair is *no better than* unilaterally connected. |
| **Weakly Connected** | **every** pair is *at least* weakly connected, and **at least one** pair is *no better than* weakly connected. |
:::

- **Takeaway**: Graph connectivity is built on top of these pairwise states — if even one pair drops to a lower state, the whole graph's connectivity level drops!

---

## Strongly Connected: A Visual Example

:::: {.columns align=center}

::: {.column width="45%"}
::: {.incremental}
- **Strong Connectivity**:
  - A directed graph is strongly connected if there is a directed path from **every node to every other node** (mutual reachability).
- **7-Node Example** (Right):
  - Consists of a complete spanning cycle ($A \rightarrow B \rightarrow C \rightarrow D \rightarrow E \rightarrow F \rightarrow G \rightarrow A$) plus chord arcs.
  - Because of the continuous cycle, any node can reach any other node, ensuring complete mutual reachability.
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_large_strong, layout = "circle") +
  geom_edge_parallel(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'),
                     color = "steelblue", edge_width = 1.2, sep = unit(3.5, 'mm')) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, color = "white", fontface="bold") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Unilaterally Connected: A Visual Example

:::: {.columns align=center}

::: {.column width="45%"}
::: {.incremental}
- **Unilateral Connectivity**:
  - A directed graph is unilaterally connected if for every pair of nodes, there is **at least one directed path** connecting them.
- **7-Node Example** (Right):
  - Same 7 nodes, but the feedback edge $G \rightarrow A$ and other backward chords are removed. Edges flow strictly forward ($A \rightarrow B \rightarrow C \dots \rightarrow G$).
  - Here, $A$ can reach $G$, but $G$ cannot reach $A$ (no mutual reachability).
  - However, for every single pair, at least one directed pathway exists.
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_large_uni, layout = "circle") +
  geom_edge_parallel(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'),
                     color = "steelblue", edge_width = 1.2, sep = unit(3.5, 'mm')) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, color = "white", fontface="bold") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Weakly Connected: A Visual Example

:::: {.columns align=center}

::: {.column width="45%"}
::: {.incremental}
- **Weak Connectivity**:
  - A directed graph that is connected only if we **ignore the direction of the arrowheads** (treating it like an undirected graph).
- **7-Node Example** (Right):
  - Uses our textbook's 7-node directed graph.
  - Nodes $E$ and $G$ are transmitter nodes (indegree 0) that cannot reach each other, nor can they be reached by anyone else.
  - Thus, $\{E, G\}$ is a mutually unreachable pair (violating unilateral connectivity).
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_large_weak, layout = "circle") +
  geom_edge_parallel(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'),
                     color = "steelblue", edge_width = 1.2, sep = unit(3.5, 'mm')) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, color = "white", fontface="bold") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Decomposing Directed Components

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- Like undirected graphs, directed graphs can be decomposed into:
- **Weakly Connected Components (WCCs)**:
  - Subgraphs connected when ignoring edge directions.
  - Yields **one WCC** containing all 7 nodes: $\{A, B, C, D, E, F, G\}$.
- **Strongly Connected Components (SCCs)**:
  - Subgraphs where every pair is mutually reachable.
  - Yields **four SCCs**:
    - $\{A, B, D, F\}$ (the core "rumor mill", highlighted in red).
    - Singletons: $\{C\}$, $\{E\}$, $\{G\}$.
:::
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_dir_7, layout = 'kk') +
  geom_edge_parallel(color = "steelblue", edge_width = 1.2,
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'), sep = unit(3, 'mm')) +
  geom_node_point(aes(color = ifelse(name %in% c("A", "B", "D", "F"), "yes", "no")), size = 18) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Wrap-Up: Connectivity & Components

::: {.incremental}
- **1. Components**: Connected vs. disconnected graphs and the giant component.
- **2. Cutsets & Robustness**: Bridges, local bridges, articulation nodes, $k$-connectivity, and the hierarchy $\kappa \le \lambda \le \delta$.
- **3. Directed Connectivity**: The weak, unilateral, strong, and recursive flavors; weakly vs. strongly connected components.
- **Next up**: Tree graphs — the minimally connected networks.
:::
