---
title: "Graph Theory: Trees, Forests & Efficiency"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Helper function to generate connected random graphs
connected_gnm <- function(n, m) {
  repeat {
    g <- play_gnm(n = n, m = m, directed = FALSE, loops = FALSE)
    if (is_connected(g)) return(g)
  }
}
```

# Tree Graphs & Forests

---

## What is a Tree Graph?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- If a graph is both **connected** *and* has no **cycles** (acyclic), then it is a **tree graph**.
- **The Cycle Restriction**:
  - A cycle is a path that begins and ends with the same node.
  - In a tree graph, you can never start with a node and trace a sequence of distinct edges that take you back to where you started!
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_prop <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(0, 1, 1, 2, 3), y = c(0, 0, 1, 0, 0))
edges_prop <- data.frame(from = c(1, 2, 2, 4), to = c(2, 3, 4, 5)) # A-B, B-C, B-D, D-E
gr_prop <- tbl_graph(nodes = nodes_prop, edges = edges_prop, directed = FALSE)

ggraph(gr_prop, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Properties of Tree Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Tree graphs possess several unique mathematical and structural properties:
  1. **The Single Path Property**:
     - *Every* pair of nodes in a tree graph is linked by exactly **one unique path**.
     - There are no redundant or alternative paths between any two actors.
  2. **The One-Cycle Property**:
     - Adding a *single* new edge connecting any pair of non-adjacent nodes always creates **exactly one cycle**, destroying its tree status.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
edges_prop_cycle <- data.frame(from = c(1, 2, 2, 4, 3), to = c(2, 3, 4, 5, 5)) # A-B, B-C, B-D, D-E, C-E (added)
gr_prop_cycle <- tbl_graph(nodes = nodes_prop, edges = edges_prop_cycle, directed = FALSE)

gr_prop_cycle %>%
  activate(edges) %>%
  mutate(type = c("tree", "tree", "tree", "tree", "added")) %>%
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = type, linetype = type, edge_width = type)) +
  scale_edge_color_manual(values = c("tree" = "steelblue", "added" = "red"), guide = "none") +
  scale_edge_linetype_manual(values = c("tree" = "solid", "added" = "dashed"), guide = "none") +
  scale_edge_width_manual(values = c("tree" = 1.2, "added" = 1.5), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Tree Graph Metrics: Vertices & Edges

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In any tree graph, the number of edges ($m$) is always exactly equal to the number of nodes ($n$) minus one:
  $$m = n - 1 \qquad \iff \qquad n = m + 1$$
- **Sum of Degrees Identity**:
  - Recall that the sum of degrees in any undirected graph is $2m$. 
  - Substituting $m = n - 1$, the sum of degrees of a tree graph is always:
    $$\sum_i k_i = 2(n - 1) = 2n - 2$$
  - *In a tree graph, the sum of degrees is always exactly twice the number of nodes minus two!*
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_metric <- data.frame(
  name = c("A", "B", "C", "D", "E"), 
  deg = c("k=1", "k=3", "k=1", "k=2", "k=1"),
  x = c(0, 1, 1, 2, 3), 
  y = c(0, 0, 1, 0, 0)
)
gr_metric <- tbl_graph(nodes = nodes_metric, edges = edges_prop, directed = FALSE)

ggraph(gr_metric, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  geom_node_text(aes(label = deg), size = 5, fontface = "italic", color = "black", vjust = -1.8) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Tree Graph Density

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Because of the strict $m = n - 1$ edge constraint, tree graph density simplifies beautifully.
- **Density of a Tree Graph**:
  $$d(T) = \frac{2m}{n(n-1)} = \frac{2(n-1)}{n(n-1)} = \frac{2}{n}$$
- *For our 5-node tree graph on the right ($n = 5$)*:
  - The density is:
    $$d(T) = \frac{2}{5} = 0.40 \quad (40\%)$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_metric, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Tree Graph Average Degree

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Average degree in a tree graph also simplifies beautifully due to the edge constraint.
- **Average Degree of a Tree Graph**:
  $$\bar{k} = \frac{\sum k_i}{n} = \frac{2m}{n} = \frac{2n - 2}{n}$$
- *For our 5-node tree graph on the right ($n = 5$)*:
  - The average degree is:
    $$\bar{k} = \frac{2(5) - 2}{5} = \frac{8}{5} = 1.6$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_metric, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Trees as Minimally Connected Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Trees are **minimally connected** graphs:
  - They are connected using the *smallest possible number of edges* needed to connect $n$ vertices.
- **The Bridges Rule**:
  - Removing even a *single* edge from a tree graph completely disconnects the graph.
  - Thus, **every edge of a tree graph is a bridge** (a minimum edge cutset of size 1)!
- Disconnecting a tree graph by removing any edge splits it into separate connected components, which are themselves trees.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_bridge_vis <- gr_prop %>%
  activate(edges) %>%
  mutate(highlight = ifelse(from == 2 & to == 4, "bridge", "normal"))

ggraph(gr_bridge_vis, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight, linetype = highlight)) +
  scale_edge_color_manual(values = c("bridge" = "red", "normal" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("bridge" = 2.0, "normal" = 1.2), guide = "none") +
  scale_edge_linetype_manual(values = c("bridge" = "dashed", "normal" = "solid"), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Numerical Example: 8-Node Tree

:::: {.columns align=center}

::: {.column width="55%"}
Let's compute the metrics for the 8-node tree shown on the right:

- **1. Nodes and Edges**:
  - Number of nodes: $n = 8$
  - Number of edges: $m = n - 1 = 7$
- **2. Degree of each node**:
  - $k_A = 2, k_B = 3, k_C = 3, k_D = 1, k_E = 2, k_F = 1, k_G = 1, k_H = 1$
  - Sum of degrees: $\sum k_i = 14$
  - Check: $2(n-1) = 2(7) = 14 \quad \checkmark$
- **3. Network Density**:
  - $d(T) = \frac{2}{n} = \frac{2}{8} = 0.25 \quad (25\%)$
- **4. Average Degree**:
  - $\bar{k} = \frac{2n - 2}{n} = \frac{14}{8} = 1.75$
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_8 <- data.frame(
  name = LETTERS[1:8],
  x = c(0, -1, 1, -2, -0.5, 0.5, 2, -0.5),
  y = c(1, 0, 0, -1, -1, -1, -1, -2)
)
edges_8 <- data.frame(
  from = c(1, 1, 2, 2, 3, 3, 5),
  to   = c(2, 3, 4, 5, 6, 7, 8)
)
gr_8 <- tbl_graph(nodes = nodes_8, edges = edges_8, directed = FALSE)

ggraph(gr_8, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Forests: Visualizing Disconnection

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **Forest**:
  - A disconnected graph whose components are all tree graphs.
- **Before** (Left):
  - A connected 9-node tree graph.
- **After** (Right):
  - Removing edge **D-H** splits the graph into two separate trees (a **forest**):
    - A **bigger tree** of 6 nodes on the left ($\{A, B, C, D, E, F\}$).
    - A **smaller tree** of 3 nodes on the right ($\{G, H, I\}$).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
nodes_f <- data.frame(
  name = LETTERS[1:9],
  x = c(0, 1, 0, 2, 3, 3, 5, 4, 5),
  y = c(0.5, 0, -0.5, 0, 0.5, -0.5, 0.5, 0, -0.5)
)
edges_f <- data.frame(
  from = c(2, 2, 2, 4, 4, 4, 8, 8),
  to   = c(1, 3, 4, 5, 6, 8, 7, 9)
)
gr_f <- tbl_graph(nodes = nodes_f, edges = edges_f, directed = FALSE)

p_before <- ggraph(gr_f, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = ifelse(from == 4 & to == 8, "yes", "no")), edge_width = 1.0) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (Connected Tree)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_f_after <- gr_f %>% 
  activate(edges) %>% 
  filter(!(from == 4 & to == 8))

p_after <- ggraph(gr_f_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Forest: 2 Disconnected Trees)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Special Types of Trees: Line, Star & Caterpillar Graphs

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- **The Line Graph ($L_n$)**:
  - A single straight path connecting $n$ nodes.
  - Used to model sequential diffusion processes (like the "Telephone Game").
- **The Star Graph ($S_n$)**:
  - A central hub connected to $n-1$ leaf nodes.
  - Represents extreme structural centralization.
- **The Caterpillar Graph**:
  - A central path ("body") with pendant leaf nodes ("legs") sticking out of it.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 4.5
gr_line <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]), edges = data.frame(from = 1:4, to = 2:5), directed = FALSE)
p_line <- ggraph(gr_line, layout = "linear") +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Line Graph (L5)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_star <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]), edges = data.frame(from = c(1, 1, 1, 1), to = 2:5), directed = FALSE)
p_star <- ggraph(gr_star, layout = "star") +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Star Graph (S5)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

fr <- c("A", "A", "C", "C", "E", "E", "G")
to <- c("B", "C", "D", "E", "F", "G", "H")
edge.dat <- data.frame(fr, to)
node.dat <- data.frame(name = union(fr, to))
gr_cat <- tbl_graph(edges = edge.dat, nodes = node.dat, directed = FALSE)
l_cat <- ggraph(gr_cat, layout = 'grid')
l_cat_mat <- as.matrix(l_cat$data[, 1:2])
l_cat_mat[,"x"] <- c(0, 1, 2, 3, 0, 1, 2, 3)
l_cat_mat[,"y"] <- c(1, 1, 1, 1, 0, 0, 0, 0)

p_cat <- ggraph(gr_cat, layout = l_cat_mat) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Caterpillar Graph (Order 8)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_line + p_star + p_cat
```
:::

::::

---

## The Graph Efficiency

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Concept** (Krackhardt, 1994):
  - Measures how **economical** the connectivity of a graph is.
  - A graph is efficiently connected if it uses the *minimum* possible number of edges to remain connected.
- **Formula**:
  $$F(G) = 1 - \left[\frac{2(m - n + 1)}{(n-1)(n-2)}\right]$$
  - Where $m$ is the number of edges and $n$ is the number of nodes.
- **Numerical Proof (The Tree on the Right)**:
  - Pictured on the right is a tree of $n = 5$ nodes and $m = 4$ edges.
  - Substituting $n = 5$ and $m = 4$ into the formula:
    $$F(G) = 1 - \left[\frac{2(4 - 5 + 1)}{(5-1)(5-2)}\right]$$
    $$F(G) = 1 - \left[\frac{2(0)}{4 \times 3}\right] = 1 - 0 = 1.0$$
- **Beyond Trees (Decreasing Efficiency)**:
  - As edges are added beyond $m = n - 1$, the numerator $(m - n + 1) > 0$, causing $F(G)$ to decrease toward $0.0$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_tree <- data.frame(name = LETTERS[1:5])
edges_tree <- data.frame(from = c(1, 2, 2, 3), to = c(2, 3, 4, 5))
gr_tree_eff <- tbl_graph(nodes = nodes_tree, edges = edges_tree, directed = FALSE)
ggraph(gr_tree_eff, layout = 'nicely') +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Maximally Efficient (Tree: F(G) = 1.0)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::

---

## Detailed Example: Graph 1 (Less Efficient)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph 1 Structure**:
  - Contains $n = 5$ nodes and $m = 7$ edges.
- **Efficiency Calculation**:
  - We substitute $n=5$ and $m=7$ into the formula:
    $$F(G) = 1 - \left[\frac{2(7 - 5 + 1)}{(5-1)(5-2)}\right]$$
    $$F(G) = 1 - \left[\frac{2(3)}{4 \times 3}\right] = 1 - \frac{6}{12} = 0.50$$
- **Interpretation**:
  - With 5 nodes, a connected tree needs only 4 edges. 
  - This graph has 3 redundant edges ($7 - 5 + 1 = 3$), resulting in a low efficiency score of **$0.50$**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
set.seed(123)
gr1 <- connected_gnm(5, 7) %>% mutate(name = LETTERS[1:5])
ggraph(gr1, layout = 'nicely') +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Graph 1 (n = 5, m = 7, F = 0.50)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::

---

## Detailed Example: Graph 2 (More Efficient)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph 2 Structure**:
  - Contains $n = 7$ nodes and $m = 10$ edges.
- **Efficiency Calculation**:
  - We substitute $n=7$ and $m=10$ into the formula:
    $$F(G) = 1 - \left[\frac{2(10 - 7 + 1)}{(7-1)(7-2)}\right]$$
    $$F(G) = 1 - \left[\frac{2(4)}{6 \times 5}\right] = 1 - \frac{8}{30} \approx 0.73$$
- **Conclusion**:
  - Despite having more edges ($10$ vs $7$), Graph 2 is larger. Its redundancy is distributed over more nodes, yielding a higher efficiency of **$0.73$**.
  - **Graph 2 is more efficient than Graph 1 ($0.73 > 0.50$)**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
set.seed(123)
gr2 <- connected_gnm(7, 10) %>% mutate(name = LETTERS[1:7])
ggraph(gr2, layout = 'nicely') +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Graph 2 (n = 7, m = 10, F = 0.73)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::

---

## Graph Efficiency: Complete Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Maximally Inefficient**:
  - The complete graph ($K_n$) is minimally efficient because it has the maximum possible number of edges.
  - Number of edges: $m = \frac{n(n-1)}{2}$
- **Example ($n=5$ Complete Graph)**:
  - Edges: $m = \frac{5 \times 4}{2} = 10$
  - Calculation:
    $$F(G) = 1 - \left[\frac{2(10 - 5 + 1)}{(5-1)(5-2)}\right]$$
    $$= 1 - \left[\frac{2(6)}{4 \times 3}\right] = 1 - \frac{12}{12} = 0.0$$
  - Thus, any complete graph yields $F(G) = 0.0$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_complete <- play_gnm(n = 5, m = 10, directed = FALSE, loops = FALSE) %>%
  mutate(name = LETTERS[1:5])
ggraph(gr_complete, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Complete Graph (F(G) = 0.0)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::

---

## Wrap-Up: Trees, Forests & Efficiency

::: {.incremental}
- **Trees** are minimally connected: exactly one path between any two nodes, every edge a bridge, and $m = n - 1$.
- **Forests** are what remains when a tree loses an edge — disjoint collections of trees.
- **Efficiency** (Krackhardt, 1994) scores how economical connectivity is, from trees ($F = 1$) to complete graphs ($F = 0$).
- **Next up**: We translate this graph structure into the language of matrices.
:::
