---
title: "Graph Theory: Paths & Connectivity"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Helper function to generate connected random graphs
connected_gnm <- function(n, m) {
  repeat {
    g <- play_gnm(n = n, m = m, directed = FALSE, loops = FALSE)
    if (is_connected(g)) return(g)
  }
}

# Base 5-node undirected graph (House layout pointing right)
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# Base 5-node directed graph
edges_dir <- data.frame(
  from = c("A", "A", "C", "D", "E", "D", "E", "B", "B", "E"),
  to   = c("C", "D", "E", "C", "D", "E", "B", "E", "D", "C")
)
l_dir <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
gr_dir <- tbl_graph(nodes = l_dir, edges = edges_dir, directed = TRUE)

# Disconnected graph showing components and giant component
nodes_disconnected <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 1, 1, 0, -1, -1, 3, 4, 3.5),
  y    = c(1, 0.5, -0.5, -1, -0.5, 0.5, 0.5, 0, -1)
)
edges_disconnected <- data.frame(
  from = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 7, 7, 8),
  to   = c(2, 6, 3, 5, 4, 5, 5, 6, 6, 8, 9, 9)
)
gr_disconnected <- tbl_graph(nodes = nodes_disconnected, edges = edges_disconnected, directed = FALSE)

# 7-node directed graphs for connectivity flavors (large, non-toy)
nodes_7 <- data.frame(name = LETTERS[1:7])

edges_large_strong <- data.frame(
  from = c(1, 2, 3, 4, 5, 6, 7, 1, 3, 6), 
  to   = c(2, 3, 4, 5, 6, 7, 1, 4, 6, 2)
)
gr_large_strong <- tbl_graph(nodes = nodes_7, edges = edges_large_strong, directed = TRUE)

edges_large_uni <- data.frame(
  from = c(1, 2, 3, 4, 5, 6, 1, 3, 2), 
  to   = c(2, 3, 4, 5, 6, 7, 4, 6, 5)
)
gr_large_uni <- tbl_graph(nodes = nodes_7, edges = edges_large_uni, directed = TRUE)

edges_large_weak <- data.frame(
  from = c(1, 1, 2, 2, 2, 4, 5, 5, 6, 7, 7), 
  to   = c(2, 6, 1, 4, 3, 2, 3, 4, 1, 6, 4)
)
gr_large_weak <- tbl_graph(nodes = nodes_7, edges = edges_large_weak, directed = TRUE)

# 7-node directed graph for directed components slide
edges_dir_7 <- data.frame(
  from = c("A", "A", "B", "B", "B", "D", "E", "E", "F", "G", "G"),
  to   = c("B", "F", "A", "D", "C", "B", "C", "D", "A", "F", "D")
)
node_dir_7 <- data.frame(name = c("A", "B", "C", "D", "E", "F", "G"))
gr_dir_7 <- tbl_graph(nodes = node_dir_7, edges = edges_dir_7, directed = TRUE)


# Local bridge graph
nodes_bridge_local <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 0.5, 1, 2, 2.5, 3, 0.5, 1.5, 2.5),
  y    = c(0.5, -0.5, 0, 0, -0.5, 0.5, -1.2, -1.5, -1.2)
)
edges_bridge_local <- data.frame(
  from = c(1, 1, 2, 3, 4, 4, 5, 2, 7, 8, 9),
  to   = c(2, 3, 3, 4, 5, 6, 6, 7, 8, 9, 5)
)
gr_bridge_local <- tbl_graph(nodes = nodes_bridge_local, edges = edges_bridge_local, directed = FALSE)

# Bridge graph (Strict bridge)
gr_bridge <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

# Symmetrical 5-node undirected graph for Hamiltonian Path & Cycle comparisons
nodes_ham <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(0, -1, -0.6, 0.6, 1), y = c(1, 0.2, -1, -1, 0.2))
edges_ham <- data.frame(from = c(1, 2, 3, 4, 5, 2), to = c(2, 3, 4, 5, 1, 5))
gr_ham <- tbl_graph(nodes = nodes_ham, edges = edges_ham, directed = FALSE)
```

# Part 1: Indirect Connections & Paths

---

## Adjacency vs. Indirect Connections

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Adjacency** captures direct connections between nodes:
  - If $A$ is adjacent to $B$, they share an edge.
- But many network processes operate **indirectly**:
  - We must look beyond immediate contacts to the "friends of our friends."
  - **Diffusion Channels**: Indirect connections act as channels for gossip, information transmission, disease contagion, or social influence.
- In graph theory, indirect connections are formalized using three core concepts:
  - **Path**, **Connectivity**, and **Reachability**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Path?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **path** between $u$ and $v$ is an alternating sequence of nodes and edges that:
  - Begins with an **origin node** ($u$).
  - Ends with a **destination node** ($v$).
  - Does **not repeat any nodes** (which implies edges are not repeated either!).
- **Anatomy of a Path** (e.g. Red path from $B$ to $D$ on the right):
  - Written as: $B-D = \{BE, ED\}$ (length $l = 2$).
  - **Origin**: $B$, **Destination**: $D$ (together called the **end nodes**).
  - **Inner Nodes**: $E$ (the nodes in between).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Paths & Geodesics

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Multiple distinct paths can connect the same pair of nodes.
- Let's trace paths from $B$ to $A$:
  - $P_1 = \{BE, ED, DA\} \quad (l = 3)$
  - $P_2 = \{BE, EC, CA\} \quad (l = 3)$
  - $P_3 = \{BE, ED, DC, CA\} \quad (l = 4)$
  - $P_4 = \{BE, EC, CD, DA\} \quad (l = 4)$
- **Shortest Path (Geodesic)**:
  - The path of minimum length between two nodes.
  - **Geodesic Distance**: The length of the geodesic path (here, geodesic distance $d(B, A) = 3$, with tied geodesics $P_1$ and $P_2$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Shortest Paths (Geodesic Ties)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In real networks, **geodesic distances are often tied**:
  - There can be multiple distinct shortest paths of equal minimum length connecting the exact same pair of nodes.
- **Example**: Shortest paths from $B$ to $A$ in our undirected graph:
  - Both paths have a geodesic distance of $d(B, A) = 3$.
  - **Geodesic Path 1**: $B \rightarrow E \rightarrow D \rightarrow A$ (Top)
  - Let's visualize this path highlighted in red on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
p1 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Geodesic 1 (B-E-D-A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
p1
```
:::

::::

---

## Geodesic Ties: Redundancy & Reliability

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at the second geodesic path from $B$ to $A$:
  - **Geodesic Path 2**: $B \rightarrow E \rightarrow C \rightarrow A$ (Bottom)
  - Let's visualize this path highlighted in red on the right.
- **Redundancy and Reliability**:
  - Having multiple geodesics of equal minimum length provides alternative routes of maximum efficiency.
  - **Structural Robustness**: If a node or edge on one path fails (e.g. node $D$), communication is completely unaffected because Path 2 remains open!
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
p2 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Geodesic 2 (B-E-C-A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
p2
```
:::

::::

---

## Sociological Significance of Geodesics

::: {.incremental}
- Why do social scientists care so much about shortest paths (geodesics) in real networks?
- **1. Speed and Bandwidth of Transmission**:
  - Information, trust, and influence decay as they travel along long paths.
  - Geodesic paths represent the channels through which diffusion (of gossip, innovation, or virus) happens **fastest** and with **least loss**.
- **2. Power, Brokerage, and Betweenness**:
  - Actors who lie on the geodesics of *many other pairs* of nodes in the network play a major role as **information brokers** or gatekeepers.
  - They control information flow, filter messages, and can extract rents or gain social influence (this is the basis of **Betweenness Centrality**).
- **3. Search and the "Small World" Phenomenon**:
  - Stanley Milgram's classic "six degrees of separation" experiment showed that human social networks have surprisingly short geodesic distances.
  - Even in massive networks of millions of people, any two individuals are connected by incredibly short paths of friendship!
:::

---


## Node Eccentricity

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: The **eccentricity** of a node is the maximum geodesic distance from that node to any other node in the network.
  - It measures the "worst-case" distance to reach any other actor.
- **Example 1: Node E (Central Hub)**:
  - Distances from $E$: $d(E, B) = 1$, $d(E, C) = 1$, $d(E, D) = 1$, $d(E, A) = 2$.
  - The maximum distance is $2$ (to node $A$).
  - **Eccentricity of $E$ = 2**.
- **Example 2: Node B (Peripheral Leaf)**:
  - Distances from $B$: $d(B, E) = 1$, $d(B, C) = 2$, $d(B, D) = 2$, $d(B, A) = 3$.
  - The maximum distance is $3$ (to node $A$).
  - **Eccentricity of $B$ = 3**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Cycle?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **cycle** is a path of length 3 or larger that starts and ends on the same node.
- **Key Concepts**:
  - Origin and destination are identical (the same vertex).
  - All inner nodes and all edges must be completely unique.
  - Cycles are the fundamental building blocks of network redundancy and feedback loops.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Visualizing a Simple Cycle

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at a simple **cycle** highlighted in our graph:
  - **Cycle**: $C \rightarrow D \rightarrow A \rightarrow C$.
  - It has a length of **3** (consisting of 3 edges).
- **Verification of Properties**:
  - Origin and destination are identical (both are node $C$).
  - All intermediate nodes ($D$, $A$) are unique.
  - All traversed edges are unique.
  - This closed loop creates a fundamental feedback pathway.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 3 & to == 5) | (from == 4 & to == 5) | (from == 3 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Girth

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: The **girth** of a graph is the length of its **shortest cycle**.
- **Properties**:
  - If a graph contains no cycles at all (like a tree graph), its girth is defined as **infinity** ($\infty$).
  - For our 5-node undirected graph on the right, the shortest cycle is the triangle:
    $$E \rightarrow D \rightarrow C \rightarrow E$$
  - Thus, the **girth** of this graph is **3**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Circumference

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: The **circumference** of a graph is the length of its **longest cycle**.
- **Properties**:
  - If a graph is acyclic, its circumference is defined as **zero** or undefined.
  - For our 5-node undirected graph on the right, the longest cycle is the four-loop:
    $$E \rightarrow D \rightarrow A \rightarrow C \rightarrow E$$
  - Thus, the **circumference** of this graph is **4**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 5) | (from == 4 & to == 5) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Special Paths: Hamiltonian Paths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Hamiltonian Path**: A path that visits **every single node** in the graph exactly once.
- **Our Example Graph**: Shown on the right, it has 5 nodes $\{A, B, C, D, E\}$.
- **A Hamiltonian Path**: Let's trace one in red:
  $$A \rightarrow B \rightarrow C \rightarrow D \rightarrow E$$
  - Every node is visited exactly once.
  - The diagonal edge $B-E$ remains untraversed.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_ham |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Special Paths: Hamiltonian Cycles

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Hamiltonian Cycle**: A cycle that visits **every single node** in the graph exactly once and returns to the origin.
- **Our Hamiltonian Cycle**: Let's trace it in red on the right:
  $$A \rightarrow B \rightarrow C \rightarrow D \rightarrow E \rightarrow A$$
  - This cycle forms the complete outer pentagonal ring.
  - Visits all 5 nodes exactly once and closes the loop!
  - **Degree Condition**: Since every node has degree $\ge 2$, a Hamiltonian cycle is successfully enabled.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_ham |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 4 & to == 5) | (from == 1 & to == 5) | (from == 5 & to == 1), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Walks, Trails, and Paths

::: {.incremental}
- **Walk**: Any sequence of alternating nodes and edges connecting $u$ to $v$.
  - *No restrictions*: Nodes and edges can be repeated as much as you want!
  - e.g., $\{B \rightarrow E \rightarrow C \rightarrow E \rightarrow D\}$ (traverses $E$ and edge $E-C$ twice).
- **Trail**: A walk that does not repeat any **edges**.
  - *Repeated nodes are allowed*, but every edge in the sequence must be unique.
  - e.g., $\{A \rightarrow C \rightarrow D \rightarrow E \rightarrow C \rightarrow B\}$ (repeats node $C$ but all edges are unique).
- **Path**: A walk that does not repeat any **nodes** (which means edges cannot be repeated either).
- **The Nested Hierarchy**: All paths are trails, and all trails are walks, but the reverse is not true:
  $$Paths \subset Trails \subset Walks$$
:::

---

## Walks: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Walk (B-E-C-E-D)**:
  - We start at node $B$, walk to $E$, then $C$, then *back* to $E$, and finally to $D$.
  - **Repeated Node**: Node $E$ is visited twice in this sequence.
  - **Repeated Edge**: Edge $E-C$ (and $C-E$) is traversed twice.
  - Highlighted in red on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Walk (Repeats Node E and Edge E-C)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
```
:::

::::

---

## Trails: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Trail (B-E-C-D-E-A)**:
  - We start at $B \rightarrow E \rightarrow C \rightarrow D \rightarrow E \rightarrow A$.
  - **Repeated Node**: Node $E$ is visited twice.
  - **Repeated Edges**: **None!** Every edge in the sequence is completely unique.
  - Highlighted in red on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Trail (Repeats Node E, Edges Unique)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
```
:::

::::

---

## Paths in Directed Graphs (Digraphs)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, paths must strictly **respect the direction of the arrows**!
- Let's trace directed paths from $A$ to $B$ in our digraph:
  - $P_1 = \{AC, CE, EB\} \quad (l = 3)$
  - $P_2 = \{AD, DE, EB\} \quad (l = 3)$
  - $P_3 = \{AC, CD, DE, EB\} \quad (l = 4)$
- But can we find any directed path from $B$ to $A$?
  - **No!** All arrows lead away from $A$ and toward $B$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Directed Reachability

::: {.incremental}
- **Asymmetric Reachability**:
  - Node $A$ can reach $B$ via directed path $\{AD, DE, EB\}$ ($l_{AB} = 3$).
  - But node $B$ **cannot** reach $A$ because we cannot travel against the arrows ($l_{BA} = \infty$).
- **Mutual Reachability**:
  - Two nodes $u$ and $v$ are mutually reachable if there exists a directed path from $u \rightarrow v$ AND from $v \rightarrow u$.
  - These two paths **do not** have to be the same length!
  - **Example**: Nodes $B$ and $C$ are mutually reachable:
    - $B$ reaches $C$ via $\{BE, ED, DC\} \quad (l = 3)$
    - $C$ reaches $B$ via shortest path $\{CE, EB\} \quad (l = 2)$
:::

---

## Directed Reachability: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "D") | (from_name == "D" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Asymmetric (A reaches B, B cannot reach A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p2 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Mutual (B and C reach each other via paths of length 3 & 2)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p1 + p2
```

---

## Directed Cycles & DAGs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **directed cycle** is a cycle where all arrows flow in a single, continuous direction.
  - **Example**: Cycle starting and ending at $C$:
    - $\{CE, ED, DC\} \quad (l = 3)$
- **Directed Acyclic Graphs (DAGs)**:
  - Directed graphs that contain **no directed cycles**.
  - e.g. strict organizational command structures, directed trees.
  - Anti-symmetric relations (like "is boss of") structurally forbid cycles to maintain consistent hierarchy.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Ignoring the Arrows: Semipaths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- What if we want to trace connections in a directed graph but **ignore the arrows**?
- This is called a **semipath** (or semicycle if it is a closed loop).
- A semipath treats directed arcs as undirected edges.
- **Example**: Semipath connecting $B$ and $A$:
  - $\{BE, EC, CA\}$ (highlighted in red on the right).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "C") | (from_name == "A" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 3: Graph Connectivity & Components

---

## Connected vs. Disconnected Graphs

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Connected Graph (Single Component)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5

    gr_l_conn <- tbl_graph(
        nodes = data.frame(name = as.character(1:9)),
        edges = data.frame(
            from = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 3, 2),
            to   = c(2, 3, 4, 5, 6, 7, 8, 9, 1, 5, 7, 8)
        ),
        directed = FALSE
    )
    
    ggraph(gr_l_conn, layout = "circle") +
        geom_edge_link(color = "steelblue", edge_width = 1.25) +
        geom_node_point(size = 15, color = "tan2") +
        geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
        theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph (Two Components)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5

    gr_l_disconn <- tbl_graph(
        nodes = data.frame(name = as.character(1:9)),
        edges = data.frame(
            from = c(1, 2, 3, 4, 5, 6, 1, 7, 8, 9),
            to   = c(2, 3, 4, 5, 6, 1, 4, 8, 9, 7)
        ),
        directed = FALSE
    )
    
    ggraph(gr_l_disconn, layout = "circle") +
        geom_edge_link(color = "steelblue", edge_width = 1.25) +
        geom_node_point(size = 15, color = "tan2") +
        geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
        theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

- **Connected**: A path of some finite length exists between *every* pair of nodes in the graph (left).
- **Disconnected**: At least one pair of nodes is unreachable; the graph is split into separate pieces (right).

---

## Components & The Giant Component

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph with 2 Components**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
gr_disconnected_colors <- gr_disconnected |>
  activate(edges) |>
  mutate(highlight = ifelse(from <= 6 & to <= 6, "giant", "small"))

ggraph(gr_disconnected_colors, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("giant" = "red", "small" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("giant" = 1.8, "small" = 0.8), guide = "none") +
  geom_node_point(aes(color = ifelse(name %in% c("G", "H", "I"), "small", "giant")), size = 18) +
  scale_color_manual(values = c("giant" = "tan2", "small" = "skyblue"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: flex-start;"}
- **Component**: A maximally connected subgraph within a larger disconnected graph.
  - The graph on the left contains exactly **two components**.
- **Giant Component**: The largest connected component in a disconnected graph.
  - The red component $\{A, B, C, D, E, F\}$ is the **giant component** (of order 6).
  - The blue component $\{G, H, I\}$ is a smaller component (of order 3).
:::

::::

# Part 4: Edge and Node Connectivity (Cutsets & Bridges)

---

## Edge Cutsets & Connectivity

::: {.incremental}
- We can disconnect a connected graph by removing a subset of its edges.
- **Edge Cutset**:
  - Any set of edges whose removal disconnects the graph into two or more components.
- **Minimum Edge Cutset**:
  - An edge cutset containing the minimum possible number of edges needed to disconnect the graph.
- **Edge-Connectivity** ($\lambda(G)$):
  - The number of edges in a minimum edge cutset.
- **Bridge**:
  - A single edge whose removal disconnects the graph (a minimum edge cutset of size 1, so $\lambda(G) = 1$).
:::

---

## Edge Cutsets: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- In our 5-node graph on the right:
  - Edges **A-C** and **B-C** are highlighted in red (Before).
- **Removing both edges**:
  - Disconnects $\{C, D, E\}$ from $\{A, B\}$ (After).
  - This is an edge cutset of size **2**.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_cutset <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

p_before <- ggraph(gr_cutset, layout = 'kk') +
  geom_edge_link(aes(color = ifelse((from == 1 & to == 3) | (from == 2 & to == 3), "yes", "no")), edge_width = 1.25) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_cutset_after <- gr_cutset |>
  activate(edges) |>
  filter(!((from == 1 & to == 3) | (from == 2 & to == 3)))

p_after <- ggraph(gr_cutset_after, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Bridges: A Cutset of Size 1

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **Bridge**:
  - A single edge whose removal disconnects the graph (so $\lambda(G) = 1$).
- In our graph on the right, the edge **C-D** is highlighted in red.
  - **Removing edge C-D** completely disconnects $\{A, B, C\}$ from $\{D, E\}$.
  - Bridges represent the most critical, yet vulnerable transmission links in any network.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_bridge <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

p_before <- ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = ifelse(from == 3 & to == 4, "yes", "no")), edge_width = 1.25) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Bridge)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_bridge_after <- gr_bridge |>
  activate(edges) |>
  filter(!(from == 3 & to == 4))

p_after <- ggraph(gr_bridge_after, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Bridge Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::
---

## Special Properties of Bridges

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In a connected graph, a **bridge** (such as the red **C-D** edge on the right) possesses two interesting mathematical properties ([Buckley and Harary 1990](https://doi.org/10.1007/978-3-642-75242-1)):
- **1. The No-Cycle Rule**:
  - A bridge **never lies on any cycle** of the graph.
  - *If it did*, removing it would still leave the rest of the cycle intact, failing to disconnect the graph.
- **2. The Intermediation Bottleneck**:
  - The bridge is an edge on **every path** connecting at least one pair of nodes (e.g., any node in $\{A, B, C\}$ to any node in $\{D, E\}$).
  - *Every packet of information flowing between these clusters must pass through this single bottleneck.*
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = ifelse(from == 3 & to == 4, "yes", "no")), edge_width = 1.5) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::


---

## Local Bridges

::: {.incremental}
- In a classic paper on "The Strength of Weak Ties" (1973), Mark Granovetter developed the concept of a **local bridge**:
  - An edge that, if removed, does *not* disconnect the entire graph, but significantly increases the shortest path length between its end nodes.
- **Sociological Power**:
  - Local bridges act as crucial "shortcuts" that facilitate efficient, cross-group communication.
  - They are the primary channels through which novel information diffuses between separate social clusters.
:::

---

## Local Bridge: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Let's look at the edge **C-D** highlighted in red:
  - It connects the left cluster $\{A, B, C\}$ to the right cluster $\{D, E, F\}$.
- **Is it a bridge?** No. If removed, the graph remains connected via the bottom path ($B-G-H-I-E$).
- **Is it a local bridge?** Yes! 
  - Observed distance: $d(C, D) = 1$.
  - If removed, the distance increases to:
    $$d(C, D) = 5 \quad (C-B-G-H-I-E-D)$$
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_bridge_local |>
  activate(edges) |>
  mutate(highlight = ifelse(from == 3 & to == 4, "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Node Cutsets & Connectivity

::: {.incremental}
- We can also disconnect a graph by **removing nodes** (vertices).
- Removing a node also removes all edges incident to it—nodes "take their links with them"!
- **Node Cutset (Vertex Cutset)**:
  - Any set of nodes whose removal disconnects the remaining graph into two or more components.
- **Minimum Node Cutset**:
  - A node cutset containing the minimum possible number of nodes needed to disconnect the graph.
- **Node-Connectivity** ($\kappa(G)$):
  - The number of nodes in a minimum node cutset.
- **Articulation Node (Cutpoint)**:
  - A single node whose removal disconnects the graph (so $\kappa(G) = 1$).
:::

---

## Articulation Nodes: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- In our house graph on the right, node **E** is highlighted in red.
- **Removing node E**:
  - Removes all of its incident edges: $B-E, E-C, E-D$.
  - Completely disconnects node $B$ from the rest of the network.
  - This is a node cutset of size **1** (Node $E$ is an **articulation node**).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "E", "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Node Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_node <- gr_undir |>
  filter(name != "E")

p_after <- ggraph(gr_undir_after_node, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Node Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Special Properties of Articulation Nodes

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- An **articulation node** (or cutpoint, such as **E** on the right) has unique properties mirroring those of bridges:
- **1. Node Equivalent of a Bridge**:
  - A single actor upon whom the entire connectivity of the network depends ($\kappa(G) = 1$).
- **2. The Path Intermediator**:
  - It stands as an **inner node** on *every single path* connecting some node $v$ to the rest of the network (e.g., node $E$ is on every path connecting $B$ to $\{A, C, D\}$).
- **3. The End-Node Rule**:
  - A node of degree $1$ (an **end-node**, like $B$) can *never* be an articulation node; removing it always leaves the remaining graph connected.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "E", "yes", "no")), size = 20) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Node Cutset: An Edge Cutset Comparison

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Let's look at node $A$ in our house graph on the right.
- To disconnect $A$ by removing edges, we must remove **2 edges**: $C-A$ and $D-A$ (red).
- But to disconnect $A$ by removing nodes, we only need to remove **1 node**: $E$ (the cutpoint that isolates $B$).
- This illustrates how node and edge connectivity can differ in the same graph.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 3 & to == 5) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 1.8, "no" = 1.0), guide = "none") +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Edge Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_edges <- gr_undir |>
  activate(edges) |>
  filter(!((from == 3 & to == 5) | (from == 4 & to == 5)))

p_after <- ggraph(gr_undir_after_edges, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Edges Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## k-Connected Graphs & Robustness

::: {.incremental}
- **$k$-Connected Graph**:
  - A graph is $k$-connected (or $k$-node connected) if the minimum number of nodes required to disconnect it is $k$.
  - **1-Connected**: Removing 1 node disconnects the graph (contains an articulation point).
  - **2-Connected**: Requires removing at least 2 nodes to disconnect it.
  - **3-Connected**: Requires removing at least 3 nodes to disconnect it.
- **Menger's Theorem (1927)**:
  - If a graph is $k$-connected, there are at least $k$ node-independent paths (paths that do not share any common inner nodes) connecting each non-adjacent pair of vertices.
- **Social Cohesion**: Higher $k$-connectivity indicates a more robust, cohesive, and redundant communication structure.
:::

---

## 1-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **1-Connected Graph**:
  - A graph where removing **exactly 1 node** can disconnect the network.
- **Before** (Left):
  - A 7-node symmetric bow-tie graph, with the central bridge node (node 7) highlighted in red.
- **After** (Right):
  - The central node 7 is removed, leaving two completely separate connected components (triangles).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_k1 <- tbl_graph(
  nodes = data.frame(
    name = as.character(1:7),
    x = c(-2, -2, -1, 1, 2, 2, 0),
    y = c(1, -1, 0, 0, 1, -1, 0)
  ),
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 3, 4),
    to   = c(2, 3, 1, 5, 6, 4, 7, 7)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k1, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "7", "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (7-Node Bow-Tie with Cutpoint 7)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_k1_after <- gr_k1 |> filter(name != "7")
p_after <- ggraph(gr_k1_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Node 7 Removed, Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## 2-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **2-Connected Graph**:
  - A graph where it requires removing at least **2 nodes** to disconnect it.
- **Before** (Left):
  - A 6-node cycle with a diagonal chord (edge 2-5), with nodes 2 and 5 highlighted in red.
- **After** (Right):
  - Removing nodes 2 and 5 leaves the graph split into two disconnected components: {1, 6} and {3, 4}.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_k2 <- tbl_graph(
  nodes = data.frame(name = as.character(1:6)),
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 2),
    to   = c(2, 3, 4, 5, 6, 1, 5)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k2, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(aes(color = ifelse(name %in% c("2", "5"), "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Before (6-Node Ring with Chord, Cutset {2, 5})") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

gr_k2_after <- gr_k2 |> filter(!name %in% c("2", "5"))
p_after <- ggraph(gr_k2_after, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("After (Nodes 2 & 5 Removed, Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## 3-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **3-Connected Graph**:
  - A graph where it requires removing at least **3 nodes** to disconnect it.
- **Before** (Left):
  - A 3-regular 6-node prism graph (two concentric triangles), with nodes 1, 2, and 6 highlighted in red.
- **After** (Right):
  - Removing nodes 1, 2, and 6 leaves the single edge 4-5 and completely isolates node 3 (disconnected).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
nodes_prism <- data.frame(
  name = as.character(1:6),
  x = c(0, -0.43, 0.43, 0, -1.04, 1.04),
  y = c(0.5, -0.25, -0.25, 1.2, -0.6, -0.6)
)
gr_k3 <- tbl_graph(
  nodes = nodes_prism,
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 1, 2, 3),
    to   = c(2, 3, 1, 5, 6, 4, 4, 5, 6)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k3, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.1) +
  geom_node_point(aes(color = ifelse(name %in% c("1", "2", "6"), "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Before (6-Node Prism, Cutset {1, 2, 6})") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

gr_k3_after <- gr_k3 |> filter(!name %in% c("1", "2", "6"))
p_after <- ggraph(gr_k3_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.1) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("After (Nodes Removed, Node 3 Isolated)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Graph Cutsets: Quantitative Cohesion

::: {.incremental}
- **Cohesion** can be directly translated to graph-theoretic cutsets:
  - **Cohesive Groups**: Have large edge-cutsets and node-cutsets. Many connections or individuals must be removed to disrupt or fragment the group.
  - **Fragmented Groups**: Have small cutsets (such as bridges or single cutpoints). They are vulnerable to rapid disruption if key facilitators are removed.
- Any network can be decomposed into hierarchical subgraphs with specific levels of $k$-connectivity.
:::

## The Connectivity Inequality

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- For any undirected graph $G$, node-connectivity ($\kappa$), edge-connectivity ($\lambda$), and minimum degree ($\delta$) obey a fundamental structural hierarchy:
  $$\kappa(G) \le \lambda(G) \le \delta(G)$$
- **A Strict Inequality Example ($1 < 2 < 3$)**:
  - In our 9-node "double-clique" graph on the right:
    - **Minimum Degree ($\delta = 3$)**: The least connected nodes ($3, 4, 7, 8$) have exactly 3 edges.
    - **Edge-Connectivity ($\lambda = 2$)**: Removing $\{1-9, 2-9\}$ separates the left clique from the waist. No single bridge exists.
    - **Node-Connectivity ($\kappa = 1$)**: Removing the central cutpoint (node 9) completely disconnects the graph.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_ineq <- tbl_graph(
  nodes = data.frame(
    name = as.character(1:9),
    x = c(-1, -1, -2, -2, 1, 1, 2, 2, 0),
    y = c(0.5, -0.5, 0.5, -0.5, 0.5, -0.5, 0.5, -0.5, 0)
  ),
  edges = data.frame(
    from = c(1, 1, 1, 2, 2, 3, 5, 5, 5, 6, 6, 7, 1, 2, 5, 6),
    to   = c(2, 3, 4, 3, 4, 4, 6, 7, 8, 7, 8, 8, 9, 9, 9, 9)
  ),
  directed = FALSE
)

ggraph(gr_ineq, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

# Part 5: Directed Connectivity & Components

---

## Flavors of Connectivity in Directed Graphs

::: {.incremental}
- Because of edge directionality, directed graphs have **three major levels** of connectivity:
1. **Strongly Connected**: Every pair of nodes is **mutually reachable** via directed paths.
   - **Recursively Connected** (Subtype): Strongly connected, AND the directed paths in both directions share the exact same set of intermediate nodes (sharing both vertices and edges).
2. **Unilaterally Connected**: Every pair is connected by **at least one directed path** (from $u \rightarrow v$ OR $v \rightarrow u$, but not necessarily both).
3. **Weakly Connected**: Every pair of nodes is connected by at least a **semipath** (connected if we ignore arrow directions).
:::

---

## Strongly Connected: A Visual Example

:::: {.columns align=center}

::: {.column width="45%"}
::: {.incremental}
- **Strong Connectivity**:
  - A directed graph is strongly connected if there is a directed path from **every node to every other node** (mutual reachability).
- **7-Node Example** (Right):
  - Consists of a complete spanning cycle ($A \rightarrow B \rightarrow C \rightarrow D \rightarrow E \rightarrow F \rightarrow G \rightarrow A$) plus chord arcs.
  - Because of the continuous cycle, any node can reach any other node, ensuring complete mutual reachability.
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_large_strong, layout = "circle") +
  geom_edge_parallel(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'),
                     color = "steelblue", edge_width = 1.1, sep = unit(3.5, 'mm')) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Unilaterally Connected: A Visual Example

:::: {.columns align=center}

::: {.column width="45%"}
::: {.incremental}
- **Unilateral Connectivity**:
  - A directed graph is unilaterally connected if for every pair of nodes, there is **at least one directed path** connecting them.
- **7-Node Example** (Right):
  - Same 7 nodes, but the feedback edge $G \rightarrow A$ and other backward chords are removed. Edges flow strictly forward ($A \rightarrow B \rightarrow C \dots \rightarrow G$).
  - Here, $A$ can reach $G$, but $G$ cannot reach $A$ (no mutual reachability).
  - However, for every single pair, at least one directed pathway exists.
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_large_uni, layout = "circle") +
  geom_edge_parallel(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'),
                     color = "steelblue", edge_width = 1.1, sep = unit(3.5, 'mm')) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Weakly Connected: A Visual Example

:::: {.columns align=center}

::: {.column width="45%"}
::: {.incremental}
- **Weak Connectivity**:
  - A directed graph that is connected only if we **ignore the direction of the arrowheads** (treating it like an undirected graph).
- **7-Node Example** (Right):
  - Uses our textbook's 7-node directed graph.
  - Nodes $E$ and $G$ are transmitter nodes (indegree 0) that cannot reach each other, nor can they be reached by anyone else.
  - Thus, $\{E, G\}$ is a mutually unreachable pair (violating unilateral connectivity).
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_large_weak, layout = "circle") +
  geom_edge_parallel(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'),
                     color = "steelblue", edge_width = 1.1, sep = unit(3.5, 'mm')) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Decomposing Directed Components

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- Like undirected graphs, directed graphs can be decomposed into:
- **Weakly Connected Components (WCCs)**:
  - Subgraphs connected when ignoring edge directions.
  - Yields **one WCC** containing all 7 nodes: $\{A, B, C, D, E, F, G\}$.
- **Strongly Connected Components (SCCs)**:
  - Subgraphs where every pair is mutually reachable.
  - Yields **four SCCs**:
    - $\{A, B, D, F\}$ (the core "rumor mill", highlighted in red).
    - Singletons: $\{C\}$, $\{E\}$, $\{G\}$.
:::
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_dir_7, layout = 'kk') +
  geom_edge_parallel(color = "steelblue", edge_width = 1.0,
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'), sep = unit(3, 'mm')) +
  geom_node_point(aes(color = ifelse(name %in% c("A", "B", "D", "F"), "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::



# Part 6: Tree Graphs & Forests

---

## What is a Tree Graph?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- If a graph is both **connected** *and* has no **cycles** (acyclic), then it is a **tree graph**.
- **The Cycle Restriction**:
  - A cycle is a path that begins and ends with the same node.
  - In a tree graph, you can never start with a node and trace a sequence of distinct edges that take you back to where you started!
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_prop <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(0, 1, 1, 2, 3), y = c(0, 0, 1, 0, 0))
edges_prop <- data.frame(from = c(1, 2, 2, 4), to = c(2, 3, 4, 5)) # A-B, B-C, B-D, D-E
gr_prop <- tbl_graph(nodes = nodes_prop, edges = edges_prop, directed = FALSE)

ggraph(gr_prop, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Properties of Tree Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Tree graphs possess several unique mathematical and structural properties:
  1. **The Single Path Property**:
     - *Every* pair of nodes in a tree graph is linked by exactly **one unique path**.
     - There are no redundant or alternative paths between any two actors.
  2. **The One-Cycle Property**:
     - Adding a *single* new edge connecting any pair of non-adjacent nodes always creates **exactly one cycle**, destroying its tree status.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
edges_prop_cycle <- data.frame(from = c(1, 2, 2, 4, 3), to = c(2, 3, 4, 5, 5)) # A-B, B-C, B-D, D-E, C-E (added)
gr_prop_cycle <- tbl_graph(nodes = nodes_prop, edges = edges_prop_cycle, directed = FALSE)

gr_prop_cycle %>%
  activate(edges) %>%
  mutate(type = c("tree", "tree", "tree", "tree", "added")) %>%
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = type, linetype = type, edge_width = type)) +
  scale_edge_color_manual(values = c("tree" = "steelblue", "added" = "red"), guide = "none") +
  scale_edge_linetype_manual(values = c("tree" = "solid", "added" = "dashed"), guide = "none") +
  scale_edge_width_manual(values = c("tree" = 1.2, "added" = 1.5), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Tree Graph Metrics: Vertices & Edges

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In any tree graph, the number of edges ($m$) is always exactly equal to the number of nodes ($n$) minus one:
  $$m = n - 1 \qquad \iff \qquad n = m + 1$$
- **Sum of Degrees Identity**:
  - Recall that the sum of degrees in any undirected graph is $2m$. 
  - Substituting $m = n - 1$, the sum of degrees of a tree graph is always:
    $$\sum_i k_i = 2(n - 1) = 2n - 2$$
  - *In a tree graph, the sum of degrees is always exactly twice the number of nodes minus two!*
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_metric <- data.frame(
  name = c("A", "B", "C", "D", "E"), 
  deg = c("k=1", "k=3", "k=1", "k=2", "k=1"),
  x = c(0, 1, 1, 2, 3), 
  y = c(0, 0, 1, 0, 0)
)
gr_metric <- tbl_graph(nodes = nodes_metric, edges = edges_prop, directed = FALSE)

ggraph(gr_metric, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  geom_node_text(aes(label = deg), size = 5, fontface = "italic", color = "black", vjust = -1.8) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Tree Graph Density

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Because of the strict $m = n - 1$ edge constraint, tree graph density simplifies beautifully.
- **Density of a Tree Graph**:
  $$d(T) = \frac{2m}{n(n-1)} = \frac{2(n-1)}{n(n-1)} = \frac{2}{n}$$
- *For our 5-node tree graph on the right ($n = 5$)*:
  - The density is:
    $$d(T) = \frac{2}{5} = 0.40 \quad (40\%)$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_metric, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Tree Graph Average Degree

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Average degree in a tree graph also simplifies beautifully due to the edge constraint.
- **Average Degree of a Tree Graph**:
  $$\bar{k} = \frac{\sum k_i}{n} = \frac{2m}{n} = \frac{2n - 2}{n}$$
- *For our 5-node tree graph on the right ($n = 5$)*:
  - The average degree is:
    $$\bar{k} = \frac{2(5) - 2}{5} = \frac{8}{5} = 1.6$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_metric, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

## Trees as Minimally Connected Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Trees are **minimally connected** graphs:
  - They are connected using the *smallest possible number of edges* needed to connect $n$ vertices.
- **The Bridges Rule**:
  - Removing even a *single* edge from a tree graph completely disconnects the graph.
  - Thus, **every edge of a tree graph is a bridge** (a minimum edge cutset of size 1)!
- Disconnecting a tree graph by removing any edge splits it into separate connected components, which are themselves trees.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_bridge_vis <- gr_prop %>%
  activate(edges) %>%
  mutate(highlight = ifelse(from == 2 & to == 4, "bridge", "normal"))

ggraph(gr_bridge_vis, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight, linetype = highlight)) +
  scale_edge_color_manual(values = c("bridge" = "red", "normal" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("bridge" = 2.0, "normal" = 1.2), guide = "none") +
  scale_edge_linetype_manual(values = c("bridge" = "dashed", "normal" = "solid"), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white", vjust = 0.3) +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Numerical Example: 8-Node Tree

:::: {.columns align=center}

::: {.column width="55%"}
Let's compute the metrics for the 8-node tree shown on the right:

- **1. Nodes and Edges**:
  - Number of nodes: $n = 8$
  - Number of edges: $m = n - 1 = 7$
- **2. Degree of each node**:
  - $k_A = 2, k_B = 3, k_C = 3, k_D = 1, k_E = 2, k_F = 1, k_G = 1, k_H = 1$
  - Sum of degrees: $\sum k_i = 14$
  - Check: $2(n-1) = 2(7) = 14 \quad \checkmark$
- **3. Network Density**:
  - $d(T) = \frac{2}{n} = \frac{2}{8} = 0.25 \quad (25\%)$
- **4. Average Degree**:
  - $\bar{k} = \frac{2n - 2}{n} = \frac{14}{8} = 1.75$
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_8 <- data.frame(
  name = LETTERS[1:8],
  x = c(0, -1, 1, -2, -0.5, 0.5, 2, -0.5),
  y = c(1, 0, 0, -1, -1, -1, -1, -2)
)
edges_8 <- data.frame(
  from = c(1, 1, 2, 2, 3, 3, 5),
  to   = c(2, 3, 4, 5, 6, 7, 8)
)
gr_8 <- tbl_graph(nodes = nodes_8, edges = edges_8, directed = FALSE)

ggraph(gr_8, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Forests: Visualizing Disconnection

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **Forest**:
  - A disconnected graph whose components are all tree graphs.
- **Before** (Left):
  - A connected 9-node tree graph.
- **After** (Right):
  - Removing edge **D-H** splits the graph into two separate trees (a **forest**):
    - A **bigger tree** of 6 nodes on the left ($\{A, B, C, D, E, F\}$).
    - A **smaller tree** of 3 nodes on the right ($\{G, H, I\}$).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
nodes_f <- data.frame(
  name = LETTERS[1:9],
  x = c(0, 1, 0, 2, 3, 3, 5, 4, 5),
  y = c(0.5, 0, -0.5, 0, 0.5, -0.5, 0.5, 0, -0.5)
)
edges_f <- data.frame(
  from = c(2, 2, 2, 4, 4, 4, 8, 8),
  to   = c(1, 3, 4, 5, 6, 8, 7, 9)
)
gr_f <- tbl_graph(nodes = nodes_f, edges = edges_f, directed = FALSE)

p_before <- ggraph(gr_f, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = ifelse(from == 4 & to == 8, "yes", "no")), edge_width = 1.25) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (Connected Tree)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

gr_f_after <- gr_f %>% 
  activate(edges) %>% 
  filter(!(from == 4 & to == 8))

p_after <- ggraph(gr_f_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Forest: 2 Disconnected Trees)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

## Special Types of Trees: Line, Star & Caterpillar Graphs

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- **The Line Graph ($L_n$)**:
  - A single straight path connecting $n$ nodes.
  - Used to model sequential diffusion processes (like the "Telephone Game").
- **The Star Graph ($S_n$)**:
  - A central hub connected to $n-1$ leaf nodes.
  - Represents extreme structural centralization.
- **The Caterpillar Graph**:
  - A central path ("body") with pendant leaf nodes ("legs") sticking out of it.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 4.5
gr_line <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]), edges = data.frame(from = 1:4, to = 2:5), directed = FALSE)
p_line <- ggraph(gr_line, layout = "linear") +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Line Graph (L5)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_star <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]), edges = data.frame(from = c(1, 1, 1, 1), to = 2:5), directed = FALSE)
p_star <- ggraph(gr_star, layout = "star") +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Star Graph (S5)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

fr <- c("A", "A", "C", "C", "E", "E", "G")
to <- c("B", "C", "D", "E", "F", "G", "H")
edge.dat <- data.frame(fr, to)
node.dat <- data.frame(name = union(fr, to))
gr_cat <- tbl_graph(edges = edge.dat, nodes = node.dat, directed = FALSE)
l_cat <- ggraph(gr_cat, layout = 'grid')
l_cat_mat <- as.matrix(l_cat$data[, 1:2])
l_cat_mat[,"x"] <- c(0, 1, 2, 3, 0, 1, 2, 3)
l_cat_mat[,"y"] <- c(1, 1, 1, 1, 0, 0, 0, 0)

p_cat <- ggraph(gr_cat, layout = l_cat_mat) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Caterpillar Graph (Order 8)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_line + p_star + p_cat
```
:::

::::

---

## The Graph Efficiency

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Concept** (Krackhardt, 1994):
  - Measures how **economical** the connectivity of a graph is.
  - A graph is efficiently connected if it uses the *minimum* possible number of edges to remain connected.
- **Formula**:
  $$F(G) = 1 - \left[\frac{2(m - n + 1)}{(n-1)(n-2)}
ight]$$
  - Where $m$ is the number of edges and $n$ is the number of nodes.
- **Numerical Proof (The Tree on the Right)**:
  - Pictured on the right is a tree of $n = 5$ nodes and $m = 4$ edges.
  - Substituting $n = 5$ and $m = 4$ into the formula:
    $$F(G) = 1 - \left[\frac{2(4 - 5 + 1)}{(5-1)(5-2)}
ight]$$
    $$F(G) = 1 - \left[\frac{2(0)}{4 \times 3}
ight] = 1 - 0 = 1.0$$
- **Beyond Trees (Decreasing Efficiency)**:
  - As edges are added beyond $m = n - 1$, the numerator $(m - n + 1) > 0$, causing $F(G)$ to decrease toward $0.0$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
nodes_tree <- data.frame(name = LETTERS[1:5])
edges_tree <- data.frame(from = c(1, 2, 2, 3), to = c(2, 3, 4, 5))
gr_tree_eff <- tbl_graph(nodes = nodes_tree, edges = edges_tree, directed = FALSE)
ggraph(gr_tree_eff, layout = 'nicely') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Maximally Efficient (Tree: F(G) = 1.0)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::

---

## Detailed Example: Graph 1 (Less Efficient)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph 1 Structure**:
  - Contains $n = 5$ nodes and $m = 7$ edges.
- **Efficiency Calculation**:
  - We substitute $n=5$ and $m=7$ into the formula:
    $$F(G) = 1 - \left[\frac{2(7 - 5 + 1)}{(5-1)(5-2)}
ight]$$
    $$F(G) = 1 - \left[\frac{2(3)}{4 \times 3}
ight] = 1 - \frac{6}{12} = 0.50$$
- **Interpretation**:
  - With 5 nodes, a connected tree needs only 4 edges. 
  - This graph has 3 redundant edges ($7 - 5 + 1 = 3$), resulting in a low efficiency score of **$0.50$**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
set.seed(123)
gr1 <- connected_gnm(5, 7) %>% mutate(name = LETTERS[1:5])
ggraph(gr1, layout = 'nicely') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Graph 1 (n = 5, m = 7, F = 0.50)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::

---

## Detailed Example: Graph 2 (More Efficient)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph 2 Structure**:
  - Contains $n = 7$ nodes and $m = 10$ edges.
- **Efficiency Calculation**:
  - We substitute $n=7$ and $m=10$ into the formula:
    $$F(G) = 1 - \left[\frac{2(10 - 7 + 1)}{(7-1)(7-2)}
ight]$$
    $$F(G) = 1 - \left[\frac{2(4)}{6 \times 5}
ight] = 1 - \frac{8}{30} \approx 0.73$$
- **Conclusion**:
  - Despite having more edges ($10$ vs $7$), Graph 2 is larger. Its redundancy is distributed over more nodes, yielding a higher efficiency of **$0.73$**.
  - **Graph 2 is more efficient than Graph 1 ($0.73 > 0.50$)**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
set.seed(123)
gr2 <- connected_gnm(7, 10) %>% mutate(name = LETTERS[1:7])
ggraph(gr2, layout = 'nicely') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Graph 2 (n = 7, m = 10, F = 0.73)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::

---

## Graph Efficiency: Complete Graphs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Maximally Inefficient**:
  - The complete graph ($K_n$) is minimally efficient because it has the maximum possible number of edges.
  - Number of edges: $m = \frac{n(n-1)}{2}$
- **Example ($n=5$ Complete Graph)**:
  - Edges: $m = \frac{5 \times 4}{2} = 10$
  - Calculation:
    $$F(G) = 1 - \left[\frac{2(10 - 5 + 1)}{(5-1)(5-2)}\right]$$
    $$= 1 - \left[\frac{2(6)}{4 \times 3}\right] = 1 - \frac{12}{12} = 0.0$$
  - Thus, any complete graph yields $F(G) = 0.0$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_complete <- play_gnm(n = 5, m = 10, directed = FALSE, loops = FALSE) %>%
  mutate(name = LETTERS[1:5])
ggraph(gr_complete, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() + coord_cartesian(clip = "off") +
  ggtitle("Complete Graph (F(G) = 0.0)") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```
:::

::::
