---
title: "Undirected Graph Metrics"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(ggplot2)

# Base graph and layout for neighborhood slides
gr_base <- tbl_graph(
  nodes = data.frame(name = 1:7),
  edges = data.frame(
    from = c(1, 4, 4, 2, 5, 6, 3),
    to   = c(3, 3, 7, 5, 6, 2, 2)
  ),
  directed = FALSE
)

set.seed(42)
layout_fr <- create_layout(gr_base, layout = "fr")
```

# Node Neighborhoods

---

## Node Neighborhoods

::: {.incremental}
- **Neighborhood** ($\mathcal{N}(v)$):
  - The set of nodes that are directly connected (adjacent) to node $v$.
  - These are a node's immediate **neighbors**.
  - Isolates have an empty neighborhood set ($\emptyset$).
:::

---

## Node Neighborhoods: Example

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Let's look at the graph on the right:
  - **Focal Node**: **4** (in tomato)
  - **Neighborhood**: $\mathcal{N}(4) = \{3, 7\}$ (in gold)
- Another example:
  - $\mathcal{N}(1) = \{3\}$
:::
:::

::: {.column width="45%"}
```{r}
#| label: plt-node-neighborhood-intro
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_base |>
  activate(nodes) |>
  mutate(
    highlight = case_when(
      name == 4 ~ "focal",
      name %in% c(3, 7) ~ "neighbor",
      TRUE ~ "other"
    )
  ) |>
  ggraph(layout = "manual", x = layout_fr$x, y = layout_fr$y) +
    geom_edge_link(color = "gray80", width = 1.5) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), color = "black", size = 6) +
    scale_color_manual(
      values = c("focal" = "tomato", "neighbor" = "gold", "other" = "skyblue"),
      guide = "none"
    ) +
    theme_graph() +
    coord_cartesian(clip = "off")
```
:::

::::

---

## Neighborhood Intersection

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Neighborhood Intersection** ($\mathcal{N}(u) \cap \mathcal{N}(v)$) is the set of **common neighbors** connected to both $u$ and $v$.
- In our graph:
  - $N(1) = \{3\}$
  - $N(4) = \{3, 7\}$
- The intersection is the single node they share:
  - $N(1) \cap N(4) = \{3\}$
:::
:::

::: {.column width="40%"}
```{r}
#| label: plt-neighborhood-intersection
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_base |>
  activate(nodes) |>
  mutate(
    highlight = case_when(
      name %in% c(1, 4) ~ "focal",
      name == 3 ~ "common",
      TRUE ~ "other"
    )
  ) |>
  ggraph(layout = "manual", x = layout_fr$x, y = layout_fr$y) +
    geom_edge_link(color = "gray80", width = 1.5) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), color = "black", size = 6) +
    scale_color_manual(
      values = c("focal" = "tomato", "common" = "gold", "other" = "skyblue"),
      name = "", labels = c("Common Neighbor", "Focal Nodes", "Other")
    ) +
    theme_graph() +
    theme(legend.position = "bottom") +
    coord_cartesian(clip = "off")
```
:::

::::

---

## Neighborhood Intersection: Example 2

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Let's find the neighborhood intersection of nodes **5** and **3**.
- First, list their individual neighborhood sets:
  - $\mathcal{N}(5) = \{2, 6\}$
  - $\mathcal{N}(3) = \{1, 2, 4\}$
- Nodes **5** and **3** (in tomato) are not directly connected, but they share a common neighbor: node **2** (in gold).
- Therefore, their neighborhood intersection is:
  $$\mathcal{N}(5) \cap \mathcal{N}(3) = \{2\}$$
:::
:::

::: {.column width="45%"}
```{r}
#| label: plt-neighborhood-intersection-2
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_base |>
  activate(nodes) |>
  mutate(
    highlight = case_when(
      name %in% c(5, 3) ~ "focal",
      name == 2 ~ "common",
      TRUE ~ "other"
    )
  ) |>
  ggraph(layout = "manual", x = layout_fr$x, y = layout_fr$y) +
    geom_edge_link(color = "gray80", width = 1.5) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), color = "black", size = 6) +
    scale_color_manual(
      values = c("focal" = "tomato", "common" = "gold", "other" = "skyblue"),
      name = "", labels = c("Common Neighbor", "Focal Nodes", "Other")
    ) +
    theme_graph() +
    theme(legend.position = "bottom") +
    coord_cartesian(clip = "off")
```
:::

::::

---

## Neighborhood Union

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Neighborhood Union** ($\mathcal{N}(u) \cup \mathcal{N}(v)$) is the set of all nodes connected to *either* $u$ or $v$.
- In our graph:
  - $N(1) = \{3\}$
  - $N(4) = \{3, 7\}$
- The union contains all unique neighbors from both sets:
  - $N(1) \cup N(4) = \{3, 7\}$
:::
:::

::: {.column width="40%"}
```{r}
#| label: plt-neighborhood-union
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_base |>
  activate(nodes) |>
  mutate(
    highlight = case_when(
      name %in% c(1, 4) ~ "focal",
      name %in% c(3, 7) ~ "neighbor",
      TRUE ~ "other"
    )
  ) |>
  ggraph(layout = "manual", x = layout_fr$x, y = layout_fr$y) +
    geom_edge_link(color = "gray80", width = 1.5) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), color = "black", size = 6) +
    scale_color_manual(
      values = c("focal" = "tomato", "neighbor" = "gold", "other" = "skyblue"),
      name = "", labels = c("Focal Nodes", "Union of Neighbors", "Other")
    ) +
    theme_graph() +
    theme(legend.position = "bottom") +
    coord_cartesian(clip = "off")
```
:::

::::

---

## Neighborhood Union: Example 2

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Let's find the neighborhood union of nodes **5** and **3**.
- First, list their individual neighborhood sets:
  - $\mathcal{N}(5) = \{2, 6\}$
  - $\mathcal{N}(3) = \{1, 2, 4\}$
- Focal nodes **5** and **3** are shown in tomato; the union of their neighbors (nodes **1, 2, 4, 6**) is shown in gold.
- Therefore, the union of neighbors is:
  $$\mathcal{N}(5) \cup \mathcal{N}(3) = \{1, 2, 4, 6\}$$
:::
:::

::: {.column width="45%"}
```{r}
#| label: plt-neighborhood-union-2
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_base |>
  activate(nodes) |>
  mutate(
    highlight = case_when(
      name %in% c(5, 3) ~ "focal",
      name %in% c(1, 2, 4, 6) ~ "neighbor",
      TRUE ~ "other"
    )
  ) |>
  ggraph(layout = "manual", x = layout_fr$x, y = layout_fr$y) +
    geom_edge_link(color = "gray80", width = 1.5) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), color = "black", size = 6) +
    scale_color_manual(
      values = c("focal" = "tomato", "neighbor" = "gold", "other" = "skyblue"),
      name = "", labels = c("Focal Nodes", "Union of Neighbors", "Other")
    ) +
    theme_graph() +
    theme(legend.position = "bottom") +
    coord_cartesian(clip = "off")
```
:::

::::

---

# Node Degree

---

## Node Degree in Undirected Graphs

::: {.incremental}
- In an undirected graph, a node's **degree** ($k_i$) has two equivalent definitions:
  1. **Neighborhood Cardinality**: The number of neighbors of node $i$ ($k_i = |\mathcal{N}(i)|$).
  2. **Incident Edges Count**: The number of edges connected to node $i$.
- Sociological meaning: **Degree Centrality** (reflects popularity, influence, or activity).
:::

---

## Node Degree: An Example

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Let's calculate the degree for Node **A** (in tomato) on the right:
  - Neighbors of **A**: $\mathcal{N}(A) = \{B, C, D, F\}$
  - Neighborhood size: $k_A = |\mathcal{N}(A)| = 4$
  - Incident edges: $AB$, $AC$, $AD$, and $AF$ ($k_A = 4$)
- Thus, the degree of Node **A** is **4**.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_deg_undir <- tbl_graph(
  nodes = data.frame(name = c("A", "B", "C", "D", "F")),
  edges = data.frame(from = c(1, 1, 1, 1, 2), to = c(2, 3, 4, 5, 4)),
  directed = FALSE
)

ggraph(gr_deg_undir, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(aes(color = (name == "A")), size = 22, show.legend = FALSE) +
  geom_node_text(aes(label = name), color = "white", size = 10, fontface = "bold") +
  scale_color_manual(values = c("TRUE" = "tomato", "FALSE" = "tan2")) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Endpoints in Undirected Graphs

::: {.incremental}
- **Endpoints** (or leaves/periphery):
  - Nodes that have a degree equal to **one** ($k_i = 1$).
  - They have just a single neighbor in the graph.
:::

---

## Endpoints: An Example

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Let's identify the endpoints in the graph on the right:
  - Nodes **C** and **F** (in tomato) are connected only to node **A**.
  - Since $k_C = 1$ and $k_F = 1$, they have degree one.
  - Therefore, **C** and **F** are the endpoints of this graph.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_deg_undir, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(aes(color = (name %in% c("C", "F"))), size = 22, show.legend = FALSE) +
  geom_node_text(aes(label = name), color = "white", size = 10, fontface = "bold") +
  scale_color_manual(values = c("TRUE" = "tomato", "FALSE" = "tan2")) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

# Basic Graph Metrics

---

## Graph Order and Size

::: {.incremental}
- **Graph Order** ($n$):
  - The number of nodes in the graph ($n = |V|$).
  - Cardinality of the vertex set.
- **Graph Size** ($m$):
  - The number of edges in the graph ($m = |E|$).
  - Cardinality of the edge set.
:::

---

## Order and Size: Undirected Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at the undirected graph on the right:
  - **Node Set** ($V$): $\{A, B, C, D, E, F, G\}$
  - **Edge Set** ($E$): $\{AC, CD, DG,$ <br> $BE, EF, FB, CB\}$
  - **Graph Order** ($n = |V|$): **7** nodes
  - **Graph Size** ($m = |E|$): **7** edges
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_letters <- gr_base |>
  activate(nodes) |>
  mutate(name = LETTERS[as.integer(name)])

ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Maximum Size

::: {.incremental}
- The **maximum size** of a simple graph of order $n$ is the number of edges in a **complete graph** (where every node is connected to every other node).
- Undirected: $Max(E) = \frac{n(n - 1)}{2}$
- Directed: $Max(E) = n(n - 1)$
:::

---

## Maximum Size: Undirected Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at the undirected graph on the right:
  - **Order** ($n$): 7 nodes
- **Formula**:
  $$Max(E) = \frac{n(n - 1)}{2}$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_letters <- gr_base |>
  activate(nodes) |>
  mutate(name = LETTERS[as.integer(name)])

ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Maximum Size: Undirected Calculation

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Order** ($n$): 7 nodes (shown on the right)
- **Maximum possible size** calculation:
  $$Max(E) = \frac{7 \times (7 - 1)}{2} = \frac{42}{2} = 21\text{ edges}$$
- **Interpretation**:
  - A complete undirected graph with 7 nodes would have exactly **21** edges.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Density: Formulas (Undirected)

::: {.incremental}
- **Density** ($d$) is the ratio of the observed number of edges ($m$) to the maximum possible number of edges.
- It is a value between 0 (completely empty) and 1 (complete graph).
- **Undirected Graph**:
  $$d = \frac{2m}{n(n-1)}$$
:::

---

## Graph Density: Interpretation

::: {.incremental}
- **Connection Probability**:
  - Density is the probability that any two randomly chosen nodes are connected.
- **Network Resilience**:
  - Denser networks are generally more resilient and less vulnerable to breakdowns when key nodes are removed.
:::

---

## Density: Undirected Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at the undirected graph on the right:
  - **Order** ($n$): 7 nodes
  - **Size** ($m$): 7 edges
  - **Maximum possible edges**:
    $$Max(E) = \frac{7 \times 6}{2} = 21$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Density: Undirected Calculation

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Observed edges** ($m$): 7 (shown on the right)
- **Maximum possible edges**: 21
- **Density Formula**:
  $$d = \frac{m}{Max(E)} = \frac{7}{21} \approx 0.333$$
- **Interpretation**:
  - Exactly **33.3%** of all possible undirected ties are present in this network.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

# Degree-Based Graph Metrics

---

## Graph Degree Set and Sequence

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph Degree Set**: A list of the degrees of every node in the graph.
- **Graph Degree Sequence**: The degree set sorted in descending order.
- This provides a "fingerprint" of the graph's overall connectivity structure.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_letters <- gr_base |>
  activate(nodes) |>
  mutate(name = LETTERS[as.integer(name)])

ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Degree Set & Sequence: Undirected Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Degrees**: $k_A=1, k_B=3, k_C=3, k_D=2, k_E=2, k_F=2, k_G=1$ (shown on the right)
- **Degree Set** (unordered list of node degrees):
  $$\{1, 3, 3, 2, 2, 2, 1\}$$
- **Degree Sequence** (sorted degrees):
  $$(3, 3, 2, 2, 2, 1, 1)$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Degree Range & Sum

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Maximum Degree** ($k_{max}$) and **Minimum Degree** ($k_{min}$).
- **Degree Range** ($k_r$): $k_{max} - k_{min}$.
- **Sum of Degrees** ($\sum k_i$):
  - In an undirected graph, this is always twice the number of edges: $\sum k_i = 2m$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Degree Range & Sum: Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Degrees**: $k_A=1, k_B=3, k_C=3, k_D=2, k_E=2, k_F=2, k_G=1$ (shown on the right)
- **Max & Min Degree**:
  - $k_{max} = 3$ (nodes B and C)
  - $k_{min} = 1$ (nodes A and G)
- **Degree Range**:
  - $k_r = 3 - 1 = 2$
- **Sum of Degrees**:
  - $\sum k_i = 1+3+3+2+2+2+1 = 14 = 2m$ (Matches!)
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Graphical vs. Non-Graphical Sequences

::: {.incremental}
- A sequence of non-negative integers is **graphical** if there exists at least one simple, undirected graph (no loops, no multiple edges) that has it as its degree sequence.
- Otherwise, the sequence is **non-graphical**—no such simple graph can possibly exist in the real world.
- Why can't certain sequences exist? 
  - They violate core mathematical rules of graph structure!
:::

---

## Two Quick Tests for Graphicality

Before using complex algorithms, we can apply two fast screening tests:

::: {.incremental}
1. **The Even Sum Rule (Handshake Lemma)**:
   - The sum of degrees must be **even**: $\sum k_i = 2m$.
   - *If the sum of a sequence is odd, it is automatically non-graphical!*
2. **The Maximum Degree Bound**:
   - In a simple graph of order $n$, no node can have a degree larger than $n-1$.
   - *If any degree in the sequence exceeds $n-1$, it is automatically non-graphical!*
:::

---

## Graphicality: Concrete Examples

::: {.incremental}
- **Graphical Sequence**: $\mathbf{d} = (4, 3, 3, 2, 2, 1, 1)$ with $n = 7$
  - Sum is $16$ (even). Max degree is $4 \le 6$.
  - A real simple graph can easily be constructed with these degrees.
- **Non-Graphical Sequence (Odd Sum Violation)**: $\mathbf{d} = (4, 3, 3, 3, 2, 1, 1)$ with $n = 7$
  - Sum is $17$ (odd). Violates Handshake Lemma. 
  - *No simple graph can ever have this sequence.*
- **Non-Graphical Sequence (Degree Bound Violation)**: $\mathbf{d} = (7, 2, 2, 1)$ with $n = 4$
  - Max degree is $7$, which is greater than $n-1 = 3$.
  - *A node in a 4-person group cannot have 7 unique friends!*
:::

## Average Degree

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Average Degree** ($\bar{k}$): The average of all node degrees in the graph.
  
  $$\bar{k} = \frac{1}{n} \sum_{i=1}^{n} k_i = \frac{2m}{n}$$
  
- It represents the expected degree of a randomly chosen node.
- It is also directly related to density, which we will derive on the next slides!
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Connecting Density & Average Degree

::: {.incremental}
- There is an **intimate mathematical relationship** between density ($d$) and average degree ($\bar{k}$).
- Recall the undirected density formula:
  $$d = \frac{2m}{n(n-1)}$$
- We can rewrite this by pulling out the average degree term ($\bar{k} = \frac{2m}{n}$):
  $$d = \left(\frac{2m}{n}\right) \times \frac{1}{n-1} = \bar{k} \times \frac{1}{n-1} = \frac{\bar{k}}{n-1} \qquad \iff \qquad \bar{k} = d(n-1)$$
:::

---

## Interpreting the Density-Degree Connection

::: {.incremental}
- **Density is Average Degree Scaled by Group Size**:
  - The density of an undirected graph is simply its average degree divided by the maximum possible connections of any single node ($n-1$).
- **For Example**:
  - In a group of $n = 10$ people, each person can have at most $n-1 = 9$ connections.
  - If the average number of friends ($\bar{k}$) is $4.5$, then the density of friendship is:
    $$d = \frac{4.5}{9} = 0.50 \quad (50\%)$$
- **Directed Case**:
  - The exact same identity holds! Since $d^d = \frac{m}{n(n-1)}$ and $\bar{k}$ is $\frac{m}{n}$, we get:
    $$d^d = \frac{\bar{k}}{n-1}$$
:::

---

## Degree Variance

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Degree Variance** ($\sigma^2_k$): A measure of how much the node degrees vary from the average degree.
  
  $$\sigma^2_k = \frac{1}{n} \sum_{i=1}^{n} (k_i - \bar{k})^2$$
  
- A low variance means less inequality in connectivity (more homogeneous), while a high variance indicates more inequality (more heterogeneous). 
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Degree Metrics: Basic Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Node Degrees**: $k_A=1, k_B=3, k_C=3, k_D=2, k_E=2, k_F=2, k_G=1$ (shown on the right)
- **Degree Set**: $\{1, 3, 3, 2, 2, 2, 1\}$ (degrees of all nodes in the graph)
- **Degree Sequence**: $(3, 3, 2, 2, 2, 1, 1)$ (ordered from highest to lowest)
- **Degree Sum**: 
  $$\sum k_i = 1+3+3+2+2+2+1 = 14 = 2m$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Degree Metrics: Average Degree

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Degrees**: $k_A=1, k_B=3, k_C=3, k_D=2, k_E=2, k_F=2, k_G=1$ (shown on the right)
- **Order** ($n$): 7 nodes
- **Average Degree** ($\bar{k}$):
  $$\bar{k} = \frac{\sum k_i}{n} = \frac{14}{7} = 2.0$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::


---

## Degree Metrics: Degree Variance

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Degrees**: $k_A=1, k_B=3, k_C=3, k_D=2, k_E=2, k_F=2, k_G=1$ (shown on the right)
- **Degree Variance** ($\sigma^2_k$):
  $$\sigma^2_k = \frac{\sum (k_i - \bar{k})^2}{n}$$
- **Variance Calculation**:
  $$\sigma^2_k = \frac{3 \times (2-2)^2 + 2 \times (3-2)^2 + 2 \times (1-2)^2}{7} = \frac{4}{7} \approx 0.571$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::


---

## Regular Graphs

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- A **regular graph** is one where every node has the same degree $k$ (called **$k$-regular**).
- Since all node degrees are identical to the average, the degree variance is zero ($\sigma^2_k = 0$).
- These are rare in real-world social networks but are used as baseline models.
- Example: A **3-regular** graph on 6 nodes (shown on the right).
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
gr_regular <- tbl_graph(
  nodes = data.frame(name = LETTERS[1:6]),
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 1, 2, 3),
    to   = c(2, 3, 4, 5, 6, 1, 4, 5, 6)
  ),
  directed = FALSE
)

ggraph(gr_regular, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, color = "white", fontface = "bold") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Regular Graphs: Metrics Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Node Degrees**: $k_A=3, k_B=3, k_C=3, k_D=3, k_E=3, k_F=3$ (shown on the right)
- **Order** ($n$): 6 nodes
- **Average Degree** ($\bar{k}$):
  $$\bar{k} = \frac{\sum k_i}{n} = \frac{18}{6} = 3.0$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
ggraph(gr_regular, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, color = "white", fontface = "bold") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Regular Graphs: Variance Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Average Degree** ($\bar{k}$): 3.0 (shown on the right)
- **Degree Variance** ($\sigma^2_k$):
  $$\sigma^2_k = \frac{6 \times (3 - 3)^2}{6} = 0$$
- **Conclusion**:
  - Since every node's degree matches the average degree, the deviation is zero for all nodes, resulting in **zero variance**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
ggraph(gr_regular, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, color = "white", fontface = "bold") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::
---

## Average Nearest Neighbor Degree

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Average Nearest Neighbor Degree** ($\bar{k}_{nn(i)}$) measures the average degree of node $i$'s neighbors.
- Formula:
  $$\bar{k}_{nn(i)} = \frac{1}{k_i} \sum_{j \in \mathcal{N}(i)} k_j$$
- It tells us whether a node's connections are, on average, highly connected or peripheral.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_letters, layout = "manual", x = layout_fr$x, y = layout_fr$y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 22, color = "skyblue") +
  geom_node_text(aes(label = name), color = "black", size = 6) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_graph() +
  theme(plot.margin = margin(10, 10, 10, 10, "pt")) +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Feld's Friendship Paradox

::: {.incremental}
- **Feld's Friendship Paradox** (Scott Feld, 1991):
  - In almost all social networks, the average degree of a node's neighbors ($\bar{k}_{nn}$) is **greater** than the average degree of the nodes in the network ($\bar{k}$).
- Why does this happen?
  - Highly connected individuals (popular hubs) belong to many people's neighborhood sets, disproportionately inflating neighbor degrees across the network.
- **Your friends, on average, have more friends than you do!**
:::

---

## Friendship Paradox: Node Calculations (A-D)

::: {.incremental}
- For the undirected graph shown previously (average degree $\bar{k} = 2.0$):
  - **Node A** (neighbor C): $\bar{k}_{nn(A)} = \frac{k_C}{1} = \frac{3}{1} = 3.0$
  - **Node B** (neighbors C, E, F): $\bar{k}_{nn(B)} = \frac{k_C + k_E + k_F}{3} = \frac{3 + 2 + 2}{3} \approx 2.33$
  - **Node C** (neighbors A, D, B): $\bar{k}_{nn(C)} = \frac{k_A + k_D + k_B}{3} = \frac{1 + 2 + 3}{3} = 2.0$
  - **Node D** (neighbors C, G): $\bar{k}_{nn(D)} = \frac{k_C + k_G}{2} = \frac{3 + 1}{2} = 2.0$
:::

---

## Friendship Paradox: Node Calculations (E-G)

::: {.incremental}
- Continuing calculations for the remaining nodes:
  - **Node E** (neighbors B, F): $\bar{k}_{nn(E)} = \frac{k_B + k_F}{2} = \frac{3 + 2}{2} = 2.5$
  - **Node F** (neighbors E, B): $\bar{k}_{nn(F)} = \frac{k_E + k_B}{2} = \frac{2 + 3}{2} = 2.5$
  - **Node G** (neighbor D): $\bar{k}_{nn(G)} = \frac{k_D}{1} = \frac{2}{1} = 2.0$
- **Summary of average neighbor degrees**:
  - $\bar{k}_{nn} = (3.0, 2.33, 2.0, 2.0, 2.5, 2.5, 2.0)$
:::


---

## Friendship Paradox: Network Average

::: {.incremental}
- From the node calculations on the previous slide, the average nearest neighbor degrees are:
  - $\bar{k}_{nn} = (3.0, 2.33, 2.0, 2.0, 2.5, 2.5, 2.0)$
- **Average neighbor degree** for the network:
  $$\text{Mean}(\bar{k}_{nn}) = \frac{3.0 + 2.33 + 2.0 + 2.0 + 2.5 + 2.5 + 2.0}{7} \approx 2.33$$
- **Demonstrating the Paradox**:
  - Average degree of nodes' neighbors ($\approx 2.33$) is **greater** than the average degree of nodes ($\bar{k} = 2.0$).
:::
