---
title: "Graph Theory: Paths & Connectivity"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Base 5-node undirected graph (House layout pointing right)
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# Base 5-node directed graph
edges_dir <- data.frame(
  from = c("A", "A", "C", "D", "E", "D", "E", "B", "B", "E"),
  to   = c("C", "D", "E", "C", "D", "E", "B", "E", "D", "C")
)
l_dir <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
gr_dir <- tbl_graph(nodes = l_dir, edges = edges_dir, directed = TRUE)

# Disconnected graph showing components and giant component
nodes_disconnected <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 1, 1, 0, -1, -1, 3, 4, 3.5),
  y    = c(1, 0.5, -0.5, -1, -0.5, 0.5, 0.5, 0, -1)
)
edges_disconnected <- data.frame(
  from = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 7, 7, 8),
  to   = c(2, 6, 3, 5, 4, 5, 5, 6, 6, 8, 9, 9)
)
gr_disconnected <- tbl_graph(nodes = nodes_disconnected, edges = edges_disconnected, directed = FALSE)

# 8-node directed graphs for connectivity flavors
nodes_8 <- data.frame(name = paste0("v", 1:8))
edges_8_strong <- data.frame(
  from = c("v1", "v2", "v3", "v4", "v5", "v6", "v7", "v8"),
  to   = c("v2", "v3", "v4", "v5", "v6", "v7", "v8", "v1")
)
gr_strong <- tbl_graph(nodes = nodes_8, edges = edges_8_strong, directed = TRUE)

edges_8_uni <- data.frame(
  from = c("v1", "v2", "v3", "v4", "v5", "v6", "v7"),
  to   = c("v2", "v3", "v4", "v5", "v6", "v7", "v8")
)
gr_uni <- tbl_graph(nodes = nodes_8, edges = edges_8_uni, directed = TRUE)

edges_8_weak <- data.frame(
  from = c("v1", "v2", "v3", "v4", "v5", "v6", "v8"),
  to   = c("v2", "v3", "v4", "v5", "v6", "v7", "v7")
)
gr_weak <- tbl_graph(nodes = nodes_8, edges = edges_8_weak, directed = TRUE)

# Local bridge graph
nodes_bridge_local <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 0.5, 1, 2, 2.5, 3, 0.5, 1.5, 2.5),
  y    = c(0.5, -0.5, 0, 0, -0.5, 0.5, -1.2, -1.5, -1.2)
)
edges_bridge_local <- data.frame(
  from = c(1, 1, 2, 3, 4, 4, 5, 2, 7, 8, 9),
  to   = c(2, 3, 3, 4, 5, 6, 6, 7, 8, 9, 5)
)
gr_bridge_local <- tbl_graph(nodes = nodes_bridge_local, edges = edges_bridge_local, directed = FALSE)
```

# Part 1: Indirect Connections & Paths

---

## Adjacency vs. Indirect Connections

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Adjacency** captures direct connections between nodes:
  - If $A$ is adjacent to $B$, they share an edge.
- But many network processes operate **indirectly**:
  - We must look beyond immediate contacts to the "friends of our friends."
  - **Diffusion Channels**: Indirect connections act as channels for gossip, information transmission, disease contagion, or social influence.
- In graph theory, indirect connections are formalized using three core concepts:
  - **Path**, **Connectivity**, and **Reachability**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Path?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **path** between $u$ and $v$ is an alternating sequence of nodes and edges that:
  - Begins with an **origin node** ($u$).
  - Ends with a **destination node** ($v$).
  - Does **not repeat any nodes** (which implies edges are not repeated either!).
- **Anatomy of a Path** (e.g. Red path from $B$ to $D$ on the right):
  - Written as: $B-D = \{BE, ED\}$ (length $l = 2$).
  - **Origin**: $B$, **Destination**: $D$ (together called the **end nodes**).
  - **Inner Nodes**: $E$ (the nodes in between).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Paths & Geodesics

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Multiple distinct paths can connect the same pair of nodes.
- Let's trace paths from $B$ to $A$:
  - $P_1 = \{BE, ED, DA\} \quad (l = 3)$
  - $P_2 = \{BE, EC, CA\} \quad (l = 3)$
  - $P_3 = \{BE, ED, DC, CA\} \quad (l = 4)$
  - $P_4 = \{BE, EC, CD, DA\} \quad (l = 4)$
- **Shortest Path (Geodesic)**:
  - The path of minimum length between two nodes.
  - **Geodesic Distance**: The length of the geodesic path (here, geodesic distance $d(B, A) = 3$, with tied geodesics $P_1$ and $P_2$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Cycle?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **cycle** is a path of length 3 or larger that starts and ends on the same node.
- **Key Concepts**:
  - Origin and destination are identical, but inner nodes are unique.
  - **Graph Girth**: The length of the **shortest** cycle in a graph (here, girth is $3$, the triangle $\{ED, DC, CE\}$).
  - **Graph Circumference**: The length of the **longest** cycle in a graph (here, circumference is $4$, $\{ED, DA, AC, CE\}$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Cycle Length: Girth vs. Circumference

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Girth (Shortest Cycle, Length 3)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p2 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 5) | (from == 4 & to == 5) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Circumference (Longest Cycle, Length 4)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p1 + p2
```

---

## Walks, Trails, and Paths

::: {.incremental}
- **Walk**: Any sequence of alternating nodes and edges connecting $u$ to $v$.
  - *No restrictions*: Nodes and edges can be repeated as much as you want!
  - e.g., $\{B \rightarrow E \rightarrow C \rightarrow E \rightarrow D\}$ (traverses $E$ and edge $E-C$ twice).
- **Trail**: A walk that does not repeat any **edges**.
  - *Repeated nodes are allowed*, but every edge in the sequence must be unique.
  - e.g., $\{A \rightarrow C \rightarrow D \rightarrow E \rightarrow C \rightarrow B\}$ (repeats node $C$ but all edges are unique).
- **Path**: A walk that does not repeat any **nodes** (which means edges cannot be repeated either).
- **The Nested Hierarchy**: All paths are trails, and all trails are walks, but the reverse is not true:
  $$Paths \subset Trails \subset Walks$$
:::

---

## Walks & Trails: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Walk (Repeats Node E and Edge E-C)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p2 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Trail (Repeats Node C, Edges Unique)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p1 + p2
```

# Part 2: Directed Paths & Reachability

---

## Paths in Directed Graphs (Digraphs)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, paths must strictly **respect the direction of the arrows**!
- Let's trace directed paths from $A$ to $B$ in our digraph:
  - $P_1 = \{AC, CE, EB\} \quad (l = 3)$
  - $P_2 = \{AD, DE, EB\} \quad (l = 3)$
  - $P_3 = \{AC, CD, DE, EB\} \quad (l = 4)$
- But can we find any directed path from $B$ to $A$?
  - **No!** All arrows lead away from $A$ and toward $B$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Directed Reachability

::: {.incremental}
- **Asymmetric Reachability**:
  - Node $A$ can reach $B$ via directed path $\{AD, DE, EB\}$ ($l_{AB} = 3$).
  - But node $B$ **cannot** reach $A$ because we cannot travel against the arrows ($l_{BA} = \infty$).
- **Mutual Reachability**:
  - Two nodes $u$ and $v$ are mutually reachable if there exists a directed path from $u \rightarrow v$ AND from $v \rightarrow u$.
  - These two paths **do not** have to be the same length!
  - **Example**: Nodes $B$ and $C$ are mutually reachable:
    - $B$ reaches $C$ via $\{BE, ED, DC\} \quad (l = 3)$
    - $C$ reaches $B$ via shortest path $\{CE, EB\} \quad (l = 2)$
:::

---

## Directed Reachability: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "D") | (from_name == "D" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Asymmetric (A reaches B, B cannot reach A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p2 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Mutual (B and C reach each other via paths of length 3 & 2)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p1 + p2
```

---

## Directed Cycles & DAGs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **directed cycle** is a cycle where all arrows flow in a single, continuous direction.
  - **Example**: Cycle starting and ending at $C$:
    - $\{CE, ED, DC\} \quad (l = 3)$
- **Directed Acyclic Graphs (DAGs)**:
  - Directed graphs that contain **no directed cycles**.
  - e.g. strict organizational command structures, directed trees.
  - Anti-symmetric relations (like "is boss of") structurally forbid cycles to maintain consistent hierarchy.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Ignoring the Arrows: Semipaths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- What if we want to trace connections in a directed graph but **ignore the arrows**?
- This is called a **semipath** (or semicycle if it is a closed loop).
- A semipath treats directed arcs as undirected edges.
- **Example**: Semipath connecting $B$ and $A$:
  - $\{BE, EC, CA\}$ (highlighted in red on the right).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "C") | (from_name == "A" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 3: Graph Connectivity & Components

---

## Connected vs. Disconnected Graphs

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Connected Graph**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
edges_connected <- rbind(edges_disconnected, data.frame(from = 6, to = 7))
gr_connected <- tbl_graph(nodes = nodes_disconnected, edges = edges_connected, directed = FALSE)

ggraph(gr_connected, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
ggraph(gr_disconnected, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

- **Connected**: A path of some finite length exists between *every* pair of nodes.
- **Disconnected**: At least one pair of nodes is unreachable (e.g. no path between A and G).

---

## Components & The Giant Component

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph with 2 Components**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
gr_disconnected_colors <- gr_disconnected |>
  activate(edges) |>
  mutate(highlight = ifelse(from <= 6 & to <= 6, "giant", "small"))

ggraph(gr_disconnected_colors, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("giant" = "red", "small" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("giant" = 1.8, "small" = 0.8), guide = "none") +
  geom_node_point(aes(color = ifelse(name %in% c("G", "H", "I"), "small", "giant")), size = 18) +
  scale_color_manual(values = c("giant" = "tan2", "small" = "skyblue"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: flex-start;"}
- **Component**: A maximally connected subgraph within a larger disconnected graph.
  - The graph on the left contains exactly **two components**.
- **Giant Component**: The largest connected component in a disconnected graph.
  - The red component $\{A, B, C, D, E, F\}$ is the **giant component** (of order 6).
  - The blue component $\{G, H, I\}$ is a smaller component (of order 3).
:::

::::

# Part 4: Node and Edge Connectivity (Bridges & Cutsets)

---

## Bridges & Local Bridges

::: {.incremental}
- We can disconnect a graph by removing edges.
- **Bridge**:
  - A single edge whose removal disconnects the graph, splitting it into separate components.
  - Represents the most critical, yet vulnerable transmission links in a network.
- **Mark Granovetter's "Strength of Weak Ties" (1973)**:
  - Bridges represent critical pathways for novel information diffusing between separate social clusters.
- **Local Bridge**:
  - An edge that, if removed, does not necessarily disconnect the entire graph, but significantly increases the shortest path length between its end nodes (acting as a key shortcut).
:::

---

## Bridges: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = ifelse(from == 3 & to == 4, "red", "normal"),
                     edge_width = ifelse(from == 3 & to == 4, "red", "normal"))) +
  scale_edge_color_manual(values = c("red" = "red", "normal" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("red" = 2.0, "normal" = 1.0), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("A Strict Bridge (C-D, removal disconnects graph)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p2 <- gr_bridge_local |>
  activate(edges) |>
  mutate(highlight = ifelse(from == 3 & to == 4, "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("A Local Bridge (C-D, removal increases distance from 1 to 5)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p1 + p2
```

---

## Node Connectivity & k-Connected Graphs

::: {.incremental}
- We can also disconnect a graph by **removing nodes** (vertices).
- Removing a node also removes all edges incident to it—nodes "take their links with them"!
- **Node Cutset (Vertex Cutset)**:
  - The minimum set of nodes whose removal disconnects the graph.
  - Nodes in the cutset represent critical **brokerage** or "connector" positions.
- **$k$-Connected Graph**:
  - A graph is $k$-connected if the minimum number of nodes required to disconnect it is $k$:
    - **1-Connected**: Removing 1 node can disconnect the graph (contains a cutpoint).
    - **2-Connected**: Requires removing at least 2 nodes to disconnect it.
    - **3-Connected**: Requires removing at least 3 nodes to disconnect it.
- **Menger's Theorem (1927)**: If a graph is $k$-connected, there are at least $k$ node-independent paths connecting each non-adjacent pair of vertices.
:::

---

## Edge Cutset: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- An **edge cutset** is a set of edges whose removal disconnects the graph.
- Removing both **C-A** and **D-A** (red) completely disconnects node $A$.
- This is an edge cutset of size **2**.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 3 & to == 5) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 1.8, "no" = 1.0), guide = "none") +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Edge Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_edges <- gr_undir |>
  activate(edges) |>
  filter(!((from == 3 & to == 5) | (from == 4 & to == 5)))

p_after <- ggraph(gr_undir_after_edges, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Edges Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Node Cutset: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- A **node cutset** is a set of nodes whose removal disconnects the graph.
- Removing node **E** (red) also removes its incident edges, completely isolating node $B$.
- This is a node cutset of size **1** (Node $E$ is a **cutpoint**).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "E", "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Node Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_node <- gr_undir |>
  filter(name != "E")

p_after <- ggraph(gr_undir_after_node, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Node Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Levels of k-Connectivity: Visual Examples

```{r}
#| label: fig-kconn-slides
#| fig-cap: "Graphs with different levels of k-connectivity."
#| fig-cap-location: bottom
#| fig-height: 4
#| fig-width: 12

# 1-connected star
gr_k1 <- tbl_graph(nodes = data.frame(name = 1:4), edges = data.frame(from = c(1, 1, 1), to = c(2, 3, 4)), directed = FALSE)
p_k1 <- ggraph(gr_k1, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, color = "white", fontface="bold") +
  theme_graph() + ggtitle("1-Connected (Star)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

# 2-connected ring
gr_k2 <- tbl_graph(nodes = data.frame(name = 1:4), edges = data.frame(from = c(1, 2, 3, 4), to = c(2, 3, 4, 1)), directed = FALSE)
p_k2 <- ggraph(gr_k2, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, color = "white", fontface="bold") +
  theme_graph() + ggtitle("2-Connected (Ring)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

# 3-connected complete K4
gr_k3 <- tbl_graph(nodes = data.frame(name = 1:4), edges = data.frame(from = c(1, 1, 1, 2, 2, 3), to = c(2, 3, 4, 3, 4, 4)), directed = FALSE)
p_k3 <- ggraph(gr_k3, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, color = "white", fontface="bold") +
  theme_graph() + ggtitle("3-Connected (Complete)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

p_k1 + p_k2 + p_k3
```

---

## Graph Cutsets: Quantitative Cohesion

::: {.incremental}
- **Cohesion** can be directly translated to graph-theoretic cutsets:
  - **Cohesive Groups**: Have large edge-cutsets and node-cutsets. Many connections or individuals must be removed to disrupt or fragment the group.
  - **Fragmented Groups**: Have small cutsets (such as bridges or single cutpoints). They are vulnerable to rapid disruption if key facilitators are removed.
- Any network can be decomposed into hierarchical subgraphs with specific levels of $k$-connectivity.
:::

# Part 5: Directed Connectivity & Matrices

---

## Flavors of Connectivity in Directed Graphs

::: {.incremental}
- Because of edge directionality, directed graphs have three levels of connectivity:
1. **Strongly Connected**: Every pair of nodes is **mutually reachable** via directed paths.
2. **Unilaterally Connected**: Every pair is connected by **at least one directed path** (from $u \rightarrow v$ OR $v \rightarrow u$, but not necessarily both).
3. **Weakly Connected**: Every pair of nodes is connected by at least a **semipath** (connected if we ignore arrow directions).
:::

---

## Directed Connectivity: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 4.5
p1 <- ggraph(gr_strong, layout = "circle") +
  geom_edge_fan(arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                start_cap = circle(4, 'mm'), end_cap = circle(6, 'mm'),
                color = "steelblue", edge_width = 1) +
  geom_node_point(size = 10, color = "tan2") +
  geom_node_text(aes(label = name), size = 5, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Strongly Connected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p2 <- ggraph(gr_uni, layout = "circle") +
  geom_edge_fan(arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                start_cap = circle(4, 'mm'), end_cap = circle(6, 'mm'),
                color = "steelblue", edge_width = 1) +
  geom_node_point(size = 10, color = "tan2") +
  geom_node_text(aes(label = name), size = 5, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Unilaterally Connected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p3 <- ggraph(gr_weak, layout = "circle") +
  geom_edge_fan(arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                start_cap = circle(4, 'mm'), end_cap = circle(6, 'mm'),
                color = "steelblue", edge_width = 1) +
  geom_node_point(size = 10, color = "tan2") +
  geom_node_text(aes(label = name), size = 5, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Weakly Connected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p1 + p2 + p3
```

---

## The Reachability Matrix ($D_r$)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- The **reachability matrix** ($D_r$) of a graph records whether each pair of nodes can reach each other ($D_r(i,j) = 1$ if node $i$ can reach $j$ via a path of any length).
- **Matrix Row & Column Patterns**:
  - **Receiver Nodes**: Have all zeros in their corresponding row of $D_r$ (cannot reach anyone, e.g. Node $D$ on the right).
  - **Transmitter Nodes**: Have all zeros in their corresponding column of $D_r$ (cannot be reached by anyone, e.g. Nodes $A, B, C$ on the right).
:::
:::

::: {.column width="40%"}
| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 0 | 1 | 1 | 1 |
| **B** | 0 | -- | 0 | 1 | 1 |
| **C** | 0 | 0 | -- | 1 | 1 |
| **D** | 0 | 0 | 0 | -- | 0 |
| **E** | 0 | 0 | 0 | 1 | -- |

: Reachability Matrix for our Directed Graph {#tbl-reach-slides}
:::

::::

---

## Reachability & Directed Connectivity

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- The reachability matrix reveals the global connectivity type of the digraph:
  - **Strongly Connected**: The matrix is the **all-ones matrix** (excluding diagonal).
  - **Unilaterally Connected**: The matrix contains some zeros, but for any pair, either $D_r(i,j) = 1$ or $D_r(j,i) = 1$. There are no mutually unreachable pairs.
  - **Weakly Connected**: Contains at least one pair of nodes that are mutually unreachable ($D_r(i,j) = 0$ AND $D_r(j,i) = 0$).
:::
:::

::: {.column width="40%"}
| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 0 | 1 | 1 | 1 |
| **B** | 0 | -- | 0 | 1 | 1 |
| **C** | 0 | 0 | -- | 1 | 1 |
| **D** | 0 | 0 | 0 | -- | 0 |
| **E** | 0 | 0 | 0 | 1 | -- |

: Reachability Matrix for our Directed Graph {#tbl-reach-slides-2}
:::

::::

---

## Graph Connectedness ($C$)

::: {.incremental}
- **Graph Connectedness** ($C$) measures the extent to which a network approaches the ideal where *everyone can reach everyone else*.
- Mathematically, it is the **density of the directed reachability matrix**:
  $$C = \frac{\sum_{i \neq j} D_r(i,j)}{n(n-1)}$$
- **Example Calculation**:
  - Sum of ones in our reachability matrix ($D_r$): $3 + 2 + 2 + 1 = 8$.
  - Total possible connections: $n(n-1) = 5(4) = 20$.
  - Connectedness: $C = \frac{8}{20} = 0.40 \quad (40\%)$.
:::

---

## The Geodesic Distance Matrix ($D_g$)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- The **geodesic distance matrix** ($D_g$) records the length of the **shortest path** (geodesic) connecting each pair of nodes.
- **Rules**:
  - Adjacent nodes get a distance of $1$.
  - Non-adjacent nodes get the length of their shortest path.
  - Unconnected nodes (unreachable) get a distance of **infinity** ($\infty$ or `Inf`).
- In undirected graphs, $D_g$ is symmetric. In directed graphs, $D_g$ is asymmetric.
:::
:::

::: {.column width="40%"}
| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 3 | 1 | 1 | 2 |
| **B** | Inf | -- | 3 | 2 | 1 |
| **C** | Inf | 2 | -- | 1 | 1 |
| **D** | Inf | 3 | 1 | -- | 2 |
| **E** | Inf | 1 | 2 | 1 | -- |

: Geodesic Distance Matrix for our Digraph {#tbl-dist-slides}
:::

::::

---

## Node Eccentricity, Diameter & Girth

From the geodesic distance matrix, we extract several fundamental whole-graph metrics:

::: {.incremental}
- **Node Eccentricity**: The maximum geodesic distance from a focal node to any other reachable node in the graph.
- **Graph Diameter** ($d$): The *maximum* geodesic distance across all pairs of connected nodes (longest shortest path).
  - For our undirected graph, $d = 3$.
- **Graph Girth**: The length of the *shortest* cycle in a graph (smallest loop).
  - For our undirected graph, the girth is $3$ (the triangle $\{ED, DC, CE\}$).
:::

---

## Conclusion & Core Takeaways

::: {.incremental}
- **Indirect connection is the engine of diffusion**:
  - Paths trace how resources, information, and influence flow through networks.
- **Directed path flows are asymmetric**:
  - Edge directions create unequal access, where some nodes can reach everyone, but are completely unreachable themselves.
- **Matrices provide global signatures**:
  - The **reachability matrix** ($D_r$) and the **distance matrix** ($D_g$) let us calculate connectedness, diameter, and eccentricity, transforming qualitative network structures into rigorous quantitative indices.
:::

---

## References {.unnumbered}
