---
title: "Graph Theory: Paths & Connectivity"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Base 5-node undirected graph (House layout pointing right)
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# Base 5-node directed graph
edges_dir <- data.frame(
  from = c("A", "A", "C", "D", "E", "D", "E", "B", "B", "E"),
  to   = c("C", "D", "E", "C", "D", "E", "B", "E", "D", "C")
)
l_dir <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
gr_dir <- tbl_graph(nodes = l_dir, edges = edges_dir, directed = TRUE)

# Disconnected graph showing components and giant component
nodes_disconnected <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 1, 1, 0, -1, -1, 3, 4, 3.5),
  y    = c(1, 0.5, -0.5, -1, -0.5, 0.5, 0.5, 0, -1)
)
edges_disconnected <- data.frame(
  from = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 7, 7, 8),
  to   = c(2, 6, 3, 5, 4, 5, 5, 6, 6, 8, 9, 9)
)
gr_disconnected <- tbl_graph(nodes = nodes_disconnected, edges = edges_disconnected, directed = FALSE)

# 5-node directed hourglass graphs for connectivity flavors
nodes_hour <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(-0.75, 0.75, 0, -0.75, 0.75), y = c(1, 1, 0, -1, -1))
edges_strong <- data.frame(
  from = c(1, 2, 4, 5, 1, 2, 3, 3, 4), 
  to   = c(2, 1, 5, 4, 3, 3, 4, 5, 1)
)
gr_strong <- tbl_graph(nodes = nodes_hour, edges = edges_strong, directed = TRUE)

edges_uni <- data.frame(
  from = c(1, 2, 4, 5, 1, 2, 3, 3), 
  to   = c(2, 1, 5, 4, 3, 3, 4, 5)
)
gr_uni <- tbl_graph(nodes = nodes_hour, edges = edges_uni, directed = TRUE)

edges_weak <- data.frame(
  from = c(1, 2, 4, 5, 1, 2, 4, 5), 
  to   = c(2, 1, 5, 4, 3, 3, 3, 3)
)
gr_weak <- tbl_graph(nodes = nodes_hour, edges = edges_weak, directed = TRUE)

# Local bridge graph
nodes_bridge_local <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 0.5, 1, 2, 2.5, 3, 0.5, 1.5, 2.5),
  y    = c(0.5, -0.5, 0, 0, -0.5, 0.5, -1.2, -1.5, -1.2)
)
edges_bridge_local <- data.frame(
  from = c(1, 1, 2, 3, 4, 4, 5, 2, 7, 8, 9),
  to   = c(2, 3, 3, 4, 5, 6, 6, 7, 8, 9, 5)
)
gr_bridge_local <- tbl_graph(nodes = nodes_bridge_local, edges = edges_bridge_local, directed = FALSE)

# Bridge graph (Strict bridge)
gr_bridge <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])
```

# Part 1: Indirect Connections & Paths

---

## Adjacency vs. Indirect Connections

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Adjacency** captures direct connections between nodes:
  - If $A$ is adjacent to $B$, they share an edge.
- But many network processes operate **indirectly**:
  - We must look beyond immediate contacts to the "friends of our friends."
  - **Diffusion Channels**: Indirect connections act as channels for gossip, information transmission, disease contagion, or social influence.
- In graph theory, indirect connections are formalized using three core concepts:
  - **Path**, **Connectivity**, and **Reachability**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Path?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **path** between $u$ and $v$ is an alternating sequence of nodes and edges that:
  - Begins with an **origin node** ($u$).
  - Ends with a **destination node** ($v$).
  - Does **not repeat any nodes** (which implies edges are not repeated either!).
- **Anatomy of a Path** (e.g. Red path from $B$ to $D$ on the right):
  - Written as: $B-D = \{BE, ED\}$ (length $l = 2$).
  - **Origin**: $B$, **Destination**: $D$ (together called the **end nodes**).
  - **Inner Nodes**: $E$ (the nodes in between).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Paths & Geodesics

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Multiple distinct paths can connect the same pair of nodes.
- Let's trace paths from $B$ to $A$:
  - $P_1 = \{BE, ED, DA\} \quad (l = 3)$
  - $P_2 = \{BE, EC, CA\} \quad (l = 3)$
  - $P_3 = \{BE, ED, DC, CA\} \quad (l = 4)$
  - $P_4 = \{BE, EC, CD, DA\} \quad (l = 4)$
- **Shortest Path (Geodesic)**:
  - The path of minimum length between two nodes.
  - **Geodesic Distance**: The length of the geodesic path (here, geodesic distance $d(B, A) = 3$, with tied geodesics $P_1$ and $P_2$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Cycle?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **cycle** is a path of length 3 or larger that starts and ends on the same node.
- **Key Concepts**:
  - Origin and destination are identical, but inner nodes are unique.
  - **Graph Girth**: The length of the **shortest** cycle in a graph (here, girth is $3$, the triangle $\{ED, DC, CE\}$).
  - **Graph Circumference**: The length of the **longest** cycle in a graph (here, circumference is $4$, $\{ED, DA, AC, CE\}$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Cycle Length: Girth vs. Circumference

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Girth (Shortest Cycle, Length 3)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p2 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 5) | (from == 4 & to == 5) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Circumference (Longest Cycle, Length 4)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p1 + p2
```

---

## Special Paths: Hamiltonian Paths & Cycles

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Hamiltonian Path**: A path that visits **every single node** in the graph exactly once.
  - *Example*: The red path on the right:
    $$B \rightarrow E \rightarrow C \rightarrow A \rightarrow D$$
- **Hamiltonian Cycle**: A cycle that visits **every single node** in the graph exactly once and returns to the origin.
  - *Question*: Is there a Hamiltonian cycle in this graph?
  - **No!** Since node $B$ is a "leaf node" ($k_B = 1$), any cycle must enter and leave $B$ via different edges, requiring $k_B \ge 2$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Walks, Trails, and Paths

::: {.incremental}
- **Walk**: Any sequence of alternating nodes and edges connecting $u$ to $v$.
  - *No restrictions*: Nodes and edges can be repeated as much as you want!
  - e.g., $\{B \rightarrow E \rightarrow C \rightarrow E \rightarrow D\}$ (traverses $E$ and edge $E-C$ twice).
- **Trail**: A walk that does not repeat any **edges**.
  - *Repeated nodes are allowed*, but every edge in the sequence must be unique.
  - e.g., $\{A \rightarrow C \rightarrow D \rightarrow E \rightarrow C \rightarrow B\}$ (repeats node $C$ but all edges are unique).
- **Path**: A walk that does not repeat any **nodes** (which means edges cannot be repeated either).
- **The Nested Hierarchy**: All paths are trails, and all trails are walks, but the reverse is not true:
  $$Paths \subset Trails \subset Walks$$
:::

---

## Walks & Trails: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Walk (Repeats Node E and Edge E-C)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p2 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Trail (Repeats Node C, Edges Unique)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p1 + p2
```

# Part 2: Directed Paths & Reachability

---

## Paths in Directed Graphs (Digraphs)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, paths must strictly **respect the direction of the arrows**!
- Let's trace directed paths from $A$ to $B$ in our digraph:
  - $P_1 = \{AC, CE, EB\} \quad (l = 3)$
  - $P_2 = \{AD, DE, EB\} \quad (l = 3)$
  - $P_3 = \{AC, CD, DE, EB\} \quad (l = 4)$
- But can we find any directed path from $B$ to $A$?
  - **No!** All arrows lead away from $A$ and toward $B$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Directed Reachability

::: {.incremental}
- **Asymmetric Reachability**:
  - Node $A$ can reach $B$ via directed path $\{AD, DE, EB\}$ ($l_{AB} = 3$).
  - But node $B$ **cannot** reach $A$ because we cannot travel against the arrows ($l_{BA} = \infty$).
- **Mutual Reachability**:
  - Two nodes $u$ and $v$ are mutually reachable if there exists a directed path from $u \rightarrow v$ AND from $v \rightarrow u$.
  - These two paths **do not** have to be the same length!
  - **Example**: Nodes $B$ and $C$ are mutually reachable:
    - $B$ reaches $C$ via $\{BE, ED, DC\} \quad (l = 3)$
    - $C$ reaches $B$ via shortest path $\{CE, EB\} \quad (l = 2)$
:::

---

## Directed Reachability: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "D") | (from_name == "D" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Asymmetric (A reaches B, B cannot reach A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p2 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Mutual (B and C reach each other via paths of length 3 & 2)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))

p1 + p2
```

---

## Directed Cycles & DAGs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **directed cycle** is a cycle where all arrows flow in a single, continuous direction.
  - **Example**: Cycle starting and ending at $C$:
    - $\{CE, ED, DC\} \quad (l = 3)$
- **Directed Acyclic Graphs (DAGs)**:
  - Directed graphs that contain **no directed cycles**.
  - e.g. strict organizational command structures, directed trees.
  - Anti-symmetric relations (like "is boss of") structurally forbid cycles to maintain consistent hierarchy.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Ignoring the Arrows: Semipaths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- What if we want to trace connections in a directed graph but **ignore the arrows**?
- This is called a **semipath** (or semicycle if it is a closed loop).
- A semipath treats directed arcs as undirected edges.
- **Example**: Semipath connecting $B$ and $A$:
  - $\{BE, EC, CA\}$ (highlighted in red on the right).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "C") | (from_name == "A" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 3: Graph Connectivity & Components

---

## Connected vs. Disconnected Graphs

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Connected Graph (Single Component)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5

    gr_l_conn <- tbl_graph(
        nodes = data.frame(name = as.character(1:9)),
        edges = data.frame(
            from = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 3, 2),
            to   = c(2, 3, 4, 5, 6, 7, 8, 9, 1, 5, 7, 8)
        ),
        directed = FALSE
    )
    
    ggraph(gr_l_conn, layout = "circle") +
        geom_edge_link(color = "steelblue", edge_width = 1.25) +
        geom_node_point(size = 15, color = "tan2") +
        geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
        theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph (Two Components)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5

    gr_l_disconn <- tbl_graph(
        nodes = data.frame(name = as.character(1:9)),
        edges = data.frame(
            from = c(1, 2, 3, 4, 5, 6, 1, 7, 8, 9),
            to   = c(2, 3, 4, 5, 6, 1, 4, 8, 9, 7)
        ),
        directed = FALSE
    )
    
    ggraph(gr_l_disconn, layout = "circle") +
        geom_edge_link(color = "steelblue", edge_width = 1.25) +
        geom_node_point(size = 15, color = "tan2") +
        geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
        theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

- **Connected**: A path of some finite length exists between *every* pair of nodes in the graph (left).
- **Disconnected**: At least one pair of nodes is unreachable; the graph is split into separate pieces (right).

---

## Components & The Giant Component

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Disconnected Graph with 2 Components**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
gr_disconnected_colors <- gr_disconnected |>
  activate(edges) |>
  mutate(highlight = ifelse(from <= 6 & to <= 6, "giant", "small"))

ggraph(gr_disconnected_colors, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("giant" = "red", "small" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("giant" = 1.8, "small" = 0.8), guide = "none") +
  geom_node_point(aes(color = ifelse(name %in% c("G", "H", "I"), "small", "giant")), size = 18) +
  scale_color_manual(values = c("giant" = "tan2", "small" = "skyblue"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: flex-start;"}
- **Component**: A maximally connected subgraph within a larger disconnected graph.
  - The graph on the left contains exactly **two components**.
- **Giant Component**: The largest connected component in a disconnected graph.
  - The red component $\{A, B, C, D, E, F\}$ is the **giant component** (of order 6).
  - The blue component $\{G, H, I\}$ is a smaller component (of order 3).
:::

::::

# Part 4: Edge and Node Connectivity (Cutsets & Bridges)

---

## Edge Cutsets & Connectivity

::: {.incremental}
- We can disconnect a connected graph by removing a subset of its edges.
- **Edge Cutset**:
  - Any set of edges whose removal disconnects the graph into two or more components.
- **Minimum Edge Cutset**:
  - An edge cutset containing the minimum possible number of edges needed to disconnect the graph.
- **Edge-Connectivity** ($\lambda(G)$):
  - The number of edges in a minimum edge cutset.
- **Bridge**:
  - A single edge whose removal disconnects the graph (a minimum edge cutset of size 1, so $\lambda(G) = 1$).
:::

---

## Edge Cutsets: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- In our 5-node graph on the right:
  - Edges **A-C** and **B-C** are highlighted in red (Before).
- **Removing both edges**:
  - Disconnects $\{C, D, E\}$ from $\{A, B\}$ (After).
  - This is an edge cutset of size **2**.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_cutset <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

p_before <- ggraph(gr_cutset, layout = 'kk') +
  geom_edge_link(aes(color = ifelse((from == 1 & to == 3) | (from == 2 & to == 3), "yes", "no")), edge_width = 1.25) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_cutset_after <- gr_cutset |>
  activate(edges) |>
  filter(!((from == 1 & to == 3) | (from == 2 & to == 3)))

p_after <- ggraph(gr_cutset_after, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Bridges: A Cutset of Size 1

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **Bridge**:
  - A single edge whose removal disconnects the graph (so $\lambda(G) = 1$).
- In our graph on the right, the edge **C-D** is highlighted in red.
  - **Removing edge C-D** completely disconnects $\{A, B, C\}$ from $\{D, E\}$.
  - Bridges represent the most critical, yet vulnerable transmission links in any network.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_bridge <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

p_before <- ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = ifelse(from == 3 & to == 4, "yes", "no")), edge_width = 1.25) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Bridge)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_bridge_after <- gr_bridge |>
  activate(edges) |>
  filter(!(from == 3 & to == 4))

p_after <- ggraph(gr_bridge_after, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Bridge Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Advanced: Local Bridges

::: {.incremental}
- In a classic paper on "The Strength of Weak Ties" (1973), Mark Granovetter developed the concept of a **local bridge**:
  - An edge that, if removed, does *not* disconnect the entire graph, but significantly increases the shortest path length between its end nodes.
- **Sociological Power**:
  - Local bridges act as crucial "shortcuts" that facilitate efficient, cross-group communication.
  - They are the primary channels through which novel information diffuses between separate social clusters.
:::

---

## Local Bridge: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Let's look at the edge **C-D** highlighted in red:
  - It connects the left cluster $\{A, B, C\}$ to the right cluster $\{D, E, F\}$.
- **Is it a bridge?** No. If removed, the graph remains connected via the bottom path ($B-G-H-I-E$).
- **Is it a local bridge?** Yes! 
  - Observed distance: $d(C, D) = 1$.
  - If removed, the distance increases to:
    $$d(C, D) = 5 \quad (C-B-G-H-I-E-D)$$
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_bridge_local |>
  activate(edges) |>
  mutate(highlight = ifelse(from == 3 & to == 4, "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Node Cutsets & Connectivity

::: {.incremental}
- We can also disconnect a graph by **removing nodes** (vertices).
- Removing a node also removes all edges incident to it—nodes "take their links with them"!
- **Node Cutset (Vertex Cutset)**:
  - Any set of nodes whose removal disconnects the remaining graph into two or more components.
- **Minimum Node Cutset**:
  - A node cutset containing the minimum possible number of nodes needed to disconnect the graph.
- **Node-Connectivity** ($\kappa(G)$):
  - The number of nodes in a minimum node cutset.
- **Articulation Node (Cutpoint)**:
  - A single node whose removal disconnects the graph (so $\kappa(G) = 1$).
:::

---

## Articulation Nodes: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- In our house graph on the right, node **E** is highlighted in red.
- **Removing node E**:
  - Removes all of its incident edges: $B-E, E-C, E-D$.
  - Completely disconnects node $B$ from the rest of the network.
  - This is a node cutset of size **1** (Node $E$ is an **articulation node**).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "E", "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Node Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_node <- gr_undir |>
  filter(name != "E")

p_after <- ggraph(gr_undir_after_node, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Node Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Node Cutset: An Edge Cutset Comparison

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Let's look at node $A$ in our house graph on the right.
- To disconnect $A$ by removing edges, we must remove **2 edges**: $C-A$ and $D-A$ (red).
- But to disconnect $A$ by removing nodes, we only need to remove **1 node**: $E$ (the cutpoint that isolates $B$).
- This illustrates how node and edge connectivity can differ in the same graph.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
p_before <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 3 & to == 5) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 1.8, "no" = 1.0), guide = "none") +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (With Edge Cutset)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_undir_after_edges <- gr_undir |>
  activate(edges) |>
  filter(!((from == 3 & to == 5) | (from == 4 & to == 5)))

p_after <- ggraph(gr_undir_after_edges, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Edges Removed)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## k-Connected Graphs & Robustness

::: {.incremental}
- **$k$-Connected Graph**:
  - A graph is $k$-connected (or $k$-node connected) if the minimum number of nodes required to disconnect it is $k$.
  - **1-Connected**: Removing 1 node disconnects the graph (contains an articulation point).
  - **2-Connected**: Requires removing at least 2 nodes to disconnect it.
  - **3-Connected**: Requires removing at least 3 nodes to disconnect it.
- **Menger's Theorem (1927)**:
  - If a graph is $k$-connected, there are at least $k$ node-independent paths (paths that do not share any common inner nodes) connecting each non-adjacent pair of vertices.
- **Social Cohesion**: Higher $k$-connectivity indicates a more robust, cohesive, and redundant communication structure.
:::

---

## 1-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **1-Connected Graph**:
  - A graph where removing **exactly 1 node** can disconnect the network.
- **Before** (Left):
  - A 7-node symmetric bow-tie graph, with the central bridge node (node 7) highlighted in red.
- **After** (Right):
  - The central node 7 is removed, leaving two completely separate connected components (triangles).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_k1 <- tbl_graph(
  nodes = data.frame(
    name = as.character(1:7),
    x = c(-2, -2, -1, 1, 2, 2, 0),
    y = c(1, -1, 0, 0, 1, -1, 0)
  ),
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 3, 4),
    to   = c(2, 3, 1, 5, 6, 4, 7, 7)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k1, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "7", "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (7-Node Bow-Tie with Cutpoint 7)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_k1_after <- gr_k1 |> filter(name != "7")
p_after <- ggraph(gr_k1_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Node 7 Removed, Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## 2-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **2-Connected Graph**:
  - A graph where it requires removing at least **2 nodes** to disconnect it.
- **Before** (Left):
  - A 6-node cycle with a diagonal chord (edge 2-5), with nodes 2 and 5 highlighted in red.
- **After** (Right):
  - Removing nodes 2 and 5 leaves the graph split into two disconnected components: {1, 6} and {3, 4}.
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_k2 <- tbl_graph(
  nodes = data.frame(name = as.character(1:6)),
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 2),
    to   = c(2, 3, 4, 5, 6, 1, 5)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k2, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(aes(color = ifelse(name %in% c("2", "5"), "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Before (6-Node Ring with Chord, Cutset {2, 5})") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

gr_k2_after <- gr_k2 |> filter(!name %in% c("2", "5"))
p_after <- ggraph(gr_k2_after, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("After (Nodes 2 & 5 Removed, Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## 3-Connectivity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **3-Connected Graph**:
  - A graph where it requires removing at least **3 nodes** to disconnect it.
- **Before** (Left):
  - A 3-regular 6-node prism graph (two concentric triangles), with nodes 1, 2, and 6 highlighted in red.
- **After** (Right):
  - Removing nodes 1, 2, and 6 leaves the single edge 4-5 and completely isolates node 3 (disconnected).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
nodes_prism <- data.frame(
  name = as.character(1:6),
  x = c(0, -0.43, 0.43, 0, -1.04, 1.04),
  y = c(0.5, -0.25, -0.25, 1.2, -0.6, -0.6)
)
gr_k3 <- tbl_graph(
  nodes = nodes_prism,
  edges = data.frame(
    from = c(1, 2, 3, 4, 5, 6, 1, 2, 3),
    to   = c(2, 3, 1, 5, 6, 4, 4, 5, 6)
  ),
  directed = FALSE
)

p_before <- ggraph(gr_k3, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.1) +
  geom_node_point(aes(color = ifelse(name %in% c("1", "2", "6"), "yes", "no")), size = 15) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Before (6-Node Prism, Cutset {1, 2, 6})") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

gr_k3_after <- gr_k3 |> filter(!name %in% c("1", "2", "6"))
p_after <- ggraph(gr_k3_after, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.1) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, color = "white", fontface="bold") +
  theme_graph() + ggtitle("After (Nodes Removed, Node 3 Isolated)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Graph Cutsets: Quantitative Cohesion

::: {.incremental}
- **Cohesion** can be directly translated to graph-theoretic cutsets:
  - **Cohesive Groups**: Have large edge-cutsets and node-cutsets. Many connections or individuals must be removed to disrupt or fragment the group.
  - **Fragmented Groups**: Have small cutsets (such as bridges or single cutpoints). They are vulnerable to rapid disruption if key facilitators are removed.
- Any network can be decomposed into hierarchical subgraphs with specific levels of $k$-connectivity.
:::

## The Connectivity Inequality

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- For any undirected graph $G$, node-connectivity ($\kappa$), edge-connectivity ($\lambda$), and minimum degree ($\delta$) obey a fundamental structural hierarchy:
  $$\kappa(G) \le \lambda(G) \le \delta(G)$$
- **A Strict Inequality Example ($1 < 2 < 3$)**:
  - In our 9-node "double-clique" graph on the right:
    - **Minimum Degree ($\delta = 3$)**: The least connected nodes ($3, 4, 7, 8$) have exactly 3 edges.
    - **Edge-Connectivity ($\lambda = 2$)**: Removing $\{1-9, 2-9\}$ separates the left clique from the waist. No single bridge exists.
    - **Node-Connectivity ($\kappa = 1$)**: Removing the central cutpoint (node 9) completely disconnects the graph.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_ineq <- tbl_graph(
  nodes = data.frame(
    name = as.character(1:9),
    x = c(-1, -1, -2, -2, 1, 1, 2, 2, 0),
    y = c(0.5, -0.5, 0.5, -0.5, 0.5, -0.5, 0.5, -0.5, 0)
  ),
  edges = data.frame(
    from = c(1, 1, 1, 2, 2, 3, 5, 5, 5, 6, 6, 7, 1, 2, 5, 6),
    to   = c(2, 3, 4, 3, 4, 4, 6, 7, 8, 7, 8, 8, 9, 9, 9, 9)
  ),
  directed = FALSE
)

ggraph(gr_ineq, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

# Part 5: Tree Graphs & Forests

---

## What is a Tree Graph?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- If a graph is both **connected** *and* has no **cycles** (acyclic), then it is a **tree graph**.
- **The Cycle Restriction**:
  - A cycle is a path that begins and ends with the same node.
  - In a tree graph, you can never start with a node and trace a sequence of distinct edges that take you back to where you started!
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
set.seed(123)
gr_tree_5 <- as_tbl_graph(sample_tree(5, directed = FALSE)) %>% mutate(name = LETTERS[1:5])
ggraph(gr_tree_5, layout = 'tree') +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Properties of Tree Graphs

::: {.incremental}
- Tree graphs possess several unique mathematical and structural properties:
  1. **The Single Path Property**:
     - *Every* pair of nodes in a tree graph is linked by exactly **one unique path**.
     - There are no redundant or alternative paths between any two actors.
  2. **The One-Cycle Property**:
     - Adding a *single* new edge connecting any pair of non-adjacent nodes always creates **exactly one cycle**, destroying its tree status.
:::

---

## Tree Graph Metrics: Vertices & Edges

::: {.incremental}
- In any tree graph, the number of edges ($m$) is always exactly equal to the number of nodes ($n$) minus one:
  $$m = n - 1 \qquad \iff \qquad n = m + 1$$
- **Sum of Degrees Identity**:
  - Recall that the sum of degrees in any undirected graph is $2m$. 
  - Substituting $m = n - 1$, the sum of degrees of a tree graph is always:
    $$\sum_i k_i = 2(n - 1) = 2n - 2$$
  - *In a tree graph, the sum of degrees is always exactly twice the number of nodes minus two!*
:::

---

## Tree Graph Density & Average Degree

::: {.incremental}
- Because of the strict $m = n - 1$ edge constraint, tree graph density and average degree simplify beautifully:
  - **Density of a Tree Graph**:
    $$d(T) = rac{2m}{n(n-1)} = rac{2(n-1)}{n(n-1)} = rac{2}{n}$$
  - **Average Degree of a Tree Graph**:
    $$ar{k} = rac{\sum k_i}{n} = rac{2n - 2}{n}$$
- *For our 5-node tree graph: density is $2/5 = 0.40$ ($40\%$), and average degree is $(10-2)/5 = 1.6$.*
:::

---

## Trees as Minimally Connected Graphs

::: {.incremental}
- Trees are **minimally connected** graphs:
  - They are connected using the *smallest possible number of edges* needed to connect $n$ vertices.
- **The Bridges Rule**:
  - Removing even a *single* edge from a tree graph completely disconnects the graph.
  - Thus, **every edge of a tree graph is a bridge** (a minimum edge cutset of size 1)!
- Disconnecting a tree graph by removing any edge splits it into separate connected components, which are themselves trees.
:::

---

## Forests: Visualizing Disconnection

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **Forest**:
  - A disconnected graph whose components are all tree graphs.
- **Before** (Left):
  - A connected 6-node tree graph.
- **After** (Right):
  - Removing edge **D-F** splits the graph into two separate trees (a **forest**).
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_f <- as_tbl_graph(sample_tree(6, directed = FALSE)) %>% mutate(name = LETTERS[1:6])
# Force manual layout for consistency
layout_f <- create_layout(gr_f, layout = "tree")
x_f <- layout_f$x
y_f <- layout_f$y

# Find a valid edge to highlight and remove, say edge 3-4 (C-D) or similar
edges_f <- as.data.frame(gr_f %>% activate(edges) %>% as_tibble())
# Let's highlight the first edge
from_node <- edges_f$from[1]
to_node <- edges_f$to[1]

p_before <- ggraph(gr_f, layout = "manual", x = x_f, y = y_f) +
  geom_edge_link(aes(color = ifelse(from == from_node & to == to_node, "yes", "no")), edge_width = 1.25) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Before (Connected Tree)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

gr_f_after <- gr_f %>% 
  activate(edges) %>% 
  filter(!(from == from_node & to == to_node))

p_after <- ggraph(gr_f_after, layout = "manual", x = x_f, y = y_f) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("After (Forest: 2 Tree Components)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

p_before + p_after
```
:::

::::

---

## Special Types of Trees: Line and Star Graphs

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- **The Line Graph ($L_n$)**:
  - A single straight path connecting $n$ nodes.
  - Used to model sequential diffusion processes (like the "Telephone Game").
- **The Star Graph ($S_n$)**:
  - A central hub connected to $n-1$ leaf nodes.
  - Represents extreme structural centralization.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4.5
gr_line <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]), edges = data.frame(from = 1:4, to = 2:5), directed = FALSE)
p_line <- ggraph(gr_line, layout = "linear") +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Line Graph (L5)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

gr_star <- tbl_graph(nodes = data.frame(name = LETTERS[1:5]), edges = data.frame(from = c(1, 1, 1, 1), to = 2:5), directed = FALSE)
p_star <- ggraph(gr_star, layout = "star") +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Star Graph (S5)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_line + p_star
```
:::

::::

---

## References {.unnumbered}
