---
title: "Graph Theory"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    theme: default
    css: ucla-theme.css
---

```{r setup, include=FALSE}
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)

# Base graph and layout for neighborhood slides
gr_base <- tbl_graph(
  nodes = data.frame(name = 1:7),
  edges = data.frame(
    from = c(1, 4, 4, 2, 5, 6),
    to   = c(3, 3, 7, 5, 6, 2)
  ),
  directed = FALSE
)

set.seed(42)
layout_fr <- create_layout(gr_base, layout = "fr")
```

# Introduction to Graphs

---

## What is a Graph?

::: {.incremental}
- A **graph** ($G$) is an abstract mathematical object consisting of two main parts:
- A **vertex set**, $V$, which contains all the nodes (actors) in the network.
  - e.g., $V = \{A, B, C, D\}$
- An **edge set**, $E$, which contains all the ties (relationships) between the nodes.
  - e.g., $E = \{AB, AC\}$
- Together, they define the graph: $G = \{V, E\}$
:::

---

## The Building Blocks: Nodes

::: {.incremental}
- In social networks, **nodes** represent the actors
  - Usually drawn as circles or points in a diagram
  - Often individuals (people) or organizations (groups, companies)
- In wider applications of **network science**:
  - Nodes can represent any entities in a system
  - Examples: power generation stations, proteins, websites
:::

---

## The Building Blocks: Edges

::: {.incremental}
- **Edges** represent a connection or a **social tie** between two nodes
  - Can be a permanent relationship (e.g., "brother of")
  - Or a fleeting interaction (e.g., a text message, co-presence)
- In set theory:
  - The set of edges is a collection of pairs of nodes
  - If node $A$ is linked to node $B$, then $AB \in E$ (or written as $V_A V_B$)
:::

---

## Edges as Channels

::: {.incremental}
- Edges act as channels for **flows of content**:
  - Positive/Neutral: influence, advice, information, support, power, traffic
  - Negative: disease, bullying, gossip, conflict
:::

---

## What is a Graph? (An Example) (1)

::: {.incremental}
- Let's define a graph $G$ with 4 nodes ($A, B, C, D$) and 2 edges ($AB, AC$):
  - $V = \{A, B, C, D\}$
  - $E = \{AB, AC\}$
  - $G = \{V, E\}$
- The **cardinality** (number of members in a set) is denoted by $|N|$:
  - Vertex set cardinality: $|V| = n = 4$
  - Edge set cardinality: $|E| = m = 2$
:::

---

## What is a Graph? (An Example) (2)

::: {.incremental}
- This graph can be represented by the signature $G(n, m) = G(4, 2)$
- Note the lack of edges:
  - No edge between $B$ and $C$ means no relationship between them.
  - Node $D$ is not connected to any other node.
:::

---

## Variations of $G(4, 2)$

- The signature $G(4, 2)$ (4 nodes, 2 edges) does not uniquely define a single graph.
- It can describe multiple topologically distinct graphs.

---

## G(4, 2) Variation 1

```{r}
#| label: fig-g42-1
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = 1, to = 2:3)) %>% 
    mutate(name = letters[1:4])
l <- as.matrix(ggraph(gr, layout = 'tree')$data[, 1:2])
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## G(4, 2) Variation 2

```{r}
#| label: fig-g42-2
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 4), to = c(2, 3))) %>% 
    mutate(name = letters[1:4])    
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## G(4, 2) Variation 3

```{r}
#| label: fig-g42-3
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = c(1, 2), to = c(4, 3))) %>% 
    mutate(name = letters[1:4])    
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## G(4, 2) Variation 4 (Unlabeled)

```{r}
#| label: fig-g42-4
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6
gr <- create_empty(4) %>% 
    bind_edges(data.frame(from = 1, to = 2:3))  
ggraph(gr, layout = 'manual', x=l[,1], y=l[,2]) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 22, color = "tan2") +
  theme_graph() + coord_cartesian(clip = "off")
```

---

## Graph Labeling

::: {.incremental}
- **Labeled Graphs**:
  - Distinctions are made between nodes (e.g., A, B, C are different people).
  - Graph is defined by its specific structural representation (e.g., the last graph in the previous slide).
- **Unlabeled Graphs**:
  - Nodes are interchangeable or anonymous.
  - The focus is only on the abstract structure of connections.
:::

---

## Trivial Graphs

::: {.incremental}
- **Trivial Graph**:
  - A $G(1, 0)$ graph: a single isolated node with zero connections.
  - Social equivalent: An isolated person with no connections (e.g., the Japanese *Hikikomori*).
:::

---

## Non-Trivial Graphs

::: {.incremental}
- **Non-Trivial Graphs**:
  - Any graph larger than $G(1, 0)$ used for social network analysis.
  - **Connected Dyad**: The $G(2, 1)$ graph. Two nodes connected by a single link. The "hydrogen atom" of society.
  - **Disconnected Dyad**: A $G(2, 0)$ graph. Two nodes with no link (e.g., two people on a deserted island not speaking).
:::

---

## Basic Graph Properties

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- **Adjacent**: Joined by an edge (e.g., $A$ and $B$).
- **Nonadjacent**: Not joined by an edge (e.g., $A$ and $C$).
- **Isolate**: Node with zero incident edges (e.g., $D$).
- **Incident**: An edge "touches" a node (e.g., edge $AB$ is incident on $A$).
- **End Vertices**: Nodes at the ends of an edge (e.g., $A$ and $B$ for edge $AB$).
:::
:::

::: {.column width="40%"}
```{r}
#| label: fig-basic-props
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6

gr <- create_empty(4, directed = FALSE) %>%
  bind_edges(data.frame(from = 1:2, to = 2:3)) %>%
  mutate(name = c("A", "B", "C", "D"))

ggraph(gr, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## The Graph Complement

::: {.incremental}
- The **complement** ($G'$) of a graph $G$ contains an edge between two nodes *if and only if* they are disconnected in $G$.
- It represents the "negative" of the original graph, showing all the ties that *could* exist but don't.
:::

---

## Graph for Complement Example

```{r}
#| label: fig-complement-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

gr <- create_empty(6, directed = FALSE) %>%
      bind_edges(data.frame(from = 1, to = c(2, 4, 6))) %>%
      bind_edges(data.frame(from = 6, to = c(3, 5))) %>%
      mutate(name = toupper(letters[1:6])) 
p <- ggraph(gr, layout = 'kk') 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p1 <- p + theme_graph()
p1
```

---

## The Graph Complement

```{r}
#| label: fig-complement-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

gr.c <- complementer(gr)
l <- as.matrix(p$data[, 1:2])
p <- ggraph(gr.c, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p2 <- p + theme_graph()
p2
```

---

## Simple Graphs

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- Most social networks are represented as **simple graphs**.
- A graph is simple if it meets two requirements:
  1. **No Multiple Edges**: Only one edge can join any pair of nodes (unlike $A$ and $B$).
  2. **No Loops**: No edges connect a node to itself (unlike node $C$).
- Loops and multiple edges do not make sense for most typical sociological applications.
:::
:::

::: {.column width="40%"}
```{r}
#| label: fig-non-simple
#| echo: false
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 6

gr_non_simple <- create_empty(3, directed = FALSE) %>%
  bind_edges(data.frame(from = c(1, 1, 2, 3), to = c(2, 2, 3, 3))) %>%
  mutate(name = c("A", "B", "C"))

ggraph(gr_non_simple, layout = 'circle') +
  geom_edge_fan(color = "steelblue", edge_width = 1.5, strength = 1.5) +
  geom_edge_loop(color = "tomato", edge_width = 1.5) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 12, color = "white") +
  theme_graph() +
  coord_cartesian(clip = "off")
```
:::

::::

---

## Graphs are Not Pictures!

::: {.incremental}
- Do not confuse a mathematical "graph" (a set of two sets) with its "drawing" (a point and line plot).
- The same graph can have many different, and even misleading, drawings.
- The drawing is just a visual aid; the underlying structure is what matters.
:::

---

## Same Graph, Different Drawings (1)

```{r}
#| label: fig-random-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

set.seed(678)
l <- layout_randomly(gr)
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p <- p + theme_graph() + coord_cartesian(clip = "off")
p
```

---

## Same Graph, Different Drawings (2)

```{r}
#| label: fig-random-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

set.seed(789)
l <- layout_randomly(gr)
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", edge_width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 12, color = "white")
p <- p + theme_graph() + coord_cartesian(clip = "off")
p
```

# Graphs and Subgraphs

---

## Graphs and Subgraphs

::: {.incremental}
- Sometimes we are interested in analyzing only some parts of a **whole network**.
- A **subgraph** $G' = \{V', E'\}$ is a subset of the original graph $G = \{V, E\}$, which is written as $G' \subset G$:
  - The node set $V'$ is a subset of $V$ ($V' \subset V$).
  - The edge set $E'$ is a subset of $E$ ($E' \subset E$).
  - Edges in the subgraph must have both endpoints in the node subset $V'$.
:::

---

## Original Graph

```{r}
#| label: fig-subgraph-slides-1
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

set.seed(367)
gr <- play_gnm(n = 10, m = 20, directed = FALSE, loops = FALSE) %>% 
  mutate(name = LETTERS[1:10])
l <- ggraph(gr, layout = 'kk')
l <- as.matrix(l$data[, 1:2])
p <- ggraph(gr, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p1 <- p + theme_graph() + coord_cartesian(clip = "off")
p1
```

---

## A Subgraph of the Original

:::: {.columns}

::: {.column width="50%"}
::: {.incremental}
- This graph is a **subgraph** of the original graph from the previous slide.
- It was created by taking a **subset of nodes**: 
  - $V' = \{A, B, C, D, E\}$
- We retained only the edges from the original graph that connect these five nodes to each other.
- All other nodes ($F, G, H, I, J$) and any edges connected to them were removed.
:::
:::

::: {.column width="50%"}
```{r}
#| label: fig-subgraph-slides-2
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 6

gr2 <- induced_subgraph(gr, toupper(letters[1:5]))
p <- ggraph(gr2, layout = l[1:5, ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p2 <- p + theme_graph() + coord_cartesian(clip = "off")
p2
```
:::

::::

---

## Vertex-Induced Subgraphs

::: {.incremental}
- A **vertex-induced subgraph** is created by selecting a subset of nodes ($V' \subset V$) and including *all* the edges from the original graph that connect nodes within that subset.
- This is the most common method for creating subgraphs in social network analysis.
:::

---

## Vertex-Induced Subgraph Example

```{r}
#| label: fig-subgraph-induced
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

gr3 <- induced_subgraph(gr, c("G", "D", "E", "I"))
p <- ggraph(gr3, layout = l[c(4, 5, 7, 9), ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p3 <- p + theme_graph() + coord_cartesian(clip = "off")
p3
```

---

## Vertex-Deleted Subgraphs (1)

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- We define a subgraph by *deleting* nodes: $G' = G - \{a, b, c\}$, where $\{a, b, c\}$ is the set of nodes being deleted.
- **CRITICAL RULE**: When you delete nodes, **all edges incident to those nodes are automatically removed as well**.
- For example, deleting $\{F, G, H, I, J\}$ from our 10-node graph yields the subgraph on the right.
- This shows that a vertex-deleted subgraph can be equivalent to a vertex-induced subgraph on the remaining nodes (here, $\{A, B, C, D, E\}$).
:::
:::

::: {.column width="40%"}
```{r}
#| label: fig-subgraph-deleted-node
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 6

gr_del <- gr - c("F", "G", "H", "I", "J")
p <- ggraph(gr_del, layout = l[1:5, ]) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p + theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Vertex-Deleted Subgraphs (2)

::: {.incremental}
- Two special cases of vertex deletion are:
  - **Null Graph**: The subgraph that results from removing *all* nodes from the original graph.
  - **Singleton Graph** (or **Trivial Graph**): The subgraph that results from removing all nodes *except for one*.
:::

---

## Edge-Deleted Subgraphs

:::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- We define a subgraph by removing edges: $G' = G - \{e_1, e_2, ...\}$, where $\{e_1, e_2, ...\}$ is the set of edges being deleted.
- **CRITICAL RULE**: When you delete edges, the nodes that were connected by those edges **remain in the graph**.
- **Spanning Subgraph**: A subgraph created by removing *only* edges while leaving all original nodes intact.
- **Empty Graph**: A special case of edge deletion that results from removing *all* edges from the original graph, leaving only the set of isolated nodes.
- The example on the right removes the six edges $\{AB, AE, AC, CI, CE, GJ\}$; notice all 10 nodes remain!
:::
:::

::: {.column width="45%"}
```{r}
#| label: fig-subgraph-deleted-edges
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 6

gr4 <- gr - edge("A|B") - edge("G|J") - edge("C|I") - edge("A|E") - edge("A|C") - edge("C|E")

p <- ggraph(gr4, layout = l) 
p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
p + theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Special Subgraphs: Ego Graph

::: {.incremental}
- **Ego Graph**: A special type of subgraph defined by picking a central node (ego) and the nodes that are connected to it, along with the edges connecting those nodes.
- The central node's adjacent contacts are called its **alters**.
- Ego graphs are particularly useful for analyzing social networks at the level of individual actors.
:::

---

## Vertex Fusion (1)

::: {.incremental}
- We can construct a new graph from a graph by **fusing** two vertices.
- **Vertex Fusion**:
  - Replace two original vertices with a single vertex.
  - Keep the combined adjacency relations of both original vertices.
:::

---

## Vertex Fusion (2)

::: {.incremental}
- Vertex fusion is useful in social network analysis for:
  - Collapsing equivalence classes.
  - Analyzing blockmodels.
  - Merging groups or roles.
:::

---

## Set Operations on Graphs: Union

::: {.incremental}
- Because graphs are defined as sets, we can perform standard set operations on them.
- **Union** ($\cup$): Combine the vertices and edges of two graphs.
  - $V(G_1 \cup G_2) = V(G_1) \cup V(G_2)$
  - $E(G_1 \cup G_2) = E(G_1) \cup E(G_2)$
:::

---

## Set Operations on Graphs: Intersection

::: {.incremental}
- **Intersection** ($\cap$): Keep only the vertices and edges present in *both* graphs.
  - $V(G_1 \cap G_2) = V(G_1) \cap V(G_2)$
  - $E(G_1 \cap G_2) = E(G_1) \cap E(G_2)$
:::

---

# Node Neighborhoods

---

## Node Neighborhoods

::: {.incremental}
- **Neighborhood** ($\mathcal{N}(v)$):
  - The set of nodes that are directly connected (adjacent) to node $v$.
  - These are a node's immediate **neighbors**.
  - Isolates have an empty neighborhood set ($\emptyset$).
- Example: $\mathcal{N}(A) = \{B, C, D, F, H\}$
:::

---

## Neighborhood Intersection

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- **Neighborhood Intersection** ($\mathcal{N}(u) \cap \mathcal{N}(v)$) is the set of **common neighbors** connected to both $u$ and $v$.
- In our graph:
  - $N(1) = \{3\}$
  - $N(4) = \{3, 7\}$
- The intersection is the single node they share:
  - $N(1) \cap N(4) = \{3\}$
:::
:::

::: {.column width="40%"}
```{r echo=FALSE}
#| label: fig-neighborhood-intersection
gr_base |>
  activate(nodes) |>
  mutate(
    highlight = case_when(
      name %in% c(1, 4) ~ "focal",
      name == 3 ~ "common",
      TRUE ~ "other"
    )
  ) |>
  ggraph(layout = "manual", x = layout_fr$x, y = layout_fr$y) +
    geom_edge_link(color = "gray80", width = 1.5) +
    geom_node_point(aes(color = highlight), size = 25) +
    geom_node_text(aes(label = name), color = "black", size = 8) +
    scale_color_manual(
      values = c("focal" = "tomato", "common" = "gold", "other" = "skyblue"),
      name = "", labels = c("Common Neighbor", "Focal Nodes", "Other")
    ) +
    theme_graph() +
    theme(legend.position = "bottom")
```
:::

::::

---

## Neighborhood Union

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- **Neighborhood Union** ($\mathcal{N}(u) \cup \mathcal{N}(v)$) is the set of all nodes connected to *either* $u$ or $v$.
- In our graph:
  - $N(1) = \{3\}$
  - $N(4) = \{3, 7\}$
- The union contains all unique neighbors from both sets:
  - $N(1) \cup N(4) = \{3, 7\}$
:::
:::

::: {.column width="40%"}
```{r echo=FALSE}
#| label: fig-neighborhood-union
gr_base |>
  activate(nodes) |>
  mutate(
    highlight = case_when(
      name %in% c(1, 4) ~ "focal",
      name %in% c(3, 7) ~ "neighbor",
      TRUE ~ "other"
    )
  ) |>
  ggraph(layout = "manual", x = layout_fr$x, y = layout_fr$y) +
    geom_edge_link(color = "gray80", width = 1.5) +
    geom_node_point(aes(color = highlight), size = 25) +
    geom_node_text(aes(label = name), color = "black", size = 8) +
    scale_color_manual(
      values = c("focal" = "tomato", "neighbor" = "gold", "other" = "skyblue"),
      name = "", labels = c("Focal Nodes", "Union of Neighbors", "Other")
    ) +
    theme_graph() +
    theme(legend.position = "bottom")
```
:::

::::

---

# Directed Graph Elements & Metrics

---

## Neighborhoods in Directed Graphs (1)

::: {.incremental}
- Because links are directional, nodes have two neighborhoods.
- **In-Neighborhood** ($N^{in}(v)$):
  - Set of source nodes pointing *to* $v$.
  - Example: $N^{in}(D) = \{B, E, G\}$.
:::

---

## Neighborhoods in Directed Graphs (2)

::: {.incremental}
- **Out-Neighborhood** ($N^{out}(v)$):
  - Set of destination nodes that $v$ points *to*.
  - Example: $N^{out}(B) = \{A, C, D\}$.
- A node appears in both neighborhoods only if there is a **reciprocal** tie.
:::

---

# Node Degree

---

## Node Degree in Undirected Graphs

::: {.incremental}
- A node's **degree** ($k_i$) is the number of edges incident on it.
- It is the cardinality (size) of a node's neighborhood: $k_i = |\mathcal{N}(i)|$.
- Sociological meaning: **Popularity, Influence, or Centrality**.
:::

---

## Node Degree in Directed Graphs: Indegree

::: {.incremental}
- Directed graphs have two distinct degrees for each node.
- **Indegree** ($k^{in}_i$):
  - Number of edges pointing to node $i$ (cardinality of $N^{in}(i)$).
  - Sociological meaning: **Popularity, Prestige, or Status**.
:::

---

## Node Degree in Directed Graphs: Outdegree

::: {.incremental}
- **Outdegree** ($k^{out}_i$):
  - Number of edges pointing from node $i$ (cardinality of $N^{out}(i)$).
  - Sociological meaning: **Sociability or Activity**.
:::

---

# Basic Graph Metrics

---

## Graph Order and Size (1)

::: {.incremental}
- **Graph Order** ($n$):
  - The number of nodes in the graph ($n = |V|$).
  - Cardinality of the vertex set.
- **Graph Size** ($m$):
  - The number of edges in the graph ($m = |E|$).
  - Cardinality of the edge set.
:::

---

## Graph Order and Size (2)

::: {.incremental}
- In our undirected graph (10 nodes, 20 edges): $n = 10$, $m = 20$.
- In our directed graph (7 nodes, 11 edges): $n = 7$, $m = 11$.
:::

---

## Graph Maximum Size

::: {.incremental}
- The **maximum size** of a simple graph of order $n$ is the number of edges in a **complete graph** (where every node is connected to every other node).
- Undirected: $Max(E) = \frac{n(n - 1)}{2}$
- Directed: $Max(E) = n(n - 1)$
:::

---

## Graph Density

::: {.incremental}
- **Density** ($d$) is the ratio of the observed number of edges ($m$) to the maximum possible number of edges.
- It's a value between 0 (empty) and 1 (complete).
- Undirected: $d = \frac{2m}{n(n-1)}$
- Directed: $d = \frac{m}{n(n-1)}$
- It can be interpreted as the probability that any two randomly chosen nodes are connected.
- Denser networks are less vulnerable to breakdowns when key nodes are removed.
:::

---

# Degree-Based Graph Metrics

---

## Graph Degree Set and Sequence

::: {.incremental}
- **Graph Degree Set**: A list of the degrees of every node in the graph.
- **Graph Degree Sequence**: The degree set sorted in descending order.
- This provides a "fingerprint" of the graph's overall connectivity structure.
:::

---

## Degree Range & Sum

::: {.incremental}
- **Maximum Degree** ($k_{max}$) and **Minimum Degree** ($k_{min}$).
- **Degree Range** ($k_r$): $k_{max} - k_{min}$.
- **Sum of Degrees** ($\sum k_i$):
  - In an undirected graph, this is always twice the number of edges: $\sum k_i = 2m$.
:::

---

## Graphical Degree Sequences

::: {.incremental}
- A **graphical degree sequence** is a sequence of numbers that *can* be the degree sequence of a simple graph.
- Not every sequence of integers is graphical.
- **Havel-Hakimi Theorem** provides a recursive algorithm to check if a sequence is graphical.
:::

---

## Average Degree

::: {.incremental}
- **Average Degree** ($\bar{k}$): The average of all node degrees in the graph.
  
  $$\bar{k} = \frac{1}{n} \sum_{i=1}^{n} k_i = \frac{2m}{n}$$
  
- It represents the expected degree of a randomly chosen node.
- It is also directly related to density: $\bar{k} = d(n-1)$.
:::

---

## Degree Variance

::: {.incremental}
- **Degree Variance** ($\sigma^2_k$): A measure of how much the node degrees vary from the average degree.
  
  $$\sigma^2_k = \frac{1}{n} \sum_{i=1}^{n} (k_i - \bar{k})^2$$
  
- A low variance indicates a more homogeneous or uniform network (like a **regular graph**).
- A high variance indicates a heterogeneous network with both low-degree nodes and high-degree "hubs" (like a **scale-free** network).
:::

---

## Regular Graphs

::: {.incremental}
- A **regular graph** is a graph where every node has the same degree.
- This means the degree variance is zero.
- These are often used as baseline models but are rare in real-world social networks.
:::

---

## Feld's Friendship Paradox

::: {.incremental}
- **Average Nearest Neighbor Degree** ($\bar{k}_{nn(i)}$):
  - The average degree of node $i$'s neighbors:
    $$\bar{k}_{nn(i)} = \frac{1}{k_i} \sum_{j \in \mathcal{N}(i)} k_j$$
- **The Friendship Paradox** (Scott Feld, 1991):
  - In almost all social networks, the average degree of a person's friends ($\bar{k}_{nn}$) is **greater** than the average degree of the people in the network ($\bar{k}$).
  - **Your friends have more friends than you do!**
:::

---

# Directed Graph Elements & Metrics

---

## Directed Degree Sets & Sequences

::: {.incremental}
- Directed graphs have two distinct degree sets (In-Degree set and Out-Degree set) and two distinct sequences.
- **Sum of Degrees**:
  - The sum of in-degrees is always equal to the sum of out-degrees, which is equal to the size $m$ (number of edges):
    
    $$\sum k^{in}_i = \sum k^{out}_i = m$$

  - Since both sum to $m$, directed graphs have only **one** average degree: $\bar{k} = m/n$.
:::
