---
title: "Graph Theory: Paths & Connectivity"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Base 5-node undirected graph (House layout pointing right)
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# Base 5-node directed graph
edges_dir <- data.frame(
  from = c("A", "A", "C", "D", "E", "D", "E", "B", "B", "E"),
  to   = c("C", "D", "E", "C", "D", "E", "B", "E", "D", "C")
)
l_dir <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
gr_dir <- tbl_graph(nodes = l_dir, edges = edges_dir, directed = TRUE)

# Disconnected graph showing components and giant component
nodes_disconnected <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 1, 1, 0, -1, -1, 3, 4, 3.5),
  y    = c(1, 0.5, -0.5, -1, -0.5, 0.5, 0.5, 0, -1)
)
edges_disconnected <- data.frame(
  from = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 7, 7, 8),
  to   = c(2, 6, 3, 5, 4, 5, 5, 6, 6, 8, 9, 9)
)
gr_disconnected <- tbl_graph(nodes = nodes_disconnected, edges = edges_disconnected, directed = FALSE)

# 5-node directed graphs for connectivity flavors (non-circle graphs)
nodes_5_conn <- data.frame(name = LETTERS[1:5], x = c(0, 1, 2, 3, 4), y = c(0, 1, 0, -1, 0))

edges_5_strong <- data.frame(
  from = c("A", "B", "C", "C", "D", "E"),
  to   = c("B", "C", "A", "D", "E", "C")
)
gr_strong <- tbl_graph(nodes = nodes_5_conn, edges = edges_5_strong, directed = TRUE)

edges_5_uni <- data.frame(
  from = c("A", "B", "C", "C", "D"),
  to   = c("B", "C", "A", "D", "E")
)
gr_uni <- tbl_graph(nodes = nodes_5_conn, edges = edges_5_uni, directed = TRUE)

edges_5_weak <- data.frame(
  from = c("A", "B", "C", "D", "D"),
  to   = c("B", "C", "A", "C", "E")
)
gr_weak <- tbl_graph(nodes = nodes_5_conn, edges = edges_5_weak, directed = TRUE)

# Local bridge graph
nodes_bridge_local <- data.frame(
  name = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  x    = c(0, 0.5, 1, 2, 2.5, 3, 0.5, 1.5, 2.5),
  y    = c(0.5, -0.5, 0, 0, -0.5, 0.5, -1.2, -1.5, -1.2)
)
edges_bridge_local <- data.frame(
  from = c(1, 1, 2, 3, 4, 4, 5, 2, 7, 8, 9),
  to   = c(2, 3, 3, 4, 5, 6, 6, 7, 8, 9, 5)
)
gr_bridge_local <- tbl_graph(nodes = nodes_bridge_local, edges = edges_bridge_local, directed = FALSE)

```

# Part 1: Indirect Connections & Paths

---

## Adjacency vs. Indirect Connections

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Adjacency** captures direct connections between nodes:
  - If $A$ is adjacent to $B$, they share an edge.
- But many network processes operate **indirectly**:
  - We must look beyond immediate contacts to the "friends of our friends."
- In graph theory, indirect connections are formalized using three core concepts:
  - **Path**, **Connectivity**, and **Reachability**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Why Indirect Connections Matter

::: {.incremental}
- Indirect pathways are the engine of diffusion and flow in a network.
- They act as channels for:
  - **Information & Gossip**: Diffusing through chains of acquaintances ("six degrees of separation").
  - **Disease & Contagion**: Spreading via intermediate, infected contacts.
  - **Influence & Support**: Reaching distant nodes through brokers.
- *Two strangers meeting at an airport who discover a common contact are connected by a path of length two.*
:::

---

## What is a Path?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **path** between $u$ and $v$ is an alternating sequence of nodes and edges that:
  - Begins with an **origin node** ($u$).
  - Ends with a **destination node** ($v$).
  - Does **not repeat any nodes** (which implies edges are not repeated either!).
- Every path is actually a **subgraph** of the original graph, connected in a single line.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Anatomy of a Path

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at our red path from $B$ to $D$:
  - Written as: $B-D = \{BE, ED\}$
- **Key Elements**:
  - **Origin Node**: $B$ (the start of the sequence)
  - **Destination Node**: $D$ (the end of the sequence)
  - **Inner Nodes**: $E$ (the nodes in between)
- Together, the origin and destination are called the **end nodes** of the path.
- The length of a path ($l$) is the number of edges: $l = 2$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Paths of Different Lengths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- The same two nodes can be connected via multiple distinct paths.
- Let's trace the pathways from $B$ to $A$:
  - $P_1 = \{BE, ED, DA\} \quad (l = 3)$
  - $P_2 = \{BE, EC, CA\} \quad (l = 3)$
  - $P_3 = \{BE, ED, DC, CA\} \quad (l = 4)$
  - $P_4 = \{BE, EC, CD, DA\} \quad (l = 4)$
- Each path represents a valid, non-repeating sequence of contacts.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Shortest Paths & Geodesics

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Shortest Path (Geodesic)**:
  - The path of minimum length connecting two nodes.
- **Geodesic Distance**:
  - The length of the shortest path between two nodes (the "degrees of separation").
- **For Nodes $B$ and $A$**:
  - The geodesic distance is $l = 3$.
  - There are **multiple shortest paths** (a tie between $P_1$ and $P_2$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Cycle?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **cycle** is a path of length 3 or larger that starts and ends on the same node.
- In other words, origin and destination are identical, but inner nodes are unique.
- **Examples**:
  - $Cycle_1 = \{ED, DC, CE\} \quad (l = 3)$
    - This is a **triangle** (closed triad), the smallest possible cycle.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Girth: Shortest Cycle

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Cycles are characterized by their length (number of edges).
- **Graph Girth**:
  - The length of the **shortest** cycle in a graph.
  - In our undirected house graph on the right, the girth is **3**.
  - This is the triangle $\{ED, DC, CE\}$, highlighted in red.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Circumference: Longest Cycle

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph Circumference**:
  - The length of the **longest** cycle in a graph.
  - In our undirected house graph on the right, the circumference is **4**.
  - This is the cycle $\{ED, DA, AC, CE\}$, highlighted in red.
  - *Notice that node $D$ and node $C$ are connected directly, but we bypass that direct edge to form the longest cycle.*
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 5) | (from == 4 & to == 5) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Walks and Trails

::: {.incremental}
- **Walk**: Any sequence of alternating nodes and edges from $u$ to $v$.
  - *No restrictions*: Nodes and edges can be repeated as much as you want!
  - e.g., $\{AD, DC, CE, ED, DC, EC, EB\}$ is a walk.
- **Trail**: A walk that does not repeat any **edges**.
  - *Repeated nodes are allowed*, but every edge in the sequence must be unique.
  - e.g., $\{CA, AD, DC, CE, ED\}$ is a trail (nodes $C$ and $D$ are visited twice, but all edges are unique).
:::

---

## Walks: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Walk**: $B \rightarrow E \rightarrow C \rightarrow E \rightarrow D$ (length 4)
- **Characteristics**:
  - Traverses the highlighted edges in red.
  - Repeats node **E** (visited twice).
  - Repeats edge **E-C** (traversed back and forth).
  - *No restrictions on repetition!*
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Trails: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Trail**: $A \rightarrow C \rightarrow D \rightarrow E \rightarrow C \rightarrow B$ (length 5)
- **Characteristics**:
  - Traverses the highlighted edges in red.
  - Repeats node **C** (visited twice).
  - **No edges are repeated**: All five edges ($AC, CD, DE, EC, EB$) are completely unique!
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## The Sequence Hierarchy

::: {.incremental}
- **Path**: A walk that does not repeat any **nodes** (which means edges cannot be repeated either).
- **The Nested Hierarchy**:
  - Every **path** is a **trail**.
  - Every **trail** is a **walk**.
  - *All paths are trails, and all trails are walks, but the reverse is not true!*
- Paths are the most restrictive, while walks are the least restricted.
:::

# Part 2: Directed Paths & Reachability

---

## Paths in Directed Graphs (Digraphs)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, paths must strictly **respect the direction of the arrows**!
- Let's trace directed paths from $A$ to $B$ in our digraph:
  - $P_1 = \{AC, CE, EB\} \quad (l = 3)$
  - $P_2 = \{AD, DE, EB\} \quad (l = 3)$
  - $P_3 = \{AC, CD, DE, EB\} \quad (l = 4)$
- But can we find any directed path from $B$ to $A$?
  - **No!** All arrows lead away from $A$ and toward $B$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Asymmetric Reachability

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed networks, **reachability is asymmetric**:
  - $A$ can reach $B$ via path $\{AD, DE, EB\}$ ($l_{AB} = 3$).
  - $B$ cannot reach $A$ ($l_{BA} = \infty$).
- This asymmetric structure is a primary driver of inequality and unilateral flows in social networks (e.g. hierarchical power).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "D") | (from_name == "D" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Mutual Reachability

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Mutual Reachability**:
  - Two nodes $u$ and $v$ are mutually reachable if there exists a directed path from $u \rightarrow v$ AND from $v \rightarrow u$.
- The two paths **do not** have to be the same length!
- **Example**: Nodes $B$ and $C$ are mutually reachable:
  - $B$ reaches $C$ via $\{BE, ED, DC\} \quad (l = 3)$
  - $C$ reaches $B$ via shortest path $\{CE, EB\} \quad (l = 2)$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Directed Cycles

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **directed cycle** is a cycle where all arrows flow in a single, continuous direction.
- **Example**: Cycle starting and ending at $C$
  - $\{CE, ED, DC\} \quad (l = 3)$
- **Directed Acyclic Graphs (DAGs)**:
  - Directed graphs that contain **no directed cycles**.
  - e.g. strict organizational command structures.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Ignoring the Arrows: Semipaths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- What if we want to trace connections in a directed graph but **ignore the arrows**?
- This is called a **semipath** (or semicycle if it is a closed loop).
- A semipath treats directed arcs as undirected edges.
- **Example**: Semipath connecting $B$ and $A$
  - $\{BE, EC, CA\}$ (visualized in red on the right).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "C") | (from_name == "A" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 3: Graph Connectivity & Components

---

## Connected vs. Disconnected Graphs

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- **Connected Graph**:
  - An undirected graph where there exists a finite path between **every** pair of nodes (left).
- **Disconnected Graph**:
  - An undirected graph where at least one pair of nodes is not connected by any path (right).
  - There is a "gap" splitting the graph into separate components.
:::
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4
gr_conn <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])

gr_disconn <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 4), to = c(2, 3, 3, 5))) %>% 
  mutate(name = LETTERS[1:5])

p1 <- ggraph(gr_conn, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Connected") + coord_cartesian(clip = "off")

p2 <- ggraph(gr_disconn, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Disconnected") + coord_cartesian(clip = "off")

p1 + p2
```
:::

::::

---

## Components and the Giant Component

:::: {.columns align=center}

::: {.column width="50%"}
::: {.incremental}
- **Component**:
  - A connected subgraph within a larger disconnected graph.
  - If a graph has more than one component, it is disconnected.
- **Giant Component**:
  - The largest connected component in a disconnected graph.
  - In our 9-node graph (right), the component of order 6 $\{A, B, C, D, E, F\}$ is the giant component.
:::
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
# Highlight giant component in red, others in steelblue
gr_disconnected_colors <- gr_disconnected |>
  activate(edges) |>
  mutate(highlight = ifelse(from <= 6 & to <= 6, "giant", "small"))

ggraph(gr_disconnected_colors, layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("giant" = "red", "small" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("giant" = 1.8, "small" = 0.8), guide = "none") +
  geom_node_point(aes(color = ifelse(name %in% c("G", "H", "I"), "small", "giant")), size = 18) +
  scale_color_manual(values = c("giant" = "tan2", "small" = "skyblue"), guide = "none") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 4: Node and Edge Connectivity (Bridges & Cutsets)

---

## Edge Cutsets

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: flex-start;"}
- **Edge Cutset**: Any set of edges whose removal disconnects the graph into two or more components.
- **Example**: The set of edges **{A-C, B-C}** (highlighted in red) is an edge cutset of size 2. 
  - Removing both edges isolates node **C** (and $\{D, E\}$) from $\{A, B\}$.
- **Edge Connectivity** ($\lambda(G)$): The size of the *minimum* edge cutset.
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
gr_cutset <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5]) %>%
  activate(edges) %>%
  mutate(highlight = ifelse((from == 1 & to == 3) | (from == 2 & to == 3), "red", "normal"))

ggraph(gr_cutset, layout = 'kk') +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("red" = "red", "normal" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("red" = 2.0, "normal" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Bridges

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: flex-start;"}
- **Bridge**: A single edge whose removal disconnects the graph (a cutset of size 1).
- **Example**: The edge **C-D** (highlighted in red) is a bridge.
  - Removing **C-D** splits the network into two separate components: $\{A, B, C\}$ and $\{D, E\}$.
- Bridges represent the most critical, yet vulnerable transmission links in a network.
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.5
#| fig-height: 4.5
gr_bridge <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5]) %>%
  activate(edges) %>%
  mutate(highlight = ifelse(from == 3 & to == 4, "red", "normal"))

ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("red" = "red", "normal" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("red" = 2.0, "normal" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## The Sociological Power of Bridges

::: {.incremental}
- Mark Granovetter's seminal paper **"The Strength of Weak Ties"** (1973):
  - Bridges represent critical pathways for novel information diffusing between separate social clusters.
- **Local Bridge**:
  - An edge that, if removed, does not necessarily disconnect the entire graph, but significantly increases the shortest path length between its end nodes.
  - Local bridges act as shortcuts that facilitate efficient, cross-group communication.
:::

---

## Local Bridge: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at the edge **C-D** highlighted in red:
  - It connects the left cluster $\{A, B, C\}$ to the right cluster $\{D, E, F\}$.
- **Is it a bridge?** No. If removed, the graph remains connected via the bottom path ($B-G-H-I-E$).
- **Is it a local bridge?** Yes! 
  - Observed distance: $d(C, D) = 1$.
  - If removed, the distance increases to:
    $$d(C, D) = 5 \quad (C-B-G-H-I-E-D)$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_bridge_local |>
  activate(edges) |>
  mutate(highlight = ifelse(from == 3 & to == 4, "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Vertex Deletion & Node Connectivity

::: {.incremental}
- We can also disconnect a graph by **removing nodes** (vertices).
- Removing a node also removes all edges incident to it—nodes "take their links with them"!
- Consequently, it is often much easier to disrupt or disconnect a network by removing key intermediate nodes than by removing edges.
- **Node Cutset (Vertex Cutset)**:
  - The minimum set of nodes whose removal disconnects the graph (or a specific pair of non-adjacent nodes $u, v$).
- Nodes in the cutset represent critical **brokerage** or "connector" positions.
:::

---

## Edge Cutset: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- An **edge cutset** is a set of edges whose removal disconnects the graph.
- In our house graph on the right, the edges **C-A** and **D-A** are highlighted in red.
- **Removing both edges** completely disconnects node $A$ from the rest of the network.
- This is a valid edge cutset of size **2**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 3 & to == 5) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Node Cutset: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **node cutset** is a set of nodes whose removal disconnects the graph.
- In our house graph on the right, node **E** is highlighted in red.
- **Removing node E** (which also removes all of its incident edges: $B-E, E-C, E-D$) completely disconnects node $B$ from the rest of the network.
- This is a minimum node cutset of size **1** (Node $E$ is a **cutpoint** or **articulation point**).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.0) +
  geom_node_point(aes(color = ifelse(name == "E", "yes", "no")), size = 20) +
  scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## K-Connected Graphs

::: {.incremental}
- A graph is **$k$-connected** (or $k$-node connected) if the minimum number of nodes required to disconnect it is $k$:
  - **1-Connected**: Removing 1 node can disconnect the graph (contains an articulation point or cutpoint).
  - **2-Connected**: Requires removing at least 2 nodes to disconnect it.
  - **3-Connected**: Requires removing at least 3 nodes to disconnect it.
- Higher $k$-connectivity indicates a more robust, cohesive, and redundant network.
:::

---

## 1-Connected Graph: Vulnerable Structure

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Before (Star Graph)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
nodes_star <- data.frame(name = c("A", "B", "C", "D"), x = c(0, 0, -0.86, 0.86), y = c(0, 1, -0.5, -0.5))
edges_star <- data.frame(from = c(1, 1, 1), to = c(2, 3, 4))
gr_star <- tbl_graph(nodes = nodes_star, edges = edges_star, directed = FALSE)

ggraph(gr_star, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**After (Removing Node A)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
nodes_star_after <- data.frame(name = c("B", "C", "D"), x = c(0, -0.86, 0.86), y = c(1, -0.5, -0.5))
gr_star_after <- tbl_graph(nodes = nodes_star_after, edges = data.frame(from = integer(0), to = integer(0)), directed = FALSE)

ggraph(gr_star_after, layout = "manual", x = x, y = y) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

- **1-Connected**: Only requires removing $k=1$ node to disconnect the graph.
- Node **A** is the **articulation point** (cutpoint). Removing it isolates nodes B, C, and D.

---

## 2-Connected Graph: Moderate Redundancy

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Before (Ring Graph)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
nodes_ring <- data.frame(name = c("A", "B", "C", "D"), x = c(0, 1, 0, -1), y = c(1, 0, -1, 0))
edges_ring <- data.frame(from = c(1, 2, 3, 4), to = c(2, 3, 4, 1))
gr_ring <- tbl_graph(nodes = nodes_ring, edges = edges_ring, directed = FALSE)

ggraph(gr_ring, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**After (Removing Nodes A and C)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
nodes_ring_after <- data.frame(name = c("B", "D"), x = c(1, -1), y = c(0, 0))
gr_ring_after <- tbl_graph(nodes = nodes_ring_after, edges = data.frame(from = integer(0), to = integer(0)), directed = FALSE)

ggraph(gr_ring_after, layout = "manual", x = x, y = y) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

- **2-Connected**: Requires removing $k=2$ nodes to disconnect the graph.
- Removing only $1$ node (e.g. A) leaves a connected path (D-C-B), so it does not disconnect the graph.
- Removing $2$ nodes (e.g. A and C) breaks the ring, completely isolating nodes **B** and **D**.

---

## 3-Connected Graph: High Cohesion

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Before (Wheel Graph)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
nodes_wheel <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(0, 0, 1, 0, -1), y = c(0, 1, 0, -1, 0))
edges_wheel <- data.frame(from = c(2, 3, 4, 5, 1, 1, 1, 1), to = c(3, 4, 5, 2, 2, 3, 4, 5))
gr_wheel <- tbl_graph(nodes = nodes_wheel, edges = edges_wheel, directed = FALSE)

ggraph(gr_wheel, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**After (Removing Nodes A, B, and D)**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
nodes_wheel_after <- data.frame(name = c("C", "E"), x = c(1, -1), y = c(0, 0))
gr_wheel_after <- tbl_graph(nodes = nodes_wheel_after, edges = data.frame(from = integer(0), to = integer(0)), directed = FALSE)

ggraph(gr_wheel_after, layout = "manual", x = x, y = y) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

- **3-Connected**: Requires removing $k=3$ nodes to disconnect the graph.
- Removing $1$ or $2$ nodes (e.g. A and B) still leaves path E-D-C connected, meaning it remains connected.
- Removing $3$ nodes (e.g. Center **A** and opposite base nodes **B** and **D**) isolates **C** and **E**.

---

## Graph Cutsets: Quantitative Cohesion

::: {.incremental}
- **Cohesion** can be directly translated to graph-theoretic cutsets:
  - **Cohesive Groups**: Have large edge-cutsets and node-cutsets. Many connections or individuals must be removed to disrupt or fragment the group.
  - **Fragmented Groups**: Have small cutsets (such as bridges or single cutpoints). They are vulnerable to rapid disruption if key facilitators are removed.
- Any network can be decomposed into hierarchical subgraphs with specific levels of $k$-connectivity.
:::

---

## Menger's Theorem

::: {.incremental}
- A foundational result in graph theory proposed by Karl Menger (1927):
  - **Menger's Theorem**: If a graph is $k$-connected, there are at least $k$ node-independent paths (paths that do not share any common inner nodes) connecting each non-adjacent pair of vertices.
- **Sociological Implication**:
  - In a 3-connected cohesive subgroup, any two non-adjacent members are connected in at least **three separate, independent ways**.
  - This ensures highly redundant, reliable information flow that cannot be blocked by a single broker!
:::

# Part 5: Directed Connectivity & Matrices

---

## The Reachability Matrix ($D_r$)

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Directed Graph**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
edges_reach <- data.frame(
  from = c("A", "A", "B", "C", "C", "E"),
  to   = c("C", "D", "E", "D", "E", "D")
)
gr_reach_ref <- tbl_graph(nodes = nodes_5, edges = edges_reach, directed = TRUE)

ggraph(gr_reach_ref, layout = "manual", x = x, y = y) +
  geom_edge_fan(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                start_cap = circle(6, 'mm'), end_cap = circle(7, 'mm'),
                color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Reachability Matrix ($D_r$)**

::: {.small}

| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 0 | 1 | 1 | 1 |
| **B** | 0 | -- | 0 | 1 | 1 |
| **C** | 0 | 0 | -- | 1 | 1 |
| **D** | 0 | 0 | 0 | -- | 0 |
| **E** | 0 | 0 | 0 | 1 | -- |

:::
:::

::::

- **Definition**: $D_r(i,j) = 1$ if node $i$ can reach node $j$ via a path of any length ($0$ otherwise).
- **Examples**: 
  - $A$ can reach $E$ via path $A \rightarrow C \rightarrow E$, so $D_r(A, E) = 1$.
  - $B$ cannot reach $A$, so $D_r(B, A) = 0$.

---

## Matrix Row & Column Patterns

The reachability matrix reveals structural node behaviors:

::: {.incremental}
- **Receiver Nodes**:
  - Have all zeros in their corresponding row of $D_r$ (cannot reach anyone, e.g. Node $D$ above).
- **Transmitter Nodes**:
  - Have all zeros in their corresponding column of $D_r$ (cannot be reached by anyone, e.g. Node $A$ or $B$ or $C$ above).
:::

---

## Flavors of Connectivity in Directed Graphs

Because of edge directionality, directed graphs have three distinct levels of connectivity:

::: {.incremental}
1. **Strongly Connected**:
   - Every pair of nodes is **mutually reachable** via directed paths.
2. **Unilaterally Connected**:
   - Every pair of nodes is connected by **at least one directed path** (from $u \rightarrow v$ OR $v \rightarrow u$, but not necessarily both).
3. **Weakly Connected**:
   - Every pair of nodes is connected by at least a **semipath** (connected if we ignore arrow directions).
:::

---

## Directed Connectivity: Visual Examples

:::: {.columns align=center}

::: {.column width="33%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
ggraph(gr_strong, layout = "manual", x = nodes_5_conn$x, y = nodes_5_conn$y) +
  geom_edge_fan(arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                start_cap = circle(5, 'mm'), end_cap = circle(5, 'mm'),
                color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Strongly Connected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))
```
:::

::: {.column width="33%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
ggraph(gr_uni, layout = "manual", x = nodes_5_conn$x, y = nodes_5_conn$y) +
  geom_edge_fan(arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                start_cap = circle(5, 'mm'), end_cap = circle(5, 'mm'),
                color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Unilaterally Connected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))
```
:::

::: {.column width="33%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
ggraph(gr_weak, layout = "manual", x = nodes_5_conn$x, y = nodes_5_conn$y) +
  geom_edge_fan(arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                start_cap = circle(5, 'mm'), end_cap = circle(5, 'mm'),
                color = "steelblue", edge_width = 1) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, color = "white", fontface="bold") +
  theme_graph() + ggtitle("Weakly Connected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))
```
:::

::::

---

## Reachability Matrices of Connectivity Flavors

:::: {.columns align=center}

::: {.column width="33%"}
### Strongly Connected

::: {.small}

| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 1 | 1 | 1 | 1 |
| **B** | 1 | -- | 1 | 1 | 1 |
| **C** | 1 | 1 | -- | 1 | 1 |
| **D** | 1 | 1 | 1 | -- | 1 |
| **E** | 1 | 1 | 1 | 1 | -- |

:::
:::

::: {.column width="33%"}
### Unilaterally Connected

::: {.small}

| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 1 | 1 | 1 | 1 |
| **B** | 1 | -- | 1 | 1 | 1 |
| **C** | 1 | 1 | -- | 1 | 1 |
| **D** | 0 | 0 | 0 | -- | 1 |
| **E** | 0 | 0 | 0 | 0 | -- |

:::
:::

::: {.column width="33%"}
### Weakly Connected

::: {.small}

| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 1 | 1 | 0 | 0 |
| **B** | 1 | -- | 1 | 0 | 0 |
| **C** | 1 | 1 | -- | 0 | 0 |
| **D** | 1 | 1 | 1 | -- | 1 |
| **E** | 0 | 0 | 0 | 0 | -- |

:::
:::

::::

---

## Reachability and Connectivity Types

The reachability matrix reveals the global connectivity type of the digraph:

::: {.incremental}
- **Strongly Connected**:
  - The reachability matrix is the **all-ones matrix** (excluding diagonal). Every node can reach every other node.
- **Unilaterally Connected**:
  - The matrix contains some zeros, but for any pair, either $D_r(i,j) = 1$ or $D_r(j,i) = 1$. There are no mutually unreachable pairs.
- **Weakly Connected**:
  - Contains at least one pair of nodes that are mutually unreachable ($D_r(i,j) = 0$ AND $D_r(j,i) = 0$, e.g., Nodes $A$ and $E$ are mutually unreachable in a weak graph).
:::

---

## Graph Connectedness ($C$): Concept

- **Graph Connectedness** ($C$) measures the extent to which a network approaches the ideal where *everyone can reach everyone else*.
- Mathematically, it is the **density of the directed reachability matrix** ($D_r$):
  $$C = \frac{\sum_{i \neq j} D_r(i,j)}{n(n-1)}$$
  - Where the numerator is the sum of all ones in the reachability matrix (excluding the diagonal).
  - The denominator $n(n-1)$ is the total possible off-diagonal cells (all possible paths).
- It ranges from $0$ (completely disconnected) to $1.0$ (strongly connected, where everyone can reach everyone).

---

## Graph Connectedness ($C$): Example

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Directed Graph**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
ggraph(gr_reach_ref, layout = "manual", x = x, y = y) +
  geom_edge_fan(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                start_cap = circle(6, 'mm'), end_cap = circle(7, 'mm'),
                color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Reachability Matrix ($D_r$)**

::: {.small}

| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 0 | 1 | 1 | 1 |
| **B** | 0 | -- | 0 | 1 | 1 |
| **C** | 0 | 0 | -- | 1 | 1 |
| **D** | 0 | 0 | 0 | -- | 0 |
| **E** | 0 | 0 | 0 | 1 | -- |

:::
:::

::::

- **Step-by-Step Calculation**:
  - Sum of ones in our reachability matrix ($D_r$): $3 \text{ (Row A)} + 2 \text{ (Row B)} + 2 \text{ (Row C)} + 0 \text{ (Row D)} + 1 \text{ (Row E)} = 8$.
  - Total possible connections: $n(n-1) = 5(4) = 20$.
  - Connectedness: $C = \frac{8}{20} = 0.40 \quad (40\%)$.

---

## The Geodesic Distance Matrix ($D_g$)

::: {.incremental}
- The **geodesic distance matrix** ($D_g$) records the length of the **shortest path** (geodesic) connecting each pair of nodes.
- **Rules**:
  - Adjacent nodes get a distance of $1$.
  - Non-adjacent nodes get the length of their shortest path.
  - Unconnected nodes (unreachable) get a distance of **infinity** ($\infty$ or `Inf`).
- In undirected graphs, $D_g$ is symmetric. In directed graphs, $D_g$ is asymmetric.
:::

---

## Undirected Geodesic Distance Matrix

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Undirected Graph**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Geodesic Distance Matrix ($D_g$)**

::: {.small}

| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 3 | 1 | 1 | 2 |
| **B** | 3 | -- | 2 | 2 | 1 |
| **C** | 1 | 2 | -- | 1 | 1 |
| **D** | 1 | 2 | 1 | -- | 1 |
| **E** | 2 | 1 | 1 | 1 | -- |

:::
:::

::::

- **Symmetric Matrix**: $D_g(i,j) = D_g(j,i)$
- **Example**: The distance $d(B, A) = 3$ (via path $B-E-C-A$ or $B-E-D-A$).

---

## Directed Geodesic Distance Matrix

:::: {.columns align=center}

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Directed Graph**

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4
ggraph(gr_dir, layout = "manual", x = x, y = y) +
  geom_edge_fan(arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                start_cap = circle(6, 'mm'), end_cap = circle(7, 'mm'),
                color = "steelblue", edge_width = 1.25) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column style="width: 50%; display: flex; flex-direction: column; justify-content: center; align-items: center;"}
**Geodesic Distance Matrix ($D_g$)**

::: {.small}

| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 3 | 1 | 1 | 2 |
| **B** | Inf | -- | 3 | 2 | 1 |
| **C** | Inf | 2 | -- | 1 | 1 |
| **D** | Inf | 3 | 1 | -- | 2 |
| **E** | Inf | 1 | 2 | 1 | -- |

:::
:::

::::

- **Asymmetric Matrix**: $D_g(i,j) \neq D_g(j,i)$
- **Example**: $d(B, A) = \infty$ (Node $A$ is unreachable from $B$ due to arrow directions).

---

## Node Eccentricity & Graph Diameter

From the geodesic distance matrix, we extract several fundamental whole-graph metrics:

::: {.incremental}
- **Node Eccentricity**: The maximum geodesic distance from a focal node to any other reachable node in the graph.
- **Graph Diameter** ($d$): The *maximum* geodesic distance across all pairs of connected nodes (longest shortest path).
  - For our undirected graph, $d = 3$.
:::

---

## Conclusion & Core Takeaways

::: {.incremental}
- **Indirect connection is the engine of diffusion**:
  - Paths trace how resources, information, and influence flow through networks.
- **Directed path flows are asymmetric**:
  - Edge directions create unequal access, where some nodes can reach everyone, but are completely unreachable themselves.
- **Matrices provide global signatures**:
  - The **reachability matrix** ($D_r$) and the **distance matrix** ($D_g$) let us calculate connectedness, diameter, and eccentricity, transforming qualitative network structures into rigorous quantitative indices.
:::
