---
title: "Graph Theory: Paths & Connectivity"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Base 5-node undirected graph (House layout pointing right)
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# Base 5-node directed graph
edges_dir <- data.frame(
  from = c("A", "A", "C", "D", "E", "D", "E", "B", "B", "E"),
  to   = c("C", "D", "E", "C", "D", "E", "B", "E", "D", "C")
)
gr_dir_base <- tbl_graph(nodes = data.frame(name = c("A", "B", "C", "D", "E")), edges = edges_dir, directed = TRUE)
# Layout coordinates for directed graph matching the slides
l_dir <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
gr_dir <- tbl_graph(nodes = l_dir, edges = edges_dir, directed = TRUE)
layout_dir <- create_layout(gr_dir, layout = "manual", x = x, y = y)

# 8-node directed graph for connectivity flavors
nodes_8 <- data.frame(name = paste0("v", 1:8))
edges_8_strong <- data.frame(
  from = c("v1", "v2", "v3", "v4", "v5", "v6", "v7", "v8"),
  to   = c("v2", "v3", "v4", "v5", "v6", "v7", "v8", "v1")
)
gr_strong <- tbl_graph(nodes = nodes_8, edges = edges_8_strong, directed = TRUE)
layout_8 <- create_layout(gr_strong, layout = "circle")
```

# Part 1: Indirect Connections & Paths

---

## Adjacency vs. Indirect Connections

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Adjacency** captures direct connections between nodes:
  - If $A$ is adjacent to $B$, they share an edge.
- But many network processes operate **indirectly**:
  - Information diffuses through a chain of acquaintances.
  - Gossip reaches you from "friends of friends."
  - Disease spreads via intermediate contacts.
- In graph theory, indirect connections are formalized using three core concepts:
  - **Path**, **Connectivity**, and **Reachability**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Path?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **path** between $u$ and $v$ is an alternating sequence of nodes and edges that:
  - Begins with an **origin node** ($u$).
  - Ends with a **destination node** ($v$).
  - Does **not repeat any nodes** (which implies edges are not repeated either!).
- **Example**: Path from $B$ to $D$
  - Written as: $B-D = \{BE, ED\}$
  - $B$ is the origin, $D$ is the destination, and $E$ is an **inner node**.
  - Length ($l$) is the number of edges: $l = 2$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Paths of Different Lengths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Multiple paths can connect the same pair of nodes.
- Let's trace paths from $B$ to $A$:
  - $P_1 = \{BE, ED, DA\} \quad (l = 3)$
  - $P_2 = \{BE, EC, CA\} \quad (l = 3)$
  - $P_3 = \{BE, ED, DC, CA\} \quad (l = 4)$
  - $P_4 = \{BE, EC, CD, DA\} \quad (l = 4)$
- **Shortest Path (Geodesic)**:
  - The path(s) of minimum length between two nodes.
  - Here, the geodesic distance between $B$ and $A$ is $l = 3$ (with two tied shortest paths: $P_1$ and $P_2$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Cycle?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **cycle** is a path of length 3 or larger that starts and ends on the same node.
- In other words, origin and destination are identical, but inner nodes are unique.
- **Examples**:
  - $Cycle_1 = \{ED, DC, CE\} \quad (l = 3)$
    - This is a **triangle** (closed triad), the smallest possible cycle.
  - $Cycle_2 = \{ED, DA, AC, CE\} \quad (l = 4)$
- Cycle structures represent closed loops where information or gossip can flow back to its source.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Walks and Trails: The Structural Hierarchy

To understand paths fully, we must compare them to other edge sequences:

::: {.incremental}
- **Walk**: An arbitrary sequence of nodes and edges connecting $u$ to $v$.
  - *No restrictions*: Nodes and edges can be repeated as much as you want!
- **Trail**: A walk that is forbidden from repeating any **edges**.
  - *Repeated nodes are allowed*, but every edge in the sequence must be unique.
- **Path**: A walk that is forbidden from repeating any **nodes**.
  - Since nodes cannot be repeated, edges cannot be repeated either. This is the most restrictive sequence.
- **Hierarchy**: All paths are trails, and all trails are walks, but the reverse is not true!
:::

# Part 2: Directed Paths & Reachability

---

## Paths in Directed Graphs (Digraphs)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, paths must strictly **respect the direction of the arrows**!
- Let's trace directed paths from $A$ to $B$ in our digraph:
  - $P_1 = \{AC, CE, EB\} \quad (l = 3)$
  - $P_2 = \{AD, DE, EB\} \quad (l = 3)$
  - $P_3 = \{AC, CD, DE, EB\} \quad (l = 4)$
- But can we find any directed path from $B$ to $A$?
  - **No!** All arrows lead away from $A$ and toward $B$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Reachability is Asymmetric

::: {.incremental}
- In directed networks, **reachability is asymmetric**:
  - $A$ can reach $B$ ($l_{AB} = 3$).
  - $B$ cannot reach $A$ ($l_{BA} = \infty$).
- **Mutual Reachability**:
  - Two nodes $u$ and $v$ are mutually reachable if there exists a directed path from $u$ to $v$ AND a directed path from $v$ to $u$.
  - These two paths **do not** have to be the same length!
  - **Example**: Nodes $B$ and $C$ are mutually reachable:
    - $B$ reaches $C$ via $\{BE, ED, DC\} \quad (l = 3)$.
    - $C$ reaches $B$ via shortest path $\{CE, EB\} \quad (l = 2)$.
:::

---

## Directed Cycles & Acyclic Graphs (DAGs)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **directed cycle** is a cycle where all arrows flow in a single, continuous direction.
- **Example**: Cycle starting and ending at $C$
  - $\{CE, ED, DC\} \quad (l = 3)$
- **Directed Acyclic Graphs (DAGs)**:
  - Directed graphs that contain **no directed cycles**.
  - Directed trees are always DAGs.
  - Anti-symmetric relations (like "is boss of") structurally forbid cycles to maintain consistent hierarchy.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Ignoring the Arrows: Semipaths

::: {.incremental}
- What if we want to trace connections in a directed graph but **ignore the arrows**?
- This is called a **semipath** (or semicycle if it is a closed loop).
- A semipath treats directed arcs as undirected edges.
- **Example**: Semipath connecting $B$ and $A$
  - $\{BE, EC, CA\}$
- Semipaths are critical for defining the baseline structural continuity of directed networks.
:::

# Part 3: Graph Connectivity & Components

---

## Connected vs. Disconnected Graphs

::: {.incremental}
- In an undirected graph:
  - A graph is **connected** if there exists a finite path between *every* pair of nodes.
  - A graph is **disconnected** if at least one pair of nodes is not connected by any path.
- **Component**:
  - A connected subgraph within a larger disconnected graph.
  - If a graph has more than one component, it is disconnected by definition.
- **Giant Component**:
  - The largest connected component in a disconnected graph (contains the maximum number of nodes).
:::

---

## Edge cuts & Bridges

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- We can disconnect a graph by removing edges.
- **Edge Cutset**: Any set of edges whose removal disconnects the graph.
- **Bridge**:
  - A single edge whose removal disconnects the graph, splitting it into separate components.
  - Bridges are highly important; they represent critical linkages connecting otherwise isolated groups.
  - Bridges are the ultimate bottleneck for network flow.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_bridge <- create_empty(5, directed = FALSE) %>% 
  bind_edges(data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 3, 4, 5))) %>% 
  mutate(name = LETTERS[1:5])
ggraph(gr_bridge, layout = 'kk') +
  geom_edge_link(aes(color = ifelse(from == 3 & to == 4, "red", "normal"),
                     edge_width = ifelse(from == 3 & to == 4, "red", "normal"))) +
  scale_edge_color_manual(values = c("red" = "red", "normal" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("red" = 2.0, "normal" = 1.0), guide = "none") +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Flavors of Connectivity in Directed Graphs

Because of edge directionality, directed graphs have three distinct levels of connectivity:

::: {.incremental}
1. **Strongly Connected**:
   - Every pair of nodes is **mutually reachable** via directed paths.
2. **Unilaterally Connected**:
   - Every pair of nodes is connected by **at least one directed path** (from $u \rightarrow v$ OR $v \rightarrow u$, but not necessarily both).
3. **Weakly Connected**:
   - Every pair of nodes is connected by at least a **semipath** (connected if we ignore arrow directions).
:::

# Part 4: Reachability & Distance Matrices

---

## The Reachability Matrix ($D_r$)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- The **reachability matrix** ($D_r$) of a graph records whether each pair of nodes can reach each other:
  - $D_r(i,j) = 1$ if node $i$ can reach node $j$ via a path of any length.
  - $D_r(i,j) = 0$ otherwise.
  - Diagonal cells are ignored (`--`).
- **Matrix Patterns**:
  - **Receiver nodes** have all zeros in their corresponding row (cannot reach anyone).
  - **Transmitter nodes** have all zeros in their corresponding column (cannot be reached).
:::
:::

::: {.column width="40%"}
| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 0 | 1 | 1 | 1 |
| **B** | 0 | -- | 0 | 1 | 1 |
| **C** | 0 | 0 | -- | 1 | 1 |
| **D** | 0 | 0 | 0 | -- | 0 |
| **E** | 0 | 0 | 0 | 1 | -- |

: Reachability Matrix for our Directed Graph {#tbl-reach}
:::

::::

---

## Reachability and Connectivity Types

The reachability matrix reveals the global connectivity type of the digraph:

::: {.incremental}
- **Strongly Connected**:
  - The reachability matrix is the **all-ones matrix** (excluding diagonal). Every node can reach every other node.
- **Unilaterally Connected**:
  - The matrix contains some zeros, but for any pair, either $D_r(i,j) = 1$ or $D_r(j,i) = 1$. There are no mutually unreachable pairs.
- **Weakly Connected**:
  - Contains at least one pair of nodes that are mutually unreachable ($D_r(i,j) = 0$ AND $D_r(j,i) = 0$).
:::

---

## Graph Connectedness ($C$)

::: {.incremental}
- **Graph Connectedness** ($C$) measures the extent to which a network approaches the ideal where *everyone can reach everyone else*.
- Mathematically, it is the **density of the directed reachability matrix**:
  $$C = \frac{\sum_{i \neq j} D_r(i,j)}{n(n-1)}$$
- **Example Calculation**:
  - Sum of ones in our reachability matrix ($D_r$): $1 + 1 + 1 + 1 + 1 + 1 + 1 + 1 = 8$ (Wait! Let's sum the table ones: row A has 3 ones, row B has 2 ones, row C has 2 ones, row E has 1 one. Sum is 8.)
  - Total possible connections: $n(n-1) = 5(4) = 20$.
  - Connectedness: $C = \frac{8}{20} = 0.40 \quad (40\%)$.
:::

---

## The Geodesic Distance Matrix ($D_g$)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- The **geodesic distance matrix** ($D_g$) records the length of the **shortest path** (geodesic) connecting each pair of nodes.
- **Rules**:
  - Adjacent nodes get a distance of $1$.
  - Non-adjacent nodes get the length of their shortest path.
  - Unconnected nodes (unreachable) get a distance of **infinity** ($\infty$ or `Inf`).
- In undirected graphs, $D_g$ is symmetric.
- In directed graphs, $D_g$ is asymmetric.
:::
:::

::: {.column width="40%"}
| | A | B | C | D | E |
|---|---|---|---|---|---|
| **A** | -- | 3 | 1 | 1 | 2 |
| **B** | Inf | -- | 3 | 2 | 1 |
| **C** | Inf | 2 | -- | 1 | 1 |
| **D** | Inf | 3 | 1 | -- | 2 |
| **E** | Inf | 1 | 2 | 1 | -- |

: Geodesic Distance Matrix for our Digraph {#tbl-dist}
:::

::::

---

## Node Eccentricity, Diameter & Girth

From the geodesic distance matrix, we extract several fundamental whole-graph metrics:

::: {.incremental}
- **Node Eccentricity**: The maximum geodesic distance from a focal node to any other reachable node in the graph.
- **Graph Diameter** ($d$): The *maximum* geodesic distance across all pairs of connected nodes (longest shortest path).
  - For our undirected graph, $d = 3$.
- **Graph Girth**: The length of the *shortest* cycle in a graph (smallest loop).
  - For our undirected graph, the girth is $3$ (the triangle $\{ED, DC, CE\}$).
:::

---

## Conclusion & Core Takeaways

::: {.incremental}
- **Indirect connection is the engine of diffusion**:
  - Paths trace how resources, information, and influence flow through networks.
- **Directed path flows are asymmetric**:
  - Edge directions create unequal access, where some nodes can reach everyone, but are completely unreachable themselves.
- **Matrices provide global signatures**:
  - The **reachability matrix** ($D_r$) and the **distance matrix** ($D_g$) let us calculate connectedness, diameter, and eccentricity, transforming qualitative network structures into rigorous quantitative indices.
:::

---

## References {.unnumbered}
