---
title: "Graph Theory: Paths, Walks & Directed Connections"
subtitle: "The Basics of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Base 5-node undirected graph (House layout pointing right)
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# Base 5-node directed graph
edges_dir <- data.frame(
  from = c("A", "A", "C", "D", "E", "D", "E", "B", "B", "E", "C"),
  to   = c("C", "D", "E", "C", "D", "E", "B", "E", "D", "C", "D")
)
l_dir <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
gr_dir <- tbl_graph(nodes = l_dir, edges = edges_dir, directed = TRUE)

# Symmetrical 5-node undirected graph for Hamiltonian Path & Cycle comparisons
nodes_ham <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(0, -1, -0.6, 0.6, 1), y = c(1, 0.2, -1, -1, 0.2))
edges_ham <- data.frame(from = c(1, 2, 3, 4, 5, 2), to = c(2, 3, 4, 5, 1, 5))
gr_ham <- tbl_graph(nodes = nodes_ham, edges = edges_ham, directed = FALSE)
```

# Part 1: Indirect Connections & Paths

---

## Adjacency vs. Indirect Connections

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Adjacency** captures direct connections between nodes:
  - If $A$ is adjacent to $B$, they share an edge.
- But many network processes operate **indirectly**:
  - We must look beyond immediate contacts to the "friends of our friends."
  - **Diffusion Channels**: Indirect connections act as channels for gossip, information transmission, disease contagion, or social influence.
- In graph theory, indirect connections are formalized using three core concepts:
  - **Path**, **Connectivity**, and **Reachability**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Path?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **path** between $u$ and $v$ is an alternating sequence of nodes and edges that:
  - Begins with an **origin node** ($u$).
  - Ends with a **destination node** ($v$).
  - Does **not repeat any nodes** (which implies edges are not repeated either!).
- **Anatomy of a Path** (e.g. Red path from $B$ to $D$ on the right):
  - Written as: $B-D = \{BE, ED\}$ (length $l = 2$).
  - **Origin**: $B$, **Destination**: $D$ (together called the **end nodes**).
  - **Inner Nodes**: $E$ (the nodes in between).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Paths & Geodesics

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Multiple distinct paths can connect the same pair of nodes.
- Let's trace paths from $B$ to $A$:
  - $P_1 = \{BE, ED, DA\} \quad (l = 3)$
  - $P_2 = \{BE, EC, CA\} \quad (l = 3)$
  - $P_3 = \{BE, ED, DC, CA\} \quad (l = 4)$
  - $P_4 = \{BE, EC, CD, DA\} \quad (l = 4)$
- **Shortest Path (Geodesic)**:
  - The path of minimum length between two nodes.
  - **Geodesic Distance**: The length of the geodesic path (here, geodesic distance $d(B, A) = 3$, with tied geodesics $P_1$ and $P_2$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Multiple Shortest Paths (Geodesic Ties)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In real networks, **geodesic distances are often tied**:
  - There can be multiple distinct shortest paths of equal minimum length connecting the exact same pair of nodes.
- **Example**: Shortest paths from $B$ to $A$ in our undirected graph:
  - Both paths have a geodesic distance of $d(B, A) = 3$.
  - **Geodesic Path 1**: $B \rightarrow E \rightarrow D \rightarrow A$ (Bottom)
  - Let's visualize this path highlighted in red on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
p1 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Geodesic 1 (B-E-D-A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
p1
```
:::

::::

---

## Geodesic Ties: Redundancy & Reliability

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at the second geodesic path from $B$ to $A$:
  - **Geodesic Path 2**: $B \rightarrow E \rightarrow C \rightarrow A$ (Top)
  - Let's visualize this path highlighted in red on the right.
- **Redundancy and Reliability**:
  - Having multiple geodesics of equal minimum length provides alternative routes of maximum efficiency.
  - **Structural Robustness**: If a node or edge on one path fails (e.g. node $D$), communication is completely unaffected because Path 2 remains open!
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
p2 <- gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Geodesic 2 (B-E-C-A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
p2
```
:::

::::

---

## Sociological Significance of Geodesics

::: {.incremental}
- Why do social scientists care so much about shortest paths (geodesics) in real networks?
- **1. Speed and Bandwidth of Transmission**:
  - Information, trust, and influence decay as they travel along long paths.
  - Geodesic paths represent the channels through which diffusion (of gossip, innovation, or virus) happens **fastest** and with **least loss**.
- **2. Power, Brokerage, and Betweenness**:
  - Actors who lie on the geodesics of *many other pairs* of nodes in the network play a major role as **information brokers** or gatekeepers.
  - They control information flow, filter messages, and can extract rents or gain social influence (this is the basis of **Betweenness Centrality**).
- **3. Search and the "Small World" Phenomenon**:
  - Stanley Milgram's classic "six degrees of separation" experiment showed that human social networks have surprisingly short geodesic distances.
  - Even in massive networks of millions of people, any two individuals are connected by incredibly short paths of friendship!
:::

---


## Node Eccentricity

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: The **eccentricity** of a node is the maximum geodesic distance from that node to any other node in the network.
  - It measures the "worst-case" distance to reach any other actor.
- **Example 1: Node E (Central Hub)**:
  - Distances from $E$: $d(E, B) = 1$, $d(E, C) = 1$, $d(E, D) = 1$, $d(E, A) = 2$.
  - The maximum distance is $2$ (to node $A$).
  - **Eccentricity of $E$ = 2**.
- **Example 2: Node B (Peripheral Leaf)**:
  - Distances from $B$: $d(B, E) = 1$, $d(B, C) = 2$, $d(B, D) = 2$, $d(B, A) = 3$.
  - The maximum distance is $3$ (to node $A$).
  - **Eccentricity of $B$ = 3**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Radius and Diameter

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Eccentricity describes a single node; two measures summarize the eccentricities of the whole graph:
- **Radius** $\mathrm{rad}(G)$: The **minimum** eccentricity across all nodes.
  - Here, $\mathrm{rad}(G) = 2$, attained by $C$, $D$, and $E$ — the **center** of the network.
- **Diameter** $\mathrm{diam}(G)$: The **maximum** eccentricity across all nodes.
  - Here, $\mathrm{diam}(G) = 3$, attained by $A$ and $B$ — the **periphery** of the network.
- The red path on the right is a **diametral path**: a geodesic whose length equals the diameter ($B \rightarrow E \rightarrow D \rightarrow A$, $l = 3$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## What is a Cycle?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **cycle** is a path of length 3 or larger that starts and ends on the same node.
- **Key Concepts**:
  - Origin and destination are identical (the same vertex).
  - All inner nodes and all edges must be completely unique.
  - Cycles are the fundamental building blocks of network redundancy and feedback loops.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.2) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Visualizing a Simple Cycle

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Let's look at a simple **cycle** highlighted in our graph:
  - **Cycle**: $C \rightarrow D \rightarrow A \rightarrow C$.
  - It has a length of **3** (consisting of 3 edges).
- **Verification of Properties**:
  - Origin and destination are identical (both are node $C$).
  - All intermediate nodes ($D$, $A$) are unique.
  - All traversed edges are unique.
  - This closed loop creates a fundamental feedback pathway.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 3 & to == 5) | (from == 4 & to == 5) | (from == 3 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Girth

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: The **girth** of a graph is the length of its **shortest cycle**.
- **Properties**:
  - If a graph contains no cycles at all (like a tree graph), its girth is defined as **infinity** ($\infty$).
  - For our 5-node undirected graph on the right, the shortest cycle is the triangle:
    $$E \rightarrow D \rightarrow C \rightarrow E$$
  - Thus, the **girth** of this graph is **3**.
- **Sociological Note**: The girth indexes how tight the smallest closed circle of actors is — a girth of 3 means friends-of-friends are directly tied, the signature of strong local closure (Simmelian ties).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 4) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Graph Circumference

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: The **circumference** of a graph is the length of its **longest cycle**.
- **Properties**:
  - If a graph is acyclic, its circumference is defined as **zero** or undefined.
  - For our 5-node undirected graph on the right, the longest cycle is the four-loop:
    $$E \rightarrow D \rightarrow A \rightarrow C \rightarrow E$$
  - Thus, the **circumference** of this graph is **4**.
- **Sociological Note**: The circumference captures the longest chain of distinct actors that closes a feedback loop — a gauge of how far redundancy can stretch across a network.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 2 & to == 3) | (from == 3 & to == 5) | (from == 4 & to == 5) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Special Paths: Hamiltonian Paths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Hamiltonian Path**: A path that visits **every single node** in the graph exactly once.
- **Our Example Graph**: Shown on the right, it has 5 nodes $\{A, B, C, D, E\}$.
- **A Hamiltonian Path**: Let's trace one in red:
  $$A \rightarrow B \rightarrow C \rightarrow D \rightarrow E$$
  - Every node is visited exactly once.
  - The diagonal edge $B-E$ remains untraversed.
- **Sociological Note**: Hamiltonian structure matters whenever a chain of introductions must reach every actor exactly once — the logic behind round-robin designs and snowball-sampling routes that avoid revisits.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_ham |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 4 & to == 5), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Special Paths: Hamiltonian Cycles

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Hamiltonian Cycle**: A cycle that visits **every single node** in the graph exactly once and returns to the origin.
- **Our Hamiltonian Cycle**: Let's trace it in red on the right:
  $$A \rightarrow B \rightarrow C \rightarrow D \rightarrow E \rightarrow A$$
  - This cycle forms the complete outer pentagonal ring.
  - Visits all 5 nodes exactly once and closes the loop!
  - **Degree Condition**: Every node has degree $\ge 2$ — a necessary condition for a Hamiltonian cycle (though not, by itself, sufficient).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_ham |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 3 & to == 4) | (from == 4 & to == 5) | (from == 1 & to == 5) | (from == 5 & to == 1), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

# Part 2: Walks, Trails, and Directed Connections

---

## Walks, Trails, and Paths

::: {.incremental}
- **Walk**: Any sequence of alternating nodes and edges connecting $u$ to $v$.
  - *No restrictions*: Nodes and edges can be repeated as much as you want!
  - e.g., $\{B \rightarrow E \rightarrow C \rightarrow E \rightarrow D\}$ (traverses $E$ and edge $E-C$ twice).
- **Trail**: A walk that does not repeat any **edges**.
  - *Repeated nodes are allowed*, but every edge in the sequence must be unique.
  - e.g., $\{B \rightarrow E \rightarrow D \rightarrow A \rightarrow C \rightarrow D\}$ (repeats node $D$ but all edges are unique).
- **Path**: A walk that does not repeat any **nodes** (which means edges cannot be repeated either).
- **The Nested Hierarchy**: All paths are trails, and all trails are walks, but the reverse is not true:
  $$Paths \subset Trails \subset Walks$$
:::

---

## Walks: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Walk (B-E-C-E-D)**:
  - We start at node $B$, walk to $E$, then $C$, then *back* to $E$, and finally to $D$.
  - **Repeated Node**: Node $E$ is visited twice in this sequence.
  - **Repeated Edge**: Edge $E-C$ (and $C-E$) is traversed twice.
  - Highlighted in red on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 3) | (from == 2 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Walk (Repeats Node E and Edge E-C)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
```
:::

::::

---

## Trails: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Trail (B-E-D-A-C-D)**:
  - We start at $B \rightarrow E \rightarrow D \rightarrow A \rightarrow C \rightarrow D$.
  - **Repeated Node**: Node $D$ is visited twice.
  - **Repeated Edges**: **None!** Every edge in the sequence is completely unique.
  - Highlighted in red on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_undir |>
  activate(edges) |>
  mutate(highlight = ifelse((from == 1 & to == 2) | (from == 2 & to == 4) | (from == 4 & to == 5) | (from == 3 & to == 5) | (from == 3 & to == 4), "yes", "no")) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(color = highlight, edge_width = highlight)) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Trail (Repeats Node D, Edges Unique)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face="bold", hjust=0.5))
```
:::

::::

---

## Paths in Directed Graphs (Digraphs)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, paths must strictly **respect the direction of the arrows**!
- Let's trace directed paths from $A$ to $B$ in our digraph:
  - $P_1 = \{AC, CE, EB\} \quad (l = 3)$
  - $P_2 = \{AD, DE, EB\} \quad (l = 3)$
  - $P_3 = \{AC, CD, DE, EB\} \quad (l = 4)$
- But can we find any directed path from $B$ to $A$?
  - **No!** All arrows lead away from $A$ and toward $B$.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Directed Reachability

::: {.incremental}
- **Asymmetric Reachability**:
  - Node $A$ can reach $B$ via directed path $\{AD, DE, EB\}$ ($l_{AB} = 3$).
  - But node $B$ **cannot** reach $A$ because we cannot travel against the arrows ($l_{BA} = \infty$).
- **Mutual Reachability**:
  - Two nodes $u$ and $v$ are mutually reachable if there exists a directed path from $u \rightarrow v$ AND from $v \rightarrow u$.
  - These two paths **do not** have to be the same length!
  - **Example**: Nodes $B$ and $C$ are mutually reachable:
    - $B$ reaches $C$ via $\{BE, ED, DC\} \quad (l = 3)$
    - $C$ reaches $B$ via shortest path $\{CE, EB\} \quad (l = 2)$
:::

---

## Directed Reachability: Visual Examples

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
p1 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "A" & to_name == "D") | (from_name == "D" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Asymmetric (A reaches B, B cannot reach A)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust=0.5))

p2 <- gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C") | (from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "B"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(5, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Mutual (B and C reach each other via paths of length 3 & 2)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 11, face="bold", hjust=0.5))

p1 + p2
```

---

## Directed Cycles & DAGs

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **directed cycle** is a cycle where all arrows flow in a single, continuous direction.
  - **Example**: Cycle starting and ending at $C$:
    - $\{CE, ED, DC\} \quad (l = 3)$
- **Directed Acyclic Graphs (DAGs)**:
  - Directed graphs that contain **no directed cycles**.
  - e.g. strict organizational command structures, directed trees.
  - Anti-symmetric relations (like "is boss of") structurally forbid cycles to maintain consistent hierarchy.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "C" & to_name == "E") | (from_name == "E" & to_name == "D") | (from_name == "D" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Ignoring the Arrows: Semipaths

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- What if we want to trace connections in a directed graph but **ignore the arrows**?
- This is called a **semipath** (or semicycle if it is a closed loop).
- A semipath treats directed arcs as undirected edges.
- **Example**: Semipath connecting $B$ and $A$:
  - $\{BE, EC, CA\}$ (highlighted in red on the right).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dir |>
  activate(edges) |>
  mutate(
    from_name = .N()$name[from],
    to_name = .N()$name[to],
    highlight = ifelse((from_name == "B" & to_name == "E") | (from_name == "E" & to_name == "C") | (from_name == "A" & to_name == "C"), "yes", "no")
  ) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = highlight, width = highlight),
                     arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.5, 'mm')) +
  scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
  scale_edge_width_manual(values = c("yes" = 2.0, "no" = 1.0), guide = "none") +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Wrap-Up: Paths & Directed Connections

::: {.incremental}
- **Paths, Geodesics & Cycles**: Indirect connections, shortest paths, and why geodesics matter for diffusion, brokerage, and the small world.
- **Walks, Trails & Directed Ties**: The $Paths \subset Trails \subset Walks$ hierarchy, plus reachability, directed cycles, and semipaths.
- **Next up**: What happens when networks come apart — components, cutsets, and robustness.
:::
