---
title: "Matrix Algebra in Social Networks"
subtitle: "Basic Operations, Conformability, and Matrix Multiplication"
format:
  revealjs:
    incremental: false
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(expm)
library(patchwork)

# Set seed for reproducible layout of 6-node graphs
set.seed(444)

# Define exact 6x6 symmetric adjacency matrices from the Google Slides (Slide 10)
# Parties With (P)
P <- matrix(c(
  0, 1, 1, 0, 1, 0,
  1, 0, 1, 1, 0, 0,
  1, 1, 0, 1, 1, 1,
  0, 1, 1, 0, 1, 0,
  1, 0, 1, 1, 0, 1,
  0, 0, 1, 0, 1, 0
), nrow = 6, byrow = TRUE, dimnames = list(LETTERS[1:6], LETTERS[1:6]))

# Plays Games With (G)
G <- matrix(c(
  0, 1, 0, 0, 1, 1,
  1, 0, 1, 0, 0, 0,
  0, 1, 0, 1, 0, 0,
  0, 0, 1, 0, 1, 1,
  1, 0, 0, 1, 0, 1,
  1, 0, 0, 1, 1, 0
), nrow = 6, byrow = TRUE, dimnames = list(LETTERS[1:6], LETTERS[1:6]))

# Create tidygraph objects from these matrices
gr_p <- as_tbl_graph(graph_from_adjacency_matrix(P, mode = "undirected"))
gr_g <- as_tbl_graph(graph_from_adjacency_matrix(G, mode = "undirected"))

# Define a directed adjacency matrix A_dir for the Transpose and Dyad Census (Slide 21 / 38)
A_dir <- matrix(c(
  0, 0, 1, 0, 1, 0,
  1, 0, 1, 1, 0, 0,
  1, 0, 0, 1, 1, 0,
  0, 0, 0, 0, 0, 0,
  0, 0, 0, 1, 0, 1,
  0, 0, 0, 1, 1, 0
), nrow = 6, byrow = TRUE, dimnames = list(LETTERS[1:6], LETTERS[1:6]))

gr_adir <- as_tbl_graph(graph_from_adjacency_matrix(A_dir, mode = "directed"))

# Define 5-node undirected matrix for Matrix Powers (Slide 109)
A_undir_5 <- matrix(c(
  0, 1, 0, 0, 0,
  1, 0, 1, 0, 0,
  0, 1, 0, 1, 1,
  0, 0, 1, 0, 1,
  0, 0, 1, 1, 0
), nrow = 5, byrow = TRUE, dimnames = list(letters[1:5], letters[1:5]))

gr_undir_5 <- as_tbl_graph(graph_from_adjacency_matrix(A_undir_5, mode = "undirected"))

# Circular layout coordinates for 6 nodes
l_coords_6 <- matrix(c(
  0, 1,
  0.866, 0.5,
  0.866, -0.5,
  0, -1,
  -0.866, -0.5,
  -0.866, 0.5
), ncol = 2, byrow = TRUE)

# Manual coordinates for 5-node house-kite graph (Slide 109)
# Node order: a, b, c, d, e
l_coords_5 <- matrix(c(
  -1.5, 0,    # a
  -0.5, 0,    # b
  0.5, 0,     # c
  1.5, 0.7,   # d
  1.5, -0.7   # e
), ncol = 2, byrow = TRUE)

# Plots for Parties With (P) and Plays Games With (G)
p_p <- ggraph(gr_p, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 2.0) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  labs(title = "Parties With (P)") +
  theme_graph() + theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))

p_g <- ggraph(gr_g, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "firebrick", width = 2.0) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  labs(title = "Plays Games With (G)") +
  theme_graph() + theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
```

# Part 1: Basic Operations & Network Overlap

---

## Combining Social Relations

- In social network analysis, we often record **multiple types of relationships** for the same set of nodes (e.g., *Parties With* vs. *Plays Games With*).
- How do we combine, subtract, or intersect these overlapping structures?
- We use **Basic Matrix Operations** performed cell-by-cell on adjacency matrices of the same dimensions:
  1. **Matrix Addition ($\mathbf{P} + \mathbf{G}$)**: Measures overall relationship volume (uniplex vs. multiplex).
  2. **Subtracting from a Matrix ($\mathbf{1} - \mathbf{P}$)**: Flips ties to find the **graph complement** (*Does Not Party With*).
  3. **Hadamard Product ($\mathbf{P} \circ \mathbf{G}$)**: Extracts multiplex overlap (*Parties and Plays Games With*).

---

## Adjacency Matrices: $\mathbf{P}$ and $\mathbf{G}$

::: {.columns}
::: {.column width="50%"}
### Parties With ($\mathbf{P}$)
```{r}
#| echo: false
options(knitr.kable.NA = '--')
P_display <- P
diag(P_display) <- NA
kbl(P_display, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_p
```
</center>
:::

::: {.column width="50%"}
### Plays Games With ($\mathbf{G}$)
```{r}
#| echo: false
G_display <- G
diag(G_display) <- NA
kbl(G_display, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_g
```
</center>
:::
:::

---

## Step-by-Step Addition ($\mathbf{P} + \mathbf{G}$)

- **Addition ($s_{ij} = p_{ij} + g_{ij}$)** adds corresponding cells of each matrix:
- Let's compute two example cells:
  - $s_{\text{A, B}} = p_{\text{A, B}} + g_{\text{A, B}} = 1 + 1 = \mathbf{2}$ (Multiplex)
  - $s_{\text{A, C}} = p_{\text{A, C}} + g_{\text{A, C}} = 1 + 0 = \mathbf{1}$ (Uniplex)

. . .

$\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix} + \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix} = \begin{pmatrix} 0 & 2 \\ 2 & 0 \end{pmatrix}$

- **Sociological Interpretation**: 
  - Value **2** represents a **multiplex tie** (connected by *both* relationships).
  - Value **1** represents a **uniplex tie** (connected by *exactly one* relationship).
  - Value **0** represents **no relationship**.

---

## Visualizing Matrix Addition ($\mathbf{P} + \mathbf{G}$)

::: {.columns}
::: {.column width="50%"}
### Parties or Plays Games With or Both
```{r}
#| echo: false
P_plus_G <- P + G
diag(P_plus_G) <- NA
kbl(P_plus_G, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- **1 (Blue)** = Uniplex bond.
- **2 (Purple)** = Multiplex bond.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_sum <- graph_from_adjacency_matrix(P + G, mode = "undirected", weighted = TRUE)
ggraph(gr_sum, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(aes(color = as.factor(weight), edge_width = weight*1.5)) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  scale_edge_color_manual(values = c("1" = "steelblue", "2" = "purple")) +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
</center>
:::
:::

---

## Subtracting a Number From a Matrix ($\mathbf{1} - \mathbf{P}$)

- We can subtract an adjacency matrix from 1 (excluding the diagonal) to find the **graph complement**.
- Subtracting from 1 flips 1s into 0s and 0s into 1s:
  - $c_{ij} = 1 - p_{ij}$

. . .

$\mathbf{1} - \mathbf{P} = 1 - \begin{pmatrix} 0 & 1 & 1 \\ 1 & 0 & 1 \\ 1 & 1 & 0 \end{pmatrix} = \begin{pmatrix} 1 & 0 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 1 \end{pmatrix}$

- **Sociological Meaning**:
  - The resulting matrix records the **absence** of the original relationship.
  - The graph corresponding to this matrix is called the **complement** of the original graph (*Does Not Party With*).

---

## Visualizing Graph Complement ($\mathbf{1} - \mathbf{P}$)

::: {.columns}
::: {.column width="50%"}
### Does Not Party With
```{r}
#| echo: false
P_comp <- 1 - P
diag(P_comp) <- NA
kbl(P_comp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Holds $1$s where there was no original tie.
- Diagonal remains blocked.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_comp <- graph_from_adjacency_matrix(1 - P, mode = "undirected", diag = FALSE)
ggraph(gr_comp, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "forestgreen", width = 2.0) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
</center>
:::
:::

---

## Step-by-Step Hadamard Product ($\mathbf{P} \circ \mathbf{G}$)

- **Hadamard / Element-Wise Product ($c_{ij} = p_{ij} \times g_{ij}$)** represents **relational intersection**:
- Let's compute two example cells:
  - $c_{\text{A, B}} = p_{\text{A, B}} \times g_{\text{A, B}} = 1 \times 1 = \mathbf{1}$ (Both active)
  - $c_{\text{A, C}} = p_{\text{A, C}} \times g_{\text{A, C}} = 1 \times 0 = \mathbf{0}$ (Only one active)

. . .

$\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix} \circ \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix} = \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$

- **Sociological Meaning**:
  - Cell value is **1** *only if* both $p_{ij} = 1$ and $g_{ij} = 1$.
  - It extracts **multiplex ties** (the overlap/intersection) and erases all uniplex ties.

---

## Visualizing the Hadamard Product ($\mathbf{P} \circ \mathbf{G}$)

::: {.columns}
::: {.column width="50%"}
### Parties and Plays Games With
```{r}
#| echo: false
P_dot_G <- P * G
diag(P_dot_G) <- NA
kbl(P_dot_G, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Retains only the shared ties.
- All uniplex ties are erased, leaving a sparse overlap network.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_dot <- graph_from_adjacency_matrix(P * G, mode = "undirected")
ggraph(gr_dot, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "purple", width = 2.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
</center>
:::
:::

---

# Part 2: The Matrix Transpose & Directed Graphs

---

## The Matrix Transpose ($\mathbf{A}^T$)

- The **Transpose** of a matrix $\mathbf{A}$ is the same matrix with its rows and columns flipped:
  - Written as $\mathbf{A}^T$ or $\mathbf{A}'$.
  - **The Rule**: Row $i$ of the original matrix becomes Column $i$ of the transpose.
  - **Formally**: $a^T_{ij} = a_{ji}$ for all cells.

. . .

- **Symmetric vs. Asymmetric Adjacency Matrices**:
  - For **undirected graphs**, relationships are reciprocal by definition, so the matrix is symmetric ($\mathbf{A}^T = \mathbf{A}$).
  - For **directed graphs (digraphs)**, ties can be one-way, resulting in an asymmetric matrix ($\mathbf{A}^T \neq \mathbf{A}$).

---

## Visualizing the Matrix Transpose

::: {.columns}
::: {.column width="50%"}
### Original Adjacency Matrix ($\mathbf{A}$)
```{r}
#| echo: false
A_disp <- A_dir
diag(A_disp) <- NA
kbl(A_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
# Create directed graph layout for A
p_adir <- ggraph(gr_adir, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 1.5,
                 arrow = arrow(length = unit(4, 'mm')),
                 end_cap = circle(6, 'mm')) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
p_adir
```
</center>
:::

::: {.column width="50%"}
### Transposed Adjacency Matrix ($\mathbf{A}^T$)
```{r}
#| echo: false
AT_disp <- t(A_dir)
diag(AT_disp) <- NA
kbl(AT_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Row 1 of $\mathbf{A}$ `(0, 0, 1, 0, 1, 0)` is Column 1 of $\mathbf{A}^T$.
- Note how the direction of every arrow is reversed in the transpose graph!
:::
:::

---

# Part 3: Directed Dyads & The Dyad Census

---

## Dyads in Directed Graphs

- In a directed graph, any pair of nodes (a **dyad**) can have one of three configurations:
  1. **Mutual (M)**: Both directed edges are present ($i \leftrightarrow j$).
  2. **Asymmetric (A)**: Only one directed edge is present ($i \rightarrow j$ or $j \rightarrow i$).
  3. **Null (N)**: Neither directed edge is present (disconnected).
- An easy way to remember these three types of dyads is the mnemonic **MAN** (Mutual, Asymmetric, Null).

. . .

<center>
```{r}
#| echo: false
#| fig-height: 2.2
#| fig-width: 8
# Create simple 3 dyads representation
gr_m <- tbl_graph(nodes = data.frame(name = c("B", "E")), edges = data.frame(from = c(1, 2), to = c(2, 1)), directed = TRUE)
gr_a <- tbl_graph(nodes = data.frame(name = c("C", "E")), edges = data.frame(from = c(1), to = c(2)), directed = TRUE)
gr_n <- tbl_graph(nodes = data.frame(name = c("A", "E")), edges = data.frame(from = numeric(0), to = numeric(0)), directed = TRUE)

p1 <- ggraph(gr_m, layout = "linear") +
  geom_edge_link(color = "steelblue", width = 1.5, arrow = arrow(length = unit(3, 'mm')), end_cap = circle(4, 'mm')) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Mutual (M)") + theme_graph() + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

p2 <- ggraph(gr_a, layout = "linear") +
  geom_edge_link(color = "steelblue", width = 1.5, arrow = arrow(length = unit(3, 'mm')), end_cap = circle(4, 'mm')) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Asymmetric (A)") + theme_graph() + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

p3 <- ggraph(gr_n, layout = "linear") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Null (N)") + theme_graph() + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

p1 + p2 + p3
```
</center>

---

## The Dyad Census

- The **Dyad Census** is a count of the total number of Mutuals ($N_M$), Asymmetrics ($N_A$), and Nulls ($N_N$) in a directed graph.
- We can compute these counts systematically using cell-by-cell matrix operations (the Hadamard product and matrix subtraction) on the directed adjacency matrix $\mathbf{A}$ and its transpose $\mathbf{A}^T$!

. . .

- We will perform these calculations on our 6-node directed reference graph $\mathbf{A}$:
  - $N_M$ = number of mutual dyads
  - $N_A$ = number of asymmetric dyads
  - $N_N$ = number of null dyads

---

## 1. Mutual Dyads Matrix ($\mathbf{M} = \mathbf{A} \circ \mathbf{A}^T$)

- Multiplying $\mathbf{A}$ by its transpose $\mathbf{A}^T$ element-wise extracts only mutual dyads where $a_{ij}=1$ and $a_{ji}=1$:

::: {.columns}
::: {.column width="50%"}
### Mutuals Matrix ($\mathbf{M}$)
```{r}
#| echo: false
M <- A_dir * t(A_dir)
M_disp <- M
diag(M_disp) <- NA
kbl(M_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Only reciprocal ties are kept (A-C and E-F).
:::

::: {.column width="50%"}
- Since each mutual dyad contains two directed edges ($M_{ij}=1$ and $M_{ji}=1$), we sum all cells in $\mathbf{M}$ and divide by two:
  $N_M = \frac{\sum M_{ij}}{2}$

. . .

- **Calculation**:
  - Sum of cells in $\mathbf{M}$ is 4.
  - $N_M = \frac{4}{2} = \mathbf{2} \text{ mutual dyads}$
:::
:::

---

## 2. Asymmetric Dyads Matrix ($\mathbf{S} = \mathbf{A} \circ (\mathbf{1} - \mathbf{A}^T)$)

- To isolate unreciprocated ties, we multiply $\mathbf{A}$ by the complement of the transpose ($\mathbf{1} - \mathbf{A}^T$). This keeps $a_{ij}=1$ only if $a_{ji}=0$:

::: {.columns}
::: {.column width="50%"}
### Asymmetrics Matrix ($\mathbf{S}$)
```{r}
#| echo: false
S <- A_dir * (1 - t(A_dir))
S_disp <- S
diag(S_disp) <- NA
kbl(S_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Retains only unreciprocated outgoing ties.
:::

::: {.column width="50%"}
- Since each asymmetric dyad has exactly one directed edge, the total count $N_A$ is simply the sum of all cells in $\mathbf{S}$:
  $N_A = \sum S_{ij}$

. . .

- **Calculation**:
  - Sum of cells in $\mathbf{S}$ is 8.
  - $N_A = \mathbf{8} \text{ asymmetric dyads}$
:::
:::

---

## 3. Null Dyads Matrix ($\mathbf{N} = (\mathbf{1} - \mathbf{A}) \circ (\mathbf{1} - \mathbf{A}^T)$)

- To isolate disconnected dyads, we multiply the complement of $\mathbf{A}$ by the complement of its transpose. This keeps a cell as 1 only if $a_{ij}=0$ and $a_{ji}=0$:

::: {.columns}
::: {.column width="50%"}
### Nulls Matrix ($\mathbf{N}$)
```{r}
#| echo: false
N_mat <- (1 - A_dir) * (1 - t(A_dir))
N_disp <- N_mat
diag(N_disp) <- NA
kbl(N_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Retains completely disconnected dyads.
:::

::: {.column width="50%"}
- Since each null dyad has 1s in both directions ($N_{ij}=1$ and $N_{ji}=1$), we sum all cells in $\mathbf{N}$ and divide by two:
  $N_N = \frac{\sum N_{ij}}{2}$

. . .

- **Calculation**:
  - Sum of cells in $\mathbf{N}$ is 10.
  - $N_N = \frac{10}{2} = \mathbf{5} \text{ null dyads}$
:::
:::

---

## The Dyad Census Identity Check

- In any directed graph of size $n$, the total number of dyads (pairs) is $\frac{n(n-1)}{2}$.
- For our $n=6$ graph, we must have exactly $\frac{6 \times 5}{2} = 15$ dyads.

. . .

- **Check**:
  - $N_M + N_A + N_N = 2 + 8 + 5 = \mathbf{15}$
  - The mathematics accounts for every single pair in the network perfectly!

. . .

- **The Matrix Identity**:
  - The cell-wise sum of our three matrices always equals the complete graph minus the diagonal ($\mathbf{J} - \mathbf{I}$):
  - $\mathbf{M} + \mathbf{S} + \mathbf{N} = \mathbf{J} - \mathbf{I}$

---

## Graph Reciprocity ($R$)

- **Graph Reciprocity ($R$)** measures the extent to which a directed graph approximates the ideal of complete reciprocity.
- It is defined as the proportion of connected dyads that are mutual:
  $R = \frac{N_M}{N_M + N_A}$

. . .

- **Calculation**:
  - We know $N_M = 2$ and $N_A = 8$.
  - $R = \frac{2}{2 + 8} = \frac{2}{10} = \mathbf{0.20}$
- **Sociological Meaning**:
  - **20%** of the connected dyads in this network are mutual. This represents a relatively low level of reciprocity!

---

# Part 4: Conformability Rules

---

## Standard Matrix Multiplication ($\mathbf{A} \times \mathbf{B}$)

- Unlike the element-wise Hadamard product, standard matrix multiplication is **non-elementary**.
- It involves multiplying the **rows of the left matrix** by the **columns of the right matrix**.
- Because rows and columns are matched, the sizes of the two multiplier matrices must follow a strict structural rule.
- This rule is called **Conformability**.

---

## The Conformability Rule

- *You can only multiply two matrices if the number of columns of the first matrix equals the number of rows of the second matrix.*
- Put their dimensions side-by-side:

$$\mathbf{A}_{3 \times \mathbf{5}} \times \mathbf{B}_{\mathbf{5} \times 6}$$

- **Inner Dimensions**: The bold "fives" are the inner dimensions. They **must be identical** for the multiplication to be defined (**conformable**).
- **Outer Dimensions**: The "three" and "six" are the outer dimensions. They dictate the size of the resulting product matrix:

$$\mathbf{A}_{3 \times \mathbf{5}} \times \mathbf{B}_{\mathbf{5} \times 6} = \mathbf{C}_{3 \times 6}$$

---

## Conformable vs. Non-Conformable

::: {.columns}
::: {.column width="50%"}
### Conformable Example
- Line up dimensions:
  $$\mathbf{A}_{4 \times \mathbf{2}} \times \mathbf{B}_{\mathbf{2} \times 3}$$
- **Inner dimensions match** (2 = 2).
- The multiplication is fully defined.
- **Product Matrix size**:
  $$\mathbf{C}_{4 \times 3}$$
:::

::: {.column width="50%"}
### Non-Conformable Example
- Line up dimensions:
  $$\mathbf{A}_{3 \times \mathbf{4}} \times \mathbf{B}_{\mathbf{2} \times 3}$$
- **Inner dimensions clash** (4 $\neq$ 2).
- This operation is **mathematically impossible**.
- **The Visual Reason**: When you try to dot-multiply a row of 4 by a column of 2, you "run out of numbers" on the right, making the sum undefined.
:::
:::

---

## Matrix Multiplication is Non-Commutative

- In scalar math, order doesn't matter: $4 \times 3 = 3 \times 4$.
- In matrix algebra, order **completely changes the structure**:

. . .

$$\mathbf{A}_{2 \times \mathbf{3}} \times \mathbf{B}_{\mathbf{3} \times 2} = \mathbf{C}_{2 \times 2} \quad \text{(Square 2x2 Matrix)}$$

. . .

$$\mathbf{B}_{3 \times \mathbf{2}} \times \mathbf{A}_{\mathbf{2} \times 3} = \mathbf{D}_{3 \times 3} \quad \text{(Square 3x3 Matrix)}$$

- Changing the multiplication order changes both the conformability checks and the resulting network's dimensions!

---

# Part 5: Step-by-Step Multiplication Mechanics

---

## Row-by-Column Vector Multiplication

- Let's multiply a $2 \times 3$ matrix $\mathbf{A}$ by a $3 \times 2$ matrix $\mathbf{B}$ to produce a $2 \times 2$ matrix $\mathbf{C}$:

$$\mathbf{A} = \begin{pmatrix} \mathbf{3} & \mathbf{4} & \mathbf{5} \\ 7 & 6 & 3 \end{pmatrix} \quad \mathbf{B} = \begin{pmatrix} \mathbf{3} & 7 \\ \mathbf{4} & 9 \\ \mathbf{5} & 3 \end{pmatrix}$$

- To compute cell $c_{ij}$ in the target matrix, we multiply row $i$ of the left matrix by column $j$ of the right matrix, then sum the products.
- Let's trace each of the 4 cells in the target matrix $\mathbf{C}_{2 \times 2}$ progressively.

---

## Target Cell $c_{11}$ (Row 1, Column 1)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 1 of A**: $\color{red}{\{3, 4, 5\}}$
* **Column 1 of B**: $\color{blue}{\{3, 4, 5\}}$

### Calculation
$$c_{11} = (\color{red}{3} \times \color{blue}{3}) + (\color{red}{4} \times \color{blue}{4}) + (\color{red}{5} \times \color{blue}{5})$$
$$c_{11} = 9 + 16 + 25 = \mathbf{50}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} \mathbf{50} & ? \\ ? & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{12}$ (Row 1, Column 2)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 1 of A**: $\color{red}{\{3, 4, 5\}}$
* **Column 2 of B**: $\color{blue}{\{7, 9, 3\}}$

### Calculation
$$c_{12} = (\color{red}{3} \times \color{blue}{7}) + (\color{red}{4} \times \color{blue}{9}) + (\color{red}{5} \times \color{blue}{3})$$
$$c_{12} = 21 + 36 + 15 = \mathbf{72}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & \mathbf{72} \\ ? & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{21}$ (Row 2, Column 1)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 2 of A**: $\color{red}{\{7, 6, 3\}}$
* **Column 1 of B**: $\color{blue}{\{3, 4, 5\}}$

### Calculation
$$c_{21} = (\color{red}{7} \times \color{blue}{3}) + (\color{red}{6} \times \color{blue}{4}) + (\color{red}{3} \times \color{blue}{5})$$
$$c_{21} = 21 + 24 + 15 = \mathbf{60}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & 72 \\ \mathbf{60} & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{22}$ (Row 2, Column 2)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 2 of A**: $\color{red}{\{7, 6, 3\}}$
* **Column 2 of B**: $\color{blue}{\{7, 9, 3\}}$

### Calculation
$$c_{22} = (\color{red}{7} \times \color{blue}{7}) + (\color{red}{6} \times \color{blue}{9}) + (\color{red}{3} \times \color{blue}{3})$$
$$c_{22} = 49 + 54 + 9 = \mathbf{112}$$
:::

::: {.column width="50%"}
### Finished Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & 72 \\ 60 & \mathbf{112} \end{pmatrix}$$
:::
:::

---

# Part 6: Dynamic Network Applications

---

## Transpose Multiplication: Common Neighbors

- If we multiply a network's adjacency matrix $\mathbf{A}$ by its transpose $\mathbf{A}^T$:
  $\mathbf{A} \times \mathbf{A}^T = \mathbf{B}$
- Why does this standard matrix multiplication reveal structural overlaps?

. . .

- **The Diagonal Cells ($b_{ii}$)**: 
  - Corresponds to multiplying Row $i$ of $\mathbf{A}$ by Column $i$ of $\mathbf{A}^T$ (which is identical to Row $i$).
  - It squares each binary entry and sums them up.
  - **Result**: Counts the total number of neighbors of node $i$ (the **node degree**).
- **The Off-Diagonal Cells ($b_{ij}$)**:
  - Multiplies Row $i$ (node $i

---

## Tracing Common Neighbors Step-by-Step

- Let's use our 6-node symmetric "Parties With" adjacency matrix $\mathbf{P}$ ($\mathbf{P}^T = \mathbf{P}$):

::: {.columns}
::: {.column width="50%"}
### Adjacency Matrix $\mathbf{P}$
$\mathbf{P} = \begin{pmatrix} 0 & \mathbf{1} & 1 & 0 & \mathbf{1} & 0 \\ 1 & 0 & 1 & 1 & 0 & 0 \\ 1 & \mathbf{1} & 0 & 1 & \mathbf{1} & 1 \\ 0 & 1 & 1 & 0 & 1 & 0 \\ 1 & 0 & 1 & 1 & 0 & 1 \\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix}$
- Let's check common neighbors between **node A (Row 1)** and **node C (Col 3)**.
:::

::: {.column width="50%"}
### Vector Multiplication
* **Row A (Node A)**: $\color{red}{\{0, 1, 1, 0, 1, 0\}}$
* **Column C (Node C)**: $\color{blue}{\{1, 1, 0, 1, 1, 1\}}$

### Calculation
$(\color{red}{0}\times\color{blue}{1}) + (\color{red}{1}\times\color{blue}{1}) + (\color{red}{1}\times\color{blue}{0}) + (\color{red}{0}\times\color{blue}{1}) + (\color{red}{1}\times\color{blue}{1}) + (\color{red}{0}\times\color{blue}{1}) = \mathbf{2}$

- **Result**: Nodes A and C share exactly **2** common neighbors.
- Tracing the 1s, we see the shared neighbors are **B** and **E**!
:::
:::

---

## Matrix Powers ($\mathbf{A}^n$) & Paths in Networks

- Multiplying a square adjacency matrix by itself recursively reveals **indirect connections**:
  - $\mathbf{A}^2 = \mathbf{A} \times \mathbf{A}$
  - $\mathbf{A}^3 = \mathbf{A} \times \mathbf{A} \times \mathbf{A}$
- In network science, the entries of the $n^{th}$ power of an adjacency matrix have a precise graph-theoretic meaning:
  - **Off-Diagonal Cells ($a^{n}_{ij}$)**: Record the exact number of **walks of length $n$** connecting node $i$ and node $j$.
  - **Diagonal Cells ($a^{n}_{ii}$)**: Record the exact number of **cycles of length $n$** that start and end at node $i$.

---

## Walks of Length 2: $\mathbf{A}^2$

- Let's use our 5-node undirected reference graph (Slides 109-119):

::: {.columns}
::: {.column width="50%"}
### Adjacency Matrix ($\mathbf{A}$)
```{r}
#| echo: false
A_u5_disp <- A_undir_5
diag(A_u5_disp) <- NA
kbl(A_u5_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 4.5
ggraph(gr_undir_5, layout = "manual", x = l_coords_5[,1], y = l_coords_5[,2]) +
  geom_edge_link(color = "steelblue", width = 2.0) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
</center>
:::

::: {.column width="50%"}
### Squared Matrix ($\mathbf{A}^2$)
```{r}
#| echo: false
A2 <- A_undir_5 %*% A_undir_5
kbl(A2, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- **Diagonals** equal node degrees: a (1), b (2), c (3), d (2), e (2).
- **Off-diagonals** count walks of length 2 (common neighbors). For example, $a^2_{a,c} = 1$, meaning exactly one length 2 walk connects a and c.
:::
:::

---

## Walks of Length 3: $\mathbf{A}^3$

- Now, let's examine the cubed adjacency matrix ($\mathbf{A}^3$):

::: {.columns}
::: {.column width="50%"}
### Cubed Matrix ($\mathbf{A}^3$)
```{r}
#| echo: false
A3 <- A_undir_5 %*% A_undir_5 %*% A_undir_5
kbl(A3, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- **Off-diagonals** represent length 3 walks.
- **Diagonals** count cycles of length 3 (triangles).
:::

::: {.column width="50%"}
- A cycle of length 3 is a **triangle**.
- Since each triangle can be traversed in two directions (clockwise & counterclockwise), we can find the exact number of triangles containing node $i$ as:
  $\text{Triangles}(i) = \frac{a^3_{ii}}{2}$

. . .

- **Calculation**:
  - Only c, d, and e have non-zero diagonals ($2$).
  - Each belongs to $\frac{2}{2} = \mathbf{1}$ triangle ({c, d, e}).
  - Nodes a and b have diagonal 0 (they form no triangles).
:::
:::

---

## Walks of Length 3: Weighted Graph Representation

- Plotting the network with edge sizes and color intensities proportional to the entries of $\mathbf{A}^3$ reveals distinct clique structures sharing multiple indirect pathways:

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5.5
m_val <- max(A3)
gr_pows <- graph.adjacency(A3, mode="undirected", weighted=TRUE) %>% 
  as_tbl_graph() 

p_pows <- ggraph(gr_pows, layout = "manual", x = l_coords_5[,1], y = l_coords_5[,2]) +
  geom_edge_link(aes(color = weight/m_val, width = weight/m_val)) +
  geom_node_point(color = "tan2", size = 20) +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + 
  scale_edge_color_steps(low = "grey90", high = "steelblue4") +
  theme(legend.position = "none") + coord_cartesian(clip = "off")
p_pows
```

---

# Part 7: Summary & Conclusion

---

## Summary of Matrix Algebraic Power

- **Basic operations combine relationships**:
  - Addition ($\mathbf{P} + \mathbf{G}$) counts uniplex vs. multiplex relationships.
  - Subtraction from 1 ($\mathbf{1} - \mathbf{P}$) flips relations to yield the graph complement (*Does Not Party With*).
  - Hadamard Product ($\mathbf{P} \circ \mathbf{G}$) extracts multiplex overlaps.
- **The Matrix Transpose ($\mathbf{A}^T$)**:
  - Flips rows and columns; essential for analyzing directed relations.
- **The Directed Dyad Census**:
  - We use $\mathbf{A}$ and $\mathbf{A}^T$ cell-by-cell to compute Mutuals ($M$), Asymmetrics ($S$), and Nulls ($N$), spelling the **MAN** census.
  - Graph Reciprocity is computed as $R = \frac{N_M}{N_M + N_A}$.
- **Conformability dictates standard multiplication**:
  - Inner dimensions must be equal.
- **Standard multiplication and powers reveal structures**:
  - $\mathbf{A} \times \mathbf{A}^T$ computes common neighbors (off-diagonals) and degrees (diagonals).
  - $\mathbf{A}^n$ maps indirect walks of length $n$. $\mathbf{A}^2$ diagonals are degrees; $\mathbf{A}^3$ diagonals are twice the number of triangles.
