---
title: "Matrix Algebra in Social Networks"
subtitle: "Basic and Advanced Matrix Operations"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(expm)
library(patchwork)

# Set seed for reproducibility
set.seed(859)

# 1. 12-node company network (Hanging out vs Co-working)
gr1 <- play_islands(3, 4, m = 1, p_within = 0.8) %>% 
  activate(nodes) %>% 
  mutate(name = toupper(letters[1:12]))

# Get stable layout coordinates
l_tree <- ggraph(gr1, layout = 'tree')
l_coords <- as.matrix(l_tree$data[, 1:2])

# Hanging out graph
p_hang <- ggraph(gr1, layout = l_coords) +
  geom_edge_link(color = "steelblue", edge_width = 1.3) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph()

# Co-working graph
gr2 <- play_islands(4, 3, m = 1, p_within = 1) %>% 
  activate(nodes) %>% 
  mutate(name = toupper(letters[1:12]))

p_work <- ggraph(gr2, layout = l_coords) +
  geom_edge_link(color = "firebrick", edge_width = 1.3) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph()

A1 <- as.matrix(as_adjacency_matrix(gr1))
A2 <- as.matrix(as_adjacency_matrix(gr2))
```

# Part 1: Basic Matrix Operations

---

## The Company Social System

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Suppose we study 12 employees at a small company and record two separate relationships:
  1. **Hanging Out ($\mathbf{H}$)**: Whether they hang out together after work (Blue edges).
  2. **Co-working ($\mathbf{C}$)**: Whether they are assigned to work together in team projects (Red edges).
- How do we combine, compare, or intersect these overlapping structures?
- We use **Basic Matrix Operations**: addition, subtraction, and the dot (element-wise) product.
:::
:::

::: {.column width="40%"}
### Hanging Out ($\mathbf{H}$)
```{r}
#| echo: false
#| fig-height: 2.8
#| fig-width: 3.5
#| fig-align: "center"
p_hang
```

### Co-working ($\mathbf{C}$)
```{r}
#| echo: false
#| fig-height: 2.8
#| fig-width: 3.5
#| fig-align: "center"
p_work
```
:::

::::

---

## Matrix Addition ($\mathbf{H} + \mathbf{C}$)

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- **Matrix Addition**:
  - Done element-by-element: $S_{ij} = H_{ij} + C_{ij}$.
  - Requires **identical dimensions** ($12 \times 12$).
- **Sociological Meaning**:
  - Cell value $2$: Share **both** hangout and co-working ties.
  - Cell value $1$: Share **either** hangout or co-working ties, but not both.
  - Cell value $0$: Share no ties.
  - Represents a weighted network of **overall relationship volume**.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_sum <- graph_from_adjacency_matrix(A1 + A2, mode = "undirected", weighted = TRUE)
ggraph(gr_sum, layout = l_coords) +
  geom_edge_link(aes(color = as.factor(weight), edge_width = weight*1.2)) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  scale_edge_color_manual(values = c("1" = "steelblue", "2" = "purple")) +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
:::

::::

---

## Matrix Subtraction & Absolute Value ($|\mathbf{H} - \mathbf{C}|$)

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- **Matrix Subtraction**:
  - Done element-by-element: $D_{ij} = |H_{ij} - C_{ij}|$.
  - Isolates cases where the two relations **disagree**.
- **Sociological Meaning**:
  - Cell value $1$: Share **exactly one** of the ties (hangout or co-working), but **not both**.
  - Cell value $0$: Share either both ties ($1-1=0$), or neither ($0-0=0$).
  - Represents the **symmetric difference** (green edges).
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_sub <- graph_from_adjacency_matrix(abs(A1 - A2), mode = "undirected")
ggraph(gr_sub, layout = l_coords) +
  geom_edge_link(color = "forestgreen", edge_width = 1.3) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Matrix Dot (Hadamard) Product ($\mathbf{H} \cdot \mathbf{C}$)

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- **Element-wise Multiplication**:
  - Cell product: $P_{ij} = H_{ij} \times C_{ij}$.
  - Requires identical dimensions.
  - Also called the **Hadamard product** (distinct from standard matrix multiplication).
- **Sociological Meaning**:
  - Cell value $1$ only if **both** $H_{ij} = 1$ and $C_{ij} = 1$ ($1 \times 1 = 1$).
  - Cell value $0$ otherwise.
  - Represents the **overlap / intersection** of networks (purple edges).
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dot <- graph_from_adjacency_matrix(A1 * A2, mode = "undirected")
ggraph(gr_dot, layout = l_coords) +
  geom_edge_link(color = "purple", edge_width = 2.0) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 2: Matrix Multiplication

---

## Standard Matrix Multiplication ($\mathbf{A} \times \mathbf{B}$)

::: {.incremental}
- Unlike the dot product, standard matrix multiplication is **not element-wise**; it involves multiplying rows of the left matrix by columns of the right matrix.
- **The Conformability Rule**:
  - You can only multiply two matrices if the **inner dimensions match**.
  - In other words, the number of columns in the first matrix ($\mathbf{A}$) must equal the number of rows in the second matrix ($\mathbf{B}$).
  - $$\mathbf{A}_{3 \times \mathbf{5}} \times \mathbf{B}_{\mathbf{5} \times 6} = \mathbf{C}_{3 \times 6}$$
- **The Product Dimensions**:
  - The resulting matrix has dimensions equal to the **outer dimensions** (Rows of left, Columns of right).
- **Non-Commutative**:
  - Matrix multiplication is order-dependent: $\mathbf{A} \times \mathbf{B} \neq \mathbf{B} \times \mathbf{A}$.
:::

---

## Step-by-Step Multiplication

:::: {.columns}

::: {.column width="55%"}
- To compute cell $c_{ij}$ in the product matrix $\mathbf{C}$:
  - Take row $i$ of left matrix $\mathbf{A}$ and column $j$ of right matrix $\mathbf{B}$.
  - Multiply corresponding elements and **sum the products** (calculating their dot product).
- **Example**: Cell $c_{11}$ of $\mathbf{A}_{5 \times 3} \times \mathbf{B}_{3 \times 5}$
  - Row 1 of $\mathbf{A}$: $\{3, 4, 5\}$
  - Column 1 of $\mathbf{B}$: $\{3, 4, 5\}$
  - $$c_{11} = (3 \times 3) + (4 \times 4) + (5 \times 5) = 9 + 16 + 25 = 50$$
:::

::: {.column width="45%"}
### Matrix A ($5 \times 3$)
```{r}
#| echo: false
A_prod <- matrix(c(3, 7, 4, 5, 2, 4, 9, 6, 3, 5, 5, 3, 2, 4, 4), nrow = 5)
kbl(A_prod, align = "c", format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

### Matrix B ($3 \times 5$)
```{r}
#| echo: false
B_prod <- t(A_prod)
kbl(B_prod, align = "c", format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```
:::

::::

---

## The Identity Matrix ($\mathbf{I}$)

::: {.incremental}
- The **Identity Matrix ($\mathbf{I}$)** is a special square matrix of order $N$ with $1$s on the main diagonal and $0$s on the off-diagonal cells.
- e.g., a $4 \times 4$ Identity Matrix:
  - $$
    \mathbf{I} = 
    \begin{pmatrix}
    1 & 0 & 0 & 0 \\
    0 & 1 & 0 & 0 \\
    0 & 0 & 1 & 0 \\
    0 & 0 & 0 & 1
    \end{pmatrix}
    $$
- It acts as the **multiplicative identity** in matrix algebra (equivalent to the number $1$ in standard math):
  - $$\mathbf{A} \times \mathbf{I} = \mathbf{A}$$
  - $$\mathbf{I} \times \mathbf{A} = \mathbf{A}$$
:::

# Part 3: Advanced Matrix Operations

---

## The Transpose of a Matrix ($\mathbf{A}^T$)

::: {.incremental}
- **The Transpose Operation**:
  - Reverses the row and column dimensions of a matrix.
  - Rows of the original matrix $\mathbf{A}$ become columns of its transpose $\mathbf{A}^T$, and columns of $\mathbf{A}$ become rows of $\mathbf{A}^T$.
  - If $\mathbf{A}$ is of dimensions $R \times C$, then $\mathbf{A}^T$ is of dimensions $C \times R$.
- **Crucial Property**:
  - The product of any matrix multiplied by its transpose is **always defined and conformable**:
    - $\mathbf{A}_{R \times C} \times \mathbf{A}^T_{C \times R} = \mathbf{B}_{R \times R}$ (Symmetric Square Matrix).
    - $\mathbf{A}^T_{C \times R} \times \mathbf{A}_{R \times C} = \mathbf{D}_{C \times C}$ (Symmetric Square Matrix).
:::

---

## Matrix Transpose Multiplication: Common Neighbors

:::: {.columns}

::: {.column width="60%"}
::: {.incremental}
- In social network analysis, multiplying an adjacency matrix $\mathbf{A}$ by its transpose $\mathbf{A}^T$ yields an incredibly intuitive result:
  - **The Diagonal Cells** ($b_{ii}$): Count the total number of neighbors of node $i$ (the **node degree**).
  - **The Off-Diagonal Cells** ($b_{ij}$): Count the exact **number of common neighbors** shared by row node $i$ and column node $j$.
  - $$\mathbf{A} \times \mathbf{A}^T = \mathbf{O}^{Common}$$
- Essential for identifying structural similarity and cliques!
:::
:::

::: {.column width="40%"}
### Common Neighbors ($\mathbf{A} \times \mathbf{A}^T$)
```{r}
#| echo: false
A_mat <- as.matrix(as_adjacency_matrix(gr1))
A_tA <- A_mat %*% t(A_mat)
kbl(A_tA, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 14)
```
:::

::::

---

## Matrix Powers ($\mathbf{A}^n$)

::: {.incremental}
- You can multiply a square matrix by itself recursively to get **matrix powers**:
  - $\mathbf{A}^2 = \mathbf{A} \times \mathbf{A}$
  - $\mathbf{A}^3 = \mathbf{A} \times \mathbf{A} \times \mathbf{A}$
- In network science, the entries of the $n^{th}$ power of an adjacency matrix have a brilliant graph-theoretic meaning:
  - **Off-Diagonal Cells ($a^{n}_{ij}$)**: Record the exact number of **indirect connections (walks) of length $n$** connecting node $i$ and node $j$.
  - **Diagonal Cells ($a^{n}_{ii}$)**: Record the exact number of **cycles of length $n$** that start and end at node $i$.
:::

---

## Walks and Cycles: $\mathbf{A}^2$ & $\mathbf{A}^3$

::: {.incremental}
- **Adjacency Matrix Squared ($\mathbf{A}^2$)**:
  - **Off-diagonals** represent the number of **walks of length 2** between nodes.
  - **Diagonals** represent the number of walks of length 2 starting and ending at the same node. This is equal to the **node degree**!
- **Adjacency Matrix Cubed ($\mathbf{A}^3$)**:
  - **Off-diagonals** represent the number of **walks of length 3** between nodes.
  - **Diagonals** represent cycles of length 3 (triangles).
  - Since each triangle can be traversed in two directions, the diagonal cell divided by 2 ($\frac{a^3_{ii}}{2}$) gives the exact **number of triangles (cliques of size 3)** that node $i$ belongs to!
:::

---

## Walks of Length 3: Weighted Graph Representation

- Plotting the network with edge sizes and color intensities proportional to the entries of $\mathbf{A}^3$ reveals distinct hangout cliques sharing multiple indirect pathways:

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5.5
A3 <- A_mat %^% 3
m_val <- max(A3)
gr_pows <- graph.adjacency(A3, mode="undirected", weighted=TRUE) %>% 
  as_tbl_graph() 
ggraph(gr_pows, layout = l_coords) +
  geom_edge_link(aes(color = weight/m_val, width = weight/m_val)) +
  geom_node_point(color = "tan2", size = 20) +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + 
  scale_edge_color_steps(low = "grey90", high = "steelblue4") +
  theme(legend.position = "none") + coord_cartesian(clip = "off")
```

# Part 4: Conclusion & Summary

---

## Summary of Matrix Algebraic Power

::: {.incremental}
- **Basic operations combine relationships**:
  - Addition ($\mathbf{H} + \mathbf{C}$) counts tie volume; Subtraction ($|\mathbf{H} - \mathbf{C}|$) highlights relational disagreements; Dot product ($\mathbf{H} \cdot \mathbf{C}$) extracts overlap.
- **Conformability dictates multiplication**:
  - Multiplications are only defined if inner dimensions match.
- **Multiplying by Transpose reveals overlaps**:
  - $\mathbf{A} \times \mathbf{A}^T$ computes common neighbors (off-diagonals) and degrees (diagonals).
- **Matrix Powers map paths**:
  - $\mathbf{A}^n$ enumerates walks of length $n$. $\mathbf{A}^2$ diagonals are degrees; $\mathbf{A}^3$ diagonals are twice the number of triangles.
:::
