---
title: "Advanced Network Matrices"
subtitle: "Beyond the Adjacency Matrix"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Set seed for reproducible layout of 5-node graph
set.seed(123)

# 1. 5-node directed graph for reachability step-by-step (matching PPT slides 79-91)
nodes_5 <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(0, -1, 1, 0, 1.2), y = c(1, 0.2, 0.2, -1, -0.8))
edges_reach <- data.frame(from = c(1, 1, 2, 3, 5), to = c(3, 4, 5, 5, 4))
gr_reach_5 <- tbl_graph(nodes = nodes_5, edges = edges_reach, directed = TRUE)

# Layout coordinates for manual placement of 5-node graph
layout_coords_5 <- as.matrix(nodes_5[, c("x", "y")])

# Helper function to plot 5-node reachability graph with highlighted paths
plot_reach_highlight <- function(active_edges = NULL, active_nodes = NULL) {
  edges_mod <- edges_reach
  edges_mod$highlight <- "no"
  if (!is.null(active_edges)) {
    for (i in seq_len(nrow(edges_mod))) {
      f <- edges_mod$from[i]
      t <- edges_mod$to[i]
      f_name <- nodes_5$name[f]
      t_name <- nodes_5$name[t]
      if (paste(f_name, t_name, sep="-") %in% active_edges) {
        edges_mod$highlight[i] <- "yes"
      }
    }
  }
  
  nodes_mod <- nodes_5
  nodes_mod$highlight <- "no"
  if (!is.null(active_nodes)) {
    nodes_mod$highlight[nodes_mod$name %in% active_nodes] <- "yes"
  }
  
  gr <- tbl_graph(nodes = nodes_mod, edges = edges_mod, directed = TRUE)
  
  p <- ggraph(gr, layout = layout_coords_5) +
    geom_edge_link(aes(color = highlight, edge_width = highlight),
                   arrow = arrow(length = unit(4, 'mm')),
                   end_cap = circle(6, 'mm')) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
    scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
    scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
    scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
    theme_graph() + coord_cartesian(clip = "off")
  return(p)
}

# 2. 8-node directed graphs for the three flavors of connectivity
# strongly connected (all-ones reachability)
edges_strong <- data.frame(from = c(1,2,3,4,5,6,7,8), to = c(2,3,4,5,6,7,8,1))
gr_strong <- tbl_graph(nodes = data.frame(name = 1:8), edges = edges_strong, directed = TRUE)

# unilaterally connected
edges_unilateral <- data.frame(from = c(2,3,4,5,6,7,8, 1,3,4,5,6,7,1), to = c(3,4,5,6,7,8,1, 3,1,1,1,1,1,5))
gr_unilateral <- tbl_graph(nodes = data.frame(name = 1:8), edges = edges_unilateral, directed = TRUE)

# weakly connected
edges_weak <- data.frame(from = c(1,2,3,4,5,6,8, 7,1,3,4,5,6,1), to = c(3,4,5,6,7,8,1, 8,3,1,1,1,1,5))
gr_weak <- tbl_graph(nodes = data.frame(name = 1:8), edges = edges_weak, directed = TRUE)

# 3. 5-node undirected graph for geodesic distance step-by-step (matching PPT slides 132-138)
nodes_undir_5 <- data.frame(name = c("A", "B", "C", "D", "E"), x = c(0, 1.5, -0.6, 0.6, 0.6), y = c(1, -0.5, 0, 0, -1))
edges_undir_5 <- data.frame(from = c(1, 1, 2, 3, 3, 4), to = c(3, 4, 5, 4, 5, 5))
gr_undir_5 <- tbl_graph(nodes = nodes_undir_5, edges = edges_undir_5, directed = FALSE)
layout_coords_undir_5 <- as.matrix(nodes_undir_5[, c("x", "y")])

# Helper function to plot 5-node undirected graph with highlighted shortest paths
plot_undir_highlight <- function(active_edges = NULL, active_nodes = NULL) {
  edges_mod <- edges_undir_5
  edges_mod$highlight <- "no"
  if (!is.null(active_edges)) {
    for (i in seq_len(nrow(edges_mod))) {
      f <- edges_mod$from[i]
      t <- edges_mod$to[i]
      f_name <- nodes_undir_5$name[f]
      t_name <- nodes_undir_5$name[t]
      if (paste(f_name, t_name, sep="-") %in% active_edges || paste(t_name, f_name, sep="-") %in% active_edges) {
        edges_mod$highlight[i] <- "yes"
      }
    }
  }
  
  nodes_mod <- nodes_undir_5
  nodes_mod$highlight <- "no"
  if (!is.null(active_nodes)) {
    nodes_mod$highlight[nodes_mod$name %in% active_nodes] <- "yes"
  }
  
  gr <- tbl_graph(nodes = nodes_mod, edges = edges_mod, directed = FALSE)
  
  p <- ggraph(gr, layout = layout_coords_undir_5) +
    geom_edge_link(aes(color = highlight, edge_width = highlight)) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
    scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
    scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
    scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
    theme_graph() + coord_cartesian(clip = "off")
  return(p)
}

# 4. 5-node directed graph for geodesic distance step-by-step
edges_dir_5 <- data.frame(from = c(1, 1, 2, 3, 5), to = c(3, 4, 5, 5, 4))
gr_dir_5 <- tbl_graph(nodes = nodes_undir_5, edges = edges_dir_5, directed = TRUE)

# Helper function to plot 5-node directed graph with highlighted geodesic paths
plot_dir_highlight_geo <- function(active_edges = NULL, active_nodes = NULL) {
  edges_mod <- edges_dir_5
  edges_mod$highlight <- "no"
  if (!is.null(active_edges)) {
    for (i in seq_len(nrow(edges_mod))) {
      f <- edges_mod$from[i]
      t <- edges_mod$to[i]
      f_name <- nodes_undir_5$name[f]
      t_name <- nodes_undir_5$name[t]
      if (paste(f_name, t_name, sep="-") %in% active_edges) {
        edges_mod$highlight[i] <- "yes"
      }
    }
  }
  
  nodes_mod <- nodes_undir_5
  nodes_mod$highlight <- "no"
  if (!is.null(active_nodes)) {
    nodes_mod$highlight[nodes_mod$name %in% active_nodes] <- "yes"
  }
  
  gr <- tbl_graph(nodes = nodes_mod, edges = edges_mod, directed = TRUE)
  
  p <- ggraph(gr, layout = layout_coords_undir_5) +
    geom_edge_link(aes(color = highlight, edge_width = highlight),
                   arrow = arrow(length = unit(4, 'mm')),
                   end_cap = circle(6, 'mm')) +
    geom_node_point(aes(color = highlight), size = 18) +
    geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
    scale_edge_color_manual(values = c("yes" = "red", "no" = "steelblue"), guide = "none") +
    scale_edge_width_manual(values = c("yes" = 2.2, "no" = 1.0), guide = "none") +
    scale_color_manual(values = c("yes" = "red", "no" = "tan2"), guide = "none") +
    theme_graph() + coord_cartesian(clip = "off")
  return(p)
}

# 5. Helper function for neighborhood overlap highlighting
plot_neighbors_highlight <- function(active_nodes = NULL, common_nodes = NULL) {
  nodes_mod <- nodes_undir_5
  nodes_mod$color_type <- "no"
  if (!is.null(active_nodes)) {
    nodes_mod$color_type[nodes_mod$name %in% active_nodes] <- "active"
  }
  if (!is.null(common_nodes)) {
    nodes_mod$color_type[nodes_mod$name %in% common_nodes] <- "common"
  }
  
  edges_mod <- edges_undir_5
  edges_mod$color_type <- "no"
  if (!is.null(active_nodes)) {
    for (i in seq_len(nrow(edges_mod))) {
      f <- edges_mod$from[i]
      t <- edges_mod$to[i]
      f_name <- nodes_undir_5$name[f]
      t_name <- nodes_undir_5$name[t]
      if (any(f_name %in% active_nodes) && any(t_name %in% active_nodes)) {
        edges_mod$color_type[i] <- "yes"
      }
    }
  }
  
  gr <- tbl_graph(nodes = nodes_mod, edges = edges_mod, directed = FALSE)
  
  p <- ggraph(gr, layout = layout_coords_undir_5) +
    geom_edge_link(color = "steelblue", edge_width = 1.0) +
    geom_node_point(aes(color = color_type), size = 18) +
    geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
    scale_color_manual(values = c("active" = "red", "common" = "forestgreen", "no" = "tan2"), guide = "none") +
    theme_graph() + coord_cartesian(clip = "off")
  return(p)
}
```

# Part 1: The Reachability Matrix Step-by-Step

---

## What is Reachability?

::: {.incremental}
- In a directed graph, the adjacency matrix only records paths of **length 1** (direct arrows).
- **Reachability** expands our view to indirect paths of **any length**:
  - Node $j$ is **reachable** from node $i$ if there exists a directed path starting at $i$ and terminating at $j$.
  - Example: If we have $A \rightarrow C \rightarrow E \rightarrow D$, then $A$ can reach $D$ via a path of length 3, even if no direct arrow $A \rightarrow D$ exists.
- Representing these reachability relations systematically yields the **Reachability Matrix ($\mathbf{D^r}$)**:
  - Binary matrix where $D^r_{ij} = 1$ if $i$ can reach $j$ via a path of any length, else $0$.
:::

---

## Reference Graph & Empty Matrix

We have 5 nodes $\{A, B, C, D, E\}$. Let's build the $5 \times 5$ reachability matrix. Diagonals are blocked (`--`).

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_reach_highlight()
```
:::

::: {.column width="50%"}
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | -- | ? | ? | ? | ? |
| **B** | ? | -- | ? | ? | ? |
| **C** | ? | ? | -- | ? | ? |
| **D** | ? | ? | ? | -- | ? |
| **E** | ? | ? | ? | ? | -- |
:::

::::

---

## Step 1: Filling Row A (Actor A's Reachability)

Let's trace which nodes can be reached by starting a directed walk at node **A**.

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_reach_highlight(active_edges = c("A-C", "A-D", "C-E", "E-D"), active_nodes = c("A", "C", "D", "E"))
```
:::

::: {.column width="50%"}
- **Can A reach B?**
  - **No** ($d^r_{AB} = 0$). No directed path goes against $B \rightarrow E$.
- **Can A reach C?**
  - **Yes** ($d^r_{AC} = 1$). Direct link $A \rightarrow C$.
- **Can A reach D?**
  - **Yes** ($d^r_{AD} = 1$). Direct link $A \rightarrow D$.
- **Can A reach E?**
  - **Yes** ($d^r_{AE} = 1$). Path $A \rightarrow C \rightarrow E$ exists.
- **Row A**: `(--,  0,  1,  1,  1)`
:::

::::

---

## Step 2: Filling Row B (Actor B's Reachability)

Let's trace which nodes can be reached by starting a directed walk at node **B**.

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_reach_highlight(active_edges = c("B-E", "E-D"), active_nodes = c("B", "E", "D"))
```
:::

::: {.column width="50%"}
- **Can B reach A or C?**
  - **No** ($d^r_{BA} = 0$, $d^r_{BC} = 0$). No paths go left from E.
- **Can B reach D?**
  - **Yes** ($d^r_{BD} = 1$). Path $B \rightarrow E \rightarrow D$ of length 2.
- **Can B reach E?**
  - **Yes** ($d^r_{BE} = 1$). Direct link $B \rightarrow E$.
- **Row B**: `(0,  --,  0,  1,  1)`
:::

::::

---

## Step 3: Completed Reachability Matrix

Repeating this row-by-row analysis yields the completed reachability matrix.

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_reach_highlight()
```
:::

::: {.column width="50%"}
### Reachability Matrix ($\mathbf{D^r}$)
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | -- | 0 | 1 | 1 | 1 |
| **B** | 0 | -- | 0 | 1 | 1 |
| **C** | 0 | 0 | -- | 1 | 1 |
| **D** | 0 | 0 | 0 | -- | 0 |
| **E** | 0 | 0 | 0 | 1 | -- |

- **Receiver Node (D)**: Row D is all $0$s (D can't reach anyone).
- **Transmitter Node (B)**: Column B is all $0$s (B can't be reached by anyone).
:::

::::

# Part 2: Directed Graphs: Flavors of Connectivity

---

## Directed Graph Connectivity

::: {.incremental}
- In undirected graphs, connectivity is simple: either there is a path between every pair (connected) or there is not (disconnected).
- In directed graphs, edge directionality creates **three distinct major levels of connectivity**:
  1. **Strongly Connected**: Every node can reach every other node.
  2. **Unilaterally Connected**: For every pair $\{u, v\}$, at least one node can reach the other ($u \rightarrow v$ OR $v \rightarrow u$).
  3. **Weakly Connected**: Nodes are connected only if we ignore the directions of the arrows (connected via semipaths).
- Let's see how these levels map directly to **zeros and ones in the reachability matrix**.
:::

---

## Case 1: Strongly Connected Graphs

Every pair of nodes is mutually reachable.

:::: {.columns align=center}

::: {.column width="50%"}
### Strongly Connected Graph
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
ggraph(gr_strong, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.15,
                 arrow = arrow(length = unit(3, 'mm')),
                 end_cap = circle(5, 'mm')) +
  geom_node_point(color = "tan2", size = 14) +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column width="50%"}
### Reachability Matrix ($\mathbf{D^r}$)
| | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **1** | -- | 1 | 1 | 1 | 1 | 1 | 1 | 1 |
| **2** | 1 | -- | 1 | 1 | 1 | 1 | 1 | 1 |
| **3** | 1 | 1 | -- | 1 | 1 | 1 | 1 | 1 |
| **4** | 1 | 1 | 1 | -- | 1 | 1 | 1 | 1 |
| **5** | 1 | 1 | 1 | 1 | -- | 1 | 1 | 1 |
| **6** | 1 | 1 | 1 | 1 | 1 | -- | 1 | 1 |
| **7** | 1 | 1 | 1 | 1 | 1 | 1 | -- | 1 |
| **8** | 1 | 1 | 1 | 1 | 1 | 1 | 1 | -- |

- **No Zeros**: Since everyone can reach everyone, the reachability matrix is the **all-ones** matrix!
:::

::::

---

## Case 2: Unilaterally Connected Graphs

At least one directed path exists between every pair, but not necessarily in both directions.

:::: {.columns align=center}

::: {.column width="50%"}
### Unilaterally Connected Graph
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
ggraph(gr_unilateral, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.15,
                 arrow = arrow(length = unit(3, 'mm')),
                 end_cap = circle(5, 'mm')) +
  geom_node_point(color = "tan2", size = 14) +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column width="50%"}
### Reachability Matrix ($\mathbf{D^r}$)
| | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **1** | -- | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | 1 | 1 | 1 |
| **2** | 1 | -- | 1 | 1 | 1 | 1 | 1 | 1 |
| **3** | 1 | <span style="color:red; font-weight:bold;">0</span> | -- | 1 | 1 | 1 | 1 | 1 |
| **4** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | -- | 1 | 1 | 1 | 1 |
| **5** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | -- | 1 | 1 | 1 |
| **6** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | -- | 1 | 1 |
| **7** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | 1 | -- | 1 |
| **8** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | 1 | 1 | -- |

- **Some Zeros**: Matrix contains some $0$s (specifically Column 2, node 2 is a transmitter).
- **No Symmetric Zeros**: For any pair $\{u, v\}$, we never have *both* $D^r_{uv}=0$ and $D^r_{vu}=0$. At least one direction is always $1$.
:::

::::

---

## Case 3: Weakly Connected Graphs

At least one pair of nodes is completely mutually unreachable via directed paths.

:::: {.columns align=center}

::: {.column width="50%"}
### Weakly Connected Graph
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
ggraph(gr_weak, layout = 'circle') +
  geom_edge_link(color = "steelblue", edge_width = 1.15,
                 arrow = arrow(length = unit(3, 'mm')),
                 end_cap = circle(5, 'mm')) +
  geom_node_point(color = "tan2", size = 14) +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::: {.column width="50%"}
### Reachability Matrix ($\mathbf{D^r}$)
| | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **1** | -- | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 |
| **2** | 1 | -- | 1 | 1 | 1 | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 |
| **3** | 1 | <span style="color:red; font-weight:bold;">0</span> | -- | 1 | 1 | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 |
| **4** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | -- | 1 | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 |
| **5** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | -- | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 |
| **6** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | -- | <span style="color:red; font-weight:bold;">0</span> | 1 |
| **7** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | 1 | -- | 1 |
| **8** | 1 | <span style="color:red; font-weight:bold;">0</span> | 1 | 1 | 1 | 1 | <span style="color:red; font-weight:bold;">0</span> | -- |

- **Mutually Unreachable Pair**:
  - Node 2 can't reach 7 ($D^r_{27}=0$), and Node 7 can't reach 2 ($D^r_{72}=0$).
  - This **symmetric zero** ($D^r_{27} = D^r_{72} = 0$) proves they are mutually unreachable, making the graph **weakly connected**.
:::

::::

# Part 3: Graph Connectedness & Geodesic Distances

---

## Measuring Graph Connectedness ($C$)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph Connectedness ($C$)** measures the overall capacity of a network to facilitate connections.
- It is defined as the **density of the reachability matrix** (excluding diagonal cells):
  - $$C = \frac{\sum_{i \neq j} D^r_{ij}}{N(N-1)}$$
- **For our weakly connected 8-node graph**:
  - Sum of all cells containing a $1$ is **48**.
  - Total possible off-diagonal cells: $8(7) = 56$.
  - $$C = \frac{48}{56} \approx 0.86$$
:::
:::

::: {.column width="40%"}
### Reachability Density
| Value | Meaning |
| :--- | :--- |
| **$C = 1.0$** | **Strongly connected** network (perfect reachability, all-ones matrix). |
| **$C < 1.0$** | Unilaterally or weakly connected network (contains zeros). |
| **$C = 0.0$** | Completely disconnected network (all zeros). |
:::

::::

# Part 4: The Geodesic Distance Matrix Step-by-Step

---

## What is Geodesic Distance?

::: {.incremental}
- While reachability tells us *if* node $i$ can reach node $j$, the **Geodesic Distance Matrix ($\mathbf{D^g}$)** tells us **how far** apart they are.
- **Geodesic Distance**:
  - The length of the shortest path (in number of edges) connecting two vertices.
- **Matrix Rules**:
  - Adjacent nodes get a geodesic distance of **1.0**.
  - Nodes are at distance **0.0** from themselves: $D^g_{ii} = 0$ for all diagonal cells.
  - If no directed or indirect path connects them, their geodesic distance is **Infinity ($\infty$)** or *Inf*.
:::

---

## Step 1: Distance Matrix (Undirected) - Adjacencies

We start with our undirected reference graph. Nodes with direct connections get a geodesic distance of **1.0**.

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_undir_highlight(active_edges = c("A-C", "A-D", "B-E", "C-D", "C-E", "D-E"))
```
:::

::: {.column width="50%"}
### Matrix Progress
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | 0 | ? | **1** | **1** | ? |
| **B** | ? | 0 | ? | ? | **1** |
| **C** | **1** | ? | 0 | **1** | **1** |
| **D** | **1** | ? | **1** | 0 | **1** |
| **E** | ? | **1** | **1** | **1** | 0 |

- Symmetrical connections $a_{AC} = a_{CA} = 1.0$, etc.
- Diagonals are initialized to **0** because $d(v, v) = 0$.
:::

::::

---

## Step 2: Distance Matrix (Undirected) - Paths of Length 2

For non-adjacent nodes, we calculate the length of the shortest path connecting them.

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_undir_highlight(active_edges = c("B-E", "C-E", "D-E", "A-C"), active_nodes = c("A", "B", "C", "D", "E"))
```
:::

::: {.column width="50%"}
### Matrix Progress
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | 0 | ? | 1 | 1 | **2** |
| **B** | ? | 0 | **2** | **2** | 1 |
| **C** | 1 | **2** | 0 | 1 | 1 |
| **D** | 1 | **2** | 1 | 0 | 1 |
| **E** | **2** | 1 | 1 | 1 | 0 |

- **B to C**: Shortest path is $B \rightarrow E \rightarrow C$ (length **2**).
- **B to D**: Shortest path is $B \rightarrow E \rightarrow D$ (length **2**).
- **A to E**: Shortest path is $A \rightarrow C \rightarrow E$ (length **2**).
:::

::::

---

## Step 3: Distance Matrix (Undirected) - Paths of Length 3

We compute the remaining non-adjacent pair, representing the maximum distance in the graph.

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_undir_highlight(active_edges = c("A-C", "C-E", "B-E"), active_nodes = c("A", "B"))
```
:::

::: {.column width="50%"}
### Symmetric Distance Matrix ($\mathbf{D^g}$)
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | 0 | **3** | 1 | 1 | 2 |
| **B** | **3** | 0 | 2 | 2 | 1 |
| **C** | 1 | 2 | 0 | 1 | 1 |
| **D** | 1 | 2 | 1 | 0 | 1 |
| **E** | 2 | 1 | 1 | 1 | 0 |

- **A to B**: Shortest path is $A \rightarrow C \rightarrow E \rightarrow B$ (length **3**).
- This completed matrix is perfectly symmetric!
:::

::::

---

## Step 4: Distance Matrix (Directed) - Paths & Connections

For directed graphs, we follow the same process but respect the arrows (Sender $\rightarrow$ Receiver).

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_dir_highlight_geo()
```
:::

::: {.column width="50%"}
### Matrix Progress
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | 0 | ? | **1** | **1** | **2** |
| **B** | ? | 0 | ? | **2** | **1** |
| **C** | ? | ? | 0 | **2** | **1** |
| **D** | ? | ? | ? | 0 | ? |
| **E** | ? | ? | ? | **1** | 0 |

- **Direct links**: $A \rightarrow C = 1, A \rightarrow D = 1, B \rightarrow E = 1, C \rightarrow E = 1, E \rightarrow D = 1$.
- **Path of length 2**: $A \rightarrow C \rightarrow E$ (length 2), $B \rightarrow E \rightarrow D$ (length 2), $C \rightarrow E \rightarrow D$ (length 2).
:::

::::

---

## Step 5: Distance Matrix (Directed) - Unreachable Pairs

If a node $i$ **cannot** reach node $j$ via any directed path, their geodesic distance is **Infinity ($\infty$)**.

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_dir_highlight_geo(active_nodes = "D")
```
:::

::: {.column width="50%"}
### Asymmetric Distance Matrix ($\mathbf{D^g}$)
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | 0 | **$\infty$** | 1 | 1 | 2 |
| **B** | **$\infty$** | 0 | **$\infty$** | 2 | 1 |
| **C** | **$\infty$** | **$\infty$** | 0 | 2 | 1 |
| **D** | **$\infty$** | **$\infty$** | **$\infty$** | 0 | **$\infty$** |
| **E** | **$\infty$** | **$\infty$** | **$\infty$** | 1 | 0 |

- **Unreachable**: e.g., $B$ can't reach $A \rightarrow d_{BA} = \infty$.
- **Pure Sink (D)**: Row D is full of $\infty$ cells since node D can't reach anyone.
:::

::::

# Part 5: Node Eccentricity and Graph Diameter

---

## Node Eccentricity and Graph Diameter

::: {.incremental}
- We can extract key distance-based centralities directly from our completed Geodesic Distance Matrix:
- **Node Eccentricity ($e(v)$)**:
  - The **maximum geodesic distance** between node $v$ and any other node in the graph:
  - $$e(v) = \max_{j \neq v} D^g_{vj}$$
  - Measures the "worst-case" distance for information to reach or leave node $v$. Nodes with low eccentricity are highly central.
- **Graph Diameter ($d$)**:
  - The **maximum eccentricity** among all nodes in the graph (the longest shortest path):
  - $$d = \max_v e(v) = \max_{i, j} D^g_{ij}$$
:::

---

## Visualizing Node Eccentricity

Eccentricity is calculated as the maximum value in each row of the Geodesic Distance Matrix (excluding diagonal 0s).

:::: {.columns align=center}

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_undir_highlight()
```
:::

::: {.column width="50%"}
### Distance Matrix with Eccentricity ($e$)
| | A | B | C | D | E | **Row Max ($e$)** |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | 0 | <span style="color:red; font-weight:bold;">3</span> | 1 | 1 | 2 | **$e(A) = 3$** |
| **B** | <span style="color:red; font-weight:bold;">3</span> | 0 | 2 | 2 | 1 | **$e(B) = 3$** |
| **C** | 1 | <span style="color:orange; font-weight:bold;">2</span> | 0 | 1 | 1 | **$e(C) = 2$** |
| **D** | 1 | <span style="color:orange; font-weight:bold;">2</span> | 1 | 0 | 1 | **$e(D) = 2$** |
| **E** | <span style="color:orange; font-weight:bold;">2</span> | 1 | 1 | 1 | 0 | **$e(E) = 2$** |

- **Graph Diameter ($d$)**: The maximum eccentricity value in the column is **3** ($d = 3$).
- Nodes C, D, E are the most central (closest "worst-case" distance of 2).
:::

::::

# Part 6: The Common-Neighbors (Neighborhood Overlap) Matrix

---

## What is Neighborhood Overlap?

::: {.incremental}
- Recall that the **neighborhood** $\mathcal{N}(i)$ is the set of vertices adjacent to node $i$.
- **Neighborhood Overlap**:
  - The number of common neighbors shared by two nodes, relative to their total number of distinct neighbors.
- **Jaccard Similarity Coefficient ($J_{ij}$)**:
  - Formally, the overlap between node $i$ and node $j$ is:
  - $$J_{ij} = \frac{|\mathcal{N}(i) \cap \mathcal{N}(j)|}{|\mathcal{N}(i) \cup \mathcal{N}(j)|} = \frac{\text{Common Neighbors}}{\text{Total Unique Neighbors}}$$
- Placing these pairwise overlap scores into a matrix yields the **Neighborhood Overlap Matrix ($\mathbf{O}$)**.
:::

---

## Visualizing Neighborhood Overlap

Overlap measures the proportion of shared neighbors (Green node) relative to unique neighbors.

:::: {.columns align=center}

::: {.column width="50%"}
### Highlighted Overlap $\{A, C\}$
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.8
#| fig-height: 4.8
plot_neighbors_highlight(active_nodes = c("A", "C"), common_nodes = "D")
```
:::

::: {.column width="50%"}
### Neighborhood Overlap Matrix ($\mathbf{O}$)
| | A | B | C | D | E |
|:---:|:---:|:---:|:---:|:---:|:---:|
| **A** | -- | 0.00 | <span style="color:red; font-weight:bold;">0.25</span> | 0.50 | 0.40 |
| **B** | 0.00 | -- | 0.33 | 0.33 | 0.33 |
| **C** | <span style="color:red; font-weight:bold;">0.25</span> | 0.33 | -- | 0.50 | 0.50 |
| **D** | 0.50 | 0.33 | 0.50 | -- | 0.50 |
| **E** | 0.40 | 0.33 | 0.50 | 0.50 | -- |

- Cell $O_{AC} = 0.25$ represents the overlap between A (red) and C (red).
- Their sole common neighbor is node D (green).
:::

::::

---

## Step-by-Step Calculation: Node $A$ and $C$

Let's compute the neighborhood overlap $O_{AC}$ for nodes $A$ and $C$ using our highlighted graph:

- **$A$'s Neighborhood**: $\mathcal{N}(A) = \{C, D\}$ (degree $k_A = 2$)
- **$C$'s Neighborhood**: $\mathcal{N}(C) = \{A, D, E\}$ (degree $k_C = 3$)

::: {.incremental}
- **Intersection (Common Neighbors)**:
  - $\mathcal{N}(A) \cap \mathcal{N}(C) = \{D\}$ (Cardinality = **1**)
  - Node $D$ is connected to both $A$ and $C$ (shown in green).
- **Union (Total Unique Neighbors)**:
  - $\mathcal{N}(A) \cup \mathcal{N}(C) = \{A, C, D, E\}$ (Cardinality = **4**)
- **Jaccard Overlap**:
  - $$O_{AC} = \frac{|\{D\}|}{|\{A, C, D, E\}|} = \frac{1}{4} = 0.25$$
:::

---

## Summary of Advanced Matrices

::: {.incremental}
- **Reachability Matrix ($\mathbf{D^r}$)**:
  - Records whether node $i$ can reach $j$ via any path. Enables calculation of **Graph Connectedness ($C$)** as matrix density.
- **Geodesic Distance Matrix ($\mathbf{D^g}$)**:
  - Records shortest path lengths. Provides the basis for measuring **Node Eccentricity** (row max) and **Graph Diameter** (global max).
  - Unreachable pairs are recorded as **Infinity ($\infty$)** or *Inf*.
- **Neighborhood Overlap Matrix ($\mathbf{O}$)**:
  - Records the Jaccard overlap coefficient ($\frac{\text{intersection}}{\text{union}}$) of neighbors, capturing structural similarity.
:::
