---
title: "Matrix Algebra in Social Networks"
subtitle: "Basic Operations, Conformability, and Matrix Multiplication"
format:
  revealjs:
    incremental: false
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(expm)
library(patchwork)

# Set seed for reproducible layout of 6-node graphs
set.seed(444)

# Define exact 6x6 symmetric adjacency matrices from the Google Slides (Slide 10)
# Parties With (P)
P <- matrix(c(
  0, 1, 1, 0, 1, 0,
  1, 0, 1, 1, 0, 0,
  1, 1, 0, 1, 1, 1,
  0, 1, 1, 0, 1, 0,
  1, 0, 1, 1, 0, 1,
  0, 0, 1, 0, 1, 0
), nrow = 6, byrow = TRUE, dimnames = list(LETTERS[1:6], LETTERS[1:6]))

# Plays Games With (G)
G <- matrix(c(
  0, 1, 0, 0, 1, 1,
  1, 0, 1, 0, 0, 0,
  0, 1, 0, 1, 0, 0,
  0, 0, 1, 0, 1, 1,
  1, 0, 0, 1, 0, 1,
  1, 0, 0, 1, 1, 0
), nrow = 6, byrow = TRUE, dimnames = list(LETTERS[1:6], LETTERS[1:6]))

# Create tidygraph objects from these matrices
gr_p <- as_tbl_graph(graph_from_adjacency_matrix(P, mode = "undirected"))
gr_g <- as_tbl_graph(graph_from_adjacency_matrix(G, mode = "undirected"))

# Define a directed adjacency matrix A_dir for the Transpose and Dyad Census (Slide 21 / 38)
A_dir <- matrix(c(
  0, 0, 1, 0, 1, 0,
  1, 0, 1, 1, 0, 0,
  1, 0, 0, 1, 1, 0,
  0, 0, 0, 0, 0, 0,
  0, 0, 0, 1, 0, 1,
  0, 0, 0, 1, 1, 0
), nrow = 6, byrow = TRUE, dimnames = list(LETTERS[1:6], LETTERS[1:6]))

gr_adir <- as_tbl_graph(graph_from_adjacency_matrix(A_dir, mode = "directed"))

# Define 5-node undirected matrix for Matrix Powers (Slide 109)
A_undir_5 <- matrix(c(
  0, 1, 0, 0, 0,
  1, 0, 1, 0, 0,
  0, 1, 0, 1, 1,
  0, 0, 1, 0, 1,
  0, 0, 1, 1, 0
), nrow = 5, byrow = TRUE, dimnames = list(letters[1:5], letters[1:5]))

gr_undir_5 <- as_tbl_graph(graph_from_adjacency_matrix(A_undir_5, mode = "undirected"))

# Circular layout coordinates for 6 nodes
l_coords_6 <- matrix(c(
  0, 1,
  0.866, 0.5,
  0.866, -0.5,
  0, -1,
  -0.866, -0.5,
  -0.866, 0.5
), ncol = 2, byrow = TRUE)

# Manual coordinates for 5-node house-kite graph (Slide 109)
# Node order: a, b, c, d, e
l_coords_5 <- matrix(c(
  -1.5, 0,    # a
  -0.5, 0,    # b
  0.5, 0,     # c
  1.5, 0.7,   # d
  1.5, -0.7   # e
), ncol = 2, byrow = TRUE)

# Plots for Parties With (P) and Plays Games With (G)
p_p <- ggraph(gr_p, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 2.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  labs(title = "Parties With (P)") +
  theme_graph() + 
  theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5)) +
  coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")

p_g <- ggraph(gr_g, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "firebrick", width = 2.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  labs(title = "Plays Games With (G)") +
  theme_graph() + 
  theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5)) +
  coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```

# Part 1: Basic Operations & Network Overlap

---

## Combining Social Relations

- In social network analysis, we often record **multiple types of relationships** for the same set of nodes (e.g., *Parties With* vs. *Plays Games With*).
- How do we combine, subtract, or intersect these overlapping structures?
- We use **Basic Matrix Operations** performed cell-by-cell on adjacency matrices of the same dimensions:
  1. **Matrix Addition ($\mathbf{P} + \mathbf{G}$)**: Measures overall relationship volume (uniplex vs. multiplex).
  2. **Subtracting from a Matrix ($\mathbf{1} - \mathbf{P}$)**: Flips ties to find the **graph complement** (*Does Not Party With*).
  3. **Hadamard Product ($\mathbf{P} \circ \mathbf{G}$)**: Extracts multiplex overlap (*Parties and Plays Games With*).

---

## Adjacency Matrices: $\mathbf{P}$ and $\mathbf{G}$

::: {.columns}
::: {.column width="50%"}
### Parties With ($\mathbf{P}$)
```{r}
#| echo: false
options(knitr.kable.NA = '--')
P_display <- P
diag(P_display) <- NA
kbl(P_display, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_p
```
</center>
:::

::: {.column width="50%"}
### Plays Games With ($\mathbf{G}$)
```{r}
#| echo: false
G_display <- G
diag(G_display) <- NA
kbl(G_display, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_g
```
</center>
:::
:::

---

## Step-by-Step Addition ($\mathbf{P} + \mathbf{G}$): Setup

- **Addition ($s_{ij} = p_{ij} + g_{ij}$)** adds corresponding cells of each matrix.
- Let's trace how the resulting sum matrix is constructed from the first two:
  - $s_{\text{A, B}} = p_{\text{A, B}} + g_{\text{A, B}} = 1 + 1 = \mathbf{2}$ (Multiplex)
  - $s_{\text{A, C}} = p_{\text{A, C}} + g_{\text{A, C}} = 1 + 0 = \mathbf{1}$ (Uniplex)

. . .

$$\mathbf{P} + \mathbf{G} = \begin{pmatrix} 0 & 1 & 1 & 0 & 1 & 0 \\ 1 & 0 & 1 & 1 & 0 & 0 \\ 1 & 1 & 0 & 1 & 1 & 1 \\ 0 & 1 & 1 & 0 & 1 & 0 \\ 1 & 0 & 1 & 1 & 0 & 1 \\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix} + \begin{pmatrix} 0 & 1 & 0 & 0 & 1 & 1 \\ 1 & 0 & 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 1 & 0 & 0 \\ 0 & 0 & 1 & 0 & 1 & 1 \\ 1 & 0 & 0 & 1 & 0 & 1 \\ 1 & 0 & 0 & 1 & 1 & 0 \end{pmatrix}$$

---

## Step-by-Step Addition ($\mathbf{P} + \mathbf{G}$): Calculation

- We add the entries cell-by-cell to get the final sum matrix:

$$\mathbf{P} + \mathbf{G} = \begin{pmatrix} 0+0 & 1+1 & 1+0 & 0+0 & 1+1 & 0+1 \\ 1+1 & 0+0 & 1+1 & 1+0 & 0+0 & 0+0 \\ 1+0 & 1+1 & 0+0 & 1+1 & 1+0 & 1+0 \\ 0+0 & 1+0 & 1+1 & 0+0 & 1+1 & 0+1 \\ 1+1 & 0+0 & 1+0 & 1+1 & 0+0 & 1+1 \\ 0+1 & 0+0 & 1+0 & 0+1 & 1+1 & 0+0 \end{pmatrix}$$

. . .

$$\mathbf{P} + \mathbf{G} = \begin{pmatrix} 0 & \mathbf{2} & 1 & 0 & \mathbf{2} & 1 \\ \mathbf{2} & 0 & \mathbf{2} & 1 & 0 & 0 \\ 1 & \mathbf{2} & 0 & \mathbf{2} & 1 & 1 \\ 0 & 1 & \mathbf{2} & 0 & \mathbf{2} & 1 \\ \mathbf{2} & 0 & 1 & \mathbf{2} & 0 & \mathbf{2} \\ 1 & 0 & 1 & 1 & \mathbf{2} & 0 \end{pmatrix}$$

- **Sociological Interpretation**: 
  - Value **2** represents a **multiplex tie** (connected by *both* relationships).
  - Value **1** represents a **uniplex tie** (connected by *exactly one* relationship).
  - Value **0** represents **no relationship**.

---

## Visualizing Matrix Addition ($\mathbf{P} + \mathbf{G}$)

::: {.columns}
::: {.column width="50%"}
### Parties or Plays Games With or Both
```{r}
#| echo: false
P_plus_G <- P + G
diag(P_plus_G) <- NA
kbl(P_plus_G, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- **1 (Blue)** = Uniplex bond.
- **2 (Purple)** = Multiplex bond.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_sum <- graph_from_adjacency_matrix(P + G, mode = "undirected", weighted = TRUE)
ggraph(gr_sum, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(aes(color = as.factor(weight), edge_width = weight*1.5)) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  scale_edge_color_manual(values = c("1" = "steelblue", "2" = "purple")) +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
</center>
:::
:::

---

## Subtracting a Number From a Matrix ($\mathbf{1} - \mathbf{P}$): Setup

- We can subtract an adjacency matrix from 1 (excluding the diagonal) to find the **graph complement**.
- Subtracting from 1 flips 1s into 0s and 0s into 1s:
  - $c_{ij} = 1 - p_{ij}$

. . .

$$\mathbf{1} - \mathbf{P} = 1 - \begin{pmatrix} - & 1 & 1 & 0 & 1 & 0 \\ 1 & - & 1 & 1 & 0 & 0 \\ 1 & 1 & - & 1 & 1 & 1 \\ 0 & 1 & 1 & - & 1 & 0 \\ 1 & 0 & 1 & 1 & - & 1 \\ 0 & 0 & 1 & 0 & 1 & - \end{pmatrix}$$

---

## Subtracting a Number From a Matrix ($\mathbf{1} - \mathbf{P}$): Calculation

- Subtracting each cell from 1 yields:

$$\mathbf{1} - \mathbf{P} = \begin{pmatrix} - & 1-1 & 1-1 & 1-0 & 1-1 & 1-0 \\ 1-1 & - & 1-1 & 1-1 & 1-0 & 1-0 \\ 1-1 & 1-1 & - & 1-1 & 1-1 & 1-1 \\ 1-0 & 1-1 & 1-1 & - & 1-1 & 1-0 \\ 1-1 & 1-0 & 1-1 & 1-1 & - & 1-1 \\ 1-0 & 1-0 & 1-1 & 1-0 & 1-1 & - \end{pmatrix}$$

. . .

$$\mathbf{1} - \mathbf{P} = \begin{pmatrix} - & 0 & 0 & 1 & 0 & 1 \\ 0 & - & 0 & 0 & 1 & 1 \\ 0 & 0 & - & 0 & 0 & 0 \\ 1 & 0 & 0 & - & 0 & 1 \\ 0 & 1 & 0 & 0 & - & 0 \\ 1 & 1 & 0 & 1 & 0 & - \end{pmatrix}$$$

- **Sociological Meaning**:
  - The resulting matrix records the **absence** of the original relationship.
  - The graph corresponding to this matrix is called the **complement** of the original graph (*Does Not Party With*).

---

## Visualizing Graph Complement ($\mathbf{1} - \mathbf{P}$)

::: {.columns}
::: {.column width="50%"}
### Does Not Party With
```{r}
#| echo: false
P_comp <- 1 - P
diag(P_comp) <- NA
kbl(P_comp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Holds $1$s where there was no original tie.
- Diagonal remains blocked.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_comp <- graph_from_adjacency_matrix(1 - P, mode = "undirected", diag = FALSE)
ggraph(gr_comp, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "forestgreen", width = 2.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
</center>
:::
:::

---

## Step-by-Step Hadamard Product ($\mathbf{P} \circ \mathbf{G}$): Setup

- **Hadamard / Element-Wise Product ($c_{ij} = p_{ij} \times g_{ij}$)** represents **relational intersection**:
- Let's trace how the multiplex intersection emerges from the product of the first two matrices:
  - $c_{\text{A, B}} = p_{\text{A, B}} \times g_{\text{A, B}} = 1 \times 1 = \mathbf{1}$ (Both active)
  - $c_{\text{A, C}} = p_{\text{A, C}} \times g_{\text{A, C}} = 1 \times 0 = \mathbf{0}$ (Only one active)

. . .

$$\mathbf{P} \circ \mathbf{G} = \begin{pmatrix} 0 & 1 & 1 & 0 & 1 & 0 \\ 1 & 0 & 1 & 1 & 0 & 0 \\ 1 & 1 & 0 & 1 & 1 & 1 \\ 0 & 1 & 1 & 0 & 1 & 0 \\ 1 & 0 & 1 & 1 & 0 & 1 \\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix} \circ \begin{pmatrix} 0 & 1 & 0 & 0 & 1 & 1 \\ 1 & 0 & 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 1 & 0 & 0 \\ 0 & 0 & 1 & 0 & 1 & 1 \\ 1 & 0 & 0 & 1 & 0 & 1 \\ 1 & 0 & 0 & 1 & 1 & 0 \end{pmatrix}$$

---

## Step-by-Step Hadamard Product ($\mathbf{P} \circ \mathbf{G}$): Calculation

- Multiplying corresponding entries cell-by-cell yields:

$$\mathbf{P} \circ \mathbf{G} = \begin{pmatrix} 0\times0 & 1\times1 & 1\times0 & 0\times0 & 1\times1 & 0\times1 \\ 1\times1 & 0\times0 & 1\times1 & 1\times0 & 0\times0 & 0\times0 \\ 1\times0 & 1\times1 & 0\times0 & 1\times1 & 1\times0 & 1\times0 \\ 0\times0 & 1\times0 & 1\times1 & 0\times0 & 1\times1 & 0\times1 \\ 1\times1 & 0\times0 & 1\times0 & 1\times1 & 0\times0 & 1\times1 \\ 0\times1 & 0\times0 & 1\times0 & 0\times1 & 1\times1 & 0\times0 \end{pmatrix}$$

. . .

$$\mathbf{P} \circ \mathbf{G} = \begin{pmatrix} 0 & \mathbf{1} & 0 & 0 & \mathbf{1} & 0 \\ \mathbf{1} & 0 & \mathbf{1} & 0 & 0 & 0 \\ 0 & \mathbf{1} & 0 & \mathbf{1} & 0 & 0 \\ 0 & 0 & \mathbf{1} & 0 & \mathbf{1} & 0 \\ \mathbf{1} & 0 & 0 & \mathbf{1} & 0 & \mathbf{1} \\ 0 & 0 & 0 & 0 & \mathbf{1} & 0 \end{pmatrix}$$

- **Sociological Meaning**:
  - Cell value is **1** *only if* both $p_{ij} = 1$ and $g_{ij} = 1$.
  - It extracts **multiplex ties** (the overlap/intersection) and erases all uniplex ties.

---

## Visualizing the Hadamard Product ($\mathbf{P} \circ \mathbf{G}$)

::: {.columns}
::: {.column width="50%"}
### Parties and Plays Games With
```{r}
#| echo: false
P_dot_G <- P * G
diag(P_dot_G) <- NA
kbl(P_dot_G, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Retains only the shared ties.
- All uniplex ties are erased, leaving a sparse overlap network.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_dot <- graph_from_adjacency_matrix(P * G, mode = "undirected")
ggraph(gr_dot, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "purple", width = 2.5) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
</center>
:::
:::

---

# Part 2: The Matrix Transpose & Directed Graphs

---

## The Matrix Transpose ($\mathbf{A}^T$)

- The **Transpose** of a matrix $\mathbf{A}$ is the same matrix with its rows and columns flipped:
  - Written as $\mathbf{A}^T$ or $\mathbf{A}'$.
  - **The Rule**: Row $i$ of the original matrix becomes Column $i$ of the transpose.
  - **Formally**: $a^T_{ij} = a_{ji}$ for all cells.

. . .

- **Symmetric vs. Asymmetric Adjacency Matrices**:
  - For **undirected graphs**, relationships are reciprocal by definition, so the matrix is symmetric ($\mathbf{A}^T = \mathbf{A}$).
  - For **directed graphs (digraphs)**, ties can be one-way, resulting in an asymmetric matrix ($\mathbf{A}^T \neq \mathbf{A}$).

---

## Visualizing the Matrix Transpose

::: {.columns}
::: {.column width="50%"}
### Original Adjacency Matrix ($\mathbf{A}$)
```{r}
#| echo: false
A_disp <- A_dir
diag(A_disp) <- NA
kbl(A_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
# Create directed graph layout for A
p_adir <- ggraph(gr_adir, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 1.5,
                 arrow = arrow(length = unit(3, 'mm')),
                 end_cap = circle(7.5, 'mm')) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
p_adir
```
</center>
:::

::: {.column width="50%"}
### Transposed Adjacency Matrix ($\mathbf{A}^T$)
```{r}
#| echo: false
AT_disp <- t(A_dir)
diag(AT_disp) <- NA
kbl(AT_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Row 1 of $\mathbf{A}$ `(0, 0, 1, 0, 1, 0)` is Column 1 of $\mathbf{A}^T$.
- Note how the direction of every arrow is reversed in the transpose graph!
:::
:::

---

# Part 3: Directed Dyads & The Dyad Census

---

## Dyads in Directed Graphs

- In a directed graph, any pair of nodes (a **dyad**) can have one of three configurations:
  1. **Mutual (M)**: Both directed edges are present ($i \leftrightarrow j$).
  2. **Asymmetric (A)**: Only one directed edge is present ($i \rightarrow j$ or $j \rightarrow i$).
  3. **Null (N)**: Neither directed edge is present (disconnected).
- An easy way to remember these three types of dyads is the mnemonic **MAN** (Mutual, Asymmetric, Null).

. . .

<center>
```{r}
#| echo: false
#| fig-height: 2.2
#| fig-width: 8
# Create simple 3 dyads representation
gr_m <- tbl_graph(nodes = data.frame(name = c("B", "E")), edges = data.frame(from = c(1, 2), to = c(2, 1)), directed = TRUE)
gr_a <- tbl_graph(nodes = data.frame(name = c("C", "E")), edges = data.frame(from = c(1), to = c(2)), directed = TRUE)
gr_n <- tbl_graph(nodes = data.frame(name = c("A", "E")), edges = data.frame(from = numeric(0), to = numeric(0)), directed = TRUE)

p1 <- ggraph(gr_m, layout = "linear") +
  geom_edge_link(color = "steelblue", width = 1.5, arrow = arrow(length = unit(3, 'mm')), end_cap = circle(4, 'mm')) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Mutual (M)") + theme_graph() + 
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  coord_cartesian(xlim = c(0.6, 2.4), clip = "off")

p2 <- ggraph(gr_a, layout = "linear") +
  geom_edge_link(color = "steelblue", width = 1.5, arrow = arrow(length = unit(3, 'mm')), end_cap = circle(4, 'mm')) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Asymmetric (A)") + theme_graph() + 
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  coord_cartesian(xlim = c(0.6, 2.4), clip = "off")

p3 <- ggraph(gr_n, layout = "linear") +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  labs(title = "Null (N)") + theme_graph() + 
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  coord_cartesian(xlim = c(0.6, 2.4), clip = "off")

p1 + p2 + p3
```
</center>

---

## The Dyad Census

- The **Dyad Census** is a count of the total number of Mutuals ($N_M$), Asymmetrics ($N_A$), and Nulls ($N_N$) in a directed graph.
- We can compute these counts systematically using cell-by-cell matrix operations (the Hadamard product and matrix subtraction) on the directed adjacency matrix $\mathbf{A}$ and its transpose $\mathbf{A}^T$!

. . .

- We will perform these calculations on our 6-node directed reference graph $\mathbf{A}$:
  - $N_M$ = number of mutual dyads
  - $N_A$ = number of asymmetric dyads
  - $N_N$ = number of null dyads

---

## 1. Mutual Dyads Matrix ($\mathbf{M} = \mathbf{A} \circ \mathbf{A}^T$): Setup

- Multiplying $\mathbf{A}$ by its transpose $\mathbf{A}^T$ element-wise extracts only mutual dyads where $a_{ij}=1$ and $a_{ji}=1$:

$$\mathbf{M} = \mathbf{A} \circ \mathbf{A}^T = \begin{pmatrix} 0 & 0 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 0 \\\\ 1 & 0 & 0 & 1 & 1 & 0 \\\\ 0 & 0 & 0 & 0 & 0 & 0 \\\\ 0 & 0 & 0 & 1 & 0 & 1 \\\\ 0 & 0 & 0 & 1 & 1 & 0 \end{pmatrix} \circ \begin{pmatrix} 0 & 1 & 1 & 0 & 0 & 0 \\\\ 0 & 0 & 0 & 0 & 0 & 0 \\\\ 1 & 1 & 0 & 0 & 0 & 0 \\\\ 0 & 1 & 1 & 0 & 1 & 1 \\\\ 1 & 0 & 1 & 0 & 0 & 1 \\\\ 0 & 0 & 0 & 0 & 1 & 0 \end{pmatrix}$$

---

## 1. Mutual Dyads Matrix ($\mathbf{M} = \mathbf{A} \circ \mathbf{A}^T$): Calculation

::: {.columns}
::: {.column width="38%"}
- We multiply the corresponding cells of $\mathbf{A}$ and $\mathbf{A}^T$:
$$\mathbf{M} = \begin{pmatrix} 0\times0 & 0\times1 & 1\times1 & 0\times0 & 1\times0 & 0\times0 \\\\ 1\times0 & 0\times0 & 1\times0 & 1\times0 & 0\times0 & 0\times0 \\\\ 1\times1 & 0\times1 & 0\times0 & 1\times0 & 1\times0 & 0\times0 \\\\ 0\times0 & 0\times1 & 0\times1 & 0\times0 & 0\times1 & 0\times1 \\\\ 0\times1 & 0\times0 & 0\times1 & 1\times0 & 0\times0 & 1\times1 \\\\ 0\times0 & 0\times0 & 0\times0 & 1\times0 & 1\times1 & 0\times0 \end{pmatrix}$$
- Resulting cell sum: 4.
- $$N_M = \frac{\sum M_{ij}}{2} = \frac{4}{2} = \mathbf{2} \text{ mutual dyads}$$
:::

::: {.column width="32%"}
### Matrix Result ($\mathbf{M}$)
```{r}
#| echo: false
M <- A_dir * t(A_dir)
M_disp <- M
diag(M_disp) <- NA
kbl(M_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
:::

::: {.column width="30%"}
### Mutual ties graph
<center>
```{r}
#| echo: false
#| fig-height: 2.8
#| fig-width: 2.8
gr_m_only <- as_tbl_graph(graph_from_adjacency_matrix(A_dir * t(A_dir), mode = "directed"))
ggraph(gr_m_only, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "purple", width = 2.0, arrow = arrow(length = unit(2.5, 'mm')), end_cap = circle(5, 'mm')) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
</center>
:::
:::

---

## 2. Asymmetric Dyads Matrix ($\mathbf{S} = \mathbf{A} \circ (\mathbf{1} - \mathbf{A}^T)$): Setup

- Multiplying $\mathbf{A}$ by the complement of transpose ($\mathbf{1} - \mathbf{A}^T$) extracts unreciprocated ties:

$$\mathbf{S} = \mathbf{A} \circ (\mathbf{1} - \mathbf{A}^T) = \begin{pmatrix} 0 & 0 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 0 \\\\ 1 & 0 & 0 & 1 & 1 & 0 \\\\ 0 & 0 & 0 & 0 & 0 & 0 \\\\ 0 & 0 & 0 & 1 & 0 & 1 \\\\ 0 & 0 & 0 & 1 & 1 & 0 \end{pmatrix} \circ \begin{pmatrix} - & 0 & 0 & 1 & 0 & 1 \\\\ 1 & - & 1 & 1 & 1 & 1 \\\\ 0 & 0 & - & 1 & 1 & 1 \\\\ 1 & 0 & 0 & - & 0 & 0 \\\\ 0 & 1 & 0 & 1 & - & 0 \\\\ 1 & 1 & 1 & 1 & 0 & - \end{pmatrix}$$

---

## 2. Asymmetric Dyads Matrix ($\mathbf{S} = \mathbf{A} \circ (\mathbf{1} - \mathbf{A}^T)$): Calculation

::: {.columns}
::: {.column width="38%"}
- We multiply the corresponding cells of $\mathbf{A}$ and $(\mathbf{1} - \mathbf{A}^T)$:
$$\mathbf{S} = \begin{pmatrix} 0\times- & 0\times0 & 1\times0 & 0\times1 & 1\times0 & 0\times1 \\\\ 1\times1 & 0\times- & 1\times1 & 1\times1 & 0\times1 & 0\times1 \\\\ 1\times0 & 0\times0 & 0\times- & 1\times1 & 1\times1 & 0\times1 \\\\ 0\times1 & 0\times0 & 0\times0 & 0\times- & 0\times0 & 0\times0 \\\\ 0\times0 & 0\times1 & 0\times0 & 1\times1 & 0\times- & 1\times0 \\\\ 0\times1 & 0\times1 & 0\times1 & 1\times1 & 1\times0 & 0\times- \end{pmatrix}$$
- Resulting cell sum: 8.
- $$N_S = \sum S_{ij} = \mathbf{8} \text{ asymmetric dyads}$$
:::

::: {.column width="32%"}
### Matrix Result ($\mathbf{S}$)
```{r}
#| echo: false
S <- A_dir * (1 - t(A_dir))
S_disp <- S
diag(S_disp) <- NA
kbl(S_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
:::

::: {.column width="30%"}
### Asymmetric ties graph
<center>
```{r}
#| echo: false
#| fig-height: 2.8
#| fig-width: 2.8
gr_s_only <- as_tbl_graph(graph_from_adjacency_matrix(A_dir * (1 - t(A_dir)), mode = "directed"))
ggraph(gr_s_only, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 1.5, arrow = arrow(length = unit(2.5, 'mm')), end_cap = circle(5, 'mm')) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
</center>
:::
:::

---

## 3. Null Dyads Matrix ($\mathbf{N} = (\mathbf{1} - \mathbf{A}) \circ (\mathbf{1} - \mathbf{A}^T)$): Setup

- Multiplying the complement of $\mathbf{A}$ by the complement of transpose ($\mathbf{1} - \mathbf{A}^T$) extracts completely disconnected dyads:

$$\mathbf{N} = (\mathbf{1} - \mathbf{A}) \circ (\mathbf{1} - \mathbf{A}^T) = \begin{pmatrix} - & 1 & 0 & 1 & 0 & 1 \\\\ 0 & - & 0 & 0 & 1 & 1 \\\\ 0 & 1 & - & 0 & 0 & 1 \\\\ 1 & 1 & 1 & - & 1 & 1 \\\\ 1 & 1 & 1 & 0 & - & 0 \\\\ 1 & 1 & 1 & 0 & 0 & - \end{pmatrix} \circ \begin{pmatrix} - & 0 & 0 & 1 & 0 & 1 \\\\ 1 & - & 1 & 1 & 1 & 1 \\\\ 0 & 0 & - & 1 & 1 & 1 \\\\ 1 & 0 & 0 & - & 0 & 0 \\\\ 0 & 1 & 0 & 1 & - & 0 \\\\ 1 & 1 & 1 & 1 & 0 & - \end{pmatrix}$$

---

## 3. Null Dyads Matrix ($\mathbf{N} = (\mathbf{1} - \mathbf{A}) \circ (\mathbf{1} - \mathbf{A}^T)$): Calculation

::: {.columns}
::: {.column width="38%"}
- We multiply the corresponding cells of $(\mathbf{1} - \mathbf{A})$ and $(\mathbf{1} - \mathbf{A}^T)$:
$$\mathbf{N} = \begin{pmatrix} -\times- & 1\times0 & 0\times0 & 1\times1 & 0\times0 & 1\times1 \\\\ 0\times1 & -\times- & 0\times1 & 0\times1 & 1\times1 & 1\times1 \\\\ 0\times0 & 1\times0 & -\times- & 0\times1 & 0\times1 & 1\times1 \\\\ 1\times1 & 1\times0 & 1\times0 & -\times- & 1\times0 & 1\times0 \\\\ 1\times0 & 1\times1 & 1\times0 & 0\times1 & -\times- & 0\times0 \\\\ 1\times1 & 1\times1 & 1\times1 & 0\times1 & 0\times0 & -\times- \end{pmatrix}$$
- Resulting cell sum: 10.
- $$N_N = \frac{\sum N_{ij}}{2} = \frac{10}{2} = \mathbf{5} \text{ null dyads}$$
:::

::: {.column width="32%"}
### Matrix Result ($\mathbf{N}$)
```{r}
#| echo: false
N_mat <- (1 - A_dir) * (1 - t(A_dir))
N_disp <- N_mat
diag(N_disp) <- NA
kbl(N_disp, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
:::

::: {.column width="30%"}
### Reference Graph (A)
<center>
```{r}
#| echo: false
#| fig-height: 2.8
#| fig-width: 2.8
p_adir_small <- ggraph(gr_adir, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "grey60", width = 1.2, arrow = arrow(length = unit(2.5, 'mm')), end_cap = circle(4.5, 'mm')) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
p_adir_small
```
</center>
:::
:::

---

## The Dyad Census Identity Check

- In any directed graph of size $n$, the total number of dyads (pairs) is $\frac{n(n-1)}{2}$.
- For our $n=6$ graph, we must have exactly $\frac{6 \times 5}{2} = 15$ dyads.

. . .

- **Check**:
  - $N_M + N_A + N_N = 2 + 8 + 5 = \mathbf{15}$
  - The mathematics accounts for every single pair in the network perfectly!

. . .

- **The Matrix Identity**:
  - The cell-wise sum of our three matrices always equals the complete graph minus the diagonal ($\mathbf{J} - \mathbf{I}$):
  - $\mathbf{M} + \mathbf{S} + \mathbf{N} = \mathbf{J} - \mathbf{I}$

---

## Graph Reciprocity ($R$)

- **Graph Reciprocity ($R$)** measures the extent to which a directed graph approximates the ideal of complete reciprocity.
- It is defined as the proportion of connected dyads that are mutual:
  $R = \frac{N_M}{N_M + N_A}$

. . .

- **Calculation**:
  - We know $N_M = 2$ and $N_A = 8$.
  - $R = \frac{2}{2 + 8} = \frac{2}{10} = \mathbf{0.20}$
- **Sociological Meaning**:
  - **20%** of the connected dyads in this network are mutual. This represents a relatively low level of reciprocity!

---

# Part 4: Conformability Rules

---

## Standard Matrix Multiplication ($\mathbf{A} \times \mathbf{B}$)

- Unlike the element-wise Hadamard product, standard matrix multiplication is **non-elementary**.
- It involves multiplying the **rows of the left matrix** by the **columns of the right matrix**.
- Because rows and columns are matched, the sizes of the two multiplier matrices must follow a strict structural rule.
- This rule is called **Conformability**.

---

## The Conformability Rule

- *You can only multiply two matrices if the number of columns of the first matrix equals the number of rows of the second matrix.*
- Put their dimensions side-by-side:

$$\mathbf{A}_{3 \times \mathbf{5}} \times \mathbf{B}_{\mathbf{5} \times 6}$$

- **Inner Dimensions**: The bold "fives" are the inner dimensions. They **must be identical** for the multiplication to be defined (**conformable**).
- **Outer Dimensions**: The "three" and "six" are the outer dimensions. They dictate the size of the resulting product matrix:

$$\mathbf{A}_{3 \times \mathbf{5}} \times \mathbf{B}_{\mathbf{5} \times 6} = \mathbf{C}_{3 \times 6}$$

---

## Conformable vs. Non-Conformable

::: {.columns}
::: {.column width="50%"}
### Conformable Example
- Line up dimensions:
  $$\mathbf{A}_{4 \times \mathbf{2}} \times \mathbf{B}_{\mathbf{2} \times 3}$$
- **Inner dimensions match** (2 = 2).
- The multiplication is fully defined.
- **Product Matrix size**:
  $$\mathbf{C}_{4 \times 3}$$
:::

::: {.column width="50%"}
### Non-Conformable Example
- Line up dimensions:
  $$\mathbf{A}_{3 \times \mathbf{4}} \times \mathbf{B}_{\mathbf{2} \times 3}$$
- **Inner dimensions clash** (4 $\neq$ 2).
- This operation is **mathematically impossible**.
- **The Visual Reason**: When you try to dot-multiply a row of 4 by a column of 2, you "run out of numbers" on the right, making the sum undefined.
:::
:::

---

## Matrix Multiplication is Non-Commutative

- In scalar math, order doesn't matter: $4 \times 3 = 3 \times 4$.
- In matrix algebra, order **completely changes the structure**:

. . .

$$\mathbf{A}_{2 \times \mathbf{3}} \times \mathbf{B}_{\mathbf{3} \times 2} = \mathbf{C}_{2 \times 2} \quad \text{(Square 2x2 Matrix)}$$

. . .

$$\mathbf{B}_{3 \times \mathbf{2}} \times \mathbf{A}_{\mathbf{2} \times 3} = \mathbf{D}_{3 \times 3} \quad \text{(Square 3x3 Matrix)}$$

- Changing the multiplication order changes both the conformability checks and the resulting network's dimensions!

---

# Part 5: Step-by-Step Multiplication Mechanics

---

## Row-by-Column Vector Multiplication

- Let's multiply a $2 \times 3$ matrix $\mathbf{A}$ by a $3 \times 2$ matrix $\mathbf{B}$ to produce a $2 \times 2$ matrix $\mathbf{C}$:

$$\mathbf{A} = \begin{pmatrix} \mathbf{3} & \mathbf{4} & \mathbf{5} \\ 7 & 6 & 3 \end{pmatrix} \quad \mathbf{B} = \begin{pmatrix} \mathbf{3} & 7 \\ \mathbf{4} & 9 \\ \mathbf{5} & 3 \end{pmatrix}$$

- To compute cell $c_{ij}$ in the target matrix, we multiply row $i$ of the left matrix by column $j$ of the right matrix, then sum the products.
- Let's trace each of the 4 cells in the target matrix $\mathbf{C}_{2 \times 2}$ progressively.

---

## Target Cell $c_{11}$ (Row 1, Column 1)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 1 of A**: $\color{red}{\{3, 4, 5\}}$
* **Column 1 of B**: $\color{blue}{\{3, 4, 5\}}$

### Calculation
$$c_{11} = (\color{red}{3} \times \color{blue}{3}) + (\color{red}{4} \times \color{blue}{4}) + (\color{red}{5} \times \color{blue}{5})$$
$$c_{11} = 9 + 16 + 25 = \mathbf{50}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} \mathbf{50} & ? \\ ? & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{12}$ (Row 1, Column 2)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 1 of A**: $\color{red}{\{3, 4, 5\}}$
* **Column 2 of B**: $\color{blue}{\{7, 9, 3\}}$

### Calculation
$$c_{12} = (\color{red}{3} \times \color{blue}{7}) + (\color{red}{4} \times \color{blue}{9}) + (\color{red}{5} \times \color{blue}{3})$$
$$c_{12} = 21 + 36 + 15 = \mathbf{72}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & \mathbf{72} \\ ? & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{21}$ (Row 2, Column 1)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 2 of A**: $\color{red}{\{7, 6, 3\}}$
* **Column 1 of B**: $\color{blue}{\{3, 4, 5\}}$

### Calculation
$$c_{21} = (\color{red}{7} \times \color{blue}{3}) + (\color{red}{6} \times \color{blue}{4}) + (\color{red}{3} \times \color{blue}{5})$$
$$c_{21} = 21 + 24 + 15 = \mathbf{60}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & 72 \\ \mathbf{60} & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{22}$ (Row 2, Column 2)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 2 of A**: $\color{red}{\{7, 6, 3\}}$
* **Column 2 of B**: $\color{blue}{\{7, 9, 3\}}$

### Calculation
$$c_{22} = (\color{red}{7} \times \color{blue}{7}) + (\color{red}{6} \times \color{blue}{9}) + (\color{red}{3} \times \color{blue}{3})$$
$$c_{22} = 49 + 54 + 9 = \mathbf{112}$$
:::

::: {.column width="50%"}
### Finished Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & 72 \\ 60 & \mathbf{112} \end{pmatrix}$$
:::
:::

---

# Part 6: Dynamic Network Applications

---

## Transpose Multiplication: Common Neighbors

- If we multiply a network's adjacency matrix $\mathbf{P}$ by its transpose $\mathbf{P}^T$:
  $$\mathbf{P} \times \mathbf{P}^T = \mathbf{C}_{\text{common}}$$
- Why does this standard matrix multiplication reveal structural overlaps?

. . .

- **The Diagonal Cells ($c_{ii}$)**: 
  - Corresponds to multiplying Row $i$ of $\mathbf{P}$ by Column $i$ of $\mathbf{P}^T$ (which is identical to Row $i$).
  - **Result**: Counts the total number of neighbors of node $i$ (the **node degree**).
- **The Off-Diagonal Cells ($c_{ij}$)**:
  - Multiplies Row $i$ (node $i$'s friendships) by Column $j$ of $\mathbf{P}^T$ (which is identical to Row $j$ of $\mathbf{P}$, node $j$'s friendships).
  - **Result**: Counts exactly how many **common neighbors** node $i$ and node $j$ share!

---

## Transpose Multiplication ($\mathbf{P} \times \mathbf{P}^T$): Setup

- Let's write out the complete multiplication of our $6 \times 6$ symmetric "Parties With" adjacency matrix $\mathbf{P}$ ($\mathbf{P}^T = \mathbf{P}$):

$$\mathbf{P} \times \mathbf{P}^T = \begin{pmatrix} 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 0 \\\\ 1 & 1 & 0 & 1 & 1 & 1 \\\\ 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 1 \\\\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix} \times \begin{pmatrix} 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 0 \\\\ 1 & 1 & 0 & 1 & 1 & 1 \\\\ 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 1 \\\\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix}$$

. . .

- To find any cell $c_{ij}$ of the resulting common neighbors matrix, we multiply **Row $i$ of $\mathbf{P}$** by **Column $j$ of $\mathbf{P}^T$** (element-wise), then sum the products.
- Let's trace how node degrees and common neighbors emerge progressively across the next slides.

---

## Transpose Multiplication ($\mathbf{P} \times \mathbf{P}^T$): Diagonal Cells

- For diagonal cells, we multiply a node's row by its own transposed column (itself). 
- Let's trace **Node C** ($c_{33}$), which corresponds to **Row 3 of $\mathbf{P}$** and **Column 3 of $\mathbf{P}^T$**:

. . .

- **Row 3 (C)**: $\color{red}{\{1, 1, 0, 1, 1, 1\}}$
- **Column 3 (C)**: $\color{blue}{\{1, 1, 0, 1, 1, 1\}}$

$$c_{33} = (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{0} \times \color{blue}{0}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1})$$
$$c_{33} = 1 + 1 + 0 + 1 + 1 + 1 = \mathbf{5}$$

- **Sociological Meaning**:
  - Cell $c_{33} = 5$ represents the **degree of Node C** (it has 5 friendships).

---

## Transpose Multiplication ($\mathbf{P} \times \mathbf{P}^T$): Off-Diagonal Cells

- For off-diagonal cells, we multiply Row $i$ (node $i$'s friends) by Column $j$ of $\mathbf{P}^T$ (node $j$'s friends).
- Let's trace common neighbors between **Node A** (Row 1) and **Node C** (Column 3):

. . .

- **Row 1 (A)**: $\color{red}{\{0, 1, 1, 0, 1, 0\}}$
- **Column 3 (C)**: $\color{blue}{\{1, 1, 0, 1, 1, 1\}}$

$$c_{\text{A, C}} = (\color{red}{0} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{0}) + (\color{red}{0} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{0} \times \color{blue}{1})$$
$$c_{\text{A, C}} = 0 + \mathbf{1} + 0 + 0 + \mathbf{1} + 0 = \mathbf{2}$$

- **Sociological Meaning**:
  - Node A and Node C share exactly **2** common neighbors.
  - The non-zero products reveal they both share friendships with **B** ($1 \times 1 = 1$) and **E** ($1 \times 1 = 1$).

---

## Transpose Multiplication ($\mathbf{P} \times \mathbf{P}^T$): Output Matrix

::: {.columns}
::: {.column width="55%"}
### Common Neighbors Matrix ($\mathbf{P} \times \mathbf{P}^T$)
$$\mathbf{P} \times \mathbf{P}^T = \begin{pmatrix} \mathbf{3} & 1 & 2 & 3 & 1 & 2 \\\\ 1 & \mathbf{3} & 2 & 1 & 3 & 1 \\\\ 2 & 2 & \mathbf{5} & 2 & 3 & 1 \\\\ 3 & 1 & 2 & \mathbf{3} & 1 & 2 \\\\ 1 & 3 & 3 & 1 & \mathbf{4} & 1 \\\\ 2 & 1 & 1 & 2 & 1 & \mathbf{2} \end{pmatrix}$$

- **Diagonals**: Node degrees $(3, 3, 5, 3, 4, 2)$.
- **Off-diagonals**: Shared neighbors counts.
:::

::: {.column width="45%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.2
#| fig-width: 4.2
ggraph(gr_p, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 2.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
- Node B and E are connected to A, C, and D, matching the cell value of **3** in the table.
</center>
:::
:::

---

## Matrix Powers ($\mathbf{P}^n$) & Paths in Networks

- Multiplying a square adjacency matrix by itself recursively reveals **indirect connections (walks)**:
  - $\mathbf{P}^2 = \mathbf{P} \times \mathbf{P}$
  - $\mathbf{P}^3 = \mathbf{P} \times \mathbf{P} \times \mathbf{P}$
- In network science, the entries of the $n^{\text{th}}$ power of an adjacency matrix have a precise graph-theoretic meaning:
  - **Off-Diagonal Cells ($p^{n}_{ij}$)**: Record the exact number of **walks of length $n$** connecting node $i$ and node $j$.
  - **Diagonal Cells ($p^{n}_{ii}$)**: Record the exact number of **cycles of length $n$** that start and end at node $i$.

---

## Walks of Length 2 ($\mathbf{P}^2$): Setup

- Let's square our $6 \times 6$ symmetric adjacency matrix $\mathbf{P}$:

$$\mathbf{P}^2 = \begin{pmatrix} 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 0 \\\\ 1 & 1 & 0 & 1 & 1 & 1 \\\\ 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 1 \\\\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix} \times \begin{pmatrix} 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 0 \\\\ 1 & 1 & 0 & 1 & 1 & 1 \\\\ 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 1 \\\\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix}$$

. . .

- Since $\mathbf{P}$ is undirected, $\mathbf{P}^T = \mathbf{P}$, making $\mathbf{P}^2$ identical to $\mathbf{P} \times \mathbf{P}^T$.
- Therefore, walks of length 2 are mathematically equivalent to **common neighbors**!
- Why? To go from $i \rightarrow k \rightarrow j$ (walk of length 2), node $k$ must be a common friend connected to both $i$ and $j$.

---

## Walks of Length 2 ($\mathbf{P}^2$): Tracing Cell $p^2_{\text{A, C}}$

- Let's compute the number of walks of length 2 between **Node A** (Row 1) and **Node C** (Column 3):
  - Row 1 (A): $\color{red}{\{0, 1, 1, 0, 1, 0\}}$
  - Column 3 (C): $\color{blue}{\{1, 1, 0, 1, 1, 1\}}$

. . .

$$p^2_{\text{A, C}} = (\color{red}{0} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{0}) + (\color{red}{0} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{0} \times \color{blue}{1}) = \mathbf{2}$$

- These two distinct walks of length 2 are:
  1. $A \rightarrow \mathbf{B} \rightarrow C$ (via common friend B)
  2. $A \rightarrow \mathbf{E} \rightarrow C$ (via common friend E)

---

## Walks of Length 2 ($\mathbf{P}^2$): Tracing Diagonal Cell $p^2_{\text{C, C}}$

- Let's compute the diagonal cell $p^2_{\text{C, C}}$ (Row 3, Column 3):
  - Row 3 (C): $\color{red}{\{1, 1, 0, 1, 1, 1\}}$
  - Column 3 (C): $\color{blue}{\{1, 1, 0, 1, 1, 1\}}$

. . .

$$p^2_{\text{C, C}} = (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{0} \times \color{blue}{0}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1}) = \mathbf{5}$$

- Walks of length 2 starting and ending at C are simply round-trips from C to its friends and back:
  - $C \rightarrow A \rightarrow C$, $C \rightarrow B \rightarrow C$, $C \rightarrow D \rightarrow C$, $C \rightarrow E \rightarrow C$, and $C \rightarrow F \rightarrow C$.
- Thus, the diagonal elements in $\mathbf{P}^2$ represent the **degrees** of each node!

---

## The Completed Squared Matrix ($\mathbf{P}^2$)

::: {.columns}
::: {.column width="55%"}
### Squared Matrix ($\mathbf{P}^2$)
$$\mathbf{P}^2 = \begin{pmatrix} \mathbf{3} & 1 & 2 & 3 & 1 & 2 \\\\ 1 & \mathbf{3} & 2 & 1 & 3 & 1 \\\\ 2 & 2 & \mathbf{5} & 2 & 3 & 1 \\\\ 3 & 1 & 2 & \mathbf{3} & 1 & 2 \\\\ 1 & 3 & 3 & 1 & \mathbf{4} & 1 \\\\ 2 & 1 & 1 & 2 & 1 & \mathbf{2} \end{pmatrix}$$

- **Diagonals**: Node degrees $(3, 3, 5, 3, 4, 2)$.
- **Off-diagonals**: Number of length-2 walks (common friends).
:::

::: {.column width="45%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.2
#| fig-width: 4.2
ggraph(gr_p, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 2.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
</center>
:::
:::

---

## Walks of Length 3 ($\mathbf{P}^3$): Setup

- To find walks of length 3, we multiply the squared matrix $\mathbf{P}^2$ by the original matrix $\mathbf{P}$:

$$\mathbf{P}^3 = \begin{pmatrix} 3 & 1 & 2 & 3 & 1 & 2 \\\\ 1 & 3 & 2 & 1 & 3 & 1 \\\\ 2 & 2 & 5 & 2 & 3 & 1 \\\\ 3 & 1 & 2 & 3 & 1 & 2 \\\\ 1 & 3 & 3 & 1 & 4 & 1 \\\\ 2 & 1 & 1 & 2 & 1 & 2 \end{pmatrix} \times \begin{pmatrix} 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 0 \\\\ 1 & 1 & 0 & 1 & 1 & 1 \\\\ 0 & 1 & 1 & 0 & 1 & 0 \\\\ 1 & 0 & 1 & 1 & 0 & 1 \\\\ 0 & 0 & 1 & 0 & 1 & 0 \end{pmatrix}$$

. . .

- **Diagonal cells ($p^3_{ii}$)**: Count length-3 cycles (triangles) starting and ending at node $i$.
- Since each triangle can be walked in two directions (clockwise & counterclockwise), the number of triangles node $i$ belongs to is:
  $$\text{Triangles}(i) = \frac{p^3_{ii}}{2}$$

---

## Walks of Length 3 ($\mathbf{P}^3$): Tracing Diagonal Cell $p^3_{\text{C, C}}$

- Let's compute the diagonal cell $p^3_{\text{C, C}}$ (Row 3, Column 3):
  - Row 3 of $\mathbf{P}^2$: $\color{red}{\{2, 2, 5, 2, 3, 1\}}$
  - Column 3 of $\mathbf{P}$: $\color{blue}{\{1, 1, 0, 1, 1, 1\}}$

. . .

$$p^3_{\text{C, C}} = (\color{red}{2} \times \color{blue}{1}) + (\color{red}{2} \times \color{blue}{1}) + (\color{red}{5} \times \color{blue}{0}) + (\color{red}{2} \times \color{blue}{1}) + (\color{red}{3} \times \color{blue}{1}) + (\color{red}{1} \times \color{blue}{1})$$
$$p^3_{\text{C, C}} = 2 + 2 + 0 + 2 + 3 + 1 = \mathbf{10}$$

- The number of triangles containing Node C is:
  $$\text{Triangles}(C) = \frac{10}{2} = \mathbf{5} \text{ triangles}$$
- Looking at the graph, C is indeed part of 5 triangles: {C,A,B}, {C,A,E}, {C,B,D}, {C,D,E}, and {C,E,F}.

---

## The Completed Cubed Matrix ($\mathbf{P}^3$)

::: {.columns}
::: {.column width="55%"}
### Cubed Matrix ($\mathbf{P}^3$)
$$\mathbf{P}^3 = \begin{pmatrix} \mathbf{4} & 8 & 10 & 4 & 10 & 3 \\\\ 8 & \mathbf{4} & 9 & 8 & 5 & 5 \\\\ 10 & 9 & \mathbf{10} & 10 & 10 & 8 \\\\ 4 & 8 & 10 & \mathbf{4} & 10 & 3 \\\\ 10 & 5 & 10 & 10 & \mathbf{6} & 7 \\\\ 3 & 5 & 8 & 3 & 7 & \mathbf{2} \end{pmatrix}$$

- **Diagonals (bold)**: Twice the number of triangles. Node degrees and triangle counts are:
  - C ($10/2 = \mathbf{5}$ triangles)
  - E ($6/2 = \mathbf{3}$ triangles)
  - F ($2/2 = \mathbf{1}$ triangle)
:::

::: {.column width="45%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.2
#| fig-width: 4.2
ggraph(gr_p, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(color = "steelblue", width = 2.0) +
  geom_node_point(size = 14, color = "tan2") +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
```
</center>
:::
:::

---

## Walks of Length 3: Weighted Graph Representation

- Plotting the network with edge sizes and color intensities proportional to the entries of $\mathbf{P}^3$ reveals distinct clique structures and **dense subgroups** (like B, C, D, E) sharing multiple indirect pathways:

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5.5
# Compute live matrix power
A3_P <- P %^% 3
m_val_P <- max(A3_P)
gr_pows_P <- graph.adjacency(A3_P, mode="undirected", weighted=TRUE) %>% 
  as_tbl_graph() 

p_pows_P <- ggraph(gr_pows_P, layout = "manual", x = l_coords_6[,1], y = l_coords_6[,2]) +
  geom_edge_link(aes(color = weight/m_val_P, width = weight/m_val_P*2.5)) +
  geom_node_point(color = "tan2", size = 14) +
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph() + 
  scale_edge_color_steps(low = "grey90", high = "purple4") +
  theme(legend.position = "none") + coord_cartesian(xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), clip = "off")
p_pows_P
```

# Part 7: Summary & Conclusion

---

## Summary of Matrix Algebraic Power

- **Basic operations combine relationships**:
  - Addition ($\mathbf{P} + \mathbf{G}$) counts uniplex vs. multiplex relationships.
  - Subtraction from 1 ($\mathbf{1} - \mathbf{P}$) flips relations to yield the graph complement (*Does Not Party With*).
  - Hadamard Product ($\mathbf{P} \circ \mathbf{G}$) extracts multiplex overlaps.
- **The Matrix Transpose ($\mathbf{A}^T$)**:
  - Flips rows and columns; essential for analyzing directed relations.
- **The Directed Dyad Census**:
  - We use $\mathbf{A}$ and $\mathbf{A}^T$ cell-by-cell to compute Mutuals ($M$), Asymmetrics ($S$), and Nulls ($N$), spelling the **MAN** census.
  - Graph Reciprocity is computed as $R = \frac{N_M}{N_M + N_A}$.
- **Conformability dictates standard multiplication**:
  - Inner dimensions must be equal.
- **Standard multiplication and powers reveal structures**:
  - $\mathbf{A} \times \mathbf{A}^T$ computes common neighbors (off-diagonals) and degrees (diagonals).
  - $\mathbf{A}^n$ maps indirect walks of length $n$. $\mathbf{A}^2$ diagonals are degrees; $\mathbf{A}^3$ diagonals are twice the number of triangles.
