---
title: "Advanced Network Matrices"
subtitle: "Beyond the Adjacency Matrix"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)

# 1. 6-node directed graph for reachability / connectedness
gr_reach <- make_graph(c("B","A", "A","C", "C","A", "C","E", "E","F", "F","E", "E","D", "F","D"), directed=TRUE) %>%
  as_tbl_graph() %>%
  mutate(name = toupper(name))

# 2. 9-node undirected graph for geodesic distance and neighborhood overlap
gr_undir <- create_empty(9, directed = FALSE) %>% 
  bind_edges(data.frame(from = 1, to = 2:5)) %>% 
  bind_edges(data.frame(from = 2, to = 3:4)) %>% 
  bind_edges(data.frame(from = 3, to = 4:5)) %>% 
  bind_edges(data.frame(from = 4, to = 5:6)) %>% 
  bind_edges(data.frame(from = 6, to = c(7:9))) %>%
  bind_edges(data.frame(from = 7, to = c(8:9))) %>%
  bind_edges(data.frame(from = 8, to = c(9))) %>%
  mutate(name = toupper(letters[1:9]))
```

# Part 1: The Reachability Matrix & Connectedness

---

## What is Reachability?

::: {.incremental}
- In directed graphs, adjacency relations (direct ties) only capture paths of **length 1**.
- **Reachability** captures both direct and indirect paths:
  - Node $j$ is **reachable** by node $i$ if there exists a directed path of any length starting at origin $i$ and terminating at destination $j$.
  - In a directed graph, if $i \rightarrow k \rightarrow j$, then $i$ can reach $j$ via a path of length 2, even if no direct tie $i \rightarrow j$ exists.
- Representing reachability systematically yields the **Reachability Matrix** ($\mathbf{D^r}$).
:::

---

## Defining the Reachability Matrix ($\mathbf{D^r}$)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Formal Definition**:
  - $$D^r_{ij} = 1 \quad \text{if node } i \text{ can reach node } j \text{ via a path of any length}$$
  - $$D^r_{ij} = 0 \quad \text{otherwise}$$
- Diagonals are blocked (`--`).
- **In the directed graph on the right**:
  - $A \rightarrow C$ is a direct link ($D^r_{AC} = 1$).
  - $A$ can reach $D$ indirectly via $A \rightarrow C \rightarrow E \rightarrow D$ ($D^r_{AD} = 1$).
  - $A$ cannot reach $B$ because no path goes against the arrow $B \rightarrow A$ ($D^r_{AB} = 0$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_reach, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.25,
                 arrow = arrow(length = unit(4, 'mm')),
                 end_cap = circle(6, 'mm')) +
  geom_node_point(color = "tan2", size = 18) +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Step-by-Step Reachability Matrix construction

```{r}
#| echo: false
d_mat <- distances(gr_reach, mode = "out")
r_mat <- ifelse(d_mat < Inf & d_mat > 0, 1, 0)
rownames(r_mat) <- V(gr_reach)$name
colnames(r_mat) <- V(gr_reach)$name
r_mat <- r_mat[order(row.names(r_mat)), order(row.names(r_mat))]
diag(r_mat) <- "--"

kbl(r_mat, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"))
```

::: {.incremental}
- **Transmitter Nodes** (e.g., $B$): Can reach others but cannot be reached. Their column is full of $0$s.
- **Receiver Nodes** (e.g., $D$): Can be reached by others but cannot reach anyone. Their row is full of $0$s.
:::

---

## Reachability & Directed Connectedness

::: {.incremental}
- The reachability matrix allows us to distinguish the **three flavors of connectedness** in directed graphs:
  1. **Strongly Connected**: Every node can reach every other node.
     - The reachability matrix is the **all-ones matrix** (excluding the diagonal).
  2. **Unilaterally Connected**: For every pair of nodes, at least one can reach the other.
     - There are some $0$s in the matrix, but no mutually disconnected pairs.
  3. **Weakly Connected**: Nodes are connected only when ignoring the directions of the arrows (via semipaths).
     - The matrix contains several $0$s representing mutually unreachable pairs.
:::

---

## Measuring Graph Connectedness ($C$)

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Graph Connectedness ($C$)** measures the proportion of node pairs that can reach one another directly or indirectly.
- It is defined as the **density of the reachability matrix**:
  - $$C = \frac{\sum_{i, j} D^r_{ij}}{N(N-1)}$$
- **Calculation on our 6-node graph**:
  - Sum of all cells containing a $1$ is **17**.
  - Total possible cells: $N(N-1) = 6(5) = 30$.
  - $$C = \frac{17}{30} \approx 0.57$$
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_reach, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.25,
                 arrow = arrow(length = unit(4, 'mm')),
                 end_cap = circle(6, 'mm')) +
  geom_node_point(color = "tan2", size = 18) +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 2: The Geodesic Distance Matrix, Eccentricity, & Diameter

---

## What is Geodesic Distance?

::: {.incremental}
- While reachability tells us *if* node $i$ can reach node $j$, the **Geodesic Distance Matrix ($\mathbf{D^g}$)** tells us **how far** apart they are.
- **Geodesic Distance**:
  - The length of the shortest path (in number of edges) connecting two vertices.
- **Matrix Rules**:
  - Adjacent nodes get a geodesic distance of **1.0**.
  - Nodes are at distance **0.0** from themselves: $D^g_{ii} = 0$ for all diagonal cells.
  - If no directed or indirect path connects them, their geodesic distance is **Infinity ($\infty$)** or *Inf*.
:::

---

## Geodesic Distance Matrix: Undirected Graph

:::: {.columns align=center}

::: {.column width="55%"}
- For undirected graphs, the distance matrix is **symmetric**:
  - $$D^g_{ij} = D^g_{ji} \quad \text{for all } i, j$$
```{r}
#| echo: false
d_undir <- distances(gr_undir)
rownames(d_undir) <- V(gr_undir)$name
colnames(d_undir) <- V(gr_undir)$name

kbl(d_undir, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 14)
```
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(color = "tan2", size = 16) +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Node Eccentricity and Graph Diameter

::: {.incremental}
- Sourcing metrics directly from the Geodesic Distance Matrix:
- **Node Eccentricity ($e(v)$)**:
  - The **maximum geodesic distance** between node $v$ and any other node in the graph:
  - $$e(v) = \max_j D^g_{vj}$$
  - Measures the "worst-case" distance for information to reach or leave node $v$.
- **Graph Diameter ($d$)**:
  - The **maximum eccentricity** among all nodes in the graph (the longest shortest path):
  - $$d = \max_v e(v)$$
  - In our 9-node graph, the maximum cell entry is **3.0** (e.g., $d_{A, G} = 3$), which represents the network's diameter.
:::

---

## Geodesic Distance Matrix: Directed Graph

:::: {.columns align=center}

::: {.column width="55%"}
- For directed graphs, the distance matrix is **asymmetric**:
  - $D^g_{ij} \neq D^g_{ji}$
  - Disconnected pairs receive **_Inf_**.
```{r}
#| echo: false
d_reach_dist <- distances(gr_reach, mode = "out")
rownames(d_reach_dist) <- V(gr_reach)$name
colnames(d_reach_dist) <- V(gr_reach)$name
d_reach_dist <- d_reach_dist[order(row.names(d_reach_dist)), order(row.names(d_reach_dist))]

kbl(d_reach_dist, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 14)
```
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_reach, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.25,
                 arrow = arrow(length = unit(4, 'mm')),
                 end_cap = circle(6, 'mm')) +
  geom_node_point(color = "tan2", size = 16) +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 3: The Common-Neighbors (Neighborhood Overlap) Matrix

---

## What is Neighborhood Overlap?

::: {.incremental}
- Recall that the **neighborhood** $\mathcal{N}(i)$ is the set of vertices adjacent to node $i$.
- **Neighborhood Overlap**:
  - The number of common neighbors shared by two nodes, relative to their total number of distinct neighbors.
- **Jaccard Similarity Coefficient ($J_{ij}$)**:
  - Formally, the overlap between node $i$ and node $j$ is:
  - $$J_{ij} = \frac{|\mathcal{N}(i) \cap \mathcal{N}(j)|}{|\mathcal{N}(i) \cup \mathcal{N}(j)|} = \frac{\text{Common Neighbors}}{\text{Total Unique Neighbors}}$$
- Placing these pairwise overlap scores into a matrix yields the **Neighborhood Overlap Matrix ($\mathbf{O}$)**.
:::

---

## Neighborhood Overlap Matrix Construction

:::: {.columns align=center}

::: {.column width="55%"}
- Always symmetric ($O_{ij} = O_{ji}$).
- Ranges from **0.00** (no shared neighbors) to **1.00** (structural equivalence).
```{r}
#| echo: false
o_mat <- similarity(gr_undir, method = "jaccard")
o_mat <- round(o_mat, 2)
rownames(o_mat) <- V(gr_undir)$name
colnames(o_mat) <- V(gr_undir)$name

kbl(o_mat, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 14)
```
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = 'kk') +
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(color = "tan2", size = 16) +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Step-by-Step Calculation: Node $A$ and $C$

- Let's compute the neighborhood overlap $O_{AC}$ for nodes $A$ and $C$:
  - **$A$'s Neighborhood**: $\mathcal{N}(A) = \{B, C, D, E\}$ (degree $k_A = 4$)
  - **$C$'s Neighborhood**: $\mathcal{N}(C) = \{A, B, D, E\}$ (degree $k_C = 4$)
- **Intersection (Common Neighbors)**:
  - $\mathcal{N}(A) \cap \mathcal{N}(C) = \{B, D, E\}$ (Cardinality = **3**)
- **Union (Total Unique Neighbors)**:
  - $\mathcal{N}(A) \cup \mathcal{N}(C) = \{A, B, C, D, E\}$ (Cardinality = **5**)
- **Jaccard Overlap**:
  - $$O_{AC} = \frac{3}{5} = 0.60$$
- This high score proves that $A$ and $C$ are highly structurally similar.

---

## Summary of Advanced Matrices

::: {.incremental}
- **Reachability Matrix ($\mathbf{D^r}$)**:
  - Records whether node $i$ can reach $j$ via any path. Enables calculation of **Graph Connectedness ($C$)** as matrix density.
- **Geodesic Distance Matrix ($\mathbf{D^g}$)**:
  - Records shortest path lengths. Provides the basis for measuring **Node Eccentricity** (row max) and **Graph Diameter** (global max).
- **Neighborhood Overlap Matrix ($\mathbf{O}$)**:
  - Records the Jaccard overlap coefficient ($\frac{\text{intersection}}{\text{union}}$) of neighbors, capturing structural similarity.
:::
