---
title: "Matrix Algebra in Social Networks"
subtitle: "Basic Operations, Conformability, and Matrix Multiplication"
format:
  revealjs:
    incremental: false
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(expm)
library(patchwork)

# Setup simple 3x3 toy networks for step-by-step builds
# Node set: A, B, C
# H (Hanging out): A-B, B-C
# C (Co-working): B-C, A-C

set.seed(111)
gr_h <- tbl_graph(
  nodes = data.frame(name = c("A", "B", "C")),
  edges = data.frame(from = c(1, 2), to = c(2, 3)),
  directed = FALSE
)

gr_c <- tbl_graph(
  nodes = data.frame(name = c("A", "B", "C")),
  edges = data.frame(from = c(2, 1), to = c(3, 3)),
  directed = FALSE
)

# Adjacency Matrices
H <- as.matrix(as_adjacency_matrix(gr_h))
C <- as.matrix(as_adjacency_matrix(gr_c))

# Simple coordinates for 3 nodes in a triangle
l_coords <- matrix(c(0, 1, -1, -1, 1, -1), ncol = 2, byrow = TRUE)

p_h <- ggraph(gr_h, layout = "manual", x = l_coords[,1], y = l_coords[,2]) +
  geom_edge_link(color = "steelblue", width = 2.0) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  labs(title = "Hanging Out (H)") +
  theme_graph() + theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))

p_c <- ggraph(gr_c, layout = "manual", x = l_coords[,1], y = l_coords[,2]) +
  geom_edge_link(color = "firebrick", width = 2.0) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  labs(title = "Co-working (C)") +
  theme_graph() + theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
```

# Part 1: Basic Operations & Network Overlap

---

## Combining Social Relations

- In social network analysis, we often record **multiple types of relationships** for the same set of nodes (e.g., *Hanging Out* vs. *Co-working*).
- How do we combine, subtract, or intersect these overlapping structures?
- We use **Basic Matrix Operations** performed cell-by-cell on adjacency matrices of the same dimensions:
  1. **Matrix Addition ($\mathbf{H} + \mathbf{C}$)**: Measures overall relationship volume.
  2. **Matrix Subtraction ($|\mathbf{H} - \mathbf{C}|$)**: Isolates relational disagreements.
  3. **Hadamard Product ($\mathbf{H} \circ \mathbf{C}$)**: Extracts multiplex overlap.

---

## Adjacency Matrices: $\mathbf{H}$ and $\mathbf{C}$

::: {.columns}
::: {.column width="50%"}
### Hanging Out ($\mathbf{H}$)
```{r}
#| echo: false
kbl(H, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_h
```
</center>
:::

::: {.column width="50%"}
### Co-working ($\mathbf{C}$)
```{r}
#| echo: false
kbl(C, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
<center>
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-width: 3.5
p_c
```
</center>
:::
:::

---

## Step-by-Step Addition ($\mathbf{H} + \mathbf{C}$)

- **Addition ($s_{ij} = h_{ij} + c_{ij}$)** adds corresponding cells:

. . .

$$\mathbf{H} + \mathbf{C} = \begin{pmatrix} 0 & 1 & 0 \\ 1 & 0 & 1 \\ 0 & 1 & 0 \end{pmatrix} + \begin{pmatrix} 0 & 0 & 1 \\ 0 & 0 & 1 \\ 1 & 1 & 0 \end{pmatrix}$$

. . .

$$\mathbf{H} + \mathbf{C} = \begin{pmatrix} 0+0 & 1+0 & 0+1 \\ 1+0 & 0+0 & 1+1 \\ 0+1 & 1+1 & 0+0 \end{pmatrix} = \begin{pmatrix} 0 & 1 & 1 \\ 1 & 0 & \mathbf{2} \\ 1 & \mathbf{2} & 0 \end{pmatrix}$$

- **Interpretation**: 
  - Value **2** represents a **multiplex tie** (both hang out and work together).
  - Value **1** represents a **uniplex tie** (either hang out or work together, but not both).

---

## Visualizing Matrix Addition ($\mathbf{H} + \mathbf{C}$)

::: {.columns}
::: {.column width="50%"}
### Resulting Sum Matrix
```{r}
#| echo: false
kbl(H + C, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- **Red/Blue** = Uniplex bonds.
- **Purple** = Multiplex bonds.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_sum <- graph_from_adjacency_matrix(H + C, mode = "undirected", weighted = TRUE)
ggraph(gr_sum, layout = "manual", x = l_coords[,1], y = l_coords[,2]) +
  geom_edge_link(aes(color = as.factor(weight), edge_width = weight*1.5)) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  scale_edge_color_manual(values = c("1" = "steelblue", "2" = "purple")) +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
</center>
:::
:::

---

## Step-by-Step Subtraction ($|\mathbf{H} - \mathbf{C}|$)

- **Absolute Subtraction ($d_{ij} = |h_{ij} - c_{ij}|$)** isolates **relational disagreement**:

. . .

$$|\mathbf{H} - \mathbf{C}| = \left| \begin{pmatrix} 0 & 1 & 0 \\ 1 & 0 & 1 \\ 0 & 1 & 0 \end{pmatrix} - \begin{pmatrix} 0 & 0 & 1 \\ 0 & 0 & 1 \\ 1 & 1 & 0 \end{pmatrix} \right|$$

. . .

$$|\mathbf{H} - \mathbf{C}| = \begin{pmatrix} |0-0| & |1-0| & |0-1| \\ |1-0| & |0-0| & |1-1| \\ |0-1| & |1-1| & |0-0| \end{pmatrix} = \begin{pmatrix} 0 & 1 & 1 \\ 1 & 0 & \mathbf{0} \\ 1 & \mathbf{0} & 0 \end{pmatrix}$$

- **Symmetric Difference**:
  - Value **1** = Connected by *exactly one* relationship.
  - Value **0** = Connected by *both* (multiplex ties cancel out) or *neither*.

---

## Visualizing Subtraction ($|\mathbf{H} - \mathbf{C}|$)

::: {.columns}
::: {.column width="50%"}
### Uniplex-Only Matrix
```{r}
#| echo: false
kbl(abs(H - C), format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Multiplex tie between **B and C** is erased.
- Only uniplex green edges remain.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_sub <- graph_from_adjacency_matrix(abs(H - C), mode = "undirected")
ggraph(gr_sub, layout = "manual", x = l_coords[,1], y = l_coords[,2]) +
  geom_edge_link(color = "forestgreen", width = 2.0) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
</center>
:::
:::

---

## Step-by-Step Hadamard Product ($\mathbf{H} \circ \mathbf{C}$)

- **Hadamard / Dot Product ($p_{ij} = h_{ij} \times c_{ij}$)** represents **relational intersection**:

. . .

$$\mathbf{H} \circ \mathbf{C} = \begin{pmatrix} 0 & 1 & 0 \\ 1 & 0 & 1 \\ 0 & 1 & 0 \end{pmatrix} \circ \begin{pmatrix} 0 & 0 & 1 \\ 0 & 0 & 1 \\ 1 & 1 & 0 \end{pmatrix}$$

. . .

$$\mathbf{H} \circ \mathbf{C} = \begin{pmatrix} 0 \times 0 & 1 \times 0 & 0 \times 1 \\ 1 \times 0 & 0 \times 0 & 1 \times 1 \\ 0 \times 1 & 1 \times 1 & 0 \times 0 \end{pmatrix} = \begin{pmatrix} 0 & 0 & 0 \\ 0 & 0 & \mathbf{1} \\ 0 & \mathbf{1} & 0 \end{pmatrix}$$

- **Sociological Meaning**:
  - Value **1** *only if* both $h_{ij} = 1$ and $c_{ij} = 1$ ($1 \times 1 = 1$).
  - Overlap is extracted; all uniplex ties are erased.

---

## Visualizing the Hadamard Product ($\mathbf{H} \circ \mathbf{C}$)

::: {.columns}
::: {.column width="50%"}
### Multiplex-Only Adjacency Matrix
```{r}
#| echo: false
kbl(H * C, format = "html", align = "c") %>% 
  column_spec(1, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 18)
```
- Retains *only* the shared co-working & hangout tie.
- Network density drops, leaving only the purple edge.
:::

::: {.column width="50%"}
<center>
```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 4.5
gr_dot <- graph_from_adjacency_matrix(H * C, mode = "undirected")
ggraph(gr_dot, layout = "manual", x = l_coords[,1], y = l_coords[,2]) +
  geom_edge_link(color = "purple", width = 2.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
</center>
:::
:::

---

# Part 2: Conformability Rules

---

## Standard Matrix Multiplication ($\mathbf{A} \times \mathbf{B}$)

- Unlike the element-wise Hadamard product, standard matrix multiplication is **non-elementary**.
- It involves multiplying the **rows of the left matrix** by the **columns of the right matrix**.
- Because rows and columns are matched, the sizes of the two multiplier matrices must follow a strict structural rule.
- This rule is called **Conformability**.

---

## The Conformability Rule

- *You can only multiply two matrices if the number of columns of the first matrix equals the number of rows of the second matrix.*
- Put their dimensions side-by-side:

$$\mathbf{A}_{3 \times \mathbf{5}} \times \mathbf{B}_{\mathbf{5} \times 6}$$

- **Inner Dimensions**: The bold "fives" are the inner dimensions. They **must be identical** for the multiplication to be defined (**conformable**).
- **Outer Dimensions**: The "three" and "six" are the outer dimensions. They dictate the size of the resulting product matrix:

$$\mathbf{A}_{3 \times \mathbf{5}} \times \mathbf{B}_{\mathbf{5} \times 6} = \mathbf{C}_{3 \times 6}$$

---

## Conformable vs. Non-Conformable

::: {.columns}
::: {.column width="50%"}
### Conformable Example
- Line up dimensions:
  $$\mathbf{A}_{4 \times \mathbf{2}} \times \mathbf{B}_{\mathbf{2} \times 3}$$
- **Inner dimensions match** (2 = 2).
- The multiplication is fully defined.
- **Product Matrix size**:
  $$\mathbf{C}_{4 \times 3}$$
:::

::: {.column width="50%"}
### Non-Conformable Example
- Line up dimensions:
  $$\mathbf{A}_{3 \times \mathbf{4}} \times \mathbf{B}_{\mathbf{2} \times 3}$$
- **Inner dimensions clash** (4 $\neq$ 2).
- This operation is **mathematically impossible**.
- **The Visual Reason**: When you try to dot-multiply a row of 4 by a column of 2, you "run out of numbers" on the right, making the sum undefined.
:::
:::

---

## Matrix Multiplication is Non-Commutative

- In scalar math, order doesn't matter: $4 \times 3 = 3 \times 4$.
- In matrix algebra, order **completely changes the structure**:

. . .

$$\mathbf{A}_{2 \times \mathbf{3}} \times \mathbf{B}_{\mathbf{3} \times 2} = \mathbf{C}_{2 \times 2} \quad \text{(Square 2x2 Matrix)}$$

. . .

$$\mathbf{B}_{3 \times \mathbf{2}} \times \mathbf{A}_{\mathbf{2} \times 3} = \mathbf{D}_{3 \times 3} \quad \text{(Square 3x3 Matrix)}$$

- Changing the multiplication order changes both the conformability checks and the resulting network's dimensions!

---

# Part 3: Step-by-Step Multiplication Mechanics

---

## Row-by-Column Vector Multiplication

- Let's multiply a $2 \times 3$ matrix $\mathbf{A}$ by a $3 \times 2$ matrix $\mathbf{B}$ to produce a $2 \times 2$ matrix $\mathbf{C}$:

$$\mathbf{A} = \begin{pmatrix} \mathbf{3} & \mathbf{4} & \mathbf{5} \\ 7 & 6 & 3 \end{pmatrix} \quad \mathbf{B} = \begin{pmatrix} \mathbf{3} & 7 \\ \mathbf{4} & 9 \\ \mathbf{5} & 3 \end{pmatrix}$$

- To compute cell $c_{ij}$ in the target matrix, we multiply row $i$ of the left matrix by column $j$ of the right matrix, then sum the products.
- Let's trace each of the 4 cells in the target matrix $\mathbf{C}_{2 \times 2}$ progressively.

---

## Target Cell $c_{11}$ (Row 1, Column 1)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 1 of A**: $\color{red}{\{3, 4, 5\}}$
* **Column 1 of B**: $\color{blue}{\{3, 4, 5\}}$

### Calculation
$$c_{11} = (\color{red}{3} \times \color{blue}{3}) + (\color{red}{4} \times \color{blue}{4}) + (\color{red}{5} \times \color{blue}{5})$$
$$c_{11} = 9 + 16 + 25 = \mathbf{50}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} \mathbf{50} & ? \\ ? & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{12}$ (Row 1, Column 2)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 1 of A**: $\color{red}{\{3, 4, 5\}}$
* **Column 2 of B**: $\color{blue}{\{7, 9, 3\}}$

### Calculation
$$c_{12} = (\color{red}{3} \times \color{blue}{7}) + (\color{red}{4} \times \color{blue}{9}) + (\color{red}{5} \times \color{blue}{3})$$
$$c_{12} = 21 + 36 + 15 = \mathbf{72}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & \mathbf{72} \\ ? & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{21}$ (Row 2, Column 1)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 2 of A**: $\color{red}{\{7, 6, 3\}}$
* **Column 1 of B**: $\color{blue}{\{3, 4, 5\}}$

### Calculation
$$c_{21} = (\color{red}{7} \times \color{blue}{3}) + (\color{red}{6} \times \color{blue}{4}) + (\color{red}{3} \times \color{blue}{5})$$
$$c_{21} = 21 + 24 + 15 = \mathbf{60}$$
:::

::: {.column width="50%"}
### Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & 72 \\ \mathbf{60} & ? \end{pmatrix}$$
:::
:::

---

## Target Cell $c_{22}$ (Row 2, Column 2)

::: {.columns}
::: {.column width="50%"}
### Highlighted Vectors
* **Row 2 of A**: $\color{red}{\{7, 6, 3\}}$
* **Column 2 of B**: $\color{blue}{\{7, 9, 3\}}$

### Calculation
$$c_{22} = (\color{red}{7} \times \color{blue}{7}) + (\color{red}{6} \times \color{blue}{9}) + (\color{red}{3} \times \color{blue}{3})$$
$$c_{22} = 49 + 54 + 9 = \mathbf{112}$$
:::

::: {.column width="50%"}
### Finished Target Matrix C
$$\mathbf{C} = \begin{pmatrix} 50 & 72 \\ 60 & \mathbf{112} \end{pmatrix}$$
:::
:::

---

# Part 4: Dynamic Network Applications

---

## Transpose Multiplication: Common Neighbors

- If we multiply a network's adjacency matrix $\mathbf{A}$ by its transpose $\mathbf{A}^T$:
  $$\mathbf{A} \times \mathbf{A}^T = \mathbf{B}$$
- Why does this matrix operation reveal structural overlaps?

. . .

- **The Diagonal Cells ($b_{ii}$)**: 
  - Corresponds to multiplying Row $i$ of $\mathbf{A}$ by Column $i$ of $\mathbf{A}^T$ (which is identical to Row $i$).
  - It squares each binary entry and sums them up.
  - **Result**: Counts the total number of neighbors of node $i$ (the **node degree**).
- **The Off-Diagonal Cells ($b_{ij}$)**:
  - Multiplies Row $i$ (node $i$'s friendships) by Row $j$ (node $j$'s friendships).
  - **Result**: Counts exactly how many **common neighbors** node $i$ and node $j$ share!

---

## Tracing Common Neighbors Step-by-Step

- Let's look at the adjacency matrix of a simple 4-node graph:

::: {.columns}
::: {.column width="50%"}
### Adjacency Matrix A
$$\mathbf{A} = \begin{pmatrix} 0 & \mathbf{1} & 0 & 1 \\ 1 & 0 & 1 & 0 \\ 0 & \mathbf{1} & 0 & 0 \\ 1 & 0 & 0 & 0 \end{pmatrix}$$
- Let's check common neighbors between **node 1 and node 3**.
:::

::: {.column width="50%"}
### Vector Multiplications
* **Row 1 (Node 1)**: $\color{red}{\{0, 1, 0, 1\}}$
* **Col 3 (Node 3)**: $\color{blue}{\{0, 1, 0, 0\}}$

### Calculation
$$\color{red}{0}\cdot\color{blue}{0} + \color{red}{1}\cdot\color{blue}{1} + \color{red}{0}\cdot\color{blue}{0} + \color{red}{1}\cdot\color{blue}{0} = \mathbf{1}$$

* Only node 2 ($1 \times 1 = 1$) is shared.
* **Takeaway**: Node 2 is the single common neighbor shared by 1 and 3.
:::
:::

---

## Matrix Powers ($\mathbf{A}^n$) & Paths in Networks

- Multiplying a square adjacency matrix by itself recursively reveals **indirect connections**:
  - $\mathbf{A}^2 = \mathbf{A} \times \mathbf{A}$
  - $\mathbf{A}^3 = \mathbf{A} \times \mathbf{A} \times \mathbf{A}$
- In network science, the entries of the $n^{th}$ power of an adjacency matrix have a precise graph-theoretic meaning:
  - **Off-Diagonal Cells ($a^{n}_{ij}$)**: Record the exact number of **indirect connections (walks) of length $n$** connecting node $i$ and node $j$.
  - **Diagonal Cells ($a^{n}_{ii}$)**: Record the exact number of **cycles of length $n$** that start and end at node $i$.

---

## Walks of Length 2: $\mathbf{A}^2$

- **The Squared Adjacency Matrix ($\mathbf{A}^2$)**:
  - **Off-diagonals** represent the number of **walks of length 2** between nodes.
  - **Diagonals** represent the number of walks of length 2 starting and ending at the same node.
  - This is equal to the **node degree** (edges incident to the node)!

. . .

### Simple Trace Example
* To go from node $i \to k \to j$ of length 2:
  * Node $k$ must be connected to *both* $i$ and $j$.
  * Thus, walks of length 2 between $i$ and $j$ is equivalent to counting their **common neighbors**!

---

## Walks of Length 3: $\mathbf{A}^3$

- **The Cubed Adjacency Matrix ($\mathbf{A}^3$)**:
  - **Off-diagonals** represent the number of **walks of length 3** between nodes.
  - **Diagonals** represent cycles of length 3 (triangles).
  - A cycle of length 3 represents a **clique of size 3**.
  - Since each triangle can be traversed in two directions (clockwise and counterclockwise), the diagonal cell divided by 2 gives the exact **number of triangles (cliques of size 3)** that node $i$ belongs to!
  - $$\text{Triangles}(i) = \frac{a^3_{ii}}{2}$$

---

## Walks of Length 3: Weighted Graph Representation

- Plotting the network with edge sizes and color intensities proportional to the entries of $\mathbf{A}^3$ reveals distinct hangout cliques sharing multiple indirect pathways:

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5.5
A3 <- H %^% 3
m_val <- max(A3)
gr_pows <- graph.adjacency(A3, mode="undirected", weighted=TRUE) %>% 
  as_tbl_graph() 

p_pows <- ggraph(gr_pows, layout = "manual", x = l_coords[1:3,1], y = l_coords[1:3,2]) +
  geom_edge_link(aes(color = weight/m_val, width = weight/m_val)) +
  geom_node_point(color = "tan2", size = 20) +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + 
  scale_edge_color_steps(low = "grey90", high = "steelblue4") +
  theme(legend.position = "none") + coord_cartesian(clip = "off")
p_pows
```

---

# Part 5: Summary & Conclusion

---

## Summary of Matrix Algebraic Power

- **Basic operations combine relationships**:
  - Addition ($\mathbf{H} + \mathbf{C}$) counts overall tie volume.
  - Subtraction ($|\mathbf{H} - \mathbf{C}|$) highlights relational disagreements.
  - Hadamard Product ($\mathbf{H} \circ \mathbf{C}$) extracts multiplex intersections.
- **Conformability dictates multiplication**:
  - Operations are only defined if inner dimensions match.
- **Multiplying by Transpose reveals common friends**:
  - $\mathbf{A} \times \mathbf{A}^T$ computes common neighbors (off-diagonals) and degrees (diagonals).
- **Matrix Powers map indirect paths**:
  - $\mathbf{A}^n$ enumerates walks of length $n$. $\mathbf{A}^2$ diagonals are degrees; $\mathbf{A}^3$ diagonals are twice the number of triangles (cliques).
