---
title: "Graph Theory: Equivalence & Similarity"
subtitle: "The Positional Structure of Social Networks"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(lsa)

# Reusable 6-node undirected graph for structural equivalence
gr_undir <- create_empty(6, directed = FALSE) %>%
  bind_edges(data.frame(from = c(1, 1, 2, 2, 3, 3, 5, 6), to = c(3, 4, 3, 4, 1, 2, 3, 5))) %>%
  mutate(name = toupper(letters[1:6]))
gr_undir <- simplify(gr_undir)

# Reusable 7-node directed graph
fr <- c("A", "A", "A", "A", "B", "B", "B", "C", "D", "D", "E", "E", "F", "G", "G", "D", "D", "B")
to <- c("C", "D", "B", "F", "A", "D", "C", "A", "A", "B", "C", "F", "A", "F", "C", "E", "G", "F")
edge_dat <- data.frame(fr, to)
node_dat <- data.frame(name = union(fr, to))
gr_dir <- tbl_graph(edges = edge_dat, nodes = node_dat)

# Reusable 15-node undirected graph for structural/automorphic/regular comparison
adj_15 <- matrix(0, 15, 15)
adj_15[2, c(1, 3, 4, 5)] <- 1
adj_15[c(1, 3, 4, 5), 2] <- 1
adj_15[2, 7] <- 1
adj_15[7, 2] <- 1
core_nodes <- 6:10
for(i in core_nodes) {
  for(j in core_nodes) {
    if(i != j) {
      adj_15[i, j] <- 1
    }
  }
}
adj_15[10, 11] <- 1
adj_15[11, 10] <- 1
adj_15[11, 12:15] <- 1
adj_15[12:15, 11] <- 1
gr_base <- graph_from_adjacency_matrix(adj_15, mode = "undirected")
V(gr_base)$name <- toupper(letters[1:15])

# Shared coordinates for 15-node symmetric layout
coords_15 <- matrix(0, 15, 2)
coords_15[2, ] <- c(-3, 0)
coords_15[1, ] <- c(-4, 1)
coords_15[3, ] <- c(-4, -1)
coords_15[4, ] <- c(-4.5, 0.5)
coords_15[5, ] <- c(-4.5, -0.5)
coords_15[7, ] <- c(-1, 0)
coords_15[6, ] <- c(0, 1)
coords_15[8, ] <- c(-0.5, -0.5)
coords_15[9, ] <- c(0.5, -0.5)
coords_15[10, ] <- c(1, 0)
coords_15[11, ] <- c(3, 0)
coords_15[12, ] <- c(4, 1)
coords_15[13, ] <- c(4, -1)
coords_15[14, ] <- c(4.5, 0.5)
coords_15[15, ] <- c(4.5, -0.5)
```

# Part 1: The Positional Approach & Structural Equivalence

---

## The Position Approach

::: {.incremental}
- In social network analysis, we can partition nodes using two different philosophies:
  1. **The Group Approach**: Nodes are divided into subsets that are tightly interconnected with one another.
  2. **The Positional Approach**: Nodes are divided into subsets based on how **similar** they are in their patterns of connectivity with others.
- Nodes in the same positional class are said to occupy the same **position** or **role** in the social structure.
- We distinguish between:
  - **Structural Equivalence**: Nodes are connected to the exact same other nodes.
  - **Regular Equivalence**: Nodes are connected to equivalent others (e.g., all teachers are connected to students, even if they are different students).
:::

---

## Defining Structural Equivalence

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Structural Equivalence** (Lorrain & White, 1971):
  - Two nodes are structurally equivalent if they are connected to the **same other nodes**.
  - Their patterns of connectivity are mathematically identical.
  - In the graph on the right:
    - $\{C, D\}$ are structurally equivalent because both are connected to $\{A, B, E\}$.
    - $\{A, B\}$ are structurally equivalent because both are connected to $\{C, D\}$.
    - $\{E\}$ and $\{F\}$ occupy **unique positions** (neighborhoods are not shared).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = 'kk') + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) +
  geom_node_point(aes(color = c("a", "a", "b", "b", "c", "d")), size = 20) + 
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
:::

::::

---

## Adjacency Matrix of Equivalent Positions

```{r}
#| echo: false
adj_mat <- matrix(as_adjacency_matrix(gr_undir), nrow = length(V(gr_undir)))
rownames(adj_mat) <- V(gr_undir)$name
colnames(adj_mat) <- V(gr_undir)$name
adj_mat <- adj_mat[order(row.names(adj_mat)), order(row.names(adj_mat))]

kbl(adj_mat, format = "html", align = "c") %>% 
  row_spec(1:2, background = "salmon", bold = TRUE) %>% 
  row_spec(3:4, background = "lightgreen", bold = TRUE) %>% 
  row_spec(5, background = "cyan", bold = TRUE) %>% 
  row_spec(6, background = "violet", bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"))
```

::: {.incremental}
- Structurally equivalent nodes have **identical rows (and columns)** in the adjacency matrix.
- Notice that Row A is identical to Row B ($0, 0, 1, 1, 0, 0$), and Row C is identical to Row D ($1, 1, 0, 0, 1, 0$).
- Nodes in unique positions (like $E$ and $F$) have completely unique row patterns.
:::

---

## Structural Similarity: A Necessary Relaxation

::: {.incremental}
- In real-world social networks, strict structural equivalence is **extremely rare**.
  - It only allows a binary "yes" or "no" answer: either nodes connect to the exact same neighbors, or they do not.
  - If node $A$ and node $B$ share 99 out of 100 neighbors, they are technically *not* structurally equivalent!
- **Structural Similarity**:
  - Allows for a "more or less" gradient rather than binary classification.
  - Measures the extent to which two nodes have **highly overlapping** patterns of connectivity.
  - Let's look at how we measure this.
:::

# Part 2: Measuring Structural Similarity

---

## Euclidean Distance ($d^{Euclid}_{k,l}$)

::: {.incremental}
- Ronald Burt (1976) proposed measuring structural similarity by calculating the **Euclidean Distance** between the row vectors of nodes in the adjacency matrix.
- For nodes $k$ and $l$, the distance is:
  - $$d^{Euclid}_{k,l} = \sqrt{\sum_j (a_{(k)j} - a_{(l)j})^2}$$
- **Key Properties**:
  - If nodes are structurally equivalent, their Euclidean distance is **exactly zero**.
  - The larger the distance, the **less structurally similar** the two nodes are.
:::

---

## Euclidean Distance Calculation: $A$ vs. $C$

- Let's calculate the Euclidean distance between $A$ ($0, 0, 1, 1, 0, 0$) and $C$ ($1, 1, 0, 0, 1, 0$):

```{r}
#| echo: false
a_vec <- c(0, 0, 1, 1, 0, 0)
c_vec <- c(1, 1, 0, 0, 1, 0)
diff_vec <- a_vec - c_vec
sq_vec <- diff_vec^2
c_diff <- c("(0 - 1)", "(0 - 1)", "(1 - 0)", "(1 - 0)", "(0 - 1)", "(0 - 0)")
calc_mat <- matrix(c(a_vec, c_vec, c_diff, diff_vec, sq_vec), nrow = 5, byrow = TRUE)
rownames(calc_mat) <- c("Vector A", "Vector C", "A - C", "Difference", "Squared Diff")
kbl(calc_mat, format = "html", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

- Sum of squared differences: $\sum_j (a_{(A)j} - a_{(C)j})^2 = 1 + 1 + 1 + 1 + 1 + 0 = 5$
- Euclidean Distance: $d^{Euclid}_{A, C} = \sqrt{5} \approx 2.24$

---

## The Structural Similarity Matrix ($S$)

```{r}
#| echo: false
eudist_func <- function(x) {
  r <- nrow(x)
  edist <- matrix(0, r, r)
  for (i in 1:r) {
    for (j in 1:r) {
      e <- x[i, ] - x[j, ]
      e <- e^2
      e <- round(sqrt(sum(e)), 2)
      edist[i, j] <- e
    }
  }
  return(edist)
}
S_mat <- eudist_func(adj_mat)
rownames(S_mat) <- rownames(adj_mat)
colnames(S_mat) <- colnames(adj_mat)
diag(S_mat) <- "--"

kbl(S_mat, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"))
```

- Pairs of structurally equivalent nodes ($\{A, B\}$ and $\{C, D\}$) have a distance of **0.00**.
- Unique positions like $F$ have high distances to almost everyone else (e.g., $d_{A, F} = 2.45$).

---

## Euclidean Distance in Directed Graphs

::: {.incremental}
- When networks contain **asymmetric (directed) ties**, we must distinguish between two behaviors:
  1. **Sending Ties**: Do nodes $u$ and $v$ point to the same out-neighbors? (Row similarity)
  2. **Receiving Ties**: Do nodes $u$ and $v$ receive ties from the same in-neighbors? (Column similarity)
- Two directed nodes are structurally equivalent if and only if they send ties to the same others **and** receive ties from the same others.
- The Euclidean distance formula expands to sum both row and column differences:
  - $$d^{Euclidean}_{k,l} = \sqrt{\sum_j (a_{(k)j}-a_{(l)j})^2 + \sum_i (a_{(k)i}-a_{(l)i})^2}$$
:::

---

## Directed Similarity Example

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- Consider the directed graph on the right.
- Nodes with the same color are structurally equivalent under directed Euclidean distance:
  - $\{E, G\}$ are structurally equivalent ($d_{E, G} = 0$). They send ties to $\{C, F\}$ and receive ties from $\{D\}$.
  - $\{C, F\}$ are structurally equivalent ($d_{C, F} = 0$). They send ties to $\{A\}$ and receive ties from $\{A, B, E, G\}$.
  - Structural equivalence in directed networks elegantly maps actors into **identical social roles**.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_dir, layout = 'tree') +
  geom_edge_parallel(color = "gray", edge_width = 1.2,
              arrow = arrow(length = unit(4, 'mm')),
              end_cap = circle(8, 'mm'), 
              sep = unit(4, 'mm')) +
  geom_node_point(aes(color = c("a", "b", "c", "d", "e", "c", "e")), size = 18) +
  geom_node_text(aes(label = name), size = 9, fontface = "bold", color = "white") +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
:::

::::

# Part 3: Measuring Similarity via Neighborhood Overlap

---

## The Intuition of Neighborhood Overlap

::: {.incremental}
- Recall that the **neighborhood** of a node is the set of all other nodes adjacent to it.
- **Neighborhood Overlap**:
  - The more neighbors two nodes have in common, the more structurally similar they are.
  - If you and your colleague share **100% of your friends**, you are structurally equivalent.
  - If you share **none of your friends**, you occupy completely different social worlds.
- We can formalize this overlap using three classical set-theoretic coefficients.
:::

---

## Three Classical Coefficients

::: {.incremental}
- Let $n_{ij}$ be the number of common neighbors of nodes $i$ and $j$.
- Let $k_i$ and $k_j$ be the degrees of nodes $i$ and $j$.
- **1. Jaccard Similarity**:
  - $$J_{ij} = \frac{n_{ij}}{k_i + k_j - n_{ij}}$$
- **2. Dice Similarity**:
  - $$D_{ij} = \frac{2n_{ij}}{k_i + k_j}$$
- **3. Cosine Similarity**:
  - $$C_{ij} = \frac{n_{ij}}{\sqrt{k_i k_j}}$$
- All three metrics range from **0.0** (no overlap) to **1.0** (structural equivalence).
:::

---

## Comparing Overlap Matrices

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 4.5
# Generate an undirected graph deterministically for similarity comparisons
set.seed(378)
gr_sim <- play_islands(n_islands = 2, size_islands = 5, p_within = 0.8, m_between = 2)
V(gr_sim)$name <- toupper(letters[1:10])

# Plot the graph
p_sim_graph <- ggraph(gr_sim, layout = 'kk') + 
  geom_edge_link(color = "steelblue", edge_width = 1.25) + 
  geom_node_point(size = 14, color = "tan2") + 
  geom_node_text(aes(label = name), size = 7, fontface = "bold", color = "white") +
  theme_graph()
p_sim_graph
```

---

## Cosine Similarity Matrix Example

```{r}
#| echo: false
adj_sim <- as.matrix(as_adjacency_matrix(gr_sim))
C_sim <- round(cosine(adj_sim), 2)
rownames(C_sim) <- V(gr_sim)$name
colnames(C_sim) <- V(gr_sim)$name
diag(C_sim) <- "--"
C_sim[lower.tri(C_sim)] <- "--"

kbl(C_sim, format = "html", align = "c") %>%
  column_spec(1, bold = TRUE) %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"), font_size = 14)
```

- Cells record Cosine Similarity between node neighborhoods.
- If $C_{ij} = 0.00$, $i$ and $j$ share absolutely zero friends.
- If $C_{ij} = 1.00$, $i$ and $j$ are structurally equivalent.

---

## Neighborhood Overlap in Directed Graphs

::: {.incremental}
- In directed networks, neighborhood overlap splits into:
  - **Out-Similarity**: Overlap in outgoing ties (pointing to the same others).
  - **In-Similarity**: Overlap in incoming ties (receiving nominations from the same others).
- **In Citation Networks** (where paper $A \rightarrow B$ means $A$ cites $B$):
  - **Bibliographic Coupling** (Out-Similarity):
    - The number of references two papers have in common.
    - Measures similarity in theoretical foundations or sources.
  - **Co-Citation** (In-Similarity):
    - The frequency with which two papers are cited together by subsequent papers.
    - Measures how much the scientific community associates their ideas.
:::

---

## Directed Cosine Equations

::: {.incremental}
- Let $n_{ij}^{out}$ be the number of shared out-neighbors, and $n_{ij}^{in}$ be the shared in-neighbors.
- Let $k_i^{out}$ and $k_i^{in}$ represent out-degrees and in-degrees.
- **Cosine Out-Similarity**:
  - $$C_{ij}^{out} = \frac{n_{ij}^{out}}{\sqrt{k_i^{out} k_j^{out}}}$$
- **Cosine In-Similarity**:
  - $$C_{ij}^{in} = \frac{n_{ij}^{in}}{\sqrt{k_i^{in} k_j^{in}}}$$
- These directed formulas allow researchers to partition information and communication networks based on highly specific sender and receiver roles.
:::

---

# Part 4: Structural, Automorphic, and Regular Equivalence

---

## Three Concepts of Equivalence

::: {.incremental}
- Beyond structural equivalence (sharing the *exact same* neighbors), there are **three nested notions** of relational position:
  1. **Structural Equivalence**: Identical neighborhoods ($N(u) = N(v)$). Highly strict, same-neighbor rule.
  2. **Automorphic Equivalence**: Globally symmetrical/interchangeable nodes. Nodes can be swapped without changing the graph structure.
  3. **Regular Equivalence**: Connection to *equivalent others*. Focuses on general roles (e.g., all doctors connect to patients).
- **The Nested Hierarchy**:
  - $\text{Structural} \subseteq \text{Automorphic} \subseteq \text{Regular}$
- Let's explore these definitions using a symmetrical 15-node network.
:::

---

## Symmetrical 15-Node Network

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- This 15-node network has:
  - Two star-like components centered at hubs $B$ and $K$.
  - Symmetrical leaf sets $\{A,C,D,E\}$ and $\{L,M,N,O\}$.
  - A tightly integrated central core $\{F,G,H,I,J\}$.
  - $G$ connects the core to hub $B$; $J$ connects the core to hub $K$.
  - Let's look at how we partition this network using the three equivalence criteria.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_base, layout = coords_15) +
  geom_edge_link(color = "steelblue", edge_width = 1.1) +
  geom_node_point(color = "tan2", size = 12) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## 1. Structural Equivalence

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- **Structural Equivalence Partition**:
  - Requires connection to the **exact same** neighbors.
  - Groups:
    - **Class 1**: Leaf set $\{A,C,D,E\}$ (connected only to $B$).
    - **Class 2**: Leaf set $\{L,M,N,O\}$ (connected only to $K$).
    - **All other nodes** (colored separately) are in **unique single-node classes**.
  - Note: Symmetrical hubs $B$ and $K$ connect to different leaves, so they are *not* structurally equivalent!
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
V(gr_base)$structural <- c("1", "unique_b", "1", "1", "1", "unique_f", "unique_g", "unique_h", "unique_i", "unique_j", "unique_k", "2", "2", "2", "2")
ggraph(gr_base, layout = coords_15) +
  geom_edge_link(color = "steelblue", alpha = 0.5, edge_width = 1.0) +
  geom_node_point(aes(color = structural), size = 12) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  scale_color_brewer(palette = "Set1") +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
:::

::::

---

## 2. Automorphic Equivalence

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- **Automorphic Equivalence Partition**:
  - Nodes that are **globally interchangeable** (symmetric swap).
  - groups nodes into **five classes**:
    - **Leaves**: $\{A,C,D,E, L,M,N,O\}$
    - **Hubs**: $\{B,K\}$ (symmetric counterparts)
    - **Core Bridges**: $\{G,J\}$ (gates to hubs)
    - **Core Mid-points**: $\{H,I\}$ (internally symmetric)
    - **Core Top**: $\{F\}$ (symmetric only to itself)
  - Successfully captures the structural similarity between hubs $B$ and $K$.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
V(gr_base)$automorphic <- c("1", "2", "1", "1", "1", "5", "3", "4", "4", "3", "2", "1", "1", "1", "1")
ggraph(gr_base, layout = coords_15) +
  geom_edge_link(color = "steelblue", alpha = 0.5, edge_width = 1.0) +
  geom_node_point(aes(color = automorphic), size = 12) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  scale_color_brewer(palette = "Set2") +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
:::

::::

---

## 3. Regular Equivalence

:::: {.columns align=center}

::: {.column width="55%"}
::: {.incremental}
- **Regular Equivalence Partition**:
  - Connects to **equivalent others** (general roles).
  - Groups the 15 nodes into **four roles**:
    - **Class 1 (Leaves)**: $\{A,C,D,E, L,M,N,O\}$ (connect *only* to Hubs)
    - **Class 2 (Hubs)**: $\{B,K\}$ (connect to Leaves and Core Bridges)
    - **Class 3 (Core Bridges)**: $\{G,J\}$ (connect to Hubs and Core nodes)
    - **Class 4 (Core Nodes)**: $\{F,H,I\}$ (connect to Core Bridges and Core nodes)
  - Elegant and highly flexible role-mapping.
:::
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
V(gr_base)$regular <- c("1", "2", "1", "1", "1", "4", "3", "4", "4", "3", "2", "1", "1", "1", "1")
ggraph(gr_base, layout = coords_15) +
  geom_edge_link(color = "steelblue", alpha = 0.5, edge_width = 1.0) +
  geom_node_point(aes(color = regular), size = 12) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  scale_color_brewer(palette = "Set3") +
  theme_graph() + theme(legend.position = "none") + coord_cartesian(clip = "off")
```
:::

::::

---

## Summary of Positional Concepts

::: {.incremental}
- **Positions represent roles**:
  - Unlike cohesive subgroups, actors do not need to connect to each other to be in the same position; they just need to connect to others similarly.
- **Structural Equivalence is the ideal limit**:
  - Perfect row/column matching is highly elegant but rare in real social networks.
- **Structural Similarity relaxes the limit**:
  - Distance metrics (Euclidean) and Overlap coefficients (Jaccard, Dice, Cosine) allow us to measure degrees of positional equivalence.
- **Directionality adds analytical depth**:
  - Directed similarity splits into In-Similarity (Co-citation) and Out-Similarity (Bibliographic coupling), providing robust tools for citation and communication analysis.
:::
