---
title: "Graph Theory: Sub-Groups & Cliques"
subtitle: "The Cohesive Structures of Social Networks"
format:
  revealjs:
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Reusable 5-node undirected graph for subgraphs/cliques
nodes_5 <- data.frame(name = c("B", "E", "C", "D", "A"), x = c(0, 1, 2, 2, 3), y = c(0, 0, 1, -1, 0))
edges_undir <- data.frame(from = c(1, 2, 2, 3, 4, 3), to = c(2, 3, 4, 5, 5, 4)) # B-E, E-C, E-D, C-A, D-A, C-D
gr_undir <- tbl_graph(nodes = nodes_5, edges = edges_undir, directed = FALSE)

# 4-node clique
gr_clique_4 <- create_complete(4) %>% mutate(name = 1:4)
# 5-node clique
gr_clique_5 <- create_complete(5) %>% mutate(name = 1:5)
```

# Part 1: Cohesion & Subgroups

---

## Levels of Analysis in Social Networks

::: {.incremental}
- In social network analysis, we navigate several levels of structure:
  1. **Individual (Micro-Node)**: Node centrality and actor attributes.
  2. **Dyads (Subgraphs of Order 2)**: Core ties and direct reciprocity.
  3. **Triads (Subgraphs of Order 3)**: Brokerage, balance, and small-group dynamics.
  4. **Subgroups, Cliques, & Communities**: Meso-level clusters of cohesive actors.
  5. **Whole Network (Macro)**: Global density, diameter, and centralization.
- The **Subgroup Level** acts as the crucial bridge between micro-interactions and macro-network systems.
:::

---

## Why Study Subgroups?

::: {.incremental}
- **Solidarity and Social Support**:
  - Subgroups are primary units of social support, shared identity, and face-to-face interaction (e.g. family cliques, support groups).
- **Functional Units in Organizations**:
  - Teams, departments, and cross-functional task forces drive productivity, collaboration, and creative innovation.
- **Engines of Influence and Conformity**:
  - High internal connectivity fosters strong social pressure, leading to high ideational homogeneity, shared values, and rapid behavioral conformity.
:::

---

## Translating "Groups" to Graph Theory

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- "Group," "community," or "clique" are intuitive but fuzzy sociological concepts.
- **Graph-Theoretic Translation**:
  - A cohesive subgroup is a relatively dense subgraph featuring more links among the subset of vertices that belong to the subgraph than links to vertices in the rest of the graph.
- We can define these subgraphs formally using criteria with varying degrees of strictness.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_undir, layout = "manual", x = x, y = y) +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

# Part 2: Cliques & Relaxed Cliques

---

## Graphs and Subgraphs: A Review

::: {.incremental}
- Recall that a subgraph $G' = (V', E')$ of a graph $G = (V, E)$ exists when:
  - $V' \subseteq V$ (the node subset)
  - $E' \subseteq E$ (the edge subset)
- A subgraph is **maximal** with respect to a given property if:
  - No additional nodes or edges from the larger graph can be added to the subgraph without losing that property.
  - e.g., a "maximal connected subgraph" is a **component**.
- We use the principle of **maximality** to define cohesive subgroups.
:::

---

## What is a Clique?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Clique**:
  - A maximally complete subgraph where every node is directly connected to every other node (density = 1.0).
  - e.g. a closed triad is a clique of size 3!
- **Size of a Clique**:
  - The number of vertices in it (clique of size 3, size 4, size 5, etc.).
  - Overlapping cliques are common; the same actor can belong to multiple cliques simultaneously, acting as a crucial bridge.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
ggraph(gr_clique_4, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Clique of Size 4") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```
:::

::::

---

## The Limitation of Strict Cliques

::: {.incremental}
- While theoretically elegant, strict cliques are **notoriously rare** in real-world social networks.
  - Social groups are often friendly and cohesive without being perfectly complete.
  - A single missing tie in a 10-person group immediately disqualifies the entire group from being a clique!
- To handle this, network analysts use **relaxed clique criteria**, focusing on distance or degree bounds.
:::

---

## $n$-Cliques: Relaxing by Distance

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- An **$n$-clique** is a maximal subgraph where the geodesic (shortest path) distance between any pair of nodes *within* the subgraph is at most $n$:
  - $$d(u, v) \le n \quad \text{for all } u, v \in V'$$
- **Types of $n$-cliques**:
  - **1-Clique**: A standard strict clique ($d_{ij} = 1$).
  - **2-Clique**: Every pair can reach each other in at most 2 steps ($d_{ij} \le 2$).
  - **3-Clique**: Every pair can reach each other in at most 3 steps ($d_{ij} \le 3$).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
# A 2-clique of size 5 (a line of 5 nodes has distance up to 4, but a circular layout with some chords is 2-clique)
gr_2clique <- tbl_graph(
  nodes = data.frame(name = 1:5),
  edges = data.frame(from = c(1, 1, 2, 3, 4), to = c(2, 3, 4, 5, 5)),
  directed = FALSE
)
ggraph(gr_2clique, layout = "circle") +
  geom_edge_link(color = "steelblue", edge_width = 1.5) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("2-Clique (Max Distance = 2)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```
:::

::::

# Part 3: Cohesive Blocking

---

## $K$-Connectivity as Social Cohesion

::: {.incremental}
- In 2003, sociologists James Moody and Douglas White proposed translating **social cohesion** directly to graph-theoretic **$k$-connectivity**:
  - A group is structurally cohesive to the extent that it can remain connected despite the removal of members.
- **The Core Intuition**:
  - The larger the **node-cutset** (the minimum number of nodes required to disconnect the subgraph), the more structurally integrated and robust the group is.
  - A group that requires removing 3 nodes to fragment is more cohesive than one that can be split by removing a single broker.
:::

---

## Cohesive Blocking Algorithm

::: {.incremental}
- Moody and White developed a recursive algorithm to identify nested, hierarchically organized cohesive subgroups (cohesive blocks):
  1. Identify the **node $k$-connectivity** of the input graph ($G$).
  2. Find and remove the **minimum $k$-node cutset** (the specific vertices that disconnect the graph).
  3. This splits the graph into separate connected subgraphs.
  4. Repeat steps 1–3 on each resulting subgraph recursively until no more subdivisions can be obtained.
:::

---

## Hierarchical Nested Cohesive Blocks

::: {.incremental}
- Cohesive blocking reveals the **structural nest** of a network:
  - The whole network might be 1-connected (vulnerable to single brokers).
  - Inside, we find 2-connected blocks (more robust).
  - Deep within those, we find highly cohesive, 3-connected cores.
- This provides an elegant quantitative map of social groups, moving from loose, sprawling associations down to highly integrated, resilient social cliques.
:::

---

## Conclusion & Summary

::: {.incremental}
- **Subgroups capture the meso-level of society**:
  - They represent departments, friendship circles, and task units.
- **Cliques are highly strict**:
  - While complete ($d = 1.0$), they are rare in the real world, necessitating relaxed definitions like **$n$-cliques** based on distance.
- **Cohesive Blocking operationalizes structural cohesion**:
  - By recursively identifying $k$-node cutsets, we map the nested, hierarchical, and robust core subgroups of a social system.
:::

---

## References {.unnumbered}
