---
title: "Ties, Dyads, and Triads"
subtitle: "The Motifs of Social Network Analysis"
format:
  revealjs:
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Perfect triangle coordinates for 3-node triad drawings
x_triad <- c(0, 2, 1)
y_triad <- c(0, 0, 1.2)

# Helper function to plot individual directed triads
gr_c <- create_empty(3, directed = TRUE)
plot_tr <- function(edges, title) {
  g <- gr_c %>% mutate(name = toupper(letters[1:3]))
  if (nrow(edges) > 0) {
    g <- g %>% bind_edges(edges)
  }
  p <- ggraph(g, layout = "manual", x = x_triad, y = y_triad) +
    geom_edge_link(color = "steelblue", edge_width = 1.1,
                   arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                   end_cap = circle(5, 'mm')) +
    geom_node_point(size = 12, color = "tan2") +
    geom_node_text(aes(label = name), size = 5, color = "white") +
    theme_graph() + ggtitle(title) + coord_cartesian(clip = "off") +
    theme(plot.title = element_text(size = 10, face="bold", hjust=0.5))
  return(p)
}
```

# Part 1: The Taxonomy of Ties

---

## The Relational Basis of Networks

::: {.incremental}
- In social network analysis, a **graph** is more than an abstract structure.
- It represents concrete **relationships** (ties) between actors.
- Not all relationships are created equal!
  - Some ties are naturally mutual (e.g., being siblings).
  - Others are directional (e.g., seeking advice).
  - Some strictly forbid mutuality (e.g., being a supervisor).
- To study networks, we must separate the **mathematical property of the relationship** (the tie type) from its **empirical realization** (the dyad or triad).
:::

---

## The Sociological Classification of Types of Ties

::: {.incremental}
- Instead of treating all connections identically, sociologists classify social ties into **four major types** ([Borgatti et al. 2009](https://doi.org/10.1126/science.1165821)):
  1. **Similarities**: Having something in common (e.g., co-location, shared attributes).
  2. **Social Relations**: Culturally recognized labels or sentiments (e.g., friend, spouse, boss).
  3. **Interactions**: Regular communication or behavior (e.g., texting, talking).
  4. **Flows**: Transmission of resources or states (e.g., virus infection, lending money).
:::

---

## States vs. Events in Social Networks

::: {.incremental}
- We can divide these four types of ties into two fundamental clusters:
- **States (Static Situations)**:
  - **Similarities** & **Social Relations**.
  - Represent continuous, long-standing conditions.
  - *Examples*: Being in the same class, being siblings, being coworker/boss.
- **Events (Discrete Activities)**:
  - **Interactions** & **Flows**.
  - Occur at specific points in time and are transient.
  - *Examples*: Texting someone, lending money, transmitting a virus.
:::

---

## 1. Symmetric Ties

::: {.incremental}
- **Definition**: A relation where connection is built-in and mutual by construction:
  - Formally: $AB \iff BA$
- **Representation**: Represented by **undirected graphs**.
- **Structural Constraint**: 
  - Pairs of actors can only be **disconnected** or **connected** by a single undirected edge.
- **Key Examples**:
  - Siblinghood, marriage, co-membership, co-presence.
:::

---

## Symmetric Ties: Visual Example

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- A single undirected edge represents a mutual relationship.
- Both actors share the tie equally and simultaneously.
- e.g., $A$ is married to $B$ implies $B$ is married to $A$.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr_sym <- create_empty(2, directed = FALSE) %>% 
  bind_edges(data.frame(from = 1, to = 2)) %>% 
  mutate(name = toupper(letters[1:2]))
ggraph(gr_sym, layout = 'linear') +
  geom_edge_link(color = "steelblue", width = 2) +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Symmetric Similarities

::: {.incremental}
- Many symmetric relations are based on **similarities** or **co-presence**:
  - *Similarities are necessarily symmetric*:
    - If you are in the same place at the same time as someone, they are there with you.
    - If you are zooming with someone, they are zooming with you.
    - If you go to UCLA with someone, they also go to UCLA with you.
- Sharing similarities creates the structural opportunity for interaction to occur.
:::

---

## 2. Asymmetric Ties

::: {.incremental}
- **Definition**: A relation with natural directionality where reciprocity is possible but *not guaranteed*:
  - Formally: $AB \implies$ nothing about $BA$
- **Representation**: Represented by **directed graphs** (digraphs).
- **Structural Constraint**:
  - Reciprocity is an **empirical event** (it may or may not occur), yielding three possible dyad configurations.
- **Key Examples**:
  - Friendship nominations, advice-seeking, texting.
:::

---

## Asymmetric Ties: Visual Example

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- A single directed arrow represents a one-way relationship.
- Reciprocity is **optional** (empirical).
- e.g., $A$ likes $B$ (solid arrow), but $B$ liking $A$ is optional (dashed arrow with question mark).
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr_asym <- create_empty(2, directed = TRUE) %>% 
  bind_edges(data.frame(from = c(1, 2), to = c(2, 1), type = c("solid", "dashed"))) %>% 
  mutate(name = toupper(letters[1:2]))
ggraph(gr_asym, layout = 'linear') +
  geom_edge_parallel(aes(linetype = type, color = type), edge_width = 1.5,
                  arrow = arrow(length = unit(5, 'mm'), type = "closed"),
                  end_cap = circle(8, 'mm'), sep = unit(4, 'mm')) +
  scale_edge_color_manual(values = c("solid" = "steelblue", "dashed" = "gray60"), guide = "none") +
  scale_edge_linetype_manual(values = c("solid" = "solid", "dashed" = "dashed"), guide = "none") +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  annotate("text", x = 1.5, y = -0.3, label = "?", size = 8, fontface = "bold", color = "gray40") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Asymmetric Interactions & Sentiments

::: {.incremental}
- Many directed ties are based on **interactions** or **sentiments**:
  - *Interactions/sentiments are naturally asymmetric*:
    - You can follow someone on Twitter, but they don't have to follow you back.
    - You can send a text to someone, but they don't have to respond.
    - You can seek advice from someone without them seeking advice from you.
- Reciprocity is therefore an empirical puzzle to explain, not a definition.
:::

---

## 3. Anti-Symmetric Ties

::: {.incremental}
- **Definition**: An inherently directional relation where reciprocity is **strictly forbidden**:
  - Formally: $AB \implies \neg BA$
- **Representation**: Represented by **directed graphs** (oriented graphs, DAGs, or trees).
- **Structural Constraint**:
  - Reciprocity is structurally impossible, yielding only two possible dyad configurations.
- **Key Examples**:
  - Parent-of, boss-of, lending money.
:::

---

## Anti-Symmetric Ties: Visual Example

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- Reciprocity is forbidden by logical or structural definition.
- e.g., $A$ is the parent of $B$ (solid arrow).
- The reverse arrow $B \rightarrow A$ is strictly forbidden (red dashed arrow crossed out with "✕").
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr_anti <- create_empty(2, directed = TRUE) %>% 
  bind_edges(data.frame(from = c(1, 2), to = c(2, 1), type = c("solid", "dashed"))) %>% 
  mutate(name = toupper(letters[1:2]))
ggraph(gr_anti, layout = 'linear') +
  geom_edge_parallel(aes(linetype = type, color = type), edge_width = 1.5,
                  arrow = arrow(length = unit(5, 'mm'), type = "closed"),
                  end_cap = circle(8, 'mm'), sep = unit(4, 'mm')) +
  scale_edge_color_manual(values = c("solid" = "steelblue", "dashed" = "red"), guide = "none") +
  scale_edge_linetype_manual(values = c("solid" = "solid", "dashed" = "dashed"), guide = "none") +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  annotate("text", x = 1.5, y = -0.3, label = "✕", size = 8, fontface = "bold", color = "red") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Anti-Symmetric Hierarchies

::: {.incremental}
- Many anti-symmetric ties are based on **formal hierarchies** or **kinship**:
  - *Hierarchical role relations are anti-symmetric*:
    - Kinship: If you are someone's parent, they cannot be your parent.
    - Formal: If you are someone's boss, they cannot be your boss at the same time.
- These relations represent irreversible flows or asymmetric structures.
:::

---

## Mapping Social Ties to Graph Representations

| Sociological Category | Relational Property | Allowed Dyads | Visual Graph representation |
|:---|:---|:---|:---|
| **Co-membership, Propinquity** | **Symmetric** | • Disconnected<br>• Connected | Undirected Graph ($A - B$) |
| **Sentiments, Advice-seeking** | **Asymmetric** | • Disconnected<br>• Non-reciprocal<br>• Reciprocal | Directed Graph ($A \rightarrow B$ or $A \leftrightarrow B$) |
| **Formal Hierarchies, Kinship** | **Anti-symmetric** | • Disconnected<br>• Non-reciprocal | Oriented Graph/Tree ($A \rightarrow B$, no cycles) |

---

## Multiplexity in Social Relations

::: {.incremental}
- In real-world social networks, actors are rarely linked by only one relationship.
- **Multiplexity** refers to the presence of multiple, distinct types of ties linking the same pair of nodes simultaneously.
  - e.g., $A$ and $B$ are coworkers (**symmetric** tie) AND $A$ is the supervisor of $B$ (**anti-symmetric** tie) AND they are friends (**asymmetric** tie).
- Multiplexity adds rich, multi-layered complexity to networks, as the behavior or pattern in one relation heavily influences and moderates others.
:::

---

## Multiplexity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Three nodes connected by multiple parallel relations:
  - **Symmetric Sibling**: (blue dotted lines).
  - **Anti-Symmetric Parent-of**: (solid black arrows).
  - **Asymmetric Advice**: (dashed purple arrows).
- *One event can affect multiple layers simultaneously.*
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6.5
#| fig-height: 5.5
edges_mult <- data.frame(
  from = c(1, 1, 2, 3, 2),
  to   = c(2, 3, 3, 2, 1),
  type = c("Parent-of", "Parent-of", "Sibling", "Sibling", "Advice")
)
gr_mult <- tbl_graph(nodes = data.frame(name = c("A", "B", "C"), x = c(1, 0, 2), y = c(1, 0, 0)), edges = edges_mult, directed = TRUE)
ggraph(gr_mult, layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = type, linetype = type),
                     arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm'),
                     sep = unit(4.5, 'mm'), edge_width = 1.25) +
  scale_edge_color_manual(values = c("Parent-of" = "black", "Sibling" = "royalblue", "Advice" = "purple")) +
  scale_edge_linetype_manual(values = c("Parent-of" = "solid", "Sibling" = "dotted", "Advice" = "dashed")) +
  geom_node_point(size = 18, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## The Growth of Social Networks

::: {.incremental}
- Social networks are highly dynamic, growing and evolving over time.
- A classic example is the formation of a new family unit:
  1. A romantic tie (an asymmetric, non-kin relation) forms between two individuals.
  2. This culminates in a marriage tie (a symmetric, kin relation).
  3. This single event triggers a cascade of new "in-law" ties (both symmetric same-generation ties and anti-symmetric cross-generation ties).
  4. This merges two separate kinship trees into a single, highly complex, interconnected network.
:::

---

## Network Growth: Visualizing the Marriage Cascade

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 5.5
  gr_g1 <- tbl_graph(
    nodes = data.frame(name = c("C", "A", "B", "D"), x = c(0, 1, 2.5, 3.5), y = c(0, 0, 0, 0)),
    edges = data.frame(from = c(1, 2, 4), to = c(2, 3, 3), type = c("kin", "rom", "kin")),
    directed = TRUE
  )
  p_g1 <- ggraph(gr_g1, layout="manual", x=x, y=y) +
    geom_edge_link(aes(color=type, linetype=type), arrow = arrow(length = unit(3, 'mm')), end_cap = circle(5, 'mm')) +
    scale_edge_color_manual(values = c("kin" = "black", "rom" = "red"), guide = "none") +
    scale_edge_linetype_manual(values = c("kin" = "solid", "rom" = "dashed"), guide = "none") +
    geom_node_point(size = 12, color = "tan2") +
    geom_node_text(aes(label = name), size = 6, color = "white") +
    theme_graph() + ggtitle("1. Romantic Tie (Event)") + coord_cartesian(clip = "off") +
    theme(plot.title = element_text(size = 11, face="bold", hjust=0.5))

  gr_g2 <- tbl_graph(
    nodes = data.frame(name = c("C", "A", "B", "D"), x = c(0, 1, 2.5, 3.5), y = c(0, 0, 0, 0)),
    edges = data.frame(from = c(1, 2, 4), to = c(2, 3, 3), type = c("kin", "mar", "kin")),
    directed = FALSE
  )
  p_g2 <- ggraph(gr_g2, layout="manual", x=x, y=y) +
    geom_edge_link(aes(color=type), edge_width = 1.2) +
    scale_edge_color_manual(values = c("kin" = "black", "mar" = "red"), guide = "none") +
    geom_node_point(size = 12, color = "tan2") +
    geom_node_text(aes(label = name), size = 6, color = "white") +
    theme_graph() + ggtitle("2. Marriage State (Tie)") + coord_cartesian(clip = "off") +
    theme(plot.title = element_text(size = 11, face="bold", hjust=0.5))

  gr_g3 <- tbl_graph(
    nodes = data.frame(name = c("C", "A", "B", "D"), x = c(0, 1, 2.5, 3.5), y = c(0, 0, 0, 0)),
    edges = data.frame(from = c(1, 2, 4, 1, 2, 4), to = c(2, 3, 3, 4, 4, 1), type = c("kin", "mar", "kin", "inlaw", "inlaw", "inlaw")),
    directed = FALSE
  )
  p_g3 <- ggraph(gr_g3, layout="manual", x=x, y=y) +
    geom_edge_link(aes(color=type, linetype=type), edge_width = 1.2) +
    scale_edge_color_manual(values = c("kin" = "black", "mar" = "red", "inlaw" = "blue"), guide = "none") +
    scale_edge_linetype_manual(values = c("kin" = "solid", "mar" = "solid", "inlaw" = "dotted"), guide = "none") +
    geom_node_point(size = 12, color = "tan2") +
    geom_node_text(aes(label = name), size = 6, color = "white") +
    theme_graph() + ggtitle("3. In-Law Cascade") + coord_cartesian(clip = "off") +
    theme(plot.title = element_text(size = 11, face="bold", hjust=0.5))

  p_g1 + p_g2 + p_g3
```

# Part 2: Dyads and Their Realizations

---

## What is a Dyad?

::: {.incremental}
- A **dyad** is any pair of nodes in a graph, whether they are joined by an edge or not.
- Mathematically, a dyad is *any subgraph of order two* ($n = 2$).
- Crucially, a dyad represents the **realized configuration** of ties between two nodes.
- **Sociological Insight**: Ties that do *not* exist (disconnected dyads) are often just as socially meaningful as those that do!
- The possible dyadic states are strictly constrained by the relational type of the tie.
:::

---

## Dyads: The Building Blocks

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- The simplest possible subgraph of order two.
- Can contain zero, one, or two directed edges depending on relational constraints.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr_dyad_base <- create_empty(2, directed = FALSE) %>% mutate(name = toupper(letters[1:2]))
ggraph(gr_dyad_base, layout = 'linear') +
  geom_node_point(size = 25, color = "tan2") +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Dyads under Symmetric Ties

::: {.incremental}
- Because undirected graphs do not allow directionality, a symmetric relation can only take on **two** dyadic states:
  1. **Disconnected Dyads** (historically called **null dyads**): The two nodes do not share an edge (size zero).
  2. **Connected Dyads**: The two nodes share an edge (size one).
- Every pair of actors in an undirected graph is either connected or disconnected.
:::

---

## Symmetric Dyads: Visual Examples

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- **Connected**: A single undirected edge (size one).
- **Disconnected**: No edge (size zero).
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr1 <- create_empty(2, directed = FALSE) %>% mutate(name = c("A", "B"))
gr2 <- create_empty(2, directed = FALSE) %>% bind_edges(data.frame(from = 1, to = 2)) %>% mutate(name = c("C", "D"))

p1 <- ggraph(gr1, layout = 'linear') +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Disconnected (Size Zero)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

p2 <- ggraph(gr2, layout = 'linear') +
  geom_edge_link(color = "steelblue", width = 1.5) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Connected (Size One)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

p1 / p2
```
:::

::::

---

## Dyads under Asymmetric Ties

::: {.incremental}
- In directed graphs, reciprocity is possible but optional.
- Thus, asymmetric ties can yield **three** distinct kinds of dyads:
  1. **Disconnected Dyads** (Null): Neither node sends a directed tie to the other (size zero).
  2. **Non-Reciprocal Dyads** (Asymmetric): One node sends a tie to another, but it is not reciprocated (size one).
  3. **Reciprocal Dyads** (Mutual): Both nodes send directed ties to each other (size two).
- In standard SNA, this three-fold classification is called the **MAN classification**: **M**utual, **A**symmetric, **N**ull.
:::

---

## Asymmetric Dyads: Visual Examples

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- **Null**: 0 edges.
- **Asymmetric**: 1 directed edge.
- **Mutual**: 2 reciprocal directed edges.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr1 <- create_empty(2, directed = TRUE) %>% mutate(name = c("A", "B"))
gr2 <- create_empty(2, directed = TRUE) %>% bind_edges(data.frame(from = 1, to = 2)) %>% mutate(name = c("C", "D"))
gr3 <- create_empty(2, directed = TRUE) %>% bind_edges(data.frame(from = c(1, 2), to = c(2, 1))) %>% mutate(name = c("E", "F"))

p1 <- ggraph(gr1, layout = 'linear') +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Null (Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

p2 <- ggraph(gr2, layout = 'linear') +
  geom_edge_link(color = "steelblue", edge_width = 1,
                  arrow = arrow(length = unit(3, 'mm'), type = "closed"), end_cap = circle(5, 'mm')) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Asymmetric (Non-reciprocal)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

p3 <- ggraph(gr3, layout = 'linear') +
  geom_edge_parallel(color = "steelblue", edge_width = 1,
                  arrow = arrow(length = unit(3, 'mm'), type = "closed"), end_cap = circle(5, 'mm'), sep = unit(4, 'mm')) +
  geom_node_point(size = 12, color = "tan2") +
  geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Mutual (Reciprocal)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

p1 / p2 / p3
```
:::

::::

---

## Separating Tie Properties from Dyad Realizations

::: {.incremental}
- It is critical to keep the property of the **tie** separate from the property of the **dyad**:
  - **Asymmetric Tie**: Describes the type of *relation* (directionality is allowed; reciprocity is optional).
  - **Non-Reciprocal Dyad**: Describes a realized *configuration* of a pair of nodes where reciprocity failed to occur empirically.
  - **Reciprocal Dyad**: Describes an empirical configuration composed of a *pair* of directed edges under an asymmetric relation.
- Confusing "asymmetric tie" with "asymmetric dyad" is a common conceptual error!
:::

---

## Dyads under Anti-Symmetric Ties

::: {.incremental}
- Because reciprocity is strictly forbidden under anti-symmetric ties, mutual (reciprocal) dyads cannot occur.
- Therefore, anti-symmetric ties can only yield **two** types of dyads:
  1. **Disconnected Dyads**: The two nodes do not share any directed edge.
  2. **Non-Reciprocal Dyads**: Only a single one-way directed tie exists between them ($A \rightarrow B$).
- This simplifies the dyad census of hierarchical networks: every pair of nodes is either completely disconnected or connected by a strictly non-reciprocal arrow.
:::

---

## Anti-Symmetric Dyads: Visual Examples

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- Mutual dyads are strictly impossible.
- Every pair of nodes is either disconnected or connected by a strictly non-reciprocal arrow.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr1 <- create_empty(2, directed = TRUE) %>% mutate(name = c("A", "B"))
gr2 <- create_empty(2, directed = TRUE) %>% bind_edges(data.frame(from = 1, to = 2)) %>% mutate(name = c("C", "D"))

p1 <- ggraph(gr1, layout = 'linear') +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Disconnected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

p2 <- ggraph(gr2, layout = 'linear') +
  geom_edge_link(color = "steelblue", edge_width = 1.5,
                  arrow = arrow(length = unit(4, 'mm'), type = "closed"), end_cap = circle(6, 'mm')) +
  geom_node_point(size = 15, color = "tan2") +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "white") +
  theme_graph() + ggtitle("Non-Reciprocal") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

p1 / p2
```
:::

::::

# Part 3: Triads and Motifs

---

## What is a Triad?

::: {.incremental}
- A **triad** is any set of three nodes, whether joined by edges or not.
- Mathematically, a triad is *any subgraph of order three* ($n = 3$).
- Triads have incredibly powerful sociological properties (e.g., structural balance, brokerage, coalitions).
- A triad can be thought of as a **concatenation of three dyads**: $\{A, B\}$, $\{B, C\}$, and $\{A, C\}$.
- Just like dyads, the allowed configurations of a triad depend entirely on the relational properties of its ties.
:::

---

## Triads under Symmetric Ties

::: {.incremental}
- Since undirected dyads can only be connected or disconnected, there are only **four** possible undirected triad configurations:
  1. **Null Triad**: 0 connected dyads ($DDD$).
  2. **Disconnected Triad**: 1 connected dyad ($CDD$).
  3. **Open Triad**: 2 connected dyads ($CCD$).
  4. **Closed Triad**: 3 connected dyads ($CCC$).
- These constitute the classic undirected subgraphs shown on the right.
:::

---

## Symmetric Triads: Visual Examples

:::: {.columns align=center}

::: {.column width="30%"}
::: {.incremental}
- Subgraphs of order three are built from combinations of undirected connected and disconnected dyads.
- Zero, one, two, or three edges can be present, creating distinct micro-structures.
:::
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5.5
gr <- create_empty(3, directed = FALSE) %>% mutate(name = toupper(letters[1:3]))
p3_ex <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
    geom_node_point(size = 15, color = "tan2") +
    geom_node_text(aes(label = name), size = 6, color = "white") +
    theme_graph() + coord_cartesian(clip = "off")
p3_ex
```
:::

::::

---

## The Four Undirected Triads

```{r}
#| label: fig-undir-triads
#| fig-cap: "The four types of undirected triads."
#| fig-cap-location: bottom
#| fig-height: 5.5
#| fig-width: 11

    gr <- create_empty(3, directed = FALSE) %>% mutate(name = toupper(letters[1:3]))
    p1 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Null") + coord_cartesian(clip = "off")
    
    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        mutate(name = toupper(letters[1:3]))
    p2 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_edge_link(color = "steelblue", width = 1.25) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Disconnected") + coord_cartesian(clip = "off")

    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        bind_edges(data.frame(from = 2, to = 3)) %>%
        mutate(name = toupper(letters[1:3]))
    p3 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_edge_link(color = "steelblue", width = 1.25) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Open") + coord_cartesian(clip = "off")
    
    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        bind_edges(data.frame(from = 2, to = 3)) %>%
        bind_edges(data.frame(from = 3, to = 1)) %>%
        mutate(name = toupper(letters[1:3]))
    p4 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_edge_link(color = "steelblue", width = 1.25) +
        geom_node_point(size = 20, color = "tan2") +
        geom_node_text(aes(label = name), size = 8, color = "white") +
        theme_graph() + ggtitle("Closed") + coord_cartesian(clip = "off")
    
    p1 + p2 + p3 + p4 + plot_layout(nrow = 1)
```

---

## Triads under Anti-Symmetric Ties

::: {.incremental}
- Because reciprocity is strictly forbidden, any triad containing a mutual connection is impossible.
- This prunes the directed triad census from **16 down to exactly seven** allowed configurations:
  1. **Null Triad** ($NNN$, panel 1): All three dyads are completely disconnected.
  2. **One-tie Triad** ($NNA$, panel 2): Contains a single non-reciprocal directed tie.
  3. **Directed Line** ($AAN(L)$, panel 3): Non-reciprocal ties form a directed path.
  4. **Out-Star Triad** ($AAN(O)$, panel 4): A single node sends ties to both other nodes.
  5. **In-Star Triad** ($AAN(I)$, panel 5): Both other nodes send ties to a single node.
  6. **Cycle Triad** ($AAA(C)$, panel 6): A directed cycle of length three ($A \rightarrow B \rightarrow C \rightarrow A$).
  7. **Transitive Hierarchy** ($AAA(I/O)$, panel 7): The classic transitive triad ($A \rightarrow B \rightarrow C$ and $A \rightarrow C$).
:::

---

## The Seven Anti-Symmetric Triads

```{r}
#| label: fig-antisym-triads-slides
#| fig-cap: "The Seven Permitted Triads Under Anti-Symmetric Ties"
#| fig-cap-location: bottom
#| fig-height: 5.5
#| fig-width: 11

    # We reuse plot_tr helper defined in setup (perfect triangle coordinates)
    p1 <- plot_tr(data.frame(), "1. NNN (Null)")
    p2 <- plot_tr(data.frame(from=1, to=2), "2. NNA (One-tie)")
    p3 <- plot_tr(data.frame(from=c(1,2), to=c(2,3)), "3. AAN(L) (Line)")
    p4 <- plot_tr(data.frame(from=c(1,1), to=c(2,3)), "4. AAN(O) (Out-Star)")
    p5 <- plot_tr(data.frame(from=c(2,3), to=c(1,1)), "5. AAN(I) (In-Star)")
    p6 <- plot_tr(data.frame(from=c(1,2,3), to=c(2,3,1)), "6. AAA(C) (Cycle)")
    p7 <- plot_tr(data.frame(from=c(1,1,2), to=c(2,3,3)), "7. AAA(I/O) (Transitive)")
    
    (p1 + p2 + p3 + p4) / (p5 + p6 + p7 + plot_spacer())
```

---

## Anti-Symmetric Triads and Command Chains

::: {.incremental}
- This mathematical pruning simplifies the analysis of hierarchies.
- In formal organization command structures:
  - We expect to see **transitive triads** ($AAA(I/O)$), representing clean, unified chains of command (e.g., "the boss of my boss is my boss").
  - **Cycles** ($AAA(C)$) are structurally avoided because they generate circular authority paradoxes, where everyone takes orders from someone else in a loop.
:::

---

## Hierarchies: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- On the right is a **formal organization hierarchy** built strictly from anti-symmetric ties (supervisor of):
  - Every connection flows downward.
  - The graph is completely acyclic (no loops).
  - Every node (except the President root) has an indegree of exactly 1.
- Removing the transitive and cyclic "noise" results in a clean **directed tree**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_hierarchy <- tbl_graph(
  nodes = data.frame(name = c("Pres", "VP 1", "VP 2", "Staff 1", "Staff 2")),
  edges = data.frame(from = c(1, 1, 2, 3), to = c(2, 3, 4, 5)),
  directed = TRUE
)
ggraph(gr_hierarchy, layout = "tree") +
  geom_edge_link(color = "steelblue", edge_width = 1.5,
                 arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                 end_cap = circle(8, 'mm'), start_cap = circle(8, 'mm')) +
  geom_node_point(size = 20, color = "tan2") +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "white") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Triads under Asymmetric Ties

::: {.incremental}
- When relations are asymmetric, reciprocity is allowed but optional.
- Because each dyadic slot can take on any of the **three** MAN states, we can generate all **sixteen** directed triad configurations.
- On the next slide, we show the **nine directed configurations** that require at least one **reciprocal (mutual) dyad** to exist:
  - **Single Mutual**: $NNM$ (panel 1), $MAN1$ (panel 2), $MAN2$ (panel 3).
  - **Double Mutual**: $MMN$ (panel 4).
  - **Mutual-Mixed Closed**: $MAA(C)$ (panel 5), $MAA(O)$ (panel 6), $MAA(I)$ (panel 7).
  - **Cohesive Closed**: $MMA$ (panel 8), $MMM$ (panel 9).
:::

---

## The Nine Reciprocal Triads

```{r}
#| label: fig-recip-triads-slides
#| fig-cap: "The Nine Directed Triads Requiring Reciprocal (Mutual) Ties"
#| fig-cap-location: bottom
#| fig-height: 5.5
#| fig-width: 9.5

    p1 <- plot_tr(data.frame(from=c(1,2), to=c(2,1)), "1. NNM")
    p2 <- plot_tr(data.frame(from=c(2,1,3), to=c(1,2,1)), "2. MAN1")
    p3 <- plot_tr(data.frame(from=c(2,1,1), to=c(1,2,3)), "3. MAN2")
    p4 <- plot_tr(data.frame(from=c(2,1,3,2), to=c(1,2,2,3)), "4. MMN")
    p5 <- plot_tr(data.frame(from=c(1,2,2,3), to=c(2,1,3,1)), "5. MAA(C)")
    p6 <- plot_tr(data.frame(from=c(1,2,3,3), to=c(2,1,1,2)), "6. MAA(O)")
    p7 <- plot_tr(data.frame(from=c(1,2,1,2), to=c(2,1,3,3)), "7. MAA(I)")
    p8 <- plot_tr(data.frame(from=c(2,1,3,2,1), to=c(1,2,2,3,3)), "8. MMA")
    p9 <- plot_tr(data.frame(from=c(2,1,3,2,1,3), to=c(1,2,2,3,3,1)), "9. MMM")
    
    (p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
```

---

## Triad Enumeration Summary

* **0 connected dyads**:
  - **NNN** (Strictly non-reciprocal, panel 1)
* **1 connected dyad**:
  - **NNA** (Non-reciprocal, panel 2), **NNM** (Reciprocal, panel 3)
* **2 connected dyads (Open)**:
  - Non-reciprocal: **AAN(L)**, **AAN(O)**, **AAN(I)** (panels 4-6)
  - Mixed: **MAN1**, **MAN2** (panels 7-8)
  - Reciprocal: **MMN** (panel 9)
* **3 connected dyads (Closed)**:
  - Non-reciprocal: **AAA(C)**, **AAA(I/O)** (panels 10-11)
  - Mixed: **MAA(C)**, **MAA(O)**, **MAA(I)** (panels 12-14)
  - Reciprocal: **MMA** (panel 15), **MMM** (panel 16)

---

## Conclusion & Core Takeaways

::: {.incremental}
- **Ties define dyads, and dyads define triads**:
  - The mathematical type of relationship strictly controls the possible micro-structures in a network.
- **Symmetric Ties** $\implies$ connected vs. disconnected dyads $\implies$ 4 undirected triads.
- **Anti-Symmetric Ties** $\implies$ disconnected vs. non-reciprocal dyads $\implies$ 7 directed triads.
- **Asymmetric Ties** $\implies$ disconnected, non-reciprocal, or reciprocal dyads $\implies$ 16 directed triads (Triad Census).
- Understanding these micro-motifs is the key to decoding the macro-structures of power, friendship, and organization in society.
:::

---

## References {.unnumbered}
