---
title: "Ties, Dyads, and Triads"
subtitle: "The Motifs of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Perfect triangle coordinates for 3-node triad drawings
x_triad <- c(0, 2, 1)
y_triad <- c(0, 0, 1.2)

# Canonical triad orientations: the pivot node (A) sits on top for the
# out-star (021D) and in-star (021U); 030T uses the classic vertical chain
xy_top   <- list(x = c(1, 0, 2),   y = c(1.2, 0, 0))
xy_chain <- list(x = c(1, 0.3, 1), y = c(1.2, 0.55, -0.05))

# Coordinates for 2-node (dyad) drawings
x <- c(1, 2)
y <- c(1, 1)

# Helper function to plot individual directed triads
plot_tr <- function(edges, title, node_size = 18, text_size = 7, title_color = "#2c3e50",
                    x = x_triad, y = y_triad) {
  g <- create_empty(3, directed = TRUE) %>% mutate(name = toupper(letters[1:3]))
  if (nrow(edges) > 0) {
    g <- g %>% bind_edges(edges)
  }
  cap_size <- node_size / 2
  arrow_len <- node_size / 4
  p <- ggraph(g, layout = "manual", x = x, y = y) +
    geom_edge_link(color = "#2c3e50", edge_width = 1.1,
                   arrow = arrow(length = unit(arrow_len, 'mm'), type = "closed"),
                   end_cap = circle(cap_size, 'mm'), start_cap = circle(cap_size, 'mm')) +
    geom_node_point(size = node_size, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.25) +
    geom_node_text(aes(label = name), size = text_size, fontface = "bold", color = "#2c3e50") +
    theme_graph() + ggtitle(title) + coord_cartesian(xlim = c(-0.5, 2.5), ylim = c(-0.2, 1.4), clip = "off") +
    theme(
      plot.title = element_text(size = 10, face="bold", hjust=0.5, color = title_color),
      plot.margin = margin(5, 5, 5, 5)
    )
  return(p)
}

# Function to create a clean, minimalist 2-node graph plot from the book chapter
plot_tie_type <- function(nodes, edges, title, subtitle, directed = TRUE, has_cross = FALSE,
                           node_size = 25, text_size = 10) {
  g <- tbl_graph(nodes = nodes, edges = edges)
  p <- ggraph(g, layout = "manual", x = x, y = y)
  cap_size <- node_size / 2.1
  arrow_len <- node_size / 5
  
  if (directed) {
    p <- p + geom_edge_arc(aes(strength = strength, color = color, linetype = linetype),
                           arrow = arrow(length = unit(arrow_len, 'mm'), type = "closed"),
                           end_cap = circle(cap_size, 'mm'),
                           start_cap = circle(cap_size, 'mm'),
                           width = 1.1,
                           show.legend = FALSE) +
             scale_edge_color_identity() +
             scale_edge_linetype_identity()
  } else {
    p <- p + geom_edge_link(color = "#2c3e50",
                            width = 1.1,
                            end_cap = circle(cap_size, 'mm'),
                            start_cap = circle(cap_size, 'mm'),
                            show.legend = FALSE)
  }
  
  pBase <- p + geom_node_point(size = node_size, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
    geom_node_text(aes(label = name), size = text_size, fontface = "bold", color = "#2c3e50")
  
  if (has_cross) {
    pBase <- pBase + annotate("text", x = 1.5, y = 0.90, label = "✕", color = "#e74c3c", size = 8, fontface = "bold")
  }
  
  pBase <- pBase + 
    labs(title = title, subtitle = subtitle) +
    theme_graph() +
    theme(
      plot.title = element_text(size = 12, face = "bold", color = "#2c3e50", hjust = 0.5),
      plot.subtitle = element_text(size = 9.5, color = "#7f8c8d", hjust = 0.5, lineheight = 1.1),
      plot.margin = margin(10, 10, 10, 10)
    ) +
    coord_fixed(ratio = 1, xlim = c(0.6, 2.4), ylim = c(0.65, 1.35), clip = "off")
  
  return(pBase)
}

# Helper: a single 2-node dyad panel
plot_dyad <- function(edges = NULL, title = NULL, directed = TRUE,
                      node_size = 18, text_size = 7, title_size = 10) {
  g <- create_empty(2, directed = directed) %>% mutate(name = toupper(letters[1:2]))
  if (!is.null(edges) && nrow(edges) > 0) {
    g <- g %>% bind_edges(edges)
  }
  cap_size <- node_size / 2.5
  p <- ggraph(g, layout = 'linear') +
    geom_node_point(size = node_size, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.25) +
    geom_node_text(aes(label = name), size = text_size, fontface = "bold", color = "#2c3e50") +
    theme_graph() + coord_cartesian(xlim = c(0.25, 2.75), clip = "off") +
    theme(plot.margin = margin(5, 5, 5, 5))
  if (!is.null(edges) && nrow(edges) > 0) {
    if (directed && nrow(edges) == 2) {
      p <- p + geom_edge_parallel(color = "#2c3e50", edge_width = 1,
                                  arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                                  start_cap = circle(cap_size, 'mm'),
                                  end_cap = circle(cap_size, 'mm'),
                                  sep = unit(5, 'mm'))
    } else {
      p <- p + geom_edge_link(color = "#2c3e50", edge_width = 1.1,
                              arrow = if (directed) arrow(length = unit(4, 'mm'), type = "closed") else NULL,
                              start_cap = circle(cap_size, 'mm'),
                              end_cap = circle(cap_size, 'mm'))
    }
  }
  if (!is.null(title)) {
    p <- p + ggtitle(title) +
      theme(plot.title = element_text(size = title_size, face = "bold", hjust = 0.5, color = "#2c3e50"))
  }
  p
}

# Helper: a single undirected triad panel
plot_triad_u <- function(edges = NULL, title = NULL, node_size = 20, text_size = 8, title_size = 11) {
  g <- create_empty(3, directed = FALSE) %>% mutate(name = toupper(letters[1:3]))
  if (!is.null(edges) && nrow(edges) > 0) {
    g <- g %>% bind_edges(edges)
  }
  p <- ggraph(g, layout = "manual", x = x_triad, y = y_triad) +
    geom_edge_link(color = "#2c3e50", edge_width = 1.1, start_cap = circle(8, 'mm'), end_cap = circle(8, 'mm')) +
    geom_node_point(size = node_size, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
    geom_node_text(aes(label = name), size = text_size, fontface = "bold", color = "#2c3e50") +
    theme_graph() +
    coord_cartesian(xlim = c(-0.5, 2.5), ylim = c(-0.2, 1.4), clip = "off") +
    theme(plot.margin = margin(5, 5, 5, 5))
  if (!is.null(title)) {
    p <- p + ggtitle(title) +
      theme(plot.title = element_text(size = title_size, face = "bold", hjust = 0.5, color = "#2c3e50"))
  }
  p
}

# Marriage cascade: two kinship families merging via marriage
fam_nodes <- data.frame(
  name    = c("Mom", "Dad", "You", "Bro", "Mom", "Dad", "Riley", "Sis"),
  x       = c(-0.5, 0.3, 0.6, -0.9, 2.7, 3.5, 2.4, 3.9),
  y       = c(2.3, 2.3, 0.7, 1.2, 2.3, 2.3, 0.7, 1.2),
  label_y = c(2.68, 2.68, 0.42, 0.92, 2.68, 2.68, 0.42, 0.92)
)

plot_cascade <- function(stage, title) {
  extra <- switch(
    stage,
    data.frame(from = integer(), to = integer(), style = character()),
    data.frame(from = 3, to = 7, style = "romance"),
    data.frame(from = c(3, 1, 2, 5, 6, 4, 3),
               to   = c(7, 7, 7, 3, 3, 7, 8),
               style = c("marriage", "inlaw_dir", "inlaw_dir", "inlaw_dir",
                         "inlaw_dir", "inlaw_sym", "inlaw_sym"))
  )
  edges <- bind_rows(
    data.frame(from = c(1, 1, 2, 2, 5, 5, 6, 6),
               to   = c(3, 4, 3, 4, 7, 8, 7, 8), style = "kin"),
    extra
  )
  g <- tbl_graph(nodes = fam_nodes, edges = edges, directed = TRUE)
  ggraph(g, layout = "manual", x = fam_nodes$x, y = fam_nodes$y) +
    geom_edge_link(aes(filter = style == "kin"),
                   color = "#2c3e50", edge_width = 1) +
    geom_edge_link(aes(filter = style == "inlaw_sym"),
                   color = "#c0392b", edge_width = 1, linetype = "dashed") +
    geom_edge_link(aes(filter = style == "inlaw_dir"),
                   color = "#c0392b", edge_width = 1, linetype = "dashed",
                   arrow = arrow(length = unit(2.5, 'mm'), type = "closed"),
                   end_cap = circle(5, 'mm'), start_cap = circle(5, 'mm')) +
    geom_edge_link(aes(filter = style == "romance"),
                   color = "#8e44ad", edge_width = 1, linetype = "dashed",
                   arrow = arrow(length = unit(2.5, 'mm'), type = "closed"),
                   end_cap = circle(5, 'mm'), start_cap = circle(5, 'mm')) +
    geom_edge_link(aes(filter = style == "marriage"),
                   color = "#c0392b", edge_width = 1.4) +
    geom_node_point(size = 10, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.25) +
    geom_node_text(aes(label = name, y = label_y), size = 4.5,
                   fontface = "bold", color = "#2c3e50") +
    theme_graph() + ggtitle(title) +
    coord_cartesian(xlim = c(-1.4, 4.4), ylim = c(0.2, 2.9), clip = "off") +
    theme(
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5, color = "#2c3e50"),
      plot.margin = margin(5, 5, 5, 5)
    )
}

# Define Nodes
nodes_2 <- data.frame(name = c("A", "B"), x = c(1, 2), y = c(1, 1))

# Define Edges for symmetric, reciprocal, non-reciprocal, and anti-symmetric
edges_sym <- data.frame(from = 1, to = 2)
edges_rec <- data.frame(from = c(1, 2), to = c(2, 1), strength = c(0.1, 0.1), color = c("#2c3e50", "#2c3e50"), linetype = c("solid", "solid"))
edges_nonrec <- data.frame(from = 1, to = 2, strength = 0, color = "#2c3e50", linetype = "solid")
edges_anti <- data.frame(
  from = c(1, 2), 
  to = c(2, 1), 
  strength = c(0.1, 0.1), 
  color = c("#2c3e50", "#e74c3c"), 
  linetype = c("solid", "dashed")
)
```

## Today's Roadmap

::: {.incremental}
- Today's guiding principle: **the type of tie strictly controls which micro-structures are possible**.
- **Part 1 — The Taxonomy of Ties**: how sociologists classify ties (similarities, social relations, interactions, flows), and how graph theory classifies them by directionality (symmetric, asymmetric, anti-symmetric).
- **Part 2 — Dyads and Their Realizations**: the possible configurations of a pair of nodes under each tie type.
- **Part 3 — Triads and Motifs**: scaling from pairs to triples — the triad census that underlies balance, brokerage, and hierarchy.
- Keep one question in mind throughout: *what does this tie type allow two (or three) people to do?*
:::

---

# Part 1: The Taxonomy of Ties

---

## The Relational Basis of Networks

::: {.incremental}
- In social network analysis, a **graph** represents concrete **relationships** (ties) between actors.
- Not all relationships are created equal!
  - Some ties are naturally mutual (e.g., being siblings).
  - Others are directional (e.g., seeking advice).
  - Some strictly forbid mutuality (e.g., being a supervisor).
- To study networks, we must separate the **mathematical property of the relationship** (the tie type) from its **empirical realization** (the dyad or triad).
:::

---

## The Sociological Classification of Types of Ties

::: {.incremental}
- Instead of treating all connections identically, sociologists classify social ties into **four major types** ([Borgatti et al. 2009](https://doi.org/10.1126/science.1165821)):
  1. **Similarities**: Having something in common (e.g., co-location, shared attributes).
  2. **Social Relations**: Culturally recognized labels or sentiments (e.g., friend, spouse, boss).
  3. **Interactions**: Regular communication or behavior (e.g., texting, talking).
  4. **Flows**: Transmission of resources or states (e.g., virus infection, lending money).
:::

---

## States vs. Events in Social Networks

::: {.incremental}
- We can divide these four types of ties into two fundamental clusters:
- **States (Static Situations)**:
  - **Similarities** & **Social Relations**.
  - Represent continuous, long-standing conditions.
  - *Examples*: Being in the same class, being siblings, being coworker/boss.
- **Events (Discrete Activities)**:
  - **Interactions** & **Flows**.
  - Occur at specific points in time and are transient.
  - *Examples*: Texting someone, lending money, transmitting a virus.
:::

---

## Graph-Theoretic Classification of Ties

::: {.incremental}
- While sociologists classify ties by their **substantive content** (similarities, relations, interactions, flows)...
- Mathematicians and graph theorists classify ties by their **formal, directionality properties**:
  - **Symmetric Ties**: Inherently mutual and undirected ($AB \iff BA$).
  - **Asymmetric Ties**: Directional, where reciprocity is optional ($AB \implies$ optional $BA$).
  - **Anti-Symmetric Ties**: Directional, where reciprocity is forbidden ($AB \implies \neg BA$).
- **The bridge**: To represent a network, we must map our sociological tie types (e.g. boss-of, friend-of) onto these three graph-theoretic classes!
:::

---

## 1. Symmetric Ties & Similarities

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: A relation where connection is built-in and mutual by construction:
  - Formally: $AB \iff BA$
- **Representation**: Represented by **undirected graphs**.
- **Structural Constraint**: Pairs of actors can only be **disconnected** or **connected** by a single undirected edge.
- **Symmetric Similarities**:
  - *Similarities are necessarily symmetric*:
    - If you are in the same place at the same time as someone, they are there with you.
    - If you are zooming with someone, they are zooming with you.
    - If you go to UCLA with someone, they also go to UCLA with you.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
plot_tie_type(nodes_2, edges_sym, "Symmetric Tie", "e.g., 'is sibling of'", directed = FALSE)
```
:::

::::

---

## 2. Asymmetric Ties & Interactions

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: A relation with natural directionality where reciprocity is possible but *not guaranteed*:
  - Formally: $AB \implies$ nothing about $BA$
- **Representation**: Represented by **directed graphs** (digraphs).
- **Structural Constraint**: Reciprocity is an **empirical event** (it may or may not occur), yielding three possible dyad configurations.
- **Asymmetric Relations**:
  - *Note*: Asymmetric relations can be either **states** (sentiments like liking) or **events** (interactions like texting) — both are naturally asymmetric:
    - You can follow someone on Twitter, but they don't have to follow you back.
    - You can send a text to someone, but they don't have to respond.
    - You can seek advice from someone without them seeking advice from you.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 7
p1 <- plot_tie_type(nodes_2, edges_rec, "Asymmetric: Reciprocal Dyad", "e.g., 'A likes B' & 'B likes A'", directed = TRUE, node_size = 24, text_size = 9)
p2 <- plot_tie_type(nodes_2, edges_nonrec, "Asymmetric: Non-Reciprocal Dyad", "e.g., 'A sends email to B'", directed = TRUE, node_size = 24, text_size = 9)
p1 / p2
```
:::

::::

---

## 3. Anti-Symmetric Ties & Hierarchies

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: An inherently directional relation where reciprocity is **strictly forbidden**:
  - Formally: $AB \implies \neg BA$
- **Representation**: Represented by **directed graphs** (oriented graphs, DAGs, or trees).
- **Structural Constraint**: Reciprocity is structurally impossible, yielding only two possible dyad configurations.
- **Anti-Symmetric Hierarchies**:
  - *Hierarchical role relations are anti-symmetric*:
    - Kinship: If you are someone's parent, they cannot be your parent.
    - Formal: If you are someone's boss, they cannot be your boss at the same time.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
plot_tie_type(nodes_2, edges_anti, "Anti-Symmetric Tie", "e.g., 'A is boss of B'", directed = TRUE, has_cross = TRUE)
```
:::

::::

---

## Mapping Sociological to Graph-Theoretic Classifications

| Sociological Classification (Types of Ties) | Graph-Theoretic Classification (Mathematical Properties) | Allowed Dyads | Visual Graph Representation |
|:---|:---|:---|:---|
| **Co-membership, Propinquity** | **Symmetric** | • Disconnected<br>• Connected | Undirected Graph ($A - B$) |
| **Sentiments, Advice-seeking** | **Asymmetric** | • Disconnected<br>• Non-reciprocal<br>• Reciprocal | Directed Graph ($A \rightarrow B$ or $A \leftrightarrow B$) |
| **Formal Hierarchies, Kinship** | **Anti-symmetric** | • Disconnected<br>• Non-reciprocal | Oriented Graph/Tree ($A \rightarrow B$, no cycles) |

---

## Multiplexity in Social Relations

::: {.incremental}
- In real-world social networks, actors are rarely linked by only one relationship.
- **Multiplexity** refers to the presence of multiple, distinct types of ties linking the same pair of nodes simultaneously.
  - e.g., $A$ and $B$ are coworkers (**symmetric** tie) AND $A$ is the supervisor of $B$ (**anti-symmetric** tie) AND they are friends (**asymmetric** tie).
- Multiplexity adds rich, multi-layered complexity to networks, as the behavior or pattern in one relation heavily influences and moderates others.
:::

---

## Multiplexity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- **Four nodes** (diamond layout) connected by **four distinct relationship layers**:
  - **Symmetric Coworker**: links *every* pair of nodes — a fully connected baseline.
  - **Anti-Symmetric Supervisor-of**
  - **Asymmetric Advice-Seeking**
  - **Asymmetric/Mutual Friendship**
- *The friendship, advice, and supervisor layers are multiplex on top of this fully connected coworker baseline (e.g., supervisors who are also friends).*
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6.5
#| fig-height: 5.5
nodes_mult <- data.frame(
  name = c("A", "B", "C", "D"),
  x = c(1, 0, 2, 1),
  y = c(2, 1, 1, 0)
)
edge_types <- c(
  "Supervisor-of", "Friendship", "Friendship",
  "Supervisor-of", "Advice",
  "Coworker", "Advice",
  "Friendship", "Friendship",
  "Supervisor-of", "Coworker",
  "Coworker", "Advice",
  "Coworker", "Coworker", "Coworker"
)
edges_mult <- data.frame(
  from = c(1, 1, 2,  1,  3,  2, 2,  2, 4,  4,  3,  1, 4,  2,  1, 4),
  to   = c(2, 2, 1,  3,  1,  3, 3,  4, 2,  3,  4,  4, 1,  1,  3, 2),
  type = edge_types,
  # Coworker is symmetric: drawn as an undirected arc, no arrows
  arrowed = edge_types != "Coworker"
)
gr_mult <- tbl_graph(nodes = nodes_mult, edges = edges_mult, directed = TRUE)
ggraph(gr_mult, layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(filter = arrowed, color = type),
                     arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                     start_cap = circle(9, 'mm'), end_cap = circle(9, 'mm'),
                     sep = unit(3.0, 'mm'), edge_width = 1.0) +
  # Coworker ties run between node pairs that also carry arrowed ties in the
  # same direction; curving them into arcs (and drawing them last) keeps them
  # from being occluded by the arrowed edges
  geom_edge_arc(aes(filter = !arrowed, color = type), strength = 0.25,
                edge_width = 1.2) +
  scale_edge_color_manual(values = c(
    "Coworker" = "#7f8c8d",
    "Supervisor-of" = "#c0392b",
    "Friendship" = "#27ae60",
    "Advice" = "#8e44ad"
  )) +
  geom_node_point(size = 20, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "#2c3e50") +
  theme_graph() +
  coord_cartesian(xlim = c(-0.5, 2.5), ylim = c(-0.5, 2.5), clip = "off") +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 10))
```
:::

::::

---

## The Growth of Social Networks: The Marriage Cascade

::: {.incremental}
- Social networks are highly dynamic, growing and evolving over time.
- A classic example of network growth is **the marriage cascade**:
  1. A **romantic tie** (asymmetric) forms between two people from separate families: `You` $\rightarrow$ `Riley`.
  2. It solidifies into **marriage** (a symmetric kin tie): `You` $\leftrightarrow$ `Riley`.
  3. The marriage triggers a cascade of new **in-law ties**:
     - **Symmetric sibling-in-law** ties: `You` $\leftrightarrow$ `Sis` and `Bro` $\leftrightarrow$ `Riley`.
     - **Anti-symmetric parent-in-law** ties: your parents $\rightarrow$ `Riley`, and Riley's parents $\rightarrow$ `You`.
  4. Two previously separate kinship trees merge into a single **multiplex** network.
:::

---

## The Marriage Cascade: A Visual Example

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 11
#| fig-height: 3.8

plot_cascade(1, "1. Two Separate Families") +
  plot_cascade(2, "2. A Romantic Tie Forms") +
  plot_cascade(3, "3. Marriage + In-Law Cascade") +
  plot_layout(ncol = 3)
```

# Part 2: Dyads and Their Realizations

---

## What is a Dyad?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **dyad** is any pair of nodes in a graph, whether they are joined by an edge or not.
- Mathematically, a dyad is *any subgraph of order two* ($n = 2$).
- Crucially, a dyad represents the **realized configuration** of ties between two nodes.
- **Sociological Insight**: Ties that do *not* exist (disconnected dyads) are often just as socially meaningful as those that do!
- **Dyads: The Building Blocks**:
  - The simplest possible subgraph of order two.
  - Can contain zero, one, or two directed edges depending on relational constraints.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
plot_dyad(directed = FALSE, node_size = 25, text_size = 10)
```
:::

::::

---

## Dyads under Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Because undirected graphs do not allow directionality, a symmetric relation can only take on **two** dyadic states:
  1. **Disconnected Dyads** (historically called **null dyads**): The two nodes do not share an edge (size zero).
  2. **Connected Dyads**: The two nodes share an edge (size one).
- Every pair of actors in an undirected graph is either connected or disconnected.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5.5
plot_dyad(NULL, "Disconnected (Size Zero)", directed = FALSE,
          node_size = 20, text_size = 8, title_size = 12) /
  plot_dyad(data.frame(from = 1, to = 2), "Connected (Size One)", directed = FALSE,
            node_size = 20, text_size = 8, title_size = 12)
```
:::

::::

---

## Dyads under Asymmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, reciprocity is possible but optional.
- Thus, asymmetric ties can yield **three** distinct kinds of dyads (the classic **MAN classification**):
  1. **Disconnected Dyads** (Null): Neither node sends a directed tie to the other (size zero).
  2. **Non-Reciprocal Dyads** (Asymmetric): One node sends a tie to another, but it is not reciprocated (size one).
  3. **Reciprocal Dyads** (Mutual): Both nodes send directed ties to each other (size two).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5.5
plot_dyad(NULL, "Null (Disconnected)") /
  plot_dyad(data.frame(from = 1, to = 2), "Asymmetric (Non-reciprocal)") /
  plot_dyad(data.frame(from = c(1, 2), to = c(2, 1)), "Mutual (Reciprocal)")
```
:::

::::

---

## Dyads under Anti-Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Because reciprocity is strictly forbidden under anti-symmetric ties, mutual (reciprocal) dyads cannot occur.
- Therefore, anti-symmetric ties can only yield **two** types of dyads:
  1. **Disconnected Dyads**: The two nodes do not share any directed edge.
  2. **Non-Reciprocal Dyads**: Only a single one-way directed tie exists between them ($A \rightarrow B$).
- This simplifies the dyad census of hierarchical networks: every pair of nodes is either completely disconnected or connected by a strictly non-reciprocal arrow.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5.5
plot_dyad(NULL, "Disconnected", node_size = 20, text_size = 8, title_size = 12) /
  plot_dyad(data.frame(from = 1, to = 2), "Non-Reciprocal", node_size = 20, text_size = 8, title_size = 12)
```
:::

::::

---

## Separating Tie Properties from Dyad Realizations

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- It is critical to keep the property of the **tie** separate from the property of the **dyad**:
  - **Asymmetric Tie**: Describes the type of *relation* (directionality is allowed; reciprocity is optional).
  - **Non-Reciprocal Dyad**: Describes a realized *configuration* of a pair of nodes where reciprocity failed to occur empirically.
  - **Reciprocal Dyad**: Describes an empirical configuration composed of a *pair* of directed edges under an asymmetric relation.
- Confusing "asymmetric tie" with "asymmetric dyad" is a common conceptual error!
- With the dyad census complete for all three tie types, we are ready to scale up from pairs to **triads**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5.5

# One tie property (header), two realized dyad configurations (panels)
plot_dyad(data.frame(from = 1, to = 2), "Non-Reciprocal Dyad",
          node_size = 20, text_size = 8, title_size = 11) /
  plot_dyad(data.frame(from = c(1, 2), to = c(2, 1)), "Reciprocal Dyad",
            node_size = 20, text_size = 8, title_size = 11) +
  plot_annotation(
    title = "Tie Property: Asymmetric (reciprocity optional)",
    theme = theme(plot.title = element_text(size = 13, face = "bold",
                                            hjust = 0.5, color = "#2c3e50"))
  )
```
:::

::::

---

# Part 3: Triads and Motifs

---

## What is a Triad?

::: {.incremental}
- A **triad** is any set of three nodes, whether joined by edges or not.
- Mathematically, a triad is *any subgraph of order three* ($n = 3$).
- Triads have incredibly powerful sociological properties (e.g., structural balance, brokerage, coalitions).
- A triad can be thought of as a **concatenation of three dyads**: $\{A, B\}$, $\{B, C\}$, and $\{A, C\}$.
- Just like dyads, the allowed configurations of a triad depend entirely on the relational properties of its ties.
:::

---

## Triads under Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Since undirected dyads can only be connected or disconnected, there are only **four** possible undirected triad configurations:
  1. **Null Triad**: 0 connected dyads ($DDD$).
  2. **Disconnected Triad**: 1 connected dyad ($CDD$).
  3. **Open Triad**: 2 connected dyads ($CCD$).
  4. **Closed Triad**: 3 connected dyads ($CCC$).
- These constitute the classic undirected subgraphs shown on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
plot_triad_u(NULL, "Null", node_size = 16, text_size = 6, title_size = 10) +
  plot_triad_u(data.frame(from = 1, to = 2), "Disconnected",
               node_size = 16, text_size = 6, title_size = 10) +
  plot_triad_u(data.frame(from = c(1, 2), to = c(2, 3)), "Open",
               node_size = 16, text_size = 6, title_size = 10) +
  plot_triad_u(data.frame(from = c(1, 2, 3), to = c(2, 3, 1)), "Closed",
               node_size = 16, text_size = 6, title_size = 10) +
  plot_layout(ncol = 2)
```
:::

::::

---

## The Four Undirected Triads

```{r}
#| label: fig-undir-triads
#| fig-cap: "The four types of undirected triads."
#| fig-cap-location: bottom
#| fig-height: 5.5
#| fig-width: 11

plot_triad_u(NULL, "Null") +
  plot_triad_u(data.frame(from = 1, to = 2), "Disconnected") +
  plot_triad_u(data.frame(from = c(1, 2), to = c(2, 3)), "Open") +
  plot_triad_u(data.frame(from = c(1, 2, 3), to = c(2, 3, 1)), "Closed") +
  plot_layout(nrow = 1)
```

---

## The Holland & Leinhardt (MAN) Labeling System

::: {.incremental}
- Directed triads are classified using the standard Holland and Leinhardt (1970) system.
- Labels are composed of **three digits** representing:
  - **M**: Number of **Mutual** (reciprocal) dyads.
  - **A**: Number of **Asymmetric** (non-reciprocal) dyads.
  - **N**: Number of **Null** (disconnected) dyads.
- Since there are 3 node pairs, **M + A + N = 3** always.
- A trailing letter distinguishes triads sharing identical M-A-N counts — coming up on the next slide.
:::

---

## MAN Labels: The Trailing Letter (D, U, C, T)

:::: {.columns align=center}

::: {.column width="55%"}
- The trailing letter distinguishes triads sharing identical M-A-N counts, based on how the **asymmetric arrows** are arranged:
  - **D** (Down): an **out-star** — a single node sends both arrows ($021D$); when there is a mutual pair, the lone arrow points *into* it ($111D$).
  - **U** (Up): an **in-star** — a single node receives both arrows ($021U$); when there is a mutual pair, the lone arrow points *out of* it ($111U$).
  - **C** (Cyclic): arrows form a directed chain ($021C$) or a closed loop ($030C$).
  - **T** (Transitive): arrows form a transitive hierarchy: $A \rightarrow B \rightarrow C$ with the shortcut $A \rightarrow C$ ($030T$).
- Don't read "up"/"down" literally: the names come from Holland and Leinhardt's canonical drawings, and the arrows' direction on screen depends on how a triad happens to be drawn.
- Keep these four letters in mind: they label every directed triad configuration in the slides that follow.
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.2
#| fig-height: 5.2
# Canonical orientations (xy_top, xy_chain defined in the setup chunk)

pD <- plot_tr(data.frame(from = c(1, 1), to = c(2, 3)), "D: 021D (Down)",
              node_size = 16, text_size = 6, x = xy_top$x, y = xy_top$y)
pU <- plot_tr(data.frame(from = c(2, 3), to = c(1, 1)), "U: 021U (Up)",
              node_size = 16, text_size = 6, x = xy_top$x, y = xy_top$y)
pC <- plot_tr(data.frame(from = c(1, 2, 3), to = c(2, 3, 1)), "C: 030C (Cyclic)",
              node_size = 16, text_size = 6)
pT <- plot_tr(data.frame(from = c(1, 1, 2), to = c(2, 3, 3)), "T: 030T (Transitive)",
              node_size = 16, text_size = 6, x = xy_chain$x, y = xy_chain$y)

pD + pU + pC + pT + plot_layout(nrow = 2)
```
:::

::::

---

## Triads under Anti-Symmetric Ties

::: {.incremental}
- Because reciprocity is strictly forbidden, any triad containing a mutual connection is impossible.
- This prunes the directed triad census from **16 down to exactly seven** allowed configurations:
  1. **Null Triad** ($003$): All three dyads are completely disconnected.
  2. **One-tie Triad** ($012$): Contains a single non-reciprocal directed tie.
  3. **Directed Line** ($021C$): Non-reciprocal ties form a directed path.
  4. **Out-Star Triad** ($021D$): A single node sends ties to both other nodes.
  5. **In-Star Triad** ($021U$): Both other nodes send ties to a single node.
  6. **Cycle Triad** ($030C$): A directed cycle of length three ($A \rightarrow B \rightarrow C \rightarrow A$).
  7. **Transitive Hierarchy** ($030T$): The classic transitive triad ($A \rightarrow B \rightarrow C$ and $A \rightarrow C$).
:::

---

## Anti-Symmetric Triads

```{r}
#| label: fig-antisym-triads
#| fig-align: "center"
#| fig-height: 6.5
#| fig-width: 11

p1 <- plot_tr(data.frame(), "1. 003", node_size = 20, text_size = 8)
p2 <- plot_tr(data.frame(from = 1, to = 2), "2. 012", node_size = 20, text_size = 8)
p3 <- plot_tr(data.frame(from = c(1, 2), to = c(2, 3)), "3. 021C", node_size = 20, text_size = 8)
p4 <- plot_tr(data.frame(from = c(1, 1), to = c(2, 3)), "4. 021D",
              node_size = 20, text_size = 8, x = xy_top$x, y = xy_top$y)
p5 <- plot_tr(data.frame(from = c(2, 3), to = c(1, 1)), "5. 021U",
              node_size = 20, text_size = 8, x = xy_top$x, y = xy_top$y)
p6 <- plot_tr(data.frame(from = c(1, 2, 3), to = c(2, 3, 1)), "6. 030C", node_size = 20, text_size = 8)
p7 <- plot_tr(data.frame(from = c(1, 1, 2), to = c(2, 3, 3)), "7. 030T",
              node_size = 20, text_size = 8, x = xy_chain$x, y = xy_chain$y)

p1 + p2 + p3 + p4 + p5 + p6 + p7 + plot_spacer() + plot_layout(ncol = 4)
```

---

## Anti-Symmetric Triads and Command Chains

::: {.incremental}
- This mathematical pruning simplifies the analysis of hierarchies.
- In formal organization command structures:
  - We expect to see **transitive triads** ($030T$), representing clean, unified chains of command (e.g., "the boss of my boss is my boss").
  - **Cycles** ($030C$) are structurally avoided because they generate circular authority paradoxes, where everyone takes orders from someone else in a loop.
:::

---

## Hierarchies: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- On the right is a **formal organization hierarchy** built strictly from anti-symmetric ties (supervisor of):
  - Every connection flows downward.
  - The graph is completely acyclic (no loops).
  - Every node (except the President root) has an indegree of exactly 1.
- Removing the transitive and cyclic "noise" results in a clean **directed tree**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_hierarchy <- tbl_graph(
  nodes = data.frame(name = c("Pres", "VP 1", "VP 2", "Staff 1", "Staff 2")),
  edges = data.frame(from = c(1, 1, 2, 3), to = c(2, 3, 4, 5)),
  directed = TRUE
)
ggraph(gr_hierarchy, layout = "tree") +
  geom_edge_link(color = "#2c3e50", edge_width = 1.5,
                 arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                 end_cap = circle(13, 'mm'), start_cap = circle(13, 'mm')) +
  geom_node_point(size = 38, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 5.0, fontface = "bold", color = "#2c3e50") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Triads under Asymmetric Ties

::: {.incremental}
- When relations are asymmetric, reciprocity is allowed but optional.
- Because each dyadic slot can take on any of the **three** MAN states, we can generate all **sixteen** directed triad configurations.
- On the next slide, we show the **nine directed configurations** that require at least one **reciprocal (mutual) dyad** to exist:
  - **Single Mutual**: $102$, $111D$, $111U$.
  - **Double Mutual**: $201$.
  - **Mutual-Mixed Closed**: $120C$, $120D$, $120U$.
  - **Cohesive Closed**: $210$, $300$.
:::

---

## Reciprocal Triads (Part 1: Single Mutual)

```{r}
#| label: fig-recip-triads-slides-1
#| fig-align: "center"
#| fig-height: 5.5
#| fig-width: 11

    p1 <- plot_tr(data.frame(from=c(1,2), to=c(2,1)), "1. 102")
    p2 <- plot_tr(data.frame(from=c(2,1,3), to=c(1,2,1)), "2. 111D")
    p3 <- plot_tr(data.frame(from=c(2,1,1), to=c(1,2,3)), "3. 111U")
    
    p1 + p2 + p3 + plot_spacer() + plot_layout(nrow = 1)
```

---

## Reciprocal Triads (Part 2: Double Mutual & Mixed Closed)

```{r}
#| label: fig-recip-triads-slides-2
#| fig-align: "center"
#| fig-height: 5.5
#| fig-width: 11

    p4 <- plot_tr(data.frame(from=c(2,1,3,2), to=c(1,2,2,3)), "4. 201")
    p5 <- plot_tr(data.frame(from=c(1,2,2,3), to=c(2,1,3,1)), "5. 120C")
    p6 <- plot_tr(data.frame(from=c(1,2,3,3), to=c(2,1,1,2)), "6. 120D")
    
    p4 + p5 + p6 + plot_spacer() + plot_layout(nrow = 1)
```

---

## Reciprocal Triads (Part 3: In-Star Closed & Cohesive)

```{r}
#| label: fig-recip-triads-slides-3
#| fig-align: "center"
#| fig-height: 5.5
#| fig-width: 11

    p7 <- plot_tr(data.frame(from=c(1,2,1,2), to=c(2,1,3,3)), "7. 120U")
    p8 <- plot_tr(data.frame(from=c(2,1,3,2,1), to=c(1,2,2,3,3)), "8. 210")
    p9 <- plot_tr(data.frame(from=c(2,1,3,2,1,3), to=c(1,2,2,3,3,1)), "9. 300")
    
    p7 + p8 + p9 + plot_spacer() + plot_layout(nrow = 1)
```

---

## The Full Directed Triad Census

```{r}
#| label: fig-triad-census
#| fig-cap: "The 16 directed triad isomorphism classes. Titles in green are permitted under anti-symmetric relations (no mutual dyads); titles in red require at least one reciprocal (mutual) dyad."
#| fig-cap-location: bottom
#| fig-align: "center"
#| fig-width: 11
#| fig-height: 9

triad_census <- list(
  list("003 — Null", data.frame()),
  list("012 — One Tie", data.frame(from = 1, to = 2)),
  list("021C — Line", data.frame(from = c(1, 2), to = c(2, 3))),
  list("021D — Out-Star", data.frame(from = c(1, 1), to = c(2, 3)), xy = xy_top),
  list("021U — In-Star", data.frame(from = c(2, 3), to = c(1, 1)), xy = xy_top),
  list("030C — Cycle", data.frame(from = c(1, 2, 3), to = c(2, 3, 1))),
  list("030T — Transitive", data.frame(from = c(1, 1, 2), to = c(2, 3, 3)), xy = xy_chain),
  list("102 — One Mutual", data.frame(from = c(1, 2), to = c(2, 1))),
  list("111D — Mutual + In", data.frame(from = c(2, 1, 3), to = c(1, 2, 1))),
  list("111U — Mutual + Out", data.frame(from = c(2, 1, 1), to = c(1, 2, 3))),
  list("201 — Two Mutuals", data.frame(from = c(2, 1, 3, 2), to = c(1, 2, 2, 3))),
  list("120C — Mutual + Cyclic", data.frame(from = c(1, 2, 2, 3), to = c(2, 1, 3, 1))),
  list("120D — Mutual + Down", data.frame(from = c(1, 2, 3, 3), to = c(2, 1, 1, 2))),
  list("120U — Mutual + Up", data.frame(from = c(1, 2, 1, 2), to = c(2, 1, 3, 3))),
  list("210 — Two Mutuals + Tie", data.frame(from = c(2, 1, 3, 2, 1), to = c(1, 2, 2, 3, 3))),
  list("300 — All Mutual", data.frame(from = c(2, 1, 3, 2, 1, 3), to = c(1, 2, 2, 3, 3, 1)))
)

panels <- lapply(seq_along(triad_census), function(i) {
  color <- if (i <= 7) "#27ae60" else "#c0392b"
  tc <- triad_census[[i]]
  if (is.null(tc$xy)) {
    plot_tr(tc[[2]], tc[[1]], node_size = 14, text_size = 5.5, title_color = color)
  } else {
    plot_tr(tc[[2]], tc[[1]], node_size = 14, text_size = 5.5, title_color = color,
            x = tc$xy$x, y = tc$xy$y)
  }
})

Reduce(`+`, panels) + plot_layout(ncol = 4)
```

---

## Conclusion & Core Takeaways

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 9
#| fig-height: 2.8

p_sym <- plot_triad_u(data.frame(from = c(1, 2, 3), to = c(2, 3, 1)),
                      "Symmetric: 4 Triads", node_size = 14, text_size = 5.5, title_size = 10)
p_antisym <- plot_tr(data.frame(from = c(1, 1, 2), to = c(2, 3, 3)),
                     "Anti-Symmetric: 7 Triads", node_size = 14, text_size = 5.5)
p_asym <- plot_tr(data.frame(from = c(2, 1, 3, 2, 1, 3), to = c(1, 2, 2, 3, 3, 1)),
                  "Asymmetric: 16 Triads", node_size = 14, text_size = 5.5)

p_sym + p_antisym + p_asym + plot_layout(nrow = 1)
```

::: {.incremental}
- **Ties define dyads, and dyads define triads**:
  - The mathematical type of relationship strictly controls the possible micro-structures in a network.
- **Symmetric Ties** $\implies$ connected vs. disconnected dyads $\implies$ 4 undirected triads.
- **Anti-Symmetric Ties** $\implies$ disconnected vs. non-reciprocal dyads $\implies$ 7 directed triads.
- **Asymmetric Ties** $\implies$ disconnected, non-reciprocal, or reciprocal dyads $\implies$ 16 directed triads (Triad Census).
- Understanding these micro-motifs is the key to decoding the macro-structures of power, friendship, and organization in society.
:::
