---
title: "Ties, Dyads, and Triads"
subtitle: "The Motifs of Social Network Analysis"
format:
  revealjs:
    scrollable: true
    theme: default
    css: ucla-theme.css
    width: 1280
    height: 720
    margin: 0.1
    min-scale: 0.2
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggraph)
library(tidygraph)
library(igraph)
library(dplyr)
library(kableExtra)
library(patchwork)

# Perfect triangle coordinates for 3-node triad drawings
x_triad <- c(0, 2, 1)
y_triad <- c(0, 0, 1.2)

# Helper function to plot individual directed triads
gr_c <- create_empty(3, directed = TRUE)
plot_tr <- function(edges, title) {
  g <- gr_c %>% mutate(name = toupper(letters[1:3]))
  if (nrow(edges) > 0) {
    g <- g %>% bind_edges(edges)
  }
  p <- ggraph(g, layout = "manual", x = x_triad, y = y_triad) +
    geom_edge_link(color = "#2c3e50", edge_width = 1.1,
                   arrow = arrow(length = unit(3, 'mm'), type = "closed"),
                   end_cap = circle(6, 'mm'), start_cap = circle(6, 'mm')) +
    geom_node_point(size = 12, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.25) +
    geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "#2c3e50") +
    theme_graph() + ggtitle(title) + coord_cartesian(clip = "off") +
    theme(
      plot.title = element_text(size = 10, face="bold", hjust=0.5, color = "#2c3e50"),
      plot.margin = margin(5, 5, 5, 5)
    )
  return(p)
}

# Function to create a clean, minimalist 2-node graph plot from the book chapter
plot_tie_type <- function(nodes, edges, title, subtitle, directed = TRUE, has_cross = FALSE) {
  g <- tbl_graph(nodes = nodes, edges = edges)
  p <- ggraph(g, layout = "manual", x = x, y = y)
  
  if (directed) {
    p <- p + geom_edge_arc(aes(strength = strength, color = color, linetype = linetype),
                           arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                           end_cap = circle(8, 'mm'),
                           start_cap = circle(8, 'mm'),
                           width = 1.1,
                           show.legend = FALSE) +
             scale_edge_color_identity() +
             scale_edge_linetype_identity()
  } else {
    p <- p + geom_edge_link(color = "#2c3e50",
                            width = 1.1,
                            end_cap = circle(8, 'mm'),
                            start_cap = circle(8, 'mm'),
                            show.legend = FALSE)
  }
  
  pBase <- p + geom_node_point(size = 14, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
    geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "#2c3e50")
  
  if (has_cross) {
    pBase <- pBase + annotate("text", x = 1.5, y = 0.90, label = "✕", color = "#e74c3c", size = 8, fontface = "bold")
  }
  
  pBase <- pBase + 
    labs(title = title, subtitle = subtitle) +
    theme_graph() +
    theme(
      plot.title = element_text(size = 12, face = "bold", color = "#2c3e50", hjust = 0.5),
      plot.subtitle = element_text(size = 9.5, color = "#7f8c8d", hjust = 0.5, lineheight = 1.1),
      plot.margin = margin(10, 10, 10, 10)
    ) +
    coord_fixed(ratio = 1, xlim = c(0.6, 2.4), ylim = c(0.65, 1.35), clip = "off")
  
  return(pBase)
}

# Define Nodes
nodes_2 <- data.frame(name = c("A", "B"), x = c(1, 2), y = c(1, 1))

# Define Edges for symmetric, reciprocal, non-reciprocal, and anti-symmetric
edges_sym <- data.frame(from = 1, to = 2)
edges_rec <- data.frame(from = c(1, 2), to = c(2, 1), strength = c(0.1, 0.1), color = c("#2c3e50", "#2c3e50"), linetype = c("solid", "solid"))
edges_nonrec <- data.frame(from = 1, to = 2, strength = 0, color = "#2c3e50", linetype = "solid")
edges_anti <- data.frame(
  from = c(1, 2), 
  to = c(2, 1), 
  strength = c(0.1, 0.1), 
  color = c("#2c3e50", "#e74c3c"), 
  linetype = c("solid", "dashed")
)
```

# Part 1: The Taxonomy of Ties

---

## The Relational Basis of Networks

::: {.incremental}
- In social network analysis, a **graph** represents concrete **relationships** (ties) between actors.
- Not all relationships are created equal!
  - Some ties are naturally mutual (e.g., being siblings).
  - Others are directional (e.g., seeking advice).
  - Some strictly forbid mutuality (e.g., being a supervisor).
- To study networks, we must separate the **mathematical property of the relationship** (the tie type) from its **empirical realization** (the dyad or triad).
:::

---

## The Sociological Classification of Types of Ties

::: {.incremental}
- Instead of treating all connections identically, sociologists classify social ties into **four major types** ([Borgatti et al. 2009](https://doi.org/10.1126/science.1165821)):
  1. **Similarities**: Having something in common (e.g., co-location, shared attributes).
  2. **Social Relations**: Culturally recognized labels or sentiments (e.g., friend, spouse, boss).
  3. **Interactions**: Regular communication or behavior (e.g., texting, talking).
  4. **Flows**: Transmission of resources or states (e.g., virus infection, lending money).
:::

---

## States vs. Events in Social Networks

::: {.incremental}
- We can divide these four types of ties into two fundamental clusters:
- **States (Static Situations)**:
  - **Similarities** & **Social Relations**.
  - Represent continuous, long-standing conditions.
  - *Examples*: Being in the same class, being siblings, being coworker/boss.
- **Events (Discrete Activities)**:
  - **Interactions** & **Flows**.
  - Occur at specific points in time and are transient.
  - *Examples*: Texting someone, lending money, transmitting a virus.
:::

---

## Graph-Theoretic Classification of Ties

::: {.incremental}
- While sociologists classify ties by their **substantive content** (similarities, relations, interactions, flows)...
- Mathematicians and graph theorists classify ties by their **formal, directionality properties**:
  - **Symmetric Ties**: Inherently mutual and undirected ($AB \iff BA$).
  - **Asymmetric Ties**: Directional, where reciprocity is optional ($AB \implies$ optional $BA$).
  - **Anti-Symmetric Ties**: Directional, where reciprocity is forbidden ($AB \implies \neg BA$).
- **The Link**: To represent a network, we must map our sociological tie types (e.g. boss-of, friend-of) onto these three graph-theoretic classes!
:::

---

## 1. Symmetric Ties & Similarities

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: A relation where connection is built-in and mutual by construction:
  - Formally: $AB \iff BA$
- **Representation**: Represented by **undirected graphs**.
- **Structural Constraint**: Pairs of actors can only be **disconnected** or **connected** by a single undirected edge.
- **Symmetric Similarities**:
  - *Similarities are necessarily symmetric*:
    - If you are in the same place at the same time as someone, they are there with you.
    - If you are zooming with someone, they are zooming with you.
    - If you go to UCLA with someone, they also go to UCLA with you.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
plot_tie_type(nodes_2, edges_sym, "Symmetric Tie", "e.g., 'is sibling of'", directed = FALSE)
```
:::

::::

---

## 2. Asymmetric Ties & Interactions

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: A relation with natural directionality where reciprocity is possible but *not guaranteed*:
  - Formally: $AB \implies$ nothing about $BA$
- **Representation**: Represented by **directed graphs** (digraphs).
- **Structural Constraint**: Reciprocity is an **empirical event** (it may or may not occur), yielding three possible dyad configurations.
- **Asymmetric Interactions**:
  - *Interactions/sentiments are naturally asymmetric*:
    - You can follow someone on Twitter, but they don't have to follow you back.
    - You can send a text to someone, but they don't have to respond.
    - You can seek advice from someone without them seeking advice from you.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6.5
#| fig-height: 5.5
p1 <- plot_tie_type(nodes_2, edges_rec, "Asymmetric: Reciprocal Dyad", "e.g., 'A likes B' & 'B likes A'", directed = TRUE)
p2 <- plot_tie_type(nodes_2, edges_nonrec, "Asymmetric: Non-Reciprocal Dyad", "e.g., 'A sends email to B'", directed = TRUE)
p1 + p2
```
:::

::::

---

## 3. Anti-Symmetric Ties & Hierarchies

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- **Definition**: An inherently directional relation where reciprocity is **strictly forbidden**:
  - Formally: $AB \implies \neg BA$
- **Representation**: Represented by **directed graphs** (oriented graphs, DAGs, or trees).
- **Structural Constraint**: Reciprocity is structurally impossible, yielding only two possible dyad configurations.
- **Anti-Symmetric Hierarchies**:
  - *Hierarchical role relations are anti-symmetric*:
    - Kinship: If you are someone's parent, they cannot be your parent.
    - Formal: If you are someone's boss, they cannot be your boss at the same time.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
plot_tie_type(nodes_2, edges_anti, "Anti-Symmetric Tie", "e.g., 'A is boss of B'", directed = TRUE, has_cross = TRUE)
```
:::

::::

---

## Mapping Sociological to Graph-Theoretic Classifications

| Sociological Classification (Types of Ties) | Graph-Theoretic Classification (Mathematical Properties) | Allowed Dyads | Visual Graph Representation |
|:---|:---|:---|:---|
| **Co-membership, Propinquity** | **Symmetric** | • Disconnected<br>• Connected | Undirected Graph ($A - B$) |
| **Sentiments, Advice-seeking** | **Asymmetric** | • Disconnected<br>• Non-reciprocal<br>• Reciprocal | Directed Graph ($A \rightarrow B$ or $A \leftrightarrow B$) |
| **Formal Hierarchies, Kinship** | **Anti-symmetric** | • Disconnected<br>• Non-reciprocal | Oriented Graph/Tree ($A \rightarrow B$, no cycles) |

---

## Multiplexity in Social Relations

::: {.incremental}
- In real-world social networks, actors are rarely linked by only one relationship.
- **Multiplexity** refers to the presence of multiple, distinct types of ties linking the same pair of nodes simultaneously.
  - e.g., $A$ and $B$ are coworkers (**symmetric** tie) AND $A$ is the supervisor of $B$ (**anti-symmetric** tie) AND they are friends (**asymmetric** tie).
- Multiplexity adds rich, multi-layered complexity to networks, as the behavior or pattern in one relation heavily influences and moderates others.
:::

---

## Multiplexity: A Visual Example

:::: {.columns align=center}

::: {.column width="40%"}
::: {.incremental}
- Three nodes connected by multiple parallel relations:
  - **Symmetric Sibling**: (blue dotted lines).
  - **Anti-Symmetric Parent-of**: (solid black arrows).
  - **Asymmetric Advice**: (dashed purple arrows).
- *One event can affect multiple layers simultaneously.*
:::
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6.5
#| fig-height: 5.5
edges_mult <- data.frame(
  from = c(1, 1, 2, 3, 2),
  to   = c(2, 3, 3, 2, 1),
  type = c("Parent-of", "Parent-of", "Sibling", "Sibling", "Advice")
)
gr_mult <- tbl_graph(nodes = data.frame(name = c("A", "B", "C"), x = c(1, 0, 2), y = c(1, 0, 0)), edges = edges_mult, directed = TRUE)
ggraph(gr_mult, layout = "manual", x = x, y = y) +
  geom_edge_parallel(aes(color = type, linetype = type),
                     arrow = arrow(length = unit(3.5, 'mm'), type = "closed"),
                     start_cap = circle(8, 'mm'), end_cap = circle(8, 'mm'),
                     sep = unit(4.5, 'mm'), edge_width = 1.25) +
  scale_edge_color_manual(values = c("Parent-of" = "#2c3e50", "Sibling" = "#2980b9", "Advice" = "#8e44ad")) +
  scale_edge_linetype_manual(values = c("Parent-of" = "solid", "Sibling" = "dotted", "Advice" = "dashed")) +
  geom_node_point(size = 18, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "#2c3e50") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## The Growth of Social Networks

::: {.incremental}
- Social networks are highly dynamic, growing and evolving over time.
- A classic example of network growth is **the marriage cascade**:
  1. A romantic tie (an asymmetric, non-kin relation) forms between two individuals from separate families: `You` $\rightarrow$ `Riley`.
  2. This culminates in a marriage tie (a symmetric, kin relation): `You` $\leftrightarrow$ `Riley`.
  3. This single event immediately triggers a cascade of new "in-law" ties:
     - **Symmetric Sibling-in-Law**: between `You` $\leftrightarrow$ `Bro` and `Riley` $\leftrightarrow$ `Sis`.
     - **Anti-Symmetric Parent-in-Law**: `Parent 1` $\rightarrow$ `Riley` and `Parent 2` $\rightarrow$ `You`.
  4. This merges two previously separate, independent kinship trees into a single, cohesive, multiplex social network.
:::

# Part 2: Dyads and Their Realizations

---

## What is a Dyad?

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- A **dyad** is any pair of nodes in a graph, whether they are joined by an edge or not.
- Mathematically, a dyad is *any subgraph of order two* ($n = 2$).
- Crucially, a dyad represents the **realized configuration** of ties between two nodes.
- **Sociological Insight**: Ties that do *not* exist (disconnected dyads) are often just as socially meaningful as those that do!
- **Dyads: The Building Blocks**:
  - The simplest possible subgraph of order two.
  - Can contain zero, one, or two directed edges depending on relational constraints.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_dyad_base <- create_empty(2, directed = FALSE) %>% mutate(name = toupper(letters[1:2]))
ggraph(gr_dyad_base, layout = 'linear') +
  geom_node_point(size = 25, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 10, fontface = "bold", color = "#2c3e50") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Dyads under Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Because undirected graphs do not allow directionality, a symmetric relation can only take on **two** dyadic states:
  1. **Disconnected Dyads** (historically called **null dyads**): The two nodes do not share an edge (size zero).
  2. **Connected Dyads**: The two nodes share an edge (size one).
- Every pair of actors in an undirected graph is either connected or disconnected.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5.5
gr1 <- create_empty(2, directed = FALSE) %>% mutate(name = c("A", "B"))
gr2 <- create_empty(2, directed = FALSE) %>% bind_edges(data.frame(from = 1, to = 2)) %>% mutate(name = c("C", "D"))

p1 <- ggraph(gr1, layout = 'linear') +
  geom_node_point(size = 15, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "#2c3e50") +
  theme_graph() + ggtitle("Disconnected (Size Zero)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5, color = "#2c3e50"))

p2 <- ggraph(gr2, layout = 'linear') +
  geom_edge_link(color = "#2c3e50", width = 1.25, start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm')) +
  geom_node_point(size = 15, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "#2c3e50") +
  theme_graph() + ggtitle("Connected (Size One)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5, color = "#2c3e50"))

p1 / p2
```
:::

::::

---

## Dyads under Asymmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- In directed graphs, reciprocity is possible but optional.
- Thus, asymmetric ties can yield **three** distinct kinds of dyads (the classic **MAN classification**):
  1. **Disconnected Dyads** (Null): Neither node sends a directed tie to the other (size zero).
  2. **Non-Reciprocal Dyads** (Asymmetric): One node sends a tie to another, but it is not reciprocated (size one).
  3. **Reciprocal Dyads** (Mutual): Both nodes send directed ties to each other (size two).
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5.5
gr1 <- create_empty(2, directed = TRUE) %>% mutate(name = c("A", "B"))
gr2 <- create_empty(2, directed = TRUE) %>% bind_edges(data.frame(from = 1, to = 2)) %>% mutate(name = c("C", "D"))
gr3 <- create_empty(2, directed = TRUE) %>% bind_edges(data.frame(from = c(1, 2), to = c(2, 1))) %>% mutate(name = c("E", "F"))

p1 <- ggraph(gr1, layout = 'linear') +
  geom_node_point(size = 12, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.25) +
  geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "#2c3e50") +
  theme_graph() + ggtitle("Null (Disconnected)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5, color = "#2c3e50"))

p2 <- ggraph(gr2, layout = 'linear') +
  geom_edge_link(color = "#2c3e50", edge_width = 1,
                  arrow = arrow(length = unit(3, 'mm'), type = "closed"), 
                  start_cap = circle(5, 'mm'), end_cap = circle(5, 'mm')) +
  geom_node_point(size = 12, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.25) +
  geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "#2c3e50") +
  theme_graph() + ggtitle("Asymmetric (Non-reciprocal)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5, color = "#2c3e50"))

p3 <- ggraph(gr3, layout = 'linear') +
  geom_edge_parallel(color = "#2c3e50", edge_width = 1,
                  arrow = arrow(length = unit(3, 'mm'), type = "closed"), 
                  start_cap = circle(5, 'mm'), end_cap = circle(5, 'mm'), sep = unit(4, 'mm')) +
  geom_node_point(size = 12, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.25) +
  geom_node_text(aes(label = name), size = 5, fontface = "bold", color = "#2c3e50") +
  theme_graph() + ggtitle("Mutual (Reciprocal)") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5, color = "#2c3e50"))

p1 / p2 / p3
```
:::

::::

---

## Separating Tie Properties from Dyad Realizations

::: {.incremental}
- It is critical to keep the property of the **tie** separate from the property of the **dyad**:
  - **Asymmetric Tie**: Describes the type of *relation* (directionality is allowed; reciprocity is optional).
  - **Non-Reciprocal Dyad**: Describes a realized *configuration* of a pair of nodes where reciprocity failed to occur empirically.
  - **Reciprocal Dyad**: Describes an empirical configuration composed of a *pair* of directed edges under an asymmetric relation.
- Confusing "asymmetric tie" with "asymmetric dyad" is a common conceptual error!
:::

---

## Dyads under Anti-Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Because reciprocity is strictly forbidden under anti-symmetric ties, mutual (reciprocal) dyads cannot occur.
- Therefore, anti-symmetric ties can only yield **two** types of dyads:
  1. **Disconnected Dyads**: The two nodes do not share any directed edge.
  2. **Non-Reciprocal Dyads**: Only a single one-way directed tie exists between them ($A \rightarrow B$).
- This simplifies the dyad census of hierarchical networks: every pair of nodes is either completely disconnected or connected by a strictly non-reciprocal arrow.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 5.5
gr1 <- create_empty(2, directed = TRUE) %>% mutate(name = c("A", "B"))
gr2 <- create_empty(2, directed = TRUE) %>% bind_edges(data.frame(from = 1, to = 2)) %>% mutate(name = c("C", "D"))

p1 <- ggraph(gr1, layout = 'linear') +
  geom_node_point(size = 15, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "#2c3e50") +
  theme_graph() + ggtitle("Disconnected") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5, color = "#2c3e50"))

p2 <- ggraph(gr2, layout = 'linear') +
  geom_edge_link(color = "#2c3e50", edge_width = 1.25,
                  arrow = arrow(length = unit(4, 'mm'), type = "closed"), 
                  start_cap = circle(6, 'mm'), end_cap = circle(6, 'mm')) +
  geom_node_point(size = 15, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "#2c3e50") +
  theme_graph() + ggtitle("Non-Reciprocal") + coord_cartesian(clip = "off") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5, color = "#2c3e50"))

p1 / p2
```
:::

::::

# Part 3: Triads and Motifs

---

## What is a Triad?

::: {.incremental}
- A **triad** is any set of three nodes, whether joined by edges or not.
- Mathematically, a triad is *any subgraph of order three* ($n = 3$).
- Triads have incredibly powerful sociological properties (e.g., structural balance, brokerage, coalitions).
- A triad can be thought of as a **concatenation of three dyads**: $\{A, B\}$, $\{B, C\}$, and $\{A, C\}$.
- Just like dyads, the allowed configurations of a triad depend entirely on the relational properties of its ties.
:::

---

## Triads under Symmetric Ties

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- Since undirected dyads can only be connected or disconnected, there are only **four** possible undirected triad configurations:
  1. **Null Triad**: 0 connected dyads ($DDD$).
  2. **Disconnected Triad**: 1 connected dyad ($CDD$).
  3. **Open Triad**: 2 connected dyads ($CCD$).
  4. **Closed Triad**: 3 connected dyads ($CCC$).
- These constitute the classic undirected subgraphs shown on the right.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr <- create_empty(3, directed = FALSE) %>% mutate(name = toupper(letters[1:3]))
p3_ex <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
    geom_node_point(size = 15, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
    geom_node_text(aes(label = name), size = 6, fontface = "bold", color = "#2c3e50") +
    theme_graph() + coord_cartesian(clip = "off")
p3_ex
```
:::

::::

---

## The Four Undirected Triads

```{r}
#| label: fig-undir-triads
#| fig-cap: "The four types of undirected triads."
#| fig-cap-location: bottom
#| fig-height: 5.5
#| fig-width: 11

    gr <- create_empty(3, directed = FALSE) %>% mutate(name = toupper(letters[1:3]))
    p1 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_node_point(size = 20, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
        geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "#2c3e50") +
        theme_graph() + ggtitle("Null") + coord_cartesian(clip = "off") +
        theme(plot.title = element_text(size = 11, face="bold", hjust=0.5, color = "#2c3e50"))
    
    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        mutate(name = toupper(letters[1:3]))
    p2 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_edge_link(color = "#2c3e50", width = 1.1, start_cap = circle(8, 'mm'), end_cap = circle(8, 'mm')) +
        geom_node_point(size = 20, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
        geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "#2c3e50") +
        theme_graph() + ggtitle("Disconnected") + coord_cartesian(clip = "off") +
        theme(plot.title = element_text(size = 11, face="bold", hjust=0.5, color = "#2c3e50"))

    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        bind_edges(data.frame(from = 2, to = 3)) %>%
        mutate(name = toupper(letters[1:3]))
    p3 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_edge_link(color = "#2c3e50", width = 1.1, start_cap = circle(8, 'mm'), end_cap = circle(8, 'mm')) +
        geom_node_point(size = 20, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
        geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "#2c3e50") +
        theme_graph() + ggtitle("Open") + coord_cartesian(clip = "off") +
        theme(plot.title = element_text(size = 11, face="bold", hjust=0.5, color = "#2c3e50"))
    
    gr <- create_empty(3, directed = FALSE) %>% 
        bind_edges(data.frame(from = 1, to = 2)) %>%
        bind_edges(data.frame(from = 2, to = 3)) %>%
        bind_edges(data.frame(from = 3, to = 1)) %>%
        mutate(name = toupper(letters[1:3]))
    p4 <- ggraph(gr, layout = "manual", x = x_triad, y = y_triad) +
        geom_edge_link(color = "#2c3e50", width = 1.1, start_cap = circle(8, 'mm'), end_cap = circle(8, 'mm')) +
        geom_node_point(size = 20, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
        geom_node_text(aes(label = name), size = 8, fontface = "bold", color = "#2c3e50") +
        theme_graph() + ggtitle("Closed") + coord_cartesian(clip = "off") +
        theme(plot.title = element_text(size = 11, face="bold", hjust=0.5, color = "#2c3e50"))
    
    p1 + p2 + p3 + p4 + plot_layout(nrow = 1)
```

---

## Triads under Anti-Symmetric Ties

::: {.incremental}
- Because reciprocity is strictly forbidden, any triad containing a mutual connection is impossible.
- This prunes the directed triad census from **16 down to exactly seven** allowed configurations:
  1. **Null Triad** ($NNN$, panel 1): All three dyads are completely disconnected.
  2. **One-tie Triad** ($NNA$, panel 2): Contains a single non-reciprocal directed tie.
  3. **Directed Line** ($AAN(L)$, panel 3): Non-reciprocal ties form a directed path.
  4. **Out-Star Triad** ($AAN(O)$, panel 4): A single node sends ties to both other nodes.
  5. **In-Star Triad** ($AAN(I)$, panel 5): Both other nodes send ties to a single node.
  6. **Cycle Triad** ($AAA(C)$, panel 6): A directed cycle of length three ($A \rightarrow B \rightarrow C \rightarrow A$).
  7. **Transitive Hierarchy** ($AAA(I/O)$, panel 7): The classic transitive triad ($A \rightarrow B \rightarrow C$ and $A \rightarrow C$).
:::

---

## The Seven Anti-Symmetric Triads

```{r}
#| label: fig-antisym-triads-slides
#| fig-cap: "The Seven Permitted Triads Under Anti-Symmetric Ties (Strict Hierarchies)"
#| fig-cap-location: bottom
#| fig-height: 4.5
#| fig-width: 10

    p1 <- plot_tr(data.frame(), "1. NNN")
    p2 <- plot_tr(data.frame(from=1, to=2), "2. NNA")
    p3 <- plot_tr(data.frame(from=c(1,2), to=c(2,3)), "3. AAN(L)")
    p4 <- plot_tr(data.frame(from=c(1,1), to=c(2,3)), "4. AAN(O)")
    p5 <- plot_tr(data.frame(from=c(2,3), to=c(1,1)), "5. AAN(I)")
    p6 <- plot_tr(data.frame(from=c(1,2,3), to=c(2,3,1)), "6. AAA(C)")
    p7 <- plot_tr(data.frame(from=c(1,1,2), to=c(2,3,3)), "7. AAA(I/O)")
    
    (p1 + p2 + p3 + p4) / (p5 + p6 + p7 + plot_spacer())
```

---

## Anti-Symmetric Triads and Command Chains

::: {.incremental}
- This mathematical pruning simplifies the analysis of hierarchies.
- In formal organization command structures:
  - We expect to see **transitive triads** ($AAA(I/O)$), representing clean, unified chains of command (e.g., "the boss of my boss is my boss").
  - **Cycles** ($AAA(C)$) are structurally avoided because they generate circular authority paradoxes, where everyone takes orders from someone else in a loop.
:::

---

## Hierarchies: A Visual Example

:::: {.columns align=center}

::: {.column width="60%"}
::: {.incremental}
- On the right is a **formal organization hierarchy** built strictly from anti-symmetric ties (supervisor of):
  - Every connection flows downward.
  - The graph is completely acyclic (no loops).
  - Every node (except the President root) has an indegree of exactly 1.
- Removing the transitive and cyclic "noise" results in a clean **directed tree**.
:::
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 5.5
#| fig-height: 5.5
gr_hierarchy <- tbl_graph(
  nodes = data.frame(name = c("Pres", "VP 1", "VP 2", "Staff 1", "Staff 2")),
  edges = data.frame(from = c(1, 1, 2, 3), to = c(2, 3, 4, 5)),
  directed = TRUE
)
ggraph(gr_hierarchy, layout = "tree") +
  geom_edge_link(color = "#2c3e50", edge_width = 1.5,
                 arrow = arrow(length = unit(4, 'mm'), type = "closed"),
                 end_cap = circle(13, 'mm'), start_cap = circle(13, 'mm')) +
  geom_node_point(size = 38, fill = "#ecf0f1", color = "#2c3e50", shape = 21, stroke = 1.5) +
  geom_node_text(aes(label = name), size = 5.0, fontface = "bold", color = "#2c3e50") +
  theme_graph() + coord_cartesian(clip = "off")
```
:::

::::

---

## Triads under Asymmetric Ties

::: {.incremental}
- When relations are asymmetric, reciprocity is allowed but optional.
- Because each dyadic slot can take on any of the **three** MAN states, we can generate all **sixteen** directed triad configurations.
- On the next slide, we show the **nine directed configurations** that require at least one **reciprocal (mutual) dyad** to exist:
  - **Single Mutual**: $NNM$ (panel 1), $MAN1$ (panel 2), $MAN2$ (panel 3).
  - **Double Mutual**: $MMN$ (panel 4).
  - **Mutual-Mixed Closed**: $MAA(C)$ (panel 5), $MAA(O)$ (panel 6), $MAA(I)$ (panel 7).
  - **Cohesive Closed**: $MMA$ (panel 8), $MMM$ (panel 9).
:::

---

## The Nine Reciprocal Triads

```{r}
#| label: fig-recip-triads-slides
#| fig-cap: "The Nine Directed Triads Requiring Reciprocal (Mutual) Ties"
#| fig-cap-location: bottom
#| fig-height: 4.5
#| fig-width: 10

    p1 <- plot_tr(data.frame(from=c(1,2), to=c(2,1)), "1. NNM")
    p2 <- plot_tr(data.frame(from=c(2,1,3), to=c(1,2,1)), "2. MAN1")
    p3 <- plot_tr(data.frame(from=c(2,1,1), to=c(1,2,3)), "3. MAN2")
    p4 <- plot_tr(data.frame(from=c(2,1,3,2), to=c(1,2,2,3)), "4. MMN")
    p5 <- plot_tr(data.frame(from=c(1,2,2,3), to=c(2,1,3,1)), "5. MAA(C)")
    p6 <- plot_tr(data.frame(from=c(1,2,3,3), to=c(2,1,1,2)), "6. MAA(O)")
    p7 <- plot_tr(data.frame(from=c(1,2,1,2), to=c(2,1,3,3)), "7. MAA(I)")
    p8 <- plot_tr(data.frame(from=c(2,1,3,2,1), to=c(1,2,2,3,3)), "8. MMA")
    p9 <- plot_tr(data.frame(from=c(2,1,3,2,1,3), to=c(1,2,2,3,3,1)), "9. MMM")
    
    (p1 + p2 + p3 + p4 + p5) / (p6 + p7 + p8 + p9 + plot_spacer())
```

---

## Triad Summary: Strictly Non-Reciprocal

These 7 configurations contain **zero reciprocal (mutual) ties** (referenced in the 7-panel anti-symmetric figure):

* **0 connected dyads**:
  - **NNN** (Null, Anti-Symmetric panel 1)
* **1 connected dyad**:
  - **NNA** (Non-reciprocal, Anti-Symmetric panel 2)
* **2 connected dyads (Open)**:
  - **AAN(L)** (Directed Line, Anti-Symmetric panel 3)
  - **AAN(O)** (Out-Star, Anti-Symmetric panel 4)
  - **AAN(I)** (In-Star, Anti-Symmetric panel 5)
* **3 connected dyads (Closed)**:
  - **AAA(C)** (Directed Cycle, Anti-Symmetric panel 6)
  - **AAA(I/O)** (Transitive Hierarchy, Anti-Symmetric panel 7)

---

## Triad Summary: Requiring Reciprocal Ties

These 9 configurations contain **at least one reciprocal (mutual) tie** (referenced in the 9-panel reciprocal figure):

* **1 connected dyad**:
  - **NNM** (Reciprocal, Reciprocal panel 1)
* **2 connected dyads (Open)**:
  - **MAN1** & **MAN2** (Mixed-reciprocity, Reciprocal panels 2 & 3)
  - **MMN** (Double reciprocal, Reciprocal panel 4)
* **3 connected dyads (Closed)**:
  - **MAA(C)**, **MAA(O)**, & **MAA(I)** (Mixed closed, Reciprocal panels 5, 6, & 7)
  - **MMA** (Double reciprocal, Reciprocal panel 8)
  - **MMM** (Triple reciprocal, Reciprocal panel 9)

---

## Conclusion & Core Takeaways

::: {.incremental}
- **Ties define dyads, and dyads define triads**:
  - The mathematical type of relationship strictly controls the possible micro-structures in a network.
- **Symmetric Ties** $\implies$ connected vs. disconnected dyads $\implies$ 4 undirected triads.
- **Anti-Symmetric Ties** $\implies$ disconnected vs. non-reciprocal dyads $\implies$ 7 directed triads.
- **Asymmetric Ties** $\implies$ disconnected, non-reciprocal, or reciprocal dyads $\implies$ 16 directed triads (Triad Census).
- Understanding these micro-motifs is the key to decoding the macro-structures of power, friendship, and organization in society.
:::

---

## References {.unnumbered}
