---
title: "Simmelian Ties"
date: "`r Sys.Date()`"
output:
  bookdown::tufte_html2: default
bibliography: networks.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
```


# Structural Embededness 
`r newthought('Imagine you have a really close friend, but that friend does know or interact with any other of your close friends.')` Now imagine the same close friend who is also very close to another one of your close friends. It seems like we are talking about two very different kinds of relationships. In the first case, the tie between you and your close friend is separate from your other close (or even weak!) ties. In the second case, the tie between you and your close friend fits snuggly with the other ties between your other close friends. Social network analysts use a special word to refer to this property of social ties: **embeddedness**. 

```{r simmelian, fig.margin=TRUE, fig.cap="A graph with three embedded dyads."}
    gr <- create_empty(4, directed = FALSE)
    gr <- gr %>% 
        bind_edges(data.frame(from = 1, to = 2:4)) %>% 
        bind_edges(data.frame(from = 2, to = 3)) %>% 
        mutate(name = toupper(letters[1:4]))
    p <- ggraph(gr, layout = 'auto') 
    ew <- c("y", "y", "n", "y")
    ec <- c("y", "y", "n", "y")
    cols <- c("n" = "steelblue", "y" = "red")
    wdts <- c("n" = 1, "y" = 3)
    p <- p + geom_edge_link(aes(color = ec, edge_width = ew))
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() 
    p <- p + ylim(-0.6,0.6) + xlim(-0.8, 1.3) + theme(legend.position = "none")
    p
```


A tie between you and a close friend who is also close with another one of your close friends is said to be **embedded** in your network. A "standalone" tie between you and someone else, regardless of how close the two of you are, is said to be **disembedded**. Note that when a tie between you and someone else is embedded, we can draw the three relationships as a **closed triad**. An embedded tie is a **connected dyad** embedded within a closed triad.^[If you know the story of Romeo and Juliet, you know that regardless of how much they loved one another, their tie was **disembedded** from their other relationships to their respective families. We know how that ended!] In general, sociologists theorize that embeddedness make strong ties even stronger.  

For instance, in Figure \@ref(fig:simmelian), the dyads $\{AB, AC, BC\}$ (pictured in thick red lines) are embedded within the $\{ABC\}$ triad. The dyad $\{AD\}$, on the other hand, is a disembedded dyad (pictured as a thinner blue line).

# Simmelian Ties
But why does embeddedness make strong ties stronger? The social network theorist [**David Krackhardt**](https://en.wikipedia.org/wiki/David_Krackhardt), drawing on the work of a turn-of-the-century German sociologist named [**Georg Simmel**](https://en.wikipedia.org/wiki/Georg_Simmel). Developed a theory that tells you why. Krackhardt calls strong ties embedded in a triads **Simmelian Ties** (in honor of Georg Simmel, who first theorized the importance of the transition from a two-person relationship to a three-person **clique**). The difference between standalone ties and embedded ties, comes down to three things.

1. If you are tied to someone but that person does not have any ties with anyone else you know, then you have more *individuality* and *independence* within that relationship. If you are tied to someone who is also tied to someone else you know you have less individuality within the three-person clique.
1. If you are tied to someone but that person does not have any ties with anyone else you know, then you have more *bargaining power* within that relationship. If you are tied to someone who is also tied to someone else you know you have less bargaining power, because the other two people can threaten to dissolve the relationship with you and keep being friends with one another. You stay alone, and they still have each other. 
1. If there is conflict within a disembedded tie, it is more difficult to fix it. Ties embedded in triads have a greater capacity for *conflict resolution* because a third person can always *mediate* the any two of the other members of the three-person clique. 

Imagine that you and your disembedded friend (who does not any one of your other friends so you are hanging out by yourselves by definition) are trying to decide where to go for dinner, but are having a hard time. You want Pizza, they want Korean Barbecue. Because each one of you have exactly one vote you are at a standstill. That means that if the whole thing escalates into an argument all you can do is shout at one another and stomp out of the room (remaining hungry or going to dinner by yourself). 

Now imagine the same situation, but it is you and your *two* other close friends. The idea is because now you are trying to decide what to do in the context of your embedded ties, the situation should play you very differently. This illustrates Krackhardt's three principles described above. First, in this situation your personal preferences and opinions count for less, because now you have to coordinate with two other people who could theoretically outvote you (if they both want Pizza your preference for Korean matters less). So the fact that you have *less bargaining power* also reduces your *individuality* in the three-person clique, forcing you toward *group conformity*. 

Now let's say you are stubborn and get into an argument with them, shouting "it's Korean or nothing!" They can just shrug, go out to dinner by themselves, leaving you angry, hungry, and alone in your room. Once again, this demonstrates the idea that you have less power within the three-person clique, because the other people always have their own relationship to fall back to. 

Finally, it could be that one of your other friends who is indifferent between Pizza and Korean helps assuage the conflict. Maybe they point out that they heard about this cool Pizza place that has Korean BBQ slices! This demonstrates the fact that ties embedded in triads have greater potential for conflict resolution via the inter-mediation of third parties.

Because of this, Krackhardt predicts that disembedded ties are more *brittle* than embedded ties. They are prone to conflict and can easily be broken without anyone helping save the relationship. Embedded ties, on the other hand, are **strong** and **sticky**. That means that most long-lasting, close ties we observe in the world should be of the embedded (Simmelian) type, not of the disembedded kind. 