---
title: "Homework VI: Directed Graph Metrics"
date: "`r Sys.Date()`"
output: 
  bookdown::tufte_html2: 
    tufte_variant: "envisioned"
    number_sections: true
    toc: true
    smart: true
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

```{r grex1, fig.cap="A directed graph.", fig.height=10, fig.width=10}
    set.seed(123)
    gr <- play_blocks(10, 10, 0.4, directed = TRUE, loops = FALSE)
    V(gr)$name <- toupper(letters[1:length(V(gr))])
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.05,
                            arrow = arrow(length = unit(4, 'mm')),
                            end_cap = circle(7, 'mm'), 
                            sep = unit(5, 'mm'))
    p <- p + geom_node_point(aes(x = x, y = y), size = 16, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() 
    p
```

This is the **adjacency matrix** of the **directed graph** shown in Figure \@ref(fig:grex1): 

```{r adj}
    a <- matrix(as_adjacency_matrix(gr), nrow = length(V(gr)))
    rownames(a) <- V(gr)$name
    colnames(a) <- V(gr)$name
    diag(a) <- "--"
    a <- a[order(row.names(a)), order(row.names(a))]
    kable(a, format = "pipe", padding = 2, caption = "Adjancency Matrix of a Directed Graph") 
```

(@) Write down the cell entries for the **transpose** ($A'$) of the **adjacency matrix** ($A$) shown in Table \@ref(tab:adj) (2pts):

```{r}
    kable(t(a), format = "pipe", padding = 2) 
```

(@) Write down the cell entries for the matrix corresponding to the **element-wise product** of the **adjacency matrix** ($A$) and the **transpose** of the adjacency matrix ($A'$) corresponding to the graph shown in Figure \@ref(fig:grex1) ($A \circ A'$) (2pts):

```{r}
    a <- matrix(as_adjacency_matrix(gr), nrow = length(V(gr)))
    rownames(a) <- V(gr)$name
    colnames(a) <- V(gr)$name
    c <- a*t(a)
    diag(c) <- "--"
    kable(c, format = "pipe", padding = 2) 
```

Using the information from the answers obtained for questions 1 and 2 answer the following:

(@) What is the number of **mutual dyads** in the graph?


- $M = `r dyad_census(gr)$mut`$


(@) What is the number of **asymmetric dyads** in the graph?


- $M = `r dyad_census(gr)$asym`$


(@) What is the number of **null dyads** in the graph?


- $M = `r dyad_census(gr)$null`$


(@) From the answers to questions 3 and 4, Compute the **Graph Reciprocity**:


- $R = `r round(reciprocity(gr, mode = "ratio"),2)`$


```{r grex2, fig.cap="A directed graph.", fig.height=10, fig.width=10}
    set.seed(123)
    gr <- play_blocks(12, 12, 0.2, directed = TRUE, loops = FALSE)
    V(gr)$name <- toupper(letters[1:length(V(gr))])
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.05,
                            arrow = arrow(length = unit(4, 'mm')),
                            end_cap = circle(7, 'mm'), 
                            sep = unit(5, 'mm'))
    p <- p + geom_node_point(aes(x = x, y = y), size = 16, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() 
    p
```


This is the **reachability matrix** of the graph shown in Figure \@ref(fig:grex2): 

```{r reach}
    r <-distances(gr, mode = "out")
    r <- ifelse(r == Inf, 0, 1)
    rownames(r) <- V(gr)$name
    colnames(r) <- V(gr)$name
    diag(r) <- "--"
    kable(r, format = "pipe", padding = 2, caption = "Reachability Matrix of a Directed Graph") 
```

(@) What is the **Graph Connectedness**?

```{r radj}
    rn <-distances(gr, mode = "out")
    rn <- ifelse(rn == Inf, 0, 1)
    diag(rn) <- 0
    rg <- graph_from_adjacency_matrix(rn)
```

- $C = `r round(edge_density(rg), 2)`$


(@) Write down the cell entries for the **transpose** ($R'$) of the **reachability matrix** ($R$) corresponding to the graph shown in Figure \@ref(fig:grex2) (2pts):

```{r}
    kable(t(r), format = "pipe", padding = 2) 
```


(@) Write down the cell entries for the matrix corresponding to the **element-wise product** of the **reachability matrix** ($R$) and the **transpose** of the reachability matrix ($R'$) corresponding to the graph shown in Figure \@ref(fig:grex2) ($R \circ R'$) (2pts):

```{r}
    c <- rn*t(rn)
    diag(c) <- "--"
    kable(c, format = "pipe", padding = 2) 
```

Using the information from the answers obtained for questions 8 and 9 answer the following:


(@) What is the number of **mutually reachable dyads** in the graph?


- $M = `r dyad_census(rg)$mut`$


(@) What is the number of **asymmetrically reachable dyads** in the graph?


- $A = `r dyad_census(rg)$asym`$


(@) From the answers to questions 10 and 11, Compute the **Graph Hierarchy**:

- $N = `r 1- round(reciprocity(rg, mode = "ratio"), 2)`$




