---
title: "Homework III: Graph Metrics"
date: "`r Sys.Date()`"
output:
  bookdown::tufte_html2: 
    tufte_variant: "envisioned"
    number_sections: true
---

```{r setup, include=FALSE}
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

Consider the graph shown in Figure \@ref(fig:grex1):

```{r grex1, fig.cap="A simple graph.", fig.height=8, fig.width=10}
    set.seed(934)
    d <- c(4, 4, 2, 8, 6, 4, 2, 10, 2, 4, 4, 8, 4, 6)
    gr <- play_degree(d, method = "simple")
    gr <- as_tbl_graph(simplify(gr))
    gr <- mutate(gr, name = toupper(letters[1:length(d)]))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 16, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 7, color = "white")
    p <- p + theme_graph() 
    p
```

(@) What is the **order** of the graph?

- $n =  `r gorder(gr)`$

(@) What is the **size** of the graph?

- $m =  `r gsize(gr)`$

(@) Write down the graph's **degree sequence**:

- `r kbl(t(sort(degree(gr), decreasing  = TRUE)))`

(@) What is the graph's **maximum degree**?

- $k_{max} = `r max(degree(gr))`$

(@) What is the graph's **minimum degree**?

- $k_{min} = `r min(degree(gr))`$

(@) What is the graph's **degree range**?

- $k_r = `r max(degree(gr))` - `r min(degree(gr))` = `r max(degree(gr)) -  min(degree(gr))`$

(@) What is the graph's **average degree**?

- $\bar{k} = `r round(sum(degree(gr)) / length(degree(gr)), 1)`$

(@) What is the graph's **maximum size**?

- $Max(E) = `r (gorder(gr) * (gorder(gr) - 1))/2`$

(@) Compute the **density** of the graph:

- $d = `r round(gsize(gr) / ((gorder(gr) * (gorder(gr) - 1))/2), 2)`$


Consider the graph shown in Figure \@ref(fig:grex2):


```{r grex2, fig.cap="A directed graph.", fig.height=10, fig.width=10}
    set.seed(934)
    gr <- play_erdos_renyi(n = 14, m = 29, directed = TRUE, loops = FALSE)
    gr <- mutate(gr, name = toupper(letters[1:14]))
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.1,
                            arrow = arrow(length = unit(4, 'mm')),
                            end_cap = circle(7, 'mm'), 
                            sep = unit(8, 'mm'))    
    p <- p + geom_node_point(aes(x = x, y = y), size = 16, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 7, color = "white")
    p <- p + theme_graph()
    p
```

(@) What is the graph's **maximum out-degree**?

- `r kbl(t(degree(gr, mode = "out")))`

- $k_{max}^{out} = `r max(degree(gr, mode = "out"))`$

(@) What is the graph's **minimum in-degree**?

- `r kbl(t(degree(gr, mode = "in")))`

- $k_{min}^{in} = `r min(degree(gr, mode = "in"))`$

(@) What is the graph's **maximum in-degree**?

- $k_{max}^{in} = `r max(degree(gr, mode = "in"))`$

(@) What is the graph's **minimum out-degree**?

- $k_{min}^{out} = `r min(degree(gr, mode = "out"))`$

(@) What is the graph's **maximum size**?

- $Max(E) = `r gorder(gr) * (gorder(gr) - 1)`$

(@) Compute the **density** of the graph:

- $d = `r round(gsize(gr) / (gorder(gr) * (gorder(gr) - 1)), 2)`$



