---
title: "Homework I: Graph Theory"
date: "`r Sys.Date()`"
output:
  bookdown::tufte_html2: 
    tufte_variant: "envisioned"
    number_sections: true
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
```

# Vertex and edge sets

```{r grex1, fig.cap="A simple graph.", fig.height=4}
    set.seed(529)
    gr <- play_erdos_renyi(n = 5, m = 5, directed = FALSE, loops = FALSE)
    gr <- mutate(gr, name = toupper(letters[1:5]))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() + ylim(-1, 1)
    p
```
Consider the graph shown in Figure \@ref(fig:grex1):

(@) Write down the **vertex set** of the graph:

- $V = \{A, B, C, D, E\}$

or

- $V = \{V_A, V_B, V_C, V_D, V_E\}$ 

- (Same alternative for the answers below that involving listing node subsets)



(@) Write down the **edge set** of the graph:

- $E = \{AB, AC, BC, BE, CD\}$ 

or

- $V = \{V_AV_B, V_AV_C, V_BV_C, V_BV_E, V_CV_D\}$ 

- With the understanding that the order of listing node labels in the edge doesn't matter in the undirected case: $AB = BA$ and $V_AV_B = V_BV_A$.

# Node Neighborhoods

```{r grex2, fig.cap="A simple graph.", fig.width=8, fig.height=6.75}
    set.seed(471)
    gr <- play_erdos_renyi(n = 8, m = 13, directed = FALSE, loops = FALSE)
    gr <- mutate(gr, name = toupper(letters[1:8]))
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() 
    p
    
```
Consider the graph shown in Figure \@ref(fig:grex2):

(@) Write down the **neighborhood** of node *D*

- $N(D) = \{A, B, C, H\}$

(@) Write down the **neighborhood** of node *E*

- $N(E) = \{A, C, G, H\}$

(@) What is the **intersection** of the neighborhoods of nodes *D* and *E*?

- $N(D) \cap N(E) = \{A, C, H\}$

(@) What is the **intersection** of the neighborhoods of nodes *C* and *G*?

- $N(C) \cap N(G) = \{E\}$

(@) What is the **union** of the neighborhoods of nodes *B* and *C*? 

- $N(B) \cup N(C) = \{D, E, F, G\}$

# Node Degree

Consider the graph shown in Figure \@ref(fig:grex2):

(@) What is the **degree** of node *B*?

- $k_B = 3$

(@) What are the nodes with the largest **degree**?

- $\{D, E, G\}$

(@) What is the node with the smallest **degree**?

- $\{F\}$