---
title: "Homework VII: Two Mode Networks"
date: "`r Sys.Date()`"
output: 
  bookdown::tufte_html2: 
    tufte_variant: "envisioned"
    number_sections: true
    toc: true
    smart: true
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

```{r}
    A <- matrix(c(1, 1, 1, 1, 1, 1, 0, 1, 1, 0, 0, 0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 1, 0, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 1, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 0, 1, 1, 1, 0, 0, 0, 0, 0, 1, 1, 0, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 0, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 0, 0, 0), ncol = 14)
w <- c("EVELYN", "LAURA", "THERESA", "BRENDA", "CHARLOTTE", "FRANCES", "ELEANOR", "PEARL", "RUTH", "VERNE", "MYRNA", "KATHERINE", "SYLVIA", "NORA", "HELEN", "DOROTHY", "OLIVIA", "FLORA")
e <- paste("E", c(1:14), sep = "")
rownames(A) <- w
colnames(A) <- e
    affil.el <- data.frame(matrix(ncol = 2, nrow = 0))
    k <- 1
    for (i in 1:nrow(A)) {
        for(j in 1:ncol(A)) {
            if (A[i, j] == 1) {
              affil.el[k, 1] = w[i]
              affil.el[k, 2] = e[j]
            k <- k + 1
            }
        }
    }
```

```{r bipart, fig.width=20, fig.height=8, fig.fullwidth=TRUE, caption = "Southern women bipartite graph."}
    gr <- graph_from_data_frame(data.frame(affil.el), directed = FALSE)
    V(gr)$type <- bipartite_mapping(gr)$type
    p <- ggraph(gr, layout = 'bipartite')  
    p <- p + geom_edge_link(color = "steelblue", edge_width = 1.05) 
    p <- p + geom_node_point(color = c(rep("white", 18), rep("tan2", 14)), 
                             shape = c(rep("square", 18), rep("circle", 14)), 
                             size = c(rep(15, 18), rep(20, 14)))
    p <- p + geom_node_text(aes(label = name), 
                            size = c(rep(4.5, 18), rep(5, 14)), 
                            color = "black")   
    p <- p + theme_graph() + theme(legend.position = "none") 
    p
```

The Figure above shows the **bipartite graph** corresponding to the *Southern Women Events Data* a classic social network data set collected in the 1930s by the sociologist Allison Davis and collaborators and included in book [*Deep South*](https://www.amazon.com/Deep-South-Anthropological-Southern-Classics/dp/1570038155). The graph represents observed attendance at 14 social events by 18 Southern "high society" women from the South. 

(@) Write down the **affiliation matrix** corresponding to the Southern women network bipartite graph (2pts).

```{r}
    kbl(A, format = "html", padding = 2) %>% 
        kable_classic(lightable_options = "striped")
```

```{r}
    kbl(t(rowSums(A)), format = "html", padding = 2) %>% 
        kable_classic(lightable_options = "striped")
```

```{r}
    kbl(t(colSums(A)), format = "html", padding = 2) %>% 
        kable_classic(lightable_options = "striped")
```

(@) Which women attend the most events?

- Evelyn

(@) Which women attend the least events?

- Verne

(@) What is the average number of events attended by the women?

- $\bar{C}_{women} = `r round(mean(rowSums(A)), 1)`$

(@) Which are the largest event(s)?

- E1 and E2

(@) Which are the smallest event(s)?

- E14

(@) What is the average event size?

- $\bar{C}_{events} = `r round(mean(colSums(A)), 1)`$

(@) Write down the *transpose* of the **affiliation matrix** corresponding to the Southern women network bipartite graph (2pts).

```{r}
    kbl(t(A), format = "html", padding = 2) %>% 
        kable_classic(lightable_options = "striped")
```

Using the information in 1 and 2 and what you know about **matrix multiplication** (5pts):

(@) Write down the entries corresponding to the **co-membership matrix** of the Southern women.

```{r}
    c.m <- A %*% t(A)
    kbl(c.m, format = "html", padding = 2) %>% 
        kable_classic(lightable_options = "striped")
```

(@) Write down the entries corresponding to the **group overlap matrix** of the Southern women.

```{r}
    g.o <- t(A) %*% A
    kbl(g.o, format = "html", padding = 2) %>% 
        kable_classic(lightable_options = "striped")
```