---
title: "Homework II: Directed Graphs"
date: "`r Sys.Date()`"
output:
  bookdown::tufte_html2: 
    tufte_variant: "envisioned"
    number_sections: true
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
```


```{r grex1, fig.cap="A directed graph.", fig.width=10, fig.height=10}
    set.seed(529)
    gr <- play_erdos_renyi(n = 10, m = 20, directed = TRUE, loops = FALSE)
    gr <- mutate(gr, name = toupper(letters[1:10]))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(5, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))    
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph()
    p
```
Consider the graph shown in Figure \@ref(fig:grex1):

(@) Write down the **edge set** of the graph:

- $E = \{AB, AD, AE, BA, BE, CH, EA, EC, ED, EH, FC, FE, FJ, FI, GC, GI, HD, IC, JD, JH\}$

- (Also accept $V_X$ notation for all that follows) 


(@) Who are the **out-neighbors** of node *F*?

- $N_{out}(F) = \{C, E, I, J\}$


(@) Who are the **in-neighbors** of node *D*?

- $N_{in}(D) = \{A, E, H, J\}$

(@) What is the **intersection** of the **out-neighborhoods** of vertices *E* and *A*?

- $N_{out}(A) = \{B, D, E\}$

- $N_{out}(E) = \{A, C, D, H\}$

- $N_{out}(A) \cap N_{out}(E) = \{D\}$


(@) What is the **union** of the **in-neighborhoods** of vertices *D* and *H*?

- $N_{in}(D) = \{A, E, H, J\}$

- $N_{in}(H) = \{C, E, J\}$

- $N_{in}(D) \cup N_{in}(H) = \{A, C, E, H, J\}$


(@) What is the **out-degree** of node *F*?

- $k_F^{out} = 4$


(@) What is the **in-degree** of node *D*?


- $k_F^{in} = 4$


(@) What are the nodes with the largest **out-degree**?


- $\{E, F\}$


(@) What are the nodes with the largest **in-degree**?


- $\{C, D\}$


(@) Which one is the **receiver node** in the graph?


- $\{D\}$


(@) Which are the **transmitter nodes** in the graph?


- $\{F, G\}$


(@) What type of **dyad** is formed by the subgraph containing nodes *A* and *B*?


- *Mutual*


(@) What type of **dyad** is formed by the subgraph containing nodes *C* and *F*?


- *Asymmetric*


(@) What type of **dyad** is formed by the subgraph containing nodes *B* and *C*?


- *Null*


(@) How many **mutual dyads** are there in the graph?

- $M = 2$

