---
title: "Homework II: Directed Graphs"
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
```

```{r grex1, fig.cap="A directed graph.", fig.width=10, fig.height=10}
#| label: fig-grex1
#| fig-cap: "A directed graph."
#| fig-cap-location: margin   
#| fig-height: 8
#| fig-width: 8
    set.seed(539)
    gr <- play_gnm(n = 10, m = 20, directed = TRUE, loops = FALSE)
    gr <- mutate(gr, name = toupper(letters[1:10]))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(5, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))    
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph()
    p
```
Consider the graph shown in Figure \@ref(fig:grex1):

1. Write down the **edge set** of the graph:
<br>
<br>
<br>

1. Who are the **out-neighbors** of node *F*?
<br>
<br>
<br>

1. Who are the **in-neighbors** of node *D*?
<br>
<br>
<br>

1. What is the **intersection** of the **out-neighborhoods** of vertices *E* and *A*?
<br>
<br>
<br>

1. What is the **union** of the **in-neighborhoods** of vertices *D* and *H*?
<br>
<br>
<br>

1. What is the **out-degree** of node *F*?
<br>
<br>
<br>

1. What is the **in-degree** of node *D*?
<br>
<br>
<br>

1. What are the nodes with the largest **out-degree**?
<br>
<br>
<br>

1. What are the nodes with the largest **in-degree**?
<br>
<br>
<br>

1. Which are the **receiver nodes** in the graph?
<br>
<br>
<br>

1. Which are the **transmitter nodes** in the graph?
<br>
<br>
<br>

1. Write down the set of **mutual dyads** in the graph:
<br>
<br>
<br>

1. What type of **dyad** is formed by the subgraph containing nodes *C* and *F*?
<br>
<br>
<br>

1. What type of **dyad** is formed by the subgraph containing nodes *B* and *C*?
<br>
<br>
<br>

1. How many **mutual dyads** are there in the graph?
<br>
<br>
<br>


