---
title: "Homework III: "
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
```

Consider the graph shown in Figure \@ref(fig:grex1):

```{r grex1, fig.cap="A simple graph.", fig.height=8, fig.width=10}
    set.seed(964)
    d <- c(5, 4, 3, 7, 7, 5, 1, 9, 3, 5, 4, 6, 2, 2)
    gr <- play_degree(d, method = "simple")
    gr <- mutate(gr, name = toupper(letters[1:length(d)]))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 16, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 7, color = "white")
    p <- p + theme_graph() 
    p
```

(@) What is the **order** of the graph?

(@) What is the **size** of the graph?

(@) Write down the graph's **degree sequence**:

(@) What is the graph's **maximum degree**?

(@) What is the graph's **minimum degree**?

(@) What is the graph's **degree range**?

(@) What is the graph's **average degree**?

(@) What is the graph's **maximum size**?

(@) Compute the **density** of the graph:



Consider the graph shown in Figure \@ref(fig:grex2):

```{r grex2, fig.cap="A directed graph.", fig.height=10, fig.width=10}
    set.seed(998)
    gr <- play_erdos_renyi(n = 14, m = 29, directed = TRUE, loops = FALSE)
    gr <- mutate(gr, name = toupper(letters[1:14]))
    p <- ggraph(gr, layout = 'dh') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.1,
                            arrow = arrow(length = unit(4, 'mm')),
                            end_cap = circle(7, 'mm'), 
                            sep = unit(8, 'mm'))    
    p <- p + geom_node_point(aes(x = x, y = y), size = 16, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 7, color = "white")
    p <- p + theme_graph()
    p
```

(@) What is the graph's **maximum out-degree**?

(@) What is the graph's **minimum in-degree**?

(@) What is the graph's **maximum in-degree**?

(@) What is the graph's **minimum out-degree**?

(@) What is the graph's **maximum size**?

(@) Compute the **density** of the graph:



