---
title: "Homework V: Indirect Connections and Matrices"
date: "`r Sys.Date()`"
output:
    bookdown::tufte_html2: 
    tufte_variant: "envisioned"
number_sections: true
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
```

Consider the graph shown in Figure \@ref(fig:grex1):

```{r grex1, fig.cap="An undirected graph.", fig.height=10, fig.width=10}
    set.seed(237)
    d <- c(sample(c(2, 4, 6, 8), 9, replace = TRUE))
    gr <- play_degree(d, method = "simple")
    gr <- mutate(gr, name = toupper(letters[1:length(d)]))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() 
    p
```

(@) Write down the **path(s)** connecting node *A* and node *G* (3pts):
<br>
<br>
<br>
<br>
<br>
<br>
<br>

(@) What is the length of the **shortest path(s)** connecting nodes *A* and *G*?

- 

(@) What is the length of the longest **path** connecting nodes *A* and *G*?

- 

Consider the graph shown in Figure \@ref(fig:grex2):

```{r grex2, fig.cap="A directed graph.", fig.height=10, fig.width=10}
    set.seed(123)
    fr <- sample(toupper(letters[1:9]), 24, replace = TRUE)
    to <- sample(toupper(letters[1:9]), 24, replace = TRUE)
    el <- matrix(c(fr, to), nc = 2, byrow = TRUE)
    gr <- as_tbl_graph(simplify(graph_from_edgelist(el)))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_parallel(color = "steelblue", edge_width = 1.25,
                            arrow = arrow(length = unit(7, 'mm')),
                            end_cap = circle(10, 'mm'), 
                            sep = unit(8, 'mm'))
    p <- p + geom_node_point(aes(x = x, y = y), size = 22, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 10, color = "white")
    p <- p + theme_graph() 
    p
```

(@) Write down the **directed path(s)** going *from* node *E* *to* node *B* (2pts):

<br>
<br>
<br>
<br>
<br>
<br>

(@) What is the length of the **shortest directed path(s)** connecting nodes *E* and *B*?

- 

(@) What is the length of the longest **directed path** connecting nodes *E* and *B*?

- 

(@) Write down the cell entries for the **adjacency matrix** corresponding to the graph shown in Figure \@ref(fig:grex1) (3pts):

```{r}
    a <- matrix(" ", nrow = 9, ncol = 9)
    rownames(a) <- toupper(letters[1:9])
    colnames(a) <- toupper(letters[1:9])
    diag(a) <- "--"
    kable(a, format = "pipe", padding = 2) 
```

(@) Write down the cell entries **reachability matrix** corresponding to the graph shown in Figure \@ref(fig:grex2) (3pts):

```{r}
    a <- matrix(" ", nrow = 9, ncol = 9)
    rownames(a) <- toupper(letters[1:9])
    colnames(a) <- toupper(letters[1:9])
    diag(a) <- "--"
    kable(a, format = "pipe", padding = 2) 
```
