---
title: "Types of Graphs"
date: "`r Sys.Date()`"
output: 
  bookdown::tufte_html2: 
    tufte_variant: "envisioned"
    number_sections: true
    toc: true
    smart: true
bibliography: networks.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
    require(patchwork)
```

# Practice Problems

## Balance
1.Does the following graphs represent balanced or unbalanced relations between actors? (Dotted lines are negative ties)
[picture]
1.Why is balance important sociologically? What is the name for the theory that thinks balance is something people do?
1.Name the personality trait from the Five Factor Model that describes the following situations.
	i.Someone does not want to try a food because they haven't had it before.
	i.Someone gets a bad grade, they get a little sad but generally have a good day.
	i.Someone's two friends are fighting, that someone does nothing.
	i.Someone really likes going to parties and meeting new people.
	i.Someone is totally ok trying a new food on the recommendation of the waiter.
	i.You invite a friend over for food, they bring someone else with them without asking you.
	i.Someone gets a bad grade on a homework. It ruins their entire weekend.
	i.You order appetizers for the table. You pay for them yourself because you didn't ask other people if they wanted them or not (2 answers!)


###
```{r directed_exercise_5_1, echo = FALSE, warning = FALSE, message=FALSE, fig.height=8}
from <- c("A", "A", "A", "A", "B", "B", "C", "C", "D", "D", "E", "E", "I", "F", "J", "H", "G", "K", "K", "L", "L", "L", "L", "M", "P", "M", "N", "N", "O")
to <- c("B", "D", "I", "E", "E", "I", "M", "I", "G", "F", "J", "J", "F", "I", "G", "G", "M", "M", "L", "M" , "P", "N", "O", "N", "M", "O", "P", "O", "P")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 14, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from,to))), size = 8)
p <- p + theme_graph()
p
```

a) The graph is k-connected with what k?
b) What is the diameter of the graph?
c) What is a name for node H? What does it do?
d) Let us remove node H. Notice that we have essentially done some c by removing the fewest node/edges to create network components. What are the resulting n-cliques for each of the resulting components?
e) Calculate the Connectedness score for each of the two components. Which is the more centralized component?

###
```{r directed_exercise_5_2, echo = FALSE, warning = FALSE, message=FALSE, fig.height=8}
from <- c("A", "A", "B", "D", "F", "D", "C", "C", "E", "H", "G", "K", "I", "L", "J")
to <- c("B", "D", "C", "C", "D", "E", "F", "E", "H", "F", "H", "G", "G", "K", "I")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 14, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from,to))), size = 8)
p <- p + theme_graph()
p
```

The Ministry of Silly Walks: Both J and L would like to get approval for funding the development of their new idea, which can only be approved by person A. The bureaucracy to A though is a little convoluted. What is the shortest path for person J and L to get their idea passed up the chain of command for approval?


###

```{r directed_exercise_5_3, fig.margin=FALSE, echo = FALSE, warning = FALSE, message=FALSE, fig.height=8}
from <- c("A", "A", "A", "E", "E", "I", "I", "J", "A", "B", "C", "K", "D", "L", "L", "L", "L", "L", "L", "M", "M", "M", "N", "N")
to <- c("B", "E", "I", "J", "F", "F", "J", "F", "C", "C", "D", "D", "L", "K", "M", "N", "O", "P", "N", "P", "O", "P", "O", "P")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 14, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from,to))), size = 8)
p <- p + theme_graph()
p
```

a) How many components are there in the graph?
b) The giant component is k-connected with what k?
c) What is the diameter of the graph?

Remove component G-H so now the graph is what was before the giant component and answer the following questions.
d) What is the diameter of the graph? 
e) The graph is k-connected with what k?
f) What type of clique is subgraph (L, M, N, O, P)?
g) What type of clique is subgraph (A, E, I, F, J)?



```{r directed-practice1, fig.cap="Example 1", echo=FALSE, warning=FALSE, message=FALSE}
    g.nodes <- data.frame(name = c("A", "B", "C", "D", "E", "F", "G"))
    g.edges <- data.frame(from = c(rep("A", 2), rep("B", 4), rep("C", 2), rep("D", 3), rep("E", 3),
                                   rep("F", 1), rep("G", 3)),
                          to = c("B", "C", "A", "C", "D", "G", "A", "B", "E", "G", "B", "D", "G",
                                 "F", "E", "B", "D", "E"))
    gr <- tbl_graph(nodes = g.nodes, edges = g.edges)
    
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "lightblue") 
    p <- p + geom_node_text(aes(label = name), size = 8)
    p <- p + theme_graph() + ylim(-1.25,1.25)
    p
```


1. What type of graph is pictured above? What type of relationships might the graph represent?
1. How many nodes and edges are in the above graph?
1. What is the set that describes the above graph?
1. What is the average in-degree of the above graph? The average out-degree?
1. What is modal degree in the above graph?
1. What is the degree distribution of the above graph?
1. What is the density of the above graph?


```{r directed-practice2, fig.cap="Example 2", echo=FALSE, warning=FALSE, message=FALSE}
    g.nodes <- data.frame(name = c("L", "M", "N", "O", "P", "Q", "R"))
    g.edges <- data.frame(from = c(rep("L", 3), rep("M", 3), rep("N", 2), rep("O", 3), rep("P", 3),
                                   rep("Q", 1), rep("R", 1)),
                          to = c("M", "N", "O", "Q", "P", "L", "L", "O", "L", "N", "R", "M", "Q",
                                 "M", "P", "O"))
    gr <- tbl_graph(nodes = g.nodes, edges = g.edges)
    p <- ggraph(gr, layout = 'auto')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5) 
p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "lightblue") 
p <- p + geom_node_text(aes(label = name), size = 8)
p <- p + theme_graph() + ylim(-1.25,1.25)
p
```

1. What type of graph is pictured above? What type of relationships might the graph represent?
1. How many nodes and edges are in the above graph?
1. What is the set that describes the above graph?
1. What is the average in-degree of the above graph? The average out-degree?
1. What is modal degree in the above graph?
1. What is the degree distribution of the above graph?
1. What is the density of the above graph?


```{r directed-practice3, fig.cap="Example 3", echo=FALSE, warning=FALSE, message=FALSE}
    g.nodes <- data.frame(name = c("A", "B", "C", "D", "E", "F", "G"))
    g.edges <-  data.frame(from = c(rep("A", 2), rep("B", 3), rep("C", 2), rep("D", 1), rep("E", 1),
                                   rep("F", 1), rep("G", 2)),
                          to = c("D", "C", "A", "F", "D", "E", "B", "B", "D", "A", "F", "A"))
    gr <- tbl_graph(nodes = g.nodes, edges = g.edges)
    p <- ggraph(gr, layout = 'auto')
    p <- p + geom_edge_parallel(color = "darkgray", edge_width = 1.5,
                        arrow = arrow(length = unit(6, 'mm')),
                        end_cap = circle(7, 'mm'), 
                        sep = unit(6, 'mm'))
    p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "lightblue") 
    p <- p + geom_node_text(aes(label = name), size = 8)
    p <- p + theme_graph() + ylim(-1, 1)
    p
```

1. What type of graph is pictured above? What type of relationships might the graph represent?
1. How many nodes and edges are in the above graph?
1. What is the set that describes the above graph?
1. What is the average in-degree of the above graph? The average out-degree?
1. What is modal degree in the above graph?
1. What is the degree distribution of the above graph?
1. What is the density of the above graph?





```{r directed_practice5, fig.cap="Example 5", echo=FALSE, warning=FALSE, message=FALSE}
from <- c("Hassan", "Farid", "Haifa", "Haifa", "Hisham", "Latifa", "Tamer", "Ahmed", "Nora", "Haifa", "Ahmed")
to <- c("Haifa", "Haifa", "Nora", "Hisham", "Latifa", "Nora", "Hisham", "Latifa", "Hisham", "Latifa", "Nora")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'star')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 5)
p <- p + theme_graph()
p <- p + expand_limits(x = c(-1.5,1.5), y = c(-1.5, 1.5))
p
```


```{r directed_practice6, fig.cap="Example 6", echo=FALSE, warning=FALSE, message=FALSE}
from<- c("Vishal", "Prakriti", "Prakriti", "Arjun", "Neha", "Karan", "Navv", "Navv", "Raja", "Sukriti", "Neja", "Raja")
to <- c("Arjun", "Arjun", "Neha", "Neha", "Arjun", "Neha", "Karan", "Raja", "Sukriti", "Neha", "Sukriti", "Karan")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'star')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 5)
p <- p + theme_graph() + theme(plot.margin = unit(c(0,0,0,0), "cm"))
p <- p + expand_limits(x = c(-1.5,1.5), y = c(-1.5, 1.5))
p
```


```{r undirected_practice8, , fig.cap="Example 8", echo=FALSE, warning=FALSE, message=FALSE}
from <- c("Yeeun", "Yeeun", "Sunye", "Sunmi", "Sohee", "Hyun Seung", "Hyun-a", "Hyun-a", "Hyun-a", "Hyun-a", "Hyun-a", "Hyun-a", "Hyun-a", "Hyun-a", "Ji-hyun", "Ji-hyun", "Ji-hyun","Ga-yoon", "Ji-yoon", "So-hyun", "Ji-yoon") 
to <- c("Sunmi", "Sohee", "Yeeun", "Sunye", "Sunmi", "Hyun-a", "Yeeun", "Sunye", "Sunmi", "Sohee", "Ji-hyun", "Ga-yoon", "Ji-yoon", "So-hyun", "Ga-yoon", "Ji-yoon", "So-hyun", "Ji-yoon", "So-hyun", "Ji-yoon", "So-hyun")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'star')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 24, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 5)
p <- p + theme_graph() + theme(plot.margin = unit(c(0,0,0,0), "cm"))
p <- p + expand_limits(x = c(-1.5,1.5), y = c(-1.5, 1.5))
p
```

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```


## Practice Problems

### 

```{r directed_practice3_1, fig.margin=FALSE, fig.cap="A directed graph", echo = FALSE, warning = FALSE, message=FALSE, fig.height=8}
from <- c("A", "B", "B", "D", "I", "G", "F", "J")
to <- c("B", "I", "D", "G", "G", "E", "F", "J")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 20, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 12)
p <- p + theme_graph()
p
```

a)  Create a reachability matrix describing the above graph.

### 

```{r directed_practice3_2, fig.margin=FALSE, fig.cap="An undirected graph", echo = FALSE, warning = FALSE, message=FALSE, fig.height=8}
from <- c("A", "F", "A", "B", "D", "B", "I", "G", "B", "D")
to <- c("F", "A", "B", "I", "B", "D", "J", "J", "G", "E")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 20, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 12)
p <- p + theme_graph()
p
```

### 

```{r directed_practice3_3, fig.margin=FALSE, fig.cap="A directed graph", echo = FALSE, warning = FALSE, message=FALSE, fig.height=6}
from <- c("Diane", "Lynne", "Lynne", "Lynne", "Lynne", "Lisa", "Lisa", "Kevin")
to <- c("Lynne", "Kevin", "Jacob", "Lisa", "Mike", "Mike", "Jacob", "Jacob")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 30, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 8)
p <- p + theme_graph() + expand_limits(x = c(-2,1.5), y = c(-2.2, .5))
p
```

Above we have a made-up network consisting of academics who have worked together on projects that was presented in the exercises of Chapter 1. Please conduct the following exercises using data from the above graph.

a)  Please turn the above graph into an undirected matrix
b)  Is the above matrix an asymmetric, symmetric, or antisymmetric matrix?
c)  What are examples of non-adjacent ties in the graph?

### 

Please solve the following matrix multiplication problems, if possible. $$
\begin{equation}
\begin{bmatrix} 8 & 9 & 3\\ 3 & 2 & 2\\ \end{bmatrix}
\begin{bmatrix} 5 & 4\\ 1 & 2\\ 3 & 2\\ \end{bmatrix}
=
\end{equation}
$$

$$
\begin{equation}
\begin{bmatrix} 2 & 1 & 4\\ \end{bmatrix}
\begin{bmatrix} 3\\ 3\\ 0\\ \end{bmatrix}
=
\end{equation}
$$

$$
\begin{equation}
\begin{bmatrix} 8\\ 2\\ \end{bmatrix}
\begin{bmatrix} 8 & 5\\ 7 & 6\\ 9 & 3\\ \end{bmatrix}
=
\end{equation}
$$

$$
\begin{equation}
\begin{bmatrix} 2 & 5\\ 3 & 5\\ 2 & 5\\ \end{bmatrix}
\begin{bmatrix} 2\\ 3\\ \end{bmatrix}
=
\end{equation}
$$

$$
\begin{equation}
\begin{bmatrix} 5 & 9 & 1\\ 4 & 5 & 2\\ \end{bmatrix}
\begin{bmatrix} 7 & 4\\ 1 & 2\\ 3 & 6 \end{bmatrix}
=
\end{equation}
$$

$$
\begin{equation}
\begin{bmatrix} 4 & 8 & 2\\ \end{bmatrix}
\begin{bmatrix} 3\\ 7\\ 5\\ \end{bmatrix}
=
\end{equation}
$$

$$
\begin{equation}
\begin{bmatrix} 4\\ 5\\ \end{bmatrix}
\begin{bmatrix} 4 & 8\\ 5 & 8\\ 9 & 7 \end{bmatrix}
=
\end{equation}
$$

$$
\begin{equation}
\begin{bmatrix} 4 & 8\\ 5 & 8\\ 9 & 7 \end{bmatrix}
\begin{bmatrix} 4\\ 5\\ \end{bmatrix}
=
\end{equation}
$$

### 

```{r directed_practice3_5, fig.margin=FALSE, fig.cap="An undirected graph", echo = FALSE, warning = FALSE, message=FALSE, fig.height=5}
from <- c("A", "B")
to <- c("B", "C")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 20, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from, to))), size = 12)
p <- p + theme_graph() + expand_limits(x = c(-.75,0.25), y = c(-0.75, .2))
p
```

a)  Multiply the adjacency matrix for the above graph by itself. What is the resulting matrix and interpret the results




