---
title: "Subgroups"
date: "`r Sys.Date()`"
output:
  bookdown::tufte_html2: default
  toc: true
  smart: true
#bibliography: networks.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
    # invalidate cache when the package version changes
    knitr::opts_chunk$set(cache.extra = packageVersion('tufte'), echo = FALSE, message=FALSE, warning=FALSE)
    options(htmltools.dir.version = FALSE)
    require(ggraph)
    require(tidygraph)
    require(igraph)
    require(kableExtra)
    require(patchwork)
```

# Subgroups
While we've so far examined the principles behind building graphs and matrices to represent network structure, we have yet to touch on how looking at smaller clusters of associations in the network can inform our understanding of the social world. This chapter thus explores how we can think of social activity occuring at the **group** level and not just within the network as a whole. The key graph theoretic concept we will deal with in this chapter is that of **subgraphs**. In real world networks, subgraphs are used to represent such phenomena as **groups**, **social circles**, and **communities**. 

The group level in social networks can be thought of as located between the "micro" level corresponding to individuals and their dyadic and triadic interactions and the "macro" level of the entire (whole) network. Why study groups? Social science research teaches us that groups are a fundamental, for some *the most fundamental*, phenomenon in social life. If you think about it, a lot of your every day interactions takes place in groups. Without groups, you'd be in trouble. Groups provide you with a lot of benefits, including companionship and support. People also get done most of their work in groups, which means they are useful for accomplishing collective goals. 

The rise of telecommunications technology as a key conduit allowing for connectivity without face-to-face interaction has not altered the importance of offline groups as the primary nexus of social life. In fact, new technologically-mediated interactions such as texting, online chatting, and social media complement rather than replace face-to-face interaction in small groups. Even today, the great majority of online ties ultimately relies, on or emerge from, offline face-to-face interactions

Our examination of subgraphs starts by presenting some fundamental concepts from graph theory about we can think about connectivity within networks. Second, we present ideas about how network analysts think about how to identify groups out of relational data. While groups are an intuitive part of how we think about the social world, actually defining groups is an incredibly difficult problem. Finally, we present techniques commonly used to break apart larger networks into constituent groups.


```{r, fig.margin=TRUE, fig.width = 7, fig.height=7}
    gr <- create_notable(name = "Krackhardt kite") %>% 
        mutate(name = toupper(letters[1:10]))
    p <- ggraph(gr, layout = 'auto') 
    p <- p + geom_edge_link(color = "steelblue", width = 1.25) 
    p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph()
    p
```

## Core and Periphery
```{r coreper, fig.margin=TRUE, fig.cap="A undirected graph with core-periphery structure", fig.width = 7, fig.height=7}
    gr <- create_empty(10, directed = FALSE)
    gr <- gr %>% 
        bind_edges(data.frame(from = 1, to = c(2, 3, 4, 5))) %>% 
        bind_edges(data.frame(from = 2, to = c(3, 4, 6, 7, 8))) %>% 
        bind_edges(data.frame(from = 3, to = c(4, 8, 9))) %>% 
        bind_edges(data.frame(from = 4, to = c(5, 10))) %>% 
        mutate(name = toupper(letters[1:10]))
    ewc <- c(rep("y", 3), "n", rep("y", 2), rep("n", 3), "y", "n", "n", "n", "n")
    cols <- c("n" = "steelblue", "y" = "red")
    wdts <- c("n" = 1.25, "y" = 2)
    p <- ggraph(gr, layout = 'kk') 
    p <- p + geom_edge_link(aes(color = ewc, edge_width = ewc))
    p <- p + scale_discrete_manual(values = cols, aesthetics = "edge_colour")
    p <- p + scale_discrete_manual(values = wdts, aesthetics = "edge_width")
    p <- p + geom_node_point(aes(x = x, y = y), size = 18, color = "tan2") 
    p <- p + geom_node_text(aes(label = name), size = 8, color = "white")
    p <- p + theme_graph() + theme(legend.position = "none")
    #p <- p + + xlim(-1,1) + ylim(-2,2.1)
    p
    kbl(as.matrix(gr[1:10]))
```

# Groups


## Cliques

###
```{clique, fig.margin=TRUE, fig.cap="A clique", echo = FALSE, warning = FALSE, message=FALSE, fig.height=8}
from <- c("A", "A", "A", "B", "B", "B", "C", "C", "C", "D", "D", "D")
to <- c("B", "C", "D", "A", "C", "D", "A", "B", "D", "A", "B", "C")
dat <- data.frame(from, to)
gr <- as_tbl_graph(dat)
p <- ggraph(gr, layout = 'kk')
p <- p + geom_edge_link(color = "darkgray", edge_width = 1.5)
p <- p + geom_node_point(aes(x = x, y = y), size = 14, color = "lightblue")
p <- p + geom_node_text(aes(label = unique(c(from,to))), size = 8)
p <- p + theme_graph()
p
```

K-Cliques
K-connectedness
Menger’s Theorem

## Group Detection
Connectedness
Cut-sets and Subgroups
Cohesive Blocking
Hierarchy
Block Modeling
Norfolk Gentry-regular equivalence/blocking (ch 6)
Bearman 1993 Famous Networks
Famous Networks
Gould Whiskey (ch 6)